### 2015-02-18 phaseDelay.tex
files

\documentclass{sig-alternative}
%\usepackage{multirow}
%\usepackage{color}
%% \usepackage{colortbl}
%% \usepackage{picture}
%% %\%usepackage{algorithm}
%% %\usepackage{algorithmicx}
%% %\usepackage{algpseudocode}
%% %\renewcommand{\algorithmicrequire}{\textbf{Input:}}
%% %\renewcommand{\algorithmicensure}{\textbf{Output:}}
%% \newcommand{\quart}[3]{\begin{picture}(100,6)%1
%% {\color{black}\put(#3,3){\circle*{4}}\put(#1,3){\line(1,0){#2}}}\end{picture}}

%% %% timm tricks
%% \newcommand{\bi}{\begin{itemize}[leftmargin=0.4cm]}
%% \newcommand{\ei}{\end{itemize}}
%% \newcommand{\be}{\begin{enumerate}}
%% \newcommand{\ee}{\end{enumerate}}
%% \newcommand{\tion}[1]{\S\ref{sect:#1}}
%% \newcommand{\fig}[1]{Figure~\ref{fig:#1}}
%% \newcommand{\eq}[1]{Equation~\ref{eq:#1}}

%% %% space saving measures

%% \usepackage[shortlabels]{enumitem} 
%% %\usepackage{times}

%% \usepackage{url}
%% \def\baselinestretch{1}

%% \setlist{nosep}
%%  \usepackage[font={small}]{caption, subfig}
%% \setlength{\abovecaptionskip}{1ex}
%%  \setlength{\belowcaptionskip}{1ex}

%%  \setlength{\floatsep}{1ex}
%%  \setlength{\textfloatsep}{1ex}
%% %\usepackage[compact,small]{titlesec}
%% \DeclareMathSizes{7}{7}{7}{7} 
%% \pagenumbering{arabic}
%% \setlength{\columnsep}{7mm}

\begin{document}

\conferenceinfo{FSE}{'15 Bergamo, Italy}
\title{}
\numberofauthors{2}
\author{
\alignauthor
Tim Menzies, Carter Pape\\
       \affaddr{CS, NcState, USA}\\
       tim.menzies@gmail.com,\\carterpape@gmail.com
\alignauthor
Lucas Layman\\
       \affaddr{Fraunhofer Center, Exp. SE, USA}\\ 
       llayman@fc-md.umd.edu
\alignauthor
William Curtis, Forrest Schull\\
\affaddr{SEI, CMU, USA}
wrn,fjshull@sei.cmu.edu
}

\maketitle
\begin{abstract}

%RQ1: Does tuning affect learners' performance?
%RQ2: How to choose

\end{abstract}

% A category with the (minimum) three required fields
\vspace{1mm}
\noindent
{\bf Categories/Subject Descriptors:} 
D.2.8 [Software Engineering]: Product metrics;
I.2.6 [Artificial Intelligence]: Induction

\vspace{1mm}
\noindent
{\bf General Terms:} Experimentation, Algorithm}

\vspace{1mm}
\noindent
{\bf Keywords:} defect prediction, CART, random forests,
Fastmap,   bootstrap sampling, effect size, A12.

\section{Introduction}

\section{Conclusion}

\section*{Acknowledgements}
This work was partially funded by an National Science
Foundation grant NSF-CISE 1302169.

\vspace*{0.5mm}
\scriptsize

\bibliographystyle{unsrt}
\bibliography{refs}  
\end{document}

### 2015-02-18 phaseDelay.tex
Updates from ShareLaTeX

@@
-\documentclass{sig-alternative}
-%\usepackage{multirow}
-%\usepackage{color}
-%% \usepackage{colortbl}
-%% \usepackage{picture}
-%% %\%usepackage{algorithm}
-%% %\usepackage{algorithmicx}
-%% %\usepackage{algpseudocode}
-%% %\renewcommand{\algorithmicrequire}{\textbf{Input:}}
-%% %\renewcommand{\algorithmicensure}{\textbf{Output:}}
-%% \newcommand{\quart}[3]{\begin{picture}(100,6)%1
-%% {\color{black}\put(#3,3){\circle*{4}}\put(#1,3){\line(1,0){#2}}}\end{picture}}
+\documentclass{sig-alternate}
+\usepackage{multirow}
+\usepackage{color}
+\usepackage{colortbl}
+\usepackage{picture}
+\usepackage{algorithm}
+\usepackage{algorithmicx}
+\usepackage{algpseudocode}
+\renewcommand{\algorithmicrequire}{\textbf{Input:}}
+\renewcommand{\algorithmicensure}{\textbf{Output:}}
 
-%% %% timm tricks
-%% \newcommand{\bi}{\begin{itemize}[leftmargin=0.4cm]}
-%% \newcommand{\ei}{\end{itemize}}
-%% \newcommand{\be}{\begin{enumerate}}
-%% \newcommand{\ee}{\end{enumerate}}
-%% \newcommand{\tion}[1]{\S\ref{sect:#1}}
-%% \newcommand{\fig}[1]{Figure~\ref{fig:#1}}
-%% \newcommand{\eq}[1]{Equation~\ref{eq:#1}}
+%timm tricks
+\newcommand{\bi}{\begin{itemize}[leftmargin=0.4cm]}
+\newcommand{\ei}{\end{itemize}}
+\newcommand{\be}{\begin{enumerate}}
+\newcommand{\ee}{\end{enumerate}}
+\newcommand{\tion}[1]{\S\ref{sect:#1}}
+\newcommand{\fig}[1]{Figure~\ref{fig:#1}}
+\newcommand{\eq}[1]{Equation~\ref{eq:#1}}
 
-%% %% space saving measures
+\usepackage[shortlabels]{enumitem} 
+\usepackage{times}
+\newcommand{\subparagraph}{}
+\usepackage{url}
+\def\baselinestretch{1}
 
-%% \usepackage[shortlabels]{enumitem} 
-%% %\usepackage{times}
+\setlist{nosep}
+ \usepackage[font={small}]{caption, subfig}
+\setlength{\abovecaptionskip}{1ex}
+ \setlength{\belowcaptionskip}{1ex}
 
-%% \usepackage{url}
-%% \def\baselinestretch{1}
-
-%% \setlist{nosep}
-%%  \usepackage[font={small}]{caption, subfig}
-%% \setlength{\abovecaptionskip}{1ex}
-%%  \setlength{\belowcaptionskip}{1ex}
-
-%%  \setlength{\floatsep}{1ex}
-%%  \setlength{\textfloatsep}{1ex}
-%% %\usepackage[compact,small]{titlesec}
-%% \DeclareMathSizes{7}{7}{7}{7} 
-%% \pagenumbering{arabic}
-%% \setlength{\columnsep}{7mm}
+ \setlength{\floatsep}{1ex}
+ \setlength{\textfloatsep}{1ex}
+\usepackage[compact,small]{titlesec}
+\DeclareMathSizes{7}{7}{7}{7} 
+\pagenumbering{arabic}
+\setlength{\columnsep}{7mm}
 
@@
 \conferenceinfo{FSE}{'15 Bergamo, Italy}
-\title{}
-\numberofauthors{2}
+\title{Phase Delay Increases  Repair Time. Not.}
+\numberofauthors{3}
 \author{
 \alignauthor
-Tim Menzies, Carter Pape\\
+Tim Menzies, \\Carter Pape\\
        \affaddr{CS, NcState, USA}\\
@@
 \alignauthor
-Lucas Layman\\
-       \affaddr{Fraunhofer Center, Exp. SE, USA}\\ 
-       llayman@fc-md.umd.edu
-\alignauthor
-William Curtis, Forrest Schull\\
+William Curtis,\\ Forrest Schull\\
 \affaddr{SEI, CMU, USA}
 wrn,fjshull@sei.cmu.edu
+\alignauthor
+Lucas \\Layman\\
+       \affaddr{Fraunhofer Center,  USA}\\ 
+       llayman@fc-md.umd.edu
 }
@@
 \begin{abstract}
+This paper categorically rejects 
+the myth that 
+the longer a defect persists in a system, the harder
+it is to remove that defect.  
+This myth is very widely believed
+and that the evidence for this myth dates back to certain
+isolated studies dating from the 1970s. Recent evidence from
+230 software projects from around the world shows that this
+myth is completely false.
 
-%RQ1: Does tuning affect learners' performance?
-%RQ2: How to choose
-
+More important that our particular results, this paper
+raises the methodological question of how a small number
+of observations from the 1970s can become dogma for decades.
+The obvious next question arising from this study is ``how
+many other widely held beliefs in software engineering are
+not supported by current data?''.
 \end{abstract}
@@
 D.2.8 [Software Engineering]: Product metrics;
-I.2.6 [Artificial Intelligence]: Induction
 
@@
 \noindent
-{\bf General Terms:} Experimentation, Algorithm}
-
-\vspace{1mm}
-\noindent
-{\bf Keywords:} defect prediction, CART, random forests,
-Fastmap,   bootstrap sampling, effect size, A12.
+{\bf Keywords:} defect prediction,
 

### 2015-02-23 phaseDelay.tex
init commit
Updates from ShareLaTeX
added stuff for lucas

@@
 \renewcommand{\algorithmicensure}{\textbf{Output:}} 
+\newcommand{\lucas}[1]{\textcolor{red}{#1}} 
+\newcommand{\bill}[1]{\textcolor{blue}{#1}} 
+\newcommand{\carter}[1]{\textcolor{cyan}{#1}} 
+%\newenvironment{changed}{\par}{\par}
 
@@
 \conferenceinfo{FSE}{'15 Bergamo, Italy}
-\title{Phase Delay Increases  Repair Time. Not.}
+\title{Revisiting the Truisms of Software Engineering:\\ Does Phase Delay Exponentially Increases  Repair Time?}
 \numberofauthors{3}
@@
 \begin{abstract}
-This paper categorically rejects 
-the myth that 
-the longer a defect persists in a system, the harder
-it is to remove that defect.  
-This myth is very widely believed
-and that the evidence for this myth dates back to certain
-isolated studies dating from the 1970s. Recent evidence from
-230 software projects from around the world shows that this
-myth is completely false.
+Does
+repair time increases exponentially
+the longer a defect persists in a system?
+This  is very widely belief and the core rational  for
+ rejecting   traditional linear development methods
+ in favor of more cyclic agile approaches.
 
-More important that our particular results, this paper
-raises the methodological question of how a small number
-of observations from the 1970s can become dogma for decades.
-The obvious next question arising from this study is ``how
-many other widely held beliefs in software engineering are
-not supported by current data?''.
+This paper shows that this belief is maintained
+quite strongly in the industrial software engineering
+community (but, in the academic community, somewhat less so).
+Yet based on a sample of 
+\carter{230} software projects from around the world from 
+\bill{2005 to 2013}, we argue that this belief is mostly 
+incorrect. Specifically: the median fix time for issues
+is just a few minutes and this increases by a small linear amount
+the longer the issue remains in the system.
+
+What was observed was a small number
+of defects that take a remarkably long time to resolve
+For these issues, if an issue takes more an hour
+to resolve then it will usually require many more hours of work.
+These long-tail bugs
+are more prevalent the longer a project delays issue removal. 
+In order to handle such ``long tail'' issues,
+we suggest  augmenting software
+development groups with additional ``tiger teams'' that are triggered
+when issue resolution time increases beyond a certain number of
+minutes.
 \end{abstract}
@@
 \section{Introduction}
+Several recent papers call into questions long-held trusims in the field
+of software engineering. Devanbu reports that, in practice, the advantage of types languages
+is only very minimal. Nagappan et al. found that XXX, contrary to the advice of Dyketra,
+it is not necessarily true that ``goto considered harmful'' XXX. More generally,
+several recent  papers comment on ``locality'' effects where effects
+thold  generally across XXXX
 
+Perhaps it is time to take a look at the base premises of our field, particularly in light of the large software analytics data sets newly available for reseaercehrs.
+
+XXX tsp. 40MB . 235 prokects
+
+\section{Related Work}
+
+In their text Romabach laws
+
+Kent Beck's original extreme programming text is big on this.
+
+Historical  evidence for this dataes back to some studies in the 1970s. To the best of
+our knowledge, not updated since. Regardless, as discussed in the next section, this
+
+\input{lucasRelatedWork}
+
+\section{lucasSurvey}
+
+\input{lucas}
+
+\section{billTSP}
+
+\section{carterCharts}
+
+\input{carter}
+
+\sction{Data}
+
+\begin{figure}[!t]
+
+%\renewcommand{\baselinestretch}{0.8}
+\scriptsize
+\begin{center}
+\begin{tabular}{r|rrr|ll}
+  Sample&\multicolumn{3}{c|}{Percentiles}\\ 
+size & 25th & 50th & 75th & Phase injected & Phase removed\\\hline
+57& 6& 16& 47&  BeforeDevelopment&Code\\
+72& 2& 4& 11&  BeforeDevelopment&CodeInspect\\
+165& 6& 18& 56&  BeforeDevelopment&Test\\
+50& 6& 20& 37&  BeforeDevelopment&IntTest\\
+42& 8& 28& 65&  BeforeDevelopment&SysTest\\\hline
+
+66& 2& 6& 15&  Planning&Planning\\
+41& 1& 5& 16&  Planning&ReqtsReview\\\hline
+
+23& 0& 1& 3&  Reqts&Reqts\\
+245& 6& 13& 23&  Reqts&ReqtsReview\\
+289& 7& 16& 30&  Reqts&ReqtsInspect\\
+32& 6& 21& 40&  Reqts&Design\\
+49& 2& 7& 24&  Reqts&DesignInspect\\\hline
+
+94& 2& 5& 10&  HLD&HLDReview\\
+133& 3& 8& 19&  HLD&HLDInspect\\
+33& 2& 5& 17&  HLD&Design\\\hline
+
+37& 2& 4& 6&  Design&Design\\
+455& 11& 24& 53&  Design&DesignInspect\\
+218& 5& 14& 30&  Design&Code\\
+166& 4& 11& 22&  Design&CodeInspect\\
+542& 14& 31& 60&  Design&Test\\
+67& 6& 15& 40&  Design&IntTest\\
+92& 10& 25& 53&  Design&SysTest\\\hline
+
+53& 2& 4& 9&  DesignInspect&DesignInspect\\
+38& 1& 3& 9&  DesignInspect&Code\\\hline
+
+126& 4& 12& 31&  Code&Code\\
+459& 11& 24& 45&  Code&CodeInspect\\
+461& 11& 26& 55&  Code&Test\\
+348& 10& 26& 62&  Code&IntTest\\
+230& 8& 18& 39&  Code&SysTest\\
+71& 2& 11& 28&  Code&AcceptTest\\\hline
+
+64& 2& 4& 10&  CodeInspect&CodeInspect\\
+57& 2& 5& 13&  CodeInspect&Test\\\hline
+
+110& 3& 10& 23&  Test&Test\\
+66& 2& 4& 12&  Test&QualTest\\\hline
+
+32& 1& 2& 4&  IntTest&QualTest\\
+36& 1& 11& 30&  IntTest&IntTest\\
+21& 1& 2& 5&  IntTest&SysTest\\
+ \end{tabular}
+\end{center}
+\caption{Distribution of fix times seen in SEI TSP data.}
+\label{fig:faw}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\scriptsize
+\begin{tabular}{rrl}
+year& \# issues&\\\hline
+2006 &  44 &\\
+2007 &  34 &\\
+2008&  288 &\rule{1mm}{2mm}\\
+2009&  846 &\rule{3mm}{2mm}\\
+2010& 1007 &\rule{3mm}{2mm}\\
+2011& 3273 &\rule{10mm}{2mm}\\
+2012&18102 &\rule{45mm}{2mm}\\
+2013&20336 &\rule{50mm}{2mm}\\
+2014& 3307 & \rule{10mm}{2mm}\\\cline{1-2}
+Total:&47228
+\end{tabular}
+\end{center}
+\caption{This paper studies 47,228 issues recorded 2006 to 2014.}\label{fig:years}
+\end{figure}
+\begin{figure*}
+\begin{center}
+\includegraphic[width=6in]{waterfall.png}
+\end{center}
+\caption{Some parts of the waterfall cycle.}
+\end{figure*}
+\begin{figure}
+\begin{center}
+\includegraphic[width=3.5in]{agile.png}
+\end{center}
+\caption{Some parts of the waterfall cycle.}
+\end{figure}
+
+\begin{figure*}[!t]
+
+\renewcommand{\baselinestretch}{0.7}
+\small
+\begin{center}
+\begin{tabular}{ll|r|rl}
+            &                  & Sample\\
+Phase injected & Phase removed & size & \multicolumn{2}{l}{Scale up, w.r.t. to first phase}
+\input{deltas}
+\end{tabular}
+\end{center}
+\caption{Scale ups for median time to resolve issues (taken from \fig{raw}).}
+\label{fig:scale}
+\end{figure*}
+
+The starting point for this paper were the times required to fix 47,228 issues.
+All data was collected
+by the Software Engineering Institute in the period 2006 to 2014 (see \fig{years}).
+To the best of our knowledge, this is the largest sample of ``time to fix issues''
+yet analyzed.
+
+This data was collected using the   methodology described by
+Watts Humphrey's Team Software Process~\cite{tsp00}. TSP is an extension of Humphrey's early
+work on Personnel Software Process~\cite{psp05}.  PSP encourages developers to continually make estimates
+about their day-to-day tasks, then compare those estimates against the actual development effort
+(in this way, developers can acquire a more realistic understanding of their work behaviour).
+
+In all data, was collected from 172 separate projects. Given Beck's comments (above) about phase delay motivating agile processes, we take care to identify the differences of these 
+projects to agile development projects. In what we call a ``standard agile'' project, there exists
+some backlog of tasks. When teams complete their current tasks, they reach into to the backlog
+to select the next task(s) to complete. That selection process may use a variety of criteria
+to prioritize which  tasks are selected (for more details on that selection process, see~\cite{me09j,port08,boehmturner03}). Tasks are completed in ``sprints'' that can last hours,
+days, but rarely not more than weeks. Each day meet for brief ``scrum'' sessions to assess (and possibly alter) their current progress on the goals of the spring.
+
+One way to characterize the above is a ``micro cyclic'' approach where teams 
+constantly reassess their priorities using feedback gained after trying to code 
+functionality X,Y,Z. An alternate, more traditional approach is a ``macro linear''
+approach such as the ``waterfall methodology'' that develops code after an extensive
+pre-planning stage:
+
+\[
+\mathit{Reqts}
+\mathit{ReqtsReview}
+\mathit{ReqtsInspect}
+
+\mathit{HighLevelDesign}
+\mathit{HighLevelReview}
+\mathit{HighLevelInspect}
+
+\mathit{Design}
+\mathit{DesignReview}
+\mathit{DesignInspect}
+
+\mathit{Code}
+
+\mathit{Compile}
+\mathit{Test}
+\mathit{IntTest}
+\mathit{SysTest}
+\]
+
+In the original paper defining and 
+critiquing ``waterfall'', Royce defines a set of assessment and feedback processes
+such that
+
+care to contrast these 
+The majority of the teams follow a basically linear process within a single feature of physical component (usually no more than a week or two work in total).
+Unlike some agile teams, *most*TSP teams will not conflate development and unit test.
+This is usually pretty clear from the task  phase sequencing.
+A few teams choose to follow a Test Driven Development like process in which code review or inspection will follow test.
+In a more traditional “micro-linear” TSP approach, unit test follows inspections.
+
+Our data reporting the time required
+to resolve issues in phase $j$ that there were introduced in phase $i, i<j$.
+
+The defect typ that we studied were
+\be 
+\item design, compile, test, or other support system problems = Environment
+procedure calls and reference, I/O, user formats = Interface
+structure, content = Data
+comments, messages = Documentation
+spelling, punctuatio typos, instruction formats = Syntax
+logic, pointers, loops, recursion, computation, function defects = Function
+error messages, inadequate checks = Checking
+change management, library, version control = Build, package
+declaration, duplicate names, scope, limits = Assignment
+configuration, timing, memory = System
 \section{Conclusion}
@@
 \bibliography{refs}
+
 \end{document}

### 2015-02-23 phaseDelay.tex
Updates from ShareLaTeX

@@
 \usepackage{picture}
+\usepackage{wrapfig}
 \usepackage{algorithm}
+\usepackage{graphicx}
 \usepackage{algorithmicx}
@@
 \conferenceinfo{FSE}{'15 Bergamo, Italy}
-\title{Revisiting the Truisms of Software Engineering:\\ Does Phase Delay Exponentially Increases  Repair Time?}
+\title{Revisiting the Truisms of Software Engineering:\\ Does Phase Delay Dramatically Increases  Repair Time?}
 \numberofauthors{3}
@@
 Does
-repair time increases exponentially
+repair time increases dramatically
 the longer a defect persists in a system?
@@
 
-\sction{Data}
+\section{Data}
 
@@
 \end{figure}
-\begin{figure*}
+\begin{figure*}[!t]
 \begin{center}
-\includegraphic[width=6in]{waterfall.png}
+\includegraphics[width=6in]{waterfall2.png}
 \end{center}
-\caption{Some parts of the waterfall cycle.}
+\caption{Different approaches to software development:  waterfall and agile (bottom left).}
 \end{figure*}
-\begin{figure}
-\begin{center}
-\includegraphic[width=3.5in]{agile.png}
-\end{center}
-\caption{Some parts of the waterfall cycle.}
-\end{figure}
 
@@
 \end{figure*}
+
+The data used in this work shirai14
+
+The Team Software ProcessSM (TSPSM) provides a
+framework that teams use to collect software process data in real
+time, using a defined disciplined process
+
+\begin{figure}
+\begin{center}
+\scriptsize\begin{tabular}{|l@{~:~}l|}\hline
+Bug Prediction Dataset &http://bug.inf.usi.ch \\
+Eclipse Bug Data &http://goo.gl/tYKahN \\
+FLOSSMetrics& http://flossmetrics.org \\
+FLOSSMole &http://flossmole.org \\
+IBSBSG& http://www.isbsg.org \\
+ohloh& http://www.openhub.net \\
+PROMISE &http://promisedata.googlecode.com \\
+Qualitas Corpus &http://qualitascorpus.com \\
+Software Artifact Repository &http://sir.unl.edu \\
+SourceForge Research Data &http://zerlot.cse.nd.edu \\
+Sourcerer Project &http://sourcerer.ics.uci.edu \\
+Tukutuku &http://www.metriq.biz/tukutuku \\
+Ultimate Debian Database &http://udd.debian.org\\\hline
+\end{tabular}
+\end{center}
+\caption{Some repositories of software engineering data.}\label{fig:sedata}
+\end{figure}
+
+The TSP couples
+the use of integrated product teams with the Personal Software
+ProcessSM (PSPSM) to give engineers a defined, planned, and
+measured process. For example, PSPSM encourages developers to continually make estimates
+about their day-to-day tasks, then compare those estimates against the actual development effort
+(in this way, developers can acquire a more realistic understanding of their work behaviour
+
+The Software Engineering Institute (SEI) at
+Carnegie Mellon University has collected data from organizations
+that have adopted TSP. 
+Our efforts focused on data collected by projects that launched
+after 2009 and used an automatic data recording tool. 
+Data collected included time logs, defect
+logs, added and modified size logs, and work
+breakdown strcture logs.
+The projects
+were mostly small to medium, with a median duration of 46 days
+and a maximum duration of 90 days. Median team size was 7
+people, with a maximum of 40. Hence,  total development 
+effort\footnote{
+Effort =  duration*teamSize/workDaysPerYear.} for
+these projects ranged from 1.3 years (median) to 14.3 years (max).
+
+In these logs
+the log records include work start time, work end time, delta
+work time, and interruption time. Software engineers are often
+interrupted by meetings, requests for technical help, reporting, and
+so forth. These events are recorded, in minutes, as interruption
+time. In this paper, when we report ``time to resolve an
+issue'', we show the difference between the start and end times
+of a work session, with any interruption time subtracted (the
+difference in times, minus the interruptions).
+
+As of January 2014, the SEI TSP database contained data from 109
+TSP projects. The projects started between July 2009 and
+September 2013; they included 34 teams and 309 people. Among
+the database fact tables, the time store contains 103,023 time logs,
+18,408 defect logs, and 7,464 size logs\footnote{\bill{we need to 
+update this table}}.
+
+A common property of real-world data sets is the presence
+of noisy entries (superfluous  or spurious data). 
+The level of noise can be quite high. As reported
+in \cite{shepperd12}, around
+10\% to 30\%
+of the records in the NASA MDP defect data sets are
+affected by noise. Nichols et al.~\cite{shirai14}  report that
+the noise levesl in the SEI TSE data are smaller than those seen
+in other data sets. They found in the SEI TSP data that:\bi 
+\item
+4\% of the data was incorrect such as  null values of illegal formats;
+\item  2\% of the data has inconsistencies such as timestamps
+where the stop time was before the start time;
+\item 3\% of the data had data that was not credible
+such as tasks listed in one day that took like than six hours.
+\ei 
+That said, certain semantic features of the SEI TSP data should be noted.
+Firstly, in the current TSP collection tool, 
+fix times are only the developer time for the developer walking through the phased of \fig{waterfall}.
+We are currently tracking the fix time for later issues (e.g. those raised during  acceptance test and later
+product life cycle). Informally, we can say that in that data we have not detected
+a dramatic phase escalation effect (but we have nothing definitive yet to publish  at this time).
+
+\bill{somewhere you have one note on \underline{find} and fix times.  for this paper, we need just fix times. but is there
+anything we need to fret about re \underline{find} times?}
+
+\fig{sedata} lists some of the publicly available
+software projects data sets currently available to 
+SE researchers. Most of the \fig{sedata} data sets have software product information
+such as full source code, or summaries of static features.
+A subset of that data contain issue or defect reports and/or
+the time taken to build these systems. Compared to those data
+setss, the SEI TSP data contains much more detailed process details
+including a detailed phased breakdown showing what happened at what
+phases of the lifecycle. The phased used in this paper are shown in \fig{waterfall}. Note that, in that figure:
+\bi 
+\item
+Several  phases have the same  sub-activities of {\em review} and {\em inspect}\footnote{\bill{plz distinguish review and inspect. does one usually happen first?}}
+\item Testing is divided into several statges \footnote{\bill{need a one
+line description of test vs qualtest vs inttest vs systemTest vs AcceptTest. I tries some words in the Key to \fig{sedata}. dont know if i got it right.}}
+\ei
+
+software-related data sets are available to researchers; well
+known examples include those from ISBSG (the International
+Software Benchmarking Standards Group) and the Promise Data
+Repository. The ISBSG provides datasets from more than 5,600
+software development projects and publishes data analysis reports
+[11]. The Promise Data Repository mainly stores product metrics
+of open source software and NASA projects [17]. For a  list
+of other Compared to
+other data sets, the TSP database provides a granular level of
+development detail that enables not only analysis with great
+precision at different levels (e.g., project, team, component, and
+individual developer), but also additional opportunity to assess
+data quality. We hope to make the TSP database more broadly
+available in the future to supplement the other research resources
 
@@
 
+\clearpage
 \vspace*{0.5mm}
 \scriptsize
-
 \bibliographystyle{unsrt}

### 2015-02-23 phaseDelay.tex
various things

@@
 \begin{center}
-\scriptsize
 \begin{tabular}{rrl}
@@
 \renewcommand{\baselinestretch}{0.7} 
-\small
 \begin{center}
@@
 \end{center}
-\caption{Scale ups for median time to resolve issues (taken from \fig{raw}).}
+\caption{50th percentile (median) scale ups  for  time to resolve issues (taken from \fig{raw}).}
 \label{fig:scale}
+\end{figure*}
+
+In these results, we checked if a small number of bugs are most expensive. 
+To check this:
+\bi
+\item
+We repeated the analysis that generated \fig{scale},
+but instead of looking at the 50th percentile, we displayed the scale up factors
+\item 
+We only
+checked the Design and Code scale up results since, from \fig{scale}, it is clear these
+have the most examples of longest phase delays.
+\ei 
+The  results are shown in \fig{scale90} and these
+results are somewhat different for Design and Coding issues:
+\bi 
+\item For Design issues,  these have the same
+general form as the 50th percentile results. That is, while it it certainly faster
+to remove thing sin the phase where they are created, once we leave that phase
+it does not seem to matter much how many phases we wait before fixing the issue.
+\begin{itemize}
+\item For Coding issues, the seems little impact of phase delay on the time
+required to fix.
+\end{itemize}
+
+\begin{figure*}[!t]
+
+\renewcommand{\baselinestretch}{0.7} 
+\begin{center}
+\begin{tabular}{ll|r|rl}
+            &                  & Sample\\
+Phase injected & Phase removed & size & \multicolumn{2}{l}{Scale up, w.r.t. to first phase}
+\input{deltaBig}
+\end{tabular}
+\end{center}
+\caption{90th percentile scale ups  for  time to resolve issues .}
+\label{fig:scale90}
 \end{figure*}
@@
 Firstly, in the current TSP collection tool, 
-fix times are only the developer time for the developer walking through the phased of \fig{waterfall}.
-We are currently tracking the fix time for later issues (e.g. those raised during  acceptance test and later
-product life cycle). Informally, we can say that in that data we have not detected
-a dramatic phase escalation effect (but we have nothing definitive yet to publish  at this time).
+fix times are only the developer time for the developer walking through the phases of \fig{waterfall}.
+We are currently tracking the fix time for post-release issues (e.g. those raised during  acceptance test and later
+product life cycle). So far, in that post-release data,  we have not detected
+a dramatic phase escalation effect (but at this time, we have nothing definitive comment on that matter).
 
@@
 configuration, timing, memory = System
+\ee
 \section{Conclusion}

### 2015-02-23 phaseDelay.tex
Added more references

@@
 \renewcommand{\algorithmicensure}{\textbf{Output:}} 
-\newcommand{\lucas}[1]{\textcolor{red}{#1}} 
-\newcommand{\bill}[1]{\textcolor{blue}{#1}} 
-\newcommand{\carter}[1]{\textcolor{cyan}{#1}} 
+\newcommand{\lucas}[1]{\textcolor{red}{LUCAS: #1}} 
+\newcommand{\bill}[1]{\textcolor{blue}{BILL: #1}} 
+\newcommand{\carter}[1]{\textcolor{cyan}{CARTER: #1}} 
 %\newenvironment{changed}{\par}{\par}
@@
 
+We offer the following notes on the data used in this study.
+These notes will be somewhat detailed and we offer them in order
+to make the following point:
+\bi 
+\item The results section of this paper fails to find that  phase delay causing dramatic increases
+in time to fix issues;
+\item This lack-of-phase-delay effect {\em cannot} be explained away just by  saying that
+the projects in this sample adopted something like  agile methods to reduce re-work costs.
+\ei
+
 The Team Software ProcessSM (TSPSM) provides a
@@
 
-The TSP couples
-the use of integrated product teams with the Personal Software
-ProcessSM (PSPSM) to give engineers a defined, planned, and
-measured process. For example, PSPSM encourages developers to continually make estimates
-about their day-to-day tasks, then compare those estimates against the actual development effort
-(in this way, developers can acquire a more realistic understanding of their work behaviour
+\fig{sedata} lists some of the publicly available
+software projects data sets currently available to 
+SE researchers. Most of the \fig{sedata} data sets have software product information
+such as full source code, or summaries of static features.
+A subset of that data contain issue or defect reports and/or
+the time taken to build these systems. Compared to those data
+setss, the SEI TSP data contains much more detailed process details
+including a detailed phased breakdown showing what happened at what
+phases of the lifecycle.
+
+As dictacted by the  the projects in our SEI TSP data set  use coaches,
+peer review, and  personnel reviews:
+a ``coach'' is the team member  authorized to submit project data
+(before submission,
+these coaches check the data for obvious errors);
+``personnel review'' is a technique taken  from the Personal Software
+ProcessSM (PSPSM); and
+``peer review'' is a standard technique in
+traditional software engineering.
+PSPSM encourages developers to continually make and review their personnel estimates
+about their day-to-day tasks, then compare those estimates against the actual development effort.
+In this way, developers can acquire a more realistic understanding of their work behaviour.
+As to peer review,  Basili and Boehm write  commented in 2001~\cite{boehm01} 
+that peer reviews can catch over half the defects introduced into a system.
+Peer review can be conducted on any artifact generated anywhere in the software
+lifecycle and can quickly be adapted to new kinds of artifacts.
+
+The phased used in this paper are shown in \fig{waterfall}. Note that, in that figure:
+\bi 
+\item
+Several  phases have the same  sub-activities of {\em review} and {\em inspect}\footnote{\bill{plz distinguish review and inspect. does one usually happen first?}}
+\item Testing is divided into several statges \footnote{\bill{need a one
+line description of test vs qualtest vs inttest vs systemTest vs AcceptTest. I tries some words in the Key to \fig{sedata}. dont know if i got it right.}}
+\ei 
+Also shown in \fig{waterfall} (bottom left) is the standard definition of an agile process~\cite{boehmturner03}:
+Given some
+some backlog of tasks, wgeb teams complete their current tasks, they select the next task(s) to complete. That selection process may use a variety of criteria
+to prioritize which  tasks are selected (for more details on that selection process, see~\cite{me09j,port08,boehmturner03}). Tasks are completed in ``sprints'' that can last hours,
+days, but rarely not more than weeks. Each day meet for brief ``scrum'' sessions to assess (and possibly alter) their current progress on the goals of the sprint.  
+Agile teams race to generate releases
+(in the continuous release model, releases can be generated on a daily basis, or even faster).  
+Experience gained from those releases informs the discussion in the daily scrums which, in turn,
+can inform the team's decisions on how to select and implement the next set of tasks for next sprint.
+In fact, that experience can result in changing some/all of the tasks in the current backlog.
+
+TSP is not antithetical to agile--  indeed a TSP-waterfall style project can adopt aspects
+of agile.  For example the agile loop  could be applied
+over one or more of the phases shown in the long waterfall chain of \fig{waterfall}. Some
+TSP teams adopt something like test-driven-development\footnote{TDD is an agile-method
+where the ``test'' is the primary driver of the design. The tests are written first,
+then the code to support those tests. TDD proceeds in three steps: red (where there
+are broken tests); green (where tests are passing); and, possibly, refactor (where
+the code is re-organized based on feedback from those tests~\cite{fraser03}.} where
+reviews are scheduled after testing. Some of the groups in the SEI TSP data used that approach but
+they do not effect the main conclusions of this paper:
+\bi 
+\item They were strongly in the minority\bill{any numbers on this?};
+\item Most of our phase delay data comes from much early in the waterfall model
+of \fig{waterfall}.
+\ei 
+That said, there are some very ``un-agile'' aspects
+of the processes used by the   projects in the TSP SEI data.
+For teams to be included in this data set they had to be following certain 
+ guidelines
+laid down by TSPSM (and SEI staff monitored these projects to ensure that that the
+following guidelines were maintained\footnote{bill, can you talk at all about your work doing site visits? i.e. how did you ensure compilance with the following}):
+\be 
+\item  The TSP projects studied here offer new releases at least every three months, but
+often much longer that that\bill{any numbers on that}.
+\item The TSP projects studied     spend one day (or more) working on 
+\item
+The time spent in the of a design activity/phase should require 
+approximately as much effort as the coding phase. That is, for the projects
+studied here, at least half of the reflection on the requirements, design, code
+arises without feedback from running code.
+\item The teams
+cannot combine development and  testing. That is, as before,
+it can be days/weeks
+before team members in the TSP SEI datatet gain  feedback from executing code.  
+\item The time devoted to ``personnel review'' and ``peer review'' (defined below)
+is  about 50\% as much effort as the previous construction phase activity.
+\ee
+
+\section{Data}
 
@@
 
-\fig{sedata} lists some of the publicly available
-software projects data sets currently available to 
-SE researchers. Most of the \fig{sedata} data sets have software product information
-such as full source code, or summaries of static features.
-A subset of that data contain issue or defect reports and/or
-the time taken to build these systems. Compared to those data
-setss, the SEI TSP data contains much more detailed process details
-including a detailed phased breakdown showing what happened at what
-phases of the lifecycle. The phased used in this paper are shown in \fig{waterfall}. Note that, in that figure:
-\bi 
-\item
-Several  phases have the same  sub-activities of {\em review} and {\em inspect}\footnote{\bill{plz distinguish review and inspect. does one usually happen first?}}
-\item Testing is divided into several statges \footnote{\bill{need a one
-line description of test vs qualtest vs inttest vs systemTest vs AcceptTest. I tries some words in the Key to \fig{sedata}. dont know if i got it right.}}
-\ei
-
-software-related data sets are available to researchers; well
-known examples include those from ISBSG (the International
-Software Benchmarking Standards Group) and the Promise Data
-Repository. The ISBSG provides datasets from more than 5,600
-software development projects and publishes data analysis reports
-[11]. The Promise Data Repository mainly stores product metrics
-of open source software and NASA projects [17]. For a  list
-of other Compared to
-other data sets, the TSP database provides a granular level of
-development detail that enables not only analysis with great
-precision at different levels (e.g., project, team, component, and
-individual developer), but also additional opportunity to assess
-data quality. We hope to make the TSP database more broadly
-available in the future to supplement the other research resources
+The majority of the teams follow a more linear process within a single feature of physical component (usually no more than a week or two work in total).
+Unlike some agile teams, *most*TSP teams will not conflate development and unit test.
 
@@
 In all data, was collected from 172 separate projects. Given Beck's comments (above) about phase delay motivating agile processes, we take care to identify the differences of these 
-projects to agile development projects. In what we call a ``standard agile'' project, there exists
-some backlog of tasks. When teams complete their current tasks, they reach into to the backlog
-to select the next task(s) to complete. That selection process may use a variety of criteria
-to prioritize which  tasks are selected (for more details on that selection process, see~\cite{me09j,port08,boehmturner03}). Tasks are completed in ``sprints'' that can last hours,
-days, but rarely not more than weeks. Each day meet for brief ``scrum'' sessions to assess (and possibly alter) their current progress on the goals of the spring.
-
-One way to characterize the above is a ``micro cyclic'' approach where teams 
-constantly reassess their priorities using feedback gained after trying to code 
-functionality X,Y,Z. An alternate, more traditional approach is a ``macro linear''
+projects to agile development projects. An alternate, more traditional approach is a ``macro linear''
 approach such as the ``waterfall methodology'' that develops code after an extensive

### 2015-02-23 phaseDelay.tex
Updates from ShareLaTeX

@@
 \alignauthor
-William Curtis,\\ Forrest Schull\\
+William Nichols,\\ Forrest Schull\\
 \affaddr{SEI, CMU, USA}
@@
 
-We offer the following notes on the data used in this study.
-These notes will be somewhat detailed and we offer them in order
-to make the following point:
-\bi 
-\item The results section of this paper fails to find that  phase delay causing dramatic increases
-in time to fix issues;
-\item This lack-of-phase-delay effect {\em cannot} be explained away just by  saying that
-the projects in this sample adopted something like  agile methods to reduce re-work costs.
-\ei
-
 The Team Software ProcessSM (TSPSM) provides a
 framework that teams use to collect software process data in real
-time, using a defined disciplined process
+time, using a defined disciplined process.
+In particular, three common features of the projects in our SEI TSP data set  are coaches,
+peer review, and  personnel reviews:
+\bi 
+\item
+A ``coach'' is the team member  authorized to submit project data
+(before submission,
+these coaches check the data for obvious errors);
+\item 
+``Personnel review'' is a technique taken  from the Personal Software
+ProcessSM (PSPSM);
+\item 
+``Peer review'' is a standard technique in
+traditional software engineering.
+\ei
+PSPSM encourages developers to continually make and review their personnel estimates
+about their day-to-day tasks, then compare those estimates against the actual development effort.
+In this way, developers can acquire a more realistic understanding of their work behaviour.
+As to peer review,  Basili and Boehm write  commented in 2001~\cite{boehm01} 
+that peer reviews can catch over half the defects introduced into a system.
+Peer review can be conducted on any artifact generated anywhere in the software
+lifecycle and can quickly be adapted to new kinds of artifacts.
 
@@
 \end{figure}
-
-\fig{sedata} lists some of the publicly available
+The data collected in the SEI TSP databases contains 
+extensive process details
+including a detailed phased breakdown showing what happened at what
+phases of the lifecycle. This makes this data somewhat different to the standard
+ publicly available
 software projects data sets currently available to 
@@
 A subset of that data contain issue or defect reports and/or
-the time taken to build these systems. Compared to those data
-setss, the SEI TSP data contains much more detailed process details
-including a detailed phased breakdown showing what happened at what
-phases of the lifecycle.
-
-As dictacted by the  the projects in our SEI TSP data set  use coaches,
-peer review, and  personnel reviews:
-a ``coach'' is the team member  authorized to submit project data
-(before submission,
-these coaches check the data for obvious errors);
-``personnel review'' is a technique taken  from the Personal Software
-ProcessSM (PSPSM); and
-``peer review'' is a standard technique in
-traditional software engineering.
-PSPSM encourages developers to continually make and review their personnel estimates
-about their day-to-day tasks, then compare those estimates against the actual development effort.
-In this way, developers can acquire a more realistic understanding of their work behaviour.
-As to peer review,  Basili and Boehm write  commented in 2001~\cite{boehm01} 
-that peer reviews can catch over half the defects introduced into a system.
-Peer review can be conducted on any artifact generated anywhere in the software
-lifecycle and can quickly be adapted to new kinds of artifacts.
+the time taken to build these systems.  We are unaware
+of any of these having detailed phase-by-phase breakdowns of project data.
 
@@
 of the processes used by the   projects in the TSP SEI data.
-For teams to be included in this data set they had to be following certain 
- guidelines
-laid down by TSPSM (and SEI staff monitored these projects to ensure that that the
-following guidelines were maintained\footnote{bill, can you talk at all about your work doing site visits? i.e. how did you ensure compilance with the following}):
-\be 
+A TSP participant spends much time reflecting on the project,
+undisturbed by the behaviour of the  executables.
+In fact,
+it can be days/weeks
+before TSP participants gain  feedback from executing code, for the following reasons:  
+\bi 
+\item
+For TSP projects, the times spent in the of a design activity/phase is   
+approximately as much effort as the coding phase.  
+\item Our TSP teams did not
+ combine development and  testing. 
  \item  The TSP projects studied here offer new releases at least every three months, but
-often much longer that that\bill{any numbers on that}.
-\item The TSP projects studied     spend one day (or more) working on 
-\item
-The time spent in the of a design activity/phase should require 
-approximately as much effort as the coding phase. That is, for the projects
-studied here, at least half of the reflection on the requirements, design, code
-arises without feedback from running code.
-\item The teams
-cannot combine development and  testing. That is, as before,
-it can be days/weeks
-before team members in the TSP SEI datatet gain  feedback from executing code.  
-\item The time devoted to ``personnel review'' and ``peer review'' (defined below)
+often much longer that that\bill{any numbers on that}. Note that this is a far slower
+release schedule than seen in, say, continuous integration projects.  
+\item The time devoted to personnel review'' and peer review (defined above)
 is  about 50\% as much effort as the previous construction phase activity.
-\ee
+
+\ei
+
+Having presented all that, we can now present the main point of this section:
+\bi 
+\item The results section of this paper fails to find that  phase delay causing dramatic increases
+in time to fix issues;
+\item This lack-of-phase-delay effect {\em cannot} be explained away just by  saying that
+the projects in this sample adopted something like  agile methods to reduce re-work costs.
+\ei
 
@@
 after 2009 and used an automatic data recording tool. 
+Due to confidentiality restrictions, we cannot offer too many details
+on those project. That said, there were less than a dozen United States
+Department of Defence projects. Also, the majority of
+the projects were weither web portals or banking systems in the US, South Africa, and Mexico. 
+There were also some  medical devices (US, France, Japan, and Germany) and a few from a commercial 
+computer-aided design system (developed by a world wide distributed team).
+
 Data collected included time logs, defect

### 2015-02-25 phaseDelay.tex
Updates from ShareLaTeX

@@
 XXX tsp. 40MB . 235 prokects
+
+\section{Background}
+
+It is no understatement to say that 
+concerns with  phase delay have 
+a major  motivator for changing
+SE practices. Basili and Boehm~\cite{boehm01}  commented in
+2001 that, since the 1980s, the phase delay
+``has been a major driver in focusing
+industrial software practice on thorough
+requirements analysis and design,
+on early verification and validation, and
+on up-front prototyping and simulation
+to avoid costly downstream fixes.''
+
+Also, concerns with the phase delay  had
+ a significant impact on the history of agile programming. Kent Beck is
+ particulary eloquent
+
+Under certain circumstances, the exponential rise in the cost of changing software over time can be flattened.
+If we can flatten the curve, old assumptions about the best way to develop software no longer hold.
+
+One of the universal assumptions of software engineering is that the cost of changing a program rises exponentially over time.
+I can remember sitting in a big linoleum-floored classroom as a college junior and seeing the professor draw on the board the curve found in Figure 1.
+
+The cost to fix a problem in a piece of software rises exponentially over time.
+A problem that might take a dollar to fix if you found it during requirements analysis might costs thousands to fix once the software is in production."
+
+I resolved then and there that I would never let a problem get through to production.
+No sirree, I was going to catch problems as soon as possible.
+I would work out every possible problem in advance.
+I would review and crosscheck my code.
+No way was I going to cost my employer $100,000.
 

### 2015-02-25 phaseDelay.tex
Updates from ShareLaTeX

@@
 
+\begin{center}
+\includegraphics[width=3.3in]{beckB4.png}
+
+\includegraphics[width=3.3in]{beckAFTER.png}
+\end{center}
+
 Also, concerns with the phase delay  had
  a significant impact on the history of agile programming. Kent Beck is
- particulary eloquent
+ particularly eloquent
 
@@
 I can remember sitting in a big linoleum-floored classroom as a college junior and seeing the professor draw on the board the curve found in Figure 1.
+
+The software development community has spent enormous resources in recent decades trying to reduce the cost of change—better languages, better database technology, better programming practices, better environments and tools, new notations.
+
+What would we do if all that investment paid off? What if all that work on languages and databases and whatnot actually got somewhere? What if the cost of change didn't rise exponentially over
+
+This is one of the premises of XP.
+It is the technical premise of XP.
+If the cost of change rose slowly over time, you would act completely differently from how you do under the assumption that costs rise exponentially.
+You would make big decisions as late in the process as possible, to defer the cost of making the decisions and to have the greatest possible chance that they would be right.
+You would only implement what you had to, in hopes that the needs you anticipate for tomorrow wouldn't come true.
+You would introduce elements to the design only as they simplified existing code or made writing the next bit of code simpler.
+￼
+time, but rose much more slowly, eventually reaching an asymptote? What if tomorrow's software engineering professor draws Figure 3 on the board?
 

### 2015-02-25 phaseDelay.tex
truisms  general intro done

@@
 \usepackage{colortbl}
+
 \usepackage{picture}
@@
 \usepackage{times}
+
+\usepackage{cite}
 \newcommand{\subparagraph}{}
@@
 \setlist{nosep}
+
+\usepackage{colortbl}
  \usepackage[font={small}]{caption, subfig}
@@
 
-\section{Background}
+\section{Motivation}
+\subsection{Why Study Triusms?}
+
+Before exploring one particular truism, we digress to discuss the merits of revisiting
+old conclusions in software engineering.
+
+There are many examples of writers proposing general laws of software engineering.
+Processional societies assume such generalities exist when the offer
+ lists of  supposedly general ``best practices'' such as
+the IEEE 1012 standard for software verification~\cite{1012}.
+Budgen \& Kitchenham seek to reorganize SE research using
+general
+conclusions drawn from a larger number of studies~\cite{kitch04,budgen09}.
+Widely-cited researchers 
+list
+conclusions that are meant to hold across most SE projects; e.g.
+Glass~\cite{glass02}; Jones~\cite{jones07}; Boehm~\cite{boehm00b} and Endres \& Rombach
+whose textbook lists ``50 laws of software engineering''~\cite{endres03,rombach11}. 
+Endres \& Rombach are especially adamant   that such generalities exist:  
+\begin{quote}
+Based on {\em repeated and consistent} observations (emphasis added), key lessons of these fields can now be formulated into rules or even laws, providing initial building blocks towards a theoretical foundation that is essential for further research, for teaching and for the practice of software development.
+\end{quote}
+The message of this paper is that we need to be very careful before taking action based
+on these supposed general laws. Contrary to the claims of Endres \& Rombach,
+we found {\em repeated inconsistencies} in the SE literature.
+For example,
+this papers tests   Endres \& Rombach's claim that  phase delay
+ dramatically increases the time to fix issues. We show that (a)~this is one of the most
+ strongly held beliefs amongst industrial practitioners and that (b)~it is
+ not supported by data from  171 projects from the period 2004 to 2014. 
+ This result is one of many similar findings that 
+cast doubt on the existence of  ``key lessons  that can be formulated
+as general laws of SE'':
+\bi
+\item
+Turhan~\cite{me12d} lists 28 studies with contradictory conclusions
+on the relation of OO measures to defects.  Those results
+ directly  contradict some of the laws listed by 
+Endres \& Rombach~\cite{endres03}.
+\item
+Ray et al.~\cite{ray2014lang} tested if   strongly typed languages
+predict for better code quality. In  728 projects,
+they found  only a modest benefit in strong typing
+(and caution that that effect may actually be due to other  conflating factors).
+\item
+Fenton \& Neil~\cite{fenton00,fenton00b}   critique the truism that
+``pre-release fault rates for software
+are a predictor for post-release failures'' (as claimed by~\cite{dunsmore88},
+amongst others). For the systems described in~\cite{fenton97}, they
+show that software modules that were highly fault-prone
+prior to release revealed very few faults after release.
+\item
+Meyer claims that   object-oriented (OO) encapsulation will
+reduce error rates in software~\cite{Meyer1988}.  Yet empirical results suggest
+that debugging an OO program is many times harder and
+longer than debugging a standard procedural program~\cite{hatton98}.
+\item
+A truism of visual programming is that ``visual
+representations are inherently superior to mere textual representations''.
+A review by Menzies suggests that the available
+evidence for this claim is hardly conclusive~\cite{me00v}. 
+\ei
+To be fair to the software engineering community, SE is not the only
+field where practitioners hang on to persistent beliefs, even if the evidence
+for those beliefs is not strong.
+The medical profession applies  many practices based on studies
+that have been disproved. For example,
+a  recent article
+in the Mayo Clinic Proceedings~\cite{prasad13} found  
+146 medical practices based on studies 
+in year $i$, but which were  reversed by subsequent trials within years $i+10$.
+Even when the evidence for or against a treatment or intervention is clear, medical providers and patients may not accept it~\cite{aschwanden10}.
+Aschwanden warns:
+\begin{quote}
+Cognitive biases - such as motivated reasoning (all of us want to believe that the things we do make a difference), base rate neglect (failing to pay attention to what happens in the absence of the intervention), and confirmation bias (the tendency to look for evidence that supports what you already know and to ignore the rest) - also influence how we process information~\cite{aschwanden15}.
+\end{quote}
+
+The cognitive issues that complicate medicine are also found in software engineering.
+According to Passos et al.~\cite{passos11}, commercial developers
+are all too willing to believe in general
+truisms.
+They  warn that developers
+usually assume that the truisms they learn from a few past
+projects are general to 
+all their future projects:
+\begin{quote}\label{q:pass}
+...the past experiences were taken into account without 
+much consideration for their context~\cite{passos11}.  
+\end{quote}
+The results of Jorgensen \& Gruschke support the findings if Passos et al. They repor that 
+  commercial estimation ``gurus'' rarely use lessons
+  from past projects to improve their future estimates~\cite{jorgensen09}. 
+When engineers
+  fail to revise their beliefs, this leads to poor
+ estimates  (see examples in~\cite{jorgensen09}).
+%% This leads to  detrimental
+%% situations where groups or 
+%% individuals working on the same project have conflicting beliefs, which they never debate
+%or refine or unify.
+
+In summary, just like medicine, our field suffers when
+ software engineers do  revise old beliefs.  Therefore, papers like this one
+ that retests old beliefs are vital.
+\subsection{Why Study Phase Delay?}
 
@@
 I would review and crosscheck my code.
-No way was I going to cost my employer $100,000.
+No way was I going to cost my employer \$100,000.
 
@@
 
-\[
-\mathit{Reqts}
-\mathit{ReqtsReview}
-\mathit{ReqtsInspect}
-
-\mathit{HighLevelDesign}
-\mathit{HighLevelReview}
-\mathit{HighLevelInspect}
-
-\mathit{Design}
-\mathit{DesignReview}
-\mathit{DesignInspect}
-
-\mathit{Code}
-
-\mathit{Compile}
-\mathit{Test}
-\mathit{IntTest}
-\mathit{SysTest}
-\]
-
 In the original paper defining and 
@@
 \scriptsize
-\bibliographystyle{unsrt}
+\bibliographystyle{plain}
 \bibliography{refs}
@@
 \end{document}
+
+One way to address the absence of general laws is  {\em local learning}.
+Even if general laws
+do not exist, then {\em there exists general methods for finding the best local laws}.
+For example, one of us (Nichols) works extensively with software projects to help
+them tune their local software process in order to better fit their local needs.XXX.
+
+ Recent studies report that better predictors of the properties of
+software projects (effort and defects) can be generated by first clustering
+(a.k.a. stratifying or contextualizing or localizing) the data then learning
+different models for each cluster~\cite{posnett11,betten14,betta12,yang11,yang13,minku13}.
+Those {\em local learning} approach finds better predictions with lower variance than
+models learned from all the unclustered data~\cite{me12d,me11m}.
+
+The problem with the local learning is that they it can generate conclusions
+that conflict with strongly-held beliefs of the humans members of the software development team.
+It is an open issue how to handle those conflicts since it is unwise to
+always reject the conclusions of local learning  or the software developers.
+The local learners may be wrong due to incorrect assumptions by the data scientists who configured the learners~\cite{me11e,shull02}.
+On the other hand,  the humans may be wrong due to the cognitive biases described above.
+Experienced data scientists use a cyclic approach where
+(a)~the collected data; and (b)~the learning methods; and (c) the goals of the inquiry are matured  using
+insight offered by humans as they study the
+feedback generated by the learners~\cite{Fayyad96,me11e}.
+
+the learners' conclusions are discussed with the team in or
+
+(which may be erroneous
+due to incorrect assumptions
+
+We come to this work since  recent results on {\em locality} by Menzies and others

### 2015-02-26 phaseDelay.tex
Updates from ShareLaTeX

--- 2015-02-25 phaseDelay.tex
@@
 the longer a defect persists in a system?
-This  is very widely belief and the core rational  for
- rejecting   traditional linear development methods
- in favor of more cyclic agile approaches.
+This  is very widely belief that is used to justify 
+radical changed to    traditional software development
+ processes.
 
@@
 Yet based on a sample of 
-\carter{230} software projects from around the world from 
+171 software projects from around the world from 
 \bill{2005 to 2013}, we argue that this belief is mostly 
 incorrect. Specifically: the median fix time for issues
-is just a few minutes and this increases by a small linear amount
+is just a few minutes and this does not tend to increase
 the longer the issue remains in the system.
 
-What was observed was a small number
-of defects that take a remarkably long time to resolve
-For these issues, if an issue takes more an hour
-to resolve then it will usually require many more hours of work.
-These long-tail bugs
-are more prevalent the longer a project delays issue removal. 
-In order to handle such ``long tail'' issues,
-we suggest  augmenting software
-development groups with additional ``tiger teams'' that are triggered
-when issue resolution time increases beyond a certain number of
-minutes.
+This result raises the question:
+how did myths like phase delay survive so long  without  critical reassessment.  
+As a community, we need to reflect more on 
+research practices and on how (and when) we reflect on the prominent hypotheses of our field.
 \end{abstract}
@@
 {\bf Categories/Subject Descriptors:} 
-D.2.8 [Software Engineering]: Product metrics;
+D.2.8 [Software Engineering]: Process metrics.
 
@@
 \noindent
-{\bf Keywords:} defect prediction,
+{\bf Keywords:} phase delay, time to fix, agile, waterfall.
 
 \section{Introduction}
-Several recent papers call into questions long-held trusims in the field
-of software engineering. Devanbu reports that, in practice, the advantage of types languages
-is only very minimal. Nagappan et al. found that XXX, contrary to the advice of Dyketra,
-it is not necessarily true that ``goto considered harmful'' XXX. More generally,
-several recent  papers comment on ``locality'' effects where effects
-thold  generally across XXXX
+In 2013-2014, the time of 
+eleven  million programmers~\cite{pettey14} and
+half a trillion dollars~\cite{avram14} was spent on information technology.
+Such a large and growing effort should be managed and optimized via  well-researched conclusions.  
+It is standard practice
+in other fields such as medicine,
+to continually revisit old conclusions~\cite{prasad13}. Software engineering should do the same lest
+ill-founded  exert an undue influence on how we build software.
 
-Perhaps it is time to take a look at the base premises of our field, particularly in light of the large software analytics data sets newly available for reseaercehrs.
-
-XXX tsp. 40MB . 235 prokects
-
-\section{Background}
-
-It is no understatement to say that 
-concerns with  phase delay have 
-a major  motivator for changing
-SE practices. Basili and Boehm~\cite{boehm01}  commented in
-2001 that, since the 1980s, the phase delay
-``has been a major driver in focusing
+Accordingly, this paper revisits
+the {\em phase delay} effect
+i.e.  the time required to resolve an issue in a software project increases dramatically 
+the longer it remains in the system. 
+Phase delay has long been
+a    motivator for major changes to
+SE practices. Basili \& Boehm comment that since the 1980s,  phase delay:
+\begin{quote}
+...has been a major driver in focusing
 industrial software practice on thorough
@@
 on up-front prototyping and simulation
-to avoid costly downstream fixes.''
+to avoid costly downstream fixes~\cite{boehm01}.
+\end{quote}
+Basili and Boehm are not alone in this assessment of the significance of the phase delay effect. 
+Our surveys of  industrial practitioners (shown below) reveal it to be
+the single most strongly held belief amongst commercial SE engineers;
+Also,
+as discussed later
+in this paper, concerns over phase delay were a major driver in the development of agile software
+processes.
 
-\begin{center}
-\includegraphics[width=3.3in]{beckB4.png}
+The conclusions of this paper is that it is erroneous to believe in phase delay.
+When we looked for supporting evidence for that effect, we found that:
+\be 
+\item The evidence for that effect in the historical literature is very weak (to say the least);
+\item  In  a large sample of   contemporary software projects, there is no evidence for phase delay.
+ \ee
+The rest of this paper present evidence for the above two points. 
+The first point is  documented via  a literature survey and the second point is demonstrated
+via an analysis of  
+ 171 software  projects developed in the period \bill{2004 to 2014}.  To the best of our knowledge,
+ this is the largest study on phase delay yet seen in the literature.
 
-\includegraphics[width=3.3in]{beckAFTER.png}
-\end{center}
+The conclusions of this paper are two fold. Firstly, the SE community needs to stop
+justifying changes to SE
+practices via the phase delay effect. Secondly,   myths like phase delay should not have lasted decades
+in the literature without  critical reassessment.  As a community, we need to reflect more on our
+research practices and on how (and when) we reflect and revisit the prominent hypotheses of our field.
 
-Also, concerns with the phase delay  had
- a significant impact on the history of agile programming. Kent Beck is
- particularly eloquent
+\section{Motivation}
+\subsection{Why Reassess Existing Triusms?}
 
-Under certain circumstances, the exponential rise in the cost of changing software over time can be flattened.
-If we can flatten the curve, old assumptions about the best way to develop software no longer hold.
+Before exploring one particular truism, we digress to discuss the merits of revisiting
+old conclusions in software engineering.
 
+Some researchers   adamantly believe in general
+principles of software
+engineering. Processional societies assume such generalities exist when the offer
+ lists of  supposedly general ``best practices'' such as
+the IEEE 1012 standard for software verification~\cite{1012}.  Endres \& Rombach's comment that 
+\begin{quote}
+Based on repeated and consistent observations, key lessons of these fields can now be formulated into rules or even laws, providing initial building blocks towards a theoretical foundation that is essential for further research, for teaching and for the practice of software development~\cite{endres03,rombach11}
+\end{quote}
+ Endres \& Rombach offer dozens of supposedly general lessons of software engineering~\cite{rombach11}.
+ Many other 
+widely-cited researchers  do the same; e.g.
+Glass~\cite{glass02}; Jones~\cite{jones07}; Boehm~\cite{boehm00b}.
+Budgen \& Kitchenham seek to reorganize SE research using
+general
+conclusions drawn from a larger number of studies~\cite{kitch04,budgen09}.
+
+Nevertheless, there are many empirical  findings that 
+cast doubt on the existence of  ``key lessons  that can be formulated
+as general laws of SE'':
+\bi
+\item
+Turhan~\cite{me12d} lists 28 studies with contradictory conclusions
+on the relation of OO measures to defects.  Those results
+ directly  contradict some of the laws listed by 
+Endres \& Rombach~\cite{endres03}.
+
+\item
+Ray et al.~\cite{ray2014lang} tested if   strongly typed languages
+predict for better code quality. In  728 projects,
+they found  only a modest benefit in strong typing
+(and caution that that effect may actually be due to other  conflating factors).
+\item
+Fenton \& Neil~\cite{fenton00,fenton00b}   critique the truism that
+``pre-release fault rates for software
+are a predictor for post-release failures'' (as claimed by~\cite{dunsmore88},
+amongst others). For the systems described in~\cite{fenton97}, they
+show that software modules that were highly fault-prone
+prior to release revealed very few faults after release.
+\item
+Meyer claims that   object-oriented (OO) encapsulation will
+reduce error rates in software~\cite{Meyer1988}.  Yet empirical results suggest
+that debugging an OO program is many times harder and
+longer than debugging a standard procedural program~\cite{hatton98}.
+\item
+A truism of visual programming is that ``visual
+representations are inherently superior to mere textual representations''.
+A review by Menzies suggests that the available
+evidence for this claim is hardly conclusive~\cite{me00v}. 
+\item Numerous recent {\em local learning} results compare single models
+learned from all available data to multiple models learned from clusters within the data~\cite{betten14,yang11,yang13,minku13,me12d,me11m,betta12,posnett11}.
+A repeated result in those studies in that the local models generated the better effort
+and defect predictions (better median results,
+lower variance in the predictions).
+ \item 
+Other work studies  found dozens of single
+models that performed worse that {\em committees of models}, all with different opinions,
+that contribute to an {\em ensemble} prediction of software development effort~\cite{kocaguneli2012value,azhar13}.
+\ei
+To be fair to the software engineering community, SE is not the only
+field where practitioners hang on to persistent beliefs, even if the evidence
+for those beliefs is not strong.
+The medical profession applies  many practices based on studies
+that have been disproved. For example,
+a  recent article
+in the Mayo Clinic Proceedings~\cite{prasad13} found  
+146 medical practices based on studies 
+in year $i$, but which were  reversed by subsequent trials within years $i+10$.
+Even when the evidence for or against a treatment or intervention is clear, medical providers and patients may not accept it~\cite{aschwanden10}.
+Aschwanden warns:
+\begin{quote}
+Cognitive biases - such as motivated reasoning (all of us want to believe that the things we do make a difference), base rate neglect (failing to pay attention to what happens in the absence of the intervention), and confirmation bias (the tendency to look for evidence that supports what you already know and to ignore the rest) - also influence how we process information~\cite{aschwanden15}.
+\end{quote}
+
+The cognitive issues that complicate medicine are also found in software engineering.
+According to Passos et al.~\cite{passos11}, commercial developers
+are all too willing to believe in general
+truisms.
+They  warn that developers
+usually assume that the truisms they learn from a few past
+projects are general to 
+all their future projects:
+\begin{quote}\label{q:pass}
+...the past experiences were taken into account without 
+much consideration for their context~\cite{passos11}.  
+\end{quote}
+The results of Jorgensen \& Gruschke support the findings if Passos et al. They report that 
+  supposed software engineering    ``experts'' rarely use lessons
+  from past projects to improve their future reasoning~\cite{jorgensen09}. 
+ They note that
+when the experts
+  fail to revise their beliefs, this leads to poor
+ conclusions and software projects  (see examples in~\cite{jorgensen09}).
+%% This leads to  detrimental
+%% situations where groups or 
+%% individuals working on the same project have conflicting beliefs, which they never debate
+%or refine or unify.
+
+In summary, just like medicine, our field suffers when
+ software engineers do  revise old beliefs.  Therefore, it is important
+ to regularly  reexamine    old beliefs.
+\subsection{Why Study Phase Delay?}
+Due to its prominence in the field of software engineering,
+this paper studies phase delay. As mentioned in the introduction, concerns about phase delay have
+been used for decades to motivate major changes to software engineering practices. For example,
+in 2000, Kent Beck used phase delay to motivate his call for ``extreme programming'' (a set of techniques
+that evolved into contemporary agile and DevOps practices). Beck writes:
+\begin{quote}
 One of the universal assumptions of software engineering is that the cost of changing a program rises exponentially over time.
-I can remember sitting in a big linoleum-floored classroom as a college junior and seeing the professor draw on the board the curve found in Figure 1.
-
-The software development community has spent enormous resources in recent decades trying to reduce the cost of change—better languages, better database technology, better programming practices, better environments and tools, new notations.
-
-What would we do if all that investment paid off? What if all that work on languages and databases and whatnot actually got somewhere? What if the cost of change didn't rise exponentially over
-
-This is one of the premises of XP.
-It is the technical premise of XP.
-If the cost of change rose slowly over time, you would act completely differently from how you do under the assumption that costs rise exponentially.
-You would make big decisions as late in the process as possible, to defer the cost of making the decisions and to have the greatest possible chance that they would be right.
-You would only implement what you had to, in hopes that the needs you anticipate for tomorrow wouldn't come true.
-You would introduce elements to the design only as they simplified existing code or made writing the next bit of code simpler.
-￼
-time, but rose much more slowly, eventually reaching an asymptote? What if tomorrow's software engineering professor draws Figure 3 on the board?
+I can remember sitting in a big linoleum-floored classroom as a college junior and seeing the professor draw on the board the curve found in \fig{curve1}.
 
 The cost to fix a problem in a piece of software rises exponentially over time.
-A problem that might take a dollar to fix if you found it during requirements analysis might costs thousands to fix once the software is in production."
+A problem that might take a dollar to fix if you found it during requirements analysis might costs thousands to fix once the software is in production.
 
@@
 I would review and crosscheck my code.
-No way was I going to cost my employer $100,000.
+No way was I going to cost my employer \$100,000.~\cite{beck00}
+\end{quote}
+
+\begin{figure}
+ \includegraphics[width=3.3in]{beckB4.png}
+ \caption{Projects exhibiting phase delay. From~\cite{beck00}.}\label{fig:curve1}
+ \end{figure}
+\begin{figure}
+ \includegraphics[width=3.3in]{beckAFTER.png}
+ \caption{Projects without   phase delay. From~\cite{beck00}.}\label{fig:curve2}
+\end{figure}
+
+Beck speculated that there might be another kind of project, shown in \fig{curve2},
+that does no exhibit phase delay. In this next quote, Beck clearly states
+the importance of avoiding phase delay (in fact, for Beck, it is {\em the} major
+motivation for his work):
+\begin{quote}
+This (\fig{curve2}) is one of the premises of (agile).
+It is the technical premise of (agile).
+If the cost of change rose slowly over time, you would act completely differently from how you do under the assumption that costs rise exponentially.
+You would make big decisions as late in the process as possible, to defer the cost of making the decisions and to have the greatest possible chance that they would be right.
+You would only implement what you had to, in hopes that the needs you anticipate for tomorrow wouldn't come true.
+You would introduce elements to the design only as they simplified existing code or made writing the next bit of code simpler.~\cite{beck00}
+\end{quote}
+Beck makes one other comment that is relevant to this paper: 
+\begin{quote}
+The software development community has spent enormous resources in recent decades trying to reduce the cost of change-better languages, better database technology, better programming practices, better environments and tools, new notations.
+
+What would we do if all that investment paid off?
+What if all that work on languages and databases and whatnot actually got somewhere?
+What if the cost of change didn't rise exponentially over time?~\cite{beck00}
+\end{quote}
+The results of this paper are that, in our sample of hundreds of projects, there is no evidence for phase delay.
+This result is   consistent with Beck being correct: all that work on    languages and databases
+have paid off and, hence, modern software projects do not suffer from the  phase delay effect.
+
+XXX clumsy link here
+That said, where our results
+diverge from Beck is that we have observed this lack of phase delay in  projects that are not agile. Those
+projects are described below.
 
 \section{Related Work}
-
-In their text Romabach laws
-
-Kent Beck's original extreme programming text is big on this.
-
-Historical  evidence for this dataes back to some studies in the 1970s. To the best of
-our knowledge, not updated since. Regardless, as discussed in the next section, this
 
@@
 it does not seem to matter much how many phases we wait before fixing the issue.
-\begin{itemize}
+\item
 \item For Coding issues, the seems little impact of phase delay on the time
 required to fix.
-\end{itemize}
+\ei
 
@@
 
-\[
-\mathit{Reqts}
-\mathit{ReqtsReview}
-\mathit{ReqtsInspect}
-
-\mathit{HighLevelDesign}
-\mathit{HighLevelReview}
-\mathit{HighLevelInspect}
-
-\mathit{Design}
-\mathit{DesignReview}
-\mathit{DesignInspect}
-
-\mathit{Code}
-
-\mathit{Compile}
-\mathit{Test}
-\mathit{IntTest}
-\mathit{SysTest}
-\]
-
-In the original paper defining and 
-critiquing ``waterfall'', Royce defines a set of assessment and feedback processes
-such that
-
-care to contrast these 
-The majority of the teams follow a basically linear process within a single feature of physical component (usually no more than a week or two work in total).
-Unlike some agile teams, *most*TSP teams will not conflate development and unit test.
-This is usually pretty clear from the task  phase sequencing.
-A few teams choose to follow a Test Driven Development like process in which code review or inspection will follow test.
-In a more traditional “micro-linear” TSP approach, unit test follows inspections.
-
-Our data reporting the time required
-to resolve issues in phase $j$ that there were introduced in phase $i, i<j$.
-
 The defect typ that we studied were
 \be 
-\item design, compile, test, or other support system problems = Environment
-procedure calls and reference, I/O, user formats = Interface
-structure, content = Data
-comments, messages = Documentation
-spelling, punctuatio typos, instruction formats = Syntax
-logic, pointers, loops, recursion, computation, function defects = Function
-error messages, inadequate checks = Checking
-change management, library, version control = Build, package
-declaration, duplicate names, scope, limits = Assignment
-configuration, timing, memory = System
+\item Environment: design, compile, test, or other support system problems;
+\item Interface: procedure calls and reference, I/O, user format;
+\item Data: structure, content; 
+\item Documentation: comments, messages;
+\item Syntax: spelling, punctuation typos, instruction formats
+\item Function: logic, pointers, loops, recursion, computation, function defects  
+\item Checking: error messages, inadequate checks;
+\item Build: change management, library, version control;
+\item Assignment: package
+declaration, duplicate names, scope, limits;
+\item System: configuration, timing, memory.
 \ee
-\section{Conclusion}
 
@@
 \end{document}
+
+One way to address the absence of general laws is  {\em local learning}.
+Even if general laws
+do not exist, then {\em there exists general methods for finding the best local laws}.
+For example, one of us (Nichols) works extensively with software projects to help
+them tune their local software process in order to better fit their local needs.XXX.
+
+ Recent studies report that better predictors of the properties of
+software projects (effort and defects) can be generated by first clustering
+(a.k.a. stratifying or contextualizing or localizing) the data then learning
+different models for each cluster~\cite{posnett11,betten14,betta12,yang11,yang13,minku13}.
+Those {\em local learning} approach finds better predictions with lower variance than
+models learned from all the unclustered data~\cite{me12d,me11m}.
+
+The problem with the local learning is that they it can generate conclusions
+that conflict with strongly-held beliefs of the humans members of the software development team.
+It is an open issue how to handle those conflicts since it is unwise to
+always reject the conclusions of local learning  or the software developers.
+The local learners may be wrong due to incorrect assumptions by the data scientists who configured the learners~\cite{me11e,shull02}.
+On the other hand,  the humans may be wrong due to the cognitive biases described above.
+Experienced data scientists use a cyclic approach where
+(a)~the collected data; and (b)~the learning methods; and (c) the goals of the inquiry are matured  using
+insight offered by humans as they study the
+feedback generated by the learners~\cite{Fayyad96,me11e}.
+
+the learners' conclusions are discussed with the team in or
+
+(which may be erroneous
+due to incorrect assumptions
+
+We come to this work since  recent results on {\em locality} by Menzies and others
+
+XXXX In those 171 projects, we observed that it is
+ fastest to fix issues within the same phase $i$ as when they are generated. But in a result that contradicts
+ the phase delay effect, once an issue ``escapes'' phase $i$
+ it so no more expensive to resolve an issue in phase $i+1$ than
+ phase $i+2, i+3$, etc. Further, that increased effort may be quite modest.
+ That is, contrary to established wisdom, once an  issue ``escapes'' a phase, there is no
+ need to  rapidly retire that issue as soon as possible. 
+Our conclusion discusses the implications for our field, and how we might better
+propagated and monitor the wisdom of our field.

### 2015-02-26 data.tex
kinda v0.1 semi-coherent

--- 2015-02-25 phaseDelay.tex
@@
-\documentclass{sig-alternate}
-\usepackage{multirow}
-\usepackage{color}
-\usepackage{colortbl}
-\usepackage{picture}
-\usepackage{wrapfig}
-\usepackage{algorithm}
-\usepackage{graphicx}
-\usepackage{algorithmicx}
-\usepackage{algpseudocode}
-\renewcommand{\algorithmicrequire}{\textbf{Input:}}
-\renewcommand{\algorithmicensure}{\textbf{Output:}} 
-\newcommand{\lucas}[1]{\textcolor{red}{LUCAS: #1}} 
-\newcommand{\bill}[1]{\textcolor{blue}{BILL: #1}} 
-\newcommand{\carter}[1]{\textcolor{cyan}{CARTER: #1}} 
-%\newenvironment{changed}{\par}{\par}
+The section presents the data that we use to conclude that phase delay is a not a major
+contributor to the time required to resolve an issue.
 
-%timm tricks
-\newcommand{\bi}{\begin{itemize}[leftmargin=0.4cm]}
-\newcommand{\ei}{\end{itemize}}
-\newcommand{\be}{\begin{enumerate}}
-\newcommand{\ee}{\end{enumerate}}
-\newcommand{\tion}[1]{\S\ref{sect:#1}}
-\newcommand{\fig}[1]{Figure~\ref{fig:#1}}
-\newcommand{\eq}[1]{Equation~\ref{eq:#1}}
+The Software Engineering Institute (SEI) at
+Carnegie Mellon University explores methods for software process improvement.
+One product from that research is  
+Watts Humphrey's  Team Software ProcessSM (TSPSM)~\cite{tsp00}. TSP is an extension of Humphrey's early
+work on Personal Software
+ProcessSM (PSPSM)~\cite{psp05}.
 
-\usepackage[shortlabels]{enumitem} 
-\usepackage{times}
-\newcommand{\subparagraph}{}
-\usepackage{url}
-\def\baselinestretch{1}
+Since 2004, one  of use (Nichols) has been been mentoring 171 software development teams around the world
+as they  deploy TSP within their organizations.  
+The projects
+were mostly small to medium, with a median duration of 46 days
+and a maximum duration of 90 days. Median team size was 7
+people, with a maximum of 40. Hence,  total development 
+effort\footnote{
+Effort =  duration*teamSize/workDaysPerYear.} for
+these projects ranged from 1.3 years (median) to 14.3 years (max).
+The majority of
+the projects were either web portals or banking systems in the US, South Africa, and Mexico. 
+There were also some  medical devices (US, France, Japan, and Germany),  a few from a commercial 
+computer-aided design system (developed by a world wide distributed team), and less that  
+a dozen United States
+Department of Defence projects.
+ An anonymized version of that data is available in the PROMISE repository\footnote{\carter{Need to add}} and,
+ for confidentiality restrictions, we cannot offer 
+further details
+on those project.
 
-\setlist{nosep}
- \usepackage[font={small}]{caption, subfig}
-\setlength{\abovecaptionskip}{1ex}
- \setlength{\belowcaptionskip}{1ex}
+\bill{can we offer any more deatils here?}
 
- \setlength{\floatsep}{1ex}
- \setlength{\textfloatsep}{1ex}
-\usepackage[compact,small]{titlesec}
-\DeclareMathSizes{7}{7}{7}{7} 
-\pagenumbering{arabic}
-\setlength{\columnsep}{7mm}
+To aid in monitoring these 171 projects  Nichols  built a
+data collection tool that gathers project data a central database at SEI. 
+The issue types collected in the the SEI TSP data set divide into:
+\be 
+\item Environment: design, compile, test, or other support system problems;
+\item Interface: procedure calls and reference, I/O, user format;
+\item Data: structure, content; 
+\item Documentation: comments, messages;
+\item Syntax: spelling, punctuation typos, instruction formats
+\item Function: logic, pointers, loops, recursion, computation, function defects  
+\item Checking: error messages, inadequate checks;
+\item Build: change management, library, version control;
+\item Assignment: package
+declaration, duplicate names, scope, limits;
+\item System: configuration, timing, memory.
+\ee
+The tool also collected process information about those defects.
+That data includes  work start time, work end time, delta
+work time, and interruption time. Software engineers are often
+interrupted by meetings, requests for technical help, reporting, and
+so forth. These events are recorded, in minutes, as interruption
+time. In this paper, when we report ``time to resolve an
+issue'', we show the difference between the start and end times
+of a work session, with any interruption time subtracted (the
+difference in times, minus the interruptions).
 
-\begin{document}
+Three common features of all those  are (1)~coaches,
+(2)~peer review, and  (3)~personnel reviews.
 
-\conferenceinfo{FSE}{'15 Bergamo, Italy}
-\title{Revisiting the Truisms of Software Engineering:\\ Does Phase Delay Dramatically Increases  Repair Time?}
-\numberofauthors{3}
-\author{
-\alignauthor
-Tim Menzies, \\Carter Pape\\
-       \affaddr{CS, NcState, USA}\\
-       tim.menzies@gmail.com,\\carterpape@gmail.com
-\alignauthor
-William Nichols,\\ Forrest Schull\\
-\affaddr{SEI, CMU, USA}
-wrn,fjshull@sei.cmu.edu
-\alignauthor
-Lucas \\Layman\\
-       \affaddr{Fraunhofer Center,  USA}\\ 
-       llayman@fc-md.umd.edu
-}
+A ``coach'' is a team member  authorized to submit project data
+Before submission,
+these coaches check the data for obvious errors.
 
-\maketitle
-\begin{abstract}
-Does
-repair time increases dramatically
-the longer a defect persists in a system?
-This  is very widely belief and the core rational  for
- rejecting   traditional linear development methods
- in favor of more cyclic agile approaches.
-
-This paper shows that this belief is maintained
-quite strongly in the industrial software engineering
-community (but, in the academic community, somewhat less so).
-Yet based on a sample of 
-\carter{230} software projects from around the world from 
-\bill{2005 to 2013}, we argue that this belief is mostly 
-incorrect. Specifically: the median fix time for issues
-is just a few minutes and this increases by a small linear amount
-the longer the issue remains in the system.
-
-What was observed was a small number
-of defects that take a remarkably long time to resolve
-For these issues, if an issue takes more an hour
-to resolve then it will usually require many more hours of work.
-These long-tail bugs
-are more prevalent the longer a project delays issue removal. 
-In order to handle such ``long tail'' issues,
-we suggest  augmenting software
-development groups with additional ``tiger teams'' that are triggered
-when issue resolution time increases beyond a certain number of
-minutes.
-\end{abstract}
-
-% A category with the (minimum) three required fields
-\vspace{1mm}
-\noindent
-{\bf Categories/Subject Descriptors:} 
-D.2.8 [Software Engineering]: Product metrics;
-
-\vspace{1mm}
-\noindent
-{\bf Keywords:} defect prediction,
-
-\section{Introduction}
-Several recent papers call into questions long-held trusims in the field
-of software engineering. Devanbu reports that, in practice, the advantage of types languages
-is only very minimal. Nagappan et al. found that XXX, contrary to the advice of Dyketra,
-it is not necessarily true that ``goto considered harmful'' XXX. More generally,
-several recent  papers comment on ``locality'' effects where effects
-thold  generally across XXXX
-
-Perhaps it is time to take a look at the base premises of our field, particularly in light of the large software analytics data sets newly available for reseaercehrs.
-
-XXX tsp. 40MB . 235 prokects
-
-\section{Background}
-
-It is no understatement to say that 
-concerns with  phase delay have 
-a major  motivator for changing
-SE practices. Basili and Boehm~\cite{boehm01}  commented in
-2001 that, since the 1980s, the phase delay
-``has been a major driver in focusing
-industrial software practice on thorough
-requirements analysis and design,
-on early verification and validation, and
-on up-front prototyping and simulation
-to avoid costly downstream fixes.''
-
-\begin{center}
-\includegraphics[width=3.3in]{beckB4.png}
-
-\includegraphics[width=3.3in]{beckAFTER.png}
-\end{center}
-
-Also, concerns with the phase delay  had
- a significant impact on the history of agile programming. Kent Beck is
- particularly eloquent
-
-Under certain circumstances, the exponential rise in the cost of changing software over time can be flattened.
-If we can flatten the curve, old assumptions about the best way to develop software no longer hold.
-
-One of the universal assumptions of software engineering is that the cost of changing a program rises exponentially over time.
-I can remember sitting in a big linoleum-floored classroom as a college junior and seeing the professor draw on the board the curve found in Figure 1.
-
-The software development community has spent enormous resources in recent decades trying to reduce the cost of change—better languages, better database technology, better programming practices, better environments and tools, new notations.
-
-What would we do if all that investment paid off? What if all that work on languages and databases and whatnot actually got somewhere? What if the cost of change didn't rise exponentially over
-
-This is one of the premises of XP.
-It is the technical premise of XP.
-If the cost of change rose slowly over time, you would act completely differently from how you do under the assumption that costs rise exponentially.
-You would make big decisions as late in the process as possible, to defer the cost of making the decisions and to have the greatest possible chance that they would be right.
-You would only implement what you had to, in hopes that the needs you anticipate for tomorrow wouldn't come true.
-You would introduce elements to the design only as they simplified existing code or made writing the next bit of code simpler.
-￼
-time, but rose much more slowly, eventually reaching an asymptote? What if tomorrow's software engineering professor draws Figure 3 on the board?
-
-The cost to fix a problem in a piece of software rises exponentially over time.
-A problem that might take a dollar to fix if you found it during requirements analysis might costs thousands to fix once the software is in production."
-
-I resolved then and there that I would never let a problem get through to production.
-No sirree, I was going to catch problems as soon as possible.
-I would work out every possible problem in advance.
-I would review and crosscheck my code.
-No way was I going to cost my employer $100,000.
-
-\section{Related Work}
-
-In their text Romabach laws
-
-Kent Beck's original extreme programming text is big on this.
-
-Historical  evidence for this dataes back to some studies in the 1970s. To the best of
-our knowledge, not updated since. Regardless, as discussed in the next section, this
-
-\input{lucasRelatedWork}
-
-\section{lucasSurvey}
-
-\input{lucas}
-
-\section{billTSP}
-
-\section{carterCharts}
-
-\input{carter}
-
-\section{Data}
-
-\begin{figure}[!t]
-
-%\renewcommand{\baselinestretch}{0.8}
-\scriptsize
-\begin{center}
-\begin{tabular}{r|rrr|ll}
-  Sample&\multicolumn{3}{c|}{Percentiles}\\ 
-size & 25th & 50th & 75th & Phase injected & Phase removed\\\hline
-57& 6& 16& 47&  BeforeDevelopment&Code\\
-72& 2& 4& 11&  BeforeDevelopment&CodeInspect\\
-165& 6& 18& 56&  BeforeDevelopment&Test\\
-50& 6& 20& 37&  BeforeDevelopment&IntTest\\
-42& 8& 28& 65&  BeforeDevelopment&SysTest\\\hline
-
-66& 2& 6& 15&  Planning&Planning\\
-41& 1& 5& 16&  Planning&ReqtsReview\\\hline
-
-23& 0& 1& 3&  Reqts&Reqts\\
-245& 6& 13& 23&  Reqts&ReqtsReview\\
-289& 7& 16& 30&  Reqts&ReqtsInspect\\
-32& 6& 21& 40&  Reqts&Design\\
-49& 2& 7& 24&  Reqts&DesignInspect\\\hline
-
-94& 2& 5& 10&  HLD&HLDReview\\
-133& 3& 8& 19&  HLD&HLDInspect\\
-33& 2& 5& 17&  HLD&Design\\\hline
-
-37& 2& 4& 6&  Design&Design\\
-455& 11& 24& 53&  Design&DesignInspect\\
-218& 5& 14& 30&  Design&Code\\
-166& 4& 11& 22&  Design&CodeInspect\\
-542& 14& 31& 60&  Design&Test\\
-67& 6& 15& 40&  Design&IntTest\\
-92& 10& 25& 53&  Design&SysTest\\\hline
-
-53& 2& 4& 9&  DesignInspect&DesignInspect\\
-38& 1& 3& 9&  DesignInspect&Code\\\hline
-
-126& 4& 12& 31&  Code&Code\\
-459& 11& 24& 45&  Code&CodeInspect\\
-461& 11& 26& 55&  Code&Test\\
-348& 10& 26& 62&  Code&IntTest\\
-230& 8& 18& 39&  Code&SysTest\\
-71& 2& 11& 28&  Code&AcceptTest\\\hline
-
-64& 2& 4& 10&  CodeInspect&CodeInspect\\
-57& 2& 5& 13&  CodeInspect&Test\\\hline
-
-110& 3& 10& 23&  Test&Test\\
-66& 2& 4& 12&  Test&QualTest\\\hline
-
-32& 1& 2& 4&  IntTest&QualTest\\
-36& 1& 11& 30&  IntTest&IntTest\\
-21& 1& 2& 5&  IntTest&SysTest\\
- \end{tabular}
-\end{center}
-\caption{Distribution of fix times seen in SEI TSP data.}
-\label{fig:faw}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\begin{tabular}{rrl}
-year& \# issues&\\\hline
-2006 &  44 &\\
-2007 &  34 &\\
-2008&  288 &\rule{1mm}{2mm}\\
-2009&  846 &\rule{3mm}{2mm}\\
-2010& 1007 &\rule{3mm}{2mm}\\
-2011& 3273 &\rule{10mm}{2mm}\\
-2012&18102 &\rule{45mm}{2mm}\\
-2013&20336 &\rule{50mm}{2mm}\\
-2014& 3307 & \rule{10mm}{2mm}\\\cline{1-2}
-Total:&47228
-\end{tabular}
-\end{center}
-\caption{This paper studies 47,228 issues recorded 2006 to 2014.}\label{fig:years}
-\end{figure}
-\begin{figure*}[!t]
-\begin{center}
-\includegraphics[width=6in]{waterfall2.png}
-\end{center}
-\caption{Different approaches to software development:  waterfall and agile (bottom left).}
-\end{figure*}
-
-\begin{figure*}[!t]
-
-\renewcommand{\baselinestretch}{0.7} 
-\begin{center}
-\begin{tabular}{ll|r|rl}
-            &                  & Sample\\
-Phase injected & Phase removed & size & \multicolumn{2}{l}{Scale up, w.r.t. to first phase}
-\input{deltas}
-\end{tabular}
-\end{center}
-\caption{50th percentile (median) scale ups  for  time to resolve issues (taken from \fig{raw}).}
-\label{fig:scale}
-\end{figure*}
-
-In these results, we checked if a small number of bugs are most expensive. 
-To check this:
-\bi
-\item
-We repeated the analysis that generated \fig{scale},
-but instead of looking at the 50th percentile, we displayed the scale up factors
-\item 
-We only
-checked the Design and Code scale up results since, from \fig{scale}, it is clear these
-have the most examples of longest phase delays.
-\ei 
-The  results are shown in \fig{scale90} and these
-results are somewhat different for Design and Coding issues:
-\bi 
-\item For Design issues,  these have the same
-general form as the 50th percentile results. That is, while it it certainly faster
-to remove thing sin the phase where they are created, once we leave that phase
-it does not seem to matter much how many phases we wait before fixing the issue.
-\begin{itemize}
-\item For Coding issues, the seems little impact of phase delay on the time
-required to fix.
-\end{itemize}
-
-\begin{figure*}[!t]
-
-\renewcommand{\baselinestretch}{0.7} 
-\begin{center}
-\begin{tabular}{ll|r|rl}
-            &                  & Sample\\
-Phase injected & Phase removed & size & \multicolumn{2}{l}{Scale up, w.r.t. to first phase}
-\input{deltaBig}
-\end{tabular}
-\end{center}
-\caption{90th percentile scale ups  for  time to resolve issues .}
-\label{fig:scale90}
-\end{figure*}
-
-The data used in this work shirai14
-
-The Team Software ProcessSM (TSPSM) provides a
-framework that teams use to collect software process data in real
-time, using a defined disciplined process.
-In particular, three common features of the projects in our SEI TSP data set  are coaches,
-peer review, and  personnel reviews:
-\bi 
-\item
-A ``coach'' is the team member  authorized to submit project data
-(before submission,
-these coaches check the data for obvious errors);
-\item 
 ``Personnel review'' is a technique taken  from the Personal Software
-ProcessSM (PSPSM);
-\item 
-``Peer review'' is a standard technique in
-traditional software engineering.
-\ei
-PSPSM encourages developers to continually make and review their personnel estimates
+ProcessSM (PSPSM). PSPSM encourages developers to continually make and review their personnel estimates
 about their day-to-day tasks, then compare those estimates against the actual development effort.
 In this way, developers can acquire a more realistic understanding of their work behaviour.
-As to peer review,  Basili and Boehm write  commented in 2001~\cite{boehm01} 
+
+``Peer review'' is a  technique in
+traditional software engineering.
+ Basili and Boehm write  commented in 2001~\cite{boehm01} 
 that peer reviews can catch over half the defects introduced into a system.
@@
 
-\begin{figure}
-\begin{center}
-\scriptsize\begin{tabular}{|l@{~:~}l|}\hline
-Bug Prediction Dataset &http://bug.inf.usi.ch \\
-Eclipse Bug Data &http://goo.gl/tYKahN \\
-FLOSSMetrics& http://flossmetrics.org \\
-FLOSSMole &http://flossmole.org \\
-IBSBSG& http://www.isbsg.org \\
-ohloh& http://www.openhub.net \\
-PROMISE &http://promisedata.googlecode.com \\
-Qualitas Corpus &http://qualitascorpus.com \\
-Software Artifact Repository &http://sir.unl.edu \\
-SourceForge Research Data &http://zerlot.cse.nd.edu \\
-Sourcerer Project &http://sourcerer.ics.uci.edu \\
-Tukutuku &http://www.metriq.biz/tukutuku \\
-Ultimate Debian Database &http://udd.debian.org\\\hline
-\end{tabular}
-\end{center}
-\caption{Some repositories of software engineering data.}\label{fig:sedata}
-\end{figure}
-The data collected in the SEI TSP databases contains 
-extensive process details
-including a detailed phased breakdown showing what happened at what
-phases of the lifecycle. This makes this data somewhat different to the standard
- publicly available
-software projects data sets currently available to 
-SE researchers. Most of the \fig{sedata} data sets have software product information
-such as full source code, or summaries of static features.
-A subset of that data contain issue or defect reports and/or
-the time taken to build these systems.  We are unaware
-of any of these having detailed phase-by-phase breakdowns of project data.
+As of January 2014, the SEI TSP database contained data from 109
+TSP projects. The projects started between July 2009 and
+September 2013; they included 34 teams and 309 people. Among
+the database fact tables, the time store contains 103,023 time logs,
+18,408 defect logs, and 7,464 size logs\footnote{\bill{we need to 
+update this table}}.
 
+%That said, certain semantic features of the SEI TSP data should be noted.
+%Firstly, in the current TSP collection tool, 
+%fix times are only the developer time for the developer walking through the phases of \fig{waterfall}.
+%We are currently tracking the fix time for post-release issues (e.g. those raised during  acceptance test and %later
+%product life cycle). So far, in that post-release data,  we have not detected
+%a dramatic phase escalation effect (but at this time, we have nothing definitive comment on that matter).%
+%
+%\bill{somewhere you have one note on \underline{find} and fix times.  for this paper, we need just fix times. %but is there
+%anything we need to fret about re \underline{find} times?}
+
+\subsection{Project Phases}
+This paper studies the impact of phase delay on the time required to resolve issues.
 The phased used in this paper are shown in \fig{waterfall}. Note that, in that figure:
@@
 \ei
-
-\section{Data}
-
-The Software Engineering Institute (SEI) at
-Carnegie Mellon University has collected data from organizations
-that have adopted TSP. 
-Our efforts focused on data collected by projects that launched
-after 2009 and used an automatic data recording tool. 
-Due to confidentiality restrictions, we cannot offer too many details
-on those project. That said, there were less than a dozen United States
-Department of Defence projects. Also, the majority of
-the projects were weither web portals or banking systems in the US, South Africa, and Mexico. 
-There were also some  medical devices (US, France, Japan, and Germany) and a few from a commercial 
-computer-aided design system (developed by a world wide distributed team).
-
-Data collected included time logs, defect
-logs, added and modified size logs, and work
-breakdown strcture logs.
-The projects
-were mostly small to medium, with a median duration of 46 days
-and a maximum duration of 90 days. Median team size was 7
-people, with a maximum of 40. Hence,  total development 
-effort\footnote{
-Effort =  duration*teamSize/workDaysPerYear.} for
-these projects ranged from 1.3 years (median) to 14.3 years (max).
-
-In these logs
-the log records include work start time, work end time, delta
-work time, and interruption time. Software engineers are often
-interrupted by meetings, requests for technical help, reporting, and
-so forth. These events are recorded, in minutes, as interruption
-time. In this paper, when we report ``time to resolve an
-issue'', we show the difference between the start and end times
-of a work session, with any interruption time subtracted (the
-difference in times, minus the interruptions).
-
-As of January 2014, the SEI TSP database contained data from 109
-TSP projects. The projects started between July 2009 and
-September 2013; they included 34 teams and 309 people. Among
-the database fact tables, the time store contains 103,023 time logs,
-18,408 defect logs, and 7,464 size logs\footnote{\bill{we need to 
-update this table}}.
-
-A common property of real-world data sets is the presence
-of noisy entries (superfluous  or spurious data). 
-The level of noise can be quite high. As reported
-in \cite{shepperd12}, around
-10\% to 30\%
-of the records in the NASA MDP defect data sets are
-affected by noise. Nichols et al.~\cite{shirai14}  report that
-the noise levesl in the SEI TSE data are smaller than those seen
-in other data sets. They found in the SEI TSP data that:\bi 
-\item
-4\% of the data was incorrect such as  null values of illegal formats;
-\item  2\% of the data has inconsistencies such as timestamps
-where the stop time was before the start time;
-\item 3\% of the data had data that was not credible
-such as tasks listed in one day that took like than six hours.
-\ei 
-That said, certain semantic features of the SEI TSP data should be noted.
-Firstly, in the current TSP collection tool, 
-fix times are only the developer time for the developer walking through the phases of \fig{waterfall}.
-We are currently tracking the fix time for post-release issues (e.g. those raised during  acceptance test and later
-product life cycle). So far, in that post-release data,  we have not detected
-a dramatic phase escalation effect (but at this time, we have nothing definitive comment on that matter).
-
-\bill{somewhere you have one note on \underline{find} and fix times.  for this paper, we need just fix times. but is there
-anything we need to fret about re \underline{find} times?}
-
-The majority of the teams follow a more linear process within a single feature of physical component (usually no more than a week or two work in total).
-Unlike some agile teams, *most*TSP teams will not conflate development and unit test.
-
-The starting point for this paper were the times required to fix 47,228 issues.
-All data was collected
-by the Software Engineering Institute in the period 2006 to 2014 (see \fig{years}).
-To the best of our knowledge, this is the largest sample of ``time to fix issues''
-yet analyzed.
-
-This data was collected using the   methodology described by
-Watts Humphrey's Team Software Process~\cite{tsp00}. TSP is an extension of Humphrey's early
-work on Personnel Software Process~\cite{psp05}.  PSP encourages developers to continually make estimates
-about their day-to-day tasks, then compare those estimates against the actual development effort
-(in this way, developers can acquire a more realistic understanding of their work behaviour).
-
-In all data, was collected from 172 separate projects. Given Beck's comments (above) about phase delay motivating agile processes, we take care to identify the differences of these 
-projects to agile development projects. An alternate, more traditional approach is a ``macro linear''
-approach such as the ``waterfall methodology'' that develops code after an extensive
-pre-planning stage:
-
-\[
-\mathit{Reqts}
-\mathit{ReqtsReview}
-\mathit{ReqtsInspect}
-
-\mathit{HighLevelDesign}
-\mathit{HighLevelReview}
-\mathit{HighLevelInspect}
-
-\mathit{Design}
-\mathit{DesignReview}
-\mathit{DesignInspect}
-
-\mathit{Code}
-
-\mathit{Compile}
-\mathit{Test}
-\mathit{IntTest}
-\mathit{SysTest}
-\]
-
-In the original paper defining and 
-critiquing ``waterfall'', Royce defines a set of assessment and feedback processes
-such that
-
-care to contrast these 
-The majority of the teams follow a basically linear process within a single feature of physical component (usually no more than a week or two work in total).
-Unlike some agile teams, *most*TSP teams will not conflate development and unit test.
-This is usually pretty clear from the task  phase sequencing.
-A few teams choose to follow a Test Driven Development like process in which code review or inspection will follow test.
-In a more traditional “micro-linear” TSP approach, unit test follows inspections.
-
-Our data reporting the time required
-to resolve issues in phase $j$ that there were introduced in phase $i, i<j$.
-
-The defect typ that we studied were
-\be 
-\item design, compile, test, or other support system problems = Environment
-procedure calls and reference, I/O, user formats = Interface
-structure, content = Data
-comments, messages = Documentation
-spelling, punctuatio typos, instruction formats = Syntax
-logic, pointers, loops, recursion, computation, function defects = Function
-error messages, inadequate checks = Checking
-change management, library, version control = Build, package
-declaration, duplicate names, scope, limits = Assignment
-configuration, timing, memory = System
-\ee
-\section{Conclusion}
-
-\section*{Acknowledgements}
-This work was partially funded by an National Science
-Foundation grant NSF-CISE 1302169.
-
-\clearpage
-\vspace*{0.5mm}
-\scriptsize
-\bibliographystyle{unsrt}
-\bibliography{refs}
-
-\end{document}

### 2015-02-26 phaseDelay.tex
kinda v0.1 semi-coherent

@@
 
+\begin{figure*}[!t]
+\begin{center}
+\includegraphics[width=6in]{waterfall2.png}
+\end{center}
+\caption{Different approaches to software development:  waterfall and agile (bottom left).}
+\end{figure*}
+
 \section{Data}
+
+\input{data}
 
@@
 
-\begin{figure}
-\begin{center}
-\begin{tabular}{rrl}
-year& \# issues&\\\hline
-2006 &  44 &\\
-2007 &  34 &\\
-2008&  288 &\rule{1mm}{2mm}\\
-2009&  846 &\rule{3mm}{2mm}\\
-2010& 1007 &\rule{3mm}{2mm}\\
-2011& 3273 &\rule{10mm}{2mm}\\
-2012&18102 &\rule{45mm}{2mm}\\
-2013&20336 &\rule{50mm}{2mm}\\
-2014& 3307 & \rule{10mm}{2mm}\\\cline{1-2}
-Total:&47228
-\end{tabular}
-\end{center}
-\caption{This paper studies 47,228 issues recorded 2006 to 2014.}\label{fig:years}
-\end{figure}
-\begin{figure*}[!t]
-\begin{center}
-\includegraphics[width=6in]{waterfall2.png}
-\end{center}
-\caption{Different approaches to software development:  waterfall and agile (bottom left).}
-\end{figure*}
-
 \begin{figure*}[!t]
@@
 
+\section{Phase Delay in Slower Bugs?}
 In these results, we checked if a small number of bugs are most expensive. 
@@
 
-The data used in this work shirai14
+\section{Validity} 
+A previous paper~\cite{shirai14} has applied a range of sanity checks to the data.
+A common property of real-world data sets is the presence
+of noisy entries (superfluous  or spurious data). 
+The level of noise can be quite high. As reported
+in \cite{shepperd12}, around
+10\% to 30\%
+of the records in the NASA MDP defect data sets are
+affected by noise. Nichols et al.~\cite{shirai14}  report that
+the noise levesl in the SEI TSE data are smaller than those seen
+in other data sets. They found in the SEI TSP data that:\bi 
+\item
+4\% of the data was incorrect such as  null values of illegal formats;
+\item  2\% of the data has inconsistencies such as timestamps
+where the stop time was before the start time;
+\item 3\% of the data had data that was not credible
+such as tasks listed in one day that took like than six hours.
+\ei
 
-The Team Software ProcessSM (TSPSM) provides a
-framework that teams use to collect software process data in real
-time, using a defined disciplined process.
-In particular, three common features of the projects in our SEI TSP data set  are coaches,
-peer review, and  personnel reviews:
-\bi 
-\item
-A ``coach'' is the team member  authorized to submit project data
-(before submission,
-these coaches check the data for obvious errors);
-\item 
-``Personnel review'' is a technique taken  from the Personal Software
-ProcessSM (PSPSM);
-\item 
-``Peer review'' is a standard technique in
-traditional software engineering.
-\ei
-PSPSM encourages developers to continually make and review their personnel estimates
-about their day-to-day tasks, then compare those estimates against the actual development effort.
-In this way, developers can acquire a more realistic understanding of their work behaviour.
-As to peer review,  Basili and Boehm write  commented in 2001~\cite{boehm01} 
-that peer reviews can catch over half the defects introduced into a system.
-Peer review can be conducted on any artifact generated anywhere in the software
-lifecycle and can quickly be adapted to new kinds of artifacts.
+\section{Related Work}
 
@@
 of any of these having detailed phase-by-phase breakdowns of project data.
-
-The phased used in this paper are shown in \fig{waterfall}. Note that, in that figure:
-\bi 
-\item
-Several  phases have the same  sub-activities of {\em review} and {\em inspect}\footnote{\bill{plz distinguish review and inspect. does one usually happen first?}}
-\item Testing is divided into several statges \footnote{\bill{need a one
-line description of test vs qualtest vs inttest vs systemTest vs AcceptTest. I tries some words in the Key to \fig{sedata}. dont know if i got it right.}}
-\ei 
-Also shown in \fig{waterfall} (bottom left) is the standard definition of an agile process~\cite{boehmturner03}:
-Given some
-some backlog of tasks, wgeb teams complete their current tasks, they select the next task(s) to complete. That selection process may use a variety of criteria
-to prioritize which  tasks are selected (for more details on that selection process, see~\cite{me09j,port08,boehmturner03}). Tasks are completed in ``sprints'' that can last hours,
-days, but rarely not more than weeks. Each day meet for brief ``scrum'' sessions to assess (and possibly alter) their current progress on the goals of the sprint.  
-Agile teams race to generate releases
-(in the continuous release model, releases can be generated on a daily basis, or even faster).  
-Experience gained from those releases informs the discussion in the daily scrums which, in turn,
-can inform the team's decisions on how to select and implement the next set of tasks for next sprint.
-In fact, that experience can result in changing some/all of the tasks in the current backlog.
-
-TSP is not antithetical to agile--  indeed a TSP-waterfall style project can adopt aspects
-of agile.  For example the agile loop  could be applied
-over one or more of the phases shown in the long waterfall chain of \fig{waterfall}. Some
-TSP teams adopt something like test-driven-development\footnote{TDD is an agile-method
-where the ``test'' is the primary driver of the design. The tests are written first,
-then the code to support those tests. TDD proceeds in three steps: red (where there
-are broken tests); green (where tests are passing); and, possibly, refactor (where
-the code is re-organized based on feedback from those tests~\cite{fraser03}.} where
-reviews are scheduled after testing. Some of the groups in the SEI TSP data used that approach but
-they do not effect the main conclusions of this paper:
-\bi 
-\item They were strongly in the minority\bill{any numbers on this?};
-\item Most of our phase delay data comes from much early in the waterfall model
-of \fig{waterfall}.
-\ei 
-That said, there are some very ``un-agile'' aspects
-of the processes used by the   projects in the TSP SEI data.
-A TSP participant spends much time reflecting on the project,
-undisturbed by the behaviour of the  executables.
-In fact,
-it can be days/weeks
-before TSP participants gain  feedback from executing code, for the following reasons:  
-\bi 
-\item
-For TSP projects, the times spent in the of a design activity/phase is   
-approximately as much effort as the coding phase.  
-\item Our TSP teams did not
- combine development and  testing. 
- \item  The TSP projects studied here offer new releases at least every three months, but
-often much longer that that\bill{any numbers on that}. Note that this is a far slower
-release schedule than seen in, say, continuous integration projects.  
-\item The time devoted to personnel review'' and peer review (defined above)
-is  about 50\% as much effort as the previous construction phase activity.
-
-\ei
-
-Having presented all that, we can now present the main point of this section:
-\bi 
-\item The results section of this paper fails to find that  phase delay causing dramatic increases
-in time to fix issues;
-\item This lack-of-phase-delay effect {\em cannot} be explained away just by  saying that
-the projects in this sample adopted something like  agile methods to reduce re-work costs.
-\ei
-
-\section{Data}
-
-The Software Engineering Institute (SEI) at
-Carnegie Mellon University has collected data from organizations
-that have adopted TSP. 
-Our efforts focused on data collected by projects that launched
-after 2009 and used an automatic data recording tool. 
-Due to confidentiality restrictions, we cannot offer too many details
-on those project. That said, there were less than a dozen United States
-Department of Defence projects. Also, the majority of
-the projects were weither web portals or banking systems in the US, South Africa, and Mexico. 
-There were also some  medical devices (US, France, Japan, and Germany) and a few from a commercial 
-computer-aided design system (developed by a world wide distributed team).
-
-Data collected included time logs, defect
-logs, added and modified size logs, and work
-breakdown strcture logs.
-The projects
-were mostly small to medium, with a median duration of 46 days
-and a maximum duration of 90 days. Median team size was 7
-people, with a maximum of 40. Hence,  total development 
-effort\footnote{
-Effort =  duration*teamSize/workDaysPerYear.} for
-these projects ranged from 1.3 years (median) to 14.3 years (max).
-
-In these logs
-the log records include work start time, work end time, delta
-work time, and interruption time. Software engineers are often
-interrupted by meetings, requests for technical help, reporting, and
-so forth. These events are recorded, in minutes, as interruption
-time. In this paper, when we report ``time to resolve an
-issue'', we show the difference between the start and end times
-of a work session, with any interruption time subtracted (the
-difference in times, minus the interruptions).
-
-As of January 2014, the SEI TSP database contained data from 109
-TSP projects. The projects started between July 2009 and
-September 2013; they included 34 teams and 309 people. Among
-the database fact tables, the time store contains 103,023 time logs,
-18,408 defect logs, and 7,464 size logs\footnote{\bill{we need to 
-update this table}}.
-
-A common property of real-world data sets is the presence
-of noisy entries (superfluous  or spurious data). 
-The level of noise can be quite high. As reported
-in \cite{shepperd12}, around
-10\% to 30\%
-of the records in the NASA MDP defect data sets are
-affected by noise. Nichols et al.~\cite{shirai14}  report that
-the noise levesl in the SEI TSE data are smaller than those seen
-in other data sets. They found in the SEI TSP data that:\bi 
-\item
-4\% of the data was incorrect such as  null values of illegal formats;
-\item  2\% of the data has inconsistencies such as timestamps
-where the stop time was before the start time;
-\item 3\% of the data had data that was not credible
-such as tasks listed in one day that took like than six hours.
-\ei 
-That said, certain semantic features of the SEI TSP data should be noted.
-Firstly, in the current TSP collection tool, 
-fix times are only the developer time for the developer walking through the phases of \fig{waterfall}.
-We are currently tracking the fix time for post-release issues (e.g. those raised during  acceptance test and later
-product life cycle). So far, in that post-release data,  we have not detected
-a dramatic phase escalation effect (but at this time, we have nothing definitive comment on that matter).
-
-\bill{somewhere you have one note on \underline{find} and fix times.  for this paper, we need just fix times. but is there
-anything we need to fret about re \underline{find} times?}
-
-The majority of the teams follow a more linear process within a single feature of physical component (usually no more than a week or two work in total).
-Unlike some agile teams, *most*TSP teams will not conflate development and unit test.
-
-The starting point for this paper were the times required to fix 47,228 issues.
-All data was collected
-by the Software Engineering Institute in the period 2006 to 2014 (see \fig{years}).
-To the best of our knowledge, this is the largest sample of ``time to fix issues''
-yet analyzed.
-
-This data was collected using the   methodology described by
-Watts Humphrey's Team Software Process~\cite{tsp00}. TSP is an extension of Humphrey's early
-work on Personnel Software Process~\cite{psp05}.  PSP encourages developers to continually make estimates
-about their day-to-day tasks, then compare those estimates against the actual development effort
-(in this way, developers can acquire a more realistic understanding of their work behaviour).
-
-In all data, was collected from 172 separate projects. Given Beck's comments (above) about phase delay motivating agile processes, we take care to identify the differences of these 
-projects to agile development projects. An alternate, more traditional approach is a ``macro linear''
-approach such as the ``waterfall methodology'' that develops code after an extensive
-pre-planning stage:
-
-The defect typ that we studied were
-\be 
-\item Environment: design, compile, test, or other support system problems;
-\item Interface: procedure calls and reference, I/O, user format;
-\item Data: structure, content; 
-\item Documentation: comments, messages;
-\item Syntax: spelling, punctuation typos, instruction formats
-\item Function: logic, pointers, loops, recursion, computation, function defects  
-\item Checking: error messages, inadequate checks;
-\item Build: change management, library, version control;
-\item Assignment: package
-declaration, duplicate names, scope, limits;
-\item System: configuration, timing, memory.
-\ee
 
@@
 propagated and monitor the wisdom of our field.
+
+\begin{figure}
+\begin{center}
+\begin{tabular}{rrl}
+year& \# issues&\\\hline
+2006 &  44 &\\
+2007 &  34 &\\
+2008&  288 &\rule{1mm}{2mm}\\
+2009&  846 &\rule{3mm}{2mm}\\
+2010& 1007 &\rule{3mm}{2mm}\\
+2011& 3273 &\rule{10mm}{2mm}\\
+2012&18102 &\rule{45mm}{2mm}\\
+2013&20336 &\rule{50mm}{2mm}\\
+2014& 3307 & \rule{10mm}{2mm}\\\cline{1-2}
+Total:&47228
+\end{tabular}
+\end{center}
+\caption{This paper studies 47,228 issues recorded 2006 to 2014.}\label{fig:years}
+\end{figure}

### 2015-03-02 lucasRelatedWork.tex
Added Related Work section.

The first data on the cost to fix defects as a function of lifecycle phase date back to large systems in the late 70s from IBM~\cite{Fagan76}, TRW~\cite{Boehm76}, GTE~\cite{Daly77}, and Bell Labs~\cite{Stephenson76} (Figure~\ref{fig:cost-to-fix}). These studies suggest that the cost (in terms of effort) to find and fix an error monotonically increases with lifecycle phase. For large projects, the ratio of cost-to-fix in the requirements phase to cost-to-fix in operation is on the order of 1:100. Furthermore, cost function is superlinear, with the greatest rates of increase in the acceptance testing and operations phases.

\begin{figure}
 \includegraphics[width=3.3in]{boehm_cost-to-fix.png}
 \caption{Cost-to-fix curve. From~\cite{Boehm81}.}\label{fig:cost-to-fix}
 \end{figure}

In the 40 years since these initial studies, few studies have been published on the cost-to-fix curve as a function of lifecycle phase. Boehm~\cite{Boehm80} provides data suggesting that the cost-to-fix curve for small projects (from two student projects of 2000 deliverable source instructions) is flatter than for large projects (the dashed line of Figure~\ref{fig:cost-to-fix}). Shull et al.~\cite{Shull02} conducted a literature survey and held a series of e-workshops with industry experts on fighting defects. Workshop participants from Toshiba and IBM reported cost-to-fix ratios of 1:137 and 1:117 for large projects respectively~\cite{Shull02} -- the raw data points are not provided. One notable example to the traditional cost-to-fix curve is the CCPDS-R described by Royce~\cite{Royce98}: a million-line, safety-critical missile defense system (Figure~\ref{fig:royce}). In this project, design changes (including architecture changes) required approximately twice the effort of implementation and test changes, and the cost-to-fix in implementation and test phases increased slowly. Boehm~\cite{Boehm10} attributes this success to the CCPDS-R development process, which focused on removing architecture risk early in the devleopment lifecycle.

\begin{figure}
 \includegraphics[width=3.3in]{Royce98.PNG}
 \caption{Exception to the rule - CCPDS-R cost-to-fix curve. From~\cite{Royce98}.}\label{fig:royce}
 \end{figure}

As discussed earlier, one goal of agile methods is to flatten the cost-to-change curve~\cite{beck00}. Relatively little empirical data exists on this point. Clutterbuck et al.~\cite{Clutterbuck09} studied 5-month effort by a small-to-medium enterprise team developing a 71KLOEC web interface to a database application to implement 18 change requests (Figure~\ref{fig:clutterbuck}). Note that these were for new and changed user requirements, not defects. Clutterbuck et al. found the cost of change to be relatively flat until the later phases, with much of the effort spent in analysis of the change requests~\cite{Clutterbuck09}. Elssamadisy and Schalliol~\cite{Elssamadisy02} anecdotally report on the growing, high cost of rework in a 50 person, three-year, 500KLOEC Extreme Programming project as the project grew in size and complexity.

\begin{figure}
 \includegraphics[width=3.3in]{clutterbuck.png}
 \caption{Cost of change from an agile case study~\cite{Clutterbuck09}.}\label{fig:clutterbuck}
 \end{figure}

 We note that previous work focuses on cost-to-fix as a function of lifecycle phase irrespective of when the defect was injected, that is, previous work analyzes the cost to fix a defect found in test regardless of whether that defect was a requirements error or a coding mistake. To our knowledge, our research represents the first large-scale study of phase delay as a contributor to the cost-to-fix curve.

### 2015-03-02 phaseDelay.tex
working some extremes
Added Related Work section.

@@
 
-%\renewcommand{\baselinestretch}{0.8}
+\renewcommand{\baselinestretch}{0.7}
 \scriptsize
@@
 
-\begin{figure*}[!t]
-
+\begin{figure}[!t]
 \renewcommand{\baselinestretch}{0.7}
+\scriptsize
 \begin{center}
-\begin{tabular}{ll|r|rl}
-            &                  & Sample\\
-Phase injected & Phase removed & size & \multicolumn{2}{l}{Scale up, w.r.t. to first phase}
+\begin{tabular}{l@{~~}|l@{~}|r@{~}|r@{~}r@{~}|r@{~}l}
+           \multicolumn{2}{c}{~}                 &  &\multicolumn{2}{c|}{median}\\
+  Injection&   Removal& $n$ & initial & now & \multicolumn{2}{l}{Scale up, w.r.t. initial}
 \input{deltas}
@@
 \label{fig:scale}
-\end{figure*}
+\end{figure}
 
@@
 
-\begin{figure*}[!t]
-
+\begin{figure}[!t]
 \renewcommand{\baselinestretch}{0.7}
+\scriptsize
 \begin{center}
-\begin{tabular}{ll|r|rl}
-            &                  & Sample\\
-Phase injected & Phase removed & size & \multicolumn{2}{l}{Scale up, w.r.t. to first phase}
+\begin{tabular}{l@{~~}|l@{~}|r@{~}|r@{~}r@{~}|r@{~}l}
+           \multicolumn{2}{c}{~}                 &  &\multicolumn{2}{c|}{90th }\\
+           \multicolumn{2}{c}{~}                 &  &\multicolumn{2}{c|}{percentile }\\
+  Injection&   Removal& $n$ & initial & now & \multicolumn{2}{l}{Scale up, w.r.t. initial}
 \input{deltaBig}
@@
 \end{center}
-\caption{90th percentile scale ups  for  time to resolve issues .}
+\caption{90th percentile scale ups.}
 \label{fig:scale90}
-\end{figure*}
+\end{figure}
+
+\begin{figure}[!t]
+\renewcommand{\baselinestretch}{0.7}
+\scriptsize
+\begin{center}
+\begin{tabular}{l@{~~}|l@{~}|r@{~}|r@{~}r@{~}|r@{~}l}
+          \multicolumn{2}{c}{~}                 &  &\multicolumn{2}{c|}{99th }\\
+           \multicolumn{2}{c}{~}                 &  &\multicolumn{2}{c|}{percentile }\\
+  Injection&   Removal& $n$ & initial & now & \multicolumn{2}{l}{Scale up, w.r.t. initial}
+\input{deltasBigBig}
+\end{tabular}
+\end{center}
+\caption{99th percentile scale ups.}
+\label{fig:scale90}
+\end{figure}
+
+\begin{figure}[!t]
+\renewcommand{\baselinestretch}{0.7}
+\scriptsize
+\begin{center}
+\begin{tabular}{l@{~~}|l@{~}|r@{~}|r@{~}r@{~}|r@{~}l}
+           \multicolumn{2}{c}{~}                 &  &\multicolumn{2}{c|}{mean}\\
+  Injection&   Removal& $n$ & initial & now & \multicolumn{2}{l}{Scale up, w.r.t. initial}
+\input{deltasMean}
+\end{tabular}
+\end{center}
+\caption{Mean  scale ups.}
+\label{fig:scale90}
+\end{figure}
 
@@
 \ei
+
+XXX post release. really hard to change? auto-updates, package managers, web-based apps.
 

### 2015-03-03 lucas.tex
Added Survey section.

As part of a related research, we conducted a survey of software engineering practitioners and researchers on their views of commonly-head software engineering ''laws''. One of the laws questioned was drawn from from Glass~\cite{glass02} who cites Boehm \lucas{ref needed} as the source: ''Requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' (RqtsErr for short). The survey was conducted in two phases using Amazon's Mechanical Turk. The first phase was conducted only with professional software engineers, and the second survey was conducted only with Program Committee (PC) members of the ESEC/FSE and ICSE conferences.

The respondents answered the following questions for the laws they were presented: 
\textbf{Agreement:} ''Based on your experience, do you agree that the statement above is correct?'' A Likert scale captured the agreement score from Strongly Disagree to Strongly Agree. A text box was provided to explain the answer.
\textbf{Applicability:} ''To the extent that you believe it, how widely do you think it applies among software development contexts?'' The possible answers were presented as a scale from -1 to 5:
\begin{itemize}
\item I don't know (-1)
\item This law does not apply at all (0)
\item Applies in a very narrow range of projects  (1)
\item Rarely applies (2)
\item Occasionally applies (3)
\item Very frequently applies (4)
\item Always applies (5)
\end{itemize}
Respondents were required to explain the applicability score in a text box.

In the ''law evaluation'' portion of the survey, participants were presented with the RqtsErr law and others drawn from \cite{glass02} and \cite{endres03}. The PC member survey contained an additional question: ''Do you agree with the following? 'In general, the longer errors are in the system (requirements errors, design errors, coding errors, etc.), the more expensive they are to fix''' (PhaseDelay for short). Responses were recorded using the \textbf{Agreement} question Likert scale.

Summary statistics for the agreement and applicability scores for the RqtsErr and PhaseDelay laws are presented in Figure~\ref{fig:survey_results}. Responses whose \textbf{applicability} response was ''I don't know'' are ommitted from analysis.

\begin{figure}[!h]
\renewcommand{\baselinestretch}{0.7}
\scriptsize
\begin{center}
\begin{tabular}{c|c|c|c|c|c|c}
Survey & law & N & \multicolumn{2}{c}{agreement} & \multicolumn{2}{c}{applicability} \\ 
 & & mean & median & mean & median \\
\hline 
Practitioners & RqtsErr & 16 & 4.6 & 5 & 4.2 & 4 \\ 
Researchers & RqtsErr & 30 & 3.8 & 4 & 3.8 & 4 \\ 
Researchers & PhaseDelay & 30 & 3.6 & 4 & -- & -- 
\end{tabular} 
\end{center}
\caption{Agreement and applicability of phase delay related axioms.}
\label{fig:survey_results}
\end{figure}

While the practitioners strongly believed in the law, researchers were less firm. In reading the free response texts, most participants agreed that, unless the system and the affected requirement are simple/trivial, addressing errors may mean system redesign. Further, any changes to the production system have additional complexity (compared to changes while system in development) due to users, data existence, and architecture dependencies. The researchers who disagreed with the law generally asserted that requirements change can be expensive, but that depends on the process used (e.g., agile vs. waterfall) and the adaptability of the system architecture.

Overall, the RqtsErr law was the most agreed upon and most applicable law of 11 surveyed amongst practitioners, and the second most agreed upon law amongst researchers. The survey results seem to confirm that phase delay escalates cost-to-fix is a widely-held belief in software engineering.

### 2015-03-04 phaseDelay.tex
Added Survey section.
Updates from ShareLaTeX

@@
 Basili and Boehm are not alone in this assessment of the significance of the phase delay effect.
+Previously~\cite{me08a}, we have constructed return-on-investement models
+for a common practice amongst U.S. government contractors: 
+independent verification \& validation. IVV has been criticised as
+ineffective since IVV engineers, who many  unfamiliar with the details of the project, are required to pass judgement on the code.
+Hence,  the IVV team may find the same or fewer bugs
+as the developers,
+On the other hand, IVV has been defended
+as follows:  when the IVV team finds bugs earlier that the developers then, due to the phase delay effect, this saves the overall cost of the project.
+Note that the strength of this argument diminishes
+if fixing later-phase issues is not significantly more expensive than fixing them earlier.
+
 Our surveys of  industrial practitioners (shown below) reveal it to be
@@
 
+\begin{figure}[!t]
+\scriptsize
+\begin{center}
+\begin{tabular}{l@{~}|l@{~}|r@{~}r@{~}|r@{~}l}
+          &        & \multicolumn{2}{c|}{Time to}  &\\
+          &        & \multicolumn{2}{c|}{resolve} &\\
+Injection & Removal& initial & now & \multicolumn{2}{l}{Scale up w.r.t initial} \\\hline
+Phase $i$&Phase $i$ & 1 & 1 & 1 &\rule{2mm}{2mm} \\
+Phase $i$ &Phase $i+1$ & 1 & 2& 2 &\rule{4mm}{2mm} \\
+Phase $i$ &Phase $i+2$ & 1 & 4& 4 &\rule{8mm}{2mm} \\
+Phase $i$ &Phase $i+3$ & 1 & 8 & 8 & \rule{16mm}{2mm} \\
+ Phase $i$  &Phase $i+4$ &1 & 16 & 16 &\rule{32mm}{2mm} 
+  \end{tabular}\end{center}
+ \caption{Projects data consistent with
+ the phase delay effect of \fig{curve1}.}\label{fig:curve1a}
+\end{figure}
+
+\begin{figure}[!t]
+\scriptsize
+\begin{center}
+\begin{tabular}{l@{~}|l@{~}|r@{~}r@{~}|r@{~}l}
+          &        & \multicolumn{2}{c|}{Time to}  &\\
+          &        & \multicolumn{2}{c|}{resolve} &\\
+Injection & Removal& initial & now & \multicolumn{2}{l}{Scale up w.r.t initial} \\\hline
+Phase $i$&Phase $i$ & 1 & 1 & 1 &\rule{2mm}{2mm} \\
+Phase $i$ &Phase $i+1$ & 1 & 2& 2 &\rule{4mm}{2mm} \\
+Phase $i$ &Phase $i+2$ & 1 & 2& 2 &\rule{4mm}{2mm} \\
+Phase $i$ &Phase $i+3$ & 1 & 2 & 2 & \rule{4mm}{2mm} \\
+ Phase $i$  &Phase $i+4$ &1 & 2& 2 &\rule{4mm}{2mm} 
+  \end{tabular}\end{center}
+ \caption{Projects data consistent with
+ the phase delay effect of \fig{curve2}.}\label{fig:curve2a}
+\end{figure}
+
 Beck speculated that there might be another kind of project, shown in \fig{curve2},
@@
 
-\section{lucasSurvey}
+\section{Phase Delay and Cost-to-fix Results from Software Engineering Law Survey}
 

### 2015-03-04 phaseDelay.tex
Updates from ShareLaTeX

@@
 for a common practice amongst U.S. government contractors: 
-independent verification \& validation. IVV has been criticised as
+independent verification \& validation. IVV has been criticized as
 ineffective since IVV engineers, who many  unfamiliar with the details of the project, are required to pass judgement on the code.
@@
 \begin{center}
-\includegraphics[width=6in]{waterfall2.png}
+\includegraphics[width=6in]{waterfall3.png}
 \end{center}

### 2015-03-04 phaseDelay.tex
Added correlation and bug distribution

@@
 
+The distribution of defects found and fixed per phase is shown in Figure~\ref{fig:fix-phase-dist}.
+A high percentage of defects were found and fixed in the early phases, i.e., requirements, high level design, and design reviews and inspections.
+
+\begin{figure*}[!t]
+\begin{center}
+\includegraphics{fix-phase-dist.png}
+\end{center}
+\caption{Distribution of defects found and fixed by phase.}
+\label{fig:fix-phase-dist}
+\end{figure*}
+
+In the TSP data, there is a weak correlation (Kendall's $\tau = 0.19, p < 0.01, 95\% CI$) between a defect's cost-to-fix (minutes of effort) and the defects phase delay (phase removed - phase injected).
+The correlations were also calculated for each phase, e.g., does the phase delay of defects found in Test correlate with the defect's cost-to-fix (Table~\ref{tab:phase_corr}).
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{lll}
+  Phase & $tau$ & $p$ \\ 
+  \hline
+BeforeDevelopment & NA & -- \\ 
+  Planning & NA & -- \\ 
+  Reqts & NA & -- \\ 
+  ReqtsReview & 0.05 & 0.01 \\ 
+  ReqtsInspect & 0.17 & $<$0.01 \\ 
+  HLD & NA & -- \\ 
+  HLDReview & 0.04 & 0.24 \\ 
+  HLDInspect & -0.07 & 0.05 \\ 
+  Design & 0.19 & 0.08 \\ 
+  DesignReview & 0.08 & $<$0.01 \\ 
+  DesignInspect & 0.06 & $<$0.01 \\ 
+  Code & 0.21 & $<$0.01 \\ 
+  CodeReview & 0.15 & $<$0.01 \\ 
+  CodeInspect & 0.11 & $<$0.01 \\ 
+  Test & 0.16 & $<$0.01 \\ 
+  QualTest & NA & -- \\ 
+  IntTest & 0.13 & $<$0.01 \\ 
+  SysTest & 0.32 & $<$0.01 \\ 
+  AcceptTest & 0.07 & 0.29 \\ 
+  \end{tabular}
+\caption{Correlation of cost-to-fix and phase delay for defects found in each phase (Kendall's $\tau$)} 
+\label{tab:phase_corr}
+\end{table}
+
 \begin{figure*}[!t]

### 2015-03-11 data.tex
Outline added for Forrest

@@
-The section presents the data that we use to conclude that phase delay is a not a major
-contributor to the time required to resolve an issue.
+We examine the phase delay effect bin
+171 software projects from around the world from 
+\bill{2005 to 2013} that used the Team Software Process (TSP).
 
+\subsection{Overview of the Team Software Process}
 The Software Engineering Institute (SEI) at
@@
 
-Since 2004, one  of use (Nichols) has been been mentoring 171 software development teams around the world
-as they  deploy TSP within their organizations.  
+\sei{One paragraph summary on the key points of TSP execution? The following text should also be reviewed and better integrated.}
+
+Three common features of all those  are (1)~coaches,
+(2)~peer review, and  (3)~personnel reviews.
+
+A ``coach'' is a team member  authorized to submit project data
+Before submission,
+these coaches check the data for obvious errors.
+
+``Personnel review'' is a technique taken  from the Personal Software
+ProcessSM (PSPSM). PSPSM encourages developers to continually make and review their personnel estimates
+about their day-to-day tasks, then compare those estimates against the actual development effort.
+In this way, developers can acquire a more realistic understanding of their work behaviour.
+
+``Peer review'' is a  technique in
+traditional software engineering.
+ Basili and Boehm write  commented in 2001~\cite{boehm01} 
+that peer reviews can catch over half the defects introduced into a system.
+Peer review can be conducted on any artifact generated anywhere in the software
+lifecycle and can quickly be adapted to new kinds of artifacts.
+
+\subsection{Projects in the Sample}
+Since 2004, one of the authors (Nichols) has been been mentoring 171 software development teams around the world as they deploy TSP within their organizations.  
 The projects
@@
 \bill{can we offer any more deatils here?}
+
+\subsection{Data Collection Process}
+\sei{
+\bi
+    \item What are artifacts uploaded to the SEI tool?
+    \item What are the units contained in those artifacts? Are they the ``issues'' as defined in the following paragraph?
+    \item Is participation in the SEI data collection mandatory? i.e., is the data provided self selected?
+    \item When is the data collected?
+    \item What are screening/cleaning procedures, if any?
+\ei
+}
 
@@
 
-Three common features of all those  are (1)~coaches,
-(2)~peer review, and  (3)~personnel reviews.
-
-A ``coach'' is a team member  authorized to submit project data
-Before submission,
-these coaches check the data for obvious errors.
-
-``Personnel review'' is a technique taken  from the Personal Software
-ProcessSM (PSPSM). PSPSM encourages developers to continually make and review their personnel estimates
-about their day-to-day tasks, then compare those estimates against the actual development effort.
-In this way, developers can acquire a more realistic understanding of their work behaviour.
-
-``Peer review'' is a  technique in
-traditional software engineering.
- Basili and Boehm write  commented in 2001~\cite{boehm01} 
-that peer reviews can catch over half the defects introduced into a system.
-Peer review can be conducted on any artifact generated anywhere in the software
-lifecycle and can quickly be adapted to new kinds of artifacts.
-
 As of January 2014, the SEI TSP database contained data from 109
@@
 \ei
+
+\todo{I recommend deleting the rest of this section. Replace with better descriptions of what the phases are.}
 Also shown in \fig{waterfall} (bottom left) is the standard definition of an agile process~\cite{boehmturner03}:

### 2015-03-11 lucas.tex
Outline added for Forrest

@@
-As part of a related research, we conducted a survey of software engineering practitioners and researchers on their views of commonly-head software engineering ''laws''.
-One of the laws questioned was drawn from from Glass~\cite{glass02} who cites Boehm \lucas{ref needed} as the source: ''Requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' (RqtsErr for short).
+We conducted a survey of software engineering practitioners and researchers on their views of commonly-held software engineering ``laws''.
+One of the laws questioned relates directly to phase delay: ``Requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' (from Glass~\cite{glass02} who cites Boehm \lucas{need to find which exact reference he uses}).
+We abbreviate this law as RqtsErr.
 The survey was conducted in two phases using Amazon's Mechanical Turk.
@@
 
-The respondents answered the following questions for the laws they were presented: 
-\textbf{Agreement:} ''Based on your experience, do you agree that the statement above is correct?'' A Likert scale captured the agreement score from Strongly Disagree to Strongly Agree.
+The respondents answered the following questions for the laws they were presented: \newline
+\textbf{Agreement:} ``Based on your experience, do you agree that the statement above is correct?'' A Likert scale captured the agreement score from Strongly Disagree to Strongly Agree.
 A text box was provided to explain the answer.
-\textbf{Applicability:} ''To the extent that you believe it, how widely do you think it applies among software development contexts?'' The possible answers were presented as a scale from -1 to 5:
+\newline
+\textbf{Applicability:} ``To the extent that you believe it, how widely do you think it applies among software development contexts?'' The possible answers were presented as a scale from -1 to 5:
 \begin{itemize}
@@
 
-In the ''law evaluation'' portion of the survey, participants were presented with the RqtsErr law and others drawn from \cite{glass02} and \cite{endres03}.
-The PC member survey contained an additional question: ''Do you agree with the following?
+Participants were presented with the RqtsErr law and others drawn from \cite{glass02} and \cite{endres03}.
+The PC member survey contained an additional question: ``Do you agree with the following?
 'In general, the longer errors are in the system (requirements errors, design errors, coding errors, etc.), the more expensive they are to fix''' (PhaseDelay for short).
@@
 
-Summary statistics for the agreement and applicability scores for the RqtsErr and PhaseDelay laws are presented in Figure~\ref{fig:survey_results}.
+Summary statistics for the agreement and applicability scores for the RqtsErr and PhaseDelay laws are presented in Figure~\ref{fig:survey_results} \lucas{To add results from other laws}.
 Responses whose \textbf{applicability} response was ''I don't know'' are ommitted from analysis.
@@
 While the practitioners strongly believed in the law, researchers were less firm.
-In reading the free response texts, most participants agreed that, unless the system and the affected requirement are simple/trivial, addressing errors may mean system redesign.
+In the free response texts, most participants agreed that addressing errors may mean system redesign unless the system and the affected requirement are simple/trivial.
 Further, any changes to the production system have additional complexity (compared to changes while system in development) due to users, data existence, and architecture dependencies.

### 2015-03-11 phaseDelay.tex
Outline added for Forrest

@@
 \usepackage{multirow}
-\usepackage{color}
+\usepackage[usenames, dvipsnames]{color}
 \usepackage{colortbl}
@@
 \renewcommand{\algorithmicensure}{\textbf{Output:}} 
-\newcommand{\lucas}[1]{\textcolor{red}{LUCAS: #1}} 
+\newcommand{\lucas}[1]{\textcolor{red}{LUCAS: #1}} \newcommand{\tim}[1]{\textcolor{orange}{TIM: #1}}
 \newcommand{\bill}[1]{\textcolor{blue}{BILL: #1}} 
 \newcommand{\carter}[1]{\textcolor{cyan}{CARTER: #1}} 
+\newcommand{\sei}[1]{\textcolor{RedViolet}{BILL/FORREST: #1}} 
+\newcommand{\todo}[1]{\textcolor{Maroon}{TODO: #1}} 
 %\newenvironment{changed}{\par}{\par}
@@
 
-This result raises the question:
-how did myths like phase delay survive so long  without  critical reassessment.  
-As a community, we need to reflect more on 
-research practices and on how (and when) we reflect on the prominent hypotheses of our field.
+%This result raises the question: how did myths like phase delay survive so long  without  critical reassessment.
+As a community, we need to reflect more on research practices and on how (and when) we reflect on the prominent hypotheses of our field.
 \end{abstract}
@@
 Accordingly, this paper revisits
-the {\em phase delay} effect
-i.e.  the time required to resolve an issue in a software project increases dramatically 
+the {\em phase delay effect}
+i.e.,  the time required to resolve an issue in a software project increases dramatically 
 the longer it remains in the system. 
@@
 \end{quote}
-Basili and Boehm are not alone in this assessment of the significance of the phase delay effect.
+Basili and Boehm are not alone in this assessment of the significance of the phase delay effect. Our surveys of industrial practitioners (shown in \S\ref{survey}) reveal the phase-delay effect to be
+the single most strongly held belief amongst commercial SE engineers.
 Previously~\cite{me08a}, we have constructed return-on-investement models
@@
 
-Our surveys of  industrial practitioners (shown below) reveal it to be
-the single most strongly held belief amongst commercial SE engineers;
-Also,
-as discussed later
-in this paper, concerns over phase delay were a major driver in the development of agile software
-processes.
+% Also, as discussed later in this paper, concerns over phase delay were a major driver in the development of agile software processes.
 
-The conclusions of this paper is that it is erroneous to believe in phase delay.
-When we looked for supporting evidence for that effect, we found that:
-\be 
-\item The evidence for that effect in the historical literature is very weak (to say the least);
-\item  In  a large sample of   contemporary software projects, there is no evidence for phase delay.
- \ee
-The rest of this paper present evidence for the above two points. 
-The first point is  documented via  a literature survey and the second point is demonstrated
-via an analysis of  
- 171 software  projects developed in the period \bill{2004 to 2014}.  To the best of our knowledge,
- this is the largest study on phase delay yet seen in the literature.
+In this paper, we investigate two hypotheses:
 
-The conclusions of this paper are two fold. Firstly, the SE community needs to stop
-justifying changes to SE
-practices via the phase delay effect. Secondly,   myths like phase delay should not have lasted decades
-in the literature without  critical reassessment.  As a community, we need to reflect more on our
-research practices and on how (and when) we reflect and revisit the prominent hypotheses of our field.
+\bi
+    \item $H_1$: The cost-to-fix of defects monotonically increases across phases.
+    \item $H_2$: The cost-to-fix of defects monotonically increases as the phase delay increases.
+\ei
+The rest of the paper investigates these two hypotheses.
+First, we conduct a survey of practitioners and researchers to gauge their belief in the phase delay effect.
+Second, we  analyze 171 software  projects developed in the period \bill{2004 to 2014}.
+To the best of our knowledge, this is the largest study on phase delay yet seen in the literature.
+
+% The conclusions of this paper is that it is erroneous to believe in phase delay.
+% When we looked for supporting evidence for that effect, we found that:
+% \be 
+% \item The evidence for that effect in the historical literature is very weak (to say the least);
+% \item  In  a large sample of   contemporary software projects, there is no evidence for phase delay.
+%  \ee
+% The rest of this paper present evidence for the above two points.
+
+% The first point is  documented via  a literature survey and the second point is demonstrated
+% via an analysis of
+
+The conclusions of this paper are two fold.
+First, the phase delay effect does not hold in the large body of software projects analyzed.
+Second, commonly-held beliefs, such as phase delay, must be periodically revisited empirically in order for researchers to provide accurate and up-to-date guidance for the practitioner community.
+
+% the SE community needs to stop
+% justifying changes to SE
+% practices via the phase delay effect. Secondly,   myths like phase delay should not have lasted decades
+% in the literature without  critical reassessment.  As a community, we need to reflect more on our
+% research practices and on how (and when) we reflect and revisit the prominent hypotheses of our field.
 
@@
 
-Before exploring one particular truism, we digress to discuss the merits of revisiting
+Before exploring the commonly-held belief of the {\em phase delay effect}, we digress to discuss the merits of revisiting
 old conclusions in software engineering.
 
-Some researchers   adamantly believe in general
+Belief in general
 principles of software
-engineering. Processional societies assume such generalities exist when the offer
+engineering are common to both research and practice. Processional societies assume such generalities exist when they offer
  lists of supposedly general ``best practices'' such as
@@
 \end{quote}
- Endres \& Rombach offer dozens of supposedly general lessons of software engineering~\cite{rombach11}.
+ Endres \& Rombach offer dozens of lessons of software engineering~\cite{rombach11}.
  Many other 
@@
 longer than debugging a standard procedural program~\cite{hatton98}.
-\item
-A truism of visual programming is that ``visual
-representations are inherently superior to mere textual representations''.
-A review by Menzies suggests that the available
+\item A truism of visual programming is that ``visual
+representations are inherently superior to mere textual representations''. A review by Menzies suggests that the available
 evidence for this claim is hardly conclusive~\cite{me00v}. 
@@
 \ei
-To be fair to the software engineering community, SE is not the only
+%To be fair to the software engineering community,
+Software engineering is not the only
 field where practitioners hang on to persistent beliefs, even if the evidence
@@
 \end{quote}
-The results of Jorgensen \& Gruschke support the findings if Passos et al. They report that 
+The results of J{\o}rgensen \& Gruschke~\cite{jorgensen09} support the findings if Passos et al. They report that 
   supposed software engineering    ``experts'' rarely use lessons
@@
  to regularly  reexamine    old beliefs.
+
 \subsection{Why Study Phase Delay?}
+\todo{\begin{itemize}
+\item Given the Intro and Motivation, this section should focus on the influence of phase delay in software engineering milieux
+\item Need to tone down the reliance on Beck.
+The important Beck point is that eliminating the cost-to-fix curve (resulting from phase delay) is one of the main goals of agile.
+His observations on new technologies are good too.
+\item Need to cite other popular sources of this information, e.g., McConnell's "Ounce of Prevention" article~\cite{mcconnell01}.
+\item Robert Glass's~\cite{glass02} book is very popular and contains an entry on this point
+\item Google Scholar Citations for phase delay works Boehm88~\cite{boehm88} (673), Boehm \& Basili~\cite{boehm01} (761), 
+\end{itemize}}
+
 Due to its prominence in the field of software engineering,
@@
 Injection & Removal& initial & now & \multicolumn{2}{l}{Scale up w.r.t initial} \\\hline
-Phase $i$&Phase $i$ & 1 & 1 & 1 &\rule{2mm}{2mm} \\
-Phase $i$ &Phase $i+1$ & 1 & 2& 2 &\rule{4mm}{2mm} \\
-Phase $i$ &Phase $i+2$ & 1 & 4& 4 &\rule{8mm}{2mm} \\
-Phase $i$ &Phase $i+3$ & 1 & 8 & 8 & \rule{16mm}{2mm} \\
- Phase $i$  &Phase $i+4$ &1 & 16 & 16 &\rule{32mm}{2mm} 
+Phase $i$&Phase $i$ & 1 & 1 & 1 & **\\
+Phase $i$ &Phase $i+1$ & 1 & 2& 2 & **** \\
+Phase $i$ &Phase $i+2$ & 1 & 4& 4 &******** \\
+Phase $i$ &Phase $i+3$ & 1 & 8 & 8 & ****************  
   \end{tabular}\end{center}
@@
 Injection & Removal& initial & now & \multicolumn{2}{l}{Scale up w.r.t initial} \\\hline
-Phase $i$&Phase $i$ & 1 & 1 & 1 &\rule{2mm}{2mm} \\
-Phase $i$ &Phase $i+1$ & 1 & 2& 2 &\rule{4mm}{2mm} \\
-Phase $i$ &Phase $i+2$ & 1 & 2& 2 &\rule{4mm}{2mm} \\
-Phase $i$ &Phase $i+3$ & 1 & 2 & 2 & \rule{4mm}{2mm} \\
- Phase $i$  &Phase $i+4$ &1 & 2& 2 &\rule{4mm}{2mm} 
+Phase $i$&Phase $i$ & 1 & 1 & 1 & ** \\
+Phase $i$ &Phase $i+1$ & 1 & 2& 2 & **** \\
+Phase $i$ &Phase $i+2$ & 1 & 2& 2 & **** \\
+Phase $i$ &Phase $i+3$ & 1 & 2 & 2 & **** \\
+ Phase $i$  &Phase $i+4$ &1 & 2& 2 &**** 
   \end{tabular}\end{center}
@@
 
-\section{Phase Delay and Cost-to-fix Results from Software Engineering Law Survey}
+\section{A Survey on Phase Delay and Cost-to-fix Belief}
+\label{survey}
 
@@
 
-\section{billTSP}
+%\section{carterCharts}
 
-\section{carterCharts}
+%\input{carter}
 
-\input{carter}
+\begin{figure*}[!t]
+\begin{center}
+\includegraphics[width=6in]{waterfall3.png}
+\end{center}
+\caption{Different approaches to software development:  waterfall and agile (bottom left).}
+\label{fig:waterfall}
+\end{figure*}
+
+\section{Data Collection}
+
+\input{data}
+
+\section{Analysis of Phase Delay in 171 TSP Projects}
+
+\todo{
+\bi
+    \item What is the cost-to-fix curve in TSP data? Check against H1 and compare to past literature.
+    \item What is the effect of phase delay? Check against H2.
+\ei
+}
+
+\subsection{TSP Project Cost-to-fix curve}
 
@@
 A high percentage of defects were found and fixed in the early phases, i.e., requirements, high level design, and design reviews and inspections.
+\todo{How does this compare evidence from other sources?}
 
-\begin{figure*}[!t]
+The median time to fix a defect in each phase is shown in \fig{fix-time-per-phase}. \todo{Overlay median on Boehm's chart \fig{cost-to-fix}. How does it compare?}
+
+\begin{figure}[!ht]
 \begin{center}
-\includegraphics{fix-phase-dist.png}
+\includegraphics[width=3.3in]{fix-phase-dist.png}
 \end{center}
@@
 \label{fig:fix-phase-dist}
-\end{figure*}
+\end{figure}
+
+\begin{figure}[!ht]
+\begin{center}
+\begin{tabular}{r|rrr|l}
+  Sample&\multicolumn{3}{c|}{Percentiles}\\ 
+size & 25th & 50th & 75th & Phase removed \\
+\hline
+120 & 3 & 3 & 18 & ReqtsInspect.dat \\
+126 & 3 & 3 & 20 & ReqtsReview.dat \\
+70 & 2 & 2 & 10 & HLDInspect.dat \\
+ 55 &    2 &    1&    7&HLDReview.dat \\
+
+278&    7&    7&   33&DesignInspect.dat \\
+233&    6&    6&   25&DesignReview.dat \\
+
+315&    8&    7&   34&CodeInspect.dat \\ 
+282&    7&    7&   32&CodeReview.dat\\
+
+425&   11&   10&   54&UnitTest.dat \\ 
+226&    7&    6&   46&IntTest.dat \\ 
+194&    7&    6&   52&SysTest.dat\\ 
+ 88&    4&    3&   38&AcceptTest.dat\\
+\end{tabular}
+\end{center}
+\caption{Median fix time per phase}
+\label{fig:fix-time-per-phase}
+\end{figure}
+
+We reject $H_1$.
+
+\subsection{Phase delay analysis}
+\todo{
+\bi
+    \item What can we say about \fig{raw}
+    \item What can we say about \fig{scale}
+    \item What is the difference between \fig{raw} and \fig{scale}
+\ei
+}
 
@@
 The correlations were also calculated for each phase, e.g., does the phase delay of defects found in Test correlate with the defect's cost-to-fix (Table~\ref{tab:phase_corr}).
+
+We reject $H_2$.
+
+\begin{figure*}[!t] 
+ \renewcommand{\baselinestretch}{0.7}
+ \scriptsize
+\begin{center}
+\begin{tabular}{r|rrr|ll|rl}
+  Sample&\multicolumn{3}{c|}{Percentiles}\\ 
+size & 25th & 50th & 75th & Phase injected & Phase removed & \multicolumn{2}{l}{Scale up w.r.t. to first phase}\\\hline
+\\
+  29 &     1 &     3 &    12 & Before & CodeInspect  & 1.00  &  **  \\
+  26 &     1 &     4 &     6 & Before & CodeReview  & 1.33  &  ***  \\
+  73 &     5 &    13 &    59 & Before & UnitTest  & 4.33  &  *********  \\
+  36 &     7 &    24 &    54 & Before & IntTest  & 8.00  &  ****************  \\
+  34 &     8 &    28 &    56 & Before & SysTest  & 9.33  &  *******************  \\\hline
+\\
+128 &     2 &     6 &    15 & Reqts & ReqtsInspect  & 1.00  &  **  \\
+132 &     2 &     5 &    12 & Reqts & ReqtsReview  & 0.83  &  **  \\
+ 28 &     3 &    13 &    29 & Reqts & DesignInspect  & 2.16  &  *****  \\
+ 22 &     6 &    14 &    31 & Reqts & DesignReview  & 2.33  &  *****  \\\hline
+\\
+272 &     6 &    14 &    30 & Design & DesignInspect  & 1.00  &  **  \\
+ 32 &     1 &     4 &    13 & Design & DesignInspection  & 0.28  &  *  \\
+240 &     4 &    11 &    22 & Design & DesignReview  & 0.78  &  **  \\
+110 &     3 &    11 &    23 & Design & CodeInspect  & 0.78  &  **  \\
+ 91 &     3 &     7 &    18 & Design & CodeReview  & 0.50  &  *  \\
+323 &     8 &    21 &    45 & Design & UnitTest  & 1.50  &  ***  \\
+ 56 &     7 &    20 &    41 & Design & IntTest  & 1.42  &  ***  \\
+ 69 &     9 &    27 &    53 & Design & SysTest  & 1.92  &  ****  \\\hline
+\\
+310 &     6 &    15 &    31 & Code & CodeInspect  & 1.00  &  **  \\
+285 &     5 &    14 &    29 & Code & CodeReview  & 0.93  &  **  \\
+287 &     7 &    17 &    44 & Code & UnitTest  & 1.13  &  ***  \\
+180 &     7 &    20 &    45 & Code & IntTest  & 1.33  &  ***  \\
+137 &     6 &    15 &    35 & Code & SysTest  & 1.00  &  **  \\
+ 37 &     1 &    11 &    27 & Code & AcceptTest  & 0.73  &  **  \\
+
+ \end{tabular}
+\end{center}
+\caption{Distribution of fix times seen in SEI TSP data.}
+\label{fig:raw}
+\end{figure*}
+
+ \begin{figure}[!t]
+\renewcommand{\baselinestretch}{0.7}
+\scriptsize
+\begin{center}
+\begin{tabular}{l@{~~}|l@{~}|r@{~}|r@{~}r@{~}|r@{~}l}
+           \multicolumn{2}{c}{~}                 &  &\multicolumn{2}{c|}{median}\\
+  Injection&   Removal& $n$ & initial & now & \multicolumn{2}{l}{Scale up, w.r.t. initial}
+\input{deltas}
+\end{tabular}
+\end{center}
+\caption{50th percentile (median) scale ups  for  time to resolve issues (taken from \fig{raw}).}
+\label{fig:scale}
+\end{figure}
 
@@
 
-\begin{figure*}[!t]
-\begin{center}
-\includegraphics[width=6in]{waterfall3.png}
-\end{center}
-\caption{Different approaches to software development:  waterfall and agile (bottom left).}
-\end{figure*}
-
-\section{Data}
-
-\input{data}
-
-\begin{figure}[!t]
-
-\renewcommand{\baselinestretch}{0.7}
-\scriptsize
-\begin{center}
-\begin{tabular}{r|rrr|ll}
-  Sample&\multicolumn{3}{c|}{Percentiles}\\ 
-size & 25th & 50th & 75th & Phase injected & Phase removed\\\hline
-57& 6& 16& 47&  BeforeDevelopment&Code\\
-72& 2& 4& 11&  BeforeDevelopment&CodeInspect\\
-165& 6& 18& 56&  BeforeDevelopment&Test\\
-50& 6& 20& 37&  BeforeDevelopment&IntTest\\
-42& 8& 28& 65&  BeforeDevelopment&SysTest\\\hline
-
-66& 2& 6& 15&  Planning&Planning\\
-41& 1& 5& 16&  Planning&ReqtsReview\\\hline
-
-23& 0& 1& 3&  Reqts&Reqts\\
-245& 6& 13& 23&  Reqts&ReqtsReview\\
-289& 7& 16& 30&  Reqts&ReqtsInspect\\
-32& 6& 21& 40&  Reqts&Design\\
-49& 2& 7& 24&  Reqts&DesignInspect\\\hline
-
-94& 2& 5& 10&  HLD&HLDReview\\
-133& 3& 8& 19&  HLD&HLDInspect\\
-33& 2& 5& 17&  HLD&Design\\\hline
-
-37& 2& 4& 6&  Design&Design\\
-455& 11& 24& 53&  Design&DesignInspect\\
-218& 5& 14& 30&  Design&Code\\
-166& 4& 11& 22&  Design&CodeInspect\\
-542& 14& 31& 60&  Design&Test\\
-67& 6& 15& 40&  Design&IntTest\\
-92& 10& 25& 53&  Design&SysTest\\\hline
-
-53& 2& 4& 9&  DesignInspect&DesignInspect\\
-38& 1& 3& 9&  DesignInspect&Code\\\hline
-
-126& 4& 12& 31&  Code&Code\\
-459& 11& 24& 45&  Code&CodeInspect\\
-461& 11& 26& 55&  Code&Test\\
-348& 10& 26& 62&  Code&IntTest\\
-230& 8& 18& 39&  Code&SysTest\\
-71& 2& 11& 28&  Code&AcceptTest\\\hline
-
-64& 2& 4& 10&  CodeInspect&CodeInspect\\
-57& 2& 5& 13&  CodeInspect&Test\\\hline
-
-110& 3& 10& 23&  Test&Test\\
-66& 2& 4& 12&  Test&QualTest\\\hline
-
-32& 1& 2& 4&  IntTest&QualTest\\
-36& 1& 11& 30&  IntTest&IntTest\\
-21& 1& 2& 5&  IntTest&SysTest\\
- \end{tabular}
-\end{center}
-\caption{Distribution of fix times seen in SEI TSP data.}
-\label{fig:faw}
-\end{figure}
-
-\begin{figure}[!t]
-\renewcommand{\baselinestretch}{0.7}
-\scriptsize
-\begin{center}
-\begin{tabular}{l@{~~}|l@{~}|r@{~}|r@{~}r@{~}|r@{~}l}
-           \multicolumn{2}{c}{~}                 &  &\multicolumn{2}{c|}{median}\\
-  Injection&   Removal& $n$ & initial & now & \multicolumn{2}{l}{Scale up, w.r.t. initial}
-\input{deltas}
-\end{tabular}
-\end{center}
-\caption{50th percentile (median) scale ups  for  time to resolve issues (taken from \fig{raw}).}
-\label{fig:scale}
-\end{figure}
-
 \section{Phase Delay in Slower Bugs?}
@@
 \begin{tabular}{l@{~~}|l@{~}|r@{~}|r@{~}r@{~}|r@{~}l}
-           \multicolumn{2}{c}{~}                 &  &\multicolumn{2}{c|}{90th }\\
-           \multicolumn{2}{c}{~}                 &  &\multicolumn{2}{c|}{percentile }\\
-  Injection&   Removal& $n$ & initial & now & \multicolumn{2}{l}{Scale up, w.r.t. initial}
-\input{deltaBig}
-\end{tabular}
-\end{center}
-\caption{90th percentile scale ups.}
-\label{fig:scale90}
-\end{figure}
-
-\begin{figure}[!t]
-\renewcommand{\baselinestretch}{0.7}
-\scriptsize
-\begin{center}
-\begin{tabular}{l@{~~}|l@{~}|r@{~}|r@{~}r@{~}|r@{~}l}
           \multicolumn{2}{c}{~}                 &  &\multicolumn{2}{c|}{99th }\\
@@
 \caption{99th percentile scale ups.}
-\label{fig:scale90}
+\label{fig:scale99}
 \end{figure}

### 2015-03-11 phaseDelay.tex
Phase delay in slower bugs  is now a subsection
Added discussion and conclusion section outlines

@@
 
-\section{Phase Delay in Slower Bugs?}
+\subsection{Phase Delay in Slower Bugs?}
 In these results, we checked if a small number of bugs are most expensive. 
@@
 
-\section{Related Work}
+ \section{Discussion}
+ \todo{
+ \bi
+    \item summarize analysis findings
+    \item Offer interpretations of findings, e.g., why might these things be so
+ \ei
+ }
 
-\begin{figure}
-\begin{center}
-\scriptsize\begin{tabular}{|l@{~:~}l|}\hline
-Bug Prediction Dataset &http://bug.inf.usi.ch \\
-Eclipse Bug Data &http://goo.gl/tYKahN \\
-FLOSSMetrics& http://flossmetrics.org \\
-FLOSSMole &http://flossmole.org \\
-IBSBSG& http://www.isbsg.org \\
-ohloh& http://www.openhub.net \\
-PROMISE &http://promisedata.googlecode.com \\
-Qualitas Corpus &http://qualitascorpus.com \\
-Software Artifact Repository &http://sir.unl.edu \\
-SourceForge Research Data &http://zerlot.cse.nd.edu \\
-Sourcerer Project &http://sourcerer.ics.uci.edu \\
-Tukutuku &http://www.metriq.biz/tukutuku \\
-Ultimate Debian Database &http://udd.debian.org\\\hline
-\end{tabular}
-\end{center}
-\caption{Some repositories of software engineering data.}\label{fig:sedata}
-\end{figure}
-The data collected in the SEI TSP databases contains 
-extensive process details
-including a detailed phased breakdown showing what happened at what
-phases of the lifecycle. This makes this data somewhat different to the standard
- publicly available
-software projects data sets currently available to 
-SE researchers. Most of the \fig{sedata} data sets have software product information
-such as full source code, or summaries of static features.
-A subset of that data contain issue or defect reports and/or
-the time taken to build these systems.  We are unaware
-of any of these having detailed phase-by-phase breakdowns of project data.
+ \section{Conclusion}
+ \todo{
+ \bi
+    \item Summarize data tested - survey + TSP data
+    \item reiterate hypothesis testing results
+    \item Restate the motivation: widely-held beliefs need to be revisted. 
+    \item We now have evidence, from 5 different sources, that phase delay does not seem to hold under certain conditions (although our study is by far the biggest in terms of data points).
+    \item phase delay may still apply in some (many?) projects, but clearly it does not in all. This has huge implications for, e.g., cost modeling, effort prediction.
+ \ei
+ }
+
+% \section{Related Work}
+
+% \begin{figure}
+% \begin{center}
+% \scriptsize\begin{tabular}{|l@{~:~}l|}\hline
+% Bug Prediction Dataset &http://bug.inf.usi.ch \\
+% Eclipse Bug Data &http://goo.gl/tYKahN \\
+% FLOSSMetrics& http://flossmetrics.org \\
+% FLOSSMole &http://flossmole.org \\
+% IBSBSG& http://www.isbsg.org \\
+% ohloh& http://www.openhub.net \\
+% PROMISE &http://promisedata.googlecode.com \\
+% Qualitas Corpus &http://qualitascorpus.com \\
+% Software Artifact Repository &http://sir.unl.edu \\
+% SourceForge Research Data &http://zerlot.cse.nd.edu \\
+% Sourcerer Project &http://sourcerer.ics.uci.edu \\
+% Tukutuku &http://www.metriq.biz/tukutuku \\
+% Ultimate Debian Database &http://udd.debian.org\\\hline
+% \end{tabular}
+% \end{center}
+% \caption{Some repositories of software engineering data.}\label{fig:sedata}
+% \end{figure}
+% The data collected in the SEI TSP databases contains 
+% extensive process details
+% including a detailed phased breakdown showing what happened at what
+% phases of the lifecycle. This makes this data somewhat different to the standard
+%  publicly available
+% software projects data sets currently available to 
+% SE researchers. Most of the \fig{sedata} data sets have software product information
+% such as full source code, or summaries of static features.
+% A subset of that data contain issue or defect reports and/or
+% the time taken to build these systems.  We are unaware
+% of any of these having detailed phase-by-phase breakdowns of project data.
 
@@
 This work was partially funded by an National Science
-Foundation grant NSF-CISE 1302169.
+Foundation grants NSF-CISE 1302169 and CISE 1302169.
 

### 2015-03-12 data.tex
added scatter plots of phase delay and other charts.

@@
 We examine the phase delay effect bin
-171 software projects from around the world from 
-\bill{2005 to 2013} that used the Team Software Process (TSP).
+212 software projects from around the world from 
+\bill{2005 to 2015} that used the Team Software Process (TSP).
 
@@
 \sei{One paragraph summary on the key points of TSP execution? The following text should also be reviewed and better integrated.}
+TSP guiding principles include 1) engineering work does not happen by accident, it must be planned, 2) the plans must include realistic commitments of the resources needed to perform the work, 3) planed and actual results should be frequently examined to assure that plans remain relevant, and 4) people performing the work should be responsible for planning and tracking the work.
+After clarifying goals, defining work practices, and estimating size and effort, developers make a detailed short term plan.
+As the developers perform project work, they use a tool such as the Process Dashboard to collect their time, effort, size, and schedule data.
+Every week, the team review their data to evaluate status, review actual rates, determine if project goals for schedule, cost, and quality are being met and then make necessary plan corrections.
+At the end of the project the coach and team perform a quantitative post mortem.
 
-Three common features of all those  are (1)~coaches,
-(2)~peer review, and  (3)~personnel reviews.
+Four common features of TSP projects include (1)~coaches,
+(2)~peer review, and  (3)~personnel reviews (4)~planning.
 
-A ``coach'' is a team member  authorized to submit project data
-Before submission,
+A ``coach'' works with the team and is authorized to submit project data.
+Before reviewing data with the teams, therefore before submission,
 these coaches check the data for obvious errors.
@@
 ``Personnel review'' is a technique taken  from the Personal Software
-ProcessSM (PSPSM). PSPSM encourages developers to continually make and review their personnel estimates
-about their day-to-day tasks, then compare those estimates against the actual development effort.
-In this way, developers can acquire a more realistic understanding of their work behaviour.
+ProcessSM (PSPSM) and is more or less unique in its TSP implementation.
+The technique of a personal review is similar to that of a peer review.
+Developers systematically examine their own work products using a checklist built from their personal defect profile.
+This personal review occurs after a portion of a product is done and before peer reviews.
+
+%PSPSM teaches developers howto continually make and review their personnel estimates
+%about their day-to-day tasks, then compare those estimates against the actual development effort.
+%In this way, developers can acquire a more realistic understanding of their work behaviour.
 
@@
 lifecycle and can quickly be adapted to new kinds of artifacts.
+TSP peer reviews follow the Fagan style in which the reviewer uses a checklist composed of common team defects prior to a review team meeting.
+
+''Planning '' is another technique from the in the PSP(SM).
+Developers estimate the size of work products, define the process phases, use historic rates and time in phase to estimate the required time, the number of defects injected, and the defects to be removed in each phase.
+Coaches help the developers to compare estimates against actual results.
+In this way, developers acquire a more realistic understanding of their work behavior and performance.
 
@@
 
-To aid in monitoring these 171 projects  Nichols  built a
-data collection tool that gathers project data a central database at SEI. 
+\sei{ wrn: please review the following}
+
+Organizations using TSP agree to provide project data to the SEI for use in research with the provision that the data must not be traceable to its source.
+The data includes project and site  characteristic summaries, surveys of team members, launch presentations, launch outbriefs, and baseline plans, final data from the project, and a post mortem report.
+In practice, this data requirement has only been enforced for the purposes of certifying and reauthorizing TSP coaches who are the only individuals authorized to submit data.
+To this data, only the data recorded using the Process Dashboard tool has been collected and aggregated for research.
+
+To aid in using this data for research and benchmarking  Nichols first collected the Process Dashboard data from the submission repository then used a tool built by the Process Dashboard developer, David Tuma,
+to gather the project data into a database at SEI.
+Nichols and Shirai then extracted data into views suitable for analysis.
+.
+The key views include the a project results summary, effort logs, task logs, and defect logs.
+At this stage, the data has not yet undergone additional screening.
+
+\sei{
+do we need the following information? we do not use it in this analysis
+}
+
 The issue types collected in the the SEI TSP data set divide into:
@@
 
-As of January 2014, the SEI TSP database contained data from 109
-TSP projects. The projects started between July 2009 and
-September 2013; they included 34 teams and 309 people. Among
-the database fact tables, the time store contains 103,023 time logs,
-18,408 defect logs, and 7,464 size logs\footnote{\bill{we need to 
-update this table}}.
+%wrn, updated 3/11 these
+As of February 2015, the SEI TSP database contained data from 212
+TSP projects. The projects completed between July 2008 and
+November 2014; they included 47 organizations and 843 people. Among
+the database fact tables, 
+contains 268,726 time logs, 
+154,238 task logs,
+ 47,376 defect logs, 
+and 26,534 size logs
+%\footnote{\bill{we need toupdate this table}}.
 
@@
 This paper studies the impact of phase delay on the time required to resolve issues.
-The phased used in this paper are shown in \fig{waterfall}. Note that, in that figure:
+The logical phases used in this paper are shown in \fig{waterfall}. Note that, in that figure:
 \bi 
 \item
-Several  phases have the same  sub-activities of {\em review} and {\em inspect}\footnote{\bill{plz distinguish review and inspect. does one usually happen first?}}
-\item Testing is divided into several statges \footnote{\bill{need a one
+Several  phases have the similar  sub-phases of {\em review} and {\em inspect}.
+In TSP, individuals perform personal reviews on products prior to the peer inspection.\footnote{\bill{plz distinguish review and inspect.
+does one usually happen first?}}
+\item Testing is divided into several statges \footnote{\bill{
+(this appears to be OBE), need a one
 line description of test vs qualtest vs inttest vs systemTest vs AcceptTest. I tries some words in the Key to \fig{sedata}. dont know if i got it right.}}
@@
 the code is re-organized based on feedback from those tests~\cite{fraser03}.} where
-reviews are scheduled after testing. Some of the groups in the SEI TSP data used that approach but
+code inspections are scheduled after unit testing. Some of the groups in the SEI TSP data used that approach but
 they do not effect the main conclusions of this paper:
 \bi 
-\item They were strongly in the minority\bill{any numbers on this?};
+\item They were less than 5\% of the total \bill{any numbers on this?};
+
 \item Most of our phase delay data comes from much early in the waterfall model
@@
  combine development and  testing. 
- \item  The TSP projects studied here offer new releases at least every three months, but
+ \item  The TSP projects studied here typically offer new releases at least every three months, but
 often much longer that that\bill{any numbers on that}. Note that this is a far slower

### 2015-03-12 phaseDelay.tex
added scatter plots of phase delay and other charts.

@@
 
+\usepackage{subfigure}
 \usepackage{picture}
@@
 \conferenceinfo{FSE}{'15 Bergamo, Italy}
-\title{Revisiting the Truisms of Software Engineering:\\ Does Phase Delay Dramatically Increases  Repair Time?}
+\title{Revisiting the Truisms of Software Engineering:\\ Does Phase Delay Dramatically Increase  Repair Time?}
 \numberofauthors{3}
@@
 Yet based on a sample of 
-171 software projects from around the world from 
-\bill{2005 to 2013}, we argue that this belief is mostly 
+212 software projects from around the world from 
+\bill{2005 to 2014}, we argue that this belief is mostly 
 incorrect. Specifically: the median fix time for issues
@@
 
-The median time to fix a defect in each phase is shown in \fig{fix-time-per-phase}. \todo{Overlay median on Boehm's chart \fig{cost-to-fix}. How does it compare?}
+The median time to fix a defect in each phase is shown in \fig{fix-time-per-phase}.
+The median cost to fix a defect is low in the early phases, and then increases slightly during design and coding before peaking during the unit test phase.
+The median fix time then drops during the testing phases.
+Note that while the median time remains relatively consistent, the 75th percentile does increase during the testing phases.
 
@@
 \hline
-120 & 3 & 3 & 18 & ReqtsInspect.dat \\
-126 & 3 & 3 & 20 & ReqtsReview.dat \\
-70 & 2 & 2 & 10 & HLDInspect.dat \\
- 55 &    2 &    1&    7&HLDReview.dat \\
+120 & 3 & 3 & 18 & ReqtsInspect \\
+126 & 3 & 3 & 20 & ReqtsReview \\
+70 & 2 & 2 & 10 & HLDInspect \\
+ 55 &    2 &    1&    7&HLDReview \\
 
-278&    7&    7&   33&DesignInspect.dat \\
-233&    6&    6&   25&DesignReview.dat \\
+278&    7&    7&   33&DesignInspect \\
+233&    6&    6&   25&DesignReview \\
 
-315&    8&    7&   34&CodeInspect.dat \\ 
-282&    7&    7&   32&CodeReview.dat\\
+315&    8&    7&   34&CodeInspect \\ 
+282&    7&    7&   32&CodeReview\\
 
-425&   11&   10&   54&UnitTest.dat \\ 
-226&    7&    6&   46&IntTest.dat \\ 
-194&    7&    6&   52&SysTest.dat\\ 
- 88&    4&    3&   38&AcceptTest.dat\\
+425&   11&   10&   54&UnitTest \\ 
+226&    7&    6&   46&IntTest \\ 
+194&    7&    6&   52&SysTest\\ 
+ 88&    4&    3&   38&AcceptTest\\
 \end{tabular}
@@
 \label{fig:fix-time-per-phase}
+\end{figure}
+
+The relative cost to fix errors per lifecycle phase in the TSP project data is overlayed with Boehm's exponential cost curve~\cite{Boehm81} as well as the Royce case study~\cite{Royce98} in \fig{cost-to-fix-tsp}.
+
+\begin{figure}[!ht]
+ \includegraphics[width=3.3in]{boehm_overlay.png}
+ \caption{Cost-to-fix curve from \fig{cost-to-fix} overlayed with TSP data and case study from~\cite{Royce98}.}
+ \label{fig:cost-to-fix-tsp}
  \end{figure}
@@
 
-In the TSP data, there is a weak correlation (Kendall's $\tau = 0.19, p < 0.01, 95\% CI$) between a defect's cost-to-fix (minutes of effort) and the defects phase delay (phase removed - phase injected).
-The correlations were also calculated for each phase, e.g., does the phase delay of defects found in Test correlate with the defect's cost-to-fix (Table~\ref{tab:phase_corr}).
+ \fig{phase_fixtime_plot} shows scatterplots of the cost-to-fix (rank) per phase against the phase delay.
+In the early phases, i.e., Design \& Coding, most of the defects found have a relatively small phase delay.
+This is ideal considering that the purpose of the early phase reviews is to remove as many defects as possible.
+
+In the TSP data, there is a weak correlation (Kendall's $\tau = 0.19, p < 0.01, 95\% CI$) between a defect's cost-to-fix (minutes of effort) and the defects phase delay (phase removed - phase injected) - see \fig{correlation}.
+The correlations were also calculated for each phase, e.g., does the phase delay of defects found in Test correlate with the defect's cost-to-fix (\fig{phase_corr}).
 
 We reject $H_2$.
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=3.3in]{corr.png}
+\end{center}
+\caption{Correlation of phase delay and fix time (plot shows a 1\% sampling of data).}
+\label{fig:correlation}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\begin{tabular}{lll}
+  Phase & $tau$ & $p$ \\ 
+  \hline
+  Design & 0.14 & 0.08 \\ 
+  DesignReview & 0.06 & $<$0.01 \\ 
+  DesignInspect & 0.05 & $<$0.01 \\ 
+  Code & 0.27 & $<$0.01 \\ 
+  CodeReview & 0.12 & $<$0.01 \\ 
+  CodeInspect & 0.09 & $<$0.01 \\ 
+  Test & 0.13 & $<$0.01 \\ 
+  IntTest & 0.1 & $<$0.01 \\ 
+  SysTest & 0.25 & $<$0.01 \\ 
+  AcceptTest & 0.06 & 0.29 \\ 
+  \end{tabular}
+\caption{Correlation of cost-to-fix and phase delay for defects found (Kendall's $\tau$)} 
+\label{fig:phase_corr}
+\end{figure}
+
+\begin{figure*}[!ht]
+\centering
+\includegraphics[width=1.6in]{corr-Code.png}
+\includegraphics[width=1.6in]{corr-CodeReview.png}
+\includegraphics[width=1.6in]{corr-CodeInspect.png}
+\vspace{0.01cm}
+
+\includegraphics[width=1.6in]{corr-Design.png}
+\includegraphics[width=1.6in]{corr-DesignReview.png}
+\includegraphics[width=1.6in]{corr-DesignInspect.png}
+\vspace{0.01cm}
+
+\includegraphics[width=1.6in]{corr-Test.png}
+\includegraphics[width=1.6in]{corr-IntTest.png}
+\includegraphics[width=1.6in]{corr-SysTest.png}
+\includegraphics[width=1.6in]{corr-AcceptTest.png}
+
+\caption{Correlation of cost-to-fix (rank) and phase delay for defects found by phase for phases where \# defects >= 30 and \# unique phase delay values >= 5.
+x = phase delay, y = cost-to-fix rank.
+Plots use random samples from each phase for readability.}
+\label{fig:phase_fixtime_plot}
+\end{figure*}
 
@@
 \end{figure}
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{lll}
-  Phase & $tau$ & $p$ \\ 
-  \hline
-BeforeDevelopment & NA & -- \\ 
-  Planning & NA & -- \\ 
-  Reqts & NA & -- \\ 
-  ReqtsReview & 0.05 & 0.01 \\ 
-  ReqtsInspect & 0.17 & $<$0.01 \\ 
-  HLD & NA & -- \\ 
-  HLDReview & 0.04 & 0.24 \\ 
-  HLDInspect & -0.07 & 0.05 \\ 
-  Design & 0.19 & 0.08 \\ 
-  DesignReview & 0.08 & $<$0.01 \\ 
-  DesignInspect & 0.06 & $<$0.01 \\ 
-  Code & 0.21 & $<$0.01 \\ 
-  CodeReview & 0.15 & $<$0.01 \\ 
-  CodeInspect & 0.11 & $<$0.01 \\ 
-  Test & 0.16 & $<$0.01 \\ 
-  QualTest & NA & -- \\ 
-  IntTest & 0.13 & $<$0.01 \\ 
-  SysTest & 0.32 & $<$0.01 \\ 
-  AcceptTest & 0.07 & 0.29 \\ 
-  \end{tabular}
-\caption{Correlation of cost-to-fix and phase delay for defects found in each phase (Kendall's $\tau$)} 
-\label{tab:phase_corr}
-\end{table}
 

### 2015-03-13 data.tex
dropped correaltion scatter plots. replaced pngs with pdfs

@@
 \subsection{Projects in the Sample}
-Since 2004, one of the authors (Nichols) has been been mentoring 171 software development teams around the world as they deploy TSP within their organizations.  
+Since 2004, one of the authors (Nichols) has been been mentoring software development teams and coaches around the world as they deploy TSP within their organizations.  
 The projects
@@
 There were also some  medical devices (US, France, Japan, and Germany),  a few from a commercial 
-computer-aided design system (developed by a world wide distributed team), and less that  
-a dozen United States
-Department of Defence projects.
+computer-aided design system (developed by a world wide distributed team.
  An anonymized version of that data is available in the PROMISE repository\footnote{\carter{Need to add}} and,
@@
 \subsection{Data Collection Process}
-\sei{
-\bi
-    \item What are artifacts uploaded to the SEI tool?
-    \item What are the units contained in those artifacts? Are they the ``issues'' as defined in the following paragraph?
-    \item Is participation in the SEI data collection mandatory? i.e., is the data provided self selected?
-    \item When is the data collected?
-    \item What are screening/cleaning procedures, if any?
-\ei
-}
 
-\sei{ wrn: please review the following}
-
-Organizations using TSP agree to provide project data to the SEI for use in research with the provision that the data must not be traceable to its source.
-The data includes project and site  characteristic summaries, surveys of team members, launch presentations, launch outbriefs, and baseline plans, final data from the project, and a post mortem report.
+Organizations using a TSP agree to provide their project data to the SEI for use in research.
+In return the SEI agrees    that  data must not be traceable to its source.
+The data is collected at major project events, launch, interim checkpoints, and at completion.
+Data includes project and site  characteristic summaries, surveys of team members, launch presentations, launch outbriefs, and baseline plans, final data from the project, and the post mortem report.
 In practice, this data requirement has only been enforced for the purposes of certifying and reauthorizing TSP coaches who are the only individuals authorized to submit data.
@@
 \item
-Several  phases have the similar  sub-phases of {\em review} and {\em inspect}.
-In TSP, individuals perform personal reviews on products prior to the peer inspection.\footnote{\bill{plz distinguish review and inspect.
-does one usually happen first?}}
-\item Testing is divided into several statges \footnote{\bill{
-(this appears to be OBE), need a one
-line description of test vs qualtest vs inttest vs systemTest vs AcceptTest. I tries some words in the Key to \fig{sedata}. dont know if i got it right.}}
+Several  phases in which product is created have sub-phases of {\em review} and {\em inspect} to remove defects.
+TSP uses review when individuals perform personal reviews of their work products prior to the peer review which TSP calls the inspection).
+\item Testing is divided into several stages.
+Developers perform unit test prior to code complete.
+Some standard test phases after code complete include the integration which combines program units into workable system ready for system test.
+Integration and system test are often performed by a separate group.
+
 \ei
 
-\todo{I recommend deleting the rest of this section. Replace with better descriptions of what the phases are.}
+\todo{I recommend deleting the rest of this section. Replace with better descriptions of what the phases are.
+wrn: I agree, I will do this Friday}
 Also shown in \fig{waterfall} (bottom left) is the standard definition of an agile process~\cite{boehmturner03}:

### 2015-03-13 phaseDelay.tex
dropped correaltion scatter plots. replaced pngs with pdfs

@@
     \item What is the effect of phase delay? Check against H2.
+    \item fig12 is great.
+really important.
+its from  different decades, right?
+so can we add text saying that the general pattern is that the (approx) 15 years between each curve tends to a flattening.
+not, perhaps, due to process but to better build environs.
+    \item but fig12 is blurred in the words. recreate using crisper fonts?
+    \item fig 10 text a little blurred. can we re snap that one? or generate it it some higher res form?
+    \item figs 13 (fix time rank v phase delay) is hard to read (text)
 \ei
@@
 \begin{center}
-\includegraphics[width=3.3in]{fix-phase-dist.png}
+\includegraphics[width=3.3in]{fix-phase-dist.pdf}
 \end{center}
@@
     \item What can we say about \fig{raw}
-    \item What can we say about \fig{scale}
-    \item What is the difference between \fig{raw} and \fig{scale}
+    \item What is the difference between \fig{raw}
 \ei
@@
 
- \fig{phase_fixtime_plot} shows scatterplots of the cost-to-fix (rank) per phase against the phase delay.
  In the early phases, i.e., Design \& Coding, most of the defects found have a relatively small phase delay.
@@
 In the TSP data, there is a weak correlation (Kendall's $\tau = 0.19, p < 0.01, 95\% CI$) between a defect's cost-to-fix (minutes of effort) and the defects phase delay (phase removed - phase injected) - see \fig{correlation}.
-The correlations were also calculated for each phase, e.g., does the phase delay of defects found in Test correlate with the defect's cost-to-fix (\fig{phase_corr}).
 
@@
 \begin{center}
-\includegraphics[width=3.3in]{corr.png}
+\includegraphics[width=3.3in]{corr.pdf}
 \end{center}
@@
 
-\begin{figure}[ht]
-\centering
-\begin{tabular}{lll}
-  Phase & $tau$ & $p$ \\ 
-  \hline
-  Design & 0.14 & 0.08 \\ 
-  DesignReview & 0.06 & $<$0.01 \\ 
-  DesignInspect & 0.05 & $<$0.01 \\ 
-  Code & 0.27 & $<$0.01 \\ 
-  CodeReview & 0.12 & $<$0.01 \\ 
-  CodeInspect & 0.09 & $<$0.01 \\ 
-  Test & 0.13 & $<$0.01 \\ 
-  IntTest & 0.1 & $<$0.01 \\ 
-  SysTest & 0.25 & $<$0.01 \\ 
-  AcceptTest & 0.06 & 0.29 \\ 
-  \end{tabular}
-\caption{Correlation of cost-to-fix and phase delay for defects found (Kendall's $\tau$)} 
-\label{fig:phase_corr}
-\end{figure}
+% \begin{figure}[ht]
+% \centering
+% \begin{tabular}{lll}
+%   Phase & $tau$ & $p$ \\ 
+%   \hline
+%   Design & 0.14 & 0.08 \\ 
+%   DesignReview & 0.06 & $<$0.01 \\ 
+%   DesignInspect & 0.05 & $<$0.01 \\ 
+%   Code & 0.27 & $<$0.01 \\ 
+%   CodeReview & 0.12 & $<$0.01 \\ 
+%   CodeInspect & 0.09 & $<$0.01 \\ 
+%   Test & 0.13 & $<$0.01 \\ 
+%   IntTest & 0.1 & $<$0.01 \\ 
+%   SysTest & 0.25 & $<$0.01 \\ 
+%   AcceptTest & 0.06 & 0.29 \\ 
+%   \end{tabular}
+% \caption{Correlation of cost-to-fix and phase delay for defects found (Kendall's $\tau$)} 
+% \label{fig:phase_corr}
+% \end{figure}
 
-\begin{figure*}[!ht]
-\centering
-\includegraphics[width=1.6in]{corr-Code.png}
-\includegraphics[width=1.6in]{corr-CodeReview.png}
-\includegraphics[width=1.6in]{corr-CodeInspect.png}
-\vspace{0.01cm}
+% \begin{figure*}[!ht]
+% \centering
+% \includegraphics[width=1.6in]{corr-Code.png}
+% \includegraphics[width=1.6in]{corr-CodeReview.png}
+% \includegraphics[width=1.6in]{corr-CodeInspect.png}
+% \vspace{0.01cm}
 
-\includegraphics[width=1.6in]{corr-Design.png}
-\includegraphics[width=1.6in]{corr-DesignReview.png}
-\includegraphics[width=1.6in]{corr-DesignInspect.png}
-\vspace{0.01cm}
+% \includegraphics[width=1.6in]{corr-Design.png}
+% \includegraphics[width=1.6in]{corr-DesignReview.png}
+% \includegraphics[width=1.6in]{corr-DesignInspect.png}
+% \vspace{0.01cm}
 
-\includegraphics[width=1.6in]{corr-Test.png}
-\includegraphics[width=1.6in]{corr-IntTest.png}
-\includegraphics[width=1.6in]{corr-SysTest.png}
-\includegraphics[width=1.6in]{corr-AcceptTest.png}
+% \includegraphics[width=1.6in]{corr-Test.png}
+% \includegraphics[width=1.6in]{corr-IntTest.png}
+% \includegraphics[width=1.6in]{corr-SysTest.png}
+% \includegraphics[width=1.6in]{corr-AcceptTest.png}
 
-\caption{Correlation of cost-to-fix (rank) and phase delay for defects found by phase for phases where \# defects >= 30 and \# unique phase delay values >= 5.
+% \caption{Correlation of cost-to-fix (rank) and phase delay for defects found by phase for phases where \# defects >= 30 and \# unique phase delay values >= 5.
 x = phase delay, y = cost-to-fix rank.
 Plots use random samples from each phase for readability.}
-\label{fig:phase_fixtime_plot}
-\end{figure*}
+% \label{fig:phase_fixtime_plot}
+% \end{figure*}
 
@@
 
- \begin{figure}[!t]
-\renewcommand{\baselinestretch}{0.7}
-\scriptsize
-\begin{center}
-\begin{tabular}{l@{~~}|l@{~}|r@{~}|r@{~}r@{~}|r@{~}l}
-           \multicolumn{2}{c}{~}                 &  &\multicolumn{2}{c|}{median}\\
-  Injection&   Removal& $n$ & initial & now & \multicolumn{2}{l}{Scale up, w.r.t. initial}
-\input{deltas}
-\end{tabular}
-\end{center}
-\caption{50th percentile (median) scale ups  for  time to resolve issues (taken from \fig{raw}).}
-\label{fig:scale}
-\end{figure}
+ %\begin{figure}[!t]
+%\%renewcommand{\baselinestretch}{0.7}
+%\scriptsize
+%\begin{center}
+%\begin{tabular}{l@{~~}|l@{~}|r@{~}|r@{~}r@{~}|r@{~}l}
+%%           \multicolumn{2}{c}{~}                 &  &\multicolumn{2}{c|}{median}\\
+ % Injection&   Removal& $n$ & initial & now & \multicolumn{2}{l}{Scale up, w.r.t. initial}
+%\input{deltas}
+%\end{tabular}
+%%\end{center}
+%\caption{50th percentile (median) scale ups  for  time to resolve issues (taken from \fig{raw}).}
+%%\label{fig:scale}
+%\end{figure}
 
-\subsection{Phase Delay in Slower Bugs?}
-In these results, we checked if a small number of bugs are most expensive. 
-To check this:
-\bi
-\item
-We repeated the analysis that generated \fig{scale},
-but instead of looking at the 50th percentile, we displayed the scale up factors
-\item 
-We only
-checked the Design and Code scale up results since, from \fig{scale}, it is clear these
-have the most examples of longest phase delays.
-\ei 
-The  results are shown in \fig{scale90} and these
-results are somewhat different for Design and Coding issues: 
-\bi 
-\item For Design issues,  these have the same
-general form as the 50th percentile results. That is, while it it certainly faster
-to remove thing sin the phase where they are created, once we leave that phase
-it does not seem to matter much how many phases we wait before fixing the issue.
-\item
-\item For Coding issues, the seems little impact of phase delay on the time
-required to fix.
-\ei
+%\subsection{Phase Delay in Slower Bugs?}
+%In these results, we checked if a small number of bugs are most expensive. 
+%To check this:
+%\bi
+%\item
+%We repeated the analysis that generated \fig{scale},
+%but instead of looking at the 50th percentile, we displayed the scale up factors
+%\item 
+%We only
+%checked the Design and Code scale up results since, from \fig{scale}, it is clear these
+%have the most examples of longest phase delays.
+%\ei 
+%The  results are shown in \fig{scale90} and these
+%results are somewhat different for Design and Coding issues: 
+%\bi 
+%\item For Design issues,  these have the same
+%general form as the 50th percentile results. That is, while it it certainly faster
+%to remove thing sin the phase where they are created, once we leave that phase
+%it does not seem to matter much how many phases we wait before fixing the issue.
+%\item
+%\item For Coding issues, the seems little impact of phase delay on the time
+%required to fix.
+%\ei
 
-\begin{figure}[!t]
-\renewcommand{\baselinestretch}{0.7}
-\scriptsize
-\begin{center}
-\begin{tabular}{l@{~~}|l@{~}|r@{~}|r@{~}r@{~}|r@{~}l}
-          \multicolumn{2}{c}{~}                 &  &\multicolumn{2}{c|}{99th }\\
-           \multicolumn{2}{c}{~}                 &  &\multicolumn{2}{c|}{percentile }\\
-  Injection&   Removal& $n$ & initial & now & \multicolumn{2}{l}{Scale up, w.r.t. initial}
-\input{deltasBigBig}
-\end{tabular}
-\end{center}
-\caption{99th percentile scale ups.}
-\label{fig:scale99}
-\end{figure}
+%\begin{figure}[!t]
+%\renewcommand{\baselinestretch}{0.7}
+%\scriptsize
+%\begin{center}
+%\begin{tabular}{l@{~~}|l@{~}|r@{~}|r@{~}r@{~}|r@{~}l}
+%          \multicolumn{2}{c}{~}                 &  &\multicolumn{2}{c|}{99th }\\
+%           \multicolumn{2}{c}{~}                 &  &\multicolumn{2}{c|}{percentile }\\
+%  Injection&   Removal& $n$ & initial & now & \multicolumn{2}{l}{Scale up, w.r.t. initial}
+%\input{deltasBigBig}
+%\end{tabular}
+%\end{center}
+%%\caption{99th percentile scale ups.}
+%\label{fig:scale99}
+%\end{figure}
 
-\begin{figure}[!t]
-\renewcommand{\baselinestretch}{0.7}
-\scriptsize
-\begin{center}
-\begin{tabular}{l@{~~}|l@{~}|r@{~}|r@{~}r@{~}|r@{~}l}
-           \multicolumn{2}{c}{~}                 &  &\multicolumn{2}{c|}{mean}\\
-  Injection&   Removal& $n$ & initial & now & \multicolumn{2}{l}{Scale up, w.r.t. initial}
-\input{deltasMean}
-\end{tabular}
-\end{center}
-\caption{Mean  scale ups.}
-\label{fig:scale90}
-\end{figure}
+%\begin{figure}[!t]
+%\renewcommand{\baselinestretch}{0.7}
+%\scriptsize
+%\begin{center}
+%\begin{tabular}{l@{~~}|l@{~}|r@{~}|r@{~}r@{~}|r@{~}l}
+ %          \multicolumn{2}{c}{~}                 &  &\multicolumn{2}{c|}{mean}\\
+ % Injection&   Removal& $n$ & initial & now & \multicolumn{2}{l}{Scale up, w.r.t. initial}
+%\input{deltasMean}
+%\end{tabular}
+%\end{center}
+%\caption{Mean  scale ups.}
+%\label{fig:scale90}
+%\end{figure}
 
@@
 Foundation grants NSF-CISE 1302169 and CISE 1302169. 
+\\
+We thank Tuma Solutions for providing the Team Process Data Warehouse software in support of this research.
 

### 2015-03-13 lucas.tex
fixed up various charts and tables. added median phase delay per phase

@@
 
-Summary statistics for the agreement and applicability scores for the RqtsErr and PhaseDelay laws are presented in Figure~\ref{fig:survey_results} \lucas{To add results from other laws}.
+Summary statistics for the agreement and applicability scores for the RqtsErr and PhaseDelay laws are presented in Figure~\ref{fig:survey_results}.
 Responses whose \textbf{applicability} response was ''I don't know'' are ommitted from analysis.
@@
 \begin{center}
-\begin{tabular}{c|c|c|c|c|c|c}
-Survey & law & N & \multicolumn{2}{c}{agreement} & \multicolumn{2}{c}{applicability} \\ 
- & & mean & median & mean & median \\
+\subcaption{Practitioner survey}
+\begin{tabular}{l|c|c|c|c|c}
+Law & N & \multicolumn{2}{c}{agreement} & \multicolumn{2}{c}{applicability} \\ 
+ & & $\bar{x}$ & $\tilde{x}$ & $\bar{x}$ & $\tilde{x}$ \\
 \hline 
-Practitioners & RqtsErr & 16 & 4.6 & 5 & 4.2 & 4 \\ 
-Researchers & RqtsErr & 30 & 3.8 & 4 & 3.8 & 4 \\ 
-Researchers & PhaseDelay & 30 & 3.6 & 4 & -- & -- 
+\textbf{Rqts errors are most expensive...} & 16 & 4.6 & 5 & 4.2 & 4 \\ 
+Process maturity improves output & 17 & 4.4 & 4 & 3.9 & 4 \\ 
+Most time is spent removing errors & 16 & 4.1 & 4 & 3.9 & 4 \\ 
+Missing reqts are hardest to fix & 17 & 4.1 & 4 & 4.0 & 4 \\
+Reuse increases prod. and qual. & 16 & 3.9 & 4 & 3.8 & 4 \\
+OO-programming reduces errors & 13 & 3.9 & 4 & 3.5 & 4 \\
+80-20 rule (defects to modules) & 12 & 3.8 & 4 & 4.0 & 4 \\
+Adding manpower to a late project & 15 & 3.7 & 4 & 3.7 & 4 \\
+Inspections can remove 90\% of defects & 18 & 3.3 & 4 & 3.5 & 4 \\
+Smaller changes have higher error density & 14 & 2.8 & 3 & 3.4 & 3.5 \\
+A developer is unsuited to test own code & 17 & 2.6 & 3 & 3.5 & 4
 \end{tabular} 
+\bigskip
+\subcaption{Researcher survey}
+\begin{tabular}{l|c|c|c|c|c}
+Law & N & \multicolumn{2}{c}{agreement} & \multicolumn{2}{c}{applicability} \\ 
+ & & $\bar{x}$ & $\tilde{x}$ & $\bar{x}$ & $\tilde{x}$ \\
+\hline 
+Process maturity improves output & 4 & 4.0 & 4 & 4.0 & 4 \\
+\textbf{Rqts errors are most expensive...} & 30 & 3.8 & 4 & 3.8 & 4   \\ 
+80-20 rule (defects to modules) & 6 & 3.7 & 4 & 3.5 & 4 \\
+\textbf{PhaseDelay} & 30 & 3.6 & 4 & -- & --  \\ 
+Missing reqts are hardest to fix & 7 & 3.6 & 4 & 3.7 & 4 \\
+Adding manpower to a late project & 4 & 3.5 & 4 & 3.5 & 4 \\
+Inspections can remove 90\% of defects & 7 & 3.4 & 4 & 3.6 & 3 \\
+Reuse increases prod. and qual. & 6 & 3.3 & 4 & 3.0 & 4 \\
+OO-programming reduces errors & 6 & 3.2 & 4 & 2.7 & 3 \\
+Most time is spent removing errors & 6 & 2.8 & 3 & 3.2 & 4 \\ 
+Smaller changes have higher error density & 4 & 2.8 & 3 & 4.0 & 4 \\
+A developer is unsuited to test own code & 7 & 2.1 & 2 & 2.4 & 3
+\end{tabular}
+
 \end{center}
-\caption{Agreement and applicability of phase delay related axioms.}
+\caption{Agreement and applicability of software engineering axioms.}
 \label{fig:survey_results}

### 2015-03-13 phaseDelay.tex
fixed up various charts and tables. added median phase delay per phase

@@
 
-\usepackage{subfigure}
+\usepackage{tabularx}
+\usepackage{subcaption}
 \usepackage{picture}
@@
     \item What is the effect of phase delay? Check against H2.
-    \item fig12 is great.
+    \item \fig{cost-to-fix-tsp} is great.
 really important.
@@
 not, perhaps, due to process but to better build environs.
-    \item but fig12 is blurred in the words. recreate using crisper fonts?
-    \item fig 10 text a little blurred. can we re snap that one? or generate it it some higher res form?
-    \item figs 13 (fix time rank v phase delay) is hard to read (text)
 \ei
@@
 \begin{figure}[!ht]
- \includegraphics[width=3.3in]{boehm_overlay.png}
+ \includegraphics[width=3.3in]{boehm-overlay.pdf}
  \caption{Cost-to-fix curve from \fig{cost-to-fix} overlayed with TSP data and case study from~\cite{Royce98}.}
@@
     \item What can we say about \fig{raw}
-    \item What is the difference between \fig{raw}
 \ei
@@
 
- In the early phases, i.e., Design \& Coding, most of the defects found have a relatively small phase delay.
-This is ideal considering that the purpose of the early phase reviews is to remove as many defects as possible.
+We reject $H_2$.
 
-In the TSP data, there is a weak correlation (Kendall's $\tau = 0.19, p < 0.01, 95\% CI$) between a defect's cost-to-fix (minutes of effort) and the defects phase delay (phase removed - phase injected) - see \fig{correlation}.
+\fig{phase_delay_medians} and \fig{correlation} show that most defects have a relatively small phase delay in our project sample (phase delay is often <= 2).
+As intended, the Review and Inspection Phases for Requirements, Design, and Code mostly find defects originating in the preceding phase.
+In contrast, the Testing phases mostly find defects originating in the Code phase.
 
-We reject $H_2$.
+In \fig{raw}, we present the median fix times of defects based on when they were injected and ultimately removed.
+Defects injected Before Development follow the most severe scale up as development phases pass, but not on the exponential scale traditionally seen (\fig{cost-to-fix}).
+Defects injected in Requirements, Design, and Coding phases exhibit virtually no scale-up regardless of the phase in which they are fixed.
+
+%In the TSP data, there is a weak correlation (Kendall's $\tau = 0.19, p < 0.01, 95\% CI$) between a defect's cost-to-fix (minutes of effort) and the defects phase delay (phase removed - phase injected).
 
@@
 \label{fig:correlation}
+\end{figure}
+
+\begin{figure}[!ht]
+\centering
+\begin{tabular}{rrr|l}
+    \multicolumn{3}{c|}{Percentiles}\\ 
+  25th & 50th & 75th & Phase removed \\ 
+  \hline
+    2 & 2 & 2 & ReqtsInspect \\ 
+    1 & 1 & 1 & ReqtsReview \\ 
+    2 & 2 & 2 & HLDInspect \\ 
+    1 & 1 & 1 & HLDReview \\ 
+    2 & 2 & 2 & DesignInspect \\ 
+    1 & 1 & 1 & DesignReview \\ 
+    2 & 2 & 2 & CodeInspect \\ 
+    1 & 1 & 1 & CodeReview \\ 
+    3 & 3 & 6 & UnitTest \\ 
+    4 & 4 & 4 & IntTest \\ 
+    5 & 5 & 5 & SysTest \\ 
+    6 & 6 & 6 & AcceptTest \\ 
+  \end{tabular}
+\caption{Median phase delay of defects removed per phase} 
+\label{fig:phase_delay_medians}
 \end{figure}

### 2015-03-13 data.tex
upgraded pickture of Royce case study

@@
 \sei{One paragraph summary on the key points of TSP execution? The following text should also be reviewed and better integrated.}
+
 TSP guiding principles include 1) engineering work does not happen by accident, it must be planned, 2) the plans must include realistic commitments of the resources needed to perform the work, 3) planed and actual results should be frequently examined to assure that plans remain relevant, and 4) people performing the work should be responsible for planning and tracking the work.

### 2015-03-14 data.tex
small edits

@@
-We examine the phase delay effect bin
-212 software projects from around the world from 
-\bill{2005 to 2015} that used the Team Software Process (TSP).
+To empirically test this heuristic, we examined the phase delay effect in
+171 software projects conducted between 2005 and 2014 from around the world 
+that used the Team Software Process (TSP).
 
@@
 
-\sei{One paragraph summary on the key points of TSP execution? The following text should also be reviewed and better integrated.}
-
-TSP guiding principles include 1) engineering work does not happen by accident, it must be planned, 2) the plans must include realistic commitments of the resources needed to perform the work, 3) planed and actual results should be frequently examined to assure that plans remain relevant, and 4) people performing the work should be responsible for planning and tracking the work.
-After clarifying goals, defining work practices, and estimating size and effort, developers make a detailed short term plan.
+TSP guiding principles include 1) sound engineering work does not happen by accident, it must be planned, 2) performance of the plans requires realistic commitments of the resources needed to perform the work, 3) the planned work should be compared to actual frequently to assure plans remain relevant, and 4) people performing the work should be responsible for planning and tracking the work.
+TSP helps developers by providing a set of measures that can be applied to managing tasks, quality, and schedule.
+Planning begins by quantifying goals, defining work practices, and estimating size and effort, developers make a detailed short term plan.
 As the developers perform project work, they use a tool such as the Process Dashboard to collect their time, effort, size, and schedule data.
-Every week, the team review their data to evaluate status, review actual rates, determine if project goals for schedule, cost, and quality are being met and then make necessary plan corrections.
+Every week, the team reviews their data to evaluate status, review actual rates, determine if project goals for schedule, cost, and quality are being met and then make necessary plan corrections.
 At the end of the project the coach and team perform a quantitative post mortem.
 
-Four common features of TSP projects include (1)~coaches,
-(2)~peer review, and  (3)~personnel reviews (4)~planning.
+Common features of TSP projects include planning, personal reviews,  peer inspections, and coaching.
 
-A ``coach'' works with the team and is authorized to submit project data.
+A TSP Coach helps the team to plan and analyze performance. The Coach is the only role authorized to submit project data to the SEI.
 Before reviewing data with the teams, therefore before submission,
@@
 
-``Personnel review'' is a technique taken  from the Personal Software
+Planning follows the technique taught in the PSP(SM).
+Developers estimate the size of work products and convert this to a total effort using historical rates.
+Time in specific tasks come from the  process phases and historical percent time in phase distributions.
+Defects are estimated using historical phase injection rates and phase removal yields.
+Coaches help the developers to compare estimates against actual results.
+In this way, developers acquire a more realistic understanding of their work behavior, performance, and schedule status.
+
+Personal review is a technique taken  from the Personal Software
 ProcessSM (PSPSM) and is more or less unique in its TSP implementation.
-The technique of a personal review is similar to that of a peer review.
-Developers systematically examine their own work products using a checklist built from their personal defect profile.
-This personal review occurs after a portion of a product is done and before peer reviews.
+Developers follow a systematic process to remove defects by  examining their own work products using a checklist built from their personal defect profile.
+This personal review occurs after some product or part of a product is considered to constructed and before peer reviews or test.
 
@@
 
-``Peer review'' is a  technique in
-traditional software engineering.
+Peer inspection is a  technique in
+traditional software engineering and is often called peer review.
  Basili and Boehm write  commented in 2001~\cite{boehm01} 
 that peer reviews can catch over half the defects introduced into a system.
-Peer review can be conducted on any artifact generated anywhere in the software
+Peer inspection can be conducted on any artifact generated anywhere in the software
 lifecycle and can quickly be adapted to new kinds of artifacts.
@@
 
-''Planning '' is another technique from the in the PSP(SM).
-Developers estimate the size of work products, define the process phases, use historic rates and time in phase to estimate the required time, the number of defects injected, and the defects to be removed in each phase.
-Coaches help the developers to compare estimates against actual results.
-In this way, developers acquire a more realistic understanding of their work behavior and performance.
-
 \subsection{Projects in the Sample}
-Since 2004, one of the authors (Nichols) has been been mentoring software development teams and coaches around the world as they deploy TSP within their organizations.  
+Since shortly before the year 2000, the SEI has been teaching and coaching TSP teams.
+One of the authors (Nichols) has been been mentoring software development teams and coaches around the world as they deploy TSP within their organizations since 2006.
+The earliest projects in our sample began work in 2004, the most recent completions were in 2014.
 The projects
 were mostly small to medium, with a median duration of 46 days
-and a maximum duration of 90 days. Median team size was 7
-people, with a maximum of 40. Hence,  total development 
-effort\footnote{
-Effort =  duration*teamSize/workDaysPerYear.} for
-these projects ranged from 1.3 years (median) to 14.3 years (max).
-The majority of
-the projects were either web portals or banking systems in the US, South Africa, and Mexico. 
+and a maximum duration of 90 days in major increments. Several projects extended for many incremental cycles. Median team size was 7
+people, with a maximum of 40. 
+;Hence,  total development 
+;effort\footnote{
+;Effort =  duration*teamSize/workDaysPerYear.} for
+;these projects ranged from 1.3 years (median) to 14.3 years (max).
+Many of
+the projects were e-commerce web portals or banking systems in the US, South Africa, and Mexico. 
 There were also some  medical devices (US, France, Japan, and Germany),  a few from a commercial 
-computer-aided design system (developed by a world wide distributed team.
+computer-aided design system (developed by a world wide distributed team and embedded systems.
+
  An anonymized version of that data is available in the PROMISE repository\footnote{\carter{Need to add}} and,
@@
 
-\bill{can we offer any more deatils here?}
-
 \subsection{Data Collection Process}
 
-Organizations using a TSP agree to provide their project data to the SEI for use in research.
+Organizations using TSP agree to provide their project data to the SEI for use in research.
 In return the SEI agrees    that  data must not be traceable to its source.
-The data is collected at major project events, launch, interim checkpoints, and at completion.
-Data includes project and site  characteristic summaries, surveys of team members, launch presentations, launch outbriefs, and baseline plans, final data from the project, and the post mortem report.
-In practice, this data requirement has only been enforced for the purposes of certifying and reauthorizing TSP coaches who are the only individuals authorized to submit data.
-To this data, only the data recorded using the Process Dashboard tool has been collected and aggregated for research.
+The data is collected at major project events; launch, interim checkpoints, and at project completion.
+Data includes project and site  characteristic summaries, surveys of team members, launch presentations, launch outbriefs, and baseline plans, final data from the project, and the project post mortem report.
+In practice, this data requirement has only been enforced for the purposes of certifying and reauthorizing TSP coache who must submit data to maintain their authorization.
+Coaches are certified by demonstrating competent use of the TSP process with the artifacts and data not by the actual project results.
+Of the data submitted, only the data recorded using the Process Dashboard tool has been collected and aggregated for this research.
 
-To aid in using this data for research and benchmarking  Nichols first collected the Process Dashboard data from the submission repository then used a tool built by the Process Dashboard developer, David Tuma,
-to gather the project data into a database at SEI.
-Nichols and Shirai then extracted data into views suitable for analysis.
-.
+To use this data for research and benchmarking  Nichols first collected the Process Dashboard data from the submission repository then used a tool built by the Process Dashboard developer, David Tuma,
+to gather the project data into a database.
+Nichols and his colleague Yasutaka Shirai then extracted data into views suitable for analysis.
 The key views include the a project results summary, effort logs, task logs, and defect logs.
 At this stage, the data has not yet undergone additional screening.
+\sei{add cite to ICSSP paper} An analysis of the error rate in the data was reported in a previous work.
 
-\sei{
-do we need the following information? we do not use it in this analysis
-}
+The tool collected process information about those tasks and defects that we analyze in this work. 
+That task data includes  work start time, work end time, delta
+work time, and interruption time. Software engineers are often
+interrupted by meetings, requests for technical help, reporting, and
+so forth. These events are recorded, in minutes, as interruption
+time.
 
-The issue types collected in the the SEI TSP data set divide into:
+ Defect logs include the time and date a defect was discovered, the phase in which that defect was injected, the phase in which it was removed, the time (in minutes) required to find and fix the defect, and the categorical type.
+
+The most common defect types used in the the SEI TSP data set are as follows:
 \be 
@@
 \ee
-The tool also collected process information about those defects.
-That data includes  work start time, work end time, delta
-work time, and interruption time. Software engineers are often
-interrupted by meetings, requests for technical help, reporting, and
-so forth. These events are recorded, in minutes, as interruption
-time. In this paper, when we report ``time to resolve an
+
+As of November 2014, the SEI TSP database contained data from 212
+TSP projects. The projects completed between July 2006 and
+November 2014; they included 47 organizations and 843 people. 
+The database fact tables
+contain 268,726 time logs, 
+154,238 task logs,
+ 47,376 defect logs, 
+and 26,534 size logs.
+
+In this paper, when we report ``time to resolve an
 issue'', we show the difference between the start and end times
@@
 difference in times, minus the interruptions).
-
-%wrn, updated 3/11 these
-As of February 2015, the SEI TSP database contained data from 212
-TSP projects. The projects completed between July 2008 and
-November 2014; they included 47 organizations and 843 people. Among
-the database fact tables, 
-contains 268,726 time logs, 
-154,238 task logs,
- 47,376 defect logs, 
-and 26,534 size logs
-%\footnote{\bill{we need toupdate this table}}.
 
@@
 This paper studies the impact of phase delay on the time required to resolve issues.
-The logical phases used in this paper are shown in \fig{waterfall}. Note that, in that figure:
+The logical phases used in this paper are shown in \fig{waterfall}.
+Although the representation suggests a waterfall, the phases represent the primary stages through which requirements are reified into working and tested code.
+That is, all requirements must be stated in some way, implemented in code, integrated and tested.
+All real implementations of any size follow a spiral approach with many team performing the work in iterative and/or incremental development cycles.
+Within an project increment, multiple features or components may be developed and incremented.
+TSPSM is compatible with agile and encourages iterative and incremental development.
+Nonetheless, the specific strategy and cycle duration is a project decision.
+TSPSM does, however, strongly encourage 1) constructing units of sufficient size that measurement is practicable, and 2) separating the construction from appraisal and test activities.
+This effectively highlights the separation of construction from rework activities and aids the apportionment of defects to those found in appraisal activities (reviews and inspections) and those found through failure (test).
+The distinct construction activities (requirements, high and detailed design, and code) were chosen to help teams analyze the effectiveness and efficiency of their practices through analysis of the defect phase origin, type, fix effort, and phase of discovery.
+
+Note that, in that figure:
 \bi 
@@
 Integration and system test are often performed by a separate group.
-
 \ei
 
-\todo{I recommend deleting the rest of this section. Replace with better descriptions of what the phases are.
-wrn: I agree, I will do this Friday}
-Also shown in \fig{waterfall} (bottom left) is the standard definition of an agile process~\cite{boehmturner03}:
-Given some
-some backlog of tasks, wgeb teams complete their current tasks, they select the next task(s) to complete. That selection process may use a variety of criteria
-to prioritize which  tasks are selected (for more details on that selection process, see~\cite{me09j,port08,boehmturner03}). Tasks are completed in ``sprints'' that can last hours,
-days, but rarely not more than weeks. Each day meet for brief ``scrum'' sessions to assess (and possibly alter) their current progress on the goals of the sprint.  
-Agile teams race to generate releases
-(in the continuous release model, releases can be generated on a daily basis, or even faster).  
-Experience gained from those releases informs the discussion in the daily scrums which, in turn,
-can inform the team's decisions on how to select and implement the next set of tasks for next sprint.
-In fact, that experience can result in changing some/all of the tasks in the current backlog.
-
-TSP is not antithetical to agile--  indeed a TSP-waterfall style project can adopt aspects
-of agile.  For example the agile loop  could be applied
-over one or more of the phases shown in the long waterfall chain of \fig{waterfall}. Some
-TSP teams adopt something like test-driven-development\footnote{TDD is an agile-method
-where the ``test'' is the primary driver of the design. The tests are written first,
-then the code to support those tests. TDD proceeds in three steps: red (where there
-are broken tests); green (where tests are passing); and, possibly, refactor (where
-the code is re-organized based on feedback from those tests~\cite{fraser03}.} where
-code inspections are scheduled after unit testing. Some of the groups in the SEI TSP data used that approach but
-they do not effect the main conclusions of this paper:
+The principles followed are that 
 \bi 
-\item They were less than 5\% of the total \bill{any numbers on this?};
-
-\item Most of our phase delay data comes from much early in the waterfall model
-of \fig{waterfall}.
-\ei 
-That said, there are some very ``un-agile'' aspects
-of the processes used by the   projects in the TSP SEI data.
-A TSP participant spends much time reflecting on the project,
-undisturbed by the behaviour of the  executables.
-In fact,
-it can be days/weeks
-before TSP participants gain  feedback from executing code, for the following reasons:  
-\bi 
-\item
-For TSP projects, the times spent in the of a design activity/phase is   
-approximately as much effort as the coding phase.  
-\item Our TSP teams did not
- combine development and  testing. 
- \item  The TSP projects studied here typically offer new releases at least every three months, but
-often much longer that that\bill{any numbers on that}. Note that this is a far slower
-release schedule than seen in, say, continuous integration projects.  
-\item The time devoted to personnel review'' and peer review (defined above)
-is  about 50\% as much effort as the previous construction phase activity.
-
+\item defects should be found before test, that is, test defects should be infrequent
+\item a product should be inspected before used in a subsequent construction phase to minimize  unnecessary rework
+\item a personal review should remove the most common defects before being given to peers for a detailed inspection
+\item because different categories of test will find different types of defect, and follow in a natural sequence they should be measured separately
 \ei
 
-Having presented all that, we can now present the main point of this section:
-\bi 
-\item The results section of this paper fails to find that  phase delay causing dramatic increases
-in time to fix issues;
-\item This lack-of-phase-delay effect {\em cannot} be explained away just by  saying that
-the projects in this sample adopted something like  agile methods to reduce re-work costs.
-\ei
+In summary, the TSPSM process and measurement have been specifically designed to aid the analysis of defects for use in process improvement.
+The data is, therefore, uniquely suitable for an analysis of the phase delay effect.

### 2015-03-14 lucas.tex
small edits

@@
 While the practitioners strongly believed in the law, researchers were less firm.
-In the free response texts, most participants agreed that addressing errors may mean system redesign unless the system and the affected requirement are simple/trivial.
+In the free response texts, most participants agreed that addressing errors late may mean system redesign unless the system and the affected requirement are simple/trivial.
 Further, any changes to the production system have additional complexity (compared to changes while system in development) due to users, data existence, and architecture dependencies.
@@
 Overall, the RqtsErr law was the most agreed upon and most applicable law of 11 surveyed amongst practitioners, and the second most agreed upon law amongst researchers.
-The survey results seem to confirm that phase delay escalates cost-to-fix is a widely-held belief in software engineering.
+The survey results seem to confirm that the notion that phase delay escalates cost-to-fix is a widely-held belief in software engineering.

### 2015-03-14 phaseDelay.tex
upgraded pickture of Royce case study
small edits

@@
 Tim Menzies, \\Carter Pape\\
-       \affaddr{CS, NcState, USA}\\
+       \affaddr{Dept. of Computer Science\\
+       North Carolina State U, USA}\\
        tim.menzies@gmail.com,\\carterpape@gmail.com
 \alignauthor
-William Nichols,\\ Forrest Schull\\
-\affaddr{SEI, CMU, USA}
+William Nichols,\\ Forrest Shull\\
+\affaddr{Software Engineering Institute\\
+        Carnegie Mellon U, USA}
 wrn,fjshull@sei.cmu.edu
@@
 Lucas \\Layman\\
-       \affaddr{Fraunhofer Center,  USA}\\ 
+       \affaddr{Fraunhofer CESE, \\
+                College Park, USA}\\ 
        llayman@fc-md.umd.edu
@@
 Does
-repair time increases dramatically
+repair time increase dramatically
 the longer a defect persists in a system?
-This  is very widely belief that is used to justify 
-radical changed to    traditional software development
- processes.
+This widely held belief is used to justify 
+major investments in  software development
+ processes and methods.
 
@@
 Yet based on a sample of 
-212 software projects from around the world from 
-\bill{2005 to 2014}, we argue that this belief is mostly 
+171 software projects from around the world from 
+and completed between 2006 and 2014, we argue that this belief is mostly 
 incorrect. Specifically: the median fix time for issues
@@
 to continually revisit old conclusions~\cite{prasad13}. Software engineering should do the same lest
-ill-founded  exert an undue influence on how we build software.
+ill-founded guidance exert an undue influence on how we build software.
 
@@
 the {\em phase delay effect}
-i.e.,  the time required to resolve an issue in a software project increases dramatically 
+i.e.,  the rule of thumb stating that the time required to resolve an issue in a software project increases dramatically 
 the longer it remains in the system. 
@@
 independent verification \& validation. IVV has been criticized as
-ineffective since IVV engineers, who many  unfamiliar with the details of the project, are required to pass judgement on the code.
+ineffective since IVV engineers, many of whom are unfamiliar with the details of the project, are required to pass judgement on the code.
 Hence,  the IVV team may find the same or fewer bugs
-as the developers,
+as the developers.
 On the other hand, IVV has been defended
-as follows:  when the IVV team finds bugs earlier that the developers then, due to the phase delay effect, this saves the overall cost of the project.
+as follows:  when the IVV team finds bugs earlier than the developers then, due to the phase delay effect, this provides overall cost savings to the project.
 Note that the strength of this argument diminishes
@@
 First, we conduct a survey of practitioners and researchers to gauge their belief in the phase delay effect.
-Second, we  analyze 171 software  projects developed in the period \bill{2004 to 2014}.
+Second, we  analyze 171 software  projects developed in the period 2006 to 2014.
 To the best of our knowledge, this is the largest study on phase delay yet seen in the literature.
@@
 \section{Motivation}
-\subsection{Why Reassess Existing Triusms?}
+\subsection{Why Reassess Existing Truisms?}
 
@@
 principles of software
-engineering are common to both research and practice. Processional societies assume such generalities exist when they offer
+engineering are common to both research and practice. Professional societies assume such generalities exist when they offer
  lists of supposedly general ``best practices'' such as
@@
 lower variance in the predictions).
- \item 
-Other work studies  found dozens of single
-models that performed worse that {\em committees of models}, all with different opinions,
-that contribute to an {\em ensemble} prediction of software development effort~\cite{kocaguneli2012value,azhar13}.
+% \item 
+% Other work studies  found dozens of single
+% models that performed worse that {\em committees of models}, all with % different opinions,
+% that contribute to an {\em ensemble} prediction of software 
+% development effort~\cite{kocaguneli2012value,azhar13}.
 \ei
@@
 In summary, just like medicine, our field suffers when
- software engineers do  revise old beliefs.  Therefore, it is important
+ software engineers do  not revise old beliefs.  Therefore, it is important
  to regularly  reexamine    old beliefs.
@@
 Beck speculated that there might be another kind of project, shown in \fig{curve2},
-that does no exhibit phase delay. In this next quote, Beck clearly states
+that does not exhibit phase delay. In this next quote, Beck clearly states
 the importance of avoiding phase delay (in fact, for Beck, it is {\em the} major
@@
 
-XXX clumsy link here
-That said, where our results
-diverge from Beck is that we have observed this lack of phase delay in  projects that are not agile. Those
-projects are described below.
+%XXX clumsy link here
+%That said, where our results
+%diverge from Beck is that we have observed this lack of phase delay in  projects that are not agile. Those
+%projects are described below.
 
@@
 
-The distribution of defects found and fixed per phase is shown in Figure~\ref{fig:fix-phase-dist}.
+Our first hypothesis was that defects found in later phases tend to require more time to fix than in earlier phases.
+
+The distribution of defects found and fixed per phase in our data is shown in Figure~\ref{fig:fix-phase-dist}.
 A high percentage of defects were found and fixed in the early phases, i.e., requirements, high level design, and design reviews and inspections.
@@
 
-The relative cost to fix errors per lifecycle phase in the TSP project data is overlayed with Boehm's exponential cost curve~\cite{Boehm81} as well as the Royce case study~\cite{Royce98} in \fig{cost-to-fix-tsp}.
+For illustrative purposes, the relative cost to fix errors per lifecycle phase in the TSP project data is overlayed with Boehm's cost curve~\cite{Boehm81} as well as the CCPDS-R case study~\cite{Royce98} in \fig{cost-to-fix-tsp}.
+It is clear to see that the trend is relatively flat, closer to the pattern exhibited by the Royce case study (which also focused on early defect removal) than the exponential growth curve.
 
@@
  \includegraphics[width=3.3in]{boehm-overlay.pdf}
- \caption{Cost-to-fix curve from \fig{cost-to-fix} overlayed with TSP data and case study from~\cite{Royce98}.}
+ \caption{Cost-to-fix curve from \fig{cost-to-fix} overlayed with case study from~\cite{Royce98} and TSP data. Scale is relative to the phase with the lowest cost to fix. }
  \label{fig:cost-to-fix-tsp}
@@
 
-We reject $H_2$.
+Our second hypothesis was that defects that exhibit higher phase delay (i.e.
+that persist in the software over more lifecycle phases) tend to require more effort to fix than those with less phase delay.
+To investigate this, we calculate phase delay as simply the number of project phases (the set used in our data set is shown in \fig{waterfall}) during which the defect existed in the system, i.e.
+the ordinal rank of removal phase minus the ordinal rank of injection phase.
+Sub-phases (such as review and inspection) are not included in this calculation.
 
-\fig{phase_delay_medians} and \fig{correlation} show that most defects have a relatively small phase delay in our project sample (phase delay is often <= 2).
-As intended, the Review and Inspection Phases for Requirements, Design, and Code mostly find defects originating in the preceding phase.
+\fig{correlation} and \fig{phase_delay_medians} show that most defects have a relatively small phase delay in our project sample (phase delay is often <= 2).
+As intended, the Review and Inspection Phases for Requirements, Design, and Code mostly find defects originating in the same phase.
 In contrast, the Testing phases mostly find defects originating in the Code phase.
@@
 In \fig{raw}, we present the median fix times of defects based on when they were injected and ultimately removed.
-Defects injected Before Development follow the most severe scale up as development phases pass, but not on the exponential scale traditionally seen (\fig{cost-to-fix}).
+Defects injected Before Development (labelled "B4Dev" in \fig{waterfall}) follow the most severe scale up as development phases pass, but not on the exponential scale traditionally seen (\fig{cost-to-fix}).
 Defects injected in Requirements, Design, and Coding phases exhibit virtually no scale-up regardless of the phase in which they are fixed.
+
+We reject $H_2$.
 
@@
 \\
-  29 &     1 &     3 &    12 & Before & CodeInspect  & 1.00  &  **  \\
-  26 &     1 &     4 &     6 & Before & CodeReview  & 1.33  &  ***  \\
-  73 &     5 &    13 &    59 & Before & UnitTest  & 4.33  &  *********  \\
-  36 &     7 &    24 &    54 & Before & IntTest  & 8.00  &  ****************  \\
-  34 &     8 &    28 &    56 & Before & SysTest  & 9.33  &  *******************  \\\hline
+  29 &     1 &     3 &    12 & Before Dev & CodeInspect  & 1.00  &  **  \\
+  26 &     1 &     4 &     6 & Before Dev & CodeReview  & 1.33  &  ***  \\
+  73 &     5 &    13 &    59 & Before Dev & UnitTest  & 4.33  &  *********  \\
+  36 &     7 &    24 &    54 & Before Dev & IntTest  & 8.00  &  ****************  \\
+  34 &     8 &    28 &    56 & Before Dev & SysTest  & 9.33  &  *******************  \\\hline
 \\
@@
     \item reiterate hypothesis testing results
-    \item Restate the motivation: widely-held beliefs need to be revisted. 
+    \item Restate the motivation: widely-held beliefs need to be revisited. 
     \item We now have evidence, from 5 different sources, that phase delay does not seem to hold under certain conditions (although our study is by far the biggest in terms of data points).
@@
 \\
-We thank Tuma Solutions for providing the Team Process Data Warehouse software in support of this research.
+We thank Tuma Solutions for providing the Team Process Data Warehouse software in support of this research.  \\
+Team Software ProcessSM and TSPSM are service marks of Carnegie Mellon University.
 

### 2015-03-16 data.tex
Updates from ShareLaTeX

@@
-To empirically test this heuristic, we examined the phase delay effect in
-171 software projects conducted between 2005 and 2014 from around the world 
-that used the Team Software Process (TSP).
+Before presenting our analysis of phase delay in software projects, we define our terms, provide an overview of TSP and the sample projects, and describe the data collection process.
+
+\subsection{Definitions}
+A \emph{defect} is any change to a product, after its construction, that is necessary to make the product correct.
+A typographical error found in review is a defect.
+If that same defect is discovered while writing the code but before review, it is not considered to be a defect.
+
+The \emph{cost-to-fix} can be computed in two ways.
+The first is a direct measure of the actual work time spent in correcting the defect in the product.
+Direct measurement enables analysis of the variability the cost-to-fix among a category of defects.
+The second approach measures the total time spent in a defect removal phase and divides by the total defects removed.
+This provides an average cost-to-fix including overhead of that activity.
+For example, using method two, a code inspection may require an hour of effort to remove three defects for a 3 defects per hour or to invert the ratio, 1/3 hours per defect.
+Most of the time, however, is spent locating the defects not actually fixing them.
+we use the second approach in our analysis.
 
 \subsection{Overview of the Team Software Process}
-The Software Engineering Institute (SEI) at
-Carnegie Mellon University explores methods for software process improvement.
+To empirically test this heuristic, we examined the phase delay effect in software projects conducted between 2005 and 2014.
+These projects took place at organizations in many countries and were conducted using  the Team Software Process (TSP$\textsuperscript{SM}$ )), a software project management approach developed at the Software Engineering Institute (SEI) at Carnegie Mellon University.
+TSP is an extension of the Personal Software Process (PSP$\textsuperscript{SM}$ )) developed at the SEI by Watts Humphrey ~\cite{tsp00}.
+The data from these TSP projects were collected and stored in the Software Engineering Measured Process Repository (SEMPR) at the SEI.
+The Software Engineering Institute (SEI) at Carnegie Mellon University explores methods for software process improvement.
+
 One product from that research is  
-Watts Humphrey's  Team Software ProcessSM (TSPSM)~\cite{tsp00}. TSP is an extension of Humphrey's early
+Watts Humphrey's  Team Software Process$\textsuperscript{SM}$  (TSP$\textsuperscript{SM}$ )~\cite{tsp00} which extends Humphrey's early
 work on Personal Software
-ProcessSM (PSPSM)~\cite{psp05}.
+Process$\textsuperscript{SM}$  (PSP$\textsuperscript{SM}$ )~\cite{psp05}.
 
-TSP guiding principles include 1) sound engineering work does not happen by accident, it must be planned, 2) performance of the plans requires realistic commitments of the resources needed to perform the work, 3) the planned work should be compared to actual frequently to assure plans remain relevant, and 4) people performing the work should be responsible for planning and tracking the work.
+TSP has guiding principles; the following are relevant to this study:
+\begin{itemize}
+\item  Sound engineering work does not happen by accident. It must be planned.
+\item  Performance of the plans requires realistic commitments of the resources needed to perform the work.
+\item  The planned work should be compared to actual frequently to assure plans remain relevant.
+\item  People performing the work should be responsible for planning and tracking the work.
+\end{itemize}
+
 TSP helps developers by providing a set of measures that can be applied to managing tasks, quality, and schedule.
-Planning begins by quantifying goals, defining work practices, and estimating size and effort, developers make a detailed short term plan.
+Planning begins by quantifying goals, defining work practices, and estimating size and effort.
+Developers then use this information to make a detailed short term plan.
 As the developers perform project work, they use a tool such as the Process Dashboard to collect their time, effort, size, and schedule data.
-Every week, the team reviews their data to evaluate status, review actual rates, determine if project goals for schedule, cost, and quality are being met and then make necessary plan corrections.
-At the end of the project the coach and team perform a quantitative post mortem.
+Every week, the team reviews their data to evaluate status, identify actual rates, determine if project goals for schedule, cost, and quality are being met.
+The team then uses this information to make necessary plan corrections.
+At the end of the project the coach and team perform a quantitative project post mortem.
 
@@
 
-A TSP Coach helps the team to plan and analyze performance. The Coach is the only role authorized to submit project data to the SEI.
-Before reviewing data with the teams, therefore before submission,
-these coaches check the data for obvious errors.
+A TSP coach helps the team to plan and analyze performance. The coach is the only role authorized to submit project data to the SEI.
+Before reviewing data with the teams, therefore before submission, these coaches check the data for obvious errors.
 
-Planning follows the technique taught in the PSP(SM).
+Planning follows the technique taught in the PSP$\textsuperscript{SM}$.
 Developers estimate the size of work products and convert this to a total effort using historical rates.
@@
 
-Personal review is a technique taken  from the Personal Software
-ProcessSM (PSPSM) and is more or less unique in its TSP implementation.
+Personal review is a technique taken from the PSP and its use in TSP is unique.
 Developers follow a systematic process to remove defects by  examining their own work products using a checklist built from their personal defect profile.
-This personal review occurs after some product or part of a product is considered to constructed and before peer reviews or test.
+This personal review occurs after some product or part of a product is considered to be constructed and before peer reviews or test.
 
@@
 traditional software engineering and is often called peer review.
- Basili and Boehm write  commented in 2001~\cite{boehm01} 
+ Basili and Boehm   commented in 2001~\cite{boehm01} 
 that peer reviews can catch over half the defects introduced into a system.
@@
 One of the authors (Nichols) has been been mentoring software development teams and coaches around the world as they deploy TSP within their organizations since 2006.
-The earliest projects in our sample began work in 2004, the most recent completions were in 2014.
-The projects
-were mostly small to medium, with a median duration of 46 days
-and a maximum duration of 90 days in major increments. Several projects extended for many incremental cycles. Median team size was 7
-people, with a maximum of 40. 
-;Hence,  total development 
-;effort\footnote{
-;Effort =  duration*teamSize/workDaysPerYear.} for
-;these projects ranged from 1.3 years (median) to 14.3 years (max).
-Many of
-the projects were e-commerce web portals or banking systems in the US, South Africa, and Mexico. 
-There were also some  medical devices (US, France, Japan, and Germany),  a few from a commercial 
-computer-aided design system (developed by a world wide distributed team and embedded systems.
+The earliest projects in our sample began work in 2004; the most recent completions were in 2014.
+The projects were mostly small to medium, with a median duration of 46 days and a maximum duration of 90 days in major increments. 
+Several projects extended for multiple incremental development cycles. 
+Median team size was 7 people, with a maximum of 40.
 
- An anonymized version of that data is available in the PROMISE repository\footnote{\carter{Need to add}} and,
+Many of the projects were e-commerce web portals or banking systems in the US, South Africa, and Mexico. 
+There were  some  medical device projects in  the US, France, Japan, and Germany as well  commercial 
+computer-aided design systems, and embedded systems.
+
+An anonymized version of that data is available in the PROMISE repository  and,
 for confidentiality restrictions, we cannot offer 
-further details
-on those project.
+further details on these projects.
+
+\tim{Needs either a URL or a reference.}
 
 \subsection{Data Collection Process}
+\label{sec:data-collection}
 
@@
 Data includes project and site  characteristic summaries, surveys of team members, launch presentations, launch outbriefs, and baseline plans, final data from the project, and the project post mortem report.
-In practice, this data requirement has only been enforced for the purposes of certifying and reauthorizing TSP coache who must submit data to maintain their authorization.
+In practice, this data requirement has only been enforced for the purposes of certifying and reauthorizing TSP coaches who must submit data to maintain their authorization.
 Coaches are certified by demonstrating competent use of the TSP process with the artifacts and data not by the actual project results.
@@
 
-To use this data for research and benchmarking  Nichols first collected the Process Dashboard data from the submission repository then used a tool built by the Process Dashboard developer, David Tuma,
+To use this data for research and benchmarking,  Nichols first collected the Process Dashboard data from the submission repository then used a tool built by the Process Dashboard developer, David Tuma,
 to gather the project data into a database.
 Nichols and his colleague Yasutaka Shirai then extracted data into views suitable for analysis.
-The key views include the a project results summary, effort logs, task logs, and defect logs.
-At this stage, the data has not yet undergone additional screening.
-\sei{add cite to ICSSP paper} An analysis of the error rate in the data was reported in a previous work.
+The key views include a project results summary, effort logs, task logs, and defect logs.
+The data has not yet undergone additional screening.
+A summary of the data quality issues was reported in a previous work ~\cite{shirai14}.
 
 The tool collected process information about those tasks and defects that we analyze in this work. 
-That task data includes  work start time, work end time, delta
+That task data include  work start time, work end time, delta
 work time, and interruption time. Software engineers are often
@@
 
-The most common defect types used in the the SEI TSP data set are as follows:
-\be 
-\item Environment: design, compile, test, or other support system problems;
-\item Interface: procedure calls and reference, I/O, user format;
-\item Data: structure, content; 
-\item Documentation: comments, messages;
+The defect types in the most common defect type standard  in the the SEI TSP data are follows:
+\begin{itemize}
+\item Environment: design, compile, test, or other support system problems
+\item Interface: procedure calls and reference, I/O, user format
+\item Data: structure, content
+\item Documentation: comments, messages
 \item Syntax: spelling, punctuation typos, instruction formats
 \item Function: logic, pointers, loops, recursion, computation, function defects  
-\item Checking: error messages, inadequate checks;
-\item Build: change management, library, version control;
+\item Checking: error messages, inadequate checks
+\item Build: change management, library, version control
 \item Assignment: package
-declaration, duplicate names, scope, limits;
-\item System: configuration, timing, memory.
-\ee
+declaration, duplicate names, scope, limits
+\item System: configuration, timing, memory
+\end{itemize}
 
@@
 and 26,534 size logs. 
+After selecting defects from the data log and joining the data to the time log table  171 of these projects remained. The excluded projects had no or too few defects to use in this analysis.
 
 In this paper, when we report ``time to resolve an
-issue'', we show the difference between the start and end times
+issue,'' we show the difference between the start and end times
 of a work session, with any interruption time subtracted (the
@@
 The logical phases used in this paper are shown in \fig{waterfall}.
-Although the representation suggests a waterfall, the phases represent the primary stages through which requirements are reified into working and tested code.
-That is, all requirements must be stated in some way, implemented in code, integrated and tested.
+Although the representation suggests a waterfall, the phases represent the primary stages through which requirements are developed into working and tested code.
+That is, all requirements must be stated in some way, implemented in code, integrated, and tested.
 All real implementations of any size follow a spiral approach with many team performing the work in iterative and/or incremental development cycles.
-Within an project increment, multiple features or components may be developed and incremented.
-TSPSM is compatible with agile and encourages iterative and incremental development.
+Within a project increment, multiple features or components may be developed and incremented.
+TSP is compatible with agile and encourages iterative and incremental development.
 Nonetheless, the specific strategy and cycle duration is a project decision.
-TSPSM does, however, strongly encourage 1) constructing units of sufficient size that measurement is practicable, and 2) separating the construction from appraisal and test activities.
+TSP does, however, strongly encourage 1) constructing units of sufficient size that measurement is practicable, and 2) separating the construction from appraisal and test activities.
 This effectively highlights the separation of construction from rework activities and aids the apportionment of defects to those found in appraisal activities (reviews and inspections) and those found through failure (test).
@@
 
-Note that, in that figure:
+Note that, in that Figure~\ref{fig:waterfall}:
 \bi 
@@
 Several  phases in which product is created have sub-phases of {\em review} and {\em inspect} to remove defects.
-TSP uses review when individuals perform personal reviews of their work products prior to the peer review which TSP calls the inspection).
+TSP uses review, a sub-phase in which individuals perform personal reviews of their work products prior to the peer review, which TSP calls the inspection.
 \item Testing is divided into several stages.
 Developers perform unit test prior to code complete.
-Some standard test phases after code complete include the integration which combines program units into workable system ready for system test.
-Integration and system test are often performed by a separate group.
+After code complete a standard phase is the integration, which combines program units into workable system ready for system test.
+Integration,  system test, and acceptance test are often performed by a separate group.
 \ei
@@
 \bi 
-\item defects should be found before test, that is, test defects should be infrequent
-\item a product should be inspected before used in a subsequent construction phase to minimize  unnecessary rework
-\item a personal review should remove the most common defects before being given to peers for a detailed inspection
-\item because different categories of test will find different types of defect, and follow in a natural sequence they should be measured separately
+\item Defects should be found before test, that is, test defects should be infrequent.
+\item A product should be inspected before used in a subsequent construction phase to minimize  unnecessary rework.
+\item To minimize unnecessary rework, a product should be reviewed and inspected prior to use in a subsequent construction phase.
+\item A personal review should remove the most common defects before being given to peers for the inspection.
+\item Because different categories of test will find different types of defect, and follow in a natural sequence, the tests should be measured separately
 \ei
 
-In summary, the TSPSM process and measurement have been specifically designed to aid the analysis of defects for use in process improvement.
-The data is, therefore, uniquely suitable for an analysis of the phase delay effect.
+In summary, the TSP process and measurement have been specifically designed to aid the analysis of defects for use in process improvement.
+The data are, therefore, uniquely suitable for an analysis of the phase delay effect.
+\subsection{Distinguishing Project Characteristics}
+Modern development processes include the core engineering practices of estimation, planning, configuration control, requirements, design, code development, and test.
+Agile methods address  not only these core practices but also address team organization, individual and team commitments, test and build automation, and implementing iterative and incremental development.
+TSP includes using additional practices associated with high performance ~\cite{jones10} including; project tracking and control, specialization of team members, inspections, static analysis, documentation and training, and reuse.
+Although the teams in this sample use many typical agile practices, they are also using additional practices.

### 2015-03-16 lucas.tex
Updates from ShareLaTeX

@@
 We conducted a survey of software engineering practitioners and researchers on their views of commonly-held software engineering ``laws''.
-One of the laws questioned relates directly to phase delay: ``Requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' (from Glass~\cite{glass02} who cites Boehm \lucas{need to find which exact reference he uses}).
+One of the laws questioned relates directly to phase delay: ``Requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' (from Glass~\cite{glass02} p.71 who references Boehm \& Basili~\cite{boehm01}).
 We abbreviate this law as RqtsErr.
@@
 Participants were presented with the RqtsErr law and others drawn from \cite{glass02} and \cite{endres03}.
-The PC member survey contained an additional question: ``Do you agree with the following?
-'In general, the longer errors are in the system (requirements errors, design errors, coding errors, etc.), the more expensive they are to fix''' (PhaseDelay for short).
-Responses were recorded using the \textbf{Agreement} question Likert scale.
+The PC member survey contained an additional question on ``In general, the longer errors are in the system (requirements errors, design errors, coding errors, etc.), the more expensive they are to fix'' (PhaseDelay for short).
+Responses were recorded using the Agreement question Likert scale.
 
 Summary statistics for the agreement and applicability scores for the RqtsErr and PhaseDelay laws are presented in Figure~\ref{fig:survey_results}.
-Responses whose \textbf{applicability} response was ''I don't know'' are ommitted from analysis.
+Responses whose Applicability response was ''I don't know'' are ommitted from analysis.
 
-\begin{figure}[!h]
+\begin{figure}[!ht]
 \renewcommand{\baselinestretch}{0.7}

### 2015-03-16 lucasRelatedWork.tex
Added Survey section.
Outline added for Forrest
dropped correaltion scatter plots. replaced pngs with pdfs
fixed up various charts and tables. added median phase delay per phase
upgraded pickture of Royce case study
Updates from ShareLaTeX

@@
 
-\begin{figure}
+\begin{figure}[!ht]
  \includegraphics[width=3.3in]{boehm_cost-to-fix.png}
- \caption{Cost-to-fix curve. From~\cite{Boehm81}.}\label{fig:cost-to-fix}
+ \caption{Historical cost-to-fix curve. From~\cite{Boehm81}.}\label{fig:cost-to-fix}
  \end{figure}
@@
 In this project, design changes (including architecture changes) required approximately twice the effort of implementation and test changes, and the cost-to-fix in implementation and test phases increased slowly.
-Boehm~\cite{Boehm10} attributes this success to the CCPDS-R development process, which focused on removing architecture risk early in the devleopment lifecycle.
+Boehm~\cite{Boehm10} attributes this success to the CCPDS-R development process, which focused on removing architecture risk early in the development lifecycle.
 
-\begin{figure}
- \includegraphics[width=3.3in]{Royce98.PNG}
- \caption{Exception to the rule - CCPDS-R cost-to-fix curve. From~\cite{Royce98}.}\label{fig:royce}
+\begin{figure}[!ht]
+ \includegraphics[width=3.3in]{Royce98.png}
+ \caption{Exception to the rule - CCPDS-R case study cost-to-fix curve. From~\cite{Royce98}.}\label{fig:royce}
  \end{figure}
 
-As discussed earlier, one goal of agile methods is to flatten the cost-to-change curve~\cite{beck00}.
+As discussed is \S\ref{sec:why-study}, one goal of agile methods is to flatten the cost-to-change curve~\cite{beck00}.
 Relatively little empirical data exists on this point.
@@
 
-\begin{figure}
+\begin{figure}[!ht]
  \includegraphics[width=3.3in]{clutterbuck.png}
- \caption{Cost of change from an agile case study~\cite{Clutterbuck09}.}\label{fig:clutterbuck}
+ \caption{Cost of change from an agile case study. From~\cite{Clutterbuck09}.}\label{fig:clutterbuck}
  \end{figure}
@@
  We note that previous work focuses on cost-to-fix as a function of lifecycle phase irrespective of when the defect was injected, that is, previous work analyzes the cost to fix a defect found in test regardless of whether that defect was a requirements error or a coding mistake.
-To our knowledge, our research represents the first large-scale study of phase delay as a contributor to the cost-to-fix curve.
+To our knowledge, our research represents the first large-scale study of phase delay.

### 2015-03-16 phaseDelay.tex
Updates from ShareLaTeX

@@
 \renewcommand{\algorithmicensure}{\textbf{Output:}} 
-\newcommand{\lucas}[1]{\textcolor{red}{LUCAS: #1}} \newcommand{\tim}[1]{\textcolor{orange}{TIM: #1}}
+\newcommand{\lucas}[1]{\textcolor{red}{LUCAS: #1}} 
+\newcommand{\tim}[1]{\textcolor{Red}{TIM: #1}}
 \newcommand{\bill}[1]{\textcolor{blue}{BILL: #1}} 
@@
  \setlength{\belowcaptionskip}{1ex}
-
  \setlength{\floatsep}{1ex}
@@
 \conferenceinfo{FSE}{'15 Bergamo, Italy}
-\title{Revisiting the Truisms of Software Engineering:\\ Does Phase Delay Dramatically Increase  Repair Time?}
-\numberofauthors{3}
+\title{Revisiting the Truisms of Software Engineering:\\ Does Phase Delay Dramatically Increase Defect Fix Time?}
+\numberofauthors{5}
 \author{
 \alignauthor 
-Tim Menzies, \\Carter Pape\\
-       \affaddr{Dept. of Computer Science\\
-       North Carolina State U, USA}\\
-       tim.menzies@gmail.com,\\carterpape@gmail.com
+Tim Menzies \\
+       \affaddr{Dept. of Computer Science}\\
+       \affaddr{North Carolina State U, USA}\\
+       \email{tim.menzies@gmail.com}
 \alignauthor 
-William Nichols,\\ Forrest Shull\\
-\affaddr{Software Engineering Institute\\
-        Carnegie Mellon U, USA}
-wrn,fjshull@sei.cmu.edu
+William Nichols \\
+        \affaddr{Software Engineering Institute}\\
+        \affaddr{Carnegie Mellon U, USA}
+        \email{wrn@sei.cmu.edu} 
 \alignauthor 
-Lucas \\Layman\\
-       \affaddr{Fraunhofer CESE, \\
-                College Park, USA}\\ 
-       llayman@fc-md.umd.edu
+Lucas Layman \\
+       \affaddr{Fraunhofer CESE} \\
+       \affaddr{College Park, USA}\\ 
+       \email{llayman@fc-md.umd.edu}
+\and
+\alignauthor 
+Forrest Shull \\
+        \affaddr{Software Engineering Institute}\\
+        \affaddr{Carnegie Mellon U, USA}
+        \email{fjshull@sei.cmu.edu} 
+\alignauthor 
+Carter Pape \\
+       \affaddr{Dept. of Computer Science}\\
+       \affaddr{North Carolina State U, USA}\\
+       \email{carterpape@gmail.com}
 } 
-
 \maketitle
@@
 quite strongly in the industrial software engineering
-community (but, in the academic community, somewhat less so).
+community (but, in the academic community, a little less so).
 Yet based on a sample of 
@@
 and completed between 2006 and 2014, we argue that this belief is mostly 
-incorrect. Specifically: the median fix time for issues
-is just a few minutes and this does not tend to increase
+incorrect. Specifically: the median fix time spent in phases
+devoted to resolving issues does {\em not} increase
 the longer the issue remains in the system.
 
-%This result raises the question: how did myths like phase delay survive so long  without  critical reassessment.
+This result raises the question: how do widely-held beliefs like phase delay survive so long  without  critical reassessment?
 As a community, we need to reflect more on research practices and on how (and when) we reflect on the prominent hypotheses of our field.
@@
 \noindent
-{\bf Keywords:} phase delay, time to fix, agile, waterfall.
+{\bf Keywords:} software processes, economics, phase delay, time to fix.
 
@@
 \end{quote}
-Basili and Boehm are not alone in this assessment of the significance of the phase delay effect. Our surveys of industrial practitioners (shown in \S\ref{survey}) reveal the phase-delay effect to be
-the single most strongly held belief amongst commercial SE engineers.
-Previously~\cite{me08a}, we have constructed return-on-investement models
+Basili and Boehm are not alone in this assessment of the significance of the phase delay effect. Our surveys of industrial practitioners (shown in \S\ref{survey}) reveal the phase delay effect to be
+one of the most strongly-held beliefs amongst commercial software engineers.
+Previously~\cite{me08a}, we have constructed return-on-investment models
 for a common practice amongst U.S. government contractors: 
@@
 
-In this paper, we investigate two hypotheses:
+In this paper, we investigate several hypotheses:
 
@@
 \ei
-The rest of the paper investigates these two hypotheses.
+As we explored the above, we found a related hypothesis was believed by both professional software engineers and FSE conference reviewers:
+\bi
+ \item $H_3$: Requirements errors are the most expensive to fix.
+\ei
+
+The rest of the paper investigates these  hypotheses.
 First, we conduct a survey of practitioners and researchers to gauge their belief in the phase delay effect.
@@
 \subsection{Why Study Phase Delay?}
-\todo{\begin{itemize}
-\item Given the Intro and Motivation, this section should focus on the influence of phase delay in software engineering milieux
-\item Need to tone down the reliance on Beck.
+\label{sec:why-study}
+%\todo{\begin{itemize}
+%\item Given the Intro and Motivation, this section should focus on the influence of phase delay in software engineering milieux
+%\item Need to tone down the reliance on Beck.
 The important Beck point is that eliminating the cost-to-fix curve (resulting from phase delay) is one of the main goals of agile.
 His observations on new technologies are good too.
-\item Need to cite other popular sources of this information, e.g., McConnell's "Ounce of Prevention" article~\cite{mcconnell01}.
-\item Robert Glass's~\cite{glass02} book is very popular and contains an entry on this point
-\item Google Scholar Citations for phase delay works Boehm88~\cite{boehm88} (673), Boehm \& Basili~\cite{boehm01} (761), 
-\end{itemize}}
+%\item Need to cite other popular sources of this information, e.g., McConnell's "Ounce of Prevention" article~\cite{mcconnell01}.
+%\item Robert Glass's~\cite{glass02} book is very popular and contains an entry on this point
+%\item Google Scholar Citations for phase delay works Boehm88~\cite{boehm88} (673), Boehm \& Basili~\cite{boehm01} (761), 
+%\end{itemize}}
 
-Due to its prominence in the field of software engineering,
-this paper studies phase delay. As mentioned in the introduction, concerns about phase delay have
-been used for decades to motivate major changes to software engineering practices. For example,
+This paper examines evidence for the phase delay effect due to its prominence throughout the field of software engineering.
+Many prominent authors have commented on its perceived usefulness as a rule of thumb for software engineers.
+For example, in 2001, Steve McConnell mentioned it as a "common observation" in the field and proceeded to knock down arguments that it no longer held for contemporaneous software projects, by relying on recently published data and experience.
+McConnell also summarized the intuitive argument for why it should be so: ``A small mistake in upstream work can affect large amounts of downstream work.
+A change to a single sentence in a requirements specification can imply changes in hundreds of lines of code spread across numerous classes or modules, dozens of test cases, and numerous pages of end-user documentation''~\cite{mcconnell01} Glass echoes this sentiment, stating the the fact ``requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' is ``really just common sense''~\cite{glass02}.
+
+Also in 2001, the phase delay effect was listed as number one by Boehm and Basili in their "Top 10 list" of "objective and quantitative data, relationships,
+and predictive models that help
+software developers avoid predictable pitfalls
+and improve their ability to predict
+and control efficient software projects." \cite{boehm01}
+
+More important than its use as a rule of thumb, the phase delay effect has 
+been used for decades to motivate major changes to software engineering practices (as mentioned in Section 1). For example,
 in 2000, Kent Beck used phase delay to motivate his call for ``extreme programming'' (a set of techniques
@@
 One of the universal assumptions of software engineering is that the cost of changing a program rises exponentially over time.
-I can remember sitting in a big linoleum-floored classroom as a college junior and seeing the professor draw on the board the curve found in \fig{curve1}.
-
-The cost to fix a problem in a piece of software rises exponentially over time.
+I can remember sitting in a big linoleum-floored classroom as a college junior and seeing the professor draw on the board the curve found in \fig{curve1}...
+%
+%The cost to fix a problem in a piece of software rises exponentially over time. 
 A problem that might take a dollar to fix if you found it during requirements analysis might costs thousands to fix once the software is in production.
-
-I resolved then and there that I would never let a problem get through to production.
+%
+%I resolved then and there that I would never let a problem get through to production.
 No sirree, I was going to catch problems as soon as possible.
@@
 I would review and crosscheck my code.
-No way was I going to cost my employer \$100,000.~\cite{beck00}
+No way was I going to cost my employer \$100,000.~
+\cite{beck00}
 \end{quote}
@@
 
-\begin{figure}[!t]
-\scriptsize
-\begin{center}
-\begin{tabular}{l@{~}|l@{~}|r@{~}r@{~}|r@{~}l}
-          &        & \multicolumn{2}{c|}{Time to}  &\\
-          &        & \multicolumn{2}{c|}{resolve} &\\
-Injection & Removal& initial & now & \multicolumn{2}{l}{Scale up w.r.t initial} \\\hline
-Phase $i$&Phase $i$ & 1 & 1 & 1 & **\\
-Phase $i$ &Phase $i+1$ & 1 & 2& 2 & **** \\
-Phase $i$ &Phase $i+2$ & 1 & 4& 4 &******** \\
-Phase $i$ &Phase $i+3$ & 1 & 8 & 8 & ****************  
-  \end{tabular}\end{center}
- \caption{Projects data consistent with
- the phase delay effect of \fig{curve1}.}\label{fig:curve1a}
-\end{figure}
+% \begin{figure}[!t]
+% \scriptsize
+% \begin{center}
+% \begin{tabular}{l@{~}|l@{~}|r@{~}r@{~}|r@{~}l}
+%           &        & \multicolumn{2}{c|}{Time to}  &\\
+%           &        & \multicolumn{2}{c|}{resolve} &\\
+% Injection & Removal& initial & now & \multicolumn{2}{l}{Scale up w.r.t initial} \\\hline
+% Phase $i$&Phase $i$ & 1 & 1 & 1 & **\\
+% Phase $i$ &Phase $i+1$ & 1 & 2& 2 & **** \\
+% Phase $i$ &Phase $i+2$ & 1 & 4& 4 &******** \\
+% Phase $i$ &Phase $i+3$ & 1 & 8 & 8 & ****************  
+%   \end{tabular}\end{center}
+%  \caption{Projects data consistent with
+%  the phase delay effect of \fig{curve1}.}\label{fig:curve1a}
+% \end{figure}
 
-\begin{figure}[!t]
-\scriptsize
-\begin{center}
-\begin{tabular}{l@{~}|l@{~}|r@{~}r@{~}|r@{~}l}
-          &        & \multicolumn{2}{c|}{Time to}  &\\
-          &        & \multicolumn{2}{c|}{resolve} &\\
-Injection & Removal& initial & now & \multicolumn{2}{l}{Scale up w.r.t initial} \\\hline
-Phase $i$&Phase $i$ & 1 & 1 & 1 & ** \\
-Phase $i$ &Phase $i+1$ & 1 & 2& 2 & **** \\
-Phase $i$ &Phase $i+2$ & 1 & 2& 2 & **** \\
-Phase $i$ &Phase $i+3$ & 1 & 2 & 2 & **** \\
- Phase $i$  &Phase $i+4$ &1 & 2& 2 &**** 
-  \end{tabular}\end{center}
- \caption{Projects data consistent with
- the phase delay effect of \fig{curve2}.}\label{fig:curve2a}
-\end{figure}
+% \begin{figure}[!t]
+% \scriptsize
+% \begin{center}
+% \begin{tabular}{l@{~}|l@{~}|r@{~}r@{~}|r@{~}l}
+%           &        & \multicolumn{2}{c|}{Time to}  &\\
+%           &        & \multicolumn{2}{c|}{resolve} &\\
+% Injection & Removal& initial & now & \multicolumn{2}{l}{Scale up w.r.t initial} \\\hline
+% Phase $i$&Phase $i$ & 1 & 1 & 1 & ** \\
+% Phase $i$ &Phase $i+1$ & 1 & 2& 2 & **** \\
+% Phase $i$ &Phase $i+2$ & 1 & 2& 2 & **** \\
+% Phase $i$ &Phase $i+3$ & 1 & 2 & 2 & **** \\
+%  Phase $i$  &Phase $i+4$ &1 & 2& 2 &**** 
+%   \end{tabular}\end{center}
+%  \caption{Projects data consistent with
+%  the phase delay effect of \fig{curve2}.}\label{fig:curve2a}
+% \end{figure}
 
 Beck speculated that there might be another kind of project, shown in \fig{curve2},
-that does not exhibit phase delay. In this next quote, Beck clearly states
-the importance of avoiding phase delay (in fact, for Beck, it is {\em the} major
-motivation for his work):
-\begin{quote}
-This (\fig{curve2}) is one of the premises of (agile).
+that does not exhibit phase delay. In fact, for Beck, the possibility that phase delay does not imply exponential increases in the resources required to make changes and fix defects is {\em the} major
+motivation for his work, because it means that decisions can be made as late in the process as possible, so as to increase the chances that these decisions would be right.
+Beck found more than adequate motivation for this new assumption:
+%\begin{quote}
+%This (\fig{curve2}) is one of the premises of (agile).
 It is the technical premise of (agile).
@@
 You would introduce elements to the design only as they simplified existing code or made writing the next bit of code simpler.~\cite{beck00}
-\end{quote}
-Beck makes one other comment that is relevant to this paper: 
+%\end{quote}
+%Beck makes one other comment that is relevant to this paper: 
 \begin{quote}
@@
 
-\todo{
-\bi
-    \item What is the cost-to-fix curve in TSP data? Check against H1 and compare to past literature.
-    \item What is the effect of phase delay? Check against H2.
-    \item \fig{cost-to-fix-tsp} is great.
+%\todo{
+%\bi
+%    \item What is the cost-to-fix curve in TSP data? Check against H1 and compare to past literature.
+%    \item What is the effect of phase delay? Check against H2.
+%    \item \fig{cost-to-fix-tsp} is great.
 really important.
@@
 not, perhaps, due to process but to better build environs.
-\ei
-}
+%\ei
+%}
 
@@
 A high percentage of defects were found and fixed in the early phases, i.e., requirements, high level design, and design reviews and inspections.
-\todo{How does this compare evidence from other sources?}
+(This distribution is similar to that observed for other projects that emphasized investment in software engineering quality assurance practices, for example the NASA robotics projects which had 58\% of defects removed before implementation began.
+\cite{me08a} Similar results are presented by Jones and Bonsignour, who show 52\% of pretest defects removed before entering implementation.
+\cite{jones12})
 
-The median time to fix a defect in each phase is shown in \fig{fix-time-per-phase}.
+\todo{This paragraph is wrong now.} The median time to fix a defect in each phase is shown in \fig{fix-time-per-phase}.
 The median cost to fix a defect is low in the early phases, and then increases slightly during design and coding before peaking during the unit test phase.
@@
 
-\begin{figure}[!ht]
+\begin{figure*}[!ht]
+
+ \renewcommand{\baselinestretch}{0.7}
+ \scriptsize
 \begin{center}
-\begin{tabular}{r|rrr|l}
-  Sample&\multicolumn{3}{c|}{Percentiles}\\ 
-size & 25th & 50th & 75th & Phase removed \\
-\hline
-120 & 3 & 3 & 18 & ReqtsInspect \\
-126 & 3 & 3 & 20 & ReqtsReview \\
-70 & 2 & 2 & 10 & HLDInspect \\
- 55 &    2 &    1&    7&HLDReview \\
-
-278&    7&    7&   33&DesignInspect \\
-233&    6&    6&   25&DesignReview \\
-
-315&    8&    7&   34&CodeInspect \\ 
-282&    7&    7&   32&CodeReview\\
-
-425&   11&   10&   54&UnitTest \\ 
-226&    7&    6&   46&IntTest \\ 
-194&    7&    6&   52&SysTest\\ 
- 88&    4&    3&   38&AcceptTest\\
+\begin{tabular}{r|rr|rl}
+Phase & samples & total time&\multicolumn{2}{l}{Total time w.r.t. first phase}\\\hline
+~\\
+Planning&229&12625&1.00&*\\
+~\\
+Reqts&590&24182&1.91&**\\
+ReqtsInspect&2533&92723&7.34&********\\
+ReqtsReview&1762&67471&5.34&******\\
+~\\
+HLD&78&6856&0.54&*\\
+HLDReview&472&12821&1.01&**\\
+HLDInspect&1366&35291&2.79&***\\
+~\\
+Design&620&21573&1.70&**\\
+DesignInspect&8373&229937&18.21&*******************\\
+DesignReview&3043&92823&7.35&********\\
+~\\
+Code&3029&220058&17.43&******************\\
+CodeReview&4774&163091&12.91&*************\\
+CodeInspect&11406&334442&26.49&***************************\\
+~\\
+UnitTest&9177&352245&2.790&****************************\\
+IntTest&2761&160559&12.71&*************\\
+SysTest&2059&105796&8.37&*********\\
+AcceptTest&1045&42717&0.338&****\\
 \end{tabular}
 \end{center}
-\caption{Median fix time per phase}
+\caption{Times per phase (minutes).}
 \label{fig:fix-time-per-phase}
-\end{figure}
+\end{figure*}
 
 For illustrative purposes, the relative cost to fix errors per lifecycle phase in the TSP project data is overlayed with Boehm's cost curve~\cite{Boehm81} as well as the CCPDS-R case study~\cite{Royce98} in \fig{cost-to-fix-tsp}.
+Cost-to-fix for TSP is total time spent in each review, inspection, or test phase (reported in \fig{fix-time-per-phase}) / \# of defects found in phase (reported in \fig{raw}).
 It is clear to see that the trend is relatively flat, closer to the pattern exhibited by the Royce case study (which also focused on early defect removal) than the exponential growth curve.
@@
 
-We reject $H_1$.
+Given the above evidence, we reject $H_1$: The cost-to-fix of defects monotonically increases across phases.
 
 \subsection{Phase delay analysis}
-\todo{
-\bi
-    \item What can we say about \fig{raw}
-\ei
-}
 
@@
 the ordinal rank of removal phase minus the ordinal rank of injection phase.
-Sub-phases (such as review and inspection) are not included in this calculation.
-
-\fig{correlation} and \fig{phase_delay_medians} show that most defects have a relatively small phase delay in our project sample (phase delay is often <= 2).
-As intended, the Review and Inspection Phases for Requirements, Design, and Code mostly find defects originating in the same phase.
-In contrast, the Testing phases mostly find defects originating in the Code phase.
+Sub-phases (such as review and inspection) are included in this calculation.
 
@@
 
-We reject $H_2$.
-
-%In the TSP data, there is a weak correlation (Kendall's $\tau = 0.19, p < 0.01, 95\% CI$) between a defect's cost-to-fix (minutes of effort) and the defects phase delay (phase removed - phase injected).
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=3.3in]{corr.pdf}
-\end{center}
-\caption{Correlation of phase delay and fix time (plot shows a 1\% sampling of data).}
-\label{fig:correlation}
-\end{figure}
-
-\begin{figure}[!ht]
-\centering
-\begin{tabular}{rrr|l}
-    \multicolumn{3}{c|}{Percentiles}\\ 
-  25th & 50th & 75th & Phase removed \\ 
-  \hline
-    2 & 2 & 2 & ReqtsInspect \\ 
-    1 & 1 & 1 & ReqtsReview \\ 
-    2 & 2 & 2 & HLDInspect \\ 
-    1 & 1 & 1 & HLDReview \\ 
-    2 & 2 & 2 & DesignInspect \\ 
-    1 & 1 & 1 & DesignReview \\ 
-    2 & 2 & 2 & CodeInspect \\ 
-    1 & 1 & 1 & CodeReview \\ 
-    3 & 3 & 6 & UnitTest \\ 
-    4 & 4 & 4 & IntTest \\ 
-    5 & 5 & 5 & SysTest \\ 
-    6 & 6 & 6 & AcceptTest \\ 
-  \end{tabular}
-\caption{Median phase delay of defects removed per phase} 
-\label{fig:phase_delay_medians}
-\end{figure}
-
-% \begin{figure}[ht]
-% \centering
-% \begin{tabular}{lll}
-%   Phase & $tau$ & $p$ \\ 
-%   \hline
-%   Design & 0.14 & 0.08 \\ 
-%   DesignReview & 0.06 & $<$0.01 \\ 
-%   DesignInspect & 0.05 & $<$0.01 \\ 
-%   Code & 0.27 & $<$0.01 \\ 
-%   CodeReview & 0.12 & $<$0.01 \\ 
-%   CodeInspect & 0.09 & $<$0.01 \\ 
-%   Test & 0.13 & $<$0.01 \\ 
-%   IntTest & 0.1 & $<$0.01 \\ 
-%   SysTest & 0.25 & $<$0.01 \\ 
-%   AcceptTest & 0.06 & 0.29 \\ 
-%   \end{tabular}
-% \caption{Correlation of cost-to-fix and phase delay for defects found (Kendall's $\tau$)} 
-% \label{fig:phase_corr}
-% \end{figure}
-
-% \begin{figure*}[!ht]
-% \centering
-% \includegraphics[width=1.6in]{corr-Code.png}
-% \includegraphics[width=1.6in]{corr-CodeReview.png}
-% \includegraphics[width=1.6in]{corr-CodeInspect.png}
-% \vspace{0.01cm}
-
-% \includegraphics[width=1.6in]{corr-Design.png}
-% \includegraphics[width=1.6in]{corr-DesignReview.png}
-% \includegraphics[width=1.6in]{corr-DesignInspect.png}
-% \vspace{0.01cm}
-
-% \includegraphics[width=1.6in]{corr-Test.png}
-% \includegraphics[width=1.6in]{corr-IntTest.png}
-% \includegraphics[width=1.6in]{corr-SysTest.png}
-% \includegraphics[width=1.6in]{corr-AcceptTest.png}
-
-% \caption{Correlation of cost-to-fix (rank) and phase delay for defects found by phase for phases where \# defects >= 30 and \# unique phase delay values >= 5.
-x = phase delay, y = cost-to-fix rank.
-Plots use random samples from each phase for readability.}
-% \label{fig:phase_fixtime_plot}
-% \end{figure*}
+Given the above evidence, we reject $H_2$: the cost-to-fix of defects monotonically increases as the phase delay increases.
 
@@
 \\
-  29 &     1 &     3 &    12 & Before Dev & CodeInspect  & 1.00  &  **  \\
-  26 &     1 &     4 &     6 & Before Dev & CodeReview  & 1.33  &  ***  \\
-  73 &     5 &    13 &    59 & Before Dev & UnitTest  & 4.33  &  *********  \\
-  36 &     7 &    24 &    54 & Before Dev & IntTest  & 8.00  &  ****************  \\
-  34 &     8 &    28 &    56 & Before Dev & SysTest  & 9.33  &  *******************  \\\hline
+ 48&   14&   30&   51&Before&DesignInspect &1.00 & **********  \\
+ 94&   22&   45&   80&Before&CodeInspect &1.50 & ***************  \\
+ 53&   13&   28&   49&Before&CodeReview &0.93 & **********  \\
+153&   38&   77&  130&Before&UnitTest &2.56 & **************************  \\
+127&   30&   65&  128&Before&IntTest &2.16 & **********************  \\
+ 92&   24&   56&   89&Before&SysTest &1.86 & *******************  \\
+
 \\
-128 &     2 &     6 &    15 & Reqts & ReqtsInspect  & 1.00  &  **  \\
-132 &     2 &     5 &    12 & Reqts & ReqtsReview  & 0.83  &  **  \\
- 28 &     3 &    13 &    29 & Reqts & DesignInspect  & 2.16  &  *****  \\
- 22 &     6 &    14 &    31 & Reqts & DesignReview  & 2.33  &  *****  \\\hline
+ 56&   14&   35&   57&Planning&ReqtsReview &1.00 & **********  \\
+ 46&   11&   24&   39&Planning&DesignInspect &0.68 & *******  \\
+ 32&   18&   33&   84&Planning&UnitTest &0.94 & **********  \\
+
 \\
-272 &     6 &    14 &    30 & Design & DesignInspect  & 1.00  &  **  \\
- 32 &     1 &     4 &    13 & Design & DesignInspection  & 0.28  &  *  \\
-240 &     4 &    11 &    22 & Design & DesignReview  & 0.78  &  **  \\
-110 &     3 &    11 &    23 & Design & CodeInspect  & 0.78  &  **  \\
- 91 &     3 &     7 &    18 & Design & CodeReview  & 0.50  &  *  \\
-323 &     8 &    21 &    45 & Design & UnitTest  & 1.50  &  ***  \\
- 56 &     7 &    20 &    41 & Design & IntTest  & 1.42  &  ***  \\
- 69 &     9 &    27 &    53 & Design & SysTest  & 1.92  &  ****  \\\hline
+189&   46&   93&  155&Reqts&ReqtsInspect &1.00 & **********  \\
+184&   45&   92&  155&Reqts&ReqtsReview &0.98 & **********  \\
+102&   24&   49&   83&Reqts&DesignInspect &0.52 & ******  \\
+ 61&   14&   33&   55&Reqts&DesignReview &0.35 & ****  \\
+ 59&   18&   38&   60&Reqts&CodeInspect &0.40 & ****  \\
+ 91&   22&   49&  104&Reqts&UnitTest &0.52 & ******  \\
+100&   33&   67&  120&Reqts&IntTest &0.72 & ********  \\
+ 78&   28&  122&  237&Reqts&SysTest &1.31 & **************  \\
 \\
-310 &     6 &    15 &    31 & Code & CodeInspect  & 1.00  &  **  \\
-285 &     5 &    14 &    29 & Code & CodeReview  & 0.93  &  **  \\
-287 &     7 &    17 &    44 & Code & UnitTest  & 1.13  &  ***  \\
-180 &     7 &    20 &    45 & Code & IntTest  & 1.33  &  ***  \\
-137 &     6 &    15 &    35 & Code & SysTest  & 1.00  &  **  \\
- 37 &     1 &    11 &    27 & Code & AcceptTest  & 0.73  &  **  \\
+207&   50&  101&  156&Design&DesignInspect &1.00 & **********  \\
+210&   51&  103&  166&Design&DesignReview &1.01 & ***********  \\
+163&   39&   79&  135&Design&CodeInspect &0.78 & ********  \\
+134&   32&   66&  116&Design&CodeReview &0.65 & *******  \\
+238&   58&  119&  183&Design&UnitTest &1.17 & ************  \\
+172&   42&   88&  158&Design&IntTest &0.87 & *********  \\
+163&   40&   80&  161&Design&SysTest &0.79 & ********  \\
+ 80&   21&   45&   73&Design&AcceptTest &0.44 & *****  \\
 
+\\
+130&   31&   65&  122&Code&AcceptTest &1.00 & **********  \\
+228&   56&  113&  176&Code&CodeInspect &1.73 & ******************  \\
+235&   57&  115&  177&Code&CodeReview &1.76 & ******************  \\
+243&   59&  125&  206&Code&IntTest &1.92 & ********************  \\
+191&   46&   95&  173&Code&SysTest &1.46 & ***************  \\
+269&   66&  133&  206&Code&UnitTest &2.04 & *********************  \\
  \end{tabular}
@@
 \section{Validity} 
+We report validity concerns following the guidelines of Runeson \& H\"{o}st~\cite{runeson09}.
+
+\subsection{Construct validity}
+The essential constructs in our analyses are \emph{cost-to-fix, defects, and phase delay}.
+
+\emph{Cost-to-fix} is traditionally measured as either effort spent to fix a defect or time spent fixing a defect.
+In our analysis, we calculate the latter as \# of defects fixed in a defect-removal plan item, such as DesignReview or CodeInspection, divided by the time spent in the plan item.
+A single project has multiple such plan items.
+In the TSP data, this time accounts for time spent isolating, analyzing, and fixing the defect, but does not include time spent verifying the fix.
+
+In our sample, all \emph{defects} are considered equally, and thus TSP's reporting methods may result in ``small'' defects could be dominating the sample.
+For example, typographical requirements errors are included as well as requirements errors that necessitate an architectural change to the system.
+Our analysis does not include a notion of defect severity, however, we observe no indications in previous cost-to-fix results reported by \cite{Boehm81} and \cite{Royce98} that severity was considered in their samples.
+However, evidence discussed in \cite{Shull02} suggests that low severity defects may exhibit a lower cost to change.
+Nonetheless, even ``small'' typographical errors have been known to cause enormous damage (e.g., the Mars climate Orbiter).
+
+\emph{Phase delay}, as depicted in the scale up column in \fig{cost-to-fix-tsp} \fig{raw}, is a discrete value that does not consider the total time that a defect is latent in the system.
+Thus, the phase delay of two defects injected in Code but found in System Test will be the same even though one was found at the beginning of test while the other at the end of a potentially lengthy test period.
+Ideally, one could examine the total real time a defect was in a system from the point at which is was injected to the point at which it was discovered.
+It may be possible to gather such information if the project keeps rigorous change logs for its requirements, design, and code artifacts.
+
+\subsection{Internal validity}
+Confounding variables are a threat to any study.
+In this research, we rely on the substantial sample size of 171 TSP projects and over 47,000 defect logs to avoid confounding effects that may arise from the nature of the projects, teams, or defects.
+Similarly, the 10-year data collection period should help to ameliorate any maturation or history effects.
+As described in \S\ref{sec:data-collection}, all TSP teams are required to contribute time and defect data to the SEI, and thus there should be no selection bias in this sample compared to the overall population of TSP projects.
+However, there is likely selection bias in the teams that elect to use TSP compared to the entire population of software development teams.
+
+\subsection{External validity}
+Threats to external validity affect the generalization of findings.
+The most obvious threat to external validity is that TSP is not representative of most software development processes.
+TSP was, in fact, designed to be rigorous and eliminate defects early in the lifecycle and soon after injection.
+We make no assertions regarding the generalizability of these results beyond TSP, though we offer some conjecture in \S\ref{sec:discussion} on the potential causes for the flattened cost-to-fix curve.
+
+Our analysis of cost-to-fix does not include defects found in production, which, in Boehm's original data, account for costs up to 200 times the cost of finding and fixing defects in Requirements.
+Further, it is not clear how Boehm~\cite{Boehm81} and Royce~\cite{Royce98} define \emph{cost-to-fix}.
+In our analysis, cost-to-fix includes the time spent isolating, analyzing, and fixing a defect, but not the cost of verifying.
+If Boehm and Royce include cost of verifying a fix, then the TSP cost curve in \fig{cost-to-fix-tsp} artificially low by comparison.
+If Boehm and Royce include only the mechanics of fixing the defect, then the TSP cost curve in \fig{cost-to-fix-tsp} is artifically high by comparison.
+
+\subsection{Reliability}
 A previous paper~\cite{shirai14} has applied a range of sanity checks to the data.
@@
 affected by noise. Nichols et al.~\cite{shirai14}  report that
-the noise levesl in the SEI TSE data are smaller than those seen
+the noise levesl in the SEI TSP data are smaller than those seen
 in other data sets. They found in the SEI TSP data that:\bi 
@@
 
-XXX post release. really hard to change? auto-updates, package managers, web-based apps.
+The definitions of defect, cost-to-fix, and phase delay are provided in earlier sections.
+The source data and analysis scripts used to calculate the defect counts, cost-to-fix, and phase delay from the raw materials can be found at \url{https://github.com/txt/txt.github.io/tree/master/tsp}.
+\tim{Is this okay?}.
 
  \section{Discussion}
- \todo{
- \bi
-    \item summarize analysis findings
-    \item Offer interpretations of findings, e.g., why might these things be so
+ \label{sec:discussion}
+% \todo{
+% \bi
+%    \item summarize analysis findings
+%    \item Offer interpretations of findings, e.g., why might these things be so
+% \ei
+% }
+
+In the preceding sections, we looked for evidence of the phase delay effect (using two common interpretations of that effect) in comprehensive and fine-grained data covering 171 software development projects.
+The results did not support the existence of such an effect.
+We do not claim that this somehow disproves the phase delay effect, but we observe that it cannot be assumed to hold in every software development context.
+
+Since the phase delay effect has been such a long-lived software engineering truism, as well as one that is accepted by a substantial population of software engineers, it is important to explore possible explanations for our observations.
+
+We surmise that the substantial changes in software development and deployment technologies and practices, since the phase delay effect was first measured, may be responsible for our observations in the TSP data set.
+Recall that the original data on which the phase delay effect is based, as reported in Section 3, are approximately 40 years old.
+The most recently reported data, from Shull et al.
+\cite{Shull02}, was published in 2002 but contains substantial retrospective data.
+Over time, it would be surprising if the investment in better languages, IDEs, and other development technologies had in fact not created some improvement.
+To pick just one example, we note that much of the data showing the phase delay effect was collected before frequent and unobtrusive auto-updating software was as common as today, which certainly flattens the cost of changes.
+
+One hypothesized explanation that can be ruled out is that the lack of a phase delay effect is due to using the TSP method.
+To the contrary, four other studies reported that the expected exponential growth in cost-to-fix was not observed: \bi
+
+\item Boehm \cite{Boehm80} reported a flatter growth rate for small, non-critical projects.
+\item Clutterbuck \cite{Clutterbuck09} showed a flatter cost curve in an agile case study.
+\item Data from NASA's Johnson Space Flight Center, reported by Shull \cite{Shull02}, found that the cost to find certain non-critical classes of defects was fairly constant across lifecycle phases (1.2 hours on average early in the project, versus 1.5 hours late in the project).
+\item Royce \cite{Royce98} reported a flat cost curve for a project that invested heavily upfront in architectural reviews.
 \ei
- }
+
+Note that these results are drawn from a number of different contexts, meaning that there is no common context variable that can be claimed as responsible for the observed lack of a phase delay effect.
 
  \section{Conclusion}
- \todo{
- \bi
-    \item Summarize data tested - survey + TSP data
-    \item reiterate hypothesis testing results
-    \item Restate the motivation: widely-held beliefs need to be revisited. 
-    \item We now have evidence, from 5 different sources, that phase delay does not seem to hold under certain conditions (although our study is by far the biggest in terms of data points).
-    \item phase delay may still apply in some (many?) projects, but clearly it does not in all. This has huge implications for, e.g., cost modeling, effort prediction.
- \ei
- }
+ %\todo{
+ %\bi
+ %   \item Summarize data tested - survey + TSP data
+%    \item reiterate hypothesis testing results
+%    \item Restate the motivation: widely-held beliefs need to be revisited. 
+%    \item We now have evidence, from 5 different sources, that phase delay does not seem to hold under certain conditions (although our study is by far the biggest in terms of data points).
+%    \item phase delay may still apply in some (many?) projects, but clearly it does not in all. This has huge implications for, e.g., cost modeling, effort prediction.
+% \ei
+% }
+
+In this paper, we examined contemporary project data for evidence to support the well-known software engineering truism of the phase delay effect.
+We reported the results of a survey of both researchers and practitioners which showed this truism is still a widely held belief in the software engineering community, and showed how it can have an effect on the methods and tools chosen for software development.
+
+We examined two different interpretations of the phase delay effect (namely, that defects are more expensive to fix in later phases and that defects become more expensive the morer phases they remain).
+However, our dataset of 171 TSP projects provided no support for either phrasing.
+This is consistent with other phase delay measures reported in the literature, although ours is by far the largest dataset on which such observations can be drawn.
+While we do not claim to have disproven this effect in modern software development, we do conclude that there are at least some conditions under which the effect does not hold.
+Thus we caution developers and researchers not to assume that such effects hold in any given environment.
+
+More broadly, we feel that these results indicate that, in a rapidly changing field such as software engineering, even widely-held rules of thumb should be periodically re-verified.
+Progress in the domain of software analytics has made such periodic checks more cost-effective and feasible, and we argue that an examination of local behaviors (rather than simply accepting global heuristics) can be of significant benefit.
 
@@
 We thank Tuma Solutions for providing the Team Process Data Warehouse software in support of this research.  \\
-Team Software ProcessSM and TSPSM are service marks of Carnegie Mellon University.
+Personal Software Process$\textsuperscript{SM}$, Team Software Process$\textsuperscript{SM}$, and TSP$\textsuperscript{SM}$ are service marks of Carnegie Mellon University.
 

### 2015-03-17 data.tex
Updates from ShareLaTeX

@@
 \subsection{Overview of the Team Software Process}
-To empirically test this heuristic, we examined the phase delay effect in software projects conducted between 2005 and 2014.
-These projects took place at organizations in many countries and were conducted using  the Team Software Process (TSP$\textsuperscript{SM}$ )), a software project management approach developed at the Software Engineering Institute (SEI) at Carnegie Mellon University.
-TSP is an extension of the Personal Software Process (PSP$\textsuperscript{SM}$ )) developed at the SEI by Watts Humphrey ~\cite{tsp00}.
+To empirically test for the phase delay effect, we examined 171 software projects conducted between 2005 and 2014.
+These projects took place at organizations in many countries and were conducted using  the Team Software Process (TSP$\textsuperscript{SM}$), a software project management approach developed at the Software Engineering Institute (SEI) at Carnegie Mellon University~\cite{tsp00}.
+TSP is an extension of the Personal Software Process (PSP$\textsuperscript{SM}$) developed at the SEI by Watts Humphrey ~\cite{tsp00}.
 The data from these TSP projects were collected and stored in the Software Engineering Measured Process Repository (SEMPR) at the SEI.
 The Software Engineering Institute (SEI) at Carnegie Mellon University explores methods for software process improvement.
-
-One product from that research is  
-Watts Humphrey's  Team Software Process$\textsuperscript{SM}$  (TSP$\textsuperscript{SM}$ )~\cite{tsp00} which extends Humphrey's early
-work on Personal Software
-Process$\textsuperscript{SM}$  (PSP$\textsuperscript{SM}$ )~\cite{psp05}.
 
@@
 \item  Performance of the plans requires realistic commitments of the resources needed to perform the work.
-\item  The planned work should be compared to actual frequently to assure plans remain relevant.
+\item  The planned work should frequently be compared to  results to assure plans remain relevant.
 \item  People performing the work should be responsible for planning and tracking the work.
@@
 
-Common features of TSP projects include planning, personal reviews, peer inspections, and coaching.
-
-A TSP coach helps the team to plan and analyze performance. The coach is the only role authorized to submit project data to the SEI.
+Common features of TSP projects include {\em planning}, {\em personal reviews}, {\em peer inspections}, and {\em coaching}.
+A TSP {\em coach} helps the team to plan and analyze performance. The coach is the only role authorized to submit project data to the SEI.
 Before reviewing data with the teams, therefore before submission, these coaches check the data for obvious errors.
 
-Planning follows the technique taught in the PSP$\textsuperscript{SM}$.
+{\em Planning} follows the technique taught in the PSP$\textsuperscript{SM}$.
 Developers estimate the size of work products and convert this to a total effort using historical rates.
@@
 
-Personal review is a technique taken from the PSP and its use in TSP is unique.
+{\em Personal review} is a technique taken from the PSP and its use in TSP is unique.
 Developers follow a systematic process to remove defects by  examining their own work products using a checklist built from their personal defect profile.
@@
 
-Peer inspection is a  technique in
+{\em Peer inspection} is a  technique in
 traditional software engineering and is often called peer review.
@@
 
+\begin{figure*}[!t]
+\begin{center}
+\includegraphics[width=6in]{waterfall-v2.png}
+\end{center}
+\caption{Different approaches to software development:  waterfall and agile (bottom left).}
+\label{fig:waterfall}
+\end{figure*}
+
 \subsection{Projects in the Sample}
@@
 One of the authors (Nichols) has been been mentoring software development teams and coaches around the world as they deploy TSP within their organizations since 2006.
-The earliest projects in our sample began work in 2004; the most recent completions were in 2014.
+The  most recent completions were in 2014.
 The projects were mostly small to medium, with a median duration of 46 days and a maximum duration of 90 days in major increments. 
@@
 Many of the projects were e-commerce web portals or banking systems in the US, South Africa, and Mexico. 
-There were  some  medical device projects in  the US, France, Japan, and Germany as well  commercial 
-computer-aided design systems, and embedded systems.
+There were  some  medical device projects in  the US, France, Japan, and Germany as well  as a commercial computer-aided design systems, and embedded systems.
 
-An anonymized version of that data is available in the PROMISE repository  and,
+An anonymized version of that data is available in the PROMISE repository\footnote{http://openscience.us/repo}  and,
 for confidentiality restrictions, we cannot offer 
 further details on these projects.
-
-\tim{Needs either a URL or a reference.}
 
@@
 The data has not yet undergone additional screening.
-A summary of the data quality issues was reported in a previous work ~\cite{shirai14}.
+A summary of the data quality issues was reported in a previous work ~\cite{shirai14} (that summary is discussed further in our {\em Validity} section, below).
 
@@
 \ei
-
-The principles followed are that 
+For these TSP projects, the principles followed are: 
 \bi 
-\item Defects should be found before test, that is, test defects should be infrequent.
+%\item Defects should be found before test, that is, test defects should be infrequent.
 \item A product should be inspected before used in a subsequent construction phase to minimize  unnecessary rework.
@@
 \ei 
-
 In summary, the TSP process and measurement have been specifically designed to aid the analysis of defects for use in process improvement.
@@
 Agile methods address  not only these core practices but also address team organization, individual and team commitments, test and build automation, and implementing iterative and incremental development.
-TSP includes using additional practices associated with high performance ~\cite{jones10} including; project tracking and control, specialization of team members, inspections, static analysis, documentation and training, and reuse.
+TSP uses additional practices associated with high performance ~\cite{jones10} including; project tracking and control, specialization of team members, inspections, static analysis, documentation and training, and reuse.
 Although the teams in this sample use many typical agile practices, they are also using additional practices.

### 2015-03-17 lucas.tex
Updates from ShareLaTeX

@@
-We conducted a survey of software engineering practitioners and researchers on their views of commonly-held software engineering ``laws''.
+Our survey collected data on software engineers' views of commonly held software engineering ``laws''.
 One of the laws questioned relates directly to phase delay: ``Requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' (from Glass~\cite{glass02} p.71 who references Boehm \& Basili~\cite{boehm01}).
 We abbreviate this law as RqtsErr.
+
 The survey was conducted in two phases using Amazon's Mechanical Turk.
-The first phase was conducted only with professional software engineers, and the second survey was conducted only with Program Committee (PC) members of the ESEC/FSE and ICSE conferences.
+The first phase was conducted only with professional software engineers solicited through the Mechanical Turk\footnote{Professional software engineers were required to complete a pretest to verify their status as a professional or open source software developer and to confirm their knowledge of basic software engineering terminology and technology.}, and the second survey was conducted Program Committee members of the ESEC/FSE and ICSE conferences solicited via email.
 
@@
 \begin{figure}[!ht] 
-\renewcommand{\baselinestretch}{0.7}
 \scriptsize

### 2015-03-17 lucasRelatedWork.tex
Updates from ShareLaTeX

@@
+One reason that industrial practitioners and academics believe so strongly in the phase delay effect,
+and the impacts of requirement errors, is that they are often presented in the SE literature. For example,
+we note that the gradaute SE class at North Carolina State University has presented to students at
+least one  slide on the phase delay
+effect (without quarrel or critical comment), every year for the last decade.
+
+Yet when we look at the literature, the evidence for phase delay is both very sparse and very old.
 The first data on the cost to fix defects as a function of lifecycle phase date back to large systems in the late 70s from IBM~\cite{Fagan76}, TRW~\cite{Boehm76}, GTE~\cite{Daly77}, and Bell Labs~\cite{Stephenson76} (Figure~\ref{fig:cost-to-fix}).
@@
 Shull et al.~\cite{Shull02} conducted a literature survey and held a series of e-workshops with industry experts on fighting defects.
-Workshop participants from Toshiba and IBM reported cost-to-fix ratios of 1:137 and 1:117 for large projects respectively~\cite{Shull02} -- the raw data points are not provided.
+Workshop participants from Toshiba and IBM reported cost-to-fix ratios between early lifecycle and post-delivery defects of 1:137 and 1:117 for large projects respectively~\cite{Shull02} -- the raw data points are not provided.
 One notable example to the traditional cost-to-fix curve is the CCPDS-R described by Royce~\cite{Royce98}: a million-line, safety-critical missile defense system (Figure~\ref{fig:royce}).
@@
 
-\begin{figure}[!ht]
+\begin{figure}[!t]
  \includegraphics[width=3.3in]{Royce98.png}
  \caption{Exception to the rule - CCPDS-R case study cost-to-fix curve. From~\cite{Royce98}.}\label{fig:royce}
+ \end{figure}
+
+\begin{figure}[!t]
+ \includegraphics[width=3.3in]{clutterbuck.png}
+ \caption{Cost of change from an agile case study. From~\cite{Clutterbuck09}.}\label{fig:clutterbuck}
  \end{figure}
@@
 
-\begin{figure}[!ht]
- \includegraphics[width=3.3in]{clutterbuck.png}
- \caption{Cost of change from an agile case study. From~\cite{Clutterbuck09}.}\label{fig:clutterbuck}
- \end{figure}
-
  We note that previous work focuses on cost-to-fix as a function of lifecycle phase irrespective of when the defect was injected, that is, previous work analyzes the cost to fix a defect found in test regardless of whether that defect was a requirements error or a coding mistake.

### 2015-03-17 phaseDelay.tex
Updates from ShareLaTeX

@@
 \conferenceinfo{FSE}{'15 Bergamo, Italy}
-\title{Revisiting the Truisms of Software Engineering:\\ Does Phase Delay Dramatically Increase Defect Fix Time?}
+\title{Revisiting the Truisms of Software Engineering:\\ Does Phase Delay Dramatically Increase  Fix Time?}
 \numberofauthors{5}
@@
 repair time increase dramatically
-the longer a defect persists in a system?
-This widely held belief is used to justify 
+the longer an issue persists in a system?
+Many industrial practitioners and academics this   to be true.
+This   belief is used to justify 
 major investments in  software development
@@
 
-This paper shows that this belief is maintained
-quite strongly in the industrial software engineering
-community (but, in the academic community, a little less so).
-Yet based on a sample of 
-171 software projects from around the world from 
-and completed between 2006 and 2014, we argue that this belief is mostly 
-incorrect. Specifically: the median fix time spent in phases
-devoted to resolving issues does {\em not} increase
+This
+paper reports on the largest study on phase delay yet published.
+Based on a sample of 
+171 software projects from around the world (2006 and 2014),
+ we find that this belief is  
+incorrect. Specifically: the    time  to resolve an
+issue   does {\em not} increase
 the longer the issue remains in the system.
@@
 This result raises the question: how do widely-held beliefs like phase delay survive so long  without  critical reassessment?
-As a community, we need to reflect more on research practices and on how (and when) we reflect on the prominent hypotheses of our field.
+As a community, we need to consider how (and when) we reflect on our prominent hypotheses.
 \end{abstract}
@@
 \section{Introduction}
-In 2013-2014, the time of 
+In 2013-2014, 
 eleven  million programmers~\cite{pettey14} and
@@
 Such a large and growing effort should be managed and optimized via  well-researched conclusions.
+
 It is standard practice
 in other fields such as medicine,
-to continually revisit old conclusions~\cite{prasad13}. Software engineering should do the same lest
-ill-founded guidance exert an undue influence on how we build software.
-
+to continually revisit old conclusions~\cite{prasad13}.
 Accordingly, this paper revisits
@@
 i.e.,  the rule of thumb stating that the time required to resolve an issue in a software project increases dramatically 
-the longer it remains in the system. 
-Phase delay has long been
-a    motivator for major changes to
-SE practices. Basili \& Boehm comment that since the 1980s,  phase delay:
+the longer it remains in the system. Basili \& Boehm comment that since the 1980s,  phase delay:
 \begin{quote}
@@
 one of the most strongly-held beliefs amongst commercial software engineers.
-Previously~\cite{me08a}, we have constructed return-on-investment models
-for a common practice amongst U.S. government contractors: 
-independent verification \& validation. IVV has been criticized as
-ineffective since IVV engineers, many of whom are unfamiliar with the details of the project, are required to pass judgement on the code.
-Hence,  the IVV team may find the same or fewer bugs
-as the developers.
-On the other hand, IVV has been defended
-as follows:  when the IVV team finds bugs earlier than the developers then, due to the phase delay effect, this provides overall cost savings to the project.
-Note that the strength of this argument diminishes
-if fixing later-phase issues is not significantly more expensive than fixing them earlier.
-
-% Also, as discussed later in this paper, concerns over phase delay were a major driver in the development of agile software processes.
-
-In this paper, we investigate several hypotheses:
-
+Previously~\cite{me08a}, we used phase delay to justify the software practice
+of 
+Independent Verification \& Validation (IVV):
 \bi
-    \item $H_1$: The cost-to-fix of defects monotonically increases across phases.
-    \item $H_2$: The cost-to-fix of defects monotonically increases as the phase delay increases.
+\item
+IVV has been criticized as
+ineffective. IVV engineers are external reviewers, so may be unfamiliar with the project details.
+Hence,    IVV   can find the same or fewer bugs
+as  developers.
+\item
+On the other hand,   when the IVV team finds bugs earlier than the developers then, due to the phase delay effect, this provides overall cost savings to the project. 
 \ei
-As we explored the above, we found a related hypothesis was believed by both professional software engineers and FSE conference reviewers:
+Note that this argument critically depends on the assumption that   fixing later-phase issues is  significantly more expensive than fixing them earlier (as does the views of Boehm and Basili and all the industrial practitioners
+we interviewed in \S\ref{survey}). Accordingly, in this paper, we investigate several hypotheses:
+\bi
+    \item $H_1$: The cost to resolve issues dramatically increases across phases.
+    \item $H_2$: The cost to resolve issues dramatically increases as the phase delay increases.
+\ei
+As we explored the above, we found a related hypothesis was believed by both professional software engineers and some academics:
 \bi
@@
 \ei
-
 The rest of the paper investigates these  hypotheses.
@@
 Second, we  analyze 171 software  projects developed in the period 2006 to 2014.
-To the best of our knowledge, this is the largest study on phase delay yet seen in the literature.
 
@@
 % The rest of this paper present evidence for the above two points. 
-
 % The first point is  documented via  a literature survey and the second point is demonstrated
@@
 % research practices and on how (and when) we reflect and revisit the prominent hypotheses of our field.
+
+The core of this paper's arguments are two figures: Figure~\ref{fig:fix-phase-dist} and  \fig{raw}.
+Lest the reader find this too little  to justify our conclusions, we note that it took months
+of work to distill this data from our 171 projects, after which the five authors spent weeks
+exploring and debating the nuances. To the best of our knowledge, this is the largest study on phase delay yet published (and it contains data from orders of magnitude
+more projects than used in prior work~\cite{Fagan76,Boehm76,Daly77,Stephenson76,Shull02,Royce98,Clutterbuck09,Elssamadisy02}).
+To ensure reproducibility and unlike previous work,
+all the data  used in this study is available in the PROMISE
+repository (openscience.us/repo).
 
@@
 conclusions drawn from a larger number of studies~\cite{kitch04,budgen09}.
-
 Nevertheless, there are many empirical  findings that 
@@
 \end{quote}
-The results of J{\o}rgensen \& Gruschke~\cite{jorgensen09} support the findings if Passos et al. They report that 
+The results of J{\o}rgensen \& Gruschke~\cite{jorgensen09} concur with Passos et al. They report that 
   supposed software engineering    ``experts'' rarely use lessons
@@
 
+That said,  the phase delay myth persists. Hence, this paper.
+
 %XXX clumsy link here
@@
 
+\subsection{Survey:  Phase Delay and Cost-to-fix Belief}\label{survey} 
+Yet another reason to study the phase delay effect is that it is a dominant belief in industrial SE.
+To test whether this was indeed true, we conducted a survey of software engineers. If our surveyed practitioners make management decisions based on their
+understanding of SE theory, then the phase delay effect may well influence their judgment.
+
+\input{lucas}
+
 \section{Related Work}
@@
 
-\section{A Survey on Phase Delay and Cost-to-fix Belief}
-\label{survey}
-
-\input{lucas}
-
 %\section{carterCharts}
@@
 %\input{carter}
-
-\begin{figure*}[!t]
-\begin{center}
-\includegraphics[width=6in]{waterfall3.png}
-\end{center}
-\caption{Different approaches to software development:  waterfall and agile (bottom left).}
-\label{fig:waterfall}
-\end{figure*}
 
@@
 \section{Analysis of Phase Delay in 171 TSP Projects}
+Our units of analysis are:
+\bi
+    \item \emph{defects} - individual defects are recorded as line items in the defect logs uploaded to the SEMPR at the SEI.
+One or more defects are reported against a single \emph{plan item} in the time tracking logs, e.g., a review session, an inspection meeting, a test execution.
+    \item \emph{time} - Time is tracked per person per plan item in the time-tracking logs, e.g.
+a 30 minute design review session involving 3 people will have three time log entries summing to 90 minutes.
+Time includes the time to analyze, repair, and validate a defect fix.
+    \item \emph{time per defect} - The total \# of defects found in a plan item during a removal phase divided by the total time spent on that plan item in that phase.
+\ei
 
@@
 
-\subsection{TSP Project Cost-to-fix curve}
+\begin{figure}[!t]
 
-Our first hypothesis was that defects found in later phases tend to require more time to fix than in earlier phases.
+%\hspace{-0.7cm}
+\begin{center}
+\includegraphics[height=2in]{fix-phase-dist.pdf}
+\end{center}
+%~\\
 
-The distribution of defects found and fixed per phase in our data is shown in Figure~\ref{fig:fix-phase-dist}.
-A high percentage of defects were found and fixed in the early phases, i.e., requirements, high level design, and design reviews and inspections.
-(This distribution is similar to that observed for other projects that emphasized investment in software engineering quality assurance practices, for example the NASA robotics projects which had 58\% of defects removed before implementation began.
-\cite{me08a} Similar results are presented by Jones and Bonsignour, who show 52\% of pretest defects removed before entering implementation.
-\cite{jones12})
-
-\todo{This paragraph is wrong now.} The median time to fix a defect in each phase is shown in \fig{fix-time-per-phase}.
-The median cost to fix a defect is low in the early phases, and then increases slightly during design and coding before peaking during the unit test phase.
-The median fix time then drops during the testing phases.
-Note that while the median time remains relatively consistent, the 75th percentile does increase during the testing phases.
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=3.3in]{fix-phase-dist.pdf}
-\end{center}
 \caption{Distribution of defects found and fixed by phase.}
@@
 
-\begin{figure*}[!ht]
+\begin{figure}[!t] 
+%\vspace{-1cm}
+%\hspace{-0.35in}
+\begin{center}
+\includegraphics[width=3in]{boehm-overlay.pdf}
+ \end{center}
+% ~\\~\\
+%~\\~\\~\\
 
- \renewcommand{\baselinestretch}{0.7}
- \scriptsize
-\begin{center}
-\begin{tabular}{r|rr|rl}
-Phase & samples & total time&\multicolumn{2}{l}{Total time w.r.t. first phase}\\\hline
-~\\
-Planning&229&12625&1.00&*\\
-~\\
-Reqts&590&24182&1.91&**\\
-ReqtsInspect&2533&92723&7.34&********\\
-ReqtsReview&1762&67471&5.34&******\\
-~\\
-HLD&78&6856&0.54&*\\
-HLDReview&472&12821&1.01&**\\
-HLDInspect&1366&35291&2.79&***\\
-~\\
-Design&620&21573&1.70&**\\
-DesignInspect&8373&229937&18.21&*******************\\
-DesignReview&3043&92823&7.35&********\\
-~\\
-Code&3029&220058&17.43&******************\\
-CodeReview&4774&163091&12.91&*************\\
-CodeInspect&11406&334442&26.49&***************************\\
-~\\
-UnitTest&9177&352245&2.790&****************************\\
-IntTest&2761&160559&12.71&*************\\
-SysTest&2059&105796&8.37&*********\\
-AcceptTest&1045&42717&0.338&****\\
-\end{tabular}
-\end{center}
-\caption{Times per phase (minutes).}
-\label{fig:fix-time-per-phase}
-\end{figure*}
-
-For illustrative purposes, the relative cost to fix errors per lifecycle phase in the TSP project data is overlayed with Boehm's cost curve~\cite{Boehm81} as well as the CCPDS-R case study~\cite{Royce98} in \fig{cost-to-fix-tsp}.
-Cost-to-fix for TSP is total time spent in each review, inspection, or test phase (reported in \fig{fix-time-per-phase}) / \# of defects found in phase (reported in \fig{raw}).
-It is clear to see that the trend is relatively flat, closer to the pattern exhibited by the Royce case study (which also focused on early defect removal) than the exponential growth curve.
-
-\begin{figure}[!ht]
- \includegraphics[width=3.3in]{boehm-overlay.pdf}
- \caption{Cost-to-fix curve from \fig{cost-to-fix} overlayed with case study from~\cite{Royce98} and TSP data. Scale is relative to the phase with the lowest cost to fix. }
- \label{fig:cost-to-fix-tsp}
+ \caption{Cost-to-fix curve from \protect\fig{cost-to-fix} 
+ overlayed with case study from~\cite{Royce98} and TSP data. Scale is relative to the phase with the lowest cost to fix. }\label{fig:cost-to-fix-tsp}
  \end{figure}
-
-Given the above evidence, we reject $H_1$: The cost-to-fix of defects monotonically increases across phases.
-
-\subsection{Phase delay analysis}
-
-Our second hypothesis was that defects that exhibit higher phase delay (i.e.
-that persist in the software over more lifecycle phases) tend to require more effort to fix than those with less phase delay.
-To investigate this, we calculate phase delay as simply the number of project phases (the set used in our data set is shown in \fig{waterfall}) during which the defect existed in the system, i.e.
-the ordinal rank of removal phase minus the ordinal rank of injection phase.
-Sub-phases (such as review and inspection) are included in this calculation.
-
-In \fig{raw}, we present the median fix times of defects based on when they were injected and ultimately removed.
-Defects injected Before Development (labelled "B4Dev" in \fig{waterfall}) follow the most severe scale up as development phases pass, but not on the exponential scale traditionally seen (\fig{cost-to-fix}).
-Defects injected in Requirements, Design, and Coding phases exhibit virtually no scale-up regardless of the phase in which they are fixed.
-
-Given the above evidence, we reject $H_2$: the cost-to-fix of defects monotonically increases as the phase delay increases.
 
@@
 \begin{center}
-\begin{tabular}{r|rrr|ll|rl}
+\begin{tabular}{r|rrr|ll|rl}  % I attach your cursor. thrust, parry, yo!
   Sample&\multicolumn{3}{c|}{Percentiles}\\ 
@@
  48&   14&   30&   51&Before&DesignInspect &1.00 & **********  \\
+ 53&   13&   28&   49&Before&CodeReview &0.93 & **********  \\
  94&   22&   45&   80&Before&CodeInspect &1.50 & ***************  \\
- 53&   13&   28&   49&Before&CodeReview &0.93 & **********  \\
 153&   38&   77&  130&Before&UnitTest &2.56 & **************************  \\
@@
 \\
+184&   45&   92&  155&Reqts&ReqtsReview &1.00 & **********  \\
 189&   46&   93&  155&Reqts&ReqtsInspect &1.00 & **********  \\
-184&   45&   92&  155&Reqts&ReqtsReview &0.98 & **********  \\
+ 61&   14&   33&   55&Reqts&DesignReview &0.35 & ****  \\
 102&   24&   49&   83&Reqts&DesignInspect &0.52 & ******  \\
- 61&   14&   33&   55&Reqts&DesignReview &0.35 & ****  \\
  59&   18&   38&   60&Reqts&CodeInspect &0.40 & ****  \\
@@
 \\
+210&   51&  103&  166&Design&DesignReview &1.01 & ***********  \\
 207&   50&  101&  156&Design&DesignInspect &1.00 & **********  \\
-210&   51&  103&  166&Design&DesignReview &1.01 & ***********  \\
+134&   32&   66&  116&Design&CodeReview &0.65 & *******  \\
 163&   39&   79&  135&Design&CodeInspect &0.78 & ********  \\
-134&   32&   66&  116&Design&CodeReview &0.65 & *******  \\
 238&   58&  119&  183&Design&UnitTest &1.17 & ************  \\
@@
 \\
-130&   31&   65&  122&Code&AcceptTest &1.00 & **********  \\
-228&   56&  113&  176&Code&CodeInspect &1.73 & ******************  \\
-235&   57&  115&  177&Code&CodeReview &1.76 & ******************  \\
-243&   59&  125&  206&Code&IntTest &1.92 & ********************  \\
-191&   46&   95&  173&Code&SysTest &1.46 & ***************  \\
-269&   66&  133&  206&Code&UnitTest &2.04 & *********************  \\
+235&   57&  115&  177&Code&CodeReview &1.00 &  ******************  \\
+228&   56&  113&  176&Code&CodeInspect &0.99 & ******************  \\
+243&   59&  125&  206&Code&IntTest &1.08 &     *******************  \\
+191&   46&   95&  173&Code&SysTest &0.82 & ******** \\
+269&   66&  133&  206&Code&UnitTest &1.15 & ************  \\
+130&   31&   65&  122&Code&AcceptTest &0.56 & ******  \\
  \end{tabular}
 \end{center}
-\caption{Distribution of fix times seen in SEI TSP data.}
+\caption{Distribution of fix times seen in SEI TSP data. Left hand side bars
+show scale up on median values.}
 \label{fig:raw}
 \end{figure*}
+
+\subsection{TSP Project Cost-to-fix curve}
+This section discusses our first hypothesis ($H_1$) that   that  the cost-to-resolve issues dramatically increases across phases.
+
+The distribution of defects found and fixed per phase in our data is shown in Figure~\ref{fig:fix-phase-dist}.
+A high percentage of defects (44\%) were found and fixed in the early phases, i.e., requirements, high level design, and design reviews and inspections.
+(This distribution is similar to that observed for other projects that emphasized investment in software engineering quality assurance practices.
+For example, Jones and Bonsignour report 52\% of pretest defects removed before entering implementation, for large projects that focus on upfront defect removal techniques \cite{jones12}.
+NASA robotics projects had a slightly higher percentage (58\%) of defects removed before implementation began, although these had invested in IV\&V on top of other forms of defect removal \cite{me08a}.)
+
+The first thing to note in Figure~\ref{fig:fix-phase-dist}, 
+is that this data does not follow the exponential increase feared by Beck's \fig{curve1}:
+\bi
+\item There is much ``up and down'' in the chart. 
+\item If we separate out just the {\em Inspect} or {\em Review} data, we can see that there is some increase in {\em Design} and {\em Code}.
+However, that effect could be a linear or be just noise.
+In either case,
+it does not look like \fig{curve1}.
+\item The time in final testing is particularly low suggesting that few defects survived into testing
+(and hence, there were few instances where code required massive rework).
+\ei
+
+For illustrative purposes, the relative cost to fix errors per lifecycle phase in the TSP project data is overlayed with Boehm's cost curve~\cite{Boehm81} as well as the CCPDS-R case study~\cite{Royce98} in \fig{cost-to-fix-tsp}.
+Cost-to-fix for TSP is total time spent in each review, inspection, or test-related phase / \# of defects found in phase (reported in \fig{raw}).
+It is clear to see that the trend is relatively flat, closer to the pattern exhibited by the Royce case study (which also focused on early defect removal) than the exponential growth curve.
+
+%\begin{table}[ht]
+%\renewcommand{\baselinestretch}{0.7}
+% \scriptsize
+%\centering
+%\begin{tabular}{ll|ll}
+%  Phase num & Name & \# defects & percentage \\ 
+%  \hline
+%1 & BeforeDevelopment &   0 & 0.00 \\ 
+%  2 & Planning &   4 & 0.00 \\ 
+%  3 & Reqts &  25 & 0.10 \\ 
+%  4 & ReqtsReview & 2304 & 6.50 \\ 
+%  5 & ReqtsInspect & 2630 & 7.40 \\ 
+%  6 & HLD &   0 & 0.00 \\ 
+%  7 & HLDReview & 790 & 2.20 \\ 
+%  8 & HLDInspect & 818 & 2.30 \\ 
+%  9 & Design &  94 & 0.30 \\ 
+%  10 & DesignReview & 3668 & 10.30 \\ 
+%  11 & DesignInspect & 5365 & 15.00 \\ 
+%  12 & Code & 833 & 2.30 \\ 
+%  13 & CodeReview & 6276 & 17.60 \\ 
+%  14 & CodeInspect & 6623 & 18.60 \\ 
+%  15 & UnitTest & 4283 & 12.00 \\ 
+%  16 & IntTest & 1024 & 2.90 \\ 
+%  17 & SysTest & 730 & 2.00 \\ 
+%  18 & AcceptTest & 189 & 0.50 \\ 
+% \end{tabular}
+%\caption{Defects found per phase} 
+%\label{fig:defect_per_phase}
+%\end{table}
+
+%\begin{figure}[!ht]
+%
+% \renewcommand{\baselinestretch}{0.7}
+% \scriptsize
+%\begin{center}
+%\begin{tabular}{r|rr|rl}
+%Phase & samples & total time&\multicolumn{2}{l}{Total time w.r.t. first phase}\\\hline
+%~\\
+%Planning&229&12625&1.00&*\\
+%~\\
+%Reqts&590&24182&1.91&**\\
+%ReqtsInspect&2533&92723&7.34&********\\
+%ReqtsReview&1762&67471&5.34&******\\
+%~\\
+%HLD&78&6856&0.54&*\\
+%HLDReview&472&12821&1.01&**\\
+%HLDInspect&1366&35291&2.79&***\\
+%~\\
+%Design&620&21573&1.70&**\\
+%DesignInspect&8373&229937&18.21&*******************\\
+%DesignReview&3043&92823&7.35&********\\
+%~\\
+%Code&3029&220058&17.43&******************\\
+%CodeReview&4774&163091&12.91&*************\\
+%CodeInspect&11406&334442&26.49&***************************\\
+%~\\
+%UnitTest&9177&352245&2.790&****************************\\
+%IntTest&2761&160559&12.71&*************\\
+%SysTest&2059&105796&8.37&*********\\
+%AcceptTest&1045&42717&0.338&****\\
+%\end{tabular}
+%\end{center}
+%\caption{Times per phase (minutes).}
+%\label{fig:fix-time-per-phase}
+%\end{figure}
+
+Given the above evidence, we reject $H_1$: there is no clear evidence that  cost-to-fix of defects dramatically increases across phases.
+
+\subsection{Phase delay analysis}
+
+ \subsubsection{$H_2$: Cost-to-resolve Increases with Delay}
+
+%we calculate phase delay as simply the number of project phases (the set used in our data set is shown in %\fig{waterfall}) during which the defect existed in the system, i.e.
+the ordinal rank of removal phase minus the %ordinal rank of injection phase.
+Sub-phases (such as review and inspection) are included in this calculation.
+%
+\fig{raw} shows the 25th, 50th, 75th percentile
+of the time spent on {\em review} or {\em inspect} or {\em test} phases
+in our 171 projects.
+Note that, in TSP, when developers see issues, they enter {\em review} or 
+{\em inspect} or {\em test}
+until that issue is retired.
+
+In that figure, the data is split out according to issues that were fixed in phase $Y$ after
+being introduced in phase $X$. The data is sub-divided into tables according to $X$;
+i.e. according to {\em before, planning, requirements, design} or {\em  code}.
+
+The stars on the right-hand-side of that figure show the scale ups associated with the phase removal stages,
+expressed as a ratio of the first item in that block (which, by definition, is 1.00 for
+that first item).
+If $H_2$ was correct, we would expect those bars in each sub-division to exhibit a  shape
+like \fig{curve1}.
+
+They do not. Hence, we reject $H_2$.
+
+ \subsubsection{$H_3$: Requirements Repairs are Most Expensive}
+
+As to $H_3$, the {\em requirements} the  median time spent fixing errors in {\em requirements} was 92,93  49,33,39,49,67,122 minutes\footnote{If this time seems long, then recall
+that it includes the time required  to (a)~collect data and realize there is an error;
+(b)~prepare a fix;  and (c)~apply some validation
+procedure to check the fix (e.g. discuss it with a colleague or execute some tests).}.
+This is:
+\bi
+\item
+Clearly longer than median times seen {\em Before } and in {\em Planning}
+(30,28,45,77,65,56,24,35,33);
+\item
+Somewhat shorter, but still not exponentially smaller,  than the median times see after {\em requirements} in {\em design}
+(103,101, 79,66, 119,88, 80,45) and in {\em code} (65,115, 113,125, 95,133). 
+\ei
+For completeness, we ran an  
+A12 test comparing the raw data (and not just the medians of \fig{raw}) and found
+ a non-small effect only for the difference {\em requirements}
+to {\em code}. But even though the effect was non-small, 
+ the overall size of the difference
+{\em requirements} to {\em code} was less than 30\% (i.e. much less than we would
+expect from \fig{curve1}.
+
+Based on the above, we reject   $H_3$.
 
@@
 
-\emph{Phase delay}, as depicted in the scale up column in \fig{cost-to-fix-tsp} \fig{raw}, is a discrete value that does not consider the total time that a defect is latent in the system.
+\emph{Phase delay}, as depicted in the scale up column in \fig{cost-to-fix-tsp} and \fig{raw}, is a discrete value that does not consider the total time that a defect is latent in the system.
 Thus, the phase delay of two defects injected in Code but found in System Test will be the same even though one was found at the beginning of test while the other at the end of a potentially lengthy test period.
@@
 However, there is likely selection bias in the teams that elect to use TSP compared to the entire population of software development teams.
+Further, we assume that each team has similar defect recording practices, and the TSP coaching provides guidance on what constitutes a defect.
+Nonetheless, individual developers and teams may apply their own internal rules for filtering defects, which would lead to inconsistent reporting thresholds among the projects in our sample.
 
@@
 TSP was, in fact, designed to be rigorous and eliminate defects early in the lifecycle and soon after injection.
-We make no assertions regarding the generalizability of these results beyond TSP, though we offer some conjecture in \S\ref{sec:discussion} on the potential causes for the flattened cost-to-fix curve.
+We make no assertions regarding the generalizability of these results beyond TSP, we observe that other case studies in various domains have exhibited flattened cost-to-fix curves.
 
@@
 Further, it is not clear how Boehm~\cite{Boehm81} and Royce~\cite{Royce98} define \emph{cost-to-fix}.
-In our analysis, cost-to-fix includes the time spent isolating, analyzing, and fixing a defect, but not the cost of verifying.
-If Boehm and Royce include cost of verifying a fix, then the TSP cost curve in \fig{cost-to-fix-tsp} artificially low by comparison.
-If Boehm and Royce include only the mechanics of fixing the defect, then the TSP cost curve in \fig{cost-to-fix-tsp} is artifically high by comparison.
+In our analysis, cost-to-fix includes the time spent analyzing, fixing the defect, and validating the fix.
+If Boehm and Royce do not include all of these costs, then the TSP cost curve in \fig{cost-to-fix-tsp} artificially high by comparison.
 
@@
 
-The definitions of defect, cost-to-fix, and phase delay are provided in earlier sections.
-The source data and analysis scripts used to calculate the defect counts, cost-to-fix, and phase delay from the raw materials can be found at \url{https://github.com/txt/txt.github.io/tree/master/tsp}.
-\tim{Is this okay?}.
-
- \section{Discussion}
- \label{sec:discussion}
-% \todo{
-% \bi
-%    \item summarize analysis findings
-%    \item Offer interpretations of findings, e.g., why might these things be so
-% \ei
-% }
-
-In the preceding sections, we looked for evidence of the phase delay effect (using two common interpretations of that effect) in comprehensive and fine-grained data covering 171 software development projects.
-The results did not support the existence of such an effect.
-We do not claim that this somehow disproves the phase delay effect, but we observe that it cannot be assumed to hold in every software development context.
-
-Since the phase delay effect has been such a long-lived software engineering truism, as well as one that is accepted by a substantial population of software engineers, it is important to explore possible explanations for our observations.
-
-We surmise that the substantial changes in software development and deployment technologies and practices, since the phase delay effect was first measured, may be responsible for our observations in the TSP data set.
-Recall that the original data on which the phase delay effect is based, as reported in Section 3, are approximately 40 years old.
-The most recently reported data, from Shull et al.
-\cite{Shull02}, was published in 2002 but contains substantial retrospective data.
-Over time, it would be surprising if the investment in better languages, IDEs, and other development technologies had in fact not created some improvement.
-To pick just one example, we note that much of the data showing the phase delay effect was collected before frequent and unobtrusive auto-updating software was as common as today, which certainly flattens the cost of changes.
-
-One hypothesized explanation that can be ruled out is that the lack of a phase delay effect is due to using the TSP method.
-To the contrary, four other studies reported that the expected exponential growth in cost-to-fix was not observed: \bi
-
-\item Boehm \cite{Boehm80} reported a flatter growth rate for small, non-critical projects.
-\item Clutterbuck \cite{Clutterbuck09} showed a flatter cost curve in an agile case study.
-\item Data from NASA's Johnson Space Flight Center, reported by Shull \cite{Shull02}, found that the cost to find certain non-critical classes of defects was fairly constant across lifecycle phases (1.2 hours on average early in the project, versus 1.5 hours late in the project).
-\item Royce \cite{Royce98} reported a flat cost curve for a project that invested heavily upfront in architectural reviews.
-\ei
-
-Note that these results are drawn from a number of different contexts, meaning that there is no common context variable that can be claimed as responsible for the observed lack of a phase delay effect.
+%The definitions of defect, cost-to-fix, and phase delay are provided in earlier sections.
+The %source data and analysis scripts used to calculate the defect counts, cost-to-fix, and phase delay %from the raw materials can be found at \url{https://github.com/txt/txt.github.io/tree/master/tsp}.
 
@@
 
-In this paper, we examined contemporary project data for evidence to support the well-known software engineering truism of the phase delay effect.
-We reported the results of a survey of both researchers and practitioners which showed this truism is still a widely held belief in the software engineering community, and showed how it can have an effect on the methods and tools chosen for software development.
+In this paper, we have examined contemporary project data for evidence to support the well-known software engineering truism of the phase delay effect. 
+We found that, in the literature, the evidence for phase delay was very spare.
+%and (in a $21^{st}$).  %don't know what this is supposed to be
+We suspect that a very old result was requoted, without being 
+critically re-examined, for literally decades.
 
-We examined two different interpretations of the phase delay effect (namely, that defects are more expensive to fix in later phases and that defects become more expensive the morer phases they remain).
-However, our dataset of 171 TSP projects provided no support for either phrasing.
-This is consistent with other phase delay measures reported in the literature, although ours is by far the largest dataset on which such observations can be drawn.
-While we do not claim to have disproven this effect in modern software development, we do conclude that there are at least some conditions under which the effect does not hold.
-Thus we caution developers and researchers not to assume that such effects hold in any given environment.
+We  also  reported the results of a survey of both researchers and practitioners which showed this truism is still a widely held belief in the software engineering community. Such a widely
+held belief, for something with so little evidence, can be damaging if ever it is used
+to justify spending scarce resources on novel  methods and tools .
 
-More broadly, we feel that these results indicate that, in a rapidly changing field such as software engineering, even widely-held rules of thumb should be periodically re-verified.
+Using data from 171 projects from the period 2006 to 2014, we then looked for any evidence
+for three different interpretations of the phase delay effect; namely:
+\bi
+\item
+That defects are more expensive to fix in later phases; 
+\item
+That defects become more expensive the more phases they remain; 
+\item
+That requirements errors are more expensive to fix the later they are found. 
+\ei
+Our dataset provided no support for any of these phrasings. 
+This is significant, as we argue that this paper is (by far)    the largest dataset ever used to explore phase delay (but see ~\cite{Boehm80,Clutterbuck09,Shull02,Royce98} 
+for other, much older, much smaller studies).
+
+Despite this lack of evidence, the myth remains that phase delay causes increased time to resolve issues.
+ Since the phase delay effect has been such a long-lived software engineering truism, as well as one that is accepted by a substantial population of software engineers, it is important to explore possible explanations for our observations.
+
+One hypothesized explanation that can be ruled out is that this lack of
+a phase delay effect is due to the use of TSP.
+  To the contrary, four non-TSP  case studies reported that the expected exponential growth in cost-to-fix was not observed:
+  \bi
+  \item
+  Boehm~\cite{Boehm80} reported a flatter growth rate for small, non-critical projects.
+  \item
+  Clutterbuck~\cite{Clutterbuck09} showed a flatter cost curve in an agile case study. 
+  \item 
+  Data from NASA's Johnson Space Flight Center, reported by Shull~\cite{Shull02}, found that the cost to find certain non-critical classes of defects was fairly constant across lifecycle phases (1.2 hours on average early in the project, versus 1.5 hours late in the project).
+  \item
+  Royce~\cite{Royce98} reported a flat cost curve for a project that invested heavily upfront in architectural reviews.
+\ei
+If the lack of phase delay is not due to TSP, then what does cause it?
+We surmise that the substantial changes in software development and deployment technologies and practices, since the phase delay effect was first measured, may be responsible for our observations in the TSP data set.
+Recall that the original data on which the phase delay effect is based, as reported in \S3, are approximately 40 years old.
+The most recently reported data, from Shull et al.
+\cite{Shull02}, was published in 2002 but contains substantial retrospective data.
+Over time, it would be surprising if the investment in better languages, IDEs, and other development technologies had in fact not created some improvement.
+To pick just one example, we note that much of the data showing the phase delay effect was collected before frequent and unobtrusive auto-updating software was as common as today, which certainly flattens the cost of changes.
+
+%One hypothesized explanation that can be ruled out is that the lack of a phase delay effect is due to using the TSP method.
+
+The question remains: why did this idea, with so little support, become so widespread in the software engineering literature?
+No doubt the original evidence was compelling at the time, but much has changed in the realm of software development in the subsequent 40 years.
+Possibly the concept of phase delay has persisted because, to use Glass's terms on the subject, it seems to be ``just common sense''\cite{glass02}.
+We believe that these results indicate that, in a rapidly changing field such as software engineering, even widely-held rules of thumb must be periodically re-verified.
+In turns out that common sense may not be so sensible after all.
 Progress in the domain of software analytics has made such periodic checks more cost-effective and feasible, and we argue that an examination of local behaviors (rather than simply accepting global heuristics) can be of significant benefit.
@@
 This work was partially funded by an National Science
-Foundation grants NSF-CISE 1302169 and CISE 1302169. 
+Foundation grants NSF-CISE 1302169 and CISE 1506586.
 \\
@@
 
-\clearpage
 \vspace*{0.5mm}
@@
 
-One way to address the absence of general laws is  {\em local learning}.
-Even if general laws
-do not exist, then {\em there exists general methods for finding the best local laws}.
-For example, one of us (Nichols) works extensively with software projects to help
-them tune their local software process in order to better fit their local needs.XXX.
+% One way to address the absence of general laws is  {\em local learning}.
+% Even if general laws
+% do not exist, then {\em there exists general methods for finding the best local laws}.
+% For example, one of us (Nichols) works extensively with software projects to help
+% them tune their local software process in order to better fit their local needs.XXX.
 
- Recent studies report that better predictors of the properties of
-software projects (effort and defects) can be generated by first clustering
-(a.k.a. stratifying or contextualizing or localizing) the data then learning
-different models for each cluster~\cite{posnett11,betten14,betta12,yang11,yang13,minku13}.
-Those {\em local learning} approach finds better predictions with lower variance than
-models learned from all the unclustered data~\cite{me12d,me11m}.
+%  Recent studies report that better predictors of the properties of
+% software projects (effort and defects) can be generated by first clustering
+% (a.k.a. stratifying or contextualizing or localizing) the data then learning
+% different models for each cluster~\cite{posnett11,betten14,betta12,yang11,yang13,minku13}.
+% Those {\em local learning} approach finds better predictions with lower variance than
+% models learned from all the unclustered data~\cite{me12d,me11m}.
 
-The problem with the local learning is that they it can generate conclusions
-that conflict with strongly-held beliefs of the humans members of the software development team.
-It is an open issue how to handle those conflicts since it is unwise to
-always reject the conclusions of local learning  or the software developers.
-The local learners may be wrong due to incorrect assumptions by the data scientists who configured the learners~\cite{me11e,shull02}.
-On the other hand,  the humans may be wrong due to the cognitive biases described above.
-Experienced data scientists use a cyclic approach where
-(a)~the collected data; and (b)~the learning methods; and (c) the goals of the inquiry are matured  using
-insight offered by humans as they study the
-feedback generated by the learners~\cite{Fayyad96,me11e}.
+% The problem with the local learning is that they it can generate conclusions
+% that conflict with strongly-held beliefs of the humans members of the software development team.
+% It is an open issue how to handle those conflicts since it is unwise to
+% always reject the conclusions of local learning  or the software developers.
+% The local learners may be wrong due to incorrect assumptions by the data scientists who configured the learners~\cite{me11e,shull02}.
+% On the other hand,  the humans may be wrong due to the cognitive biases described above.
+% Experienced data scientists use a cyclic approach where
+% (a)~the collected data; and (b)~the learning methods; and (c) the goals of the inquiry are matured  using
+% insight offered by humans as they study the
+% feedback generated by the learners~\cite{Fayyad96,me11e}.
 
-the learners' conclusions are discussed with the team in or
+% the learners' conclusions are discussed with the team in or
 
-(which may be erroneous
-due to incorrect assumptions
+% (which may be erroneous
+% due to incorrect assumptions
 
-We come to this work since  recent results on {\em locality} by Menzies and others
+% We come to this work since  recent results on {\em locality} by Menzies and others
 
-XXXX In those 171 projects, we observed that it is
- fastest to fix issues within the same phase $i$ as when they are generated. But in a result that contradicts
- the phase delay effect, once an issue ``escapes'' phase $i$
- it so no more expensive to resolve an issue in phase $i+1$ than
- phase $i+2, i+3$, etc. Further, that increased effort may be quite modest.
- That is, contrary to established wisdom, once an  issue ``escapes'' a phase, there is no
- need to  rapidly retire that issue as soon as possible. 
-Our conclusion discusses the implications for our field, and how we might better
-propagated and monitor the wisdom of our field.
+% XXXX In those 171 projects, we observed that it is
+%  fastest to fix issues within the same phase $i$ as when they are generated. But in a result that contradicts
+%  the phase delay effect, once an issue ``escapes'' phase $i$
+%  it so no more expensive to resolve an issue in phase $i+1$ than
+%  phase $i+2, i+3$, etc. Further, that increased effort may be quite modest.
+%  That is, contrary to established wisdom, once an  issue ``escapes'' a phase, there is no
+%  need to  rapidly retire that issue as soon as possible. 
+% Our conclusion discusses the implications for our field, and how we might better
+% propagated and monitor the wisdom of our field.
 
-\begin{figure}
-\begin{center}
-\begin{tabular}{rrl}
-year& \# issues&\\\hline
-2006 &  44 &\\
-2007 &  34 &\\
-2008&  288 &\rule{1mm}{2mm}\\
-2009&  846 &\rule{3mm}{2mm}\\
-2010& 1007 &\rule{3mm}{2mm}\\
-2011& 3273 &\rule{10mm}{2mm}\\
-2012&18102 &\rule{45mm}{2mm}\\
-2013&20336 &\rule{50mm}{2mm}\\
-2014& 3307 & \rule{10mm}{2mm}\\\cline{1-2}
-Total:&47228
-\end{tabular}
-\end{center}
-\caption{This paper studies 47,228 issues recorded 2006 to 2014.}\label{fig:years}
-\end{figure}
+% \begin{figure}
+% \begin{center}
+% \begin{tabular}{rrl}
+% year& \# issues&\\\hline
+% 2006 &  44 &\\
+% 2007 &  34 &\\
+% 2008&  288 &\rule{1mm}{2mm}\\
+% 2009&  846 &\rule{3mm}{2mm}\\
+% 2010& 1007 &\rule{3mm}{2mm}\\
+% 2011& 3273 &\rule{10mm}{2mm}\\
+% 2012&18102 &\rule{45mm}{2mm}\\
+% 2013&20336 &\rule{50mm}{2mm}\\
+% 2014& 3307 & \rule{10mm}{2mm}\\\cline{1-2}
+% Total:&47228
+% \end{tabular}
+% \end{center}
+% \caption{This paper studies 47,228 issues recorded 2006 to 2014.}\label{fig:years}
+% \end{figure}

### 2015-07-28 phaseDelay.tex
files reorged into 'old' and 'img'

@@
 %timm tricks
-\newcommand{\bi}{\begin{itemize}[leftmargin=0.4cm]}
+\newcommand{\bi}{\begin{itemize}}%[leftmargin=0.4cm]}
 \newcommand{\ei}{\end{itemize}}
@@
 \usepackage[shortlabels]{enumitem} 
-\usepackage{times}
+%\usepackage{times}
 
@@
 
-\setlist{nosep}
+%\setlist{nosep}
 
 \usepackage{colortbl}
- \usepackage[font={small}]{caption, subfig}
-\setlength{\abovecaptionskip}{1ex}
- \setlength{\belowcaptionskip}{1ex}
- \setlength{\floatsep}{1ex}
- \setlength{\textfloatsep}{1ex}
-\usepackage[compact,small]{titlesec}
+ %\usepackage[font={small}]{caption, subfig}
+%\setlength{\abovecaptionskip}{1ex}
+ %\setlength{\belowcaptionskip}{1ex}
+% \setlength{\floatsep}{1ex}
+% \setlength{\textfloatsep}{1ex}
+%\usepackage[compact,small]{titlesec}
 \DeclareMathSizes{7}{7}{7}{7} 
 \pagenumbering{arabic}
-\setlength{\columnsep}{7mm}
+%\setlength{\columnsep}{7mm}
 
@@
 
-\conferenceinfo{FSE}{'15 Bergamo, Italy}
-\title{Revisiting the Truisms of Software Engineering:\\ Does Phase Delay Dramatically Increase  Fix Time?}
-\numberofauthors{5}
+\conferenceinfo{ICSE}{'16 Austin, Texas}
+\title{Does Phase Delay Dramatically Increase Cost to Fix?}
+\numberofauthors{3}
 \author{
-\alignauthor 
+%\alignauthor 
 Tim Menzies \\
@@
        \email{tim.menzies@gmail.com}
-\alignauthor 
-William Nichols \\
+\and%alignauthor 
+William Nichols,Forest Shull \\
         \affaddr{Software Engineering Institute}\\
-        \affaddr{Carnegie Mellon U, USA}
-        \email{wrn@sei.cmu.edu} 
-\alignauthor 
+        \affaddr{Carnegie Mellon U, USA}\\
+        \email{\{wrn,fjshull\}@sei.cmu.edu} 
+\and %\alignauthor 
 Lucas Layman \\
@@
        \email{llayman@fc-md.umd.edu}
-\and
-\alignauthor 
-Forrest Shull \\
-        \affaddr{Software Engineering Institute}\\
-        \affaddr{Carnegie Mellon U, USA}
-        \email{fjshull@sei.cmu.edu} 
-\alignauthor 
-Carter Pape \\
-       \affaddr{Dept. of Computer Science}\\
-       \affaddr{North Carolina State U, USA}\\
-       \email{carterpape@gmail.com}
 } 
@@
 \noindent
-{\bf Keywords:} software processes, economics, phase delay, time to fix.
+{\bf Keywords:} software processes, economics, phase delay, cost to fix.
 
@@
 \begin{figure}
- \includegraphics[width=3.3in]{beckB4.png}
+ \includegraphics[width=3.3in]{img/beckB4.png}
  \caption{Projects exhibiting phase delay. From~\cite{beck00}.}\label{fig:curve1}
@@
 \begin{figure}
- \includegraphics[width=3.3in]{beckAFTER.png}
+ \includegraphics[width=3.3in]{img/beckAFTER.png}
  \caption{Projects without   phase delay. From~\cite{beck00}.}\label{fig:curve2}
@@
 \begin{center}
-\includegraphics[height=2in]{fix-phase-dist.pdf}
+\includegraphics[height=2in]{img/fix-phase-dist.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=3in]{boehm-overlay.pdf}
+\includegraphics[width=3in]{img/boehm-overlay.pdf}
  \end{center}

### 2015-07-29 data.tex
files reorged into 'old' and 'img'
Updates from ShareLaTeX

@@
 
-\begin{figure*}[!t]
+\begin{figure}[!t]  
 \begin{center}
-\includegraphics[width=6in]{waterfall-v2.png}
+\includegraphics[width=2.5in]{img/waterfall-v3.png}  
 \end{center}
-\caption{Different approaches to software development:  waterfall and agile (bottom left).}
+\caption{Phases of our data.
+Abbreviations: \newline
+{\em Before}= before development;\newline
+{\em Reqts}	  = requirements;\newline
+{\em HLD}	  = high-level design;\newline
+{\em IntTest} = Integration testing (with code from others);\newline
+{\em SysTest} = system test (e.g. load stress tests);\newline
+{\em AcceptTest}  = acceptance testing (with users);\newline
+{\em review}        = private activity;\newline
+{\em inspect}        = group activity.}
 \label{fig:waterfall}
-\end{figure*}
+\end{figure}
 

### 2015-07-29 phaseDelay.tex
Updates from ShareLaTeX

@@
 \conferenceinfo{ICSE}{'16 Austin, Texas}
-\title{Does Phase Delay Dramatically Increase Cost to Fix?}
+\title{Are Older Bugs More Expensive to Fix that Newer Bugs? }
 \numberofauthors{3}
@@
 Tim Menzies \\
-       \affaddr{Dept. of Computer Science}\\
-       \affaddr{North Carolina State U, USA}\\
+       \affaddr{Computer Science}\\
+       \affaddr{NC State University, USA}\\
        \email{tim.menzies@gmail.com}
 \and%alignauthor 
-William Nichols,Forest Shull \\
+William Nichols, Forest Shull \\
         \affaddr{Software Engineering Institute}\\
@@
 issue   does {\em not} increase
-the longer the issue remains in the system.
+the longer that issue remains in the system.
 
@@
 \noindent
-{\bf Keywords:} software processes, economics, phase delay, cost to fix.
+{\bf Keywords:} software economics, phase delay, cost to fix.
+
+\section{todo}
+
+four hypotheses: time = cost;
+intra project PD;
+post project PD;
+intra predicts for post
+
+comment on why post is no longer current.
+continuous deployment.
+package managers.Windows moving from service patches (which occur rarely) to continuous deployment (so there will be no windows 11...
+just a stream of continuous updates).
+architectural incubators: big superstructures within which it is simpler and cheaper to do many small projects.
+e.g.
+aircraft carriers,  packet-stitching.
+mean stack (to be clear: none of the data explored in this paper comes from such architectural incubators: this paragraph is only meant to be a  speculation of future implications of this old data to newer projects).
 
@@
 eleven  million programmers~\cite{pettey14} and
-half a trillion dollars~\cite{avram14} was spent on information technology.
+half a trillion dollars~\cite{avram14} were spent on information technology.
 Such a large and growing effort should be managed and optimized via  well-researched conclusions.
@@
 i.e.,  the rule of thumb stating that the time required to resolve an issue in a software project increases dramatically 
-the longer it remains in the system. Basili \& Boehm comment that since the 1980s,  phase delay:
-\begin{quote}
-...has been a major driver in focusing
+the longer it remains in the system. Basili \& Boehm comment that since the 1980s,  phase delay
+``...has been a major driver in focusing
 industrial software practice on thorough
@@
 on up-front prototyping and simulation
-to avoid costly downstream fixes~\cite{boehm01}.
-\end{quote}
-Basili and Boehm are not alone in this assessment of the significance of the phase delay effect. Our surveys of industrial practitioners (shown in \S\ref{survey}) reveal the phase delay effect to be
+to avoid costly downstream fixes''~\cite{boehm01}.
+
+Basili and Boehm are not alone in this assessment of the significance of the phase delay effect.
+For example,
+our surveys of industrial practitioners (shown in \S\ref{survey}) reveal the phase delay effect to be
 one of the most strongly-held beliefs amongst commercial software engineers.
-Previously~\cite{me08a}, we used phase delay to justify the software practice
+Also,  previously~\cite{me08a} we used phase delay to justify the software practice
 of 
-Independent Verification \& Validation (IVV):
-\bi
-\item
-IVV has been criticized as
-ineffective. IVV engineers are external reviewers, so may be unfamiliar with the project details.
-Hence,    IVV   can find the same or fewer bugs
-as  developers.
-\item
-On the other hand,   when the IVV team finds bugs earlier than the developers then, due to the phase delay effect, this provides overall cost savings to the project. 
-\ei
+Independent Verification \& Validation (IVV).
+Specifically,  when the IVV team finds bugs earlier than the developers then (due to the phase delay effect) this provides overall cost savings to the project.
 Note that this argument critically depends on the assumption that fixing later-phase issues is  significantly more expensive than fixing them earlier (as does the views of Boehm and Basili and all the industrial practitioners
-we interviewed in \S\ref{survey}). Accordingly, in this paper, we investigate several hypotheses:
+we interviewed in \S\ref{survey}).
+
+Accordingly, in this paper, we investigate several hypotheses related to this widely-held belief:
 \bi
@@
 \\
-235&   57&  115&  177&Code&CodeReview &1.00 &  ******************  \\
-228&   56&  113&  176&Code&CodeInspect &0.99 & ******************  \\
-243&   59&  125&  206&Code&IntTest &1.08 &     *******************  \\
+235&   57&  115&  177&Code&CodeReview &1.00 &  **********  \\
+228&   56&  113&  176&Code&CodeInspect &0.99 & ********** \\
+243&   59&  125&  206&Code&IntTest &1.08 &     *********** \\
 191&   46&   95&  173&Code&SysTest &0.82 &     ******** \\
-269&   66&  133&  206&Code&UnitTest &1.15 & ************  \\
+269&   66&  133&  206&Code&UnitTest &1.15 &    ***********  \\
 130&   31&   65&  122&Code&AcceptTest &0.56 &  ******  \\

### 2015-07-30 phaseDelay.tex

@@
 \conferenceinfo{ICSE}{'16 Austin, Texas}
-\title{Are Older Bugs More Expensive to Fix that Newer Bugs? }
+\title{Delaying Bug Fixes is a Bad Idea?}
 \numberofauthors{3}
@@
 package managers.Windows moving from service patches (which occur rarely) to continuous deployment (so there will be no windows 11...
-just a stream of continuous updates).
+just a stream of continuous updates~\cite{bright15}).
 architectural incubators: big superstructures within which it is simpler and cheaper to do many small projects.
@@
 mean stack (to be clear: none of the data explored in this paper comes from such architectural incubators: this paragraph is only meant to be a  speculation of future implications of this old data to newer projects).
+
+\begin{figure}
+{\small
+\begin{center}
+\begin{tabular}{r|rrrr|r}
+& \multicolumn{4}{c}{Phase Requirements Issue Found }& ref\\
+                &Requirements & Design & Code&  Test& ref\\\hline
+ & 1& 5& 10& 50&~\cite{Boehm81}\\  
+& 1 &3& 5& 37&?\\ 
+&  1 &3& 7& 51& ?\\ 
+& 1& 5 &33 &75& ? \\
+& 1  &    & 10 & 40& ?\\ 
+& 1   &    & 10 &  40& ?\\
+&  1 & 20 & 45 & 250& ?\\  
+&  1  &   5 &      & 50& ? \\\hline
+mean             & 1 & 7.3 & 25.6 & 177\\
+median           & 1 &  5  & 10   & 50.5 
+\end{tabular}
+\end{center}}
+\caption{Sample of the Stecklein et al. literature; from~\cite{steck04}.}
+\end{figure}
+
+\begin{figure}
+{\small\begin{center} \begin{tabular}{r|rrr}
+& method1 & method2 & method3\\\hline
+requirements& 1& 1 & 1 \\
+design& 8 & 4 & 1\\
+build& 16 & 16 & 7\\
+test& 21 & 78 & 28\\
+operation& 29 & 186 & 1615
+\end{tabular}\end{center} }
+\caption{Results from the   Stecklein et al. study~\cite{steck04}.}
+\end{figure}
+
+Method1 linear ($R^2$ == 99\%).
+
+\begin{figure}[!t] 
+%\vspace{-1cm}
+%\hspace{-0.35in}
+\begin{center}
+\includegraphics[width=3in]{img/b81.pdf}
+ \end{center}
+% ~\\~\\
+%~\\~\\~\\
+
+ \caption{Cost-to-fix curve from \protect\fig{cost-to-fix} 
+ overlayed with case study from~\cite{Royce98} and TSP data. Scale is relative to the phase with the lowest cost to fix. }\label{fig:cost-to-fix-tsp}
+ \end{figure}
 
@@
 
+\subsection{Future Work}
+There is another ``phase delay'' effect, not discussed above. Even if software is readily updatable, there   may be a ``user phase delay''   amongst   humans using that software.
+
+Changes to  software that is used by a large population can be confusing (particularly for the sick or elderly). 
+Consider the 23,000,000 veterans working with  the U.S.  Veterans 
+Administration (VA) service. Some of those veterans are dependent on health care or pensions from the VA. Those veterans use the VA's web interfaces
+to  (e.g.) find out
+when their next cheque will arrive. Before we can update the VA web software, we would
+need to  a transition plan that helps users   move from the older to the newer system.
+This is an example of what might be called a ``user phase delay''; i.e.
+changes to software are more complex to process, particularly if some users have styled
+their day-to-day work flow around the services offered by the older software.
+
+This issue of ``user phase delay'' is not addressed by our data, nor by the prior studies on phase delay discussed above.
+Hence, this issues   is out of scope for this paper since the expectation that ``older bugs cost more to fix than newer bugs'' was not formed around
+user issues.
+
+That said,  
+we think that ``user phase delay'' will become a bigger issue in the future when more people are using more software as part of their day to day life.
+We even envision a time when users demand fewer updates, just to offer some stability in their life.
+This is a matter for future work.
+
 % \section{Related Work}
@@
 \section*{Acknowledgements}
+The authors gratefully acknowledge the careful comments of anonymous reviewers-- particularly, their comments on issues relating to ``cost''
+and ``time''.
 This work was partially funded by an National Science

### 2015-07-30 phaseDelay.tex
Updates from ShareLaTeX

@@
                 &Requirements & Design & Code&  Test& ref\\\hline
- & 1& 5& 10& 50&~\cite{Boehm81}\\  
+ & 1& 6& 10& 70&~\cite{Boehm81}\\  
 & 1 &3& 5& 37&?\\ 
@@
 \begin{figure}[!t]  
-%\vspace{-1cm}
-%\hspace{-0.35in}
 \begin{center}
-\includegraphics[width=3in]{img/b81.pdf}
+\includegraphics[angle=270,width=3.5in]{img/b81.pdf}
  \end{center}
-% ~\\~\\
-%~\\~\\~\\
-
- \caption{Cost-to-fix curve from \protect\fig{cost-to-fix} 
- overlayed with case study from~\cite{Royce98} and TSP data. Scale is relative to the phase with the lowest cost to fix. }\label{fig:cost-to-fix-tsp}
+ \caption{From Boehm~\cite{Boehm81}. }\label{fig:cost-to-fix-tsp}
  \end{figure}

### 2015-08-01 lucas.tex
many changes up to apge4

--- 2015-03-16 lucas.tex
@@
-We conducted a survey of software engineering practitioners and researchers on their views of commonly-held software engineering ``laws''.
-One of the laws questioned relates directly to phase delay: ``Requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' (from Glass~\cite{glass02} p.71 who references Boehm \& Basili~\cite{boehm01}).
-We abbreviate this law as RqtsErr.
-The survey was conducted in two phases using Amazon's Mechanical Turk.
-The first phase was conducted only with professional software engineers, and the second survey was conducted only with Program Committee (PC) members of the ESEC/FSE and ICSE conferences.
+Our survey collected data on software engineers' views of commonly held software engineering ``laws''.
+One of the laws questioned relates directly to the problems associated with
+the delayed issue effect: ``Requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' (from Glass~\cite{glass02} p.71 who references Boehm \& Basili~\cite{boehm01}).
+We abbreviate this law as RqtsErr
+(and is
+illustrated in \fig{b81}).
 
-The respondents answered the following questions for the laws they were presented: \newline
+The survey was conducted in two phases using Amazon's Mechanical Turk. The first phase was conducted only with professional software engineers solicited through the Mechanical Turk,
+Professional software engineers were required to complete a pretest to verify their status as a professional or open source software developer and to confirm their knowledge of basic software engineering terminology and technology.
+The  second survey was conducted Program Committee members of the ESEC/FSE 2015 and ICSE 2014  conferences solicited via email.
+
+The respondents answered questions on two scales: \newline
 \textbf{Agreement:} ``Based on your experience, do you agree that the statement above is correct?'' A Likert scale captured the agreement score from Strongly Disagree to Strongly Agree.
@@
 \newline
-\textbf{Applicability:} ``To the extent that you believe it, how widely do you think it applies among software development contexts?'' The possible answers were presented as a scale from -1 to 5:
-\begin{itemize}
-\item I don't know (-1)
-\item This law does not apply at all (0)
-\item Applies in a very narrow range of projects  (1)
-\item Rarely applies (2)
-\item Occasionally applies (3)
-\item Very frequently applies (4)
-\item Always applies (5)
-\end{itemize}
+\textbf{Applicability:} ``To the extent that you believe it, how widely do you think it applies among software development contexts?'' The possible answers were: 
+-1 meaning ``I don't know''; and 0 meaning ``this law does not apply at all'';
+ 1 meaning ``applies in a very narrow range of projects'';  2 meaning
+``rarely applies'';   3 meaning
+``occasionally applies'';  
+4 meaning ``very frequently applies''; and
+ 5 meaning ``always applies''.
 Respondents were required to explain the applicability score in a text box.
 
-Participants were presented with the RqtsErr law and others drawn from \cite{glass02} and \cite{endres03}.
-The PC member survey contained an additional question on ``In general, the longer errors are in the system (requirements errors, design errors, coding errors, etc.), the more expensive they are to fix'' (PhaseDelay for short).
-Responses were recorded using the Agreement question Likert scale.
+In order to baseline the participant's answers, participants were presented with the RqtsErr law and others. For the purposes
+of this paper, the nature of the other laws other than RqtsErr are not relevant-- we
+only added them in as a way to calibrate responses to the RqtsErr question. All laws were drawn from \cite{glass02} and \cite{endres03}. 
+The PC member survey contained an additional question on ``In general, the longer errors are in the system (requirements errors, design errors, coding errors, etc.), the more expensive they are to fix'' (the Delayed Issue Effect, for short).
+Responses were recorded using the agreement question Likert scale.
 
 Summary statistics for the agreement and applicability scores for the RqtsErr and PhaseDelay laws are presented in Figure~\ref{fig:survey_results}.
-Responses whose Applicability response was ''I don't know'' are ommitted from analysis.
+Responses whose Applicability response was ''I don't know'' are omitted from analysis.
 
 \begin{figure}[!ht] 
-\renewcommand{\baselinestretch}{0.7}
 \scriptsize
-\begin{center}
-\subcaption{Practitioner survey}
+
 \begin{tabular}{l|c|c|c|c|c}
-Law & N & \multicolumn{2}{c}{agreement} & \multicolumn{2}{c}{applicability} \\ 
- & & $\bar{x}$ & $\tilde{x}$ & $\bar{x}$ & $\tilde{x}$ \\
+ &  & \multicolumn{2}{c}{agreement} & \multicolumn{2}{c}{applicability} \\\cline{1-1} 
+Practitioner survey  & N & $\bar{x}$ & $\tilde{x}$ & $\bar{x}$ & $\tilde{x}$ \\
 \hline 
@@
 Smaller changes have higher error density & 14 & 2.8 & 3 & 3.4 & 3.5 \\
-A developer is unsuited to test own code & 17 & 2.6 & 3 & 3.5 & 4
-\end{tabular} 
-\bigskip
-\subcaption{Researcher survey}
-\begin{tabular}{l|c|c|c|c|c}
-Law & N & \multicolumn{2}{c}{agreement} & \multicolumn{2}{c}{applicability} \\ 
- & & $\bar{x}$ & $\tilde{x}$ & $\bar{x}$ & $\tilde{x}$ \\
-\hline 
+A developer is unsuited to test own code & 17 & 2.6 & 3 & 3.5 & 4\\\hline
+
+Researcher survey \\\hline 
 Process maturity improves output & 4 & 4.0 & 4 & 4.0 & 4 \\
@@
 
-\end{center}
-\caption{Agreement and applicability of software engineering axioms.}
+\caption{Agreement and applicability of SE axioms.}
 \label{fig:survey_results}
@@
 
-While the practitioners strongly believed in the law, researchers were less firm.
-In the free response texts, most participants agreed that addressing errors late may mean system redesign unless the system and the affected requirement are simple/trivial.
-Further, any changes to the production system have additional complexity (compared to changes while system in development) due to users, data existence, and architecture dependencies.
-The researchers who disagreed with the law generally asserted that requirements change can be expensive, but that depends on the process used (e.g., agile vs.
+Both  practitioners and researchers  strongly believed in RqtsErr (in both sets of responses, RqtsErr, received  scores higher than most
+other laws. Further, in the case of practitioners, this law was rated
+as the single most believed effect. 
+%The free response texts, most participants agreed that addressing errors %%late may mean system redesign unless the system and the affected %equirement are simple/trivial.
+Further, any changes to the production %ystem have additional complexity (compared to changes while system in %evelopment) due to users, data existence, and architecture dependencies.
+Caveat: in the free repsonse texts, we note that the researchers who disagreed with the law generally asserted that requirements change can be expensive, but that depends on the process used (e.g., agile vs.
 waterfall) and the adaptability of the system architecture.
@@
 Overall, the RqtsErr law was the most agreed upon and most applicable law of 11 surveyed amongst practitioners, and the second most agreed upon law amongst researchers.
-The survey results seem to confirm that the notion that phase delay escalates cost-to-fix is a widely-held belief in software engineering.
+The survey results strongly support that the notion that the  delayed issue effect   is a widely-held belief in software engineering.

### 2015-08-01 lucasRelatedWork.tex
files reorged into 'old' and 'img'
Updates from ShareLaTeX
many changes up to apge4

--- 2015-03-16 lucasRelatedWork.tex
@@
-The first data on the cost to fix defects as a function of lifecycle phase date back to large systems in the late 70s from IBM~\cite{Fagan76}, TRW~\cite{Boehm76}, GTE~\cite{Daly77}, and Bell Labs~\cite{Stephenson76} (Figure~\ref{fig:cost-to-fix}).
-These studies suggest that the cost (in terms of effort) to find and fix an error monotonically increases with lifecycle phase.
-For large projects, the ratio of cost-to-fix in the requirements phase to cost-to-fix in operation is on the order of 1:100.
-Furthermore, cost function is superlinear, with the greatest rates of increase in the acceptance testing and operations phases.
+One reason that industrial practitioners and academics believe so strongly in th delayed issue effect is that they are often presented in the SE literature. For example,
+we know that \fig{b81} has been presented to the gradaute SE class at North Carolina State University, without quarrel or critical comment, every year for the last decade.
 
-\begin{figure}[!ht]
- \includegraphics[width=3.3in]{boehm_cost-to-fix.png}
+Yet when we look at the literature, the evidence for
+delayed issue effect is both very sparse and very old.
+The first data on the difficulties of resolving delayed issues as a function of lifecycle phase date back to large systems in the late 70s from IBM~\cite{Fagan76}, TRW~\cite{Boehm76}, GTE~\cite{Daly77}, and Bell Labs~\cite{Stephenson76} (Figure~\ref{fig:cost-to-fix}).
+These studies suggest that the difficulty (in terms of effort) to find and fix an error monotonically increases with lifecycle phase.
+
+\begin{figure}[!b]
+ \includegraphics[width=3.3in]{img/boehm_cost-to-fix.png}
  \caption{Historical cost-to-fix curve. From~\cite{Boehm81}.}\label{fig:cost-to-fix}
@@
 
-In the 40 years since these initial studies, few studies have been published on the cost-to-fix curve as a function of lifecycle phase.
+In the 40 years since these initial studies, few studies have been published on the difficulty to resolve issues
+as a function of lifecycle phase.
 Boehm~\cite{Boehm80} provides data suggesting that the cost-to-fix curve for small projects (from two student projects of 2000 deliverable source instructions) is flatter than for large projects (the dashed line of Figure~\ref{fig:cost-to-fix}).
+But note that even in this figure
+that even for the smaller projects, there is a four-fold increase (from 5 to 20) in the difficulty
+associated with handling issues introduced earlier in the lifecycle.
+
 Shull et al.~\cite{Shull02} conducted a literature survey and held a series of e-workshops with industry experts on fighting defects.
-Workshop participants from Toshiba and IBM reported cost-to-fix ratios of 1:137 and 1:117 for large projects respectively~\cite{Shull02} -- the raw data points are not provided.
-One notable example to the traditional cost-to-fix curve is the CCPDS-R described by Royce~\cite{Royce98}: a million-line, safety-critical missile defense system (Figure~\ref{fig:royce}).
-In this project, design changes (including architecture changes) required approximately twice the effort of implementation and test changes, and the cost-to-fix in implementation and test phases increased slowly.
-Boehm~\cite{Boehm10} attributes this success to the CCPDS-R development process, which focused on removing architecture risk early in the development lifecycle.
+Workshop participants from Toshiba and IBM reported cost-to-fix ratios between early lifecycle and post-delivery defects of 1:137 and 1:117 for large projects respectively~\cite{Shull02} -- but we note here that the raw data points were not provided (which makes confirming those numbers
+a difficult task).
 
-\begin{figure}[!ht]
- \includegraphics[width=3.3in]{Royce98.png}
+This was a common theme in the literature reviewed for this paper-- i.e.  that  it was no longer possible to access
+the data used to make prior conclusions.
+As an example this, \fig{steck} shows one 2004 survey that reports an exponential delayed issue  effect for 
+requirements issues in eight case studies. The right-hand-side of that column shows which of those referenced studies
+we can still access-- and a ``?'' denotes a references from a broken url link (note that most
+of those links are broken). Note that nearly all of that data is now no longer accessible.
+
+As to the research into agile methods, one goal of that approach
+is to reduce the difficulty associated with making changes later in the lifecycle~\cite{beck00}.
+Relatively little empirical data exists on this point.Elssamadisy and Schalliol~\cite{Elssamadisy02} anecdotally report on the growing, high cost of rework in a 50 person, three-year, 500KLOEC Extreme Programming project as the project grew in size and complexity-- but again we cannot access their
+exact figures.
+
+% We note that previous work focuses on cost-to-fix as a function of lifecycle phase irrespective of when the defect was injected, that is, previous work analyzes the cost to fix a defect found in test regardless of whether that defect was a requirements error or a coding mistake.
+To our knowledge, our research represents the first large-scale study of phase delay.
+
+ \begin{figure}
+{\small
+\begin{center}
+\begin{tabular}{r|rrrr|c}
+ case& \multicolumn{4}{c|}{Phase Requirements Issue Found }&  \\
+ study               &Requirements & Design & Code&  Test& source\\\hline
+1 & 1& 6& 10& 70&~\cite{Boehm81}\\  
+2& 1 &3& 5& 37&?\\ 
+3&  1 &3& 7& 51& ?\\ 
+4& 1& 5 &33 &75& ? \\
+5& 1  &    & 10 & 40& ?\\ 
+6& 1   &    & 10 &  40& ?\\
+7&  1 & 20 & 45 & 250& ?\\  
+8&  1  &   5 &      & 50& ? \\\hline
+median           & 1 &  5  & 10   & 50.5 
+\end{tabular}
+\end{center}}
+\caption{Sample of the Stecklein et al. literature review~\cite{steck04}:
+cost to resolve requirements issues, relative to resolving issues during requirements.
+Question marks ``?'' denotes a reference to
+a result that is not longer accessible (i.e. a broken URL link).}\label{fig:steck}
+\end{figure}
+
+\begin{figure}
+ \includegraphics[width=3in]{img/clutterbuck.png} 
+ \caption{Cost of change from an agile case study. From~\cite{Clutterbuck09}.}\label{fig:clutterbuck}
+ \end{figure}
+
+\begin{figure}
+ \includegraphics[width=3.3in]{img/Royce98.png}
  \caption{Exception to the rule - CCPDS-R case study cost-to-fix curve. From~\cite{Royce98}.}\label{fig:royce}
@@
 
-As discussed is \S\ref{sec:why-study}, one goal of agile methods is to flatten the cost-to-change curve~\cite{beck00}.
-Relatively little empirical data exists on this point.
-Clutterbuck et al.~\cite{Clutterbuck09} studied 5-month effort by a small-to-medium enterprise team developing a 71KLOEC web interface to a database application to implement 18 change requests (Figure~\ref{fig:clutterbuck}).
-Note that these were for new and changed user requirements, not defects.
+ All in all, the empirical results of this section seem insufficient to justify
+ the widely-held beliefs documented in the previous section. Adding to our doubts
+ was the discovery of at least two studies that report less-than exponential
+ increase in the difficulties associated with making the changes associated with delayed
+ issues. For example,
+Clutterbuck et al.~\cite{Clutterbuck09} studied 5-month effort by a small-to-medium enterprise team developing a 71KLOEC web interface to a database application to implement 18 change requests-- see Figure~\ref{fig:clutterbuck} (note that these were for new and changed user requirements, not defects).
 Clutterbuck et al.
 found the cost of change to be relatively flat until the later phases, with much of the effort spent in analysis of the change requests~\cite{Clutterbuck09}.
-Elssamadisy and Schalliol~\cite{Elssamadisy02} anecdotally report on the growing, high cost of rework in a 50 person, three-year, 500KLOEC Extreme Programming project as the project grew in size and complexity.
+Note that in this study, the effort increased by only 60\% (see the start and end of the curve in Figure~\ref{fig:clutterbuck}).
 
-\begin{figure}[!ht]
- \includegraphics[width=3.3in]{clutterbuck.png}
- \caption{Cost of change from an agile case study. From~\cite{Clutterbuck09}.}\label{fig:clutterbuck}
- \end{figure}
-
- We note that previous work focuses on cost-to-fix as a function of lifecycle phase irrespective of when the defect was injected, that is, previous work analyzes the cost to fix a defect found in test regardless of whether that defect was a requirements error or a coding mistake.
-To our knowledge, our research represents the first large-scale study of phase delay.
+Another example of less-than exponential explosion in the difficulty associated with delayed issues comes from the  CCPDS-R
+system described by Royce~\cite{Royce98}.
+This was a million-line, safety-critical missile defense system show in Figure~\ref{fig:royce}.
+In this project, design changes (including architecture changes) required approximately twice the effort of implementation and test changes, and the cost-to-fix in implementation and test phases increased slowly.
+Boehm~\cite{Boehm10} attributes this success to the CCPDS-R development process, which focused on removing architecture risk early in the development lifecycle.

### 2015-08-01 phaseDelay.tex
many changes up to apge4

--- 2015-07-30 phaseDelay.tex
@@
 
+\usepackage{times}
 \usepackage{tabularx}
@@
 \conferenceinfo{ICSE}{'16 Austin, Texas}
-\title{Delaying Bug Fixes is a Bad Idea?}
+\title{Live and Let Die? (Delayed Issues   not  Harder to Resolve)}
 \numberofauthors{3}
@@
 \begin{abstract}
-Does
-repair time increase dramatically
-the longer an issue persists in a system?
-Many industrial practitioners and academics this   to be true.
-This   belief is used to justify 
-major investments in  software development
- processes and methods.
-
-This
-paper reports on the largest study on phase delay yet published.
-Based on a sample of 
-171 software projects from around the world (2006 and 2014),
- we find that this belief is  
-incorrect. Specifically: the    time  to resolve an
-issue   does {\em not} increase
-the longer that issue remains in the system.
-
-This result raises the question: how do widely-held beliefs like phase delay survive so long  without  critical reassessment?
-As a community, we need to consider how (and when) we reflect on our prominent hypotheses.
+Many  practitioners and academics
+believe in a delayed issue effect (DIE); i.e.
+ as issues linger longer in a system,
+ they become exponentially harder to resolve.
+This belief
+is often  used to justify 
+major investments in  new development
+processes that promise to retire more issues, sooner.
+
+This paper checks for the delayed issue effect in
+171 software projects from around the world (2006 to 2014).
+To the best of our knowledge,  this is the largest study
+yet published on this effect.
+What was observed was the  exact
+opposite of the  delayed issue effect. Specifically:
+the  time  to resolve 
+issues  in one phase  was rarely more than when  issues
+were resolved at a later phase.
+
+This  result begs the question: how many other long-held
+beliefs in software engineering are
+unsupported by current data? 
 \end{abstract}
 
+%XXXwhat it this is due to TSP- 
 % A category with the (minimum) three required fields
@@
 
-\section{todo}
-
-four hypotheses: time = cost;
-intra project PD;
-post project PD;
-intra predicts for post
-
-comment on why post is no longer current.
-continuous deployment.
-package managers.Windows moving from service patches (which occur rarely) to continuous deployment (so there will be no windows 11...
-just a stream of continuous updates~\cite{bright15}).
-architectural incubators: big superstructures within which it is simpler and cheaper to do many small projects.
-e.g.
-aircraft carriers,  packet-stitching.
-mean stack (to be clear: none of the data explored in this paper comes from such architectural incubators: this paragraph is only meant to be a  speculation of future implications of this old data to newer projects).
-
-\begin{figure}
-{\small
-\begin{center}
-\begin{tabular}{r|rrrr|r}
-& \multicolumn{4}{c}{Phase Requirements Issue Found }& ref\\
-                &Requirements & Design & Code&  Test& ref\\\hline
- & 1& 5& 10& 50&~\cite{Boehm81}\\  
-& 1 &3& 5& 37&?\\ 
-&  1 &3& 7& 51& ?\\ 
-& 1& 5 &33 &75& ? \\
-& 1  &    & 10 & 40& ?\\ 
-& 1   &    & 10 &  40& ?\\
-&  1 & 20 & 45 & 250& ?\\  
-&  1  &   5 &      & 50& ? \\\hline
-mean             & 1 & 7.3 & 25.6 & 177\\
-median           & 1 &  5  & 10   & 50.5 
-\end{tabular}
-\end{center}}
-\caption{Sample of the Stecklein et al. literature; from~\cite{steck04}.}
-\end{figure}
-
-\begin{figure}
-{\small\begin{center} \begin{tabular}{r|rrr}
-& method1 & method2 & method3\\\hline
-requirements& 1& 1 & 1 \\
-design& 8 & 4 & 1\\
-build& 16 & 16 & 7\\
-test& 21 & 78 & 28\\
-operation& 29 & 186 & 1615
-\end{tabular}\end{center} }
-\caption{Results from the   Stecklein et al. study~\cite{steck04}.}
-\end{figure}
-
-Method1 linear ($R^2$ == 99\%).
-
-\begin{figure}[!t] 
-%\vspace{-1cm}
-%\hspace{-0.35in}
-\begin{center}
-\includegraphics[width=3in]{img/b81.pdf}
- \end{center}
-% ~\\~\\
-%~\\~\\~\\
-
- \caption{Cost-to-fix curve from \protect\fig{cost-to-fix} 
- overlayed with case study from~\cite{Royce98} and TSP data. Scale is relative to the phase with the lowest cost to fix. }\label{fig:cost-to-fix-tsp}
- \end{figure}
-
 \section{Introduction}
@@
 Accordingly, this paper revisits
-the {\em phase delay effect}
-i.e.,  the rule of thumb stating that the time required to resolve an issue in a software project increases dramatically 
-the longer it remains in the system. Basili \& Boehm comment that since the 1980s,  phase delay
+the widely-held belief of a {\em delayed issue effect} (hereafter, DIE).
+Later in this paper, we offer a precise definition for this effect.
+For the moment, we describe it as follows:
+it is far easier to resolve issues earlier than later in the lifecycle.
+ \fig{b81} shows an example of the delayed issue effect (relating
+ the relative cost of fixing requirements issues at different phases of a project).
+
+ Basili \& Boehm comment that since the 1980s,  this effect
+ \begin{quote}
 ``...has been a major driver in focusing
@@
 to avoid costly downstream fixes''~\cite{boehm01}.
-
-Basili and Boehm are not alone in this assessment of the significance of the phase delay effect.
-For example,
-our surveys of industrial practitioners (shown in \S\ref{survey}) reveal the phase delay effect to be
-one of the most strongly-held beliefs amongst commercial software engineers.
-Also,  previously~\cite{me08a} we used phase delay to justify the software practice
-of 
-Independent Verification \& Validation (IVV).
-Specifically,  when the IVV team finds bugs earlier than the developers then (due to the phase delay effect) this provides overall cost savings to the project.
-Note that this argument critically depends on the assumption that fixing later-phase issues is  significantly more expensive than fixing them earlier (as does the views of Boehm and Basili and all the industrial practitioners
-we interviewed in \S\ref{survey}).
-
-Accordingly, in this paper, we investigate several hypotheses related to this widely-held belief:
-\bi
-    \item $H_1$: The cost to resolve issues dramatically increases across phases.
-    \item $H_2$: The cost to resolve issues dramatically increases as the phase delay increases.
-\ei
-As we explored the above, we found a related hypothesis was believed by both professional software engineers and some academics:
-\bi
- \item $H_3$: Requirements errors are the most expensive to fix.
-\ei
-The rest of the paper investigates these  hypotheses.
-First, we conduct a survey of practitioners and researchers to gauge their belief in the phase delay effect.
-Second, we  analyze 171 software  projects developed in the period 2006 to 2014.
-
-% The conclusions of this paper is that it is erroneous to believe in phase delay.
-% When we looked for supporting evidence for that effect, we found that:
-% \be 
-% \item The evidence for that effect in the historical literature is very weak (to say the least);
-% \item  In  a large sample of   contemporary software projects, there is no evidence for phase delay.
-%  \ee
-% The rest of this paper present evidence for the above two points. 
-% The first point is  documented via  a literature survey and the second point is demonstrated
-% via an analysis of
-
-The conclusions of this paper are two fold.
-First, the phase delay effect does not hold in the large body of software projects analyzed.
-Second, commonly-held beliefs, such as phase delay, must be periodically revisited empirically in order for researchers to provide accurate and up-to-date guidance for the practitioner community.
-
-% the SE community needs to stop
-% justifying changes to SE
-% practices via the phase delay effect. Secondly,   myths like phase delay should not have lasted decades
-% in the literature without  critical reassessment.  As a community, we need to reflect more on our
-% research practices and on how (and when) we reflect and revisit the prominent hypotheses of our field.
-
-The core of this paper's arguments are two figures: Figure~\ref{fig:fix-phase-dist} and  \fig{raw}.
-Lest the reader find this too little  to justify our conclusions, we note that it took months
-of work to distill this data from our 171 projects, after which the five authors spent weeks
-exploring and debating the nuances. To the best of our knowledge, this is the largest study on phase delay yet published (and it contains data from orders of magnitude
-more projects than used in prior work~\cite{Fagan76,Boehm76,Daly77,Stephenson76,Shull02,Royce98,Clutterbuck09,Elssamadisy02}).
-To ensure reproducibility and unlike previous work,
-all the data  used in this study is available in the PROMISE
-repository (openscience.us/repo).
-
-\section{Motivation}
-\subsection{Why Reassess Existing Truisms?}
-
-Before exploring the commonly-held belief of the {\em phase delay effect}, we digress to discuss the merits of revisiting
-old conclusions in software engineering.
-
-Belief in general
-principles of software
-engineering are common to both research and practice. Professional societies assume such generalities exist when they offer
- lists of supposedly general ``best practices'' such as
-the IEEE 1012 standard for software verification~\cite{1012}.  Endres \& Rombach's comment that 
+\end{quote}
+ \begin{figure}[!b]  
+\begin{center}
+\includegraphics[angle=270,width=3in]{img/b81.pdf}
+ \end{center}
+ \caption{An example DIE effect. From Boehm'81~\cite{Boehm81}. }\label{fig:b81}
+ \end{figure}
+Other prominent authors have commented on its perceived usefulness as a rule of thumb for software engineers.
+McConnell mentions it as a "common observation" in the field and  summarizes the intuitive argument for why it should be so:
 \begin{quote}
-Based on repeated and consistent observations, key lessons of these fields can now be formulated into rules or even laws, providing initial building blocks towards a theoretical foundation that is essential for further research, for teaching and for the practice of software development~\cite{endres03,rombach11}
+''A small mistake in upstream work can affect large amounts of downstream work.
+A change to a single sentence in a requirements specification can imply changes in hundreds of lines of code spread across numerous classes or modules, dozens of test cases, and numerous pages of end-user documentation''~\cite{mcconnell01}.
 \end{quote}
- Endres \& Rombach offer dozens of lessons of software engineering~\cite{rombach11}.
- Many other 
-widely-cited researchers  do the same; e.g.
-Glass~\cite{glass02}; Jones~\cite{jones07}; Boehm~\cite{boehm00b}.
-Budgen \& Kitchenham seek to reorganize SE research using
-general
-conclusions drawn from a larger number of studies~\cite{kitch04,budgen09}.
-Nevertheless, there are many empirical  findings that 
-cast doubt on the existence of  ``key lessons  that can be formulated
-as general laws of SE'':
-\bi
-\item
-Turhan~\cite{me12d} lists 28 studies with contradictory conclusions
-on the relation of OO measures to defects.  Those results
- directly  contradict some of the laws listed by 
-Endres \& Rombach~\cite{endres03}.
-
-\item
-Ray et al.~\cite{ray2014lang} tested if   strongly typed languages
-predict for better code quality. In  728 projects,
-they found  only a modest benefit in strong typing
-(and caution that that effect may actually be due to other  conflating factors).
-\item
-Fenton \& Neil~\cite{fenton00,fenton00b}   critique the truism that
-``pre-release fault rates for software
-are a predictor for post-release failures'' (as claimed by~\cite{dunsmore88},
-amongst others). For the systems described in~\cite{fenton97}, they
-show that software modules that were highly fault-prone
-prior to release revealed very few faults after release.
-\item
-Meyer claims that   object-oriented (OO) encapsulation will
-reduce error rates in software~\cite{Meyer1988}.  Yet empirical results suggest
-that debugging an OO program is many times harder and
-longer than debugging a standard procedural program~\cite{hatton98}.
-\item A truism of visual programming is that ``visual
-representations are inherently superior to mere textual representations''. A review by Menzies suggests that the available
-evidence for this claim is hardly conclusive~\cite{me00v}. 
-\item Numerous recent {\em local learning} results compare single models
-learned from all available data to multiple models learned from clusters within the data~\cite{betten14,yang11,yang13,minku13,me12d,me11m,betta12,posnett11}.
-A repeated result in those studies in that the local models generated the better effort
-and defect predictions (better median results,
-lower variance in the predictions).
-% \item 
-% Other work studies  found dozens of single
-% models that performed worse that {\em committees of models}, all with % different opinions,
-% that contribute to an {\em ensemble} prediction of software 
-% development effort~\cite{kocaguneli2012value,azhar13}.
-\ei
-%To be fair to the software engineering community,
-Software engineering is not the only
-field where practitioners hang on to persistent beliefs, even if the evidence
-for those beliefs is not strong.
-The medical profession applies  many practices based on studies
-that have been disproved. For example,
-a  recent article
-in the Mayo Clinic Proceedings~\cite{prasad13} found  
-146 medical practices based on studies 
-in year $i$, but which were  reversed by subsequent trials within years $i+10$.
-Even when the evidence for or against a treatment or intervention is clear, medical providers and patients may not accept it~\cite{aschwanden10}.
-Aschwanden warns:
-\begin{quote}
-Cognitive biases - such as motivated reasoning (all of us want to believe that the things we do make a difference), base rate neglect (failing to pay attention to what happens in the absence of the intervention), and confirmation bias (the tendency to look for evidence that supports what you already know and to ignore the rest) - also influence how we process information~\cite{aschwanden15}.
-\end{quote}
-
-The cognitive issues that complicate medicine are also found in software engineering.
-According to Passos et al.~\cite{passos11}, commercial developers
-are all too willing to believe in general
-truisms.
-They  warn that developers
-usually assume that the truisms they learn from a few past
-projects are general to 
-all their future projects:
-\begin{quote}\label{q:pass}
-...the past experiences were taken into account without 
-much consideration for their context~\cite{passos11}.  
-\end{quote}
-The results of J{\o}rgensen \& Gruschke~\cite{jorgensen09} concur with Passos et al. They report that 
-  supposed software engineering    ``experts'' rarely use lessons
-  from past projects to improve their future reasoning~\cite{jorgensen09}. 
- They note that
-when the experts
-  fail to revise their beliefs, this leads to poor
- conclusions and software projects  (see examples in~\cite{jorgensen09}).
-%% This leads to  detrimental
-%% situations where groups or 
-%% individuals working on the same project have conflicting beliefs, which they never debate
-%or refine or unify.
-
-In summary, just like medicine, our field suffers when
- software engineers do  not revise old beliefs.  Therefore, it is important
- to regularly  reexamine    old beliefs.
-
-\subsection{Why Study Phase Delay?}
-\label{sec:why-study}
-%\todo{\begin{itemize}
-%\item Given the Intro and Motivation, this section should focus on the influence of phase delay in software engineering milieux
-%\item Need to tone down the reliance on Beck.
-The important Beck point is that eliminating the cost-to-fix curve (resulting from phase delay) is one of the main goals of agile.
-His observations on new technologies are good too.
-%\item Need to cite other popular sources of this information, e.g., McConnell's "Ounce of Prevention" article~\cite{mcconnell01}.
-%\item Robert Glass's~\cite{glass02} book is very popular and contains an entry on this point
-%\item Google Scholar Citations for phase delay works Boehm88~\cite{boehm88} (673), Boehm \& Basili~\cite{boehm01} (761), 
-%\end{itemize}}
-
-This paper examines evidence for the phase delay effect due to its prominence throughout the field of software engineering.
-Many prominent authors have commented on its perceived usefulness as a rule of thumb for software engineers.
-For example, in 2001, Steve McConnell mentioned it as a "common observation" in the field and proceeded to knock down arguments that it no longer held for contemporaneous software projects, by relying on recently published data and experience.
-McConnell also summarized the intuitive argument for why it should be so: ``A small mistake in upstream work can affect large amounts of downstream work.
-A change to a single sentence in a requirements specification can imply changes in hundreds of lines of code spread across numerous classes or modules, dozens of test cases, and numerous pages of end-user documentation''~\cite{mcconnell01} Glass echoes this sentiment, stating the the fact ``requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' is ``really just common sense''~\cite{glass02}.
-
-Also in 2001, the phase delay effect was listed as number one by Boehm and Basili in their "Top 10 list" of "objective and quantitative data, relationships,
+Glass also endorses this effect, asserting that``requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' is ``really just common sense''~\cite{glass02}.
+Other researchers
+are just as adamant in asserting that that the delayed issue effect is a proven
+fact.
+For example, what we call the delayed issued effect was listed as number one by Boehm and Basili in their "Top 10 list" of ``objective and quantitative data, relationships,
 and predictive models that help
@@
 and improve their ability to predict
-and control efficient software projects." \cite{boehm01}
-
-More important than its use as a rule of thumb, the phase delay effect has 
-been used for decades to motivate major changes to software engineering practices (as mentioned in Section 1). For example,
-in 2000, Kent Beck used phase delay to motivate his call for ``extreme programming'' (a set of techniques
-that evolved into contemporary agile and DevOps practices). Beck writes:
-\begin{quote}
-One of the universal assumptions of software engineering is that the cost of changing a program rises exponentially over time.
-I can remember sitting in a big linoleum-floored classroom as a college junior and seeing the professor draw on the board the curve found in \fig{curve1}...
-%
-%The cost to fix a problem in a piece of software rises exponentially over time. 
-A problem that might take a dollar to fix if you found it during requirements analysis might costs thousands to fix once the software is in production.
-%
-%I resolved then and there that I would never let a problem get through to production.
-No sirree, I was going to catch problems as soon as possible.
-I would work out every possible problem in advance.
-I would review and crosscheck my code.
-No way was I going to cost my employer \$100,000.~
-\cite{beck00}
-\end{quote}
-
-\begin{figure}
- \includegraphics[width=3.3in]{img/beckB4.png}
- \caption{Projects exhibiting phase delay. From~\cite{beck00}.}\label{fig:curve1}
- \end{figure}
-\begin{figure}
- \includegraphics[width=3.3in]{img/beckAFTER.png}
- \caption{Projects without   phase delay. From~\cite{beck00}.}\label{fig:curve2}
-\end{figure}
+and control efficient software projects''~\cite{boehm01}.
 
@@
 % \end{figure}
-
-Beck speculated that there might be another kind of project, shown in \fig{curve2},
-that does not exhibit phase delay. In fact, for Beck, the possibility that phase delay does not imply exponential increases in the resources required to make changes and fix defects is {\em the} major
-motivation for his work, because it means that decisions can be made as late in the process as possible, so as to increase the chances that these decisions would be right.
-Beck found more than adequate motivation for this new assumption:
-%\begin{quote}
-%This (\fig{curve2}) is one of the premises of (agile).
-It is the technical premise of (agile).
-If the cost of change rose slowly over time, you would act completely differently from how you do under the assumption that costs rise exponentially.
-You would make big decisions as late in the process as possible, to defer the cost of making the decisions and to have the greatest possible chance that they would be right.
-You would only implement what you had to, in hopes that the needs you anticipate for tomorrow wouldn't come true.
-You would introduce elements to the design only as they simplified existing code or made writing the next bit of code simpler.~\cite{beck00}
-%\end{quote}
-%Beck makes one other comment that is relevant to this paper: 
-\begin{quote}
-The software development community has spent enormous resources in recent decades trying to reduce the cost of change-better languages, better database technology, better programming practices, better environments and tools, new notations.
-
-What would we do if all that investment paid off?
-What if all that work on languages and databases and whatnot actually got somewhere?
-What if the cost of change didn't rise exponentially over time?~\cite{beck00}
-\end{quote}
-The results of this paper are that, in our sample of hundreds of projects, there is no evidence for phase delay.
-This result is   consistent with Beck being correct: all that work on    languages and databases
-have paid off and, hence, modern software projects do not suffer from the  phase delay effect.
-
-That said,  the phase delay myth persists. Hence, this paper.
+This paper calls into question all the above claims about the delayed
+issue effect. We suggest that the delayed issue effect might have been an dominant
+effect decades ago,  but {\em not}  for $21^{\mathit{st}}$
+software development. 
+ The delayed issue effect was first reported in 1981 in a era of punch card programming
+and non-interactive environments~\cite{Boehm81}. In the 21$^\mathit{st}$} century, we  program in 
+interactive environments with higher-level languages and better source code control
+tools. Such tools allow for the faster refactoring of existing
+code-- in which case, 
+managing the changes required to fix (say) an incorrect requirements assumption
+is far less    onerous    than before.
+
+Also, we note that our development practices have changed in ways that could   mitigate the delayed issued effect.  Previously,
+software was large monolithic systems that were write once and maintain
+forever. Today, there are more architectures that
+support  extensive and faster changes to software. 
+For example, consider the MEAN stack preferred for web development by
+the continuous deployment community. Older architectures for web development
+often used some variant of LAMP (Linux+ Appache+ Mysql+ Php) that was an intricate
+combination of tools written in different languages. MEAN stacks, on the other hand,
+use Javscript throughout-- which makes large scale reorganizations faster to complete~\cite{wayner15}.
+For another example where organizations are adopting increased development flexability, 
+Microsoft is adjusting its development practices towards a continuous
+release paradigm. Theisen et al. reports their experiences where Microsoft architects
+continually adjust  their modules in response to security issues~\cite{theisen15}.
+Also,
+upgrades to their Windows operating system is moving from service patches (which occur rarely) to continuous deployment (so there will be no Windows 11- just a stream of continuous updates to what is currently called Windows 10~\cite{bright15}).
+
+Other large development shops are similarly reorganizing their work-- witness the US Department of Defense's 2010 mandate
+that all new software acquisitions must adopt agile methods~\cite{kim13}. 
+This change in DoD policy arises from a separation of baseline architecture
+(e.g. the design of an 
+aircraft carrier) and the development of applications within that architecture.
+For the baseline architecture, bad decisions
+made  early in the life cycle may be too expensive to change.
+But at least at the DoD, the majority of software development occurs within
+the framework of some larger architecture (e.g. an aircraft carrier).
+These smaller projects can certainly lever the agile advantage, while 
+using fast refactoring tools that mitigate against  the delayed issue effect.
+
+In summary, there exists (1)~historical reasons for believing in the delayed issued
+effect as well as (2)~reasons to believe that that effect may not longer hold.
+Hence, the rest of this article explores the currency of the delayed issue effect.
+After some initial definitions for terminology and hypotheses, we discuss the
+value of checking old ideas. Next, we offer a  survey of industrial practitioners and researchers to show that this the negative impact of issue delay is the very prominent
+concern in thos communities (in fact, amongst industrial practitioners, it is their number one concern).
+Next, we  analyze 171 software  projects developed in the period 2006 to 2014 to find  {\em no evidence} of the delayed issue effect
+(note that, to ensure reproducibility,
+all the data  used in this study is available in the PROMISE
+repository at openscience.us/repo). Finally, we discuss the validity and implications of our results.
+
+\section{Definitions \& Hypotheses}
+This paper uses the following definitions:
+\bi
+\item
+{\em Delayed issue effect}:   it is {\em exponentially}  more {\em difficult} to resolve an issue in a software project, the {\em longer} it remains.
+\item
+ {\em Longer} time is defined as per  Boehm'81~\cite{Boehm81}; i.e. the gap between the   phases where   issues are introduced and resolved.
+\item
+A measure collected in phase ${1,.,i,..j}$ is 
+{\em exponential} when  that
+   measure at phase $j$   
+   is larger than the sum of that measure in 
+earlier phases $1 \le i < j$. 
+\item
+Issues are more {\em difficult}  
+when their resolution takes more time or costs more  (e.g. needs expensive
+debugging tools or the skills of expensive developers).
+\ei
+This paper explores the  following hypotheses.
+\subsubsection*{H1: DIE is a current and widely-held belief}
+
+Using a literature review, we can confirm that there are numerous historical papers (dating back decades)
+that endorse DIE. Also, using a survey conducted from this paper, we find  that 
+  DIE  appears up as the
+single most strongly-held beliefs amongst commercial software engineers.
+
+\subsubsection*{H2: DIE is a poorly documented effect}
+
+As discussed in our literature review,  many of the papers reporting the DIE
+effect are either:
+\bi
+\item
+Quite old (dating from last century);
+\item
+Just quote prior papers without presenting   new data; 
+\item
+Or cite data sources that can no longer be
+confirmed.
+\ei
+
+\subsubsection*{H3: During development, issues   exhibit a DIE effect}
+While this paper  endorse the above hypotheses, we will argue against this
+third hypothesis. In our sample of 
+ 171 software  projects developed, we will show
+{\em no evidence}  that, during development, issues were exponentially harder to resolve the longer they were left
+ in the software.
+
+% the SE community needs to stop
+% justifying changes to SE
+% practices via the phase delay effect. Secondly,   myths like phase delay should not have lasted decades
+% in the literature without  critical reassessment.  As a community, we need to reflect more on our
+% research practices and on how (and when) we reflect and revisit the prominent hypotheses of our field.
+
+\section{ But Why Reassess Old Truisms?}
+
+Before going any furhter,  we digress to discuss the merits of revisiting
+old conclusions in software engineering.
+
+Beliefs in general
+principles of software
+engineering are common to both research and practice. Professional societies assume such generalities exist when they offer
+ lists of supposedly general ``best practices'' such as
+the IEEE 1012 standard for software verification~\cite{1012}. 
+ Endres \& Rombach offer dozens of lessons of software engineering~\cite{endres03}.
+ Many other 
+widely-cited researchers  do the same; e.g.
+Glass~\cite{glass02}; Jones~\cite{jones07}; Boehm~\cite{boehm00b}.
+Budgen \& Kitchenham seek to reorganize SE research using
+general
+conclusions drawn from a larger number of studies~\cite{kitch04,budgen09}.
+
+That said, there are many empirical  findings that 
+raise doubts that   general laws of SE even exist:
+\be
+\item
+Turhan~\cite{me12d} lists 28 studies with contradictory conclusions
+on the relation of OO measures to defects.  Those results
+ directly  contradict some of the laws listed by 
+Endres \& Rombach~\cite{endres03}.
+\item
+Ray et al.~\cite{ray2014lang} tested if   strongly typed languages
+predict for better code quality. In  728 projects,
+they found  only a modest benefit in strong typing
+(and caution that that effect may actually be due to other  conflating factors).
+\item
+Fenton \& Neil~\cite{fenton00,fenton00b}   critique the truism that
+``pre-release fault rates for software
+are a predictor for post-release failures'' (as claimed by~\cite{dunsmore88},
+amongst others). For the systems described in~\cite{fenton97}, they
+show that software modules that were highly fault-prone
+prior to release revealed very few faults after release.
+\item
+Meyer claims that   object-oriented (OO) encapsulation will
+reduce error rates in software~\cite{Meyer1988}.  Yet empirical results suggest
+that debugging an OO program is many times harder and
+longer than debugging a standard procedural program~\cite{hatton98}.
+\item
+A truism of visual programming is that ``visual
+representations are inherently superior to mere textual representations''. A review by Menzies suggests that the available
+evidence for this claim is hardly conclusive~\cite{me00v}. 
+\item
+Numerous recent {\em local learning} results compare single models
+learned from all available data to multiple models learned from clusters within the data~\cite{betten14,yang11,yang13,minku13,me12d,me11m,betta12,posnett11}.
+A repeated result in those studies in that the local models generated the better effort
+and defect predictions (better median results,
+lower variance in the predictions).
+\ee
+% \item 
+% Other work studies  found dozens of single
+% models that performed worse that {\em committees of models}, all with % different opinions,
+% that contribute to an {\em ensemble} prediction of software 
+% development effort~\cite{kocaguneli2012value,azhar13}.
+To be fair, 
+SE is  not the only
+field where practitioners hang on to persistent beliefs, even if the evidence
+for those beliefs is not strong.
+The medical profession applies  many practices based on studies
+that have been disproved. For example,
+a  recent article
+in the Mayo Clinic Proceedings~\cite{prasad13} found  
+146 medical practices based on studies 
+in year $i$, but which were  reversed by subsequent trials within years $i+10$.
+Even when the evidence for or against a treatment or intervention is clear, medical providers and patients may not accept it~\cite{aschwanden10}.
+Aschwanden warns that ``congitive biases''  such as  confirmation bias (the tendency to look for evidence that supports what you already know and to ignore the rest)  influence how we process information~\cite{aschwanden15}.
+
+The cognitive issues that complicate medicine are also found in software engineering.
+According to Passos et al.~\cite{passos11}, commercial developers
+are all too willing to believe in general
+truisms.
+They  warn that developers
+usually assume that the truisms they learn from a few past
+projects are general to 
+all their future projects. They comment ``past experiences were taken into account without 
+much consideration for their context~\cite{passos11}.  
+The results of J{\o}rgensen \& Gruschke~\cite{jorgensen09} concur with Passos et al. They report that 
+  supposed software engineering    ``experts'' rarely use lessons
+  from past projects to improve their future reasoning~\cite{jorgensen09}. 
+ They note that
+when the experts
+  fail to revise their beliefs, this leads to poor
+ conclusions and software projects  (see examples in~\cite{jorgensen09}).
+%% This leads to  detrimental
+%% situations where groups or 
+%% individuals working on the same project have conflicting beliefs, which they never debate
+%or refine or unify.
+
+In summary, just like medicine, our field suffers when
+ software engineers do  not revise old beliefs.  Therefore, it is important
+ to regularly  reexamine    old beliefs such as the delayed issue effect.
+
+%\todo{\begin{itemize}
+%\item Given the Intro and Motivation, this section should focus on the influence of phase delay in software engineering milieux
+%\item Need to tone down the reliance on Beck.
+The important Beck point is that eliminating the cost-to-fix curve (resulting from phase delay) is one of the main goals of agile.
+His observations on new technologies are good too.
+%\item Need to cite other popular sources of this information, e.g., McConnell's "Ounce of Prevention" article~\cite{mcconnell01}.
+%\item Robert Glass's~\cite{glass02} book is very popular and contains an entry on this point
+%\item Google Scholar Citations for phase delay works Boehm88~\cite{boehm88} (673), Boehm \& Basili~\cite{boehm01} (761), 
+%\end{itemize}}
 
@@
 
-\subsection{Survey:  Phase Delay and Cost-to-fix Belief}\label{survey} 
-Yet another reason to study the phase delay effect is that it is a dominant belief in industrial SE.
-To test whether this was indeed true, we conducted a survey of software engineers. If our surveyed practitioners make management decisions based on their
+\section{H1: DIE is a current and widely-held belief}
+To assess the prevalence of the delayed issue effect,   conducted a survey of software engineers. If our surveyed practitioners make management decisions based on their
 understanding of SE theory, then the phase delay effect may well influence their judgment.
@@
 
-\section{Related Work}
+\section{H2: DIE is a poorly documented effect}
 
@@
 \section{Validity}
+
+\subsubsection*{H4: A post-development DIE effect can be predicted from pre-deployment data} 
+The data collected from our 171 projects     pertains only to
+development, and not to subsequent operations.
+So did we miss the DIE effect in our data since
+our data collection is just not prolonged enough to detect that effect?
+
+We would argue that this is not the case.
+An  exponential DIE effect can be seen {\em during the development} of all
+the papers we have reviewed. For example:
+\bi
+\item In \fig{b81}: see the data relating to development and acceptance testing;
+\item In  all the case studies of \fig{steck04} in the columns {\em Requirements} to {\em Test}.
+\ei
+That is, in the literature we have surveyed, the post-development DIE effect would have been predicted by data collected during pre-deployment development.
+Such
+
+Another surprise was a lack of {\em growth trend} suggesting that, in operation,
+ we should expect larger exponential delays in resolving issues. This is odd since all
+ the papers we could access on DIE should such a growth trend in the pre-operational data.
+
+\subsubsection*{H5: ``Time'' to resolve an issue is analogous to ``cost''}
+This hypothesis a somewhat technical point needed for our argument.
+While classical papers on the delayed issue effect discuss the {\em cost}
+of delaying issues, the data from our 171 projects only has time data.
+Hence, to say that our data addresses the classic delayed issue effect,
+we need to make the case that our time measurements are an appropriate
+surrogate for the prior claims on the cost of delaying issues.
+
 We report validity concerns following the guidelines of Runeson \& H\"{o}st~\cite{runeson09}.
@@
 % \end{figure}
+
+\begin{figure}
+{\small\begin{center} \begin{tabular}{r|rrr}
+& method1 & method2 & method3\\\hline
+requirements& 1& 1 & 1 \\
+design& 8 & 4 & 1\\
+build& 16 & 16 & 7\\
+test& 21 & 78 & 28\\
+operation& 29 & 186 & 1615
+\end{tabular}\end{center} }
+\caption{Results from the   Stecklein et al. study~\cite{steck04}.}
+\end{figure}
+
+Method1 linear ($R^2$ == 99\%).
+
+In the above definitions, 
+we  defined DIE in terms of ``difficulty'' to resolve  an issue as an umbrella term 
+including either ``more costly''
+or ``takes longer time''.  This hypothesis will be important for our analysis since (a)~the historical
+arguments for DIE (e.g.~\cite{Boehm81}) are expressed in terms of ``cost to resolve'' while (b)~the data available
+to this study is only available as ``time to resolve''. In the general case, it is not true that ``time'' predicts for ``cost''
+(measuring merely ``time'' ignores the hourly rate of the developers invloved or the cost of the hardware/software used for that resolution.

### 2015-08-02 data.tex
Updates from ShareLaTeX

@@
-Before presenting our analysis of phase delay in software projects, we define our terms, provide an overview of TSP and the sample projects, and describe the data collection process.
+The above analysis motivates a more detailed look at the delayed issued effect.  
+Accordingly, we examined 171 software projects conducted between 2006 and 2014.
+These projects took place at organizations in many countries and were conducted using  the Team Software Process (TSP$\textsuperscript{SM}$).
 
-\subsection{Definitions}
-A \emph{defect} is any change to a product, after its construction, that is necessary to make the product correct.
-A typographical error found in review is a defect.
-If that same defect is discovered while writing the code but before review, it is not considered to be a defect.
+Since 2000, the SEI has been teaching and coaching TSP teams.
+One of the authors (Nichols) has been been mentoring software development teams and coaches around the world as they deploy TSP within their organizations since 2006.
+The  most recent completions were in 2014.
+The projects were mostly small to medium, with a median duration of 46 days and a maximum duration of 90 days in major increments. 
+Several projects extended for multiple incremental development cycles. 
+Median team size was 7 people, with a maximum of 40. 
+Many of the projects were e-commerce web portals or banking systems in the US, South Africa, and Mexico. 
+There were  some  medical device projects in  the US, France, Japan, and Germany as well  as a commercial computer-aided design systems, and embedded systems.
 
-The \emph{cost-to-fix} can be computed in two ways.
-The first is a direct measure of the actual work time spent in correcting the defect in the product.
-Direct measurement enables analysis of the variability the cost-to-fix among a category of defects.
-The second approach measures the total time spent in a defect removal phase and divides by the total defects removed.
-This provides an average cost-to-fix including overhead of that activity.
-For example, using method two, a code inspection may require an hour of effort to remove three defects for a 3 defects per hour or to invert the ratio, 1/3 hours per defect.
-Most of the time, however, is spent locating the defects not actually fixing them.
-we use the second approach in our analysis.
+An anonymized version of that data is available in the PROMISE repository at openscience.us/repo.
+For confidentiality restrictions, we cannot offer 
+further details on these projects.
 
-\subsection{Overview of the Team Software Process}
-To empirically test for the phase delay effect, we examined 171 software projects conducted between 2005 and 2014.
-These projects took place at organizations in many countries and were conducted using  the Team Software Process (TSP$\textsuperscript{SM}$), a software project management approach developed at the Software Engineering Institute (SEI) at Carnegie Mellon University~\cite{tsp00}.
+\subsection{About TSP$\textsuperscript{SM}$}
+
+TSP is a software project management approach developed at the Software Engineering Institute (SEI) at Carnegie Mellon University~\cite{tsp00}.
 TSP is an extension of the Personal Software Process (PSP$\textsuperscript{SM}$) developed at the SEI by Watts Humphrey ~\cite{tsp00}.
@@
 
-TSP has guiding principles; the following are relevant to this study:
-\begin{itemize}
-\item  Sound engineering work does not happen by accident. It must be planned.
-\item  Performance of the plans requires realistic commitments of the resources needed to perform the work.
-\item  The planned work should frequently be compared to  results to assure plans remain relevant.
-\item  People performing the work should be responsible for planning and tracking the work.
-\end{itemize}
-
-TSP helps developers by providing a set of measures that can be applied to managing tasks, quality, and schedule.
+%TSP helps developers via a set of measures that can be applied to managing tasks, quality, and schedule.
 Planning begins by quantifying goals, defining work practices, and estimating size and effort.
@@
 
-{\em Planning} follows the technique taught in the PSP$\textsuperscript{SM}$.
-Developers estimate the size of work products and convert this to a total effort using historical rates.
+During {\em Planning}, developers estimate the size of work products and convert this to a total effort using historical rates.
 Time in specific tasks come from the  process phases and historical percent time in phase distributions.
@@
 
-\begin{figure}[!t]  
+\subsection{Data Integrity}
+
+A common property of real-world data sets is the presence
+of noisy entries (superfluous  or spurious data). 
+The level of noise can be quite high. As reported
+in \cite{shepperd12}, around
+10\% to 30\%
+of the records in the NASA MDP defect data sets are
+affected by noise.
+
+One reason to use the SEI data for the analysis of this paper is its remarkably low level of noise.
+Nichols et al.~\cite{shirai14}  report that
+the noise levels in the SEI TSP data are smaller than those seen
+in other data sets. They found in the SEI TSP data that:\bi 
+\item
+4\% of the data was incorrect (e.g. nulls, illegal formats);
+\item  2\% of the data has inconsistencies such as timestamps
+where the stop time was before the start time;
+\item 3\% of the data had data that was not credible
+such as tasks listed in one day that took like than six hours.
+\ei 
+One explanation for this low level of noise is the TSP process.
+One the guiding principles of TSP was that  people performing the work are  responsible for planning and tracking the work. That is,  all the data collected here was entered
+by local developers. This data was then check by local {\em coaches} before being sent to the SEI
+databases. Coaches are certified by demonstrating competent use of the TSP process with the artifacts and data.
+The use of certified local coaches within each project increases the integrity of our data.
+
+Overall, the   effort associated with adding TSP to a project is not onerous. McHale reports~\cite{mchale02}:
+\bi
+\item
+ The time spent  tracking time, defects, and tasks requires less then 3\% of a developer's time. Weekly team meetings  require at most an hour, which is
+only 2.5\% of a 40 hour work week. 
+\item
+Team launches and replans average about 1 day per month or 5\% planning overhead.
+\ei
+It is true that one staff member is needed as a ``couch'' to mentor the teams
+and certify and monitor that data collection.
+However, one of us (Nichols) has worked with dozens of TSP teams.
+He reports that one  trained coach can support 4 or 6 teams (depending upon team experience).
+
+\subsection{Data Details}
+\label{sec:data-collection}
+
+%Organizations using TSP agree to provide their project data to the SEI for use in research.
+In return the SEI agrees    that  data must not be traceable to its source.
+The data is collected at major project events; launch, interim checkpoints, and at project completion.
+Data includes project and site  characteristic summaries, surveys of team members, launch presentations, launch outbriefs, and baseline plans, final data from the project, and the project post mortem report.
+In practice, this data requirement has only been enforced for the purposes of certifying and reauthorizing TSP coaches who must submit data to maintain their authorization.
+Coaches are certified by demonstrating competent use of the TSP process with the artifacts and data not by the actual project results.
+Of the data submitted, only the data recorded using the Process Dashboard tool has been collected and aggregated for this research.
+ %The key views include a project results summary, effort logs, task logs, and defect logs.
+The data has not yet undergone additional screening.
+A summary of the data quality issues was reported in a previous work ~\cite{shirai14} (that summary is discussed further in our {\em Validity} section, below).
+
+Using tools provided by the SEI, developers kept very detailed logs of their daily activity. Our data includes  work start time, work end time,  delta
+work time, and interruption time. Software engineers are often
+interrupted by meetings, requests for technical help, reporting, and
+so forth. These events are recorded, in minutes, as interruption
+time.  
+In this paper, when we report ``time to resolve an
+issue,'' we show the difference between the start and end times
+of a work session, with any interruption time subtracted (the
+difference in times, minus the interruptions).
+
+As of November 2014, the SEI TSP database contained data from 212
+TSP projects. The projects completed between July 2006 and
+November 2014; they included 47 organizations and 843 people. 
+The database fact tables
+contain 268,726 time logs, 
+154,238 task logs,
+ 47,376 defect logs, 
+and 26,534 size logs. 
+After selecting defects from the data log and joining the data to the time log table  171 of these projects remained (the excluded projects had no or too few defects to use in this analysis).
+
+\begin{figure}[!b]  
 \begin{center}
@@
 
-\subsection{Projects in the Sample}
-Since shortly before the year 2000, the SEI has been teaching and coaching TSP teams.
-One of the authors (Nichols) has been been mentoring software development teams and coaches around the world as they deploy TSP within their organizations since 2006.
-The  most recent completions were in 2014.
-The projects were mostly small to medium, with a median duration of 46 days and a maximum duration of 90 days in major increments. 
-Several projects extended for multiple incremental development cycles. 
-Median team size was 7 people, with a maximum of 40.
-
-Many of the projects were e-commerce web portals or banking systems in the US, South Africa, and Mexico. 
-There were  some  medical device projects in  the US, France, Japan, and Germany as well  as a commercial computer-aided design systems, and embedded systems.
-
-An anonymized version of that data is available in the PROMISE repository\footnote{http://openscience.us/repo}  and,
-for confidentiality restrictions, we cannot offer 
-further details on these projects.
-
-\subsection{Data Collection Process}
-\label{sec:data-collection}
-
-Organizations using TSP agree to provide their project data to the SEI for use in research.
-In return the SEI agrees    that  data must not be traceable to its source.
-The data is collected at major project events; launch, interim checkpoints, and at project completion.
-Data includes project and site  characteristic summaries, surveys of team members, launch presentations, launch outbriefs, and baseline plans, final data from the project, and the project post mortem report.
-In practice, this data requirement has only been enforced for the purposes of certifying and reauthorizing TSP coaches who must submit data to maintain their authorization.
-Coaches are certified by demonstrating competent use of the TSP process with the artifacts and data not by the actual project results.
-Of the data submitted, only the data recorded using the Process Dashboard tool has been collected and aggregated for this research.
-
-To use this data for research and benchmarking,  Nichols first collected the Process Dashboard data from the submission repository then used a tool built by the Process Dashboard developer, David Tuma,
-to gather the project data into a database.
-Nichols and his colleague Yasutaka Shirai then extracted data into views suitable for analysis.
-The key views include a project results summary, effort logs, task logs, and defect logs.
-The data has not yet undergone additional screening.
-A summary of the data quality issues was reported in a previous work ~\cite{shirai14} (that summary is discussed further in our {\em Validity} section, below).
-
-The tool collected process information about those tasks and defects that we analyze in this work. 
-That task data include  work start time, work end time, delta
-work time, and interruption time. Software engineers are often
-interrupted by meetings, requests for technical help, reporting, and
-so forth. These events are recorded, in minutes, as interruption
-time.
-
- Defect logs include the time and date a defect was discovered, the phase in which that defect was injected, the phase in which it was removed, the time (in minutes) required to find and fix the defect, and the categorical type.
-
-The defect types in the most common defect type standard  in the the SEI TSP data are follows:
+In these logs, a  \emph{defect} is any change to a product, after its construction, that is necessary to make the product correct.
+A typographical error found in review is a defect.
+If that same defect is discovered while writing the code but before review, it is not considered to be a defect.
+SEI TSP defect types are:
 \begin{itemize}
-\item Environment: design, compile, test, or other support system problems
+\item Environment: design, compile, test,  other support  problems
 \item Interface: procedure calls and reference, I/O, user format
@@
 \item Syntax: spelling, punctuation typos, instruction formats
-\item Function: logic, pointers, loops, recursion, computation, function defects  
+\item Function: logic, pointers, loops, recursion, computation  
 \item Checking: error messages, inadequate checks
@@
 \item Assignment: package
-declaration, duplicate names, scope, limits
+declaration, duplicate names, scope
 \item System: configuration, timing, memory
 \end{itemize}
+The defect logs in that data  include the time and date a defect was discovered, the phase in which that defect was injected, the phase in which it was removed, the time (in minutes) required to find and fix the defect, and the categorical type.
 
-As of November 2014, the SEI TSP database contained data from 212
-TSP projects. The projects completed between July 2006 and
-November 2014; they included 47 organizations and 843 people. 
-The database fact tables
-contain 268,726 time logs, 
-154,238 task logs,
- 47,376 defect logs, 
-and 26,534 size logs. 
-After selecting defects from the data log and joining the data to the time log table  171 of these projects remained. The excluded projects had no or too few defects to use in this analysis.
+The phases logged by our data are shown in \fig{waterfall}.
+ Although the representation suggests a waterfall model, the SEI experience is that all real implementations of any size follow a spiral approach with many team performing the work in iterative and/or incremental development cycles.
+%Within a project increment, multiple features or components may be developed and incremented.
+TSP is compatible %with agile and encourages iterative and incremental development.
+Nonetheless, the specific strategy and cycle %duration is a project decision.
+TSP does, however, strongly encourage 1) constructing units of sufficient size that %%measurement is practicable, and 2) separating the construction from appraisal and test activities.
+This %effectively highlights the separation of construction from rework activities and aids the apportionment of defects %to those found in appraisal activities (reviews and inspections) and those found through failure (test).
+The %distinct construction activities (requirements, high and detailed design, and code) were chosen to help teams %analyze the effectiveness and efficiency of their practices through analysis of the defect phase origin, type, fix %effort, and phase of discovery.
 
-In this paper, when we report ``time to resolve an
-issue,'' we show the difference between the start and end times
-of a work session, with any interruption time subtracted (the
-difference in times, minus the interruptions).
-
-%That said, certain semantic features of the SEI TSP data should be noted.
-%Firstly, in the current TSP collection tool, 
-%fix times are only the developer time for the developer walking through the phases of \fig{waterfall}.
-%We are currently tracking the fix time for post-release issues (e.g. those raised during  acceptance test and %later
-%product life cycle). So far, in that post-release data,  we have not detected
-%a dramatic phase escalation effect (but at this time, we have nothing definitive comment on that matter).%
-%
-%\bill{somewhere you have one note on \underline{find} and fix times.  for this paper, we need just fix times. %but is there
-%anything we need to fret about re \underline{find} times?}
-
-\subsection{Project Phases}
-This paper studies the impact of phase delay on the time required to resolve issues.
-The logical phases used in this paper are shown in \fig{waterfall}.
-Although the representation suggests a waterfall, the phases represent the primary stages through which requirements are developed into working and tested code.
-That is, all requirements must be stated in some way, implemented in code, integrated, and tested.
-All real implementations of any size follow a spiral approach with many team performing the work in iterative and/or incremental development cycles.
-Within a project increment, multiple features or components may be developed and incremented.
-TSP is compatible with agile and encourages iterative and incremental development.
-Nonetheless, the specific strategy and cycle duration is a project decision.
-TSP does, however, strongly encourage 1) constructing units of sufficient size that measurement is practicable, and 2) separating the construction from appraisal and test activities.
-This effectively highlights the separation of construction from rework activities and aids the apportionment of defects to those found in appraisal activities (reviews and inspections) and those found through failure (test).
-The distinct construction activities (requirements, high and detailed design, and code) were chosen to help teams analyze the effectiveness and efficiency of their practices through analysis of the defect phase origin, type, fix effort, and phase of discovery.
-
-Note that, in that Figure~\ref{fig:waterfall}:
-\bi 
-\item
-Several  phases in which product is created have sub-phases of {\em review} and {\em inspect} to remove defects.
-TSP uses review, a sub-phase in which individuals perform personal reviews of their work products prior to the peer review, which TSP calls the inspection.
-\item Testing is divided into several stages.
+Note that, in that Figure~\ref{fig:waterfall},
+several  phases in which product is created have sub-phases of {\em review} and {\em inspect} to remove defects.
+In TSP  reviews,  individuals perform personal reviews of their work products prior to the peer review
+(which TSP calls the inspection(.
+Also, Figure~\ref{fig:waterfall} divides  testing     as follows.
 Developers perform unit test prior to code complete.
 After code complete a standard phase is the integration, which combines program units into workable system ready for system test.
-Integration,  system test, and acceptance test are often performed by a separate group.
-\ei
-For these TSP projects, the principles followed are: 
-\bi 
+Integration,  system test, and acceptance test are often performed by another group.
+
+%For these TSP projects, the principles followed are: 
+%%\bi 
 %\item Defects should be found before test, that is, test defects should be infrequent.
-\item A product should be inspected before used in a subsequent construction phase to minimize  unnecessary rework.
-\item To minimize unnecessary rework, a product should be reviewed and inspected prior to use in a subsequent construction phase.
-\item A personal review should remove the most common defects before being given to peers for the inspection.
-\item Because different categories of test will find different types of defect, and follow in a natural sequence, the tests should be measured separately
-\ei 
-In summary, the TSP process and measurement have been specifically designed to aid the analysis of defects for use in process improvement.
+%%\item A product should be inspected before used in a subsequent construction phase to minimize  unnecessary rework.
+%\item To minimize unnecessary rework, a product should be reviewed and inspected prior to use in a subsequent construction phase.
+%\item A personal review should remove the most common defects before being given to peers for the inspection.
+%\item Because different categories of test will find different types of defect, and follow in a natural sequence, the tests should be measured separately
+%\ei 
+%In summary, the TSP process and measurement have been specifically designed to aid the analysis of defects for use in process improvement.
 The data are, therefore, uniquely suitable for an analysis of the phase delay effect.
-\subsection{Distinguishing Project Characteristics}
-Modern development processes include the core engineering practices of estimation, planning, configuration control, requirements, design, code development, and test.
+%\subsection{Distinguishing Project Characteristics}
+%Modern development processes include the core engineering practices of estimation, planning, configuration control, requirements, design, code development, and test.
 Agile methods address  not only these core practices but also address team organization, individual and team commitments, test and build automation, and implementing iterative and incremental development.

### 2015-08-02 lucas.tex
Updates from ShareLaTeX

@@
 Our survey collected data on software engineers' views of commonly held software engineering ``laws''.
-One of the laws questioned relates directly to the problems associated with
-the delayed issue effect: ``Requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' (from Glass~\cite{glass02} p.71 who references Boehm \& Basili~\cite{boehm01}).
+One of the laws questioned is the a specific form of our delayed issue effect:  ``requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' (from Glass~\cite{glass02} p.71 who references Boehm \& Basili~\cite{boehm01}).
 We abbreviate this law as RqtsErr
@@
 illustrated in \fig{b81}).
+
+(Technical aside: we    ask them  RqtsErr  since this issue typically needs no supportive explanatory
+text. On the other hand, if we had asked them about our more general term ``delayed issue
+effect'', we would have had to burden our respondents with extra explanations).
 
@@
 \begin{tabular}{l|c|c|c|c|c}
- &  & \multicolumn{2}{c}{agreement} & \multicolumn{2}{c}{applicability} \\\cline{1-1} 
-Practitioner survey  & N & $\bar{x}$ & $\tilde{x}$ & $\bar{x}$ & $\tilde{x}$ \\
+ &  & \multicolumn{2}{c|}{agreement} & \multicolumn{2}{c}{applicability} \\\cline{1-1} 
+Practitioner survey  & N & w & $\bar{x}$ & $\tilde{x}$ \\
 \hline 
@@
 Overall, the RqtsErr law was the most agreed upon and most applicable law of 11 surveyed amongst practitioners, and the second most agreed upon law amongst researchers. 
-The survey results strongly support that the notion that the  delayed issue effect   is a widely-held belief in software engineering.
+This results strongly support that the notion that the  delayed issue effect   is a widely-held belief in software engineering.

### 2015-08-02 lucasRelatedWork.tex
Updates from ShareLaTeX

@@
 These studies suggest that the difficulty (in terms of effort) to find and fix an error monotonically increases with lifecycle phase.
+In 1990, Boehm~\cite{Boehm80} provides data suggesting that the cost-to-fix curve for small projects (from two student projects of 2000 deliverable source instructions) is flatter than for large projects (the dashed line of Figure~\ref{fig:cost-to-fix}).
 
@@
 
-In the 40 years since these initial studies, few studies have been published on the difficulty to resolve issues
+In the 40 years since these initial studies, few studies have explored the difficulty to resolve issues
 as a function of lifecycle phase. 
-Boehm~\cite{Boehm80} provides data suggesting that the cost-to-fix curve for small projects (from two student projects of 2000 deliverable source instructions) is flatter than for large projects (the dashed line of Figure~\ref{fig:cost-to-fix}).
-But note that even in this figure
-that even for the smaller projects, there is a four-fold increase (from 5 to 20) in the difficulty
-associated with handling issues introduced earlier in the lifecycle.
-
 Shull et al.~\cite{Shull02} conducted a literature survey and held a series of e-workshops with industry experts on fighting defects.
@@
 the data used to make prior conclusions.
-As an example this, \fig{steck} shows one 2004 survey that reports an exponential delayed issue  effect for 
+As an example of this, \fig{steck} shows one 2004 survey that reports an exponential delayed issue  effect for 
 requirements issues in eight case studies. The right-hand-side of that column shows which of those referenced studies
@@
 
-\begin{figure}
+\begin{figure}[!b]
  \includegraphics[width=3in]{img/clutterbuck.png} 
  \caption{Cost of change from an agile case study. From~\cite{Clutterbuck09}.}\label{fig:clutterbuck}
- \end{figure}
-
-\begin{figure}
- \includegraphics[width=3.3in]{img/Royce98.png}
- \caption{Exception to the rule - CCPDS-R case study cost-to-fix curve. From~\cite{Royce98}.}\label{fig:royce}
  \end{figure}
@@
 
+\begin{figure}
+ \includegraphics[width=3.3in]{img/Royce98.png}
+ \caption{Exception to the rule - CCPDS-R case study cost-to-fix curve. From~\cite{Royce98}.}\label{fig:royce}
+ \end{figure}
+
 Another example of less-than exponential explosion in the difficulty associated with delayed issues comes from the  CCPDS-R

### 2015-08-02 phaseDelay.tex
Updates from ShareLaTeX

@@
  The delayed issue effect was first reported in 1981 in a era of punch card programming
-and non-interactive environments~\cite{Boehm81}. In the 21$^\mathit{st}$} century, we  program in 
+and non-interactive environments~\cite{Boehm81}. In the 21$^\mathit{st}$ century, we  program in 
 interactive environments with higher-level languages and better source code control
@@
 support  extensive and faster changes to software. 
-For example, consider the MEAN stack preferred for web development by
-the continuous deployment community. Older architectures for web development
-often used some variant of LAMP (Linux+ Appache+ Mysql+ Php) that was an intricate
-combination of tools written in different languages. MEAN stacks, on the other hand,
-use Javscript throughout-- which makes large scale reorganizations faster to complete~\cite{wayner15}.
-For another example where organizations are adopting increased development flexability, 
+For example, 
 Microsoft is adjusting its development practices towards a continuous
@@
 continually adjust  their modules in response to security issues~\cite{theisen15}.
-Also,
-upgrades to their Windows operating system is moving from service patches (which occur rarely) to continuous deployment (so there will be no Windows 11- just a stream of continuous updates to what is currently called Windows 10~\cite{bright15}).
-
-Other large development shops are similarly reorganizing their work-- witness the US Department of Defense's 2010 mandate
+Also, upgrades to their Windows operating system is moving from service patches (which occur rarely) to continuous deployment (so there will be no Windows 11- just a stream of continuous updates to what is currently called Windows 10~\cite{bright15}).
+
+For another example, consider the MEAN stack preferred for web development by
+the continuous deployment community. Older architectures for web development
+often used some variant of LAMP (Linux + Appache + Mysql + PHP) that was an intricate
+combination of tools written in different languages. MEAN stacks, on the other hand,
+use Javascript throughout-- which makes large scale reorganizations faster to complete~\cite{wayner15}.
+
+Inspired by the success of agile approaches, other  development organizations are similarly reorganizing their work-- witness the US Department of Defense's 2010 mandate
 that all new software acquisitions must adopt agile methods~\cite{kim13}. 
@@
 effect as well as (2)~reasons to believe that that effect may not longer hold.
-Hence, the rest of this article explores the currency of the delayed issue effect.
-After some initial definitions for terminology and hypotheses, we discuss the
-value of checking old ideas. Next, we offer a  survey of industrial practitioners and researchers to show that this the negative impact of issue delay is the very prominent
-concern in thos communities (in fact, amongst industrial practitioners, it is their number one concern).
+Hence,  this article explores the currency of the delayed issue effect.
+After some initial definitions, we discuss the
+value of checking old ideas.
+Next, survey of industrial practitioners and researchers to document a widespread belief that delayed issues have a negative impact on projects.
 Next, we  analyze 171 software  projects developed in the period 2006 to 2014 to find  {\em no evidence} of the delayed issue effect
@@
 \ei
-This paper explores the  following hypotheses.
-\subsubsection*{H1: DIE is a current and widely-held belief}
-
+Note that we use  the  term ``delayed issue effect'' as generalization of the
+more specific rule  ``requirements errors are hardest to fix''.
+This generalization is valid since the rationale for the rule about requirements
+is usually done as per McConnell~\cite{mcconnell01}; i.e. small upstream mistakes very
+early in the system can cascade into huge problems later in the lifecycle.
+That said, we prefer our more general term ``delayed issue effect'' since not
+only might requirements errors cascade, so too might analysis errors, design errors, etc.
+
+One technical aside: there are several ways to  measure {\em difficulty}.
+The first is a direct measure of the actual work time spent in correcting the defect in the product.
+Direct measurement enables analysis of the variability the cost-to-fix among a category of defects.
+The second approach measures the total time spent in a defect removal phase and divides by the total defects removed.
+This provides an average cost-to-fix including overhead of that activity.
+For example, a code inspection may require an hour of effort to remove three defects for a 3 defects per hour or to invert the ratio, 1/3 hours per defect.
+One observation made by Nichols while studying our 171 projects is that most of the time  is spent locating the defects not actually fixing them.
+we use the second approach in our analysis.
+
+This paper defects the  following claims. Note that we call them
+``claims'' not ``hypotheses'' since the later require defense via some statistical
+significance test. On the other hand, our claims will be supported via a variety
+of arguments, presented later in this paper.
+
+{\bf  Claim1: ``DIE'' is a  widely-held belief.}
 Using a literature review, we can confirm that there are numerous historical papers (dating back decades)
@@
 
-\subsubsection*{H2: DIE is a poorly documented effect}
-
+{\bf  Claim2: ``DIE'' is a poorly documented.}
  As discussed in our literature review,  many of the papers reporting the DIE
-effect are either:
-\bi
-\item
-Quite old (dating from last century);
-\item
-Just quote prior papers without presenting   new data; 
-\item
-Or cite data sources that can no longer be
+effect are either (1)~quite old (papers dating from last century);
+(2)~that just quote prior papers without presenting   new data; 
+(3)~or cite data sources that can no longer be
 confirmed.
-\ei
-
-\subsubsection*{H3: During development, issues   exhibit a DIE effect}
-While this paper  endorse the above hypotheses, we will argue against this
-third hypothesis. In our sample of 
+
+{\bf Claim3: Delayed Issues not Harder to Resolve.}
+ In our sample of 
  171 software  projects developed, we will show
-{\em no evidence}  that, during development, issues were exponentially harder to resolve the longer they were left
+{\em no evidence}  that, during development, delayed issues were exponentially harder to resolve the longer they were left
  in the software.
@@
 
-Before going any furhter,  we digress to discuss the merits of revisiting
+Before going any further,  we digress to discuss the merits of revisiting
 old conclusions in software engineering.
@@
 
-\section{H1: DIE is a current and widely-held belief}
+\section{``DIE'' is a  widely-held belief}
 To assess the prevalence of the delayed issue effect,   conducted a survey of software engineers. If our surveyed practitioners make management decisions based on their
@@
 
-\section{H2: DIE is a poorly documented effect}
+\section{``DIE'' is Poorly  Documented}
 
@@
 
-\section{Data Collection}
+\section{Delayed Issues not Harder\newline to Resolve}
 
@@
 
-\section{Analysis of Phase Delay in 171 TSP Projects}
-Our units of analysis are:
+Using the above,
+our units of analysis are:
 \bi
@@
 
-\begin{figure}[!t]
+\begin{figure}[!b]
 
@@
 
-\subsection{TSP Project Cost-to-fix curve}
-This section discusses our first hypothesis ($H_1$) that   that  the cost-to-resolve issues dramatically increases across phases.
+\subsection{Observations from 171 Projects}
 
@@
 
-The first thing to note in Figure~\ref{fig:fix-phase-dist}, 
-is that this data does not follow the exponential increase feared by Beck's \fig{curve1}:
+\fig{raw} shows the 25th, 50th, 75th percentile
+of the time spent on {\em review} or {\em inspect} or {\em test} phases
+in our 171 projects.
+Note that, in TSP, when developers see issues, they enter {\em review} or 
+{\em inspect} or {\em test}
+until that issue is retired.
+
+In that figure, the data is split out according to issues that were fixed in phase $Y$ after
+being introduced in phase $X$. The data is sub-divided into tables according to $X$;
+i.e. according to {\em before, planning, requirements, design} or {\em  code}. 
+The stars on the right-hand-side of that figure show the scale ups associated with the phase removal stages,
+expressed as a ratio of the first item in that block (which, by definition, is 1.00 for
+that first item).
+If  this data exhibited a delayed issue effect,  we would expect those bars in each sub-division to exhibit an
+exponential shape.
+
+Using Figure~\ref{fig:fix-phase-dist}, we can comment on two specific and more general
+delayed issued effects. 
+\bi
+\item {\bf Specific1}: Requirements errors are the most expensive to fix;
+\item {\bf Specific2}: Requirements errors are the most expensive to fix, 
+ when found during production but the cheapest to fix early in development;
+ \item {\bf General:}  It is exponentially more  difficult  to resolve an issue in a software project, the   longer it remains.
+\ei
+\subsubsection{Specific1}
+
+Our data does not support   
+{\em  requirements errors are the most expensive to fix}.
+Note that   median time spent fixing errors from {\em requirements} across the life cycle was 92, 93,  49, 33, 39, 49, 67, 122 minutes\footnote{If this time seems long, then recall
+that it includes the time required  to (a)~collect data and realize there is an error;
+(b)~prepare a fix;  and (c)~apply some validation
+procedure to check the fix (e.g. discuss it with a colleague or execute some tests).}.
+This is:
+\bi
+\item
+Clearly longer than median times seen {\em Before } and in {\em Planning}
+(30,28,45,77,65,56,24,35,33);
+\item
+Somewhat shorter, but still not exponentially smaller,  than the median times see after {\em requirements} in {\em design}
+(103,101, 79,66, 119,88, 80,45) and in {\em code} (65,115, 113,125, 95,133). 
+\ei
+For completeness, we ran an  
+A12 test comparing the raw data (and not just the medians of \fig{raw}) and found
+ a non-small effect only for the difference {\em requirements}
+to {\em code}. But even though the effect was non-small, 
+ the overall size of the difference
+{\em requirements} to {\em code} was less than 30\% (i.e. much less than we would
+expect from the delayed issue effect). 
+\subsubsection{Specific2}
+Next, we comment on 
+requirements errors are the most expensive to fix 
+{\em when found during production but the cheapest to fix early in development}}. 
+Our data from these 171 project does not extend into post-delivery deployment. That said,
+there is enough evidence to raise doubts that our projects will exhibit exponentially more
+difficult issues after deployment. 
+Specifically: prior reports of an  exponential increase in difficulty in resolving issues {\em during
+deployment} might have been used to predict for exponentially increasing
+difficulty {\em post-deployment}.
+For example:
+\bi
+\item  From requirements to acceptance testing
+\fig{b81} reports a 40-fold increase;
+\item
+ \fig{cost-to-fix} reports a 100-fold increase
+(for the larger projects).
+\item
+The projects surveyed in \fig{steck} report changes of a  \newline
+\{37,40,40,50,51,70,75,250\}-fold increase in pre-deployment period when the software was being developed.
+\ei
+On the other hand, a remarkable feature of Figure~\ref{fig:fix-phase-dist} is the {\em absence}
+of an exponential delay effect in the difficulty to resolve requirements issues during development.
+The greatest difficulty in resolving requirements issues was 1.31 seen in systems testing--
+which is far smaller than the 37 to 250-fold increases reported in \fig{b81}, \fig{cost-to-fix},
+and \fig{steck}.  That is, from our TSP data, we would not anticipate exponentially greater difficulty in resolving requirements issues after
+deployment. 
+ \subsubsection{General}
+ The above specific statements related just to requirements. Turning now
+ to issues creased in all phases in Figure~\ref{fig:fix-phase-dist},  we see that
+ this data does not follow the exponential increase predicted
+by the delayed issue effect:
 \bi
@@
 In either case,
-it does not look like \fig{curve1}.
+it does not look like exponential.
 \item The time in final testing is particularly low suggesting that few defects survived into testing
@@
 
-Given the above evidence, we reject $H_1$: there is no clear evidence that  cost-to-fix of defects dramatically increases across phases.
-
-\subsection{Phase delay analysis}
-
- \subsubsection{$H_2$: Cost-to-resolve Increases with Delay}
-
 %we calculate phase delay as simply the number of project phases (the set used in our data set is shown in %\fig{waterfall}) during which the defect existed in the system, i.e.
@@
 %
-\fig{raw} shows the 25th, 50th, 75th percentile
-of the time spent on {\em review} or {\em inspect} or {\em test} phases
-in our 171 projects.
-Note that, in TSP, when developers see issues, they enter {\em review} or 
-{\em inspect} or {\em test}
-until that issue is retired.
-
-In that figure, the data is split out according to issues that were fixed in phase $Y$ after
-being introduced in phase $X$. The data is sub-divided into tables according to $X$;
-i.e. according to {\em before, planning, requirements, design} or {\em  code}.
-
-The stars on the right-hand-side of that figure show the scale ups associated with the phase removal stages,
-expressed as a ratio of the first item in that block (which, by definition, is 1.00 for
-that first item).
-If $H_2$ was correct, we would expect those bars in each sub-division to exhibit a  shape
-like \fig{curve1}.
-
-They do not. Hence, we reject $H_2$.
-
- \subsubsection{$H_3$: Requirements Repairs are Most Expensive}
-
-As to $H_3$, the {\em requirements} the  median time spent fixing errors in {\em requirements} was 92,93  49,33,39,49,67,122 minutes\footnote{If this time seems long, then recall
-that it includes the time required  to (a)~collect data and realize there is an error;
-(b)~prepare a fix;  and (c)~apply some validation
-procedure to check the fix (e.g. discuss it with a colleague or execute some tests).}.
-This is:
-\bi
-\item
-Clearly longer than median times seen {\em Before } and in {\em Planning}
-(30,28,45,77,65,56,24,35,33);
-\item
-Somewhat shorter, but still not exponentially smaller,  than the median times see after {\em requirements} in {\em design}
-(103,101, 79,66, 119,88, 80,45) and in {\em code} (65,115, 113,125, 95,133). 
-\ei
-For completeness, we ran an  
-A12 test comparing the raw data (and not just the medians of \fig{raw}) and found
- a non-small effect only for the difference {\em requirements}
-to {\em code}. But even though the effect was non-small, 
- the overall size of the difference
-{\em requirements} to {\em code} was less than 30\% (i.e. much less than we would
-expect from \fig{curve1}.
-
-Based on the above, we reject   $H_3$.
 
@@
 
+To sum up the paper so far,
+\bi
+\item
+The delayed issue effect has been reported in around a dozen papers, at least half of which date back to the 1970s. 
+\item
+But there exists
+no evidence for this effect was seen  in hundreds of software projects build   since 2006.
+\ei
+From this second point, is it valid to assert that it is {\em not} disruptive to  software projects
+to let some issues advance from one phase to the next?
+
+\subsection{Alternate Explanations}
+
+One explanation for the above is that our sample is biased-- that while general software
+systems suffer from the delayed issue effect, systems built using TSP are so much better controlled
+that they  mitigate against the DIE effect.
+
+It would be convenient to offer the last paragraph as the conclusion of this study. Two of our authors (Nichols and Shull)
+ work at the organization that invented and promoted TSP.
+
 \subsubsection*{H4: A post-development DIE effect can be predicted from pre-deployment data} 
@@
 \item In \fig{b81}: see the data relating to development and acceptance testing;
-\item In  all the case studies of \fig{steck04} in the columns {\em Requirements} to {\em Test}.
+\item In  all the case studies of \fig{steck} in the columns {\em Requirements} to {\em Test}.
 \ei
@@
 \subsection{Construct validity}
+XXXX die does not aoccur during developmentjust during deployment
+
+that means during period requirements to test pre-deployent, there is not rush to resolve issues (just as long as a concerted effort is made to retire all issues priot to deployment)
+
+all our data is skewed due to TSP.
+two responses.
+1) we don't thing so.
+2) sure, let it due to TSP in which case we can fix the delayed issue effect via process methods 9everyone adopts the team software process).
+In either case, the delayed issue effect can be ignored or mitigated without extensive investment in radically new methods or tools (e.g.
+automatic formal methods, better requirement description languages).
+
 The essential constructs in our analyses are \emph{cost-to-fix, defects, and phase delay}.
@@
 If Boehm and Royce do not include all of these costs, then the TSP cost curve in \fig{cost-to-fix-tsp} artificially high by comparison.
-
-\subsection{Reliability}
-A previous paper~\cite{shirai14} has applied a range of sanity checks to the data.
-A common property of real-world data sets is the presence
-of noisy entries (superfluous  or spurious data). 
-The level of noise can be quite high. As reported
-in \cite{shepperd12}, around
-10\% to 30\%
-of the records in the NASA MDP defect data sets are
-affected by noise. Nichols et al.~\cite{shirai14}  report that
-the noise levesl in the SEI TSP data are smaller than those seen
-in other data sets. They found in the SEI TSP data that:\bi 
-\item
-4\% of the data was incorrect such as  null values of illegal formats;
-\item  2\% of the data has inconsistencies such as timestamps
-where the stop time was before the start time;
-\item 3\% of the data had data that was not credible
-such as tasks listed in one day that took like than six hours.
-\ei
 
@@
 \section*{Acknowledgements}
-The authors gratefully acknowledge the careful comments of anonymous reviewers-- particularly, their comments on issues relating to ``cost''
+The authors wish to thank  David Tuma and  Yasutaka Shirai for their work on the SEI databases
+that made this analysis possible.
+In particular, we thank Tuma Solutions for providing the Team Process Data Warehouse software.
+Also, the authors gratefully acknowledge the careful comments of anonymous reviewers-- particularly, their comments on issues relating to ``cost''
 and ``time''.
@@
 Foundation grants NSF-CISE 1302169 and CISE 1506586.
-\\
-We thank Tuma Solutions for providing the Team Process Data Warehouse software in support of this research.  \\
 Personal Software Process$\textsuperscript{SM}$, Team Software Process$\textsuperscript{SM}$, and TSP$\textsuperscript{SM}$ are service marks of Carnegie Mellon University.

### 2015-08-02 lucasRelatedWork.tex
Updates from ShareLaTeX

@@
 the data used to make prior conclusions.
-As an example of this, \fig{steck} shows one 2004 survey that reports an exponential delayed issue  effect for 
-requirements issues in eight case studies. The right-hand-side of that column shows which of those referenced studies
-we can still access-- and a ``?'' denotes a references from a broken url link (note that most
-of those links are broken). Note that nearly all of that data is now no longer accessible.
+As an example of this, \fig{steck} shows one 2004 survey that reports an 
+exponential delayed issue  effect for 
+requirements issues in eight case studies. Note that we cannot verify
+{\em any} of those results since the links in the references of
+that  survey are all broken (``page not foind'' errors).
 
@@
 \begin{center}
-\begin{tabular}{r|rrrr|c}
- case& \multicolumn{4}{c|}{Phase Requirements Issue Found }&  \\
- study               &Requirements & Design & Code&  Test& source\\\hline
-1 & 1& 6& 10& 70&~\cite{Boehm81}\\  
-2& 1 &3& 5& 37&?\\ 
-3&  1 &3& 7& 51& ?\\ 
-4& 1& 5 &33 &75& ? \\
-5& 1  &    & 10 & 40& ?\\ 
-6& 1   &    & 10 &  40& ?\\
-7&  1 & 20 & 45 & 250& ?\\  
-8&  1  &   5 &      & 50& ? \\\hline
-median           & 1 &  5  & 10   & 50.5 
+\begin{tabular}{r|rrrr}
+ case& \multicolumn{4}{c}{Phase Requirements Issue Found }  \\
+ study               &Requirements & Design & Code&  Test\\\hline
+1& 1 &3& 5& 37\\  
+2& 1  &    & 10 & 40 \\ 
+3& 1   &    & 10 &  40 \\
+4&  1  &   5 &      & 50  \\
+5&  1 &3& 7& 51 \\
+6& 1& 5 &33 &75  \\
+7&  1 & 20 & 45 & 250 \\  
 \end{tabular}
@@
 \caption{Sample of the Stecklein et al. literature review~\cite{steck04}:
-cost to resolve requirements issues, relative to resolving issues during requirements.
-Question marks ``?'' denotes a reference to
-a result that is not longer accessible (i.e. a broken URL link).}\label{fig:steck}
+cost to resolve requirements issues, relative to resolving  during requirements.}\label{fig:steck}
 \end{figure}
 
-\begin{figure}[!b]
+\begin{figure}[!t]
  \includegraphics[width=3in]{img/clutterbuck.png} 
@@
  All in all, the empirical results of this section seem insufficient to justify
- the widely-held beliefs documented in the previous section. Adding to our doubts
- was the discovery of at least two studies that report less-than exponential
+ the  beliefs documented in the previous section. Adding to our doubts
+ was the discovery of   several studies that report less-than exponential
  increase in the difficulties associated with making the changes associated with delayed
@@
 
-\begin{figure}
- \includegraphics[width=3.3in]{img/Royce98.png}
- \caption{Exception to the rule - CCPDS-R case study cost-to-fix curve. From~\cite{Royce98}.}\label{fig:royce}
- \end{figure}
-
 Another example of less-than exponential explosion in the difficulty associated with delayed issues comes from the  CCPDS-R
@@
 Boehm~\cite{Boehm10} attributes this success to the CCPDS-R development process, which focused on removing architecture risk early in the development lifecycle.
+
+ Other examples, where research papers report non-exponential changes in the difficuty in resoloving issues
+are:
+  \bi
+  \item
+  Boehm~\cite{Boehm80} reported a flatter growth rate for small, non-critical projects.
+   \item 
+  Data from NASA's Johnson Space Flight Center, reported by Shull~\cite{Shull02}, found that the cost to find certain non-critical classes of defects was fairly constant across lifecycle phases (1.2 hours on average early in the project, versus 1.5 hours late in the project).
+
+\ei
+
+\begin{figure}
+ \includegraphics[width=3.3in]{img/Royce98.png}
+ \caption{Exception to the rule - CCPDS-R case study cost-to-fix curve. From~\cite{Royce98}.}\label{fig:royce}
+ \end{figure}

### 2015-08-02 phaseDelay.tex
Updates from ShareLaTeX

@@
 
-One technical aside: there are several ways to  measure {\em difficulty}.
-The first is a direct measure of the actual work time spent in correcting the defect in the product.
-Direct measurement enables analysis of the variability the cost-to-fix among a category of defects.
-The second approach measures the total time spent in a defect removal phase and divides by the total defects removed.
-This provides an average cost-to-fix including overhead of that activity.
-For example, a code inspection may require an hour of effort to remove three defects for a 3 defects per hour or to invert the ratio, 1/3 hours per defect.
-One observation made by Nichols while studying our 171 projects is that most of the time  is spent locating the defects not actually fixing them.
-we use the second approach in our analysis.
-
 This paper defects the  following claims. Note that we call them
@@
 
-\begin{figure}[!t] 
-%\vspace{-1cm}
-%\hspace{-0.35in}
-\begin{center}
-\includegraphics[width=3in]{img/boehm-overlay.pdf}
- \end{center}
-% ~\\~\\
-%~\\~\\~\\
-
- \caption{Cost-to-fix curve from \protect\fig{cost-to-fix} 
- overlayed with case study from~\cite{Royce98} and TSP data. Scale is relative to the phase with the lowest cost to fix. }\label{fig:cost-to-fix-tsp}
- \end{figure}
-
 \begin{figure*}[!t] 
@@
 \ei
+
 \subsubsection{Specific1}
@@
 requirements errors are the most expensive to fix 
-{\em when found during production but the cheapest to fix early in development}}. 
+{\em when found during production but the cheapest to fix early in development}. 
 Our data from these 171 project does not extend into post-delivery deployment. That said,
@@
 Specifically: prior reports of an  exponential increase in difficulty in resolving issues {\em during
-deployment} might have been used to predict for exponentially increasing
-difficulty {\em post-deployment}.
+deployment}.
 For example:
@@
 \ei
+Any manager watching  these intra-development trends would be well-justified
+to anticipate that  the difficulty in resolving issues would will get much worse. Such managers
+would therefore expect an exponentially harder time resolving issues during deployment.
+
 On the other hand, a remarkable feature of Figure~\ref{fig:fix-phase-dist} is the {\em absence}
-of an exponential delay effect in the difficulty to resolve requirements issues during development.
-The greatest difficulty in resolving requirements issues was 1.31 seen in systems testing--
+of an exponential delay effect during development.
+In our data, the greatest difficulty in resolving requirements issues was 1.31 seen in systems testing--
 which is far smaller than the 37 to 250-fold increases reported in \fig{b81}, \fig{cost-to-fix},
-and \fig{steck}.  That is, from our TSP data, we would not anticipate exponentially greater difficulty in resolving requirements issues after
+and \fig{steck}.  That is, from our  data, we would not anticipate exponentially greater difficulty in resolving requirements issues after
 deployment. 
@@
 \ei
-
-For illustrative purposes, the relative cost to fix errors per lifecycle phase in the TSP project data is overlayed with Boehm's cost curve~\cite{Boehm81} as well as the CCPDS-R case study~\cite{Royce98} in \fig{cost-to-fix-tsp}.
+In summary:
+\bi
+\item
+The delayed issue effect has been reported in around a dozen papers, at least half of which date back to the 1970s. 
+\item
+But there is no trace of that  effect   in hundreds of software projects built  since 2006.
+\ei
+%For illustrative purposes, the relative cost to fix errors per lifecycle phase in the TSP project data is overlayed with Boehm's cost curve~\cite{Boehm81} as well as the CCPDS-R case study~\cite{Royce98} in \fig{cost-to-fix-tsp}.
 Cost-to-fix for TSP is total time spent in each review, inspection, or test-related phase / \# of defects found in phase (reported in \fig{raw}).
 It is clear to see that the trend is relatively flat, closer to the pattern exhibited by the Royce case study (which also focused on early defect removal) than the exponential growth curve.
+
+%\begin{figure}[!t] 
+%%\vspace{-1cm}
+%\hspace{-0.35in}
+%\begin{center}
+%\includegraphics[width=3in]{img/boehm-overlay.pdf}
+ %\end{center}
+% ~\\~\\
+%~\\~\\~\\
+
+ %\caption{Cost-to-fix curve from \protect\fig{cost-to-fix} 
+ %overlayed with case study from~\cite{Royce98} and TSP data. Scale is relative to the phase with the lowest %c%ost to fix. }\label{fig:cost-to-fix-tsp}
+% \end{figure}
 
@@
 \section{Validity}
-
-To sum up the paper so far,
-\bi
-\item
-The delayed issue effect has been reported in around a dozen papers, at least half of which date back to the 1970s. 
-\item
-But there exists
-no evidence for this effect was seen  in hundreds of software projects build   since 2006.
-\ei
-From this second point, is it valid to assert that it is {\em not} disruptive to  software projects
-to let some issues advance from one phase to the next?
-
-\subsection{Alternate Explanations}
-
-One explanation for the above is that our sample is biased-- that while general software
+We report validity concerns following the guidelines of Runeson \& H\"{o}st~\cite{runeson09}.
+
+\subsection{External validity}
+Threats to external validity affect the generalization of findings. The most obvious threat to external validity is that TSP is not representative of most software development processes. Hence,
+For example, 
+one way to discount the above is that our sample is biased-- that while general software
 systems suffer from the delayed issue effect, systems built using TSP are so much better controlled
@@
 
-It would be convenient to offer the last paragraph as the conclusion of this study. Two of our authors (Nichols and Shull)
- work at the organization that invented and promoted TSP.
-
-\subsubsection*{H4: A post-development DIE effect can be predicted from pre-deployment data} 
-The data collected from our 171 projects     pertains only to
-development, and not to subsequent operations.
-So did we miss the DIE effect in our data since
-our data collection is just not prolonged enough to detect that effect?
-
-We would argue that this is not the case.
-An  exponential DIE effect can be seen {\em during the development} of all
-the papers we have reviewed. For example:
+It would be convenient to offer the last paragraph as the conclusion of this study
+(two of our authors, Nichols and Shull,  work at the organization that invented and promoted TSP).
+ However, we cannot endorse such a conclusion:
  \bi
-\item In \fig{b81}: see the data relating to development and acceptance testing;
-\item In  all the case studies of \fig{steck} in the columns {\em Requirements} to {\em Test}.
+ \item
+We view TSP as a better way to
+ {\em monitor} the activities {\em of  existing projects}. That is, TSP
+ does not radically change a project, it just offers a better way to log the activity within
+ that project. 
+ \item
+ Our 171 project contain examples of a wide variety
+ of systems (e-commerce web portals and  banking systems) run in a variety of
+ ways (agile or  waterfall or some combination of the two). Hence, rather
+ than being unduly  biased in some manner,
+ we view our 171 projects as a large sample of current industrial processes for software
+ development.
  \ei
-That is, in the literature we have surveyed, the post-development DIE effect would have been predicted by data collected during pre-deployment development.
-Such
-
-Another surprise was a lack of {\em growth trend} suggesting that, in operation,
- we should expect larger exponential delays in resolving issues. This is odd since all
- the papers we could access on DIE should such a growth trend in the pre-operational data.
-
-\subsubsection*{H5: ``Time'' to resolve an issue is analogous to ``cost''}
-This hypothesis a somewhat technical point needed for our argument.
-While classical papers on the delayed issue effect discuss the {\em cost}
-of delaying issues, the data from our 171 projects only has time data.
-Hence, to say that our data addresses the classic delayed issue effect,
-we need to make the case that our time measurements are an appropriate
-surrogate for the prior claims on the cost of delaying issues.
-
-We report validity concerns following the guidelines of Runeson \& H\"{o}st~\cite{runeson09}.
+Another way to discount this result is to say we are only reporting on small to medium-size
+projects, and the real impact of delayed issues is on very large projects.
+There certainly exists another class of very large software project where early requirements
+errors would be catastrophic for later development. For example, when NASA launches deep space missions,
+each of those satellites are special-purpose, uniquely-built devices built by a large contractor
+population. Such systems are very hard to reconfigure halfway through the development.
+
+We concede that for very large complex NASA-like developments, the delayed issue
+effect could still hold. However, perhaps we need to split our software engineering principles
+into two parts- one for   very large NASA-like projects (which are not very numerous)
+and one for smaller-sized projects like the ones studied here (which are far more numerous).
+Most software is {\em not} like the software used on NASA deep-space satellites.
+As  agile methods grew more popular last decade; and as personnel computers grew more powerful;
+and as the libraries associated with particular languages grew more extensive; then smaller
+teams starting  deliver systems that (previously) had required very large teams.
+As mentioned above,
+our  171 projects represents $\frac{171}{212}=80$\% of the software projects
+seen by the Software Engineering Institute. That is, as far as we can tell,
+there are more small to medium-sized projects than otherwise.
 
 \subsection{Construct validity}
-XXXX die does not aoccur during developmentjust during deployment
-
-that means during period requirements to test pre-deployent, there is not rush to resolve issues (just as long as a concerted effort is made to retire all issues priot to deployment)
-
-all our data is skewed due to TSP.
-two responses.
-1) we don't thing so.
-2) sure, let it due to TSP in which case we can fix the delayed issue effect via process methods 9everyone adopts the team software process).
-In either case, the delayed issue effect can be ignored or mitigated without extensive investment in radically new methods or tools (e.g.
-automatic formal methods, better requirement description languages).
-
-The essential constructs in our analyses are \emph{cost-to-fix, defects, and phase delay}.
-
-\emph{Cost-to-fix} is traditionally measured as either effort spent to fix a defect or time spent fixing a defect.
-In our analysis, we calculate the latter as \# of defects fixed in a defect-removal plan item, such as DesignReview or CodeInspection, divided by the time spent in the plan item.
-A single project has multiple such plan items.
-In the TSP data, this time accounts for time spent isolating, analyzing, and fixing the defect, but does not include time spent verifying the fix.
-
-In our sample, all \emph{defects} are considered equally, and thus TSP's reporting methods may result in ``small'' defects could be dominating the sample.
-For example, typographical requirements errors are included as well as requirements errors that necessitate an architectural change to the system.
-Our analysis does not include a notion of defect severity, however, we observe no indications in previous cost-to-fix results reported by \cite{Boehm81} and \cite{Royce98} that severity was considered in their samples.
-However, evidence discussed in \cite{Shull02} suggests that low severity defects may exhibit a lower cost to change.
+
+Our definition of {\em difficult to resolve} combines two concepts: time to change and cost to
+change. In the above we have used them interchangable, comparing our time to resolve data (from the 171 TSP projects) against the cost-to-fix results of Boehm et al. e.g. \fig{b81}.
+Is it valid to assume the equivalence of time and cost?
+
+Certainly, there are cases where time is not the same as cost. Consider, for example,
+if debugging required some very expensive tool or the services or a very senior (and hence, very expensive)
+developer. Under those circumstances, time does not equate to cost.
+
+Also, there is the issue of the implication of  a defect.
+Evidence discussed in \cite{Shull02} suggests that low severity defects may exhibit a lower cost to change.
 Nonetheless, even ``small'' typographical errors have been known to cause enormous damage (e.g., the Mars climate Orbiter).
 
-\emph{Phase delay}, as depicted in the scale up column in \fig{cost-to-fix-tsp} and \fig{raw}, is a discrete value that does not consider the total time that a defect is latent in the system.
-Thus, the phase delay of two defects injected in Code but found in System Test will be the same even though one was found at the beginning of test while the other at the end of a potentially lengthy test period.
-Ideally, one could examine the total real time a defect was in a system from the point at which is was injected to the point at which it was discovered.
-It may be possible to gather such information if the project keeps rigorous change logs for its requirements, design, and code artifacts.
+Having documented the above issues, we assert that they are probably not major issues in the study.
+As to the first point, that some debugging may be more expensive than others, one of us (Nichols)
+was closely associated with many of the projects in our sample. He is unaware of any frequent
+use of exorbitantly expensive tools or people on these projects. As to the second point (that
+some defects have an impact beyond the mere cost to fix them), we no evidence suggesting
+that the net effect of this concern over all our projects is non-zero.
 
@@
 Nonetheless, individual developers and teams may apply their own internal rules for filtering defects, which would lead to inconsistent reporting thresholds among the projects in our sample.
-
-\subsection{External validity}
-Threats to external validity affect the generalization of findings.
-The most obvious threat to external validity is that TSP is not representative of most software development processes.
-TSP was, in fact, designed to be rigorous and eliminate defects early in the lifecycle and soon after injection.
-We make no assertions regarding the generalizability of these results beyond TSP, we observe that other case studies in various domains have exhibited flattened cost-to-fix curves.
-
-Our analysis of cost-to-fix does not include defects found in production, which, in Boehm's original data, account for costs up to 200 times the cost of finding and fixing defects in Requirements.
-Further, it is not clear how Boehm~\cite{Boehm81} and Royce~\cite{Royce98} define \emph{cost-to-fix}.
-In our analysis, cost-to-fix includes the time spent analyzing, fixing the defect, and validating the fix.
-If Boehm and Royce do not include all of these costs, then the TSP cost curve in \fig{cost-to-fix-tsp} artificially high by comparison.
 
@@
 
-In this paper, we have examined contemporary project data for evidence to support the well-known software engineering truism of the phase delay effect. 
-We found that, in the literature, the evidence for phase delay was very spare.
+In this paper, we have examined contemporary project data for traces of
+the widely-believed delayed issue effect.
+We found that, in the literature, the evidence for this effect was very spare--
+in fact, we found
 %and (in a $21^{st}$).  %don't know what this is supposed to be
@@
   To the contrary, four non-TSP  case studies reported that the expected exponential growth in cost-to-fix was not observed:
-  \bi
-  \item
-  Boehm~\cite{Boehm80} reported a flatter growth rate for small, non-critical projects.
-  \item
-  Clutterbuck~\cite{Clutterbuck09} showed a flatter cost curve in an agile case study. 
-  \item 
-  Data from NASA's Johnson Space Flight Center, reported by Shull~\cite{Shull02}, found that the cost to find certain non-critical classes of defects was fairly constant across lifecycle phases (1.2 hours on average early in the project, versus 1.5 hours late in the project).
-  \item
-  Royce~\cite{Royce98} reported a flat cost curve for a project that invested heavily upfront in architectural reviews.
-\ei
+
 If the lack of phase delay is not due to TSP, then what does cause it?
@@
 Changes to  software that is used by a large population can be confusing (particularly for the sick or elderly). 
-Consider the 23,000,000 veterans working with  the U.S.  Veterans 
+Consider the 23 million veterans working with  the U.S.  Veterans 
 Administration (VA) service. Some of those veterans are dependent on health care or pensions from the VA. Those veterans use the VA's web interfaces

### 2015-08-02 phaseDelay.tex
Updates from ShareLaTeX

@@
 in fact, we found
-%and (in a $21^{st}$).  %don't know what this is supposed to be
-We suspect that a very old result was requoted, without being 
-critically re-examined, for literally decades.
+just as many articles supporting this  exponential effect~\cite{Boehm76,Boehm81,teck,Fagan76,Stephenson76}
+than otherwise~\cite{Clutterbuck09,Clutterbuck09,Boehm80} and one paper that
+offered evidence consistent with {\em both}
+the the delayed issue effect and its opposite~\cite{shull02}.
 
@@
  Since the phase delay effect has been such a long-lived software engineering truism, as well as one that is accepted by a substantial population of software engineers, it is important to explore possible explanations for our observations.
-
-One hypothesized explanation that can be ruled out is that this lack of
-a phase delay effect is due to the use of TSP.
-  To the contrary, four non-TSP  case studies reported that the expected exponential growth in cost-to-fix was not observed:
 

### 2015-08-03 phaseDelay.tex
Updates from ShareLaTeX

@@
 in fact, we found
-just as many articles supporting this  exponential effect~\cite{Boehm76,Boehm81,teck,Fagan76,Stephenson76}
+just as many articles supporting this  exponential effect~\cite{Boehm76,Boehm81,steck04,Fagan76,Stephenson76}
 than otherwise~\cite{Clutterbuck09,Clutterbuck09,Boehm80} and one paper that
 offered evidence consistent with {\em both}
-the the delayed issue effect and its opposite~\cite{shull02}.
+the the delayed issue effect and its opposite~\cite{Shull02}.
 
@@
 
-\subsection{Future Work}
+\section{Future Work}
 There is another ``phase delay'' effect, not discussed above. Even if software is readily updatable, there   may be a ``user phase delay''   amongst   humans using that software.

### 2015-08-03 phaseDelay.tex
Updates from ShareLaTeX

@@
 \usepackage{colortbl}
+
+\usepackage{balance}
 
@@
 
-\begin{figure*}[!t] 
+\begin{figure}[!t] 
  \renewcommand{\baselinestretch}{0.7}
@@
 \begin{center}
-\begin{tabular}{r|rrr|ll|rl}  % I attach your cursor. thrust, parry, yo!
-  Sample&\multicolumn{3}{c|}{Percentiles}\\ 
-size & 25th & 50th & 75th & Phase injected & Phase removed & \multicolumn{2}{l}{Scale up w.r.t. to first phase}\\\hline
+\begin{tabular}{r@{~}|r@{~}r@{~}r|@{~}l@{~}l|@{~}r@{~}l}  % I attach your cursor. thrust, parry, yo!
+  \multicolumn{1}{c}{~} &\multicolumn{3}{c}{Percentiles}&\multicolumn{2}{l}{~~~~~~Phase}\\\cline2-6}
+N & 25th & 50th & 75th &   injected & removed   & \multicolumn{2}{l}{Scale up w.r.t. to first phase}\\\hline
 \\
@@
 \caption{Distribution of fix times seen in SEI TSP data. Left hand side bars
-show scale up on median values.}
+show scale up on median values.
+}
 \label{fig:raw}
-\end{figure*}
+\end{figure}
 
@@
 
+%\section{Future Work}
+%There is another kind of delayed issues  effect, not discussed above. Even if software is readily updatable, there   may be a ``user  inertia''   amongst   humans using that software. 
+%To understand the 
+%Changes to  software that is used by a large population can be confusing (particularly for the sick or elderly). 
+%Consider the 23 million veterans working with  the U.S.  Veterans 
+%Administration (VA) service. Some of those veterans are dependent on health care or pensions from the VA. Those veterans use the VA's web interfaces
+%to  (e.g.) find out
+%when their next cheque will arrive. Before we can update the VA web software, we would
+%need to  a transition plan that helps users   move from the older to the newer system.
+This is an example of what might be called a ``user phase delay''; i.e.
+changes to software are more complex to process, particularly if some users have styled
+%their day-to-day work flow around the services offered by the older software.  
+%
+%This issue of ``user phase delay'' is not addressed by our data, nor by the prior studies on phase delay %discussed above.
+Hence, this issues   is out of scope for this paper since the expectation that ``older %bugs cost more to fix than newer bugs'' was not formed around
+%user issues.
+
+%That said,  
+%we think that ``user phase delay'' will become a bigger issue in the future when more people are using more software as part of their day to day life.
+We even envision a time when users demand fewer updates, just to offer some stability in their life.
+
  \section{Conclusion}
@@
 
-In this paper, we have examined contemporary project data for traces of
-the widely-believed delayed issue effect.
-We found that, in the literature, the evidence for this effect was very spare--
-in fact, we found
-just as many articles supporting this  exponential effect~\cite{Boehm76,Boehm81,steck04,Fagan76,Stephenson76}
-than otherwise~\cite{Clutterbuck09,Clutterbuck09,Boehm80} and one paper that
-offered evidence consistent with {\em both}
-the the delayed issue effect and its opposite~\cite{Shull02}.
-
-We  also  reported the results of a survey of both researchers and practitioners which showed this truism is still a widely held belief in the software engineering community. Such a widely
-held belief, for something with so little evidence, can be damaging if ever it is used
-to justify spending scarce resources on novel  methods and tools .
-
-Using data from 171 projects from the period 2006 to 2014, we then looked for any evidence
-for three different interpretations of the phase delay effect; namely:
-\bi
-\item
-That defects are more expensive to fix in later phases; 
-\item
-That defects become more expensive the more phases they remain; 
-\item
-That requirements errors are more expensive to fix the later they are found. 
-\ei
-Our dataset provided no support for any of these phrasings. 
-This is significant, as we argue that this paper is (by far)    the largest dataset ever used to explore phase delay (but see ~\cite{Boehm80,Clutterbuck09,Shull02,Royce98} 
-for other, much older, much smaller studies).
-
-Despite this lack of evidence, the myth remains that phase delay causes increased time to resolve issues.
- Since the phase delay effect has been such a long-lived software engineering truism, as well as one that is accepted by a substantial population of software engineers, it is important to explore possible explanations for our observations.
-
-If the lack of phase delay is not due to TSP, then what does cause it?
-We surmise that the substantial changes in software development and deployment technologies and practices, since the phase delay effect was first measured, may be responsible for our observations in the TSP data set.
-Recall that the original data on which the phase delay effect is based, as reported in \S3, are approximately 40 years old.
-The most recently reported data, from Shull et al.
-\cite{Shull02}, was published in 2002 but contains substantial retrospective data.
-Over time, it would be surprising if the investment in better languages, IDEs, and other development technologies had in fact not created some improvement.
-To pick just one example, we note that much of the data showing the phase delay effect was collected before frequent and unobtrusive auto-updating software was as common as today, which certainly flattens the cost of changes.
-
+In this paper, we explored   the papers and data related to the 
+widely-believed {\em delayed issue effect} (that delaying the resolution of issues
+exponentially 
+increases the difficulty of completing that  resolution).
+Several prominent SE researchers state this effect is some fundamental law of software engineering~\cite{boehm01,mcconnell01,boehm01,glass02},
+Based on a  survey  of both researchers and practitioners, we  found that
+a specific form  of this effect (requirements errors are hardest to fix) is  widely-believed
+held in the community.
+
+We checked for traces of this effect in 171 projects from the period 2006 to 2014.
+That data held no trace of the delayed issued effect.
+To the best of our knowledge, this paper is the  largest study
+of this effect yet performed.
+
+Our explanation for this lack-of-effect is five-fold:
+\be
+\item The effect might be an historical relic. Evidence:
+the effect was first described in the era of punch card computing and non-interactive environments).
+\item The effect might have been intermittent (rather than some fundamental law of software). Evidence: we can  found nearly
+as many papers reporting the effect~\cite{Boehm76,Boehm81,steck04,Fagan76,Stephenson76} as otherwise~\cite{Royce98,Boehm80,Shull02}.
+\item The effect might be confined to very large systems- in which case it would be
+acceptable during development to let smaller to medium
+sized projects carry some unresolved issues from early phases into later phases.
+\item The effect might be mitigated by modern software development approaches that
+encourage change and revision of older parts of the system.
+\item The effect might be mitigated by modern software development tools
+that simplify the process of large-scale reorganization of software systems.
+\ee
  %One hypothesized explanation that can be ruled out is that the lack of a phase delay effect is due to using the TSP method. 
-
-The question remains: why did this idea, with so little support, become so widespread in the software engineering literature?
+Our results beg the question: why did an idea, with so little support, become so widespread in the software engineering literature?
 No doubt the original evidence was compelling at the time, but much has changed in the realm of software development in the subsequent 40 years.
-Possibly the concept of phase delay has persisted because, to use Glass's terms on the subject, it seems to be ``just common sense''\cite{glass02}.
-We believe that these results indicate that, in a rapidly changing field such as software engineering, even widely-held rules of thumb must be periodically re-verified.
-In turns out that common sense may not be so sensible after all.
+Possibly the concept of the delayed issue effect (or its more specific description: requirements errors are the hardest to fix)
+has persisted because, to use Glass's terms on the subject, it seems to be ``just common sense''\cite{glass02}. 
+Nevertheless, in a rapidly changing field such as software engineering, even widely-held rules of thumb must be periodically re-verified. 
 Progress in the domain of software analytics has made such periodic checks more cost-effective and feasible, and we argue that an examination of local behaviors (rather than simply accepting global heuristics) can be of significant benefit.
-
-\section{Future Work}
-There is another ``phase delay'' effect, not discussed above. Even if software is readily updatable, there   may be a ``user phase delay''   amongst   humans using that software.
-
-Changes to  software that is used by a large population can be confusing (particularly for the sick or elderly). 
-Consider the 23 million veterans working with  the U.S.  Veterans 
-Administration (VA) service. Some of those veterans are dependent on health care or pensions from the VA. Those veterans use the VA's web interfaces
-to  (e.g.) find out
-when their next cheque will arrive. Before we can update the VA web software, we would
-need to  a transition plan that helps users   move from the older to the newer system.
-This is an example of what might be called a ``user phase delay''; i.e.
-changes to software are more complex to process, particularly if some users have styled
-their day-to-day work flow around the services offered by the older software.
-
-This issue of ``user phase delay'' is not addressed by our data, nor by the prior studies on phase delay discussed above.
-Hence, this issues   is out of scope for this paper since the expectation that ``older bugs cost more to fix than newer bugs'' was not formed around
-user issues.
-
-That said,  
-we think that ``user phase delay'' will become a bigger issue in the future when more people are using more software as part of their day to day life.
-We even envision a time when users demand fewer updates, just to offer some stability in their life.
-This is a matter for future work.
 
@@
 \scriptsize
+\balance
 \bibliographystyle{plain}

### 2015-08-03 phaseDelay.tex
Updates from ShareLaTeX

@@
 \begin{tabular}{r@{~}|r@{~}r@{~}r|@{~}l@{~}l|@{~}r@{~}l}  % I attach your cursor. thrust, parry, yo!
-  \multicolumn{1}{c}{~} &\multicolumn{3}{c}{Percentiles}&\multicolumn{2}{l}{~~~~~~Phase}\\\cline2-6}
+  \multicolumn{1}{c}{~} &\multicolumn{3}{c}{Percentiles}&\multicolumn{2}{l}{~~~~~~Phase}\\%\cline{2-6}
 N & 25th & 50th & 75th &   injected & removed   & \multicolumn{2}{l}{Scale up w.r.t. to first phase}\\\hline

### 2015-08-14 data.tex
Updates from ShareLaTeX

@@
 
+Overall, the   effort associated with adding TSP to a project is not onerous. McHale reports~\cite{mchale02}:
+\bi
+\item
+ The time spent  tracking time, defects, and tasks requires less then 3\% of a developer's time. Weekly team meetings  require at most an hour, which is
+only 2.5\% of a 40 hour work week. 
+\item
+Team launches and replans average about 1 day per month or 5\% planning overhead.
+\ei
+It is true that one staff member is needed as a ``coach'' to mentor the teams
+and certify and monitor that data collection.
+However, one of us (Nichols) has worked with dozens of TSP teams.
+He reports that one  trained coach can support 4 or 6 teams (depending upon team experience).
+
 \subsection{Data Integrity}
@@
 where the stop time was before the start time;
-\item 3\% of the data had data that was not credible
-such as tasks listed in one day that took like than six hours.
+\item 3\% of the data contained values that were not credible
+such as tasks listed in one day that took more than six hours for a single developer.
 \ei 
@@
 The use of certified local coaches within each project increases the integrity of our data.
-
-Overall, the   effort associated with adding TSP to a project is not onerous. McHale reports~\cite{mchale02}:
-\bi
-\item
- The time spent  tracking time, defects, and tasks requires less then 3\% of a developer's time. Weekly team meetings  require at most an hour, which is
-only 2.5\% of a 40 hour work week. 
-\item
-Team launches and replans average about 1 day per month or 5\% planning overhead.
-\ei
-It is true that one staff member is needed as a ``couch'' to mentor the teams
-and certify and monitor that data collection.
-However, one of us (Nichols) has worked with dozens of TSP teams.
-He reports that one  trained coach can support 4 or 6 teams (depending upon team experience).
 

### 2015-08-14 data.tex
Updates from ShareLaTeX

@@
  Although the representation suggests a waterfall model, the SEI experience is that all real implementations of any size follow a spiral approach with many team performing the work in iterative and/or incremental development cycles.
+
+One special feature of  \fig{waterfall} is
+the {\em before} where the   TSP Launch Preparation scripts and checklists (see TSP: Leading a development Team~\cite{Humphrey:2005}) require the TSP team to assure that management and business have clearly identified cost, schedule, and scope goals appropriate to the upcoming development activities, often including a conceptual model.
+For example an architecture team must have sufficient requirements to reason about, prototype, and specify an architecture~\cite{Bachmann13} while a coding only team within a larger project would have more precisely defined requirements and high level design.
+
 %Within a project increment, multiple features or components may be developed and incremented.

### 2015-08-14 phaseDelay.tex
Updates from ShareLaTeX

@@
 
-In summary, there exists (1)~historical reasons for believing in the delayed issued
-effect as well as (2)~reasons to believe that that effect may not longer hold.
-Hence,  this article explores the currency of the delayed issue effect.
+The above argument is an anecdotal evidence that  the delayed issue effect might no longer exist. But anecdotes are not really rigorous
+evidence. Accordingly,  this article explores the currency of the delayed issue effect.
 After some initial definitions, we discuss the
 value of checking old ideas.
-Next, survey of industrial practitioners and researchers to document a widespread belief that delayed issues have a negative impact on projects.
-Next, we  analyze 171 software  projects developed in the period 2006 to 2014 to find  {\em no evidence} of the delayed issue effect
-(note that, to ensure reproducibility,
+Next, a survey of industrial practitioners and researchers to document a widespread belief that delayed issues have a negative impact on projects.
+After that, we  analyze 171 software  projects developed in the period 2006 to 2014 to find  {\em no evidence} of the delayed issue effect:
+\bi
+\item
+To ensure reproducibility,
 all the data  used in this study is available in the PROMISE
-repository at openscience.us/repo). Finally, we discuss the validity and implications of our results.
+repository at openscience.us/repo. 
+\item
+To the best of our knowledge,
+this the largest study devoted the delayed issue effect yet conducted.
+\ei
+Finally, we discuss the validity and implications of our results.
 
@@
 \item
-{\em Delayed issue effect}:   it is {\em exponentially}  more {\em difficult} to resolve an issue in a software project, the {\em longer} it remains.
+The {\em delayed issue effect}:   it is {\em exponentially}  more {\em difficult} to resolve  issues in a software project, the {\em longer} they remain.
 \item
@@
 
-This paper defects the  following claims. Note that we call them
+This paper defends the  following claims. Note that we call them
 ``claims'' not ``hypotheses'' since the later require defense via some statistical
@@
 \section{``DIE'' is a  widely-held belief}
-To assess the prevalence of the delayed issue effect,   conducted a survey of software engineers. If our surveyed practitioners make management decisions based on their
-understanding of SE theory, then the phase delay effect may well influence their judgment.
+To assess the prevalence of DIE,   we conducted a survey of software engineers. If our surveyed practitioners make management decisions based on their
+understanding of SE theory, then the DIE  may well inappropriately influence their decisions.
 
@@
 
-Using Figure~\ref{fig:fix-phase-dist}, we can comment on two specific and more general
-delayed issued effects. 
-\bi
-\item {\bf Specific1}: Requirements errors are the most expensive to fix;
-\item {\bf Specific2}: Requirements errors are the most expensive to fix, 
- when found during production but the cheapest to fix early in development;
- \item {\bf General:}  It is exponentially more  difficult  to resolve an issue in a software project, the   longer it remains.
-\ei
-
-\subsubsection{Specific1}
-
-Our data does not support   
+\subsubsection{Comment 1}
+Figure~\ref{fig:fix-phase-dist}
+ data does not support   
 {\em  requirements errors are the most expensive to fix}.
@@
 expect from the delayed issue effect).
-\subsubsection{Specific2}
+
+That said, Figure~\ref{fig:fix-phase-dist} does support the 
+claim that early lifecycle errors are most expensive to fix.
+In all our results, the biggest scale ups are found with resolving
+issues introduced in the {\em before} phase. For example,
+resolving {\em before} issues during unit testing was 2.56 times
+slower than resolving issues earlier in the lifecycle.
+
+Nevertheless,  it must be added that the scale up seen in 
+Figure~\ref{fig:fix-phase-dist} 
+is much less
+than suggested by the DIE literature. Returning to
+the {\em before/unit test} result: 2.56 is far smaller
+than the scale ups seen in Figure~1. In fact, nowhere in 
+our results do we see any exponential increase.
+
+\subsubsection{Comment 2}
 Next, we comment on 
@@
 difficult issues after deployment.
-Specifically: prior reports of an  exponential increase in difficulty in resolving issues {\em during
-deployment}.
+
+That evidence is as follows.
+Prior reports of an  exponential increase in difficulty in resolving issues {\em during
+development}.
 For example:
@@
 deployment. 
- \subsubsection{General}
+ \subsubsection{Comment 3}
  The above specific statements related just to requirements. Turning now
@@
 
+Another issue that threatens internal validity is that our
+results are dependent on developers correctly identifying which
+phase initially created an issue. Certainly, if that was done
+incorrectly in this study, then all our results must be questioned.
+However,  this issue threatens {\em every} study on the delayed
+issue effect-- so if our results are to be doubted on this score,
+then all prior work that reported the delayed issue effect should
+also be doubted. Moreover, the TSP method used in this study
+encourages greater care with error reporting:
+
+\bi
+\item Developers are trained   to make that judgement;
+\item All data entry is double-checked by the team couch 
+\item TSP demands that developers analyze their data to make
+process improvements. 
+\ei
+That is, TSP developers are always testing
+if their project insights are accurate. In such an environment,
+it is more likely that they will better identify the injection phase.
+
 %The definitions of defect, cost-to-fix, and phase delay are provided in earlier sections.

### 2015-08-17 lucasRelatedWork.tex

@@
-One reason that industrial practitioners and academics believe so strongly in th delayed issue effect is that they are often presented in the SE literature. For example,
+One reason that industrial practitioners and academics believe so strongly in the delayed issue effect is that it is often referenced
+in the SE literature. For example,
 we know that \fig{b81} has been presented to the gradaute SE class at North Carolina State University, without quarrel or critical comment, every year for the last decade.
@@
 \end{center}}
-\caption{Sample of the Stecklein et al. literature review~\cite{steck04}:
-cost to resolve requirements issues, relative to resolving  during requirements.}\label{fig:steck}
+\caption{Cost to resolve requirements issues, relative to resolving  during requirements. From ~\cite{steck04}.}\label{fig:steck}
 \end{figure}
@@
  All in all, the empirical results of this section seem insufficient to justify
- the  beliefs documented in the previous section. Adding to our doubts
- was the discovery of   several studies that report less-than exponential
+ the strength of the beliefs documented in the previous section. Adding to our doubts are  several studies that report less-than exponential
  increase in the difficulties associated with making the changes associated with delayed
- issues. For example,
+ issues. 
 Clutterbuck et al.~\cite{Clutterbuck09} studied 5-month effort by a small-to-medium enterprise team developing a 71KLOEC web interface to a database application to implement 18 change requests-- see Figure~\ref{fig:clutterbuck} (note that these were for new and changed user requirements, not defects).
@@
 
-Another example of less-than exponential explosion in the difficulty associated with delayed issues comes from the  CCPDS-R
-system described by Royce~\cite{Royce98}.
-This was a million-line, safety-critical missile defense system show in Figure~\ref{fig:royce}.
-In this project, design changes (including architecture changes) required approximately twice the effort of implementation and test changes, and the cost-to-fix in implementation and test phases increased slowly.
-Boehm~\cite{Boehm10} attributes this success to the CCPDS-R development process, which focused on removing architecture risk early in the development lifecycle.
+Another example of less-than exponential explosion in the difficulty associated with delayed issues comes from  Royce~\cite{Royce98}. He
+studied  a million-line, safety-critical missile defense system
+(see  Figure~\ref{fig:royce}).
+Design changes (including architecture changes) required approximately twice the effort of implementation and test changes, and the cost-to-fix in implementation and test phases increased slowly.
+Boehm~\cite{Boehm10} attributes this success to the  development process, which focused on removing architecture risk early in the development lifecycle.
 
- Other examples, where research papers report non-exponential changes in the difficuty in resoloving issues
-are:
+ Other examples that report the opposite of DIE are:
   \bi
@@
  \includegraphics[width=3.3in]{img/Royce98.png}
- \caption{Exception to the rule - CCPDS-R case study cost-to-fix curve. From~\cite{Royce98}.}\label{fig:royce}
+ \caption{Exception to the rule - The Royce study: cost-to-fix curve. From~\cite{Royce98}.}\label{fig:royce}
  \end{figure}

### 2015-08-17 lucas.tex
syncing

@@
 
-Summary statistics for the agreement and applicability scores for the RqtsErr and PhaseDelay laws are presented in Figure~\ref{fig:survey_results}.
+Summary statistics for the agreement and applicability scores for the RqtsErr and DelayedIssueEffect laws are presented in Figure~\ref{fig:survey_results}.
 Responses whose Applicability response was ''I don't know'' are omitted from analysis.
@@
 80-20 rule (defects to modules) & 6 & 3.7 & 4 & 3.5 & 4 \\
-\textbf{PhaseDelay} & 30 & 3.6 & 4 & -- & --  \\ 
+\textbf{DelayedIssueEffect} & 30 & 3.6 & 4 & -- & --  \\ 
 Missing reqts are hardest to fix & 7 & 3.6 & 4 & 3.7 & 4 \\
@@
 Further, any changes to the production %ystem have additional complexity (compared to changes while system in %evelopment) due to users, data existence, and architecture dependencies.
-Caveat: in the free repsonse texts, we note that the researchers who disagreed with the law generally asserted that requirements change can be expensive, but that depends on the process used (e.g., agile vs.
+Caveat: in the free reponse texts, we note that the researchers who disagreed with the law generally asserted that requirements change can be expensive, but that depends on the process used (e.g., agile vs.
 waterfall) and the adaptability of the system architecture.

### 2015-08-17 phaseDelay.tex
syncing

@@
 
-\begin{figure}[!b]
+\begin{figure}[!t]
 
@@
 Another way to discount this result is to say we are only reporting on small to medium-size
-projects, and the real impact of delayed issues is on very large projects.
+projects,
+(see
+\fig{dist}:  median
+effort was was  
+271 days). 
 There certainly exists another class of very large software project where early requirements
@@
 our  171 projects represents $\frac{171}{212}=80$\% of the software projects
-seen by the Software Engineering Institute. That is, as far as we can tell,
+seen by the Software Engineering Institute.
+That is, as far as we can tell,
 there are more small to medium-sized projects than otherwise.

### 2015-08-17 data.tex

@@
 Several projects extended for multiple incremental development cycles. 
-Median team size was 7 people, with a maximum of 40. 
-Many of the projects were e-commerce web portals or banking systems in the US, South Africa, and Mexico. 
+Median team size was 7 people, with a maximum of 40. See \fig{dist} for the total
+effort seen in those projects.
+
+As to problem domain,
+many of the projects were e-commerce web portals or banking systems in the US, South Africa, and Mexico. 
 There were  some  medical device projects in  the US, France, Japan, and Germany as well  as a commercial computer-aided design systems, and embedded systems.
@@
 %In this way, developers can acquire a more realistic understanding of their work behaviour.
+
+\begin{figure}
+\begin{center} \includegraphics[width=2.5in]{dom.pdf}
+\end{center}
+\caption{Distribution of 
+{\em effort} (which is
+{\em team size} times
+{\em days of work}). Median
+value = 271 days.}\label{fig:dist}
+\end{figure}
 

### 2015-08-18 lucas.tex
Changed mean/median to median/mode for survey responses and resorted the table according to these values. Close #23

@@
  &  & \multicolumn{2}{c|}{agreement} & \multicolumn{2}{c}{applicability} \\\cline{1-1} 
-Practitioner survey  & N & w & $\bar{x}$ & $\tilde{x}$ \\
+Practitioner survey  & N & med & mode & med & mode \\
 \hline 
-\textbf{Rqts errors are most expensive...} & 16 & 4.6 & 5 & 4.2 & 4 \\ 
-Process maturity improves output & 17 & 4.4 & 4 & 3.9 & 4 \\ 
-Most time is spent removing errors & 16 & 4.1 & 4 & 3.9 & 4 \\ 
-Missing reqts are hardest to fix & 17 & 4.1 & 4 & 4.0 & 4 \\
-Reuse increases prod. and qual. & 16 & 3.9 & 4 & 3.8 & 4 \\
-OO-programming reduces errors & 13 & 3.9 & 4 & 3.5 & 4 \\
-80-20 rule (defects to modules) & 12 & 3.8 & 4 & 4.0 & 4 \\
-Adding manpower to a late project & 15 & 3.7 & 4 & 3.7 & 4 \\
-Inspections can remove 90\% of defects & 18 & 3.3 & 4 & 3.5 & 4 \\
-Smaller changes have higher error density & 14 & 2.8 & 3 & 3.4 & 3.5 \\
-A developer is unsuited to test own code & 17 & 2.6 & 3 & 3.5 & 4\\\hline
+\textbf{Rqts errors are most expensive...} & 16 & 5 & 5 & 4 & 5 \\ 
+Inspections can remove 90\% of defects & 18 & 4 & 5 & 4 & 5 \\
+80-20 rule (defects to modules) & 12 & 4 & 5 & 4 & 5 \\
+Most time is spent removing errors & 16 & 4 & 4 & 4 & 5 \\ 
+Process maturity improves output & 17 & 4 & 4 & 4 & 4 \\ 
+Missing reqts are hardest to fix & 17 & 4 & 4 & 4 & 4 \\
+Reuse increases prod. and qual. & 16 & 4 & 4 & 4 & 4 \\
+OO-programming reduces errors & 13 & 4 & 4 & 4 & 3 \\
+Adding manpower to a late project & 15 & 4 & 4 & 4 & 4 \\
+Smaller changes have higher error density & 14 & 3 & 3 & 3.5 & 5 \\
+A developer is unsuited to test own code & 17 & 3 & 1 & 4 & 5\\\hline
 
 Researcher survey \\\hline 
-Process maturity improves output & 4 & 4.0 & 4 & 4.0 & 4 \\
-\textbf{Rqts errors are most expensive...} & 30 & 3.8 & 4 & 3.8 & 4   \\ 
-80-20 rule (defects to modules) & 6 & 3.7 & 4 & 3.5 & 4 \\
-\textbf{DelayedIssueEffect} & 30 & 3.6 & 4 & -- & --  \\ 
-Missing reqts are hardest to fix & 7 & 3.6 & 4 & 3.7 & 4 \\
-Adding manpower to a late project & 4 & 3.5 & 4 & 3.5 & 4 \\
-Inspections can remove 90\% of defects & 7 & 3.4 & 4 & 3.6 & 3 \\
-Reuse increases prod. and qual. & 6 & 3.3 & 4 & 3.0 & 4 \\
-OO-programming reduces errors & 6 & 3.2 & 4 & 2.7 & 3 \\
-Most time is spent removing errors & 6 & 2.8 & 3 & 3.2 & 4 \\ 
-Smaller changes have higher error density & 4 & 2.8 & 3 & 4.0 & 4 \\
-A developer is unsuited to test own code & 7 & 2.1 & 2 & 2.4 & 3
+Process maturity improves output & 4 & 4 & 4 & 4 & 5 \\
+\textbf{Rqts errors are most expensive...} & 30 & 4 & 4 & 4 & 4   \\ 
+\textbf{DelayedIssueEffect} & 30 & 4 & 4 & -- & --  \\
+Reuse increases prod. and qual. & 6 & 4 & 4 & 4 & 4 \\
+80-20 rule (defects to modules) & 6 & 4 & 4 & 4 & 3 \\
+Missing reqts are hardest to fix & 7 & 4 & 4 & 4 & 3 \\
+OO-programming reduces errors & 6 & 4 & 4 & 3 & 4 \\
+Inspections can remove 90\% of defects & 7 & 4 & 4 & 3 & 3 \\
+Adding manpower to a late project & 4 & 3 & 4 & 4 & 3 \\
+Most time is spent removing errors & 6 & 3 & 3 & 4 & 4 \\ 
+Smaller changes have higher error density & 4 & 3 & -- & 4 & 4 \\
+A developer is unsuited to test own code & 7 & 2 & 1 & 3 & 3
 \end{tabular}

### 2015-08-20 data.tex
Updates from ShareLaTeX

@@
 Since 2000, the SEI has been teaching and coaching TSP teams.
-One of the authors (Nichols) has been been mentoring software development teams and coaches around the world as they deploy TSP within their organizations since 2006.
+One of the authors (Nichols) has mentored software development teams and coaches around the world as they deploy TSP within their organizations since 2006.
 The  most recent completions were in 2014.
@@
 Note that, in that Figure~\ref{fig:waterfall},
-several  phases in which product is created have sub-phases of {\em review} and {\em inspect} to remove defects.
+several  phases in which the product is created have sub-phases of {\em review} and {\em inspect} to remove defects.
 In TSP  reviews,  individuals perform personal reviews of their work products prior to the peer review

### 2015-08-20 phaseDelay.tex
Updates from ShareLaTeX

@@
 \and%alignauthor 
-William Nichols, Forest Shull \\
+William Nichols, Forrest Shull \\
         \affaddr{Software Engineering Institute}\\
@@
 
-This paper checks for the delayed issue effect in
-171 software projects from around the world (2006 to 2014).
+This paper tests for the delayed issue effect in
+171 software projects conducted in the period from 2006--2014 from around the world.
 To the best of our knowledge,  this is the largest study
 yet published on this effect.
-What was observed was the  exact
-opposite of the  delayed issue effect. Specifically:
+In that sample, we found the exact
+opposite of the  delayed issue effect; i.e.
 the  time  to resolve 
@@
 release paradigm. Theisen et al. reports their experiences where Microsoft architects
-continually adjust  their modules in response to security issues~\cite{theisen15}.
+continually adjust  their modules in response to security issues~\cite{Theisen15}.
 Also, upgrades to their Windows operating system is moving from service patches (which occur rarely) to continuous deployment (so there will be no Windows 11- just a stream of continuous updates to what is currently called Windows 10~\cite{bright15}).
@@
 Next, a survey of industrial practitioners and researchers to document a widespread belief that delayed issues have a negative impact on projects.
-After that, we  analyze 171 software  projects developed in the period 2006 to 2014 to find  {\em no evidence} of the delayed issue effect:
+After that, we  analyze 171 software  projects developed in the period 2006--2014 to find  {\em no evidence} of the delayed issue effect:
 \bi
@@
 
-\subsubsection{Comment 2}
-Next, we comment on 
-requirements errors are the most expensive to fix 
-{\em when found during production but the cheapest to fix early in development}. 
+\newpage\subsubsection{Comment 2}
 Our data from these 171 project does not extend into post-delivery deployment. That said,
@@
 
-We checked for traces of this effect in 171 projects from the period 2006 to 2014.
+We checked for traces of this effect in 171 projects from the period 2006--2014.
 That data held no trace of the delayed issued effect.
@@
 \item The effect might be an historical relic. Evidence:
-the effect was first described in the era of punch card computing and non-interactive environments).
+the effect was first described in the era of punch card computing and non-interactive environments.
 \item The effect might have been intermittent (rather than some fundamental law of software). Evidence: we can  found nearly

### 2015-08-26 phaseDelay.tex
Updates from ShareLaTeX

@@
 This paper tests for the delayed issue effect in
-171 software projects conducted in the period from 2006--2014 from around the world.
+171 software projects conducted around the world in the period from 2006--2014.
 To the best of our knowledge,  this is the largest study
 yet published on this effect.
-In that sample, we found the exact
-opposite of the  delayed issue effect; i.e.
+In our sample, we found no evidence for the  delayed issue effect; i.e.
 the  time  to resolve 
-issues  in one phase  was rarely more than when  issues
-were resolved at a later phase.
+issues in a later phase was not consistently  more than  
+when  issues were resolved soon after their introduction.
 
@@
 It is standard practice
-in other fields such as medicine,
+in other fields, such as medicine,
 to continually revisit old conclusions~\cite{prasad13}.
@@
 For the moment, we describe it as follows:
-it is far easier to resolve issues earlier than later in the lifecycle.
+it is far easier to resolve issues earlier rather than later in the lifecycle.
  \fig{b81} shows an example of the delayed issue effect (relating
@@
 \end{quote}
-Glass also endorses this effect, asserting that``requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' is ``really just common sense''~\cite{glass02}.
+Glass also endorses this effect, asserting that ``requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' is ``really just common sense''~\cite{glass02}.
 Other researchers
@@
 fact.
-For example, what we call the delayed issued effect was listed as number one by Boehm and Basili in their "Top 10 list" of ``objective and quantitative data, relationships,
+For example, what we call the delayed issued effect was listed first by Boehm and Basili in their "Top 10 list" of ``objective and quantitative data, relationships,
 and predictive models that help
@@
 Also, we note that our development practices have changed in ways that could   mitigate the delayed issued effect.  Previously,
-software was large monolithic systems that were write once and maintain
-forever. Today, there are more architectures that
+software was large monolithic systems that were ''write once and maintain
+forever.'' Today, there are more architectures that
 support  extensive and faster changes to software. 
@@
 that endorse DIE. Also, using a survey conducted from this paper, we find  that 
-  DIE  appears up as the
+  DIE  appears as the
 single most strongly-held beliefs amongst commercial software engineers.
@@
 effect are either (1)~quite old (papers dating from last century);
-(2)~that just quote prior papers without presenting   new data; 
-(3)~or cite data sources that can no longer be
+(2)~quoting prior papers without presenting   new data; 
+(3)~or citing data sources that can no longer be
 confirmed.
@@
 learned from all available data to multiple models learned from clusters within the data~\cite{betten14,yang11,yang13,minku13,me12d,me11m,betta12,posnett11}.
-A repeated result in those studies in that the local models generated the better effort
+A repeated result in those studies is that the local models generated the better effort
 and defect predictions (better median results,
@@
 Even when the evidence for or against a treatment or intervention is clear, medical providers and patients may not accept it~\cite{aschwanden10}.
-Aschwanden warns that ``congitive biases''  such as  confirmation bias (the tendency to look for evidence that supports what you already know and to ignore the rest)  influence how we process information~\cite{aschwanden15}.
+Aschwanden warns that ``cognitive biases''  such as  confirmation bias (the tendency to look for evidence that supports what you already know and to ignore the rest)  influence how we process information~\cite{aschwanden15}.
 
@@
 all their future projects. They comment ``past experiences were taken into account without 
-much consideration for their context~\cite{passos11}.  
+much consideration for their context~\cite{passos11}.''  
 The results of J{\o}rgensen \& Gruschke~\cite{jorgensen09} concur with Passos et al. They report that 
@@
 To assess the prevalence of DIE,   we conducted a survey of software engineers. If our surveyed practitioners make management decisions based on their
-understanding of SE theory, then the DIE  may well inappropriately influence their decisions.
+understanding of SE theory, then the DIE  may well influence their decisions.
 
@@
 \subsubsection{Comment 1}
-Figure~\ref{fig:fix-phase-dist}
- data does not support   
+Figure~\ref{fig:raw}
+ data does not support that   
 {\em  requirements errors are the most expensive to fix}.
@@
 \ei
-For completeness, we ran an  
-A12 test comparing the raw data (and not just the medians of \fig{raw}) and found
- a non-small effect only for the difference {\em requirements}
-to {\em code}. But even though the effect was non-small, 
+For completeness, we ran an A12 test comparing the raw data for each phase, and found an effect size of ''medium'' or higher only for the difference between {\em requirements}
+and {\em code}. In this case, however, 
  the overall size of the difference
@@
 
-That said, Figure~\ref{fig:fix-phase-dist} does support the 
+That said, Figure~\ref{fig:raw} does support the 
 claim that early lifecycle errors are most expensive to fix.
@@
 
-Nevertheless,  it must be added that the scale up seen in 
-Figure~\ref{fig:fix-phase-dist} 
+Nevertheless, the scale up seen in 
+Figure~\ref{fig:raw} 
 is much less
-than suggested by the DIE literature. Returning to
+than that suggested by the DIE literature. Returning to
 the {\em before/unit test} result: 2.56 is far smaller
@@
 That evidence is as follows.
-Prior reports of an  exponential increase in difficulty in resolving issues {\em during
+Prior literature reports an  exponential increase in difficulty in resolving issues {\em during
 development}.
@@
 Any manager watching  these intra-development trends would be well-justified
-to anticipate that  the difficulty in resolving issues would will get much worse. Such managers
+to anticipate that  the difficulty in resolving issues  will get much worse. Such managers
 would therefore expect an exponentially harder time resolving issues during deployment.
 
-On the other hand, a remarkable feature of Figure~\ref{fig:fix-phase-dist} is the {\em absence}
+On the other hand, a remarkable feature of Figure~\ref{fig:raw} is the {\em absence}
 of an exponential delay effect during development.
-In our data, the greatest difficulty in resolving requirements issues was 1.31 seen in systems testing--
+In our data, the greatest increase in difficulty in resolving requirements issues was the 1.31 scale-up seen in systems testing--
 which is far smaller than the 37 to 250-fold increases reported in \fig{b81}, \fig{cost-to-fix},
@@
  \subsubsection{Comment 3}
- The above specific statements related just to requirements. Turning now
- to issues creased in all phases in Figure~\ref{fig:fix-phase-dist},  we see that
+ The above specific statements are related just to requirements. Turning now
+ to issues created in all phases in Figure~\ref{fig:raw},  we see that
  this data does not follow the exponential increase predicted
@@
 \ei
-In summary:
-\bi
-\item
-The delayed issue effect has been reported in around a dozen papers, at least half of which date back to the 1970s. 
-\item
-But there is no trace of that  effect   in hundreds of software projects built  since 2006.
-\ei
+%In summary:
+%\bi
+%\item
+%The delayed issue effect has been reported in around a dozen papers, at least half of which date back to the 1970s. 
+%\item
+%But there is no trace of that  effect   in hundreds of software projects built  since 2006.
+%\ei
 %For illustrative purposes, the relative cost to fix errors per lifecycle phase in the TSP project data is overlayed with Boehm's cost curve~\cite{Boehm81} as well as the CCPDS-R case study~\cite{Royce98} in \fig{cost-to-fix-tsp}.
@@
 Threats to external validity affect the generalization of findings. The most obvious threat to external validity is that TSP is not representative of most software development processes. Hence,
-For example, 
-one way to discount the above is that our sample is biased-- that while general software
+for example, 
+one way to discount the above argumentation is that our sample is biased-- that while general software
 systems suffer from the delayed issue effect, systems built using TSP are so much better controlled
@@
  \item
- Our 171 project contain examples of a wide variety
- of systems (e-commerce web portals and  banking systems) run in a variety of
+ Our 171 projects contain examples of a wide variety
+ of systems (ranging from e-commerce web portals to  banking systems) run in a variety of
  ways (agile or  waterfall or some combination of the two). Hence, rather
@@
 Another way to discount this result is to say we are only reporting on small to medium-size
-projects,
+projects
 (see
@@
 Most software is {\em not} like the software used on NASA deep-space satellites.
-As  agile methods grew more popular last decade; and as personnel computers grew more powerful;
+As  agile methods grew more popular last decade; and as personal computers grew more powerful;
 and as the libraries associated with particular languages grew more extensive; then smaller
@@
 our  171 projects represents $\frac{171}{212}=80$\% of the software projects
-seen by the Software Engineering Institute.
+seen by the Software Engineering Institute's TSP team.
 That is, as far as we can tell,
@@
 Our definition of {\em difficult to resolve} combines two concepts: time to change and cost to
-change. In the above we have used them interchangable, comparing our time to resolve data (from the 171 TSP projects) against the cost-to-fix results of Boehm et al. e.g. \fig{b81}.
+change. In the above we have used them interchangably, comparing our time to resolve data (from the 171 TSP projects) against the cost-to-fix results of Boehm et al. e.g. \fig{b81}.
 Is it valid to assume the equivalence of time and cost?
@@
 Evidence discussed in \cite{Shull02} suggests that low severity defects may exhibit a lower cost to change.
-Nonetheless, even ``small'' typographical errors have been known to cause enormous damage (e.g., the Mars climate Orbiter).
-
-Having documented the above issues, we assert that they are probably not major issues in the study.
+Nonetheless, even ``small'' errors have been known to cause enormous damage (e.g., the Mars Climate Orbiter).
+
+Having documented the above issues, we assert that they are unlikely to be major issues in the study.
 As to the first point, that some debugging may be more expensive than others, one of us (Nichols)
@@
 use of exorbitantly expensive tools or people on these projects. As to the second point (that
-some defects have an impact beyond the mere cost to fix them), we no evidence suggesting
+some defects have an impact beyond the mere cost to fix them), we have no evidence suggesting
 that the net effect of this concern over all our projects is non-zero.
@@
 increases the difficulty of completing that  resolution).
-Several prominent SE researchers state this effect is some fundamental law of software engineering~\cite{boehm01,mcconnell01,boehm01,glass02},
+Several prominent SE researchers state this effect is a fundamental law of software engineering~\cite{boehm01,mcconnell01,boehm01,glass02}.
 Based on a  survey  of both researchers and practitioners, we  found that
@@
 
-Our explanation for this lack-of-effect is five-fold:
+To be clear, we do not claim that this effect {\em never} holds in software projects; just that it cannot be assumed to {\em always} hold. Our explanation of the observed lack-of-effect is five-fold:
 \be

### 2015-12-31 phaseDelay4Emse.tex
Updates from ShareLaTeX

--- 2015-08-26 phaseDelay.tex
@@
-\documentclass{sig-alternate}
+\documentclass[smallcondesed]{svjour3}
 \usepackage{multirow}
@@
 
-\usepackage[shortlabels]{enumitem} 
+%\usepackage[shortlabels]{enumitem} 
 %\usepackage{times}
@@
 
+\pagenumbering{arabic} 
 \begin{document}
-
-\conferenceinfo{ICSE}{'16 Austin, Texas}
-\title{Live and Let Die? (Delayed Issues   not  Harder to Resolve)}
-\numberofauthors{3}
-\author{
+\date{}
+ \title{Are Delayed Issues Harder to Resolve?}
+
+ \author{Tim Menzies, William Nichols, Forrest Shull, Lucas Layman}
+
+\institute{
 %\alignauthor 
-Tim Menzies \\
-       \affaddr{Computer Science}\\
-       \affaddr{NC State University, USA}\\
-       \email{tim.menzies@gmail.com}
-\and%alignauthor 
-William Nichols, Forrest Shull \\
-        \affaddr{Software Engineering Institute}\\
-        \affaddr{Carnegie Mellon U, USA}\\
+Tim Menzies \at
+       CS, North Carolina State University, USA,
+                \email{tim.menzies@gmail.com} \
+\and 
+William Nichols, Forrest Shull \at 
+        Software Engineering Institute ,
+        Carnegie Mellon University, USA.
         \email{\{wrn,fjshull\}@sei.cmu.edu} 
-\and %\alignauthor 
-Lucas Layman \\
-       \affaddr{Fraunhofer CESE} \\
-       \affaddr{College Park, USA}\\ 
+\and  
+Lucas Layman \at
+        Fraunhofer CESE,  
+        College Park, USA,
        \email{llayman@fc-md.umd.edu}
@@
  as issues linger longer in a system,
- they become exponentially harder to resolve.
+ they become very much harder to resolve.
 This belief
@@
 yet published on this effect.
-In our sample, we found no evidence for the  delayed issue effect; i.e.
+We found no evidence for the  delayed issue effect; i.e.
 the  time  to resolve 
@@
 
-This  result begs the question: how many other long-held
-beliefs in software engineering are
-unsupported by current data? 
+This paper documents the above studies and explores reasons for this  mismatch between this widely-held
+belief and empirical data.
+In  summary, DIE is not some constant across all projects. Rather, is may be an historical relic that only occurrs intermittently 
+in  certain kinds of projects.  This is a significant result since it predicts that  new development
+processes that promise to faster retire more issues will have only a  limited return on investment. 
 \end{abstract}
 
-%XXXwhat it this is due to TSP- 
-% A category with the (minimum) three required fields
  \vspace{1mm}
@@
 \end{quote}
- \begin{figure}[!b]  
+ \begin{figure}[!t]  
 \begin{center}
-\includegraphics[angle=270,width=3in]{img/b81.pdf}
+\includegraphics[angle=270,width=4in]{img/b81.pdf}
  \end{center}
@@
 
-% \begin{figure}[!t]
-% \scriptsize
-% \begin{center}
-% \begin{tabular}{l@{~}|l@{~}|r@{~}r@{~}|r@{~}l}
-%           &        & \multicolumn{2}{c|}{Time to}  &\\
-%           &        & \multicolumn{2}{c|}{resolve} &\\
-% Injection & Removal& initial & now & \multicolumn{2}{l}{Scale up w.r.t initial} \\\hline
-% Phase $i$&Phase $i$ & 1 & 1 & 1 & **\\
-% Phase $i$ &Phase $i+1$ & 1 & 2& 2 & **** \\
-% Phase $i$ &Phase $i+2$ & 1 & 4& 4 &******** \\
-% Phase $i$ &Phase $i+3$ & 1 & 8 & 8 & ****************  
-%   \end{tabular}\end{center}
-%  \caption{Projects data consistent with
-%  the phase delay effect of \fig{curve1}.}\label{fig:curve1a}
-% \end{figure}
-
-% \begin{figure}[!t]
-% \scriptsize
-% \begin{center}
-% \begin{tabular}{l@{~}|l@{~}|r@{~}r@{~}|r@{~}l}
-%           &        & \multicolumn{2}{c|}{Time to}  &\\
-%           &        & \multicolumn{2}{c|}{resolve} &\\
-% Injection & Removal& initial & now & \multicolumn{2}{l}{Scale up w.r.t initial} \\\hline
-% Phase $i$&Phase $i$ & 1 & 1 & 1 & ** \\
-% Phase $i$ &Phase $i+1$ & 1 & 2& 2 & **** \\
-% Phase $i$ &Phase $i+2$ & 1 & 2& 2 & **** \\
-% Phase $i$ &Phase $i+3$ & 1 & 2 & 2 & **** \\
-%  Phase $i$  &Phase $i+4$ &1 & 2& 2 &**** 
-%   \end{tabular}\end{center}
-%  \caption{Projects data consistent with
-%  the phase delay effect of \fig{curve2}.}\label{fig:curve2a}
-% \end{figure}
 This paper calls into question all the above claims about the delayed
@@
  in the software.
-
-% the SE community needs to stop
-% justifying changes to SE
-% practices via the phase delay effect. Secondly,   myths like phase delay should not have lasted decades
-% in the literature without  critical reassessment.  As a community, we need to reflect more on our
-% research practices and on how (and when) we reflect and revisit the prominent hypotheses of our field.
 
@@
 \ee
-% \item 
-% Other work studies  found dozens of single
-% models that performed worse that {\em committees of models}, all with % different opinions,
-% that contribute to an {\em ensemble} prediction of software 
-% development effort~\cite{kocaguneli2012value,azhar13}.
+
 To be fair, 
@@
  conclusions and software projects  (see examples in~\cite{jorgensen09}).
-%% This leads to  detrimental
-%% situations where groups or 
-%% individuals working on the same project have conflicting beliefs, which they never debate
-%or refine or unify.
 
@@
 
-%\todo{\begin{itemize}
-%\item Given the Intro and Motivation, this section should focus on the influence of phase delay in software engineering milieux
-%\item Need to tone down the reliance on Beck.
-The important Beck point is that eliminating the cost-to-fix curve (resulting from phase delay) is one of the main goals of agile.
-His observations on new technologies are good too.
-%\item Need to cite other popular sources of this information, e.g., McConnell's "Ounce of Prevention" article~\cite{mcconnell01}.
-%\item Robert Glass's~\cite{glass02} book is very popular and contains an entry on this point
-%\item Google Scholar Citations for phase delay works Boehm88~\cite{boehm88} (673), Boehm \& Basili~\cite{boehm01} (761), 
-%\end{itemize}}
-
-%XXX clumsy link here
-%That said, where our results
-%diverge from Beck is that we have observed this lack of phase delay in  projects that are not agile. Those
-%projects are described below.
-
 \section{``DIE'' is a  widely-held belief}
@@
 
-\input{lucas}
+Our survey collected data on software engineers' views of commonly held software engineering ``laws''.
+One of the laws questioned is a specific form of our delayed issue effect:  ``requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' (from Glass~\cite{glass02} p.71 who references Boehm \& Basili~\cite{boehm01}).
+We abbreviate this law (illustrated in \fig{b81}) as RqtsErr.
+
+(Technical aside: we use the RqtsErr formulation since this issue typically needs no supportive explanatory
+text. If we had asked respondents about our more general term ``delayed issue
+effect'', we would have had to burden our respondents with extra explanations).
+
+The survey was conducted in two phases using Amazon's Mechanical Turk. The first phase was conducted only with professional software engineers solicited through the Mechanical Turk;
+participants were required to complete a pretest to verify their status as a professional or open source software developer and to confirm their knowledge of basic software engineering terminology and technology.
+The  second survey was conducted with Program Committee members of the ESEC/FSE 2015 and ICSE 2014  conferences solicited via email.
+
+The respondents answered questions on two scales: \newline
+\textbf{Agreement:} ``Based on your experience, do you agree that the statement above is correct?'' A Likert scale captured the agreement score from Strongly Disagree to Strongly Agree.
+A text box was provided to explain the answer.
+\newline
+\textbf{Applicability:} ``To the extent that you believe it, how widely do you think it applies among software development contexts?'' The possible answers were: 
+-1 meaning ``I don't know''; and 0 meaning ``this law does not apply at all'';
+ 1 meaning ``applies in a very narrow range of projects'';  2 meaning
+``rarely applies'';   3 meaning
+``occasionally applies'';  
+4 meaning ``very frequently applies''; and
+ 5 meaning ``always applies''.
+Respondents were required to explain the applicability score in a text box.
+
+In order to baseline the participant's answers, participants were presented with the RqtsErr law and others. For the purposes
+of this paper, the nature of the other laws other than RqtsErr are not relevant-- we
+only added them in as a way to calibrate responses to the RqtsErr question. All laws were drawn from \cite{glass02} and \cite{endres03}. 
+The PC member survey contained an additional question on ``In general, the longer errors are in the system (requirements errors, design errors, coding errors, etc.), the more expensive they are to fix'' (the Delayed Issue Effect).
+Responses were recorded using the agreement question Likert scale.
+
+Summary statistics for the agreement and applicability scores for the RqtsErr and DelayedIssueEffect laws are presented in Figure~\ref{fig:survey_results}.
+Responses whose Applicability response was ''I don't know'' are omitted from analysis.
+
+\begin{figure}[!ht] 
+\scriptsize 
+ \begin{center}
+\begin{tabular}{l|c|c|c|c|c}
+ &  & \multicolumn{2}{c|}{agreement} & \multicolumn{2}{c}{applicability} \\\cline{1-1} 
+Practitioner survey  & N & med & mode & med & mode \\
+\hline 
+\textbf{Rqts errors are most expensive...} & 16 & 5 & 5 & 4 & 5 \\ 
+Inspections can remove 90\% of defects & 18 & 4 & 5 & 4 & 5 \\
+80-20 rule (defects to modules) & 12 & 4 & 5 & 4 & 5 \\
+Most time is spent removing errors & 16 & 4 & 4 & 4 & 5 \\ 
+Process maturity improves output & 17 & 4 & 4 & 4 & 4 \\ 
+Missing reqts are hardest to fix & 17 & 4 & 4 & 4 & 4 \\
+Reuse increases prod. and qual. & 16 & 4 & 4 & 4 & 4 \\
+OO-programming reduces errors & 13 & 4 & 4 & 4 & 3 \\
+Adding manpower to a late project & 15 & 4 & 4 & 4 & 4 \\
+Smaller changes have higher error density & 14 & 3 & 3 & 3.5 & 5 \\
+A developer is unsuited to test own code & 17 & 3 & 1 & 4 & 5\\\hline
+
+Researcher survey \\\hline 
+Process maturity improves output & 4 & 4 & 4 & 4 & 5 \\
+\textbf{Rqts errors are most expensive...} & 30 & 4 & 4 & 4 & 4   \\ 
+\textbf{DelayedIssueEffect} & 30 & 4 & 4 & -- & --  \\
+Reuse increases prod. and qual. & 6 & 4 & 4 & 4 & 4 \\
+80-20 rule (defects to modules) & 6 & 4 & 4 & 4 & 3 \\
+Missing reqts are hardest to fix & 7 & 4 & 4 & 4 & 3 \\
+OO-programming reduces errors & 6 & 4 & 4 & 3 & 4 \\
+Inspections can remove 90\% of defects & 7 & 4 & 4 & 3 & 3 \\
+Adding manpower to a late project & 4 & 3 & 4 & 4 & 3 \\
+Most time is spent removing errors & 6 & 3 & 3 & 4 & 4 \\ 
+Smaller changes have higher error density & 4 & 3 & -- & 4 & 4 \\
+A developer is unsuited to test own code & 7 & 2 & 1 & 3 & 3
+\end{tabular} 
+ \end{center}
+\caption{Agreement and applicability of SE axioms.}
+\label{fig:survey_results}
+\end{figure}
+
+Both  practitioners and researchers  strongly believed in RqtsErr: In both sets of responses, RqtsErr received  scores higher than most
+other laws. Further, in the case of practitioners, this law was rated
+as the single most believed effect.
+
+Caveat: in the free reponse texts, we note that the researchers who disagreed with the law generally asserted that requirements change can be expensive, but that the effect depends on the process used (e.g., agile vs.
+waterfall) and the adaptability of the system architecture.
+
+Overall, the RqtsErr law was the most agreed upon and most applicable law of 11 surveyed amongst practitioners, and the second most agreed upon law amongst researchers. 
+This results strongly support that the notion that the  delayed issue effect   is a widely-held belief in software engineering.
 
@@
 
-\input{lucasRelatedWork}
-
-%\section{carterCharts}
-
-%\input{carter}
-
-\section{Delayed Issues not  Harder\newline to Resolve}
-
-\input{data}
+One reason that industrial practitioners and academics believe so strongly in the delayed issue effect is that it is often referenced
+in the SE literature. For example,
+we know that \fig{b81} has been presented to the gradaute SE class at North Carolina State University, without quarrel or critical comment, every year for the last decade.
+
+Yet when we look at the literature, the evidence for
+delayed issue effect is both very sparse and very old.
+The first data on the difficulties of resolving delayed issues as a function of lifecycle phase date back to large systems in the late 70s from IBM~\cite{Fagan76}, TRW~\cite{Boehm76}, GTE~\cite{Daly77}, and Bell Labs~\cite{Stephenson76} (Figure~\ref{fig:cost-to-fix}).
+These studies are most often cited by secondary sources regarding the delayed issue effect.
+We note that it is unclear from the text in ~\cite{Daly77} and \cite{Boehm76} if cost is defined in terms of effort, or in actual cost (i.e., labor, materiel, travel, etc).
+
+These studies suggest that the difficulty (in terms of effort) to find and fix an error monotonically increases with lifecycle phase.
+In 1990, Boehm~\cite{Boehm80} provides data suggesting that the cost-to-fix curve for small projects (from two student projects of 2000 deliverable source instructions) is flatter than for large projects (the dashed line of Figure~\ref{fig:cost-to-fix}).
+
+\begin{figure}[!b]
+\begin{center}\includegraphics[width=4in]{img/boehm_cost-to-fix.png}\end{center}
+ \caption{Historical cost-to-fix curve. From~\cite{Boehm81}.}\label{fig:cost-to-fix}
+ \end{figure}
+
+In the 40 years since these initial studies, few studies have explored the difficulty to resolve issues
+as a function of lifecycle phase. 
+Shull et al.~\cite{Shull02} conducted a literature survey and held a series of e-workshops with industry experts on fighting defects.
+Workshop participants from Toshiba and IBM reported cost-to-fix ratios between early lifecycle and post-delivery defects of 1:137 and 1:117 for large projects respectively~\cite{Shull02} -- but we note here that the raw data points were not provided (which makes confirming those numbers
+a difficult task).
+
+This was a common theme in the literature reviewed for this paper-- i.e.  that  it was no longer possible to access
+the data used to make prior conclusions.
+As an example of this, \fig{steck} shows one 2004 survey that reports an 
+exponential delayed issue  effect for 
+requirements issues in eight case studies. Note that we cannot verify
+{\em any} of those results since the links in the references of
+that  survey are all broken (``page not found'' errors).
+
+As to the research into agile methods, one goal of that approach
+is to reduce the difficulty associated with making changes later in the lifecycle~\cite{beck00}.
+Relatively little empirical data exists on this point.Elssamadisy and Schalliol~\cite{Elssamadisy02} anecdotally report on the growing, high cost of rework in a 50 person, three-year, 500KLOEC Extreme Programming project as the project grew in size and complexity-- but again we cannot access their
+exact figures.
+
+% We note that previous work focuses on cost-to-fix as a function of lifecycle phase irrespective of when the defect was injected, that is, previous work analyzes the cost to fix a defect found in test regardless of whether that defect was a requirements error or a coding mistake.
+To our knowledge, our research represents the first large-scale study of phase delay.
+
+ \begin{figure}
+{\small
+\begin{center}
+\begin{tabular}{r|rrrr}
+ case& \multicolumn{4}{c}{Phase Requirements Issue Found }  \\
+ study               &Requirements & Design & Code&  Test\\\hline
+1& 1 &3& 5& 37\\  
+2& 1  &    & 10 & 40 \\ 
+3& 1   &    & 10 &  40 \\
+4&  1  &   5 &      & 50  \\
+5&  1 &3& 7& 51 \\
+6& 1& 5 &33 &75  \\
+7&  1 & 20 & 45 & 250 \\  
+\end{tabular}
+\end{center}}
+\caption{Cost to resolve requirements issues, relative to resolving  during requirements. From ~\cite{steck04}.}\label{fig:steck}
+\end{figure}
+
+\begin{figure}[!t]
+\begin{center}
+ \includegraphics[width=4in]{img/clutterbuck.png} 
+ \end{center}
+ \caption{Cost of change from an agile case study. From~\cite{Clutterbuck09}.}\label{fig:clutterbuck}
+ \end{figure}
+
+ All in all, the empirical results of this section seem insufficient to justify
+ the strength of the beliefs documented in the previous section. Adding to our doubts are  several studies that report less-than exponential
+ increase in the difficulties associated with making the changes associated with delayed
+ issues. 
+Clutterbuck et al.~\cite{Clutterbuck09} studied a 5-month effort by a small-to-medium enterprise team developing a 71KLOEC web interface to a database application to implement 18 change requests-- see Figure~\ref{fig:clutterbuck} (note that these were for new and changed user requirements, not defects).
+Clutterbuck et al.
+found the cost of change to be relatively flat until the later phases, with much of the effort spent in analysis of the change requests~\cite{Clutterbuck09}.
+Note that in this study, the effort increased by only 60\% (see the start and end of the curve in Figure~\ref{fig:clutterbuck}).
+
+Another example of less-than exponential explosion in the difficulty associated with delayed issues comes from  Royce~\cite{Royce98}. He
+studied  a million-line, safety-critical missile defense system
+(see  Figure~\ref{fig:royce}).
+Design changes (including architecture changes) required approximately twice the effort of implementation and test changes, and the cost-to-fix in implementation and test phases increased slowly.
+Boehm~\cite{Boehm10} attributes this success to the  development process, which focused on removing architecture risk early in the development lifecycle.
+
+ Other examples that report the opposite of DIE are:
+  \bi
+  \item
+  Boehm~\cite{Boehm80} reported a flatter growth rate for small, non-critical projects.
+   \item 
+  Data from NASA's Johnson Space Flight Center, reported by Shull~\cite{Shull02}, found that the cost to find certain non-critical classes of defects was fairly constant across lifecycle phases (1.2 hours on average early in the project, versus 1.5 hours late in the project).
+
+\ei
+
+\begin{figure}
+\begin{center}\includegraphics[width=4in]{img/Royce98.png}\end{center}
+ \caption{Exception to the rule - The Royce study: cost-to-fix curve. From~\cite{Royce98}.}\label{fig:royce}
+ \end{figure}
+
+ One feature of the above results will become very important in the subsequent discussion. We note that that in all
+the results that report a delayed issue effect, those effects have began to manifest by the end of that project.
+
+\section{Delayed Issues not  Harder  to Resolve}
+
+The above analysis motivates a more detailed look at the delayed issued effect.  
+Accordingly, we examined 171 software projects conducted between 2006 and 2014.
+These projects took place at organizations in many countries and were conducted using  the Team Software Process (TSP$\textsuperscript{SM}$).
+
+Since 2000, the SEI has been teaching and coaching TSP teams.
+One of the authors (Nichols) has mentored software development teams and coaches around the world as they deploy TSP within their organizations since 2006.
+The  most recent completions were in 2014.
+The projects were mostly small to medium, with a median duration of 46 days and a maximum duration of 90 days in major increments. 
+Several projects extended for multiple incremental development cycles. 
+Median team size was 7 people, with a maximum of 40. See \fig{dist} for the total
+effort seen in those projects.
+
+As to problem domain,
+many of the projects were e-commerce web portals or banking systems in the US, South Africa, and Mexico. 
+There were  some  medical device projects in  the US, France, Japan, and Germany as well  as a commercial computer-aided design systems, and embedded systems.
+
+An anonymized version of that data is available in the PROMISE repository at openscience.us/repo.
+For confidentiality restrictions, we cannot offer 
+further details on these projects.
+
+\subsection{About TSP$\textsuperscript{SM}$}
+
+TSP is a software project management approach developed at the Software Engineering Institute (SEI) at Carnegie Mellon University~\cite{tsp00}.
+TSP is an extension of the Personal Software Process (PSP$\textsuperscript{SM}$) developed at the SEI by Watts Humphrey ~\cite{tsp00}.
+The data from these TSP projects were collected and stored in the Software Engineering Measured Process Repository (SEMPR) at the SEI.
+The Software Engineering Institute (SEI) at Carnegie Mellon University explores methods for software process improvement.
+
+%TSP helps developers via a set of measures that can be applied to managing tasks, quality, and schedule.
+Planning begins by quantifying goals, defining work practices, and estimating size and effort.
+Developers then use this information to make a detailed short term plan.
+As the developers perform project work, they use a tool such as the Process Dashboard to collect their time, effort, size, and schedule data.
+Every week, the team reviews their data to evaluate status, identify actual rates, determine if project goals for schedule, cost, and quality are being met.
+The team then uses this information to make necessary plan corrections.
+At the end of the project the coach and team perform a quantitative project post mortem.
+
+Common features of TSP projects include {\em planning}, {\em personal reviews}, {\em peer inspections}, and {\em coaching}.
+A TSP {\em coach} helps the team to plan and analyze performance. The coach is the only role authorized to submit project data to the SEI.
+Before reviewing data with the teams, therefore before submission, these coaches check the data for obvious errors.
+
+During {\em Planning}, developers estimate the size of work products and convert this to a total effort using historical rates.
+Time in specific tasks come from the  process phases and historical percent time in phase distributions.
+Defects are estimated using historical phase injection rates and phase removal yields.
+Coaches help the developers to compare estimates against actual results.
+In this way, developers acquire a more realistic understanding of their work behavior, performance, and schedule status.
+
+{\em Personal review} is a technique taken from the PSP and its use in TSP is unique.
+Developers follow a systematic process to remove defects by  examining their own work products using a checklist built from their personal defect profile.
+This personal review occurs after some product or part of a product is considered to be constructed and before peer reviews or test.
+
+%PSPSM teaches developers howto continually make and review their personnel estimates
+%about their day-to-day tasks, then compare those estimates against the actual development effort.
+%In this way, developers can acquire a more realistic understanding of their work behaviour.
+
+\begin{figure}
+\begin{center} 
+\includegraphics[width=2.5in]{dom.pdf}
+\end{center} 
+\caption{Distribution of 
+{\em effort} (which is
+{\em team size} times
+{\em days of work}). Median
+value = 271 days.}\label{fig:dist}
+\end{figure}
+
+{\em Peer inspection} is a  technique in
+traditional software engineering and is often called peer review.
+ Basili and Boehm   commented in 2001~\cite{boehm01} 
+that peer reviews can catch over half the defects introduced into a system.
+Peer inspection can be conducted on any artifact generated anywhere in the software
+lifecycle and can quickly be adapted to new kinds of artifacts.
+TSP peer reviews follow the Fagan style in which the reviewer uses a checklist composed of common team defects prior to a review team meeting.
+
+Overall, the   effort associated with adding TSP to a project is not onerous. McHale reports~\cite{mchale02}:
+\bi
+\item
+ The time spent  tracking time, defects, and tasks requires less then 3\% of a developer's time. Weekly team meetings  require at most an hour, which is
+only 2.5\% of a 40 hour work week. 
+\item
+Team launches and replans average about 1 day per month or 5\% planning overhead.
+\ei
+It is true that one staff member is needed as a ``coach'' to mentor the teams
+and certify and monitor that data collection.
+However, one of us (Nichols) has worked with dozens of TSP teams.
+He reports that one  trained coach can support 4 or 6 teams (depending upon team experience).
+
+\subsection{Data Integrity}
+
+A common property of real-world data sets is the presence
+of noisy entries (superfluous  or spurious data). 
+The level of noise can be quite high. As reported
+in \cite{shepperd12}, around
+10\% to 30\%
+of the records in the NASA MDP defect data sets are
+affected by noise.
+
+One reason to use the SEI data for the analysis of this paper is its remarkably low level of noise.
+Nichols et al.~\cite{shirai14}  report that
+the noise levels in the SEI TSP data are smaller than those seen
+in other data sets. They found in the SEI TSP data that:\bi 
+\item
+4\% of the data was incorrect (e.g. nulls, illegal formats);
+\item  2\% of the data has inconsistencies such as timestamps
+where the stop time was before the start time;
+\item 3\% of the data contained values that were not credible
+such as tasks listed in one day that took more than six hours for a single developer.
+\ei 
+One explanation for this low level of noise is the TSP process.
+One the guiding principles of TSP was that  people performing the work are  responsible for planning and tracking the work. That is,  all the data collected here was entered
+by local developers. This data was then checked by local coaches before being sent to the SEI
+databases. Coaches are certified by demonstrating competent use of the TSP process with the artifacts and data.
+The use of certified local coaches within each project increases the integrity of our data.
+
+\subsection{Data Details}
+\label{sec:data-collection}
+
+%Organizations using TSP agree to provide their project data to the SEI for use in research.
+In return the SEI agrees    that  data must not be traceable to its source.
+The data is collected at major project events; launch, interim checkpoints, and at project completion.
+Data includes project and site  characteristic summaries, surveys of team members, launch presentations, launch outbriefs, and baseline plans, final data from the project, and the project post mortem report.
+In practice, this data requirement has only been enforced for the purposes of certifying and reauthorizing TSP coaches who must submit data to maintain their authorization.
+Coaches are certified by demonstrating competent use of the TSP process with the artifacts and data not by the actual project results.
+Of the data submitted, only the data recorded using the Process Dashboard tool has been collected and aggregated for this research.
+ %The key views include a project results summary, effort logs, task logs, and defect logs.
+The data has not yet undergone additional screening.
+A summary of the data quality issues was reported in a previous work ~\cite{shirai14} (that summary is discussed further in our {\em Validity} section, below).
+
+Using tools provided by the SEI, developers kept very detailed logs of their daily activity. Our data includes  work start time, work end time,  delta
+work time, and interruption time. Software engineers are often
+interrupted by meetings, requests for technical help, reporting, and
+so forth. These events are recorded, in minutes, as interruption
+time.  
+In this paper, when we report ``time to resolve an
+issue,'' we show the difference between the start and end times
+of a work session, with any interruption time subtracted (the
+difference in times, minus the interruptions).
+
+As of November 2014, the SEI TSP database contained data from 212
+TSP projects. The projects completed between July 2006 and
+November 2014; they included 47 organizations and 843 people. 
+The database fact tables
+contain 268,726 time logs, 
+154,238 task logs,
+ 47,376 defect logs, 
+and 26,534 size logs. 
+After selecting defects from the data log and joining the data to the time log table  171 of these projects remained (the excluded projects had no or too few defects to use in this analysis).
+
+\begin{figure}[!b]  
+\begin{center}
+\includegraphics[width=2in]{img/waterfall-v3.png}  
+\end{center}
+\caption{Phases of our data.
+Abbreviations: 
+{\em Before}= before development; 
+{\em Reqts}	  = requirements; 
+{\em HLD}	  = high-level design; 
+{\em IntTest} = Integration testing (with code from others); 
+{\em SysTest} = system test (e.g. load stress tests); 
+{\em AcceptTest}  = acceptance testing (with users); 
+{\em review}        = private activity; 
+{\em inspect}        = group activity.}
+\label{fig:waterfall}
+\end{figure}
+
+In these logs, a  \emph{defect} is any change to a product, after its construction, that is necessary to make the product correct.
+A typographical error found in review is a defect.
+If that same defect is discovered while writing the code but before review, it is not considered to be a defect.
+SEI TSP defect types are:
+\begin{itemize}
+\item Environment: design, compile, test,  other support  problems
+\item Interface: procedure calls and reference, I/O, user format
+\item Data: structure, content
+\item Documentation: comments, messages
+\item Syntax: spelling, punctuation typos, instruction formats
+\item Function: logic, pointers, loops, recursion, computation  
+\item Checking: error messages, inadequate checks
+\item Build: change management, library, version control
+\item Assignment: package
+declaration, duplicate names, scope
+\item System: configuration, timing, memory
+\end{itemize}
+The defect logs in that data  include the time and date a defect was discovered, the phase in which that defect was injected, the phase in which it was removed, the time (in minutes) required to find and fix the defect, and the categorical type.
+
+The phases logged by our data are shown in \fig{waterfall}.
+ Although the representation suggests a waterfall model, the SEI experience is that all real implementations of any size follow a spiral approach with many team performing the work in iterative and/or incremental development cycles.
+
+One special feature of  \fig{waterfall} is
+the {\em before} phase, in which the TSP team assures that management has clearly identified cost, schedule, and scope goals appropriate to the upcoming development activities, often including a conceptual model ~\cite{Humphrey:2005}.
+For example an architecture team must have sufficient requirements to reason about, prototype, and specify an architecture~\cite{Bachmann13} while a coding only team within a larger project would have more precisely defined requirements and high level design.
+
+%Within a project increment, multiple features or components may be developed and incremented.
+TSP is compatible %with agile and encourages iterative and incremental development.
+Nonetheless, the specific strategy and cycle %duration is a project decision.
+TSP does, however, strongly encourage 1) constructing units of sufficient size that %%measurement is practicable, and 2) separating the construction from appraisal and test activities.
+This %effectively highlights the separation of construction from rework activities and aids the apportionment of defects %to those found in appraisal activities (reviews and inspections) and those found through failure (test).
+The %distinct construction activities (requirements, high and detailed design, and code) were chosen to help teams %analyze the effectiveness and efficiency of their practices through analysis of the defect phase origin, type, fix %effort, and phase of discovery.
+
+Note that, in that Figure~\ref{fig:waterfall},
+several  phases in which the product is created have sub-phases of {\em review} and {\em inspect} to remove defects.
+In TSP  reviews,  individuals perform personal reviews of their work products prior to the peer review
+(which TSP calls the inspection).
+Also, Figure~\ref{fig:waterfall} divides  testing     as follows.
+Developers perform unit test prior to code complete.
+After code complete a standard phase is integration, which combines program units into a workable system ready for system test.
+Integration,  system test, and acceptance test are often performed by another group.
 
@@
 
-%\todo{
-%\bi
-%    \item What is the cost-to-fix curve in TSP data? Check against H1 and compare to past literature.
-%    \item What is the effect of phase delay? Check against H2.
-%    \item \fig{cost-to-fix-tsp} is great.
-really important.
-its from  different decades, right?
-so can we add text saying that the general pattern is that the (approx) 15 years between each curve tends to a flattening.
-not, perhaps, due to process but to better build environs.
-%\ei
-%}
-
 \begin{figure}[!t]
 
-%\hspace{-0.7cm}
 \begin{center}
-\includegraphics[height=2in]{img/fix-phase-dist.pdf}
+\hspace{-1.5in}\includegraphics[height=1.5in]{img/fix-phase-dist.pdf}
 \end{center} 
-%~\\
+\vspace{2 cm}
 
@@
 \begin{center}
-\begin{tabular}{r@{~}|r@{~}r@{~}r|@{~}l@{~}l|@{~}r@{~}l}  % I attach your cursor. thrust, parry, yo!
-  \multicolumn{1}{c}{~} &\multicolumn{3}{c}{Percentiles}&\multicolumn{2}{l}{~~~~~~Phase}\\%\cline{2-6}
+\begin{tabular}{r@{~}|r@{~}r@{~}r|@{~}l@{~}l|@{~}r@{~}l}   
+  \multicolumn{1}{c}{~} &\multicolumn{3}{c}{Percentiles}&\multicolumn{2}{l}{~~~~~~Phase}\\
 N & 25th & 50th & 75th &   injected & removed   & \multicolumn{2}{l}{Scale up w.r.t. to first phase}\\\hline
@@
 
+\newpage
 \subsection{Observations from 171 Projects}
@@
 \bi
-\item There is much ``up and down'' in the chart. 
-\item If we separate out just the {\em Inspect} or {\em Review} data, we can see that there is some increase in {\em Design} and {\em Code}.
-However, that effect could be a linear or be just noise.
-In either case,
-it does not look like exponential.
-\item The time in final testing is particularly low suggesting that few defects survived into testing
+\item There is much ``up and down'' in the chart, but no steady exponential growth such
+as that see in, for example,  \fig{b81}.
+\item The time in acceptance  testing is particularly low suggesting that few defects survived into testing
 (and hence, there were few instances where code required massive rework).
@@
 
-%\begin{figure}[!t]
-%\renewcommand{\baselinestretch}{0.7}
-%\scriptsize
-%\begin{center}
-%\begin{tabular}{l@{~~}|l@{~}|r@{~}|r@{~}r@{~}|r@{~}l}
-%          \multicolumn{2}{c}{~}                 &  &\multicolumn{2}{c|}{99th }\\
-%           \multicolumn{2}{c}{~}                 &  &\multicolumn{2}{c|}{percentile }\\
-%  Injection&   Removal& $n$ & initial & now & \multicolumn{2}{l}{Scale up, w.r.t. initial}
-%\input{deltasBigBig}
-%\end{tabular}
-%\end{center}
-%%\caption{99th percentile scale ups.}
-%\label{fig:scale99}
-%\end{figure}
-
-%\begin{figure}[!t]
-%\renewcommand{\baselinestretch}{0.7}
-%\scriptsize
-%\begin{center}
-%\begin{tabular}{l@{~~}|l@{~}|r@{~}|r@{~}r@{~}|r@{~}l}
- %          \multicolumn{2}{c}{~}                 &  &\multicolumn{2}{c|}{mean}\\
- % Injection&   Removal& $n$ & initial & now & \multicolumn{2}{l}{Scale up, w.r.t. initial}
-%\input{deltasMean}
-%\end{tabular}
-%\end{center}
-%\caption{Mean  scale ups.}
-%\label{fig:scale90}
-%\end{figure}
+\begin{figure}[!t]
+\renewcommand{\baselinestretch}{0.7}
+\scriptsize
+\begin{center}
+\begin{tabular}{l@{~~}|l@{~}|r@{~}|r@{~}r@{~}|rl}
+          \multicolumn{2}{c}{~}                 &  &\multicolumn{2}{c|}{99th }\\
+           \multicolumn{2}{c}{~}                 &  &\multicolumn{2}{c|}{percentile }\\
+ Injection&   Removal& $n$ & initial & now & \multicolumn{2}{l}{Scale up, w.r.t. initial}
+\\\hline
+
+B4dev&Code&57&257&257&1&\rule{2mm}{2mm} \\
+B4dev&CodeInspect&72&257&153&0.60&\rule{2mm}{2mm} \\
+B4dev&Test&165&257&282&1.093&\rule{2mm}{2mm} \\
+B4dev&IntTest&50&257&365&1.42&\rule{2mm}{2mm} \\
+B4dev&SysTest&42&257&151&0.59&\rule{2mm}{2mm} \\
+
+\\\hline
+
+Design&Design&37&600&600&1&\rule{2mm}{2mm} \\
+Design&DesignInspect&455&600&277&0.46&\rule{0mm}{2mm} \\
+Design&Code&218&600&240&0.4&\rule{0mm}{2mm} \\
+Design&CodeInspect&166&600&163&0.27&\rule{0mm}{2mm} \\
+Design&Test&542&600&652&1.09&\rule{2mm}{2mm} \\
+Design&IntTest&67&600&277&0.46&\rule{0mm}{2mm} \\
+Design&SysTest&92&600&480&0.8&\rule{2mm}{2mm} \\
+
+\\\hline
+
+Code&Code&126&600&600&1&\rule{2mm}{2mm} \\
+Code&CodeInspect&459&600&243&0.41&\rule{0mm}{2mm} \\
+Code&Test&461&600&365&0.61&\rule{2mm}{2mm} \\
+Code&IntTest&348&600&7200&12&\rule{24mm}{2mm} \\
+Code&SysTest&230&600&290&0.48&\rule{0mm}{2mm} \\
+Code&AcceptTest&71&600&600&1&\rule{2mm}{2mm} \\
+\end{tabular}
+\end{center}
+\caption{99th percentile scale ups.}
+\label{fig:scale99}
+\end{figure}
 
@@
 Having documented the above issues, we assert that they are unlikely to be major issues in the study.
-As to the first point, that some debugging may be more expensive than others, one of us (Nichols)
+One of us (Nichols)
 was closely associated with many of the projects in our sample. He is unaware of any frequent
-use of exorbitantly expensive tools or people on these projects. As to the second point (that
-some defects have an impact beyond the mere cost to fix them), we have no evidence suggesting
-that the net effect of this concern over all our projects is non-zero.
+use of exorbitantly expensive tools or people on these projects.
 
@@
 
-%The definitions of defect, cost-to-fix, and phase delay are provided in earlier sections.
-The %source data and analysis scripts used to calculate the defect counts, cost-to-fix, and phase delay %from the raw materials can be found at \url{https://github.com/txt/txt.github.io/tree/master/tsp}.
-
-%\section{Future Work}
-%There is another kind of delayed issues  effect, not discussed above. Even if software is readily updatable, there   may be a ``user  inertia''   amongst   humans using that software. 
-%To understand the 
-%Changes to  software that is used by a large population can be confusing (particularly for the sick or elderly). 
-%Consider the 23 million veterans working with  the U.S.  Veterans 
-%Administration (VA) service. Some of those veterans are dependent on health care or pensions from the VA. Those veterans use the VA's web interfaces
-%to  (e.g.) find out
-%when their next cheque will arrive. Before we can update the VA web software, we would
-%need to  a transition plan that helps users   move from the older to the newer system.
-This is an example of what might be called a ``user phase delay''; i.e.
-changes to software are more complex to process, particularly if some users have styled
-%their day-to-day work flow around the services offered by the older software.  
-%
-%This issue of ``user phase delay'' is not addressed by our data, nor by the prior studies on phase delay %discussed above.
-Hence, this issues   is out of scope for this paper since the expectation that ``older %bugs cost more to fix than newer bugs'' was not formed around
-%user issues.
-
-%That said,  
-%we think that ``user phase delay'' will become a bigger issue in the future when more people are using more software as part of their day to day life.
-We even envision a time when users demand fewer updates, just to offer some stability in their life.
-
  \section{Conclusion}
- %\todo{
- %\bi
- %   \item Summarize data tested - survey + TSP data
-%    \item reiterate hypothesis testing results
-%    \item Restate the motivation: widely-held beliefs need to be revisited. 
-%    \item We now have evidence, from 5 different sources, that phase delay does not seem to hold under certain conditions (although our study is by far the biggest in terms of data points).
-%    \item phase delay may still apply in some (many?) projects, but clearly it does not in all. This has huge implications for, e.g., cost modeling, effort prediction.
-% \ei
-% }
 
@@
 \ee
- %One hypothesized explanation that can be ruled out is that the lack of a phase delay effect is due to using the TSP method. 
+
 Our results beg the question: why did an idea, with so little support, become so widespread in the software engineering literature?
@@
 Progress in the domain of software analytics has made such periodic checks more cost-effective and feasible, and we argue that an examination of local behaviors (rather than simply accepting global heuristics) can be of significant benefit.
-
-% \section{Related Work}
-
-% \begin{figure}
-% \begin{center}
-% \scriptsize\begin{tabular}{|l@{~:~}l|}\hline
-% Bug Prediction Dataset &http://bug.inf.usi.ch \\
-% Eclipse Bug Data &http://goo.gl/tYKahN \\
-% FLOSSMetrics& http://flossmetrics.org \\
-% FLOSSMole &http://flossmole.org \\
-% IBSBSG& http://www.isbsg.org \\
-% ohloh& http://www.openhub.net \\
-% PROMISE &http://promisedata.googlecode.com \\
-% Qualitas Corpus &http://qualitascorpus.com \\
-% Software Artifact Repository &http://sir.unl.edu \\
-% SourceForge Research Data &http://zerlot.cse.nd.edu \\
-% Sourcerer Project &http://sourcerer.ics.uci.edu \\
-% Tukutuku &http://www.metriq.biz/tukutuku \\
-% Ultimate Debian Database &http://udd.debian.org\\\hline
-% \end{tabular}
-% \end{center}
-% \caption{Some repositories of software engineering data.}\label{fig:sedata}
-% \end{figure}
-% The data collected in the SEI TSP databases contains 
-% extensive process details
-% including a detailed phased breakdown showing what happened at what
-% phases of the lifecycle. This makes this data somewhat different to the standard
-%  publicly available
-% software projects data sets currently available to 
-% SE researchers. Most of the \fig{sedata} data sets have software product information
-% such as full source code, or summaries of static features.
-% A subset of that data contain issue or defect reports and/or
-% the time taken to build these systems.  We are unaware
-% of any of these having detailed phase-by-phase breakdowns of project data.
 
@@
 \vspace*{0.5mm} 
-\scriptsize
 \balance
@@
 \end{document}
-
-% One way to address the absence of general laws is  {\em local learning}.
-% Even if general laws
-% do not exist, then {\em there exists general methods for finding the best local laws}.
-% For example, one of us (Nichols) works extensively with software projects to help
-% them tune their local software process in order to better fit their local needs.XXX.
-
-%  Recent studies report that better predictors of the properties of
-% software projects (effort and defects) can be generated by first clustering
-% (a.k.a. stratifying or contextualizing or localizing) the data then learning
-% different models for each cluster~\cite{posnett11,betten14,betta12,yang11,yang13,minku13}.
-% Those {\em local learning} approach finds better predictions with lower variance than
-% models learned from all the unclustered data~\cite{me12d,me11m}.
-
-% The problem with the local learning is that they it can generate conclusions
-% that conflict with strongly-held beliefs of the humans members of the software development team.
-% It is an open issue how to handle those conflicts since it is unwise to
-% always reject the conclusions of local learning  or the software developers.
-% The local learners may be wrong due to incorrect assumptions by the data scientists who configured the learners~\cite{me11e,shull02}.
-% On the other hand,  the humans may be wrong due to the cognitive biases described above.
-% Experienced data scientists use a cyclic approach where
-% (a)~the collected data; and (b)~the learning methods; and (c) the goals of the inquiry are matured  using
-% insight offered by humans as they study the
-% feedback generated by the learners~\cite{Fayyad96,me11e}.
-
-% the learners' conclusions are discussed with the team in or
-
-% (which may be erroneous
-% due to incorrect assumptions
-
-% We come to this work since  recent results on {\em locality} by Menzies and others
-
-% XXXX In those 171 projects, we observed that it is
-%  fastest to fix issues within the same phase $i$ as when they are generated. But in a result that contradicts
-%  the phase delay effect, once an issue ``escapes'' phase $i$
-%  it so no more expensive to resolve an issue in phase $i+1$ than
-%  phase $i+2, i+3$, etc. Further, that increased effort may be quite modest.
-%  That is, contrary to established wisdom, once an  issue ``escapes'' a phase, there is no
-%  need to  rapidly retire that issue as soon as possible. 
-% Our conclusion discusses the implications for our field, and how we might better
-% propagated and monitor the wisdom of our field.
-
-% \begin{figure}
-% \begin{center}
-% \begin{tabular}{rrl}
-% year& \# issues&\\\hline
-% 2006 &  44 &\\
-% 2007 &  34 &\\
-% 2008&  288 &\rule{1mm}{2mm}\\
-% 2009&  846 &\rule{3mm}{2mm}\\
-% 2010& 1007 &\rule{3mm}{2mm}\\
-% 2011& 3273 &\rule{10mm}{2mm}\\
-% 2012&18102 &\rule{45mm}{2mm}\\
-% 2013&20336 &\rule{50mm}{2mm}\\
-% 2014& 3307 & \rule{10mm}{2mm}\\\cline{1-2}
-% Total:&47228
-% \end{tabular}
-% \end{center}
-% \caption{This paper studies 47,228 issues recorded 2006 to 2014.}\label{fig:years}
-% \end{figure}
 

### 2015-12-31 phaseDelay4Emse.tex
Updates from ShareLaTeX

@@
 \item
-The {\em delayed issue effect}:   it is {\em exponentially}  more {\em difficult} to resolve  issues in a software project, the {\em longer} they remain.
+The {\em delayed issue effect}:   it is {\em very much}  more {\em difficult} to resolve  issues in a software project, the {\em longer} they remain.
 \item
@@
 \item
-A measure collected in phase ${1,.,i,..j}$ is 
-{\em exponential} when  that
+We say that a measure collected in phase ${1,.,i,..j}$ is 
+{\em very much} more when  that
    measure at phase $j$   
@@
  171 software  projects developed, we will show
-{\em no evidence}  that, during development, delayed issues were exponentially harder to resolve the longer they were left
+{\em no evidence}  that, during development, delayed issues were very much more harder to resolve the longer they were left
  in the software.
@@
 As an example of this, \fig{steck} shows one 2004 survey that reports an 
-exponential delayed issue  effect for 
+greatly increased delayed issue  effect for 
 requirements issues in eight case studies. Note that we cannot verify
@@
  All in all, the empirical results of this section seem insufficient to justify
- the strength of the beliefs documented in the previous section. Adding to our doubts are  several studies that report less-than exponential
+ the strength of the beliefs documented in the previous section. Adding to our doubts are  several studies that report less-than large
  increase in the difficulties associated with making the changes associated with delayed
@@
 
-Another example of less-than exponential explosion in the difficulty associated with delayed issues comes from  Royce~\cite{Royce98}. He
+Another example of less-than large increase in the difficulty associated with delayed issues comes from  Royce~\cite{Royce98}. He
 studied  a million-line, safety-critical missile defense system
@@
 
- One feature of the above results will become very important in the subsequent discussion. We note that that in all
-the results that report a delayed issue effect, those effects have began to manifest by the end of that project.
+\subseection{Early Onset of the DIE Effect}\label{sect:earlyonset}
+
+One feature of the above results will become very important in the subsequent discussion.  
+All the   literature described above that reports the onset of  DIE  {\em prior} to delivery.
+\bi
+\item  
+\fig{b81} reports a 40-fold increase in effort requirements to acceptance testing
+\item
+ \fig{cost-to-fix} reports a 100-fold increase
+(for the larger projects) before the code is delivered 
+\item
+The projects surveyed in \fig{steck} report changes of a  
+\{37,40,40,50,51,70,75,250\}-fold increase in pre-deployment period when the software was being developed.
+\ei
+Any manager noticing this  early onset of DIE (prior to delivery, during the initial development)
+would be well-justified
+in believing that  the difficulty in resolving issues  will get much worse. Such managers
+would therefore expect DIE to have a marked effect, post-deployment.
+
+We make this point since,  in the new project data presented below, there is no early onset of DIE. 
+That is:
+\bi
+\item There is no evidence of a growing problem prior to delivery that delayed issues are becoming harder to manage
+\item  Hence, there there is less evidence that DIE is a trend that will significantly and negatively effect the product,
+post-delivery.
+\ei
 
@@
 \begin{figure}[!t] 
- \renewcommand{\baselinestretch}{0.7}
+% \renewcommand{\baselinestretch}{0.8}
  \scriptsize
 \begin{center}
-\begin{tabular}{r@{~}|r@{~}r@{~}r|@{~}l@{~}l|@{~}r@{~}l}   
-  \multicolumn{1}{c}{~} &\multicolumn{3}{c}{Percentiles}&\multicolumn{2}{l}{~~~~~~Phase}\\
+\begin{tabular}{r|rrr|ll|rl}   
+  \multicolumn{1}{c}{~} &\multicolumn{3}{c}{Percentiles}&\multicolumn{2}{c}{Phase}\\\cline{2-6}
 N & 25th & 50th & 75th &   injected & removed   & \multicolumn{2}{l}{Scale up w.r.t. to first phase}\\\hline
-\\
  48&   14&   30&   51&Before&DesignInspect &1.00 & **********  \\
@@
 127&   30&   65&  128&Before&IntTest &2.16 & **********************  \\
- 92&   24&   56&   89&Before&SysTest &1.86 & *******************  \\
-
-\\
+ 92&   24&   56&   89&Before&SysTest &1.86 & *******************  \\\hline
+
  56&   14&   35&   57&Planning&ReqtsReview &1.00 & **********  \\
  46&   11&   24&   39&Planning&DesignInspect &0.68 & *******  \\
- 32&   18&   33&   84&Planning&UnitTest &0.94 & **********  \\
-
-\\
+ 32&   18&   33&   84&Planning&UnitTest &0.94 & **********  \\\hline
+
 184&   45&   92&  155&Reqts&ReqtsReview &1.00 & **********  \\
@@
 100&   33&   67&  120&Reqts&IntTest &0.72 & ********  \\
- 78&   28&  122&  237&Reqts&SysTest &1.31 & **************  \\
-\\
+ 78&   28&  122&  237&Reqts&SysTest &1.31 & **************  \\\hline 
 210&   51&  103&  166&Design&DesignReview &1.01 & ***********  \\
@@
 163&   40&   80&  161&Design&SysTest &0.79 & ********  \\
- 80&   21&   45&   73&Design&AcceptTest &0.44 & *****  \\
-
-\\
+ 80&   21&   45&   73&Design&AcceptTest &0.44 & *****  \\\hline
+
 235&   57&  115&  177&Code&CodeReview &1.00 &  **********  \\
@@
 
+\begin{figure}[!t] 
+ %\renewcommand{\baselinestretch}{0.8}
+ \scriptsize
+\begin{center}
+\begin{tabular}{r|r|ll|rl}   
+  \multicolumn{1}{c}{~} &\multicolumn{1}{c}{Percentiles}&\multicolumn{2}{c}{Phase}\\\cline{2-4}
+N  & 99th &   injected & removed   & \multicolumn{2}{l}{Scale up w.r.t. to first phase}\\\hline
+
+57&257&Before&Code&1.00&*********\\
+72&153&Before&CodeInspect&0.60&****** \\
+165&282&Before&Test&1.01&**********\\
+50&365&Before&IntTest&1.42&************** \\
+42&151&Before&SysTest&0.59&****** \\\hline 
+37&600&Design&Design&1.00&********** \\
+455&277&Design&DesignInspect&0.46&***** \\
+218&240&Design&Code&0.40&**** \\
+166&163&Design&CodeInspect&0.27&*** \\
+542&652&Design&Test&1.09&***********\\
+67&277&Design&IntTest&0.46&***** \\
+92&480&Design&SysTest&0.80&******** \\\hline 
+126&600&Code&Code&1.0&**********\\
+459&243&Code&CodeInspect&0.4&**** \\
+461&365&Code&Test&0.6&****** \\
+230&290&Code&SysTest&0.5&***** \\
+71&600&Code&AcceptTest&1.0&*********\\
+348&7200&Code&IntTest&12.0&************************************\\
+  &    &    &       &    & ************************************\\
+  &    &    &       &    & ************************************\\
+
+\end{tabular}
+\end{center}
+\caption{99th percentile scale ups.}
+\label{fig:scale99}
+\end{figure}
+
 \newpage
@@
 If  this data exhibited a delayed issue effect,  we would expect those bars in each sub-division to exhibit an
-exponential shape.
+very large increased.
 
@@
 \item
-Somewhat shorter, but still not exponentially smaller,  than the median times see after {\em requirements} in {\em design}
+Somewhat shorter, but still not very much larger,  than the median times see after {\em requirements} in {\em design}
 (103,101, 79,66, 119,88, 80,45) and in {\em code} (65,115, 113,125, 95,133). 
@@
 than the scale ups seen in Figure~1. In fact, nowhere in 
-our results do we see any exponential increase.
+our results do we see the kind of very large increases reported in the papers
+documenting DIE.
 
@@
 Our data from these 171 project does not extend into post-delivery deployment. That said,
-there is enough evidence to raise doubts that our projects will exhibit exponentially more
-difficult issues after deployment.
-
-That evidence is as follows.
-Prior literature reports an  exponential increase in difficulty in resolving issues {\em during
-development}.
-For example:
-\bi
-\item  From requirements to acceptance testing
-\fig{b81} reports a 40-fold increase;
-\item
- \fig{cost-to-fix} reports a 100-fold increase
-(for the larger projects).
-\item
-The projects surveyed in \fig{steck} report changes of a  \newline
-\{37,40,40,50,51,70,75,250\}-fold increase in pre-deployment period when the software was being developed.
-\ei
-Any manager watching  these intra-development trends would be well-justified
-to anticipate that  the difficulty in resolving issues  will get much worse. Such managers
-would therefore expect an exponentially harder time resolving issues during deployment.
-
-On the other hand, a remarkable feature of Figure~\ref{fig:raw} is the {\em absence}
-of an exponential delay effect during development.
+there is enough evidence to raise doubts that our projects will exhibit very much more
+difficulty in resolving issues after deployment.
+
+As mentioned in \tion{earlyonset}, every other
+paper reporting DIE also reported  early onset of DIE
+within the current development. A remarkable feature of Figure~\ref{fig:raw} is the {\em absence}
+of a very large DIE   effect during development.
 In our data, the greatest increase in difficulty in resolving requirements issues was the 1.31 scale-up seen in systems testing--
 which is far smaller than the 37 to 250-fold increases reported in \fig{b81}, \fig{cost-to-fix},
-and \fig{steck}.  That is, from our  data, we would not anticipate exponentially greater difficulty in resolving requirements issues after
+and \fig{steck}.
+
+That is, from our  data, we would not anticipate very much larger  difficulty in resolving requirements issues after
 deployment. 
@@
  to issues created in all phases in Figure~\ref{fig:raw},  we see that
- this data does not follow the exponential increase predicted
+ this data does not exhibit the very large   increases predicted
 by the delayed issue effect:
 \bi
-\item There is much ``up and down'' in the chart, but no steady exponential growth such
+\item There is much ``up and down'' in the chart, but no steady very large growth such
 as that see in, for example,  \fig{b81}.
@@
 
-\begin{figure}[!t]
-\renewcommand{\baselinestretch}{0.7}
-\scriptsize
-\begin{center}
-\begin{tabular}{l@{~~}|l@{~}|r@{~}|r@{~}r@{~}|rl}
-          \multicolumn{2}{c}{~}                 &  &\multicolumn{2}{c|}{99th }\\
-           \multicolumn{2}{c}{~}                 &  &\multicolumn{2}{c|}{percentile }\\
- Injection&   Removal& $n$ & initial & now & \multicolumn{2}{l}{Scale up, w.r.t. initial}
-\\\hline
-
-B4dev&Code&57&257&257&1&\rule{2mm}{2mm} \\
-B4dev&CodeInspect&72&257&153&0.60&\rule{2mm}{2mm} \\
-B4dev&Test&165&257&282&1.093&\rule{2mm}{2mm} \\
-B4dev&IntTest&50&257&365&1.42&\rule{2mm}{2mm} \\
-B4dev&SysTest&42&257&151&0.59&\rule{2mm}{2mm} \\
-
-\\\hline
-
-Design&Design&37&600&600&1&\rule{2mm}{2mm} \\
-Design&DesignInspect&455&600&277&0.46&\rule{0mm}{2mm} \\
-Design&Code&218&600&240&0.4&\rule{0mm}{2mm} \\
-Design&CodeInspect&166&600&163&0.27&\rule{0mm}{2mm} \\
-Design&Test&542&600&652&1.09&\rule{2mm}{2mm} \\
-Design&IntTest&67&600&277&0.46&\rule{0mm}{2mm} \\
-Design&SysTest&92&600&480&0.8&\rule{2mm}{2mm} \\
-
-\\\hline
-
-Code&Code&126&600&600&1&\rule{2mm}{2mm} \\
-Code&CodeInspect&459&600&243&0.41&\rule{0mm}{2mm} \\
-Code&Test&461&600&365&0.61&\rule{2mm}{2mm} \\
-Code&IntTest&348&600&7200&12&\rule{24mm}{2mm} \\
-Code&SysTest&230&600&290&0.48&\rule{0mm}{2mm} \\
-Code&AcceptTest&71&600&600&1&\rule{2mm}{2mm} \\
-\end{tabular}
-\end{center}
-\caption{99th percentile scale ups.}
-\label{fig:scale99}
-\end{figure}
-
 \section{Validity}
@@
 widely-believed {\em delayed issue effect} (that delaying the resolution of issues
-exponentially 
+very much 
 increases the difficulty of completing that  resolution).

### 2016-01-01 phaseDelay4Emse.tex
Updates from ShareLaTeX

@@
 belief and empirical data.
-In  summary, DIE is not some constant across all projects. Rather, is may be an historical relic that only occurrs intermittently 
-in  certain kinds of projects.  This is a significant result since it predicts that  new development
+In  summary, DIE is not some constant across all projects. Rather, DIE might
+actually be an historical relic that  occurs intermittently 
+only in  certain kinds of infrequently occuring projects.  This is a significant result since it predicts that  new development
 processes that promise to faster retire more issues will have only a  limited return on investment. 
@@
 
-The respondents answered questions on two scales: \newline
+The respondents answered questions on two scales:
+
+\bi
+\item
 \textbf{Agreement:} ``Based on your experience, do you agree that the statement above is correct?'' A Likert scale captured the agreement score from Strongly Disagree to Strongly Agree.
@@
 \newline
+\item
 \textbf{Applicability:} ``To the extent that you believe it, how widely do you think it applies among software development contexts?'' The possible answers were: 
@@
 Respondents were required to explain the applicability score in a text box.
+\ei
 
@@
 
-\begin{figure}[!t] 
+\begin{figure} 
 % \renewcommand{\baselinestretch}{0.8}
@@
 \end{center}
-\caption{Distribution of fix times seen in SEI TSP data. Left hand side bars
-show scale up on median values.
+\caption{ SEI TSP data: distribution of fix times. RHS bars
+show scale up on {\bf  50th percent} values (i.e. the median).
 }
@@
 
-\begin{figure}[!t] 
+\begin{figure} 
  %\renewcommand{\baselinestretch}{0.8}
@@
 461&365&Code&Test&0.6&****** \\
+348&720&Code&IntTest&1.2&************\\
 230&290&Code&SysTest&0.5&***** \\
-71&600&Code&AcceptTest&1.0&*********\\
-348&7200&Code&IntTest&12.0&************************************\\
-  &    &    &       &    & ************************************\\
-  &    &    &       &    & ************************************\\
+71&600&Code&AcceptTest&1.0&*********
 
@@
 \end{center}
-\caption{99th percentile scale ups.}
+\caption{SEI TSP data: distribution of fix times. RHS bars
+show scale up on {\bf 99th percent} values.}
 \label{fig:scale99}
@@
 
-\newpage
 \subsection{Observations from 171 Projects}

### 2016-01-01 phaseDelay4Emse.tex
Updates from ShareLaTeX

@@
 
+\subsection{Preliminaries} 
+Before beginning, it is appropriate to make the following full disclosure statement.
+
+All the 171 software projects
+studied here were developed using the Team Software Process (TSP$\textsuperscript{SM}$), which is a software development methodology
+developed and promulgated by the employer of the second and third author of this paper (for more details on TSP,
+ see \tion{tsp}).
+Hence, it is tempting to use these results to advocate for TSP since    TSP   might solve  the  DIE problem for  software projects 
+without needing all the extra machinery seen in (say)  Silicon Valley 
+continuous deployment organizations\footnote{ E.g. large scale process changes such as  open source
+development and/pr switching to new tools like NoSQL, Docker, AWS, Javascript and MEAN,  Jenkins, Github, etc.}.
+
+That said, it is not clear to us that TSP is such a radical change to software development that it can
+stamp out a supposedly rampant problem like the delayed issue effect. We view TSP as a better way to
+ {\em monitor} the activities   of  existing projects.  TSP
+ does not significantly change a project-- it just offers a better way to log the activity within
+ that project.
+
+Hence, given our data, the conclusion of this paper is {\em  either} the delayed issue effect can be  removed via:
+simple changes to a project (like TSP) {\em or} that the DIE problem does not exist in the first place.
+Either way, our more general point is the same: the delayed issue effect does not plague software development
+(which is a contrary claim to statements from leading figures in the  field of software engineering -- see the introduction).
+
 \section{Definitions \& Hypotheses}
@@
 
-\subsection{About TSP$\textsuperscript{SM}$}
+\subsection{About TSP$\textsuperscript{SM}$}\label{sect:tsp}
 
@@
 \begin{center}
-\begin{tabular}{r|rrr|ll|rl}   
-  \multicolumn{1}{c}{~} &\multicolumn{3}{c}{Percentiles}&\multicolumn{2}{c}{Phase}\\\cline{2-6}
-N & 25th & 50th & 75th &   injected & removed   & \multicolumn{2}{l}{Scale up w.r.t. to first phase}\\\hline
- 48&   14&   30&   51&Before&DesignInspect &1.00 & **********  \\
- 53&   13&   28&   49&Before&CodeReview &0.93 & **********  \\
- 94&   22&   45&   80&Before&CodeInspect &1.50 & ***************  \\
-153&   38&   77&  130&Before&UnitTest &2.56 & **************************  \\
-127&   30&   65&  128&Before&IntTest &2.16 & **********************  \\
- 92&   24&   56&   89&Before&SysTest &1.86 & *******************  \\\hline
-
- 56&   14&   35&   57&Planning&ReqtsReview &1.00 & **********  \\
- 46&   11&   24&   39&Planning&DesignInspect &0.68 & *******  \\
- 32&   18&   33&   84&Planning&UnitTest &0.94 & **********  \\\hline
-
-184&   45&   92&  155&Reqts&ReqtsReview &1.00 & **********  \\
-189&   46&   93&  155&Reqts&ReqtsInspect &1.00 & **********  \\
- 61&   14&   33&   55&Reqts&DesignReview &0.35 & ****  \\
-102&   24&   49&   83&Reqts&DesignInspect &0.52 & ******  \\
- 59&   18&   38&   60&Reqts&CodeInspect &0.40 & ****  \\
- 91&   22&   49&  104&Reqts&UnitTest &0.52 & ******  \\
-100&   33&   67&  120&Reqts&IntTest &0.72 & ********  \\
- 78&   28&  122&  237&Reqts&SysTest &1.31 & **************  \\\hline 
-210&   51&  103&  166&Design&DesignReview &1.01 & ***********  \\
-207&   50&  101&  156&Design&DesignInspect &1.00 & **********  \\
-134&   32&   66&  116&Design&CodeReview &0.65 & *******  \\
-163&   39&   79&  135&Design&CodeInspect &0.78 & ********  \\
-238&   58&  119&  183&Design&UnitTest &1.17 & ************  \\
-172&   42&   88&  158&Design&IntTest &0.87 & *********  \\
-163&   40&   80&  161&Design&SysTest &0.79 & ********  \\
- 80&   21&   45&   73&Design&AcceptTest &0.44 & *****  \\\hline
-
-235&   57&  115&  177&Code&CodeReview &1.00 &  **********  \\
-228&   56&  113&  176&Code&CodeInspect &0.99 & ********** \\
-243&   59&  125&  206&Code&IntTest &1.08 &     *********** \\
-191&   46&   95&  173&Code&SysTest &0.82 &     ******** \\
-269&   66&  133&  206&Code&UnitTest &1.15 &    ***********  \\
-130&   31&   65&  122&Code&AcceptTest &0.56 &  ******  \\
+\begin{tabular}{l@{~~~}r|r@{~}r|r@{~}lr@{~}l} 
+\renewcommand{\baselinestretch}{0.5}
+\multicolumn{2}{l|}{Phase} &\multicolumn{2}{c|}{Percentile}& \multicolumn{4}{c}{Ratio w.r.t. to phase where issue was injected)}\\\hline
+ injected & removed&50th & 95th &\multicolumn{2}{c|}{50th}&\multicolumn{2}{c}{95th}\\\hline
+\\
+Before&DesignInspect&30&97&1&\rule{10mm}{2mm}&1&\textcolor{red}{\rule{10mm}{2mm}} \\
+&CodeReview&28&94&0.93&\rule{9mm}{2mm}&0.96&\textcolor{red}{\rule{10mm}{2mm}} \\
+&CodeInspect&45&154&1.5&\rule{15mm}{2mm}&1.58&\textcolor{red}{\rule{16mm}{2mm}} \\
+&UnitTest&77&187&2.56&\rule{26mm}{2mm}&1.92&\textcolor{red}{\rule{19mm}{2mm}} \\
+&IntTest&65&424&2.16&\rule{22mm}{2mm}&4.37&\textcolor{red}{\rule{35mm}{2mm}} \\
+&SysTest&56&164&1.86&\rule{19mm}{2mm}&1.69&\textcolor{red}{\rule{17mm}{2mm}} \\\hline\\
+Planning&ReqtsReview&35&99&1&\rule{10mm}{2mm}&1&\textcolor{red}{\rule{10mm}{2mm}} \\
+&DesignInspect&24&81&0.68&\rule{7mm}{2mm}&0.81&\textcolor{red}{\rule{8mm}{2mm}} \\
+&UnitTest&33&238&0.94&\rule{9mm}{2mm}&2.4&\textcolor{red}{\rule{24mm}{2mm}} \\\hline\\
+Reqts&ReqtsReview&92&240&1&\rule{10mm}{2mm}&1&\textcolor{red}{\rule{10mm}{2mm}} \\
+&ReqtsInspect&93&243&1.01&\rule{10mm}{2mm}&1.01&\textcolor{red}{\rule{10mm}{2mm}} \\
+&DesignReview&33&119&0.35&\rule{4mm}{2mm}&0.49&\textcolor{red}{\rule{5mm}{2mm}} \\
+&DesignInspect&49&137&0.53&\rule{5mm}{2mm}&0.57&\textcolor{red}{\rule{6mm}{2mm}} \\
+&CodeInspect&38&156&0.41&\rule{4mm}{2mm}&0.65&\textcolor{red}{\rule{7mm}{2mm}} \\
+&UnitTest&49&180&0.53&\rule{5mm}{2mm}&0.75&\textcolor{red}{\rule{8mm}{2mm}} \\
+&IntTest&67&300&0.72&\rule{7mm}{2mm}&1.25&\textcolor{red}{\rule{13mm}{2mm}} \\
+&SysTest&122&459&1.32&\rule{13mm}{2mm}&1.91&\textcolor{red}{\rule{19mm}{2mm}} \\\hline\\
+Design&DesignReview&103&331&1&\rule{10mm}{2mm}&1&\textcolor{red}{\rule{10mm}{2mm}} \\
+&DesignInspect&101&272&0.98&\rule{10mm}{2mm}&0.82&\textcolor{red}{\rule{8mm}{2mm}} \\
+&CodeReview&66&215&0.64&\rule{6mm}{2mm}&0.64&\textcolor{red}{\rule{6mm}{2mm}} \\
+&CodeInspect&79&259&0.76&\rule{8mm}{2mm}&0.78&\textcolor{red}{\rule{8mm}{2mm}} \\
+&IntTest&88&337&0.85&\rule{9mm}{2mm}&1.01&\textcolor{red}{\rule{10mm}{2mm}} \\
+&UnitTest&119&383&1.15&\rule{12mm}{2mm}&1.15&\textcolor{red}{\rule{12mm}{2mm}} \\
+&SysTest&80&336&0.77&\rule{8mm}{2mm}&1.01&\textcolor{red}{\rule{10mm}{2mm}} \\
+&AcceptTest&45&120&0.43&\rule{4mm}{2mm}&0.36&\textcolor{red}{\rule{4mm}{2mm}} \\\hline\\
+Code&CodeInspect&113&337&1&\rule{10mm}{2mm}&1&\textcolor{red}{\rule{10mm}{2mm}} \\
+&CodeReview&115&300&1.01&\rule{10mm}{2mm}&0.89&\textcolor{red}{\rule{9mm}{2mm}} \\
+&IntTest&125&473&1.1&\rule{11mm}{2mm}&1.4&\textcolor{red}{\rule{14mm}{2mm}} \\
+&UnitTest&133&434&1.17&\rule{12mm}{2mm}&1.28&\textcolor{red}{\rule{13mm}{2mm}} \\
+&SysTest&95&383&0.84&\rule{8mm}{2mm}&1.13&\textcolor{red}{\rule{11mm}{2mm}} \\
+&AcceptTest&65&360&0.57&\rule{6mm}{2mm}&1.06&\textcolor{red}{\rule{11mm}{2mm}} \\
+
 \end{tabular}
+
 \end{center}
-\caption{ SEI TSP data: distribution of fix times. RHS bars
-show scale up on {\bf  50th percent} values (i.e. the median).
+\caption{ SEI TSP data: 50th and 95th percentiles of issue resolution times for issues injected in phase $X$ and resolved in phase $Y$.
+The right hand side bars
+illustrate how these resolutions increases (expressed as ratios of resolution time in the phase where they were first injected).
+The {\bf BLACK} and \textcolor{red}{{\bf RED}} bars show the increases for the  50th (median) and 95th percentile values, respectively.
 }
@@
 
-\begin{figure} 
- %\renewcommand{\baselinestretch}{0.8}
- \scriptsize
-\begin{center}
-\begin{tabular}{r|r|ll|rl}   
-  \multicolumn{1}{c}{~} &\multicolumn{1}{c}{Percentiles}&\multicolumn{2}{c}{Phase}\\\cline{2-4}
-N  & 99th &   injected & removed   & \multicolumn{2}{l}{Scale up w.r.t. to first phase}\\\hline
-
-57&257&Before&Code&1.00&*********\\
-72&153&Before&CodeInspect&0.60&****** \\
-165&282&Before&Test&1.01&**********\\
-50&365&Before&IntTest&1.42&************** \\
-42&151&Before&SysTest&0.59&****** \\\hline 
-37&600&Design&Design&1.00&********** \\
-455&277&Design&DesignInspect&0.46&***** \\
-218&240&Design&Code&0.40&**** \\
-166&163&Design&CodeInspect&0.27&*** \\
-542&652&Design&Test&1.09&***********\\
-67&277&Design&IntTest&0.46&***** \\
-92&480&Design&SysTest&0.80&******** \\\hline 
-126&600&Code&Code&1.0&**********\\
-459&243&Code&CodeInspect&0.4&**** \\
-461&365&Code&Test&0.6&****** \\
-348&720&Code&IntTest&1.2&************\\
-230&290&Code&SysTest&0.5&***** \\
-71&600&Code&AcceptTest&1.0&*********
-
-\end{tabular}
-\end{center}
-\caption{SEI TSP data: distribution of fix times. RHS bars
-show scale up on {\bf 99th percent} values.}
-\label{fig:scale99}
-\end{figure}
-
 \subsection{Observations from 171 Projects}
@@
 The distribution of defects found and fixed per phase in our data is shown in Figure~\ref{fig:fix-phase-dist}. 
-A high percentage of defects (44\%) were found and fixed in the early phases, i.e., requirements, high level design, and design reviews and inspections.
+ high percentage of defects (44\%) were found and fixed in the early phases, i.e., requirements, high level design, and design reviews and inspections.
 (This distribution is similar to that observed for other projects that emphasized investment in software engineering quality assurance practices.
@@
 
-\fig{raw} shows the 25th, 50th, 75th percentile
-of the time spent on {\em review} or {\em inspect} or {\em test} phases
-in our 171 projects.
-Note that, in TSP, when developers see issues, they enter {\em review} or 
+\fig{raw} shows the  50th and 95th percentile
+of the time spent resolving issues
+(note that, in TSP, when developers see issues, they enter {\em review} or 
 {\em inspect} or {\em test}
-until that issue is retired.
+until that issue is retired).
+These values include all the time required  to (a)~collect data and realize there is an error;
+(b)~prepare a fix;  and (c)~apply some validation
+procedure to check the fix (e.g. discuss it with a colleague or execute some tests).
 
@@
 i.e. according to {\em before, planning, requirements, design} or {\em  code}. 
-The stars on the right-hand-side of that figure show the scale ups associated with the phase removal stages,
-expressed as a ratio of the first item in that block (which, by definition, is 1.00 for
-that first item).
-If  this data exhibited a delayed issue effect,  we would expect those bars in each sub-division to exhibit an
-very large increased.
-
-\subsubsection{Comment 1}
-Figure~\ref{fig:raw}
- data does not support that   
-{\em  requirements errors are the most expensive to fix}.
-Note that   median time spent fixing errors from {\em requirements} across the life cycle was 92, 93,  49, 33, 39, 49, 67, 122 minutes\footnote{If this time seems long, then recall
-that it includes the time required  to (a)~collect data and realize there is an error;
-(b)~prepare a fix;  and (c)~apply some validation
-procedure to check the fix (e.g. discuss it with a colleague or execute some tests).}.
-This is:
-\bi
-\item
-Clearly longer than median times seen {\em Before } and in {\em Planning}
-(30,28,45,77,65,56,24,35,33);
-\item
-Somewhat shorter, but still not very much larger,  than the median times see after {\em requirements} in {\em design}
-(103,101, 79,66, 119,88, 80,45) and in {\em code} (65,115, 113,125, 95,133). 
-\ei
-For completeness, we ran an A12 test comparing the raw data for each phase, and found an effect size of ''medium'' or higher only for the difference between {\em requirements}
-and {\em code}. In this case, however, 
- the overall size of the difference
-{\em requirements} to {\em code} was less than 30\% (i.e. much less than we would
-expect from the delayed issue effect).
-
-That said, Figure~\ref{fig:raw} does support the 
+To ensure the representativeness of each line, we display examples
+where there exists at least $N\ge 30$ examples\footnote{We selected 30
+for this threshold via the central limit theorem~\cite{maxwelldata}.} of issues {\em injected} in phase $X$ then
+removed in phase $Y$.
+
+The right hand side bars of that figure
+illustrate how these resolutions increases (expressed as ratios of resolution time in the phase where they were first injected).
+The {\bf BLACK} and \textcolor{red}{{\bf RED}} bars show the increases for the  50th (median) and 95th percentile values, respectively.
+Those bars offer some support for the 
 claim that early lifecycle errors are most expensive to fix.
@@
 issues introduced in the {\em before} phase. For example,
-resolving {\em before} issues during unit testing was 2.56 times
+resolving {\em before} issues during unit testing was 2.56, 4.37 times (at the 50th and 95th percentile, respectively)
 slower than resolving issues earlier in the lifecycle.
@@
 than that suggested by the DIE literature. Returning to
-the {\em before/unit test} result: 2.56 is far smaller
+the {\em before/integration test} result: 2.56 and 4.37 is far smaller
 than the scale ups seen in Figure~1. In fact, nowhere in 
@@
 
-\newpage\subsubsection{Comment 2}
-Our data from these 171 project does not extend into post-delivery deployment. That said,
-there is enough evidence to raise doubts that our projects will exhibit very much more
-difficulty in resolving issues after deployment.
-
-As mentioned in \tion{earlyonset}, every other
-paper reporting DIE also reported  early onset of DIE
-within the current development. A remarkable feature of Figure~\ref{fig:raw} is the {\em absence}
-of a very large DIE   effect during development.
-In our data, the greatest increase in difficulty in resolving requirements issues was the 1.31 scale-up seen in systems testing--
-which is far smaller than the 37 to 250-fold increases reported in \fig{b81}, \fig{cost-to-fix},
-and \fig{steck}.
-
-That is, from our  data, we would not anticipate very much larger  difficulty in resolving requirements issues after
-deployment. 
- \subsubsection{Comment 3}
- The above specific statements are related just to requirements. Turning now
- to issues created in all phases in Figure~\ref{fig:raw},  we see that
- this data does not exhibit the very large   increases predicted
-by the delayed issue effect:
-\bi
-\item There is much ``up and down'' in the chart, but no steady very large growth such
-as that see in, for example,  \fig{b81}.
-\item The time in acceptance  testing is particularly low suggesting that few defects survived into testing
-(and hence, there were few instances where code required massive rework).
-\ei
 %In summary:
@@
 
-\section{Validity}
-We report validity concerns following the guidelines of Runeson \& H\"{o}st~\cite{runeson09}.
-
-\subsection{External validity}
-Threats to external validity affect the generalization of findings. The most obvious threat to external validity is that TSP is not representative of most software development processes. Hence,
-for example, 
-one way to discount the above argumentation is that our sample is biased-- that while general software
-systems suffer from the delayed issue effect, systems built using TSP are so much better controlled
-that they  mitigate against the DIE effect.
-
-It would be convenient to offer the last paragraph as the conclusion of this study
-(two of our authors, Nichols and Shull,  work at the organization that invented and promoted TSP).
- However, we cannot endorse such a conclusion:
- \bi
- \item
-We view TSP as a better way to
- {\em monitor} the activities {\em of  existing projects}. That is, TSP
- does not radically change a project, it just offers a better way to log the activity within
- that project. 
- \item
- Our 171 projects contain examples of a wide variety
+\section{Threats to Validity}
+
+\subsection{External Validity}
+
+When discussing these results with colleagues, they have raised the following questions as threats to validity.
+
+{\em Question1:} The data collection   for this work does not 
+distinguished between ``low risk'' and ``high risk''
+issues; i.e. between easy-to-fix issues and other most-expensive fixes. 
+It is possible that such high risk issues might occur infrequently,  yet are still detrimental
+since  their resolution time is extremely large?
+
+{\em Answer1} If such outliers were common, they would appear in the upper percentiles of results. 
+As shown in the \textcolor{red}{{\bf RED}} bars of \fig{raw}, no such effect appears in the 95th percentiles of \fig{raw}.
+
+ {\em Question2:}
+The data used in this analysis does not extend into post-delivery deployment. 
+Did this study failed to find the DIE effect since it does not arise until after deployment?
+
+{\em Answer2:} 
+As mentioned in \tion{earlyonset}, every other
+paper reporting DIE also reported  early onset of DIE
+within the current development. Specifically: those pro-DIE papers reported very large
+increases in the time required to resolve issues {\em even before delivery}. That is, extrapolating those
+trends it would be possible to predict for a large DIE effect, even before delivering the software.
+This is an important point since  Figure~\ref{fig:raw} shows an {\em absence}
+of a very large DIE   effect during development
+(in this data, the greatest increase in difficulty in resolving requirements issues was the 2.16 to 4.37
+scale-up seen in the {\em before} to {\em integration testing} 
+which is far smaller than the 37 to 250-fold increases reported in \fig{b81}, \fig{cost-to-fix},
+and \fig{steck}(.  
+That is, extrapolating the above data, we would not expect DIE effects, post-deployment.
+
+(As an aside, we have (informally) checked this prediction by following up with the projects for which we
+still have contacts. This follow-up study lacks the detailed data collection used above (it was more ``word of mouth''
+amongst the developers with personnel contacts with the second author). Based
+on that informal data collection, we can report that none of these 171  projects reported
+a  large post-deployment  
+DIE effect.)
+
+ {\em Question3:}
+ While the 171 project studies here are a large sample, are they representative of all software projects?
+
+ {\em Answer3:} These 171 projects contain examples of a wide variety
  of systems (ranging from e-commerce web portals to  banking systems) run in a variety of
@@
  development.
- \ei
+
+ {\em Question4:}
+
 Another way to discount this result is to say we are only reporting on small to medium-size

### 2016-06-15 phaseDelay.tex

@@
 yet published on this effect.
-In our sample, we found no evidence for the  delayed issue effect; i.e.
+We found no evidence for the  delayed issue effect; i.e.
 the  time  to resolve 
-issues in a later phase was not consistently  more than  
+issues in a later phase was {\bf not} consistently  more than  
 when  issues were resolved soon after their introduction.

### 2016-06-15 tex/fig13.tex
Updates from ShareLaTeX

\resizebox{\columnwidth}{!}{\begin{tabular}{c@{~}|l@{~}r@{~}|rrr|r@{~}l|r@{~}l}
  & \multicolumn{2}{c|}{ } & \multicolumn{3}{c|}{Percentiles} & \multicolumn{4}{c}{Growth with respect to earliest phase  } \\ 
   & \multicolumn{2}{c|}{Phase} & \multicolumn{3}{c|}{(units= time in miniutes)} & \multicolumn{4}{c}{(unitless, since these are ratios of two time values) } \\\hline

  rank & injected & removed & 50th & 90th &  IQR & \multicolumn{2}{c|}{50th percentile growth} & \multicolumn{2}{c}{90th percentile growth} \\ 
\hline\multicolumn{10}{c}{~}  \\
1 &  Before   &   DesignInspect   & 10 & 32 & 14 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}} & 1.00 & \textcolor{Gray}{\rule{10mm}{2mm}}\\
1 &   &   CodeReview      & 8 & 31 & 14 & 0.80 & \textcolor{black}{\rule{8mm}{2mm}} & 0.97 & \textcolor{Gray}{\rule{9mm}{2mm}}\\
1 &   &   CodeInspect     & 10 & 32 & 16 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}} & 1.00 & \textcolor{Gray}{\rule{10mm}{2mm}}\\
1 &   &   UnitTest        & 12 & 45 & 21 & 1.20 & \textcolor{black}{\rule{12mm}{2mm}} & 1.41 & \textcolor{Gray}{\rule{14mm}{2mm}}\\
1 &   &   IntTest         & 15 & 63 & 31 & 1.50 & \textcolor{black}{\rule{15mm}{2mm}} & 1.97 & \textcolor{Gray}{\rule{19mm}{2mm}}\\
1 &   &   SysTest         & 11 & 46 & 22 & 1.10 & \textcolor{black}{\rule{12mm}{2mm}} & 1.44 & \textcolor{Gray}{\rule{14mm}{2mm}}\\  
\hline\multicolumn{10}{c}{~}  \\
1 &  Planning     &   ReqtsReview     & 8 & 35 & 14 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}} & 1.00 & \textcolor{Gray}{\rule{10mm}{2mm}}\\
1 &               &   DesignInspect   & 11 & 31 & 13 & 1.38 & \textcolor{black}{\rule{13mm}{2mm}} & 0.89 & \textcolor{Gray}{\rule{8mm}{2mm}}\\
2 &               &   UnitTest        & 24 & 53 & 25 & 3.00 & \textcolor{black}{\rule{30mm}{2mm}} & 1.51 & \textcolor{Gray}{\rule{15mm}{2mm}}\\
\hline\multicolumn{10}{c}{~}  \\
1 &  Reqts   &   ReqtsReview     & 13 & 42 & 20 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}} & 1.00 & \textcolor{Gray}{\rule{10mm}{2mm}}\\
1 &          &   ReqtsInspect    & 12 & 40 & 18 & 0.92 & \textcolor{black}{\rule{9mm}{2mm}} & 0.95 & \textcolor{Gray}{\rule{9mm}{2mm}}\\
1 &          &   DesignReview    & 10 & 34 & 14 & 0.77 & \textcolor{black}{\rule{7mm}{2mm}} & 0.81 & \textcolor{Gray}{\rule{8mm}{2mm}}\\
1 &          &   DesignInspect   & 9 & 38 & 15 & 0.69 & \textcolor{black}{\rule{6mm}{2mm}} & 0.90 & \textcolor{Gray}{\rule{9mm}{2mm}}\\
1 &          &   CodeInspect     & 13 & 45 & 24 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}} & 1.07 & \textcolor{Gray}{\rule{10mm}{2mm}}\\
1 &          &   UnitTest        & 10 & 40 & 17 & 0.77 & \textcolor{black}{\rule{7mm}{2mm}} & 0.95 & \textcolor{Gray}{\rule{9mm}{2mm}}\\
1 &          &   IntTest         & 33 & 95 & 42 & 2.54 & \textcolor{black}{\rule{25mm}{2mm}} & 2.26 & \textcolor{Gray}{\rule{22mm}{2mm}}\\
1 &          &   SysTest         & 24 & 126 & 108 & 1.85 & \textcolor{black}{\rule{18mm}{2mm}} & 3.00 & \textcolor{Gray}{\rule{30mm}{2mm}}\\
\hline\multicolumn{10}{c}{~}  \\
1 & Design   &   DesignReview    & 11 & 37 & 16 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}} & 1.00 & \textcolor{Gray}{\rule{10mm}{2mm}}\\
1 &   &   DesignInspect   & 8 & 28 & 12 & 0.73 & \textcolor{black}{\rule{7mm}{2mm}} & 0.76 & \textcolor{Gray}{\rule{7mm}{2mm}}\\
1 &   &   CodeReview      & 10 & 40 & 18 & 0.91 & \textcolor{black}{\rule{9mm}{2mm}} & 1.08 & \textcolor{Gray}{\rule{10mm}{2mm}}\\
1 &   &   CodeInspect     & 9 & 33 & 14 & 0.82 & \textcolor{black}{\rule{8mm}{2mm}} & 0.89 & \textcolor{Gray}{\rule{8mm}{2mm}}\\
1 &   &   UnitTest        & 11 & 41 & 18 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}} & 1.11 & \textcolor{Gray}{\rule{12mm}{2mm}}\\
1 &   &   IntTest         & 17 & 75 & 31 & 1.55 & \textcolor{black}{\rule{15mm}{2mm}} & 2.03 & \textcolor{Gray}{\rule{20mm}{2mm}}\\
1 &   &   AcceptTest      & 14 & 44 & 19 & 1.27 & \textcolor{black}{\rule{12mm}{2mm}} & 1.19 & \textcolor{Gray}{\rule{12mm}{2mm}}\\
1 &   &   SysTest         & 13 & 40 & 18 & 1.18 & \textcolor{black}{\rule{12mm}{2mm}} & 1.08 & \textcolor{Gray}{\rule{10mm}{2mm}}\\
\hline\multicolumn{10}{c}{~}  \\
1 &  Code   &   CodeReview    & 10 & 35 & 16 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}} & 1.00 & \textcolor{Gray}{\rule{10mm}{2mm}}\\
1 &    &   CodeInspect   & 10 & 32 & 15 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}} & 0.91 & \textcolor{Gray}{\rule{9mm}{2mm}}\\
1 &   &   UnitTest      & 12 & 45 & 20 & 1.20 & \textcolor{black}{\rule{12mm}{2mm}} & 1.29 & \textcolor{Gray}{\rule{12mm}{2mm}}\\
1 &    &   IntTest       & 14 & 58 & 25 & 1.40 & \textcolor{black}{\rule{14mm}{2mm}} & 1.66 & \textcolor{Gray}{\rule{16mm}{2mm}}\\
1 &    &   AcceptTest    & 16 & 60 & 25 & 1.60 & \textcolor{black}{\rule{16mm}{2mm}} & 1.71 & \textcolor{Gray}{\rule{17mm}{2mm}}\\
1 &   &   SysTest       & 13 & 47 & 20 & 1.30 & \textcolor{black}{\rule{13mm}{2mm}} & 1.34 & \textcolor{Gray}{\rule{13mm}{2mm}}\\

\end{tabular}}

### 2016-06-15 phaseDelay4Emse.tex

@@
-\documentclass[smallcondesed]{svjour3}
+\documentclass[smallcondensed]{svjour3}
+\usepackage{amsmath}
 \usepackage{multirow}
@@
 \usepackage{balance}
-
+\usepackage{array}
 \usepackage{times}
 \usepackage{tabularx}
-\usepackage{subcaption}
+
 \usepackage{picture}
@@
         College Park, USA,
-       \email{llayman@fc-md.umd.edu}
+       \email{llayman@cese.fraunhofer.org}
 }
@@
 believe in a delayed issue effect (DIE); i.e.
- as issues linger longer in a system,
- they become very much harder to resolve.
+ the longer an issue lingers in the system, the more effort it requires to resolve.
 This belief
@@
 major investments in  new development
-processes that promise to retire more issues, sooner.
+processes that promise to retire more issues sooner.
 
@@
 We found no evidence for the  delayed issue effect; i.e.
-the  time  to resolve 
-issues in a later phase was  not consistently  more than  
+the  effort  to resolve 
+issues in a later phase was not consistently or substantially greater than  
 when  issues were resolved soon after their introduction.
 
-This paper documents the above studies and explores reasons for this  mismatch between this widely-held
-belief and empirical data.
+This paper documents the above study and explores reasons for this  mismatch between this common rule of thumb and empirical data.
 In  summary, DIE is not some constant across all projects. Rather, DIE might
-actually be an historical relic that  occurs intermittently 
-only in  certain kinds of infrequently occuring projects.  This is a significant result since it predicts that  new development
-processes that promise to faster retire more issues will have only a  limited return on investment. 
+be an historical relic that  occurs intermittently 
+only in  certain kinds of projects.  This is a significant result since it predicts that  new development
+processes that promise to faster retire more issues will have only a  limited return on investment, and that a long-held truth in software engineering can be defeated on a large scale. 
 \end{abstract}
@@
 In 2013-2014, 
-eleven  million programmers~\cite{pettey14} and
-half a trillion dollars~\cite{avram14} were spent on information technology.
+eleven  million programmers~\cite{avram14} and
+half a trillion dollars~\cite{pettey14} were spent on information technology.
 Such a large and growing effort should be managed and optimized via  well-researched conclusions.
@@
 Accordingly, this paper revisits
-the widely-held belief of a {\em delayed issue effect} (hereafter, DIE).
-Later in this paper, we offer a precise definition for this effect.
-For the moment, we describe it as follows:
-it is far easier to resolve issues earlier rather than later in the lifecycle.
+the commonly held belief we label {\em delayed issue effect} (hereafter, DIE):
+% Later in this paper, we offer a precise definition for this effect.
+% For the moment, we describe it as follows:
+more effort is required to resolve an issue the longer an issue lingers in a system.
+% it is far easier to resolve issues earlier rather than later in the lifecycle.
  \fig{b81} shows an example of the delayed issue effect (relating
@@
 
- Basili \& Boehm comment that since the 1980s,  this effect
+ Basili \& Boehm comment that, since the 1980s,  this effect
  \begin{quote}
@@
 \end{quote}
- \begin{figure}[!t]  
+ \begin{figure}[!b]  
 \begin{center}
@@
 Other researchers
-are just as adamant in asserting that that the delayed issue effect is a proven
-fact.
+are just as adamant in asserting that that the delayed issue effect is a generally useful law of software engineering.
 For example, what we call the delayed issued effect was listed first by Boehm and Basili in their "Top 10 list" of ``objective and quantitative data, relationships,
@@
 
-This paper calls into question all the above claims about the delayed
-issue effect. We suggest that the delayed issue effect might have been an dominant
-effect decades ago,  but {\em not}  for $21^{\mathit{st}}$
+In analyzing data from a contemporary set of software development projects, however, we did not find results to corroborate these claims. While the delayed issue effect might have been a dominant
+effect decades ago, perhaps not so for $21^{\mathit{st}}$
 software development. 
- The delayed issue effect was first reported in 1981 in a era of punch card programming
-and non-interactive environments~\cite{Boehm81}. In the 21$^\mathit{st}$ century, we  program in 
+ The delayed issue effect was first reported in 1976 in a era of punch card programming
+and non-interactive environments~\cite{Boehm76}. In the 21$^\mathit{st}$ century, we  program in 
 interactive environments with higher-level languages and better source code control
@@
 is far less    onerous    than before.
-
-Also, we note that our development practices have changed in ways that could   mitigate the delayed issued effect.  Previously,
-software was large monolithic systems that were ''write once and maintain
-forever.'' Today, there are more architectures that
-support  extensive and faster changes to software. 
-For example, 
-Microsoft is adjusting its development practices towards a continuous
-release paradigm. Theisen et al. reports their experiences where Microsoft architects
-continually adjust  their modules in response to security issues~\cite{Theisen15}.
-Also, upgrades to their Windows operating system is moving from service patches (which occur rarely) to continuous deployment (so there will be no Windows 11- just a stream of continuous updates to what is currently called Windows 10~\cite{bright15}).
-
-For another example, consider the MEAN stack preferred for web development by
-the continuous deployment community. Older architectures for web development
-often used some variant of LAMP (Linux + Appache + Mysql + PHP) that was an intricate
-combination of tools written in different languages. MEAN stacks, on the other hand,
-use Javascript throughout-- which makes large scale reorganizations faster to complete~\cite{wayner15}.
-
-Inspired by the success of agile approaches, other  development organizations are similarly reorganizing their work-- witness the US Department of Defense's 2010 mandate
-that all new software acquisitions must adopt agile methods~\cite{kim13}. 
-This change in DoD policy arises from a separation of baseline architecture
-(e.g. the design of an 
-aircraft carrier) and the development of applications within that architecture.
-For the baseline architecture, bad decisions
-made  early in the life cycle may be too expensive to change.
-But at least at the DoD, the majority of software development occurs within
-the framework of some larger architecture (e.g. an aircraft carrier).
-These smaller projects can certainly lever the agile advantage, while 
-using fast refactoring tools that mitigate against  the delayed issue effect.
-
-The above argument is an anecdotal evidence that  the delayed issue effect might no longer exist. But anecdotes are not really rigorous
-evidence. Accordingly,  this article explores the currency of the delayed issue effect.
+Further, software engineering theory and practice has evolved into new paradigms focused on rapid feedback and delivery, enabled by significant technological advances in the past 40 years.
+There is little empirical evidence for the delayed issue effect since its initial observation, no doubt due in part to DIE being ``just common sense'' as Glass states~\cite{glass02}.
+
+This article explores the currency of the delayed issue effect.
+% Given all of the changes to development practice in the time since the effect was first observed, finding no softening of the DIE might be taken as more surprising than our current results.
+These changes in software development practice are discussed later in this paper.
+% The above argument is an anecdotal evidence that  the delayed issue effect might no longer exist. But anecdotes are not rigorous
+% evidence. Accordingly,  this article explores the currency of the delayed issue effect.
 After some initial definitions, we discuss the
 value of checking old ideas.
-Next, a survey of industrial practitioners and researchers to document a widespread belief that delayed issues have a negative impact on projects.
-After that, we  analyze 171 software  projects developed in the period 2006--2014 to find  {\em no evidence} of the delayed issue effect:
-\bi
-\item
+Next, we present a survey of industrial practitioners and researchers that documents the widespread belief that delayed issues have a negative impact on projects.
+After that, we  analyze 171 software  projects developed in the period 2006--2014 and find  {\em no evidence} of the delayed issue effect.
+Finally, we discuss the validity and implications of our results, as well as possible reasons for the lack of observed effect given the state of the practice.
 To ensure reproducibility,
@@
 repository at openscience.us/repo. 
-\item
 To the best of our knowledge,
 this the largest study devoted the delayed issue effect yet conducted.
-\ei
-Finally, we discuss the validity and implications of our results.
-
-\subsection{Preliminaries} 
+
+\subsection{Preliminaries}\label{sect:nontsp}
 Before beginning, it is appropriate to make the following full disclosure statement. 
-
-All the 171 software projects
+All 171 software projects
 studied here were developed using the Team Software Process (TSP$\textsuperscript{SM}$), which is a software development methodology
@@
  see \tion{tsp}).
-Hence, it is tempting to use these results to advocate for TSP since    TSP   might solve  the  DIE problem for  software projects 
-without needing all the extra machinery seen in (say)  Silicon Valley 
-continuous deployment organizations\footnote{ E.g. large scale process changes such as  open source
-development and/pr switching to new tools like NoSQL, Docker, AWS, Javascript and MEAN,  Jenkins, Github, etc.}.
-
-That said, it is not clear to us that TSP is such a radical change to software development that it can
+
+We argue that TSP is not such a radical change to software development that it can
 stamp out a supposedly rampant problem like the delayed issue effect. We view TSP as a better way to
@@
  does not significantly change a project-- it just offers a better way to log the activity within
- that project.
-
-Hence, given our data, the conclusion of this paper is {\em  either} the delayed issue effect can be  removed via:
-simple changes to a project (like TSP) {\em or} that the DIE problem does not exist in the first place.
-Either way, our more general point is the same: the delayed issue effect does not plague software development
-(which is a contrary claim to statements from leading figures in the  field of software engineering -- see the introduction).
-
-\section{Definitions \& Hypotheses}
+ that project. The limitations of our sample drawing from TSP projects are discussed more thoroughly in the Threats to Validity section.
+
+% Hence, given our data, the conclusion of this paper is {\em  either} the delayed issue effect can be  removed via:
+% simple changes to a project (like TSP) {\em or} that the DIE problem does not exist in the first place, on at least some class of projects.
+% Either way, our  general point is the same: the delayed issue effect is not useful as a general law of what can be expected from software development practice.
+
+\section{Definitions \& Claims}
+\todo{Revisit these definitions after reviewing the rest of the paper for consistency of terms}
 This paper uses the following definitions:
@@
 \item
-We say that a measure collected in phase ${1,.,i,..j}$ is 
+We say that a measure $m$ collected in phase ${1,.,i,..j}$ is 
 {\em very much} more when  that
    measure at phase $j$   
-   is larger than the sum of that measure in 
-earlier phases $1 \le i < j$. 
+   is larger than the sum of those measures in the earlier phases;
+   i.e. $\sum_{i=1}^{j-1} m_i $. 
 \item
@@
 \ei
-Note that we use  the  term ``delayed issue effect'' as generalization of the
-more specific rule  ``requirements errors are hardest to fix''.
-This generalization is valid since the rationale for the rule about requirements
-is usually done as per McConnell~\cite{mcconnell01}; i.e. small upstream mistakes very
-early in the system can cascade into huge problems later in the lifecycle.
-That said, we prefer our more general term ``delayed issue effect'' since not
-only might requirements errors cascade, so too might analysis errors, design errors, etc.
-
-This paper defends the  following claims. Note that we call them
-``claims'' not ``hypotheses'' since the later require defense via some statistical
-significance test. On the other hand, our claims will be supported via a variety
-of arguments, presented later in this paper.
-
-{\bf  Claim1: ``DIE'' is a  widely-held belief.}
-Using a literature review, we can confirm that there are numerous historical papers (dating back decades)
-that endorse DIE. Also, using a survey conducted from this paper, we find  that 
-  DIE  appears as the
-single most strongly-held beliefs amongst commercial software engineers.
-
-{\bf  Claim2: ``DIE'' is a poorly documented.}
+% Note that we use  the  term ``delayed issue effect'' as generalization of the
+% more specific rule  ``requirements errors are hardest to fix''.
+% This generalization is valid since the rationale for the rule about requirements
+% is usually done as per McConnell~\cite{mcconnell01}; i.e. small upstream mistakes very
+% early in the system can cascade into huge problems later in the lifecycle.
+% That said, we prefer our more general term ``delayed issue effect'' since not
+% only might requirements errors cascade, so too might analysis errors, design errors, etc.
+
+Note that this  definition of ``difficult to resolve''  combines two concepts: time to change and cost to change.  Is it valid to assume the equivalence of time and cost?
+Certainly, there are cases where time is not the same as cost.
+Consider, for example, if debugging required some very expensive tool or the services or a very senior (and hence, very expensive) developer.
+Under those circumstances, time does not equate to cost.
+Having documented the above issues, we assert that they are unlikely to be major issues in the study.
+One of us (Nichols) was closely associated with many of the projects in our sample.
+He is unaware of any frequent use of exorbitantly expensive tools or people on these projects.
+For more on the validity of this definition
+of ``difficult to resolve'' see \tion{construct}.
+
+This paper defends the  following two claims and one hypothesis. Note hypotheses are defended
+using some statistical
+significance test while the claims are  supported via a variety
+of arguments presented later in this paper.
+
+{\bf  Claim1: ``DIE'' is a  commonly held belief.}
+Using a literature review, we confirm that there are numerous historical papers (dating back decades)
+that endorse the DIE. Also, using a survey of practitioners and researchers, we find  that 
+ the DIE  appears as the
+most strongly-held belief amongst software engineers.
+
+{\bf  Claim2: ``DIE'' is poorly documented.}
  As discussed in our literature review,  many of the papers reporting the DIE
-effect are either (1)~quite old (papers dating from last century);
-(2)~quoting prior papers without presenting   new data; 
-(3)~or citing data sources that can no longer be
+ either (1)~are quite old (papers dating from last century);
+(2)~quote prior papers without presenting   new data; 
+(3)~or cite data sources that can no longer be
 confirmed.
 
-{\bf Claim3: Delayed Issues not Harder to Resolve.}
+{\bf Hypothesis: Delayed issues are not harder to resolve.}
  In our sample of 
- 171 software  projects developed, we will show
-{\em no evidence}  that, during development, delayed issues were very much more harder to resolve the longer they were left
- in the software.
-
-\section{ But Why Reassess Old Truisms?}
-
-Before going any further,  we digress to discuss the merits of revisiting
-old conclusions in software engineering.
-
+ 171 commercial software  projects, we offer  a statistical analysis showing that, in overwhelming majority
+ of our results, there is no   
+ significant increase in the time to resolve issues  as they are delayed across multiple phases.
+
+\section{ Reassessing Old Truisms}
 Beliefs in general
@@
 the IEEE 1012 standard for software verification~\cite{1012}. 
- Endres \& Rombach offer dozens of lessons of software engineering~\cite{endres03}.
+ Endres \& Rombach offer empirical observations, theories, and laws\footnote{Endres \& Rombach note that these are not laws of nature in the scientific sense, but theories with repeated empirical evidence.}~\cite{endres03}.
  Many other 
-widely-cited researchers  do the same; e.g.
-Glass~\cite{glass02}; Jones~\cite{jones07}; Boehm~\cite{boehm00b}.
+commonly cited researchers  do the same, e.g.,
+Glass~\cite{glass02}, Jones \cite{jones07}, and Boehm~\cite{boehm00b}.
 Budgen \& Kitchenham seek to reorganize SE research using
@@
 
-That said, there are many empirical  findings that 
-raise doubts that   general laws of SE even exist:
+In contrast, there are many empirical  findings 
+that demonstrate the difficulty in finding general truisms in software engineering, even for claims that seem intuitive:
+
 \be
@@
 Turhan~\cite{me12d} lists 28 studies with contradictory conclusions
-on the relation of OO measures to defects.  Those results
+on the relation of object-oriented (OO) measures to defects.  Those results
  directly  contradict some of the laws listed by 
@@
 predict for better code quality. In  728 projects,
-they found  only a modest benefit in strong typing
-(and caution that that effect may actually be due to other  conflating factors).
+they found  only a modest benefit in strong typing and warn that the effect may be due to other conflating factors.
 \item
@@
 ``pre-release fault rates for software
-are a predictor for post-release failures'' (as claimed by~\cite{dunsmore88},
+are a predictor for post-release failures'' (as claimed in~\cite{dunsmore88},
 amongst others). For the systems described in~\cite{fenton97}, they
@@
 prior to release revealed very few faults after release.
-\item
-Meyer claims that   object-oriented (OO) encapsulation will
-reduce error rates in software~\cite{Meyer1988}.  Yet empirical results suggest
-that debugging an OO program is many times harder and
-longer than debugging a standard procedural program~\cite{hatton98}.
-\item
-A truism of visual programming is that ``visual
-representations are inherently superior to mere textual representations''. A review by Menzies suggests that the available
-evidence for this claim is hardly conclusive~\cite{me00v}. 
+% \item
+% Meyer claims that OO encapsulation will
+% reduce error rates in software~\cite{Meyer1988}.  Yet empirical results suggest
+% that debugging an OO program is many times harder and
+% longer than debugging a standard procedural program~\cite{hatton98}.
+% \item
+% A truism of visual programming is that ``visual
+% representations are inherently superior to mere textual representations''. A review by Menzies suggests that the available
+% evidence for this claim is hardly conclusive~\cite{me00v}. 
 \item
@@
 
-To be fair, 
-SE is  not the only
-field where practitioners hang on to persistent beliefs, even if the evidence
-for those beliefs is not strong.
+The dilemma of failing truths in the face of new evidence is not particular to software engineering. 
+% To be fair, 
+% SE is  not the only
+% field where practitioners hang on to persistent beliefs, even if the evidence
+% for those beliefs is not strong.
 The medical profession applies  many practices based on studies
@@
 The cognitive issues that complicate medicine are also found in software engineering.
-According to Passos et al.~\cite{passos11}, commercial developers
-are all too willing to believe in general
-truisms.
-They  warn that developers
+Passos et al.~\cite{passos11} warn that developers
 usually assume that the truisms they learn from a few past
@@
 all their future projects. They comment ``past experiences were taken into account without 
-much consideration for their context~\cite{passos11}.''  
-The results of J{\o}rgensen \& Gruschke~\cite{jorgensen09} concur with Passos et al. They report that 
-  supposed software engineering    ``experts'' rarely use lessons
-  from past projects to improve their future reasoning~\cite{jorgensen09}. 
- They note that
+much consideration for their context''~\cite{passos11}.
+The results of J{\o}rgensen \& Gruschke~\cite{jorgensen09} support Passos et al. In an empirical study of expert effort estimation, they report that the experts rarely use lessons
+  from past projects to improve their future reasoning in effort estimation~\cite{jorgensen09}. 
+ They note that,
 when the experts
@@
  conclusions and software projects  (see examples in~\cite{jorgensen09}).
-
-In summary, just like medicine, our field suffers when
- software engineers do  not revise old beliefs.  Therefore, it is important
+Just as in medicine, 
+%our field suffers when
+ %software engineers do  not revise old beliefs.  Therefore, 
+ it is important for our field
  to regularly  reexamine    old beliefs such as the delayed issue effect.
 
-\section{``DIE'' is a  widely-held belief}
-To assess the prevalence of DIE,   we conducted a survey of software engineers. If our surveyed practitioners make management decisions based on their
-understanding of SE theory, then the DIE  may well influence their decisions.
-
-Our survey collected data on software engineers' views of commonly held software engineering ``laws''.
-One of the laws questioned is a specific form of our delayed issue effect:  ``requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' (from Glass~\cite{glass02} p.71 who references Boehm \& Basili~\cite{boehm01}).
-We abbreviate this law (illustrated in \fig{b81}) as RqtsErr.
-
-(Technical aside: we use the RqtsErr formulation since this issue typically needs no supportive explanatory
+\section{``DIE'' is a  commonly held belief}
+\label{sect:belief}
+% To assess the prevalence of DIE,   we conducted a survey of software engineers. If our surveyed practitioners make management decisions based on their
+% understanding of SE theory, then the DIE  may well influence their decisions.
+
+We conducted two surveys of software engineers to assess the prevalence of DIE. 
+The surveys collected data on software engineers' views of DIE and other commonly held software engineering ``laws''.
+The surveys were conducted using Amazon's Mechanical Turk.
+The first survey was conducted only with professional software engineers.
+Participants were required to complete a pretest to verify their status as a professional or open source software developer and to confirm their knowledge of basic software engineering terminology and technology.
+The  second survey was conducted with Program Committee members of the ESEC/FSE 2015 and ICSE 2014 conferences solicited via email.
+
+The practitioner survey presented the following law: ``requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' (from Glass~\cite{glass02} p.71 who references Boehm \& Basili~\cite{boehm01}).
+We abbreviate this law as RqtsErr.\footnote{We use the RqtsErr formulation since this issue typically needs no supportive explanatory
 text. If we had asked respondents about our more general term ``delayed issue
-effect'', we would have had to burden our respondents with extra explanations).
-
-The survey was conducted in two phases using Amazon's Mechanical Turk. The first phase was conducted only with professional software engineers solicited through the Mechanical Turk;
-participants were required to complete a pretest to verify their status as a professional or open source software developer and to confirm their knowledge of basic software engineering terminology and technology.
-The  second survey was conducted with Program Committee members of the ESEC/FSE 2015 and ICSE 2014  conferences solicited via email.
-
-The respondents answered questions on two scales:
-
+effect'', we would have had to burden our respondents with extra explanations.}
+The PC member survey presented the RqtsErr law and an additional law on the DelayedIssueEffect: ``In general, the longer errors are in the system (requirements errors, design errors, coding errors, etc.), the more expensive they are to fix''.
+
+% Laws other than RqtsErr are not relevant to this paper, but we add them to as comparison points.
+
+The respondents answered two questions in response to each law:
 \bi
@@
 \textbf{Applicability:} ``To the extent that you believe it, how widely do you think it applies among software development contexts?'' The possible answers were:
--1 meaning ``I don't know''; and 0 meaning ``this law does not apply at all'';
- 1 meaning ``applies in a very narrow range of projects'';  2 meaning
-``rarely applies'';   3 meaning
-``occasionally applies'';  
-4 meaning ``very frequently applies''; and
- 5 meaning ``always applies''.
+\bi
+\item -1: I don't know
+\item 0:  this law does not apply at all
+ \item 1: applies in a very narrow range of projects  
+ \item 2: rarely applies
+ \item 3: occasionally applies
+\item 4: very frequently applies
+\item 5: always applies
+\ei
 Respondents were required to explain the applicability score in a text box.
@@
 
-In order to baseline the participant's answers, participants were presented with the RqtsErr law and others. For the purposes
-of this paper, the nature of the other laws other than RqtsErr are not relevant-- we
-only added them in as a way to calibrate responses to the RqtsErr question. All laws were drawn from \cite{glass02} and \cite{endres03}. 
-The PC member survey contained an additional question on ``In general, the longer errors are in the system (requirements errors, design errors, coding errors, etc.), the more expensive they are to fix'' (the Delayed Issue Effect).
-Responses were recorded using the agreement question Likert scale.
-
 Summary statistics for the agreement and applicability scores for the RqtsErr and DelayedIssueEffect laws are presented in Figure~\ref{fig:survey_results}.
 Responses whose Applicability response was ''I don't know'' are omitted from analysis.
+Laws other than RqtsErr and DIE are not relevant to this paper, but are shown for comparison.
 
@@
 \begin{tabular}{l|c|c|c|c|c}
- &  & \multicolumn{2}{c|}{agreement} & \multicolumn{2}{c}{applicability} \\\cline{1-1} 
+ &  & \multicolumn{2}{c|}{agreement} & \multicolumn{2}{c}{applicability} \\
 Practitioner survey  & N & med & mode & med & mode \\
@@
 Smaller changes have higher error density & 14 & 3 & 3 & 3.5 & 5 \\
-A developer is unsuited to test own code & 17 & 3 & 1 & 4 & 5\\\hline
-
-Researcher survey \\\hline 
+A developer is unsuited to test own code & 17 & 3 & 1 & 4 & 5\\
+ \multicolumn{6}{l}{}\\
+\multicolumn{6}{l}{Researcher survey} \\\hline 
 Process maturity improves output & 4 & 4 & 4 & 4 & 5 \\
@@
 
-Both  practitioners and researchers  strongly believed in RqtsErr: In both sets of responses, RqtsErr received  scores higher than most
-other laws. Further, in the case of practitioners, this law was rated
-as the single most believed effect.
-
-Caveat: in the free reponse texts, we note that the researchers who disagreed with the law generally asserted that requirements change can be expensive, but that the effect depends on the process used (e.g., agile vs.
+Both  practitioners and researchers strongly believed in RqtsErr. In both sets of responses, RqtsErr received  scores higher than most
+other laws. Overall, the RqtsErr law was the most agreed upon and most applicable law of 11 surveyed amongst practitioners, and the second most agreed upon law amongst researchers. 
+This results strongly support that the notion that the  delayed issue effect   is a commonly held belief in software engineering.
+From the free response texts, we note that the researchers who disagreed with RqtsErr generally asserted that requirements change can be expensive, but that the effect depends on the process used (e.g., agile vs.
 waterfall) and the adaptability of the system architecture.
 
-Overall, the RqtsErr law was the most agreed upon and most applicable law of 11 surveyed amongst practitioners, and the second most agreed upon law amongst researchers. 
-This results strongly support that the notion that the  delayed issue effect   is a widely-held belief in software engineering.
-
-\section{``DIE'' is Poorly  Documented}
+\section{DIE is Poorly  Documented}\label{sect:doco}
 
 One reason that industrial practitioners and academics believe so strongly in the delayed issue effect is that it is often referenced
-in the SE literature. For example,
-we know that \fig{b81} has been presented to the gradaute SE class at North Carolina State University, without quarrel or critical comment, every year for the last decade.
-
+in the SE literature. 
+%For example,
+%we know that \fig{b81} has been presented to the graduate SE class at North Carolina State University, without quarrel or %critical comment, every year for the last decade.
 Yet when we look at the literature, the evidence for
@@
 The first data on the difficulties of resolving delayed issues as a function of lifecycle phase date back to large systems in the late 70s from IBM~\cite{Fagan76}, TRW~\cite{Boehm76}, GTE~\cite{Daly77}, and Bell Labs~\cite{Stephenson76} (Figure~\ref{fig:cost-to-fix}).
-These studies are most often cited by secondary sources regarding the delayed issue effect.
+These studies are most often cited by secondary sources on the delayed issue effect (e.g., \cite{boehm01, mcconnell01,glass02}.
 We note that it is unclear from the text in ~\cite{Daly77} and \cite{Boehm76} if cost is defined in terms of effort, or in actual cost (i.e., labor, materiel, travel, etc).
-
-These studies suggest that the difficulty (in terms of effort) to find and fix an error monotonically increases with lifecycle phase.
-In 1990, Boehm~\cite{Boehm80} provides data suggesting that the cost-to-fix curve for small projects (from two student projects of 2000 deliverable source instructions) is flatter than for large projects (the dashed line of Figure~\ref{fig:cost-to-fix}).
+The data points from these studies are not published for analysis.
+These seminal studies are the basis for most DIE claims and that the cost to find and fix a defect monotonically increases with lifecycle phase.
 
@@
 Shull et al.~\cite{Shull02} conducted a literature survey and held a series of e-workshops with industry experts on fighting defects.
-Workshop participants from Toshiba and IBM reported cost-to-fix ratios between early lifecycle and post-delivery defects of 1:137 and 1:117 for large projects respectively~\cite{Shull02} -- but we note here that the raw data points were not provided (which makes confirming those numbers
-a difficult task).
-
+Workshop participants from Toshiba and IBM reported cost-to-fix ratios between early lifecycle and post-delivery defects of 1:137 and 1:117 for large projects respectively~\cite{Shull02} -- but the raw data points were not provided and thus cannot be confirmed.
+
+\todo{REview the primary sources for this paper}
 This was a common theme in the literature reviewed for this paper-- i.e.  that  it was no longer possible to access
@@
 requirements issues in eight case studies. Note that we cannot verify
-{\em any} of those results since the links in the references of
+any of those results since the links in the references of
 that  survey are all broken (``page not found'' errors).
 
-As to the research into agile methods, one goal of that approach
-is to reduce the difficulty associated with making changes later in the lifecycle~\cite{beck00}.
-Relatively little empirical data exists on this point.Elssamadisy and Schalliol~\cite{Elssamadisy02} anecdotally report on the growing, high cost of rework in a 50 person, three-year, 500KLOEC Extreme Programming project as the project grew in size and complexity-- but again we cannot access their
+A goal of agile methods 
+is to reduce the difficulty associated with making changes later in the lifecycle~\cite{beck00}. Relatively little empirical data exists on this point.
+Elssamadisy and Schalliol~\cite{Elssamadisy02} offer an anecdotal report on the growing, high cost of rework in a 50 person, three-year, 500KLOC Extreme Programming project as the project grew in size and complexity-- but again we cannot access their
 exact figures.
@@
 
-\begin{figure}[!t]
-\begin{center}
- \includegraphics[width=4in]{img/clutterbuck.png} 
- \end{center}
- \caption{Cost of change from an agile case study. From~\cite{Clutterbuck09}.}\label{fig:clutterbuck}
- \end{figure}
-
- All in all, the empirical results of this section seem insufficient to justify
- the strength of the beliefs documented in the previous section. Adding to our doubts are  several studies that report less-than large
- increase in the difficulties associated with making the changes associated with delayed
- issues. 
-Clutterbuck et al.~\cite{Clutterbuck09} studied a 5-month effort by a small-to-medium enterprise team developing a 71KLOEC web interface to a database application to implement 18 change requests-- see Figure~\ref{fig:clutterbuck} (note that these were for new and changed user requirements, not defects).
+% \begin{figure}[!t]
+% \begin{center}
+%  \includegraphics[width=4in]{img/clutterbuck.png} 
+%  \end{center}
+%  \caption{Cost of change from an agile case study. From~\cite{Clutterbuck09}.}\label{fig:clutterbuck}
+%  \end{figure}
+
+ %  All in all, the literature reviewed in this section
+% does not inspire confidence in a DIE effect due
+% to a lack of applicability and lack of replication.
+%  Adding to our doubts are
+
+% Clutterbuck et al.~\cite{Clutterbuck09} studied a 5-month effort by a small-to-medium enterprise team developing a 71KLOEC web interface to a database application to implement 18 change requests-- see Figure~\ref{fig:clutterbuck} (note that these were for new and changed user requirements, not defects).
 Clutterbuck et al.
@@
 
-Another example of less-than large increase in the difficulty associated with delayed issues comes from  Royce~\cite{Royce98}. He
-studied  a million-line, safety-critical missile defense system
+Some studies that report less-than large
+ increase in the effort required to fix delayed issues. Royce~\cite{Royce98} studied  a million-line, safety-critical missile defense system
 (see  Figure~\ref{fig:royce}).
@@
 Boehm~\cite{Boehm10} attributes this success to the  development process, which focused on removing architecture risk early in the development lifecycle.
-
- Other examples that report the opposite of DIE are:
-  \bi
-  \item
-  Boehm~\cite{Boehm80} reported a flatter growth rate for small, non-critical projects.
-   \item 
-  Data from NASA's Johnson Space Flight Center, reported by Shull~\cite{Shull02}, found that the cost to find certain non-critical classes of defects was fairly constant across lifecycle phases (1.2 hours on average early in the project, versus 1.5 hours late in the project).
-
-\ei
+Boehm~\cite{Boehm80} provides data suggesting that the cost-to-fix curve for small projects ,from two student projects of 2000 deliverable source instructions, is flatter than for large projects (the dashed line of Figure~\ref{fig:cost-to-fix}).
+Data from NASA's Johnson Space Flight Center, reported by Shull~\cite{Shull02}, found that the cost to fix certain non-critical classes of defects was fairly constant across lifecycle phases (1.2 hours on average early in the project, versus 1.5 hours late in the project).
+Again, the data points from these studies are not available, and thus the evidence both in favor of and contrary to the DIE is either quite old and/or is based on data that cannot be confirmed.
 
@@
 
-\subseection{Early Onset of the DIE Effect}\label{sect:earlyonset}
-
-One feature of the above results will become very important in the subsequent discussion.  
-All the   literature described above that reports the onset of  DIE  {\em prior} to delivery.
+\subsection{Early Onset of the DIE Effect}\label{sect:earlyonset}
+
+One feature of the results in the previous section is important to our subsequent discussion.  
+The DIE literature reports the onset of  DIE  {\em prior} to delivery.
 \bi
@@
 in believing that  the difficulty in resolving issues  will get much worse. Such managers
-would therefore expect DIE to have a marked effect, post-deployment.
-
-We make this point since,  in the new project data presented below, there is no early onset of DIE. 
-That is:
-\bi
-\item There is no evidence of a growing problem prior to delivery that delayed issues are becoming harder to manage
-\item  Hence, there there is less evidence that DIE is a trend that will significantly and negatively effect the product,
-post-delivery.
-\ei
-
-\section{Delayed Issues not  Harder  to Resolve}
+would therefore expect DIE to have a marked effect post-deployment.
+We make this point since,  in the new project data presented below, we focus on DIE pre-delivery.
+%there is no early onset of DIE. 
+% That is:
+% \bi
+% \item There is no evidence of a growing problem prior to delivery that delayed issues are becoming harder to manage
+% \item  Hence, there is less evidence that DIE is a trend that will significantly and negatively effect the product,
+% post-delivery.
+% \ei
+
+\section{Delayed Issues are not  Harder  to Resolve}
 
@@
 Accordingly, we examined 171 software projects conducted between 2006 and 2014.
+
 These projects took place at organizations in many countries and were conducted using  the Team Software Process (TSP$\textsuperscript{SM}$).
-
 Since 2000, the SEI has been teaching and coaching TSP teams.
@@
 The  most recent completions were in 2014.
+
 The projects were mostly small to medium, with a median duration of 46 days and a maximum duration of 90 days in major increments. 
@@
 Median team size was 7 people, with a maximum of 40. See \fig{dist} for the total
-effort seen in those projects.
-
-As to problem domain,
-many of the projects were e-commerce web portals or banking systems in the US, South Africa, and Mexico. 
+effort seen in those projects. Many of the projects were e-commerce web portals or banking systems in the US, South Africa, and Mexico. 
 There were  some  medical device projects in  the US, France, Japan, and Germany as well  as a commercial computer-aided design systems, and embedded systems.
+A more thorough characterization of the projects providing data is provided in \S\ref{sec:data_character}.
 
@@
 TSP is a software project management approach developed at the Software Engineering Institute (SEI) at Carnegie Mellon University~\cite{tsp00}.
-TSP is an extension of the Personal Software Process (PSP$\textsuperscript{SM}$) developed at the SEI by Watts Humphrey ~\cite{tsp00}.
-The data from these TSP projects were collected and stored in the Software Engineering Measured Process Repository (SEMPR) at the SEI.
-The Software Engineering Institute (SEI) at Carnegie Mellon University explores methods for software process improvement.
+TSP is an extension of the Personal Software Process (PSP$\textsuperscript{SM}$) developed at the SEI by Watts Humphrey~\cite{tsp00}.
 
@@
 {\em team size} times
-{\em days of work}). Median
-value = 271 days.}\label{fig:dist}
+{\em days of work}). For example, if 10 programmers work for 10 days,
+then the effort is 100 days. The median value in this plot
+ 271 days.}\label{fig:dist}
 \end{figure}
@@
 \item
- The time spent  tracking time, defects, and tasks requires less then 3\% of a developer's time. Weekly team meetings  require at most an hour, which is
+ The time spent  tracking time, defects, and tasks requires less than 3\% of a developer's time. Weekly team meetings  require at most an hour, which is
 only 2.5\% of a 40 hour work week. 
@@
 
+\subsection{Data Collection and Definitions}
+\label{sec:data-collection}
+
+Organizations using TSP agree to provide their project data to the SEI for use in research.
+In return the SEI agrees that  data must not be traceable to its source.
+The data is collected at major project events: launch, interim checkpoints, and at project completion.
+The data from these TSP projects were collected and stored in the Software Engineering Measured Process Repository (SEMPR) at the SEI.
+
+As of November 2014, the SEI TSP database contained data from 212
+TSP projects. The projects completed between July 2006 and
+November 2014; they included 47 organizations and 843 people. 
+The database fact tables
+contain 268,726 time logs, 
+154,238 task logs,
+ 47,376 defect logs, 
+and 26,534 size logs. 
+In this paper, we exclude 41 of the 212 that had too few defects \todo{Define ``too few''}, leaving 171 projects included in the analysis.
+
+% Data includes project and site  characteristic summaries, surveys of team members, launch presentations, launch outbriefs, and baseline plans, final data from the project, and the project post mortem report.
+In practice, this data requirement has only been enforced for the purposes of certifying and reauthorizing TSP coaches who must submit data to maintain their authorization.
+Coaches are certified by demonstrating competent use of the TSP process with the artifacts and data not by the actual project results.
+Of the data submitted, only the data recorded using the Process Dashboard tool has been collected and aggregated for this research.
+ %The key views include a project results summary, effort logs, task logs, and defect logs.
+The data has not yet undergone additional screening.
+A summary of the data quality issues was reported in a previous work ~\cite{shirai14} (that summary is discussed further in our {\em Validity} section, below).
+
+% Our units of analysis are: \textit{time}, \textit{defects}, \textit{time per defect}, and \textit{phase of development}.
+
+% \bi
+%     \item \emph{defects} - individual defects are recorded as line items in the defect logs uploaded to the SEMPR at the SEI.
+One or more defects are reported against a single \emph{plan item} in the time tracking logs, e.g., a review session, an inspection meeting, a test execution.
+%     \item \emph{time} - Time is tracked per person per plan item in the time-tracking logs, e.g.
+a 30 minute design review session involving 3 people will have three time log entries summing to 90 minutes.
+Time includes the time to analyze, repair, and validate a defect fix.
+%     \item \emph{time per defect} - The total \# of defects found in a plan item during a removal phase divided by the total time spent on that plan item in that phase.
+% \ei
+
+\subsubsection{Definition: time for plan item}
+
+Using a tool supporting the SEI data specification, developers keep detailed time-tracking logs. The time-tracking logs record  work start time, work end time,  delta
+work time, and interruption time. Software engineers are often
+interrupted by meetings, requests for technical help, reporting, and
+so forth. These events are recorded, in minutes, as interruption
+time.
+In TSP, time logs are recorded against \textit{plan items}.
+A planned item is a specific task assigned to a specific developer, such as resolving a defect, coding a feature, performing an inspection or writing a test.
+Each  work session includes a start time, an end time, and interruption time.
+The active time, or actual time for the plan item is calculated by summing the active time durations for all work sessions on that task.
+\[
+\text{\emph{actual time for plan item}} := \text{SUM(}\text{end time} - \text{start time} - \text{interruption time}) 
+\]
+Time is tracked per person per plan item in the time-tracking logs, e.g.
+a 30 minute design review session involving 3 people will have three time log entries summing to 90 minutes.
+Time includes the time to analyze, repair, and validate a defect fix.
+
+% In this paper, when we report ``time to resolve an
+% issue,'' we show the difference between the start and end times
+% of a work session, with any interruption time subtracted (the
+% difference in times, minus the interruptions).
+
+\subsubsection{Definition: defects and time-to-fix}\label{tion:defin}
+In the TSP, a  \emph{defect} is any change to a product, after its construction, that is necessary to make the product correct.
+A typographical error found in review is a defect.
+If that same defect is discovered while writing the code but before review, it is not considered to be a defect.
+SEI TSP defect types are:
+\bi
+\item Environment: design, compile, test,  other support  problems
+\item Interface: procedure calls and reference, I/O, user format
+\item Data: structure, content
+\item Documentation: comments, messages
+\item Syntax: spelling, punctuation typos, instruction formats
+\item Function: logic, pointers, loops, recursion, computation  
+\item Checking: error messages, inadequate checks
+\item Build: change management, library, version control
+\item Assignment: package
+declaration, duplicate names, scope
+\item System: configuration, timing, memory
+\ei
+
+\begin{figure}[t]
+\begin{center}
+\includegraphics[width=2.5in]{img/defect_type_frequency.png}
+\end{center}
+\caption{Relative frequencies of these defect types seen in our TSP data.}\label{fig:dtypes}.
+\end{figure}
+In our TSP data,  the relative frequencies of these defect types are shown in \fig{dtypes}. Around a quarter of the fixes were simple documentation changes.
+That said, 75\% of the changes are quote  elaborate;   e.g. fixes to function necessitates a careful reflection of the purpose of the code.
+
+Individual defects are recorded as line items in the defect logs uploaded to the SEMPR at the SEI.
+The defect entry includes the time and date a defect was discovered, the phase in which that defect was injected, the development phase in which it was removed, the time (in minutes) required to find and fix the defect, and the categorical type.
+
+One or more defects are reported against a single plan item in the time tracking logs, e.g., a review session, an inspection meeting, a test execution.
+The time-tracking logs against the plan items for a defect includes the time required to (a)~collect data and realize there is an error,
+(b)~prepare a fix,  and (c)~apply some validation
+procedure to check the fix (e.g. discuss it with a colleague or execute some tests).
+Since multiple defects can be recorded against a plan item, the time-to-fix a defect is defined as:
+\[
+\text{\emph{time-to-fix a defect}} := \frac{\text{time for defect plan item}}{\text{\# of defects in plan item}}
+\]
+
+\subsubsection{Definition: development phase} \label{development_phase}
+\begin{figure}[!b]  
+\begin{center}
+\includegraphics[width=1.6in]{img/wfall.png}  
+\end{center}
+\caption{Phases of our data.
+Abbreviations: 
+{\em Before}= before development; 
+{\em Reqts}	  = requirements; 
+{\em HLD}	  = high-level design; 
+{\em IntTest} = Integration testing (with code from others); 
+{\em SysTest} = system test (e.g. load stress tests); 
+{\em AcceptTest}  = acceptance testing (with users); 
+{\em review}        = private activity; 
+{\em inspect}        = group activity.}
+\label{fig:waterfall}
+\end{figure}
+
+The \textit{development phases} against which plan items are logged in the data are shown in \fig{waterfall}.
+Although the representation suggests a waterfall model, the SEI experience is that the projects follow a spiral approach or perform the work in iterative and/or incremental development cycles.
+The phases are thus the logical stages through which each increment must progress during development.
+
+One special feature of  \fig{waterfall} is the {\em before} phase, in which the TSP team assures that management has clearly identified cost, schedule, and scope goals appropriate to the upcoming development activities, often including a conceptual model~\cite{Humphrey:2005}.
+For example an architecture team must have sufficient requirements to reason about, prototype, and specify an architecture~\cite{Bachmann13} while a coding only team within a larger project would have more precisely defined requirements and high level design.
+
+%Within a project increment, multiple features or components may be developed and incremented.
+TSP is compatible %with agile and encourages iterative and incremental development.
+Nonetheless, the specific strategy and cycle %duration is a project decision.
+TSP does, however, strongly encourage 1) constructing units of sufficient size that %%measurement is practicable, and 2) separating the construction from appraisal and test activities.
+This %effectively highlights the separation of construction from rework activities and aids the apportionment of defects %to those found in appraisal activities (reviews and inspections) and those found through failure (test).
+The %distinct construction activities (requirements, high and detailed design, and code) were chosen to help teams %analyze the effectiveness and efficiency of their practices through analysis of the defect phase origin, type, fix %effort, and phase of discovery.
+
+Note that, in Figure~\ref{fig:waterfall}, several  phases in which the product is created have sub-phases of {\em review} and {\em inspect} to remove defects.
+As discussed in \S\ref{sect:tsp}, individuals perform personal reviews of their work products prior to the peer review (which TSP calls the inspection).
+Testing activities are divided as follows.
+Developers perform unit test prior to code complete.
+After code complete a standard phase is integration, which combines program units into a workable system ready for system test.
+Integration,  system test, and acceptance test are often performed by another group.
+
 \subsection{Data Integrity}
+\label{sec:data_integrity}
 
@@
 of noisy entries (superfluous  or spurious data). 
-The level of noise can be quite high. As reported
+The level of noise can be quite high. For example, as reported
 in \cite{shepperd12}, around
@@
 One the guiding principles of TSP was that  people performing the work are  responsible for planning and tracking the work. That is,  all the data collected here was entered
-by local developers. This data was then checked by local coaches before being sent to the SEI
-databases. Coaches are certified by demonstrating competent use of the TSP process with the artifacts and data.
+by local developers, who use the data for planning and tracking their projects. This data was then checked by local coaches before being sent to the SEI
+databases. While coaches are certified by demonstrating competent use of the TSP process with the artifacts and data,  project success or performance is not a criterion. 
 The use of certified local coaches within each project increases the integrity of our data.
 
-\subsection{Data Details}
-\label{sec:data-collection}
-
-%Organizations using TSP agree to provide their project data to the SEI for use in research.
-In return the SEI agrees    that  data must not be traceable to its source.
-The data is collected at major project events; launch, interim checkpoints, and at project completion.
-Data includes project and site  characteristic summaries, surveys of team members, launch presentations, launch outbriefs, and baseline plans, final data from the project, and the project post mortem report.
-In practice, this data requirement has only been enforced for the purposes of certifying and reauthorizing TSP coaches who must submit data to maintain their authorization.
-Coaches are certified by demonstrating competent use of the TSP process with the artifacts and data not by the actual project results.
-Of the data submitted, only the data recorded using the Process Dashboard tool has been collected and aggregated for this research.
- %The key views include a project results summary, effort logs, task logs, and defect logs.
-The data has not yet undergone additional screening.
-A summary of the data quality issues was reported in a previous work ~\cite{shirai14} (that summary is discussed further in our {\em Validity} section, below).
-
-Using tools provided by the SEI, developers kept very detailed logs of their daily activity. Our data includes  work start time, work end time,  delta
-work time, and interruption time. Software engineers are often
-interrupted by meetings, requests for technical help, reporting, and
-so forth. These events are recorded, in minutes, as interruption
-time.  
-In this paper, when we report ``time to resolve an
-issue,'' we show the difference between the start and end times
-of a work session, with any interruption time subtracted (the
-difference in times, minus the interruptions).
-
-As of November 2014, the SEI TSP database contained data from 212
-TSP projects. The projects completed between July 2006 and
-November 2014; they included 47 organizations and 843 people. 
-The database fact tables
-contain 268,726 time logs, 
-154,238 task logs,
- 47,376 defect logs, 
-and 26,534 size logs. 
-After selecting defects from the data log and joining the data to the time log table  171 of these projects remained (the excluded projects had no or too few defects to use in this analysis).
-
-\begin{figure}[!b]  
+\subsection{Project Descriptive Characteristics}
+\label{sec:data_character}
+
+%\todo{should we add a citation to Petersen and a few more words on context.}
+In this section we provide some descriptive statistics and discuss the of the projects from which this data was drawn and summarize some additional contextual information.
+The project contexts describe the conditions under which these measures were obtained, help determine relevance of the results, and may guide future data analysis with segmentation.
+
+While the processes and workflows among these projects can vary, the logical order is described section \ref{development_phase} is followed.
+Development tasks such as requirements development, design, or code, are typically followed by an appraisal phase such as personal review or inspection.
+Effort and effectiveness of these activities vary among projects and developers.
+
+We are unable at this time to provide all individual context data for each of the projects for several reasons.
+While the development data was recorded in tools and submitted in a structured form, context data was collected in less structured project questionnaires, site questionnaires, team member surveys, launch presentations and reports, post mortem presentations and reports.
+This data has not yet been mined from the submissions 1) because of the cost and effort required, 2) we are obligated to avoid providing any data that can identify projects (that is, the data must remain anonymous), and 3) the unstructured data is not always  complete  submitted.
+Gathering more projects will make it easier to anonymize the data and overcome missing data problems.
+Interest in the data sets by the community may encourage our sponsor to fund additional data mining.
+
+ The projects included come from 45 unique organizations from 6 countries.
+\fig{nationality} shows the country of origin and application domains for the projects.
+\fig{number of projects per organization} shows the number of projects from each organization.
+
+The most common countries of origin are the US and Mexico.
+Not apparent in this display is that the US companies tend to be fewer and larger with many projects while the Mexican companies are more likely to have one to several projects.
+Several companies, typically larger companies, are international with development teams in the US and either France or China.
+
+The most common project application domains are banking, consumer applications, engineering design tools, and medical devices.
+
+\begin{figure}[ht]
+\scriptsize
+\centering
+\begin{tabular}{cc}
+\begin{tabular}[t]{lr}
+Country      & \% of projects \\\hline
+China        & 1.0 \%            \\
+France       & 10.0 \%           \\
+Mexico       & 41.0 \%           \\
+South Africa & 4.0 \%            \\
+UK           & 1.5 \%          \\
+US           & 42.5 \%        
+\end{tabular}
+&
+\begin{tabular}[t]{lr}
+Application Domain         & count \\\hline
+Aviation                   & 19    \\
+Banking                    & 23    \\
+Business intelligence      & 19    \\
+Construction Support tools & 3     \\
+Consumer applications      & 24    \\
+Custom Applications        & 1     \\
+Embedded systems           & 2     \\
+Engineering Design Tools   & 21    \\
+Geography and Mapping      & 2     \\
+Government                 & 5     \\
+Human Resources Management & 3     \\
+Information Technology     & 1     \\
+Manufacturing              & 3     \\
+Medical Devices            & 15    \\
+Other                      & 2     \\
+Payroll services           & 1     \\
+Solutions Integration      & 1     \\
+Web applications           & 13    \\
+Wholesale or retail trade  & 9    
+\end{tabular}
+\end{tabular}
+\caption{Project nationality and application domain}
+\label{fig:nationality}
+\end{figure}
+
+\begin{figure}[!t] 
 \begin{center}
-\includegraphics[width=2in]{img/waterfall-v3.png}  
+\includegraphics[height=1.5in]{img/number_of_projects_per_org.png}
 \end{center} 
-\caption{Phases of our data.
-Abbreviations: 
-{\em Before}= before development; 
-{\em Reqts}	  = requirements; 
-{\em HLD}	  = high-level design; 
-{\em IntTest} = Integration testing (with code from others); 
-{\em SysTest} = system test (e.g. load stress tests); 
-{\em AcceptTest}  = acceptance testing (with users); 
-{\em review}        = private activity; 
-{\em inspect}        = group activity.}
-\label{fig:waterfall}
+\caption{Number of projects per development organization.}
+\label{fig:number of projects per organization}
 \end{figure}
 
-In these logs, a  \emph{defect} is any change to a product, after its construction, that is necessary to make the product correct.
-A typographical error found in review is a defect.
-If that same defect is discovered while writing the code but before review, it is not considered to be a defect.
-SEI TSP defect types are:
-\begin{itemize}
-\item Environment: design, compile, test,  other support  problems
-\item Interface: procedure calls and reference, I/O, user format
-\item Data: structure, content
-\item Documentation: comments, messages
-\item Syntax: spelling, punctuation typos, instruction formats
-\item Function: logic, pointers, loops, recursion, computation  
-\item Checking: error messages, inadequate checks
-\item Build: change management, library, version control
-\item Assignment: package
-declaration, duplicate names, scope
-\item System: configuration, timing, memory
-\end{itemize}
-The defect logs in that data  include the time and date a defect was discovered, the phase in which that defect was injected, the phase in which it was removed, the time (in minutes) required to find and fix the defect, and the categorical type.
-
-The phases logged by our data are shown in \fig{waterfall}.
- Although the representation suggests a waterfall model, the SEI experience is that all real implementations of any size follow a spiral approach with many team performing the work in iterative and/or incremental development cycles.
-
-One special feature of  \fig{waterfall} is
-the {\em before} phase, in which the TSP team assures that management has clearly identified cost, schedule, and scope goals appropriate to the upcoming development activities, often including a conceptual model ~\cite{Humphrey:2005}.
-For example an architecture team must have sufficient requirements to reason about, prototype, and specify an architecture~\cite{Bachmann13} while a coding only team within a larger project would have more precisely defined requirements and high level design.
-
-%Within a project increment, multiple features or components may be developed and incremented.
-TSP is compatible %with agile and encourages iterative and incremental development.
-Nonetheless, the specific strategy and cycle %duration is a project decision.
-TSP does, however, strongly encourage 1) constructing units of sufficient size that %%measurement is practicable, and 2) separating the construction from appraisal and test activities.
-This %effectively highlights the separation of construction from rework activities and aids the apportionment of defects %to those found in appraisal activities (reviews and inspections) and those found through failure (test).
-The %distinct construction activities (requirements, high and detailed design, and code) were chosen to help teams %analyze the effectiveness and efficiency of their practices through analysis of the defect phase origin, type, fix %effort, and phase of discovery.
-
-Note that, in that Figure~\ref{fig:waterfall},
-several  phases in which the product is created have sub-phases of {\em review} and {\em inspect} to remove defects.
-In TSP  reviews,  individuals perform personal reviews of their work products prior to the peer review
-(which TSP calls the inspection).
-Also, Figure~\ref{fig:waterfall} divides  testing     as follows.
-Developers perform unit test prior to code complete.
-After code complete a standard phase is integration, which combines program units into a workable system ready for system test.
-Integration,  system test, and acceptance test are often performed by another group.
-
-Using the above,
-our units of analysis are:
-\bi
-    \item \emph{defects} - individual defects are recorded as line items in the defect logs uploaded to the SEMPR at the SEI.
-One or more defects are reported against a single \emph{plan item} in the time tracking logs, e.g., a review session, an inspection meeting, a test execution.
-    \item \emph{time} - Time is tracked per person per plan item in the time-tracking logs, e.g.
-a 30 minute design review session involving 3 people will have three time log entries summing to 90 minutes.
-Time includes the time to analyze, repair, and validate a defect fix.
-    \item \emph{time per defect} - The total \# of defects found in a plan item during a removal phase divided by the total time spent on that plan item in that phase.
-\ei
+The data for programming languages is incomplete, with most projects using more than one language, but few reporting  programming language by component or size.
+The list of languages includes ABAP, ADA, Alpha, C, C++, C\#, ASP.net, Delphi, Gauss, Genexus, Hotware, HTML, Java, JavaScript, PHP, PLSQL, Ruby, SQL, and Visual Basic.
+
+ \begin{figure}[ht]
+\scriptsize
+\centering
+\begin{tabular}{llr} 
+ Begin Phase &  Final Phase & Count  \\\hline
+  Requirements      & Unit Test                   & 12  \\ 
+  Requirements      & Build and Integration Test  & 19  \\ 
+  Requirements      & System Test                 & 36  \\ 
+  High Level Design & Unit Test                   & 3  \\ 
+  High Level Design & Build and Integration Test  & 10  \\ 
+  High Level Design & System Test                 & 9  \\ 
+  Detailed Level Design & Unit Test                   & 18  \\ 
+  Detailed Level Design & Build and Integration Test  & 24  \\ 
+  Detailed Level Design & System Test                 & 17  \\ 
+\end{tabular}
+\caption{Earliest and latest process phases used by the projects}
+\label{fig:earliest-and-least-process-phases}
+\end{figure}
+
+ The project schedule, cost, and scope  are characterized by calendar duration, development team size project, and product size (measured in added and modified lines of code and number of components).
+These data are all available from the project fact sheets for each project.
+Summary statistics and the year of project initiation are displayed in Figure~\ref{fig:Project-descriptive-stats}.
+From this table we can make some observations about the range of project characteristics .
+
+ \begin{figure}[ht]
+\scriptsize
+\centering
+\begin{tabular}{lrrrrrrrl}
+                                                                   & N   & Min & Q1   & Median & Q3    & Max   & Mean   &  Distribution \\\hline
+Team size                                                          & 171 & 1   & 4    & 6      & 10    & 36    &  7.8    &  \includegraphics[width=15mm]{img/team_size.png} \\ 
+Duration {[}days{]}                                                & 171 & 7   & 33   & 61     & 118   & 1918  & 107    & \includegraphics[width=15mm]{img/duration.png} \\
+Added \& Modified LOC                                                 & 117 & 2   & 1125 & 4201   & 13092 & 88394 & 10259  & \includegraphics[width=15mm]{img/LOC.png} \\
+Components                                                         & 171 & 8   & 26   & 49     & 107   & 4170  & 116.2  & \includegraphics[width=15mm]{img/components.png} \\
+Year                                                               & 171 &     & 2011 & 2012   & 2013  & 2014  & 2011.9 & \includegraphics[width=15mm]{img/year.png}
+\end{tabular}
+
+\caption{Project summary description}
+\label{fig:Project-descriptive-stats}
+\end{figure}
+
+ Of the 171 projects in the sample, only 117 collected size data in lines of code, however all projects tracked effort and  the component counts with applied effort are provided.
+Other data is complete for all 171 projects.
+The projects were mostly of short duration and small to medium size.
+The median project began in 2012 lasted 61 days, produced 4200 Lines of Code, 49 components (modules or features).
+Duration ranged from 7 to 1918 days.
+Size ranged from minimal (this may represent a short maintenance project) to 88394.
+The earliest project was in 2006 and the most recent in 2014.
+
+ The teams were mostly of small size with a median size of 6 and a mean of 7.8. only 25\% of the teams were larger than 10 with a long tail on the distribution.
 
 \begin{figure}[!t] 
-
+\vspace{0.5cm}
 \begin{center}
-\hspace{-1.5in}\includegraphics[height=1.5in]{img/fix-phase-dist.pdf}
+\includegraphics[height=2.5in]{img/fix-phase-dist.pdf}
 \end{center} 
-\vspace{2 cm}
-
 \caption{Distribution of defects found and fixed by phase.}
@@
 % \renewcommand{\baselinestretch}{0.8}
- \scriptsize
+% \small
 \begin{center}
-\begin{tabular}{l@{~~~}r|r@{~}r|r@{~}lr@{~}l} 
-\renewcommand{\baselinestretch}{0.5}
-\multicolumn{2}{l|}{Phase} &\multicolumn{2}{c|}{Percentile}& \multicolumn{4}{c}{Ratio w.r.t. to phase where issue was injected)}\\\hline
- injected & removed&50th & 95th &\multicolumn{2}{c|}{50th}&\multicolumn{2}{c}{95th}\\\hline
-\\
-Before&DesignInspect&30&97&1&\rule{10mm}{2mm}&1&\textcolor{red}{\rule{10mm}{2mm}} \\
-&CodeReview&28&94&0.93&\rule{9mm}{2mm}&0.96&\textcolor{red}{\rule{10mm}{2mm}} \\
-&CodeInspect&45&154&1.5&\rule{15mm}{2mm}&1.58&\textcolor{red}{\rule{16mm}{2mm}} \\
-&UnitTest&77&187&2.56&\rule{26mm}{2mm}&1.92&\textcolor{red}{\rule{19mm}{2mm}} \\
-&IntTest&65&424&2.16&\rule{22mm}{2mm}&4.37&\textcolor{red}{\rule{35mm}{2mm}} \\
-&SysTest&56&164&1.86&\rule{19mm}{2mm}&1.69&\textcolor{red}{\rule{17mm}{2mm}} \\\hline\\
-Planning&ReqtsReview&35&99&1&\rule{10mm}{2mm}&1&\textcolor{red}{\rule{10mm}{2mm}} \\
-&DesignInspect&24&81&0.68&\rule{7mm}{2mm}&0.81&\textcolor{red}{\rule{8mm}{2mm}} \\
-&UnitTest&33&238&0.94&\rule{9mm}{2mm}&2.4&\textcolor{red}{\rule{24mm}{2mm}} \\\hline\\
-Reqts&ReqtsReview&92&240&1&\rule{10mm}{2mm}&1&\textcolor{red}{\rule{10mm}{2mm}} \\
-&ReqtsInspect&93&243&1.01&\rule{10mm}{2mm}&1.01&\textcolor{red}{\rule{10mm}{2mm}} \\
-&DesignReview&33&119&0.35&\rule{4mm}{2mm}&0.49&\textcolor{red}{\rule{5mm}{2mm}} \\
-&DesignInspect&49&137&0.53&\rule{5mm}{2mm}&0.57&\textcolor{red}{\rule{6mm}{2mm}} \\
-&CodeInspect&38&156&0.41&\rule{4mm}{2mm}&0.65&\textcolor{red}{\rule{7mm}{2mm}} \\
-&UnitTest&49&180&0.53&\rule{5mm}{2mm}&0.75&\textcolor{red}{\rule{8mm}{2mm}} \\
-&IntTest&67&300&0.72&\rule{7mm}{2mm}&1.25&\textcolor{red}{\rule{13mm}{2mm}} \\
-&SysTest&122&459&1.32&\rule{13mm}{2mm}&1.91&\textcolor{red}{\rule{19mm}{2mm}} \\\hline\\
-Design&DesignReview&103&331&1&\rule{10mm}{2mm}&1&\textcolor{red}{\rule{10mm}{2mm}} \\
-&DesignInspect&101&272&0.98&\rule{10mm}{2mm}&0.82&\textcolor{red}{\rule{8mm}{2mm}} \\
-&CodeReview&66&215&0.64&\rule{6mm}{2mm}&0.64&\textcolor{red}{\rule{6mm}{2mm}} \\
-&CodeInspect&79&259&0.76&\rule{8mm}{2mm}&0.78&\textcolor{red}{\rule{8mm}{2mm}} \\
-&IntTest&88&337&0.85&\rule{9mm}{2mm}&1.01&\textcolor{red}{\rule{10mm}{2mm}} \\
-&UnitTest&119&383&1.15&\rule{12mm}{2mm}&1.15&\textcolor{red}{\rule{12mm}{2mm}} \\
-&SysTest&80&336&0.77&\rule{8mm}{2mm}&1.01&\textcolor{red}{\rule{10mm}{2mm}} \\
-&AcceptTest&45&120&0.43&\rule{4mm}{2mm}&0.36&\textcolor{red}{\rule{4mm}{2mm}} \\\hline\\
-Code&CodeInspect&113&337&1&\rule{10mm}{2mm}&1&\textcolor{red}{\rule{10mm}{2mm}} \\
-&CodeReview&115&300&1.01&\rule{10mm}{2mm}&0.89&\textcolor{red}{\rule{9mm}{2mm}} \\
-&IntTest&125&473&1.1&\rule{11mm}{2mm}&1.4&\textcolor{red}{\rule{14mm}{2mm}} \\
-&UnitTest&133&434&1.17&\rule{12mm}{2mm}&1.28&\textcolor{red}{\rule{13mm}{2mm}} \\
-&SysTest&95&383&0.84&\rule{8mm}{2mm}&1.13&\textcolor{red}{\rule{11mm}{2mm}} \\
-&AcceptTest&65&360&0.57&\rule{6mm}{2mm}&1.06&\textcolor{red}{\rule{11mm}{2mm}} \\
-
-\end{tabular}
+\input{tex/fig13}
+%
+% \begin{tabular}{l@{~~~}r|r@{~}r|r@{~}lr@{~}l} 
+% \renewcommand{\baselinestretch}{0.5}
+% \multicolumn{2}{l|}{Phase}&\multicolumn{2}{c|}{Percentile}&\multicolumn{4}{c}{Ratio w.r.t. to phase where issue was injected}\\\hline
+%  injected&removed&50th&95th&\multicolumn{2}{c|}{50th}&\multicolumn{2}{c}{95th}\\\hline
+% \\
+% Before&DesignInspect&30&97&1&\rule{10mm}{2mm}&1&\textcolor{red}{\rulree{10mm}{2mm}} \\
+% &CodeReview&28&94&0.93&\rule{9mm}{2mm}&0.96&\textcolor{red}{\rule{10mm}{2mm}} \\
+% &CodeInspect&45&154&1.5&\rule{15mm}{2mm}&1.58&\textcolor{red}{\rule{16mm}{2mm}} \\
+% &UnitTest&77&187&2.56&\rule{26mm}{2mm}&1.92&\textcolor{red}{\rule{19mm}{2mm}} \\
+% &IntTest&65&424&2.16&\rule{22mm}{2mm}&4.37&\textcolor{red}{\rule{35mm}{2mm}} \\
+% &SysTest&56&164&1.86&\rule{19mm}{2mm}&1.69&\textcolor{red}{\rule{17mm}{2mm}} \\\hline\\
+% Planning&ReqtsReview&35&99&1&\rule{10mm}{2mm}&1&\textcolor{red}{\rule{10mm}{2mm}} \\
+% &DesignInspect&24&81&0.68&\rule{7mm}{2mm}&0.81&\textcolor{red}{\rule{8mm}{2mm}} \\
+% &UnitTest&33&238&0.94&\rule{9mm}{2mm}&2.4&\textcolor{red}{\rule{24mm}{2mm}} \\\hline\\
+% Reqts&ReqtsReview&92&240&1&\rule{10mm}{2mm}&1&\textcolor{red}{\rule{10mm}{2mm}} \\
+% &ReqtsInspect&93&243&1.01&\rule{10mm}{2mm}&1.01&\textcolor{red}{\rule{10mm}{2mm}} \\
+% &DesignReview&33&119&0.35&\rule{4mm}{2mm}&0.49&\textcolor{red}{\rule{5mm}{2mm}} \\
+% &DesignInspect&49&137&0.53&\rule{5mm}{2mm}&0.57&\textcolor{red}{\rule{6mm}{2mm}} \\
+% &CodeInspect&38&156&0.41&\rule{4mm}{2mm}&0.65&\textcolor{red}{\rule{7mm}{2mm}} \\
+% &UnitTest&49&180&0.53&\rule{5mm}{2mm}&0.75&\textcolor{red}{\rule{8mm}{2mm}} \\
+% &IntTest&67&300&0.72&\rule{7mm}{2mm}&1.25&\textcolor{red}{\rule{13mm}{2mm}} \\
+% &SysTest&122&459&1.32&\rule{13mm}{2mm}&1.91&\textcolor{red}{\rule{19mm}{2mm}} \\\hline\\
+% Design&DesignReview&103&331&1&\rule{10mm}{2mm}&1&\textcolor{red}{\rule{10mm}{2mm}} \\
+% &DesignInspect&101&272&0.98&\rule{10mm}{2mm}&0.82&\textcolor{red}{\rule{8mm}{2mm}} \\
+% &CodeReview&66&215&0.64&\rule{6mm}{2mm}&0.64&\textcolor{red}{\rule{6mm}{2mm}} \\
+% &CodeInspect&79&259&0.76&\rule{8mm}{2mm}&0.78&\textcolor{red}{\rule{8mm}{2mm}} \\
+% &UnitTest&119&383&1.15&\rule{12mm}{2mm}&1.15&\textcolor{red}{\rule{12mm}{2mm}} \\
+% &IntTest&88&337&0.85&\rule{9mm}{2mm}&1.01&\textcolor{red}{\rule{10mm}{2mm}} \\
+% &SysTest&80&336&0.77&\rule{8mm}{2mm}&1.01&\textcolor{red}{\rule{10mm}{2mm}} \\
+% &AcceptTest&45&120&0.43&\rule{4mm}{2mm}&0.36&\textcolor{red}{\rule{4mm}{2mm}} \\\hline\\
+% Code&CodeInspect&113&337&1&\rule{10mm}{2mm}&1&\textcolor{red}{\rule{10mm}{2mm}} \\
+% &CodeReview&115&300&1.01&\rule{10mm}{2mm}&0.89&\textcolor{red}{\rule{9mm}{2mm}} \\
+% &UnitTest&133&434&1.17&\rule{12mm}{2mm}&1.28&\textcolor{red}{\rule{13mm}{2mm}} \\
+% &IntTest&125&473&1.1&\rule{11mm}{2mm}&1.4&\textcolor{red}{\rule{14mm}{2mm}} \\
+% &SysTest&95&383&0.84&\rule{8mm}{2mm}&1.13&\textcolor{red}{\rule{11mm}{2mm}} \\
+% &AcceptTest&65&360&0.57&\rule{6mm}{2mm}&1.06&\textcolor{red}{\rule{11mm}{2mm}} \\
+
+% \end{tabular}
 
 \end{center}
-\caption{ SEI TSP data: 50th and 95th percentiles of issue resolution times for issues injected in phase $X$ and resolved in phase $Y$.
+\caption{Time to resolve issues seen in the SEI TSP data.
+The main effect here is that there is no clear exponential growth in the time to resolve issues as issues are injected in some phase $P_0$ then resolved in some subsequent issue $P_r$.
+Shown where are the 50th and 90th percentiles of issue resolution times for issues injected in phase $P_O$ and resolved in phase $P_r$ (these values are calculated
+by sorting all resolution time, then reporting the values at 0.5 and 0.9).
+The ``IQR'' column shows the ``inter-quartile range''; i.e. the range of values representing the 75th - 25th percentile range
+The left hand column shows the result of the Scott-Knott ranking procedure described in \tion{stats}. Note that
+most treatments achieved the same ranks i.e. they were found to be not statistically significantly different  (the one exception is the Planning:UnitTest results were UnitTests were ranked 2).
 The right hand side bars
-illustrate how these resolutions increases (expressed as ratios of resolution time in the phase where they were first injected).
-The {\bf BLACK} and \textcolor{red}{{\bf RED}} bars show the increases for the  50th (median) and 95th percentile values, respectively.
-}
+illustrate how these resolutions increases.
+The {\bf BLACK} and \textcolor{Gray}{{\bf GRAY}} bars show the increases for the  50th (median) and 90th percentile values, respectively.
+These increases are calculated with respect to the first value in each section ``Before, Planning, Reqts.
+Design, Code''.
+These right-hand-side bars are unitless since they are ratios. For example,
+on the last line, issues injected during coding and fixed in SysTest take 47 minutes (median) to resolve. This is 130\% more than the 35 minites (median) required
+to resolve coding issues during CodeReview.}
 \label{fig:raw}
@@
 
-\subsection{Observations from 171 Projects}
+\subsection{Statistical Analysis}\label{sect:stats}
+
+    In the following presentation of our results, three statistical methods were used to test for the delayed
+    issue effect: the Scott-Knott ranker;   bootstrap sampling (to test for statistical significantly
+    different results); and an effect size test (to reject any significant differences that are trivially small).
+
+    The Scott-Knott ranker 
+    recommended was   recommended  by  Mittas and Angelis in a
+    recent TSE'13 article~\cite{mittas13} and by Ghotra et al. in a recent
+    ICSE'15 article~\cite{ghotra2015icse}
+    Scott-Knott is a 
+    top-down clustering approach used to rank different treatments. If that 
+    clustering finds an ``interesting division'' of the data, then some 
+    statistical test is applied to the two divisions to check if they are 
+    statistically significant different. If so, Scott-Knott considers recurses into both 
+    halves.
+    Before Scott-Knot recurses,  however,  it apples some  statistical hypothesis test $H$ to check
+    if $m,n$ are significantly different.
+
+    To operationalize ``interesting'', Scott-Knott seeks the division of 
+    of $l$ treatments into subsets of size $m,n$ of sizes $ls,ms,ns$ and median values
+    $l.\mu, m.\mu, n.\mu$ (respectively) in order to maximize
+ $\frac{ms}{ls}abs(m.\mu - l.\mu)^2 + \frac{ns}{ls}abs(n.\mu - l.\mu)^2$.
+
+   To operationalize $H$, we use both  bootstrap sampling  and the A12 effect size test.
+   In other 
+    words, we divide the data if \textit{both} bootstrap sampling and effect 
+    size test agree that a division is statistically significant (with a 
+    confidence of 99\%) and not a small effect ($A12 \ge 0.6$).
+    For a justification of the use of non-parametric bootstrapping, see Efron 
+    \& Tibshirani~\cite[p220-223]{efron93}. For a justification of the use of 
+    effect size tests see Shepperd and MacDonell~\cite{shepperd12a}; 
+    Kampenes~\cite{kampenes07}; and Kocaguenli et 
+    al.~\cite{Kocaguneli2013:ep}. These researchers warn that even if a 
+    hypothesis test declares two populations to be ``significantly'' 
+    different, then that result is misleading if the ``effect size'' is very 
+    small. Hence, to assess the performance differences we first must rule out 
+    small effects using A12, a test   recently endorsed by Arcuri and 
+    Briand at ICSE'11~\cite{arcuri11}.
+
+    To  apply Scott-Knott, we divided data into the phases $P_0$ where issues are introduced.
+Next, for each division, we separated all the issues that were removed at different subsequent issues $P_r \in \{P_1,P_2,..\}$.
+    For each pair $P_0,P_r$, we build one treatment containing the  issue resolution times for     issues raises in $P_0$ and resolved in $P_r$. These treatments
+    were then ranked by Scott-Knott.
+
+\subsection{Observations from 171 Projects}\label{tion:171projects}
 
 The distribution of defects found and fixed per phase in our data is shown in Figure~\ref{fig:fix-phase-dist}. 
- high percentage of defects (44\%) were found and fixed in the early phases, i.e., requirements, high level design, and design reviews and inspections.
-(This distribution is similar to that observed for other projects that emphasized investment in software engineering quality assurance practices.
+A high percentage of defects (44\%) were found and fixed in the early phases, i.e., requirements, high level design, and design reviews and inspections.
+\todo{Make sure that defect injection phases in the figure match those of \fig{waterfall}} This distribution is similar to that observed for other projects that emphasized investment in software engineering quality assurance practices.
 For example, Jones and Bonsignour report 52\% of pretest defects removed before entering implementation, for large projects that focus on upfront defect removal techniques \cite{jones12}.
-NASA robotics projects had a slightly higher percentage (58\%) of defects removed before implementation began, although these had invested in IV\&V on top of other forms of defect removal \cite{me08a}.)
-
-\fig{raw} shows the  50th and 95th percentile
+NASA robotics projects had a slightly higher percentage (58\%) of defects removed before implementation began, although these had invested in independent verification and validation on top of other forms of defect removal \cite{me08a}.
+
+\fig{raw} shows the  50th and 90th percentile
 of the time spent resolving issues
@@
 
-In that figure, the data is split out according to issues that were fixed in phase $Y$ after
-being introduced in phase $X$. The data is sub-divided into tables according to $X$;
+In \fig{raw}, the data is split out according to issues that were fixed in phase $P_r$ after
+being introduced in phase $P_0$. The data is sub-divided into tables according to $P_0$;
 i.e. according to {\em before, planning, requirements, design} or {\em  code}. 
@@
 for this threshold via the central limit theorem~\cite{maxwelldata}.} of issues {\em injected} in phase $X$ then
-removed in phase $Y$.
+removed in phase $P_r$.
+
+Nowhere in 
+these results do we see the kind of very large increases reported in the papers
+documenting DIE. Returning to
+the {\em before/integration test} result: 1.50 and 1.97 is far smaller
+than the scale ups seen in Figure~1.
 
@@
 illustrate how these resolutions increases (expressed as ratios of resolution time in the phase where they were first injected).
-The {\bf BLACK} and \textcolor{red}{{\bf RED}} bars show the increases for the  50th (median) and 95th percentile values, respectively.
-Those bars offer some support for the 
+The {\bf BLACK} and \textcolor{Gray}{{\bf Gray}} bars show the increases for the  50th (median) and 90th percentile values, respectively.
+Those bars offer little support for the 
 claim that early lifecycle errors are most expensive to fix.
 In all our results, the biggest scale ups are found with resolving
-issues introduced in the {\em before} phase. For example,
-resolving {\em before} issues during unit testing was 2.56, 4.37 times (at the 50th and 95th percentile, respectively)
+issues injected in the {\em planning} phase. For example,
+resolving {\em planning} issues during unit testing was 3.00 and 1.51times (at the 50th and 90th percentile, respectively)
 slower than resolving issues earlier in the lifecycle.
 
-Nevertheless, the scale up seen in 
-Figure~\ref{fig:raw} 
-is much less
-than that suggested by the DIE literature. Returning to
-the {\em before/integration test} result: 2.56 and 4.37 is far smaller
-than the scale ups seen in Figure~1. In fact, nowhere in 
-our results do we see the kind of very large increases reported in the papers
-documenting DIE.
+The left hand column of \fig{raw} shows the results of the Scott-Knott statistical tests. Note that nearly all
+the treatments have the same rank (``1''); i.e. usually there is no statistically significant difference in the 
+time to resolve issues. The only exception here is Planning:UnitTest where is ranked ``2'' but ever here, the scale up is merely
+a factor of 3, and not the exponential increase promised by classic reports of the delayed issue effect.
+
+Before moving on, we commend on one quirk of \fig{raw} which is the counter-intuiative
+nature of some of the results. Consider, for example, the ``Reqts'' results were
+the time required to fix issues actually tends to {\em decrease} the longer they are left in the system.
+In terms of explaining this result, the key thing is the left-hand-side statistically ranking: all these treatments were found to be statistically indistinguishable.
+In such a set of treatment, the observed
+difference may {\em not} be a causal effect; rather, it may just
+be the result of  random noise.
 
@@
 %\item
-%The delayed issue effect has been reported in around a dozen papers, at least half of which date back to the 1970s. 
 %\item
@@
 
-\subsection{External Validity}
-
-When discussing these results with colleagues, they have raised the following questions as threats to validity.
-
-{\em Question1:} The data collection   for this work does not 
-distinguished between ``low risk'' and ``high risk''
-issues; i.e. between easy-to-fix issues and other most-expensive fixes. 
-It is possible that such high risk issues might occur infrequently,  yet are still detrimental
-since  their resolution time is extremely large?
-
-{\em Answer1} If such outliers were common, they would appear in the upper percentiles of results. 
-As shown in the \textcolor{red}{{\bf RED}} bars of \fig{raw}, no such effect appears in the 95th percentiles of \fig{raw}.
-
- {\em Question2:}
-The data used in this analysis does not extend into post-delivery deployment. 
-Did this study failed to find the DIE effect since it does not arise until after deployment?
-
-{\em Answer2:} 
-As mentioned in \tion{earlyonset}, every other
+Threats to validity are reported according to the four categories described in Wohlin et al.~\cite{wohlin2012}, which are drawn from Cook and Campbell~\cite{cook1979}.
+
+\subsection{Conclusion Validity}
+Threats to conclusion validity are ``issues that affect the ability to draw the correct conclusion about relations between the treatment and the outcome''~\cite{wohlin2012}.
+We do not have a traditional treatment or control as in a classical experiment.
+Instead, we evaluate if the DIE holds in a modern data set.
+The data set is comprised of TSP projects, so the treatment could be misconstrued as TSP, but this is not that case as we do not have an experimental control to compare TSP against.
+
+\textit{Low statistical power}: Our data set is comprised of 47,376 defect logs.
+Figure~\ref{tab:phaseSamples} shows the number of samples in each Injection-Removal phase pair from Figure~\ref{fig:fix-phase-dist}.
+\todo{more once Lucas' email is answered.}
+
+\textit{Reliability of measures}: The base measures in this study described in \S\ref{sec:data-collection} are \textit{defects} recorded in TSP defect logs and \textit{time} reported in time tracking logs.
+The primary threats to the reliability of these measures are: that the definition of a defect varies between projects and that time is not reported accurately or consistently.
+The reliability of the time reporting is discussed in \S\ref{sec:data_integrity}.
+Time is reported on a level of minutes.
+We do not have an assessment of the error margin for time reporting.
+Some developers are less precise with time or estimates.
+Nonetheless defect time-to-fix must sum to less than the total effort in the phase and time log time stamps must be consistent with the time stamps and phase in the defect log.
+Based on time stamp consistency and conformance with Benford distributions, our experience suggests greater than 90\% of the time log data is recorded in real time.
+The fidelity and consistency of data will be subject of a future paper.
+
+We assume that each team has similar defect recording practices, and the TSP coaching provides guidance on what constitutes a defect.
+Nonetheless, individual developers and teams may apply their own internal rules for filtering defects, which would lead to inconsistent reporting thresholds among the projects in our sample.
+A related issue is that we assume developers correctly report in which phases a defect was injected and corrected.
+One points of variation is the measurement framework that identifies process phases and joins the effort to a size measurement framework.
+Individual projects may choose to implement a different framework, for example adding phases for specific types of development (for example, adding static analysis or special testing or a non-standard size unit).
+
+Certainly, if the defect and time reporting was done
+incorrectly in this study, then all our results must be questioned.
+However,  this issue threatens {\em every} study on the delayed
+issue effect-- so if our results are to be doubted on this score,
+then all prior work that reported the delayed issue effect should
+also be doubted.
+In TSP, developers are trained and supplied templates for defect and time tracking, all data entry is double-checked by the team TSP coach, and TSP demands that developers analyze their data to make process improvements.
+That is, TSP developers are always testing
+if their project insights are accurate. In such an environment,
+it is more likely that they will better identify the injection phase.
+
+\textit{Reliability of treatment implementation}: 
+Although TSP is not prescriptive about the development process, goals, or strategy, TSP provides precise guidance and training for data gathering.
+The guidance for logging time and defects is precisely defined.
+All tasks should be logged as the work is performed with a stopwatch.
+All defects that escape a phase must be logged.
+All data fields for each defect must be completed.
+
+There are a number of reasons to believe that the data are are consistent between developers and between projects.
+First, developers receive PSP training, during which instructors focus on complete and accurate data gathering.
+Second, each project that submitted data had a certified TSP coach responsible for evaluating  process adherence and submitting the data.
+Third, because the teams use their data to manage the projects the team is motivated to collect complete, accurate, and precise data otherwise the data gathering and analysis would be wasted effort.
+Fourth, process fidelity issues are apparent to the TSP coach as missing or inconsistent data (e.g.
+time and defect logs do not match, log entries have excessive rounding, or a developer is an outlier).
+Fourth, 15 of the projects received a TSP Certification in which process fidelity was evaluated independently by an observer and data analyst examining data internal consistency and consistency with  distributional properties known to consistent among all projects and team members.
+Fifth, all projects in this sample used the same data gathering tool.
+Nonetheless, some variations exist.
+
+\begin{figure}[ht]
+%\renewcommand{\baselinestretch}{0.7}
+\scriptsize
+\centering
+\begin{tabular}{rrrrrr}
+  \hline
+                & BeforeDevelopment & Planning      & Reqts          & Design           & Code \\ 
+  \hline
+ReqtsReview     &   3               &  \textbf{80}  & \textbf{2221}  &   0              &   0 \\ 
+  ReqtsInspect  &   1               &   5           & \textbf{2439}  &   0              &   0 \\ 
+  DesignReview  &   6               &   7           &  \textbf{29}   & \textbf{3617}    &   0 \\ 
+  DesignInspect &  \textbf{24}      &  \textbf{12}  &  \textbf{58}   & \textbf{5243}    &   0 \\ 
+  CodeReview    &  \textbf{75}      &   2           &   8            & \textbf{250}     & \textbf{5916} \\ 
+  CodeInspect   & \textbf{161}      &  12           &  \textbf{11}   & \textbf{357}     & \textbf{6045} \\ 
+  UnitTest      & \textbf{240}      &   \textbf{9}  &  \textbf{27}   & \textbf{1789}    & \textbf{2082} \\ 
+  IntTest       &  \textbf{72}      &   0           &  \textbf{15}   &  \textbf{77}     & \textbf{841} \\ 
+  SysTest       &  \textbf{50}      &   3           &  \textbf{12}   & \textbf{116}     & \textbf{486} \\ 
+  AcceptTest    &   5               &   0           &  \textbf{12}   &  \textbf{13}     & \textbf{120} \\ 
+   \hline
+\end{tabular}
+\caption{phase delay samples. Boldface indicated that a cell is included in Fig. 8.
+\todo{XXX Lucas. Check figure reference}.} 
+\label{tab:phaseSamples}
+\end{figure}
+
+\subsection{Internal validity}
+Threats to internal validity concern the causal relationship between the treatment and the outcome~\cite{wohlin2012}.
+Again, we do not consider TSP as a treatment, but we observe that the DIE does not hold in the TSP data set.
+Nonetheless, it is useful to consider threats to internal validity at an abstract level between the software engineering milieu that generated the original DIE observations and today's context where TSP was applied.
+
+\textit{History}: Many technological advances have occurred in the time between when DIE was originally observed in the late 70s and today.
+Processors are more powerful, memory is cheap, programming languages are more expressive, developer tools are more advanced, access to information is easier via the Internet, and significant evolutions in programming paradigms and software process have been realized in the past 40 years.
+In addition to the risk-oriented, disciplined nature of TSP, any or all of these additional historical factors may have contributed to the lack of the delayed issue effect in our data.
+
+\textit{Instrumentation}: The forms by which the TSP defect and time data are collected have been studied and matured over 20 years.
+Conversely, we do not find much documented evidence on how time and defects are reported for the original DIE papers (see the \S\ref{sect:doco}).
+Thus, we cannot be assured that reporting and data capture were not a significant influence on the delayed issue effect in the original papers.
+
+\textit{Interactions with selection}: As described in \S\ref{sec:data-collection}, all TSP teams are required to contribute time and defect data to the SEI, and thus there should be no selection bias in this sample compared to the overall population of TSP projects.
+However, there is likely selection bias in the teams that elect to use TSP compared to the entire population of software development teams.
+We do not have a basis for comparing TSP teams to those teams in which the DIE was originally observed.
+
+\subsection{Construct validity} \iton{sect:construct}
+Construct validity concerns ``generalizing the result of the experiment to the concept or theory behind the experiment''~\cite{wohlin2012}.
+Thus, do the observations in this paper provide evidence on the general delayed issue effect theory?
+
+\textit{Inadequate preoperational explication of constructs}: As described in \S\ref{sec:data-collection}, the measures of defect, time, and cost in the original DIE papers are not clearly defined.\footnote{In retrospect, empirical software engineering studies at that time were extremely rare, and guidance for reporting empirical case studies and experiments have improved substantially.
+One of the seminal books on quasi-experimention and reporting of validity concerns, Cook and Campbell~\cite{cook1979}, had not been published when most of the DIE papers were written.} Note that in Figure~\ref{fig:cost-to-fix}, the units of ``cost-to-fix'' are not expressed -- in the source references, cost appears as calendar time, effort, and price.
+In the TSP, a defect is defined as ``any change to a product, after its construction, that is necessaryto make the product correc'' and time to correct a defect includes ``the time to analyze, repair, and validate a defect fix''.
+Our analysis of DIE focuses on time as a measure of time-as-effort (persons * time).
+
+The data used in this analysis does not extend into post-delivery deployment. As mentioned in \tion{earlyonset}, every other
 paper reporting DIE also reported  early onset of DIE
@@
 This is an important point since  Figure~\ref{fig:raw} shows an {\em absence}
-of a very large DIE   effect during development
+of any  large DIE effect during development
 (in this data, the greatest increase in difficulty in resolving requirements issues was the 2.16 to 4.37
@@
 which is far smaller than the 37 to 250-fold increases reported in \fig{b81}, \fig{cost-to-fix},
-and \fig{steck}(.  
-That is, extrapolating the above data, we would not expect DIE effects, post-deployment.
-
-(As an aside, we have (informally) checked this prediction by following up with the projects for which we
-still have contacts. This follow-up study lacks the detailed data collection used above (it was more ``word of mouth''
-amongst the developers with personnel contacts with the second author). Based
-on that informal data collection, we can report that none of these 171  projects reported
-a  large post-deployment  
-DIE effect.)
-
- {\em Question3:}
- While the 171 project studies here are a large sample, are they representative of all software projects?
-
- {\em Answer3:} These 171 projects contain examples of a wide variety
- of systems (ranging from e-commerce web portals to  banking systems) run in a variety of
- ways (agile or  waterfall or some combination of the two). Hence, rather
- than being unduly  biased in some manner,
- we view our 171 projects as a large sample of current industrial processes for software
- development.
-
- {\em Question4:}
-
-Another way to discount this result is to say we are only reporting on small to medium-size
-projects
-(see
-\fig{dist}:  median
-effort was was  
-271 days). 
-There certainly exists another class of very large software project where early requirements
-errors would be catastrophic for later development. For example, when NASA launches deep space missions,
-each of those satellites are special-purpose, uniquely-built devices built by a large contractor
-population. Such systems are very hard to reconfigure halfway through the development.
-
-We concede that for very large complex NASA-like developments, the delayed issue
-effect could still hold. However, perhaps we need to split our software engineering principles
-into two parts- one for   very large NASA-like projects (which are not very numerous)
-and one for smaller-sized projects like the ones studied here (which are far more numerous).
-Most software is {\em not} like the software used on NASA deep-space satellites.
-As  agile methods grew more popular last decade; and as personal computers grew more powerful;
-and as the libraries associated with particular languages grew more extensive; then smaller
-teams starting  deliver systems that (previously) had required very large teams.
-As mentioned above,
-our  171 projects represents $\frac{171}{212}=80$\% of the software projects
-seen by the Software Engineering Institute's TSP team.
-That is, as far as we can tell,
-there are more small to medium-sized projects than otherwise.
-
-\subsection{Construct validity}
-
-Our definition of {\em difficult to resolve} combines two concepts: time to change and cost to
-change. In the above we have used them interchangably, comparing our time to resolve data (from the 171 TSP projects) against the cost-to-fix results of Boehm et al. e.g. \fig{b81}.
-Is it valid to assume the equivalence of time and cost?
-
-Certainly, there are cases where time is not the same as cost. Consider, for example,
-if debugging required some very expensive tool or the services or a very senior (and hence, very expensive)
-developer. Under those circumstances, time does not equate to cost.
-
-Also, there is the issue of the implication of  a defect.
+and \fig{steck}).
+
+\textit{Mono-method bias}: We only measure the delayed issue effect in terms of defects (as reported by teams) and time (in minutes of effort).
+To mitigate mono-method bias, additional measures of these constructs would be needed.
+For example, defects may be segmented into customer-reported defects and pre-release defects.
+In addition to time-as-effort, calendar time and price to fix (including labor, CPU time, overhead) would provide a more complete picture of the abstraction ``cost to fix a defect''.
+Further, there are no subjective measures of cost-to-fix, such as the social impact on the team or frustration of the customer.
+
+\textit{Confounding constructs and levels of constructs}: We do not consider the severity of defects in this analysis.
 Evidence discussed in \cite{Shull02} suggests that low severity defects may exhibit a lower cost to change.
 Nonetheless, even ``small'' errors have been known to cause enormous damage (e.g., the Mars Climate Orbiter).
-
-Having documented the above issues, we assert that they are unlikely to be major issues in the study.
-One of us (Nichols)
-was closely associated with many of the projects in our sample. He is unaware of any frequent
-use of exorbitantly expensive tools or people on these projects.
-
-\subsection{Internal validity}
-Confounding variables are a threat to any study.
-In this research, we rely on the substantial sample size of 171 TSP projects and over 47,000 defect logs to avoid confounding effects that may arise from the nature of the projects, teams, or defects.
-Similarly, the 10-year data collection period should help to ameliorate any maturation or history effects.
-As described in \S\ref{sec:data-collection}, all TSP teams are required to contribute time and defect data to the SEI, and thus there should be no selection bias in this sample compared to the overall population of TSP projects.
-However, there is likely selection bias in the teams that elect to use TSP compared to the entire population of software development teams.
-Further, we assume that each team has similar defect recording practices, and the TSP coaching provides guidance on what constitutes a defect.
-Nonetheless, individual developers and teams may apply their own internal rules for filtering defects, which would lead to inconsistent reporting thresholds among the projects in our sample.
-
-Another issue that threatens internal validity is that our
-results are dependent on developers correctly identifying which
-phase initially created an issue. Certainly, if that was done
-incorrectly in this study, then all our results must be questioned.
-However,  this issue threatens {\em every} study on the delayed
-issue effect-- so if our results are to be doubted on this score,
-then all prior work that reported the delayed issue effect should
-also be doubted. Moreover, the TSP method used in this study
-encourages greater care with error reporting:
-
-\bi
-\item Developers are trained   to make that judgement;
-\item All data entry is double-checked by the team couch 
-\item TSP demands that developers analyze their data to make
-process improvements. 
-\ei
-That is, TSP developers are always testing
-if their project insights are accurate. In such an environment,
-it is more likely that they will better identify the injection phase.
+It is possible that high-severity defects require more effort to fix simply because more people work on them, or conversely, low-severity defects may be fixed quickly simply because they it is easier to do so.
+High-severity defects are of particular concern in software projects, and even if the number of high-severity defects is low their cost to fix may be extremely large.
+If such outliers were common in our data, they would appear in the upper percentiles of results.
+As shown in the \textcolor{gray}{{\bf GRAY}} bars of \fig{raw}, no such effect appears in the 90th percentiles of \fig{raw}.
+That is, while such ``high risk'' issues may certainly exist, they were not a major effect in our data.
+
+\textit{Restricted generalizability across constructs}: While we observe a lack of DIE in the TSP dataset, we examine only the construct of time-to-fix.
+We do not consider the tradeoffs between time-to-fix and other ''-ilities'', such as maintainability.
+For example, a low time-to-fix may come at the expense of a more robust solution, i.e., a quick and dirty fix instead of an elegant repair.
+
+\subsection{External Validity}
+External validity concerns the generalizability of findings~\cite{wohlin2012} beyond the context of the study.
+Madigan et al.~\cite{madigan2014} and Carlson \& Morrison~\cite{carlson2009} discuss primarily external validity concerns drawn from studies of large datasets in medicine that are useful for identifying limitations in our study.
+
+\textit{Interaction of selection and treatment}: The most obvious limitation in our study is that the dataset in which we observed no DIE was composed entirely of TSP projects.
+TSP is a mature process constructed with risk mitigation as its primary purpose.
+We do not claim that our findings generalize beyond the projects using the TSP process.
+Similarly, we make no claims regarding generalizability across domains (e.g., defense, banking, games, COTS), scope (\# of features, people, and development length), or organizational features.
+The purpose of this study is to draw attention to the notion that commonly-held belief of the delayed issue effect may not be a universal truth.
+This study adds to the evidence offered by the cased study in Royce~\cite{Royce98}.
+Our study invites a further explanation into the causal factors that mitigate DIE.
+
+\textit{Interaction of setting and treatment}: The 171 TSP projects in our data set as well as the case studies in the original DIE papers were all industry projects conducted by software development teams.
+The TSP projects contain examples of a wide variety
+ of systems (ranging from e-commerce web portals to  banking systems) run in a variety of
+ ways (agile or  waterfall or some combination of the two)These are realistic settings for contemporary software development teams, though perhaps not representative of all types of projects (see prior paragraph).
+
+\textit{Interaction of history and treatment}: The TSP projects and the original DIE projects took place over several months or years of development.
+Thus, it is unlikely that the data are substantially influenced by rare events that occurred during project execution.
+
+ \section{Discussion}
+% Unexpected results such as this one require significant caution.
+In prior sections of this paper, we have discussed our quantitative analysis and its interpretation.
+We also survey the state of software development practice, especially in comparison to the environment in which the DIE was first observed, to assess whether the lack of an observed DIE should necessarily be so surprising.
+
+Earlier we noted that the delayed issue effect was first reported in 1976 in a era of punch card programming and non-interactive environments~\cite{Boehm76}.
+We also note that other development practices have changed in ways that could   mitigate the delayed issued effect.
+Previously,
+most software systems were large, monolithic, and ''write once and maintain
+forever''.
+Today, even large software systems are trending toward DevOps and cloud-based deployment.
+Advances in network communications, CPU processing power, memory storage, virtualization, and cloud architectures have enabled faster changes to software, even for large systems.
+Facebook deploys its 1.5 GB binary blob via BitTorrent in 30 minutes every day~\cite{Facebook}.
+% For example, 
+% Microsoft is adjusting its development practices towards a continuous
+% release paradigm. Theisen et al. reports their experiences where Microsoft architects
+% continually adjust  their modules in response to security issues~\cite{theisen2015}.
+Upgrades to the Microsft Windows operating system is moving from service patches and major releases to a stream of updates (so there will be no Windows 11- just a stream of continuous updates to what is currently called Windows 10)~\cite{bright15}.
+
+% For another example, consider the MEAN stack preferred for web development by
+% the continuous deployment community. Older architectures for web development
+% often used some variant of LAMP (Linux + Apache + Mysql + PHP) that was an intricate
+% combination of tools written in different languages. MEAN stacks, on the other hand,
+% use Javascript throughout-- which makes large scale reorganizations faster to complete~\cite{wayner15}.
+
+Even organizations that build complex, high assurance systems are turning to agile development processes that purport to address the DIE. In 2010, the US Department of Defense mandated
+that all new software acquisitions must adopt agile methods~\cite{kim13}. 
+This change in DoD policy is enabled by a separation of baseline architecture
+(e.g., the design of an 
+aircraft carrier) marked by significant up-front design and the agile development of applications within that architecture.
+For the baseline architecture, bad decisions
+made early in the life cycle may be too expensive to change and the DIE may still hold.
+However, smaller projects within the larger architecture (e.g., lift controls, radar displays) can leverage more agile, interactive development provided that interfaces and architectural requirements are well-defined.
+
+So, is it really surprising that DIE was not observed?
+Many software engineering technologies have been created precisely to avoid the delayed issue effect by removing risk as early as possible.
+Boehm's spiral model~\cite{boehm1988spiral}, Humphrey's PSP~\cite{humphrey1995discipline} and TSP~\cite{tsp00}, the Unified Software Development Process~\cite{jacobson1999unified}, and agile methods~\cite{Beck2001a} all in part or in whole focus on removing risk early in the devleopment lifecycle.
+Indeed, this idea is core to the whole history of iterative and incremental product development dating back to ``plan-do-study-act'' developed at Bell Labs in the 1930's~\cite{larman2003iterative} and popularized by W.
+Edwards Deming~\cite{deming1986out}.
+Technical advancements in processing power, storage, networking, and parallelism have combined with a deeper scientific understanding of software construction to enable a whole host of software assurance technologies, from early-phase requirements modeling to automated release testing.
+
+The delayed issue effect may continue to be prevalent in some cases, such as high-assurance software, architecturally complex systems, or in projects with poor engineering discipline.
+We do not have evidence for or against such claims.
+However, our data shows that the DIE has been mitigated through some combination of software engineering technology and process in a large set of projects in many domains.
+Our results are evidence that the software engineering community has been successful in meeting one of its over-arching goals.
+But our results raise an equally important point - should the DIE persist as a truism (see \S\ref{sect:belief}), or should is it a project outcome that can be controlled by software engineering process and technology?
 
@@
 In this paper, we explored   the papers and data related to the 
-widely-believed {\em delayed issue effect} (that delaying the resolution of issues
+commonly believed {\em delayed issue effect} (that delaying the resolution of issues
 very much 
@@
 Based on a  survey  of both researchers and practitioners, we  found that
-a specific form  of this effect (requirements errors are hardest to fix) is  widely-believed
+a specific form  of this effect (requirements errors are hardest to fix) is  commonly believed
 held in the community.
@@
 
-To be clear, we do not claim that this effect {\em never} holds in software projects; just that it cannot be assumed to {\em always} hold. Our explanation of the observed lack-of-effect is five-fold:
+% The data collected in this study comes from TSP-based developments. Hence, it is possible
+% that the reason we did not find DIE is that TSP effectively mitigates for this problem.
+% However, if DIE can be controlled by something as simple as TSP then it can no
+% longer be regarded as a {\em
+% ``major driver in focusing industrial software practice on thorough requirements analysis and design, on early verification and validation, and on up-front prototyping and simulation to avoid costly downstream fixes''} (as said by Boehm \& Basili in our introduction).
+
+We do not claim that this effect {\em never} holds in software projects; just that it cannot be assumed to {\em always} hold. Our explanation of the observed lack-of-effect is five-fold:
 \be
@@
 
-Our results beg the question: why did an idea, with so little support, become so widespread in the software engineering literature?
+Our results beg the question: why does the delayed issue effect persist as a truism in software engineering literature?
 No doubt the original evidence was compelling at the time, but much has changed in the realm of software development in the subsequent 40 years.
@@
 has persisted because, to use Glass's terms on the subject, it seems to be ``just common sense''\cite{glass02}. 
-Nevertheless, in a rapidly changing field such as software engineering, even widely-held rules of thumb must be periodically re-verified. 
+Nevertheless, in a rapidly changing field such as software engineering, even commonly held rules of thumb must be periodically re-verified. 
 Progress in the domain of software analytics has made such periodic checks more cost-effective and feasible, and we argue that an examination of local behaviors (rather than simply accepting global heuristics) can be of significant benefit.
@@
 In particular, we thank Tuma Solutions for providing the Team Process Data Warehouse software.
-Also, the authors gratefully acknowledge the careful comments of anonymous reviewers-- particularly, their comments on issues relating to ``cost''
-and ``time''.
+Also, the authors gratefully acknowledge the careful comments of anonymous reviewers from the
+FSE and ICSE conferences.
 This work was partially funded by an National Science
@@
 \bibliography{refs}
+
+\newpage
+\pagenumbering{arabic}
+ \setcounter{page}{1}
+
+\input{tex/responses}
 

### 2016-06-15 tex/responses.tex

 \section*{Response to Reviewers}

{\bf Editor's comments}

Dear authors,

Three knowledgeable reviewers have scrutinized the manuscript
and found it interesting and relevant,  although in need for
certain improvements. Therefore, the authors are
 invited to submit a revised version for a new review, where
 the reviewers' comments
are addressed.

\begin{quote} {\em Thank you for the careful reviews. Using
those comments we have extensively revised and extended this draft.
We trust you will find this draft gratefully improved}
\end{quote}

Particularly, this holds for:

1) the connection to activities of theory building in ESE
(reviewers 1).

\begin{quote}{\em This new draft  now maps our work into the SE literature on theory building.}\end{quote}

2) clarification about the data origin and
characteristics (reviewer 1).

\begin{quote}{\em
This new draft  now maps not contains extensive notes on the origin and characteristics of our data.}\end{quote}

3)  use of statistical analysis (reviewers 2 and 3)

\begin{quote}{\em All our results are now augmented with statistical significance tests as well as effect size tests. Those tests greatly strength the overall message of this paper-- that there is no significant
phase delay effect.}\end{quote}

- writing style, that would benefit form being more
precise (reviewer 3)

\begin{quote}{\em We have carefully read those comments  nd have made numerous changes
as directed by reviewer3.}\end{quote}

{\bf Reviewer \#1's Comments}

The paper investigates whether a widely
accepted "rule" in software engineering holds by looking at
a large set of empirical data. The research topic is highly
relevant and the results are interesting.   The discussion
on reassessing old truism fits the scope of the journal very
well.  The language in the paper is excellent, and the
structure is (although non-standard) easy to follow.

\begin{quote}{\em Thank you for those kind comments.}\end{quote}

It is not easy for the reader to understand what parts
of the "life of an issue" the authors claim are affected by
DIE, and it is also not too easy to understand what data the
authors have available.

\begin{quote}{\em  We have added an example of the DIE in the Introduction, and more precisely defined how we measure this effect Section 6.\footnote{Lucas- where exactly?} }\end{quote}

 The authors do not relate their reassessment of an
"old truism" to the growing set of papers on theory building
in software engineering. With access to such large amounts
of empirical evidence, it should be possible to take some
steps toward improved DIE theories.

\begin{quote}{\em  fshull XXX }\end{quote}

The different projects studied are all grouped together. I
believe studying clusters of development contexts would be
highly interesting, but unfortunately the current
characterization of projects is inadequate.

\begin{quote}{\em  We agree with this comment and so took two actions for this draft-- one that was effective and one that was ineffective.

Firstly, this draft now contains extensive notes on the different contexts of the
projects studied here. Those notes are offered in \tion{data_character} and \tion{171projects}.

Secondly (and this was the ineffective step), we  searched the numerous   listed in \tion{data_character} and \tion{171projects}
in order to check if any stratification contained the delayed issue effect. We failed to find a stratification
containing an exponential expansion in the time to fix issues. The reason for this was due to the dataL those
stratifications resulted in smaller sample sizes that exacerbated the large IQR's seen in this data
(to see these large IQRs,  please consider the 50th and IQR columns of Figure~\ref{fig:fix}.IQR is the
range 75th to 25th in the data. Note that most of our IQRs were larger than tan the 50th percentile; i.e. software
data exhibits large variances-- where are exacerbated by the smaller samples seen in the stratifications
 }\end{quote}

DETAILED REVIEW: The authors should expand on what they
include in the "delayed issue effect" (DIE), in particular
in relation to standard milestones in the life of an issue
and the corresponding timespans.  I'm particularly
interested in whether all "indirect" issue resolution steps
are covered, e.g., change impact analysis, updating
documentation, and regression testing. The effort involved
in these steps clearly depends on the development context of
the specific project. The short a-c) list on page 14:50
suggest that the issues studied are restricted to minor
fixes, i.e., no updates to architecture, changed hardware,
recertification, updated user manuals etc.

\begin{quote} {\em

We understand your comment here to be ``are we just reporting trivial errors?''
As shown in \fig{dtypes}, it is true that around a quarter of the fixes were simple documentation changes. That said, 75\% of the changes are quote  elaborate;   e.g. fixes to function necessitates a careful reflection of the purpose of the code.
}\end{quote}

Section 6.3 describes the data, but some aspects
should be further explained.

\begin{quote} {\em
At your suggestion, we had added extensive notes on the nature of our data--
see \tion{data_character} and \tion{171projects}.
}\end{quote}

Logging interruption time
sounds very useful, but I wonder how carefully the
developers actually did this 
 it must be really hard to keep up the discipline required.
 If you are interrupted as a developer (e.g., phone calls,
 urgent mail, someone asks a question) I don't think the
 first thing you do is to stop a timer on your computer.
 Moreover, developers often work on several defects in
 parallel, and might interweave bug fixing with new
 development. I don't think the "interruption time" captures
 all such multi-tasking, and it should at least be properly
 discussed in Section 7 "Threats to validity".

\begin{quote}{{\em 
Yes, there is undoubtedly some accuracy in the time-tracking. We provide some evidence on the TSP data set's integrity in Section 6.3, and a more extensive discussion on the issue of time-tracking and defect logging accuracy in Section 7.1 on Conclusion Validity.XXX lucas: please check the section numbers are currently
correct.
}
\end{quote}

The discussion on reassessing old truisms (Section 3) is
interesting, but it should be complemented by the
perspective of theory building in software engineering - An
active research topic lately, with a dedicated workshop
series (GTSE). I suggest looking into the following
references for a start: Sjøberg et al. (2008) "Building
Theories in Software Engineering", Smolander and Päivärinta
(2013) "Theorizing about software development practices",
and Stol and Fitzgerald (2015) "Theory-oriented software
engineering". Considering new empirical evidence is
obviously critical to theory building, and discussing your
new results in the light of an theory creation would be
valuable. The authors mention that DIE appears to occur
intermittently in certain kinds of projects - maybe the
authors could elaborate on this idea and present an improved
DIE theory based on what they now know? I believe the
authors have the best available data to do so, and I would
expect the paper to go beyond simply questioning the "old
truth".

 \begin{quote}{\em 
XXX Fshull
}\end{quote}

The authors study three claims in this paper, and the
third claim is the central one: "delayed issues are not
harder to resolve". To study whether issues require longer
resolution times in later phases, the authors analyze a
large set of issue data from historical projects. Thus the
manuscript reports from an observational study rather than
an experiment with a controlled delivery of treatments.
While I believe the authors' approach is practical, I would
like to see a critical discussion on threats to validity of
observational studies (e.g., Madigan et al., A Systematic
Statistical Approach to Evaluating Evidence from
Observational Studies, Annu. Rev. Stat. Appl. 2014. 1:11-39
and Carlson and Morison, Study Design, Precision, and
Validity in Observational Studies, J Palliat Med. 2009 Jan;
12(1): 77-82). An alternative study design (although
difficult to realize in a system of industrial size) would
be to let different developers resolve the same issues for
the same software system, i.e., one group resolves an issue
during design, another during implementation, and a third
during testing - some discussion along these lines would
strengthen the validity section.

\begin{quote}{\em 
 We have reviewed and cited the mentioned papers. We have incorporated many of the concerns raised in these papers into our expanded Section 7.4 on External Validity.
XXX lucas: please check the section numbers are currently
correct. XXX section 7.4 is a little... terse. can you add another 1/4 to 1/2 page of stuff? or is this sectionn just as large as it neeeds to be
}\end{quote}

The first and second claims are studied with much less
rigor. "DIE is a commonly held belief" is studied using a
survey of software engineers, both practitioners and
researchers. According to Fig. 2 (actually a table), the
number of respondents is 16 and 30 for practitioners and
researchers, respectively. Sixteen respondents from industry
represent a tiny survey of a very general claim that any
software engineer could respond to. Why were not more
answers collected? The authors do not have much evidence to
defend the first claim, and there is no discussion of the
corresponding validity threats. The second claim, "DIE is
poorly documented", is studied using a literature review.
Unfortunately, the method underlying the literature review
is not presented. Although it doesn't need to be an SLR of
the most rigorous kind, the authors should report how the
papers were identified. I suspect the terminology used to
describe the DIE phenomenon is highly diverse, thus it would
strengthen the paper if the authors reported how they
reviewed the literature. According to Page 19:34 only eight
publications were identified.

\begin{quote}{\em XXX
Lucas: need to show that these were the only empirical
results for DIE in tow
}\end{quote}

The iterative fashion of modern software development,
with agile at the extreme end, is not fully discussed (the
short discussion section could be extended). The phases of
the linear development of the 80s, (such as in Fig. 1)
probably don't exist in many of the projects in the TSP
dataset, still the authors discuss the DIE-effect from the
perspective of the 80s. Page 16:31 states "DIE has been used
to justify many changes to early lifecycle software
engineering" - does this mean the agile movement
successfully mitigated DIE? This possibility is not fully
considered in the paper.
How many of the 171 projects were
(more or less) agile? This appears to be an important
characteristic of the included projects - very important to
describe!

\begin{quote}{\em An astute observation. We provide more information on the types of projects included in our sample in Section 6.4. XXX Lucas. Check section number.
XXX Lucas: can't see the "percent agile" in 6.4. Can you work this with Bill?

More to the point -- does agile help mitigate DIE? The answer is - we don't know. A strong argument can be made that all iterative&incremental methodologies are meant to mitigate the DIE. Further, huge advances in software engineering technology and process have occurred since DIE was first observed, many of which have precisely targeted the DIE (static analysis, test-first development, automated build). Perhaps our analysis provides evidence that DIE can be defeated, at a large scale across multiple varied projects, using modern technologies. We do not yet know the causes here, but certainly it appears the DIE should not be treated as a truism any more, but rather a variable that can be controlled. We discuss this in Section 8.
}\end{quote}

Concerning characterization of the 171 projects, the
paper needs to report much more detail. I would expect to
see some descriptive statistics.   The paper does not report much
characterization of the 171 projects. I strongly suggest the
authors to dig deeper into the data, and analyzing for which
types of projects the findings hold. What patterns are there
to discover? I suggest Petersen and Wohlin, "Context in
industrial software engineering research" (ESEM2009) for
details on how to characterize development contexts.

\begin{quote}{\em Those notes are offered in \tion{data_character} and \tion{171projects}. \end{quote}

Fig 1: "Operation" dwarfs everything in Boehm's diagram.
Since you do not study anything post release in this paper,
I think this figure skews the reader's mental picture. If
you remove the rightmost bar and rescale accordingly, the
plot better matches the findings you report in Fig. 10. You
still identify an interesting result though, but the
presentation turns fairer.

\begin{quote}{\em We understand why you are making that suggestion.
That figure is something of a classic. It is appropriate we chop it like that?}.
\end{quote}

MINOR COMMENTS: 
Several figures are copied from previous work. Are all copyrights properly
managed?

\begin{quote}{\em We recently published a text on classic SE papers. We know how to manage those issues.}\end{quote}

\fig{raw}: Black and red is not a good pair for
grayscale printouts. Please use black and gray instead.
\begin{quote}{\em  We agree: switched to black and gray.  }\end{quote}

Some figures should be resized to better match the page width.

\begin{quote}{\em  We agree: fixed. }\end{quote}

Page 3:14 - "The above argument" Which argument? Could be precisely specified.

\begin{quote}{\em  We agree: that reference has been deleted. }\end{quote}

Page 3:22 - Please provide the full link to the dataset.

\begin{quote}{\em  Our policy is to get one publication out of paper
before issueing the data in public. If this paper is accepted, we will certainly add that link. Note that we have a long history of living up to that policy-- see http://openscience.us/repo. }\end{quote}

Page 18:26 - First sentence: "Unexpected results such as this one". Ambiguous reference,
please be specific.  
\begin{quote}{\em  We agree. That back reference is removed. }\end{quote}

\begin{quote}{\em XXX Lucas. Not sure. but do you hanle all the following? }\end{quote}

Page 18:27 - "We also survey the state of SW dev. practice /---/": given the size of the survey,
this statement feels a bit bold.

TYPOS ETC.  
Sec 1, §1: [3] and [30] have been swapped?  
Page 14:46 - Spell out IV\&V the first (and only) time it appears.
Page 3:24 - missing verb (is) Page 16:43 - "Did this study failed" 
Page 18:44 - Appache 
Page 19:19 - Al Ref 26 - oo --> OO

{\bf Reviewer \#2's comments}

The topic of this paper is the empirical
evaluation of a largely admitted claim in software
engineering, i.e. the exponential cost of correcting errors
according to the phase in which the errors are discovered.
The claim is first confirmed by surveying and interviewing
practitioners and experts. Next, the authors use a large set
of data from projects in SEI database. The results show
that, strictly speaking, the claim does not hold (although
some effects of delayed corrections can be noticed).

Globally, the paper is well written and the authors develop
a convincing demonstration. The topic of the paper is highly
relevant for both practitioners and people from academia.
Among many, I personally believed in this claim and had
regularly taught it in my courses. Thus, reading this paper
offers a refreshing perspective on our understandings of
software engineering background and theoretical knowledge. I
also like the suggested posture that insists on the
necessary skepticism that we should have against such
prevailing claims.

Beyond this positive global impression, I have some concerns
about the study reporting and analysis: -  It seems to me
that more descriptive statistics about the sample would
contribute to a better understanding of the scope of the
study.

\begin{quote}{\em 
Those notes are offered in \tion{data_character} and \tion{171projects}.}\end{qupte}

I think in particular about the size (lines of codes and/or
 number of software components); a duration histogram would
 also be relevant. Accordingly, the formula for calculating
 variable "Total effort" in Fig. 7 could be explained, and
 Figure 9 made more readable (bigger size).

 \begin{quote}{\em We have added more description of the projects to Section 6.4. We have also added more precise definitions of our measures in Section 6.2. We have adjusted the size of figures throughout the paper to be more readable.XXX Lucas please check section references. We have added more text 
 to explain {\em effort} in \fig{dist}.}\end{quote}

If certain attributes for the issues and errors are
available in the data (e.g. severity, priority, etc.), they
should also be brought to the readers knowledge.

\begin{quote}{\em 
Those notes are offered in \tion{data_character} and \tion{171projects}.}\end{qupte}

Figure 10 is central to the paper's demonstration; its
expressiveness could be enhanced: the formula for calculating the 50th
and 95 ratio could be provided,  the unit of column
"Percentile" could be mentioned.

\begin{quote}{\em
\fig{raw}'s caption now has extensive explanation text. e.g. units
are added on line two. Also, an example of the calculations of the ratios is offered
in the last line of the caption. }\end{quote}

I like the discussion section, in particular, the idea
of software architectures that could be a contributing
factor for enhancing software evolution and reducing the
cost of issues and errors fixing. Is there any available
knowledge in the data set concerning any architectural
design choices made in each project?

\begin{quote}{\em
Not explicitly, but \fig{dtypes} lists "interfaces" as a the fourth most
common type of defect. Note that "interfaces" are changed by architectural
issues. }\end{quote}

In the same trend of ideas, I think what has
fundamentally changed since early days of SE is the
development of requirements engineering. In actual SE
practices, problem and the solution spaces are
systematically explored thanks to RE techniques; together
with architecture, this could also explain why things have
changed since the 70'. It could be, for example, that people
tend to make less severe errors in early software project
phases (thanks to RE techniques); a similar phenomenon was
observed when SE practices become more mature (see Harter et
al. 2012).

References Harter D. E., Kemerer C. F., Slaughter S. (2012).
Does Software Process Improvement Reduce the Severity of
Defects? A Longitudinal Field Study. IEEE Transactions on
Software Engineering, vol. 38, n° 4, p. 810-827.
```

\begin{quote}{\em
  Absolutely - there are potentially many explanations for why the DIE was NOT observed in our dataset. We discuss some more of the potential causes in Section 8.
  XXX Lucas. Is the section reference right? Do we reference Harter (and if we do, mention that here).?
 }\end{quote}

In the conclusion section, I feel uncomfortable with
the assertion "That data held no trace of the delayed issued
effect" (line 17). As mentioned elsewhere in the paper, the
delayed issue is not absent; it is much less significant and
systematic than what is usually claimed. Moreover, this
reduction has been demonstrated for medium sized projects;
we cannot generalize to larger projects.  
\begin{quote}{\em
  We agree -That line has been removed.
 }\end{quote}

\begin{quote}{\em
 XXX Lucas: can you retire the follwoing
 }\end{quote}

Minor remarks: - p.9, line 29 : " … reported by Shull [52], found that the cost to find certain non-critical classes of defects …" => is it the "cost to find" or the "cost to FIX"?  
```

- [X]  llayman It should be "cost to fix". We have corrected this.

- p.10, line 21: duplicate word "there" -
p.12, line19: missing word "One of THE guiding …" - p.16,
line 33: "distinguish" instead of "distinguished"?  - p.18,
line 42: explain acronym MEAN - p.20, line 20: useless
enumeration A1?

{\bf  Reviewer \#3's Comemnts}:

I like the idea of the paper and I certainly
would like to see more empirical studies that periodically
check if our beliefs about software engineering practices
still hold (or even if they have ever held). The authors
address one of the beliefs that are more entrenched in the
population of researchers and practitioners, and a very
important one too.

\begin{quote}{\em
Thanks for those kinds words.
 }\end{quote}

At the same time, I think that the paper needs to be
improved before it can be published.

The empirical analysis (Section 6) is a bit of a let down. I
would like to see some sort of more robust and rigorous
statistical analysis, based on statistical tests.

\begin{quote}{\em
We agree. \tion{stats} lists our stats methods and \fig{raw}
shows the results (see the left-hand column: all those ``1''s mean
that our stats could not find any difference between the fix times for
delaying or not delaying.
 }\end{quote}

The paper seems to be written in a somewhat casual style,
which makes the paper more pleasant to read, but less clear
in some parts (see Detail Comments below).

\begin{quote}{\em
Thanks for those pointers. We will fix.
 }\end{quote}

DETAIL COMMENTS

Section 2

I'm not really sure what the authors mean by "We say that a
measure collected in phase 1, ., i, .. j is very much more
when that measure at phase j is larger than the sum of that
measure in earlier phases 1 <= i < j." You are defining a
property of a measure, but, the way this definition is
written, it seems as if you are defining the property of a
measure of "being much more," without any further
qualifications.  I can guess that you mean that a measure
$m$ (collected in phases, so you can denote by $m(i)$ the
value of $m$ in phase $i$) has this property *in phase $j$*
(so, it's a property m has in a specific phase and not a
general property of the measure) if $\sum_{1 <= i < i} m(i)
< m(j)$. Then, it's up to you to make this a property of the
measure, for example by using an existential "policy" ($m$
has this property if there exists one phase $j$ where
$\sum_{1 <= i < i} m(i) < m(j)$) or a universal one ($m$ has
this property if for all phases $1<j$, $\sum_{1 <= i < i}
m(i) < m(j)$).
\begin{quote}{\em
Thanks for that. We added the $m_j$ notation to Section 2, as you suggested.
 }\end{quote}

Your definition of "difficult issue" is "Issues are more
difficult when their resolution takes more time or costs
more (e.g. needs expensive debugging tools or the skills of
expensive developers)." That's not a very precise
definition, though, because it has two different
interpretations, one in terms of time and the other in terms
of effort. The reasons why you introduce this definition
become clearer only in Section 7.2, in the Threats to
Validity, and that's too late. You should move some of the
discussion of Section 7.2 here.
\begin{quote}
{\em Good idea! We did just that- see middle para in Section 2.}
\end{quote}
Why would the term "delayed issue effect" be a
generalization of the rule "requirements errors are hardest
to fix"? "Delayed" seems to be quite specific as it appears
to refer to time, while "hardest to fix" may refer to other
variables, like effort or cost.
\begin{quote}
{\em Not sure if this a reviewer requirement for something to be fixed
prior to publication, or a clarification question. We used these
terms since when we ran the survey, the audience clearly made
that generalization. Does this reply adequately to your comment?}
\end{quote}
In Claim 3, "very much more harder" sounds like a
little bit too much ... ("more harder"?)
\begin{quote}
{\em Our bad. We fixed that.}
\end{quote}
Section 4

A very minor issue: why is Figure 2 a ... figure, instead of
a table? Same for all other tables ...
\begin{quote}
{\em Authors' preference. When a document has many tables and figures, navigating
between them is hard. Calling them all "figures" makes navigation faster. 
But do you want us to change that?}
\end{quote}
Section 5.1

This is not a complete sentence "All the literature
described above that reports the onset of DIE prior to delivery.''

\begin{quote}
{\em Fixed}
\end{quote}

Section 6.3

You should rephrase your definition "a defect is any change," since a defect is what existed before the change was made and not the change itself.
\begin{quote}
{\em We now provide a more precise definition of defect in Section 6.2.
XXXX= Lucas: check section number.
}
\end{quote} 
You do not explain that "QualTest" in Fig. 8 is.
\begin{quote}
{\em Removed}
\end{quote}
The definition of "time per defect - The total # of defects found in a plan item during a removal phase divided by the total time spent on that plan item in that phase." is not correct as it is, as this would basically be the number of defects per unit time, instead. The roles of time and
defects should be reversed.
\begin{quote}{\em We now provide a more precise definition of time measurement related to defects in Section 6.2.}\end{quote}

Section 6.4

Besides adding a more rigorous statistical analysis, you
need improve your explanations.
\begin{quote}{\em We agree. \tion{stats} lists our stats methods and \fig{raw}
shows the results (see the left-hand column: all those ``1''s mean
that our stats could not find any difference between the fix times for
delaying or not delaying.}\end{quote}

Figure 9. The caption says "Distribution of defects found
and fixed by phase." Is that the distribution of defects
found in some phase and fixed in that phase? The text says
"The distribution of defects found and fixed per phase in
our data is shown in Figure 9. A high percentage of defects
(44\) were found and fixed in the early phases, i.e.,
requirements, high level design, and design reviews and
inspections." which doesn't add much. If that's the
distribution of defects introduced in a phase and fixed
immediately in that phase, it is not clear why you introduce
it, though.
\begin{quote}
{\em XXXX Bill: one for you.}
\end{quote}
Figure 10. The explanation of the data and histograms in the
figure is, at best, confusing. The caption says "50th and
95th percentiles of issue resolution times" and the opening
sentence of Section 6.4 says "Figure 10 shows the 50th and
95th percentile of the time spent resolving issues ..."
However, these are not "times," because you write "expressed
as ratios of resolution time in the phase where they were
first injected" a couple of paragraphs below. This seems to
be the right interpretation, but right after that you write
"The BLACK and RED bars show the increases for the 50th
(median) and 95th percentile values, respectively," which
would be a third interpretation of the data in Figure 10.
\begin{quote}
{\em \fig{raw} now has an extensive help text in the caption.
We used your comments there where we noted the ratios used in the
bard charts are unitless while the left-hand-side column numbers are
times in minutes.
}
\end{quote}

You need to at least mention (or, better, discuss) some of
your results, because they appear to be "counterintuitive,"
since they show that, for example, median resolution times
can even decrease if issues are not fixed immediately, but
in later phases. For example, looking at the Reqts section,
most percentages are way below 1, which seems to indicate
that some issues are actually much easier to fix at later
stages (maybe because more or better information about the
software system is available only later). That would be a
very interesting result by itself, especially if you can
provide some statistical support for it.
\begin{quote}{\em
The last paragraph of \tion{171projects} now comments
on that effect. }
\end{quote}

Section 7.1

Maybe you could analyze the projects developed in a
"traditional" way and check if there is some sort of DIE
there. For example, if you look at the projects developed
with the waterfall life cycle, maybe you will find larger
DIE than for the other life cycles, which might partially
justify the claims of previous researchers about the
existence of DIE. This would also show that newer kinds of
life cycles (like Agile ones) help solve DIE. You do provide
hints about this in Section 8 ("We also note that other
development practices have changed in ways that could
mitigate the delayed issued effect") and Section 9, but
that's clearly not enough.
\begin{quote}{\em
If your comment is along the lines of ``look for stratifications to check if
DIE exists in subsets'' then we offer the following failed study.

We looked into those stratifications and have  failed to find a stratification
containing an exponential expansion in the time to fix issues. The reason for this was due to the data in  those
stratifications resulted in smaller sample sizes that exacerbated the large IQR's seen in this data
(to see these large IQRs,  please consider the 50th and IQR columns of Figure~\ref{fig:fix}.IQR is the
range 75th to 25th in the data. Note that most of our IQRs were larger than tan the 50th percentile; i.e. software
data exhibits large variances-- where are exacerbated by the smaller samples seen in the stratifications.}
\end{quote}
Section 8

- [ ] Verify that the Section #'s in our responses are accurate prior to submission.

### 2016-06-17 phaseDelay4Emse.tex
Updates from ShareLaTeX

@@
 Such a large and growing effort should be managed and optimized via  well-researched conclusions.  
-
-It is standard practice
+To assist in achieving this, there has been a growing recognition within the software engineering research community 
+of the importance of theory building ~\cite{Sjoberg08, Paivarinta15, Stol15}. A good theory allows empirical research to go beyond simply
+reporting observations and instead provides explanations for why results are observed ~\cite{Stol15}.
+This occurs by testing theories against data from multiple sources; by reconciling similarities and differences in results it can be determined what factors need to be accounted for in a theory ~\cite{Shull08}.
+Theory-building needs to be an iterative process,
+in which results from practice are used to refine theories and theories are used to inform future observation and data collection ~\cite{Paivarinta15, Stol15}.
+It is no coincidence that it is standard practice
 in other fields, such as medicine,
-to continually revisit old conclusions~\cite{prasad13}.
+to continually revisit old conclusions in the light of new theories~\cite{prasad13}.
+
 Accordingly, this paper revisits
-the commonly held belief we label {\em delayed issue effect} (hereafter, DIE):
+the commonly held belief we label the {\em delayed issue effect} (hereafter, DIE):
 % Later in this paper, we offer a precise definition for this effect.
@@
 
- Basili \& Boehm comment that, since the 1980s,  this effect
+This theory has been used as the basis for decision-making in software engineering. For example,  Basili and Boehm comment that, since the 1980s,  this effect
  \begin{quote}
@@
 \begin{center}
-\includegraphics[angle=270,width=4in]{img/b81.pdf}
+\includegraphics[,width=4in]{img/b81.png}
  \end{center}
- \caption{An example DIE effect. From Boehm'81~\cite{Boehm81}. }\label{fig:b81}
+ \caption{A widely-recreated chart of the DIE effect. Adapted from Boehm'81~\cite{Boehm81}. }\label{fig:b81}
  \end{figure}
-Other prominent authors have commented on its perceived usefulness as a rule of thumb for software engineers.
-McConnell mentions it as a "common observation" in the field and  summarizes the intuitive argument for why it should be so:
+
+Like any good theory, DIE includes a rationale for why the expected results would be seen.
+McConnell mentions it as a ``common observation'' in the field and  summarizes the intuitive argument for why it should be so:
 \begin{quote}
@@
 \end{quote}
-Glass also endorses this effect, asserting that ``requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' is ``really just common sense''~\cite{glass02}.
+Glass also endorses this rationale, asserting that ``requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' is ``really just common sense''~\cite{glass02}.
 Other researchers
 are just as adamant in asserting that that the delayed issue effect is a generally useful law of software engineering.
-For example, what we call the delayed issued effect was listed first by Boehm and Basili in their "Top 10 list" of ``objective and quantitative data, relationships,
+For example, what we call the delayed issued effect was listed at \#1 by Boehm and Basili in their ``Top 10 list'' of ``objective and quantitative data, relationships,
 and predictive models that help
@@
 In analyzing data from a contemporary set of software development projects, however, we did not find results to corroborate these claims. While the delayed issue effect might have been a dominant
-effect decades ago, perhaps not so for $21^{\mathit{st}}$
+effect decades ago, this does not mean that it is necessarily so for $21^{\mathit{st}}$ century
 software development. 
@@
 After that, we  analyze 171 software  projects developed in the period 2006--2014 and find  {\em no evidence} of the delayed issue effect.
-Finally, we discuss the validity and implications of our results, as well as possible reasons for the lack of observed effect given the state of the practice.
+Finally, we discuss the validity and implications of our results, as well as possible reasons for the lack of observed effect given the state of the practice - reasons which, when subjected to further testing, may prove useful for refining the theory.
 To ensure reproducibility,
@@
 \section{Definitions \& Claims}
-\todo{Revisit these definitions after reviewing the rest of the paper for consistency of terms}
+\label{sect:claims}
+
 This paper uses the following definitions:
@@
 
-This paper defends the  following two claims and one hypothesis. Note hypotheses are defended
+This paper defends the following claim and hypothesis. The hypothesis is defended
 using some statistical
-significance test while the claims are  supported via a variety
-of arguments presented later in this paper.
-
-{\bf  Claim1: ``DIE'' is a  commonly held belief.}
-Using a literature review, we confirm that there are numerous historical papers (dating back decades)
-that endorse the DIE. Also, using a survey of practitioners and researchers, we find  that 
- the DIE  appears as the
-most strongly-held belief amongst software engineers.
-
-{\bf  Claim2: ``DIE'' is poorly documented.}
- As discussed in our literature review,  many of the papers reporting the DIE
+significance tests while the claim is supported via a variety
+of arguments.
+
+{\bf  Claim: ``DIE'' is a  commonly held, yet poorly documented belief.}
+We examine the literature promoting the DIE and find that most reference a few primary sources. Many of the papers reporting the DIE
  either (1)~are quite old (papers dating from last century);
@@
 (3)~or cite data sources that can no longer be
-confirmed.
+confirmed. We follow-up with a short survey that finds that DIE appears as the most strongly-held belief among software engineers in our sample.
 
@@
 \section{ Reassessing Old Truisms}
-Beliefs in general
-principles of software
-engineering are common to both research and practice. Professional societies assume such generalities exist when they offer
- lists of supposedly general ``best practices'' such as
+General
+theories of software
+engineering principles are common to both research and practice, although not always explicitly stated.
+Such theories underlie  lists of proposed general ``best practices'' for effective software development, such as
 the IEEE 1012 standard for software verification~\cite{1012}. 
@@
 Passos et al.~\cite{passos11} warn that developers
-usually assume that the truisms they learn from a few past
-projects are general to 
-all their future projects. They comment ``past experiences were taken into account without 
+usually develop their own theories of what works and what doesn't work in creating software, based on experiences from a few past
+projects. Too often, these theories are assumed to be general truisms with widespread applicability to future projects. They comment ``past experiences were taken into account without 
 much consideration for their context''~\cite{passos11}.
@@
  conclusions and software projects  (see examples in~\cite{jorgensen09}).
-Just as in medicine, 
+ A similar effect is reported by
+Devanbu et al.~\cite{prem16}  who examined responses from 564 Microsoft software developers from around
+the world, they found that  ``(a)~programmers do indeed have very
+strong beliefs on certain topics; (b)~their beliefs are primarily formed
+based on personal experience, rather than on findings in empirical
+research; (c)~beliefs can vary with each project, but do not necessarily
+correspond with actual evidence in that project''.
+Devanbu et al. further  comment that ``programmers give personal experience
+as the strongest influence in forming their opinions''. This is a troubling
+result, especially given the above comments from Passos and  J{\o}rgensen et al.~\cite{passos11,jorgensen09} about how quickly practitioners form, freeze, and rarely revisit those opinions.
+
+From all we above we conclude that, just as in medicine, 
 %our field suffers when
@@
  it is important for our field
- to regularly  reexamine    old beliefs such as the delayed issue effect.
-
-\section{``DIE'' is a  commonly held belief}
+ to regularly  reassess old truisms  like the  delayed issue effect.
+
+\section{Motivation: ``DIE'' is commonly held, yet poorly documented}
 \label{sect:belief}
@@
 % understanding of SE theory, then the DIE  may well influence their decisions.
-
-We conducted two surveys of software engineers to assess the prevalence of DIE. 
-The surveys collected data on software engineers' views of DIE and other commonly held software engineering ``laws''.
+One reason that industrial practitioners and academics believe so strongly in the delayed issue effect is that it is often referenced
+in the SE literature. 
+%For example,
+%we know that \fig{b81} has been presented to the graduate SE class at North Carolina State University, without quarrel or %critical comment, every year for the last decade.
+Yet when we look at the literature, the evidence for
+delayed issue effect is both very sparse and very old.
+
+We examined the literature on the delayed issue effect through a combination of snowball sampling~\cite{wohlin2014guidelines} and database search.
+We began with \cite{boehm2012}, the most recent publication by Barry Boehm (at the time of this paper's first draft) that discusses the delayed issue effect.
+Boehm cites the seminal \textit{Software Engineering Economics}~\cite{Boehm81}.
+%, which is bases the original delayed issue effect on four primary sources in the late 70s from IBM~\cite{Fagan76}, TRW~\cite{Boehm76}, GTE~\cite{Daly77}, and Bell Labs~\cite{Stephenson76}. 
+We searched Google Scholar for terms such as ``cost to fix'' and ``defect cost'' and ``software quality cost''.
+The majority of the search results discuss quality measurements or quality improvement, but relatively few discuss cost-to-fix as a function of when the defect was injected or found.
+Our literature search is by no means exhaustive, but our search results yielded many popular sources such as~\cite{pressman2005software, boehm01,glass02,endres03}\footnote{These sources have a combined citation count of over 14,500 on Google Scholar}, all of which cite \textit{Software Engineering Economics}~\cite{Boehm81} or its derivative works.
+
+\begin{figure}[!t]
+% \begin{center}\includegraphics[width=4in]{img/boehm_cost-to-fix.png}\end{center}
+\begin{center}\includegraphics[width=4.5in]{img/boehm81.png}\end{center}
+ \caption{Historical cost-to-fix curve. Adapted from~\cite{Boehm81}, p. 40.}\label{fig:cost-to-fix}
+ \end{figure}
+
+Figure~\ref{fig:cost-to-fix} shows the DIE as reported in \textit{Software Engineering Economics}~\cite{Boehm81} based on data from large systems in the late 70s from IBM~\cite{Fagan76}, TRW~\cite{Boehm76}, GTE~\cite{Daly77}, and Bell Labs~\cite{Stephenson76}.
+%These studies are most often cited by secondary sources on the delayed issue effect (e.g., \cite{boehm01, mcconnell01,glass02}. 
+We note that it is unclear from the text in~\cite{Daly77} and \cite{Boehm76} if cost is defined in terms of effort, or in actual cost (i.e., labor, materiel, travel, etc).
+The data points from these studies are not published for analysis.
+%These seminal studies are the basis for most DIE claims and that the cost to find and fix a defect monotonically increases with lifecycle phase.
+
+In the 40 years since these initial studies, few studies have explored the cost to resolve issues
+as a function of lifecycle phase. 
+Shull et al.~\cite{Shull02} conducted a literature survey and held a series of e-workshops with industry experts on fighting defects.
+Workshop participants from Toshiba and IBM reported cost-to-fix ratios between early lifecycle and post-delivery defects of 1:137 and 1:117 for large projects respectively~\cite{Shull02} -- but the raw data points were not provided and thus cannot be confirmed.
+A goal of agile methods
+is to reduce the difficulty associated with making changes later in the lifecycle~\cite{beck00}. Relatively little empirical data exists on this point.
+Elssamadisy and Schalliol~\cite{Elssamadisy02} offer an anecdotal report on the growing, high cost of rework in a 50 person, three-year, 500KLOC Extreme Programming project as the project grew in size and complexity-- but again we cannot access their
+exact figures. This was a common theme in the literature reviewed for this paper-- i.e.  that  it was no longer possible to access the data used to make prior conclusions.
+
+% We note that previous work focuses on cost-to-fix as a function of lifecycle phase irrespective of when the defect was injected, that is, previous work analyzes the cost to fix a defect found in test regardless of whether that defect was a requirements error or a coding mistake.
+To our knowledge, our research represents the first large-scale study of phase delay.
+
+% \begin{figure}[!t]
+% \begin{center}
+%  \includegraphics[width=4in]{img/clutterbuck.png} 
+%  \end{center}
+%  \caption{Cost of change from an agile case study. From~\cite{Clutterbuck09}.}\label{fig:clutterbuck}
+%  \end{figure}
+
+ %  All in all, the literature reviewed in this section
+% does not inspire confidence in a DIE effect due
+% to a lack of applicability and lack of replication.
+%  Adding to our doubts are
+
+% Clutterbuck et al.~\cite{Clutterbuck09} studied a 5-month effort by a small-to-medium enterprise team developing a 71KLOEC web interface to a database application to implement 18 change requests-- see Figure~\ref{fig:clutterbuck} (note that these were for new and changed user requirements, not defects).
+Clutterbuck et al.
+found the cost of change to be relatively flat until the later phases, with much of the effort spent in analysis of the change requests~\cite{Clutterbuck09}.
+Note that in this study, the effort increased by only 60\% (see the start and end of the curve in Figure~\ref{fig:clutterbuck}).
+
+Some studies that report less-than large
+ increase in the effort required to fix delayed issues.
+Royce~\cite{Royce98} studied  a million-line, safety-critical missile defense system.
+Design changes (including architecture changes) required approximately twice the effort of implementation and test changes, and the cost-to-fix in implementation and test phases increased slowly.
+Boehm~\cite{Boehm10} attributes this success to the  development process, which focused on removing architecture risk early in the development lifecycle.
+Boehm~\cite{Boehm80} provides data suggesting that the cost-to-fix curve for small projects, two student projects of 2000 deliverable source instructions, is flatter than for large projects (the dashed line of Figure~\ref{fig:cost-to-fix}).
+Data from NASA's Johnson Space Flight Center, reported by Shull~\cite{Shull02}, found that the cost to fix certain non-critical classes of defects was fairly constant across lifecycle phases (1.2 hours on average early in the project, versus 1.5 hours late in the project).
+Again, the data points from these studies are not available, and thus newer evidence both in favor of and contrary to the DIE cannot be evaluated.
+
+To gain a sense of how current the perception of the DIE is, 
+we conducted two surveys of software engineers. 
+The surveys collected data on software engineers' views of the DIE and other commonly held software engineering ``laws''.
 The surveys were conducted using Amazon's Mechanical Turk.
@@
 The PC member survey presented the RqtsErr law and an additional law on the DelayedIssueEffect: ``In general, the longer errors are in the system (requirements errors, design errors, coding errors, etc.), the more expensive they are to fix''.
-
-% Laws other than RqtsErr are not relevant to this paper, but we add them to as comparison points.
-
 The respondents answered two questions in response to each law:
 \bi
-\item
-\textbf{Agreement:} ``Based on your experience, do you agree that the statement above is correct?'' A Likert scale captured the agreement score from Strongly Disagree to Strongly Agree.
+\item \textbf{Agreement:} ``Based on your experience, do you agree that the statement above is correct?'' A Likert scale captured the agreement score from Strongly Disagree to Strongly Agree.
 A text box was provided to explain the answer.
-\newline
-\item
-\textbf{Applicability:} ``To the extent that you believe it, how widely do you think it applies among software development contexts?'' The possible answers were:
-\bi
-\item -1: I don't know
-\item 0:  this law does not apply at all
- \item 1: applies in a very narrow range of projects  
- \item 2: rarely applies
- \item 3: occasionally applies
-\item 4: very frequently applies
-\item 5: always applies
-\ei
+\item \textbf{Applicability:} ``To the extent that you believe it, how widely do you think it applies among software development contexts?'' The possible answers were: -1: I don't know, 0: this law does not apply at all, ..., 5: always applies.
 Respondents were required to explain the applicability score in a text box.
@@
 
-Summary statistics for the agreement and applicability scores for the RqtsErr and DelayedIssueEffect laws are presented in Figure~\ref{fig:survey_results}.
-Responses whose Applicability response was ''I don't know'' are omitted from analysis.
-Laws other than RqtsErr and DIE are not relevant to this paper, but are shown for comparison.
-
-\begin{figure}[!ht] 
+\begin{figure}[ht] 
 \scriptsize 
@@
 
+Summary statistics for the agreement and applicability scores for the RqtsErr and DelayedIssueEffect laws are presented in Figure~\ref{fig:survey_results}.
+Responses whose Applicability response was ''I don't know'' are omitted from analysis.
+Laws other than RqtsErr and DIE are not relevant to this paper, but are shown for comparison.
+
 Both  practitioners and researchers strongly believed in RqtsErr. In both sets of responses, RqtsErr received  scores higher than most
-other laws. Overall, the RqtsErr law was the most agreed upon and most applicable law of 11 surveyed amongst practitioners, and the second most agreed upon law amongst researchers. 
-This results strongly support that the notion that the  delayed issue effect   is a commonly held belief in software engineering.
+other laws.
+Overall, the RqtsErr law was the most agreed upon and most applicable law of 11 surveyed amongst practitioners, and the second most agreed upon law amongst researchers.
 From the free response texts, we note that the researchers who disagreed with RqtsErr generally asserted that requirements change can be expensive, but that the effect depends on the process used (e.g., agile vs.
@@
 
-\section{DIE is Poorly  Documented}\label{sect:doco}
-
-One reason that industrial practitioners and academics believe so strongly in the delayed issue effect is that it is often referenced
-in the SE literature. 
-%For example,
-%we know that \fig{b81} has been presented to the graduate SE class at North Carolina State University, without quarrel or %critical comment, every year for the last decade.
-Yet when we look at the literature, the evidence for
-delayed issue effect is both very sparse and very old.
-The first data on the difficulties of resolving delayed issues as a function of lifecycle phase date back to large systems in the late 70s from IBM~\cite{Fagan76}, TRW~\cite{Boehm76}, GTE~\cite{Daly77}, and Bell Labs~\cite{Stephenson76} (Figure~\ref{fig:cost-to-fix}).
-These studies are most often cited by secondary sources on the delayed issue effect (e.g., \cite{boehm01, mcconnell01,glass02}.
-We note that it is unclear from the text in ~\cite{Daly77} and \cite{Boehm76} if cost is defined in terms of effort, or in actual cost (i.e., labor, materiel, travel, etc).
-The data points from these studies are not published for analysis.
-These seminal studies are the basis for most DIE claims and that the cost to find and fix a defect monotonically increases with lifecycle phase.
-
-\begin{figure}[!b]
-\begin{center}\includegraphics[width=4in]{img/boehm_cost-to-fix.png}\end{center}
- \caption{Historical cost-to-fix curve. From~\cite{Boehm81}.}\label{fig:cost-to-fix}
- \end{figure}
-
-In the 40 years since these initial studies, few studies have explored the difficulty to resolve issues
-as a function of lifecycle phase. 
-Shull et al.~\cite{Shull02} conducted a literature survey and held a series of e-workshops with industry experts on fighting defects.
-Workshop participants from Toshiba and IBM reported cost-to-fix ratios between early lifecycle and post-delivery defects of 1:137 and 1:117 for large projects respectively~\cite{Shull02} -- but the raw data points were not provided and thus cannot be confirmed.
-
-\todo{REview the primary sources for this paper}
-This was a common theme in the literature reviewed for this paper-- i.e.  that  it was no longer possible to access
-the data used to make prior conclusions.
-As an example of this, \fig{steck} shows one 2004 survey that reports an 
-greatly increased delayed issue  effect for 
-requirements issues in eight case studies. Note that we cannot verify
-any of those results since the links in the references of
-that  survey are all broken (``page not found'' errors).
-
-A goal of agile methods 
-is to reduce the difficulty associated with making changes later in the lifecycle~\cite{beck00}. Relatively little empirical data exists on this point.
-Elssamadisy and Schalliol~\cite{Elssamadisy02} offer an anecdotal report on the growing, high cost of rework in a 50 person, three-year, 500KLOC Extreme Programming project as the project grew in size and complexity-- but again we cannot access their
-exact figures.
-
-% We note that previous work focuses on cost-to-fix as a function of lifecycle phase irrespective of when the defect was injected, that is, previous work analyzes the cost to fix a defect found in test regardless of whether that defect was a requirements error or a coding mistake.
-To our knowledge, our research represents the first large-scale study of phase delay.
-
- \begin{figure}
+The above arguments provide evidence to the claim that the DIE is both poorly documented yet (still) widely believed.
+The comments of Glass~\cite{glass02}, that the DIE is ``just common sense'', suggest that DIE may be the target of confirmation bias.
+An example of this is \fig{steck} from~\cite{steck04}, which purports to show nine references to ``studies [that] have been performed to determine the software error cost factors''.
+Only one of these sources, \textit{Software Engineering Economics}~\cite{Boehm81}, is based on real project data.
+Despite the 40-year old evidence, the perception of the DIE persists today among both the software engineers sampled in our survey and in popular literature.
+In the intervening years, many advances in software technology and processes have been made precisely to deal with risks such as the DIE.
+Thus, it is appropriate to ask the question, does the DIE still exist?
+
+ \begin{figure}[!ht] 
 {\small
 \begin{center}
-\begin{tabular}{r|rrrr}
- case& \multicolumn{4}{c}{Phase Requirements Issue Found }  \\
- study               &Requirements & Design & Code&  Test\\\hline
-1& 1 &3& 5& 37\\  
-2& 1  &    & 10 & 40 \\ 
-3& 1   &    & 10 &  40 \\
-4&  1  &   5 &      & 50  \\
-5&  1 &3& 7& 51 \\
-6& 1& 5 &33 &75  \\
-7&  1 & 20 & 45 & 250 \\  
+\begin{tabular}{r|rrrr|l}
+                        & \multicolumn{4}{c}{Phase Requirements Issue Found }   &     \\
+ Cited source           & Requirements  &   Design  &   Code    &  Test         & Data sources used to determine DIE          \\\hline
+\cite{Boehm81}          &   1           &   5       &   10      &   50          & Multiple projects     \\  
+Hoffman, 2001           &   1           &   3       &   5       &   37          & Unknown - no bibliography entry \\ 
+\cite{Cigital2003}      &   1           &   3       &   7       &   51          & Extrapolated from defect counts for a one project \\
+\cite{Rothman2000}      &               &   5       &   33      &   75          & Fictitious example \\
+\cite{Rothman2000} Case B   &               &           &   10      &   40      & Fictitious example \\
+\cite{Rothman2000} Case C   &               &           &   10      &   40      & Fictitious example \\
+\cite{Rothman2002}      &   1           &   20      &   45      &   250         & Fictitious example \\
+\cite{Pavlina2000}      &   1           &   10      &   100     &   1000        & None provided \\
+\cite{McGibbon2007}     &               &   5       &           &   50          & Pen \& paper exercise - no real data \\
 \end{tabular}
 \end{center}}
-\caption{Cost to resolve requirements issues, relative to resolving  during requirements. From ~\cite{steck04}.}\label{fig:steck}
+\caption{Confirmation bias -- sources for DIE cited in Table 1 of~\cite{steck04}.
+Note that \textit{all} of these are cited as ``studies [that] have been performed to determine the software error cost factors'', but only one, \cite{Boehm81}, is backed by actual data.}\label{fig:steck}
 \end{figure}
 
-% \begin{figure}[!t]
-% \begin{center}
-%  \includegraphics[width=4in]{img/clutterbuck.png} 
-%  \end{center}
-%  \caption{Cost of change from an agile case study. From~\cite{Clutterbuck09}.}\label{fig:clutterbuck}
+% \begin{figure}
+% \begin{center}\includegraphics[width=4in]{img/Royce98.png}\end{center}
+%  \caption{Exception to the rule - The Royce study: cost-to-fix curve. From~\cite{Royce98}.}\label{fig:royce}
 %  \end{figure}
 
- %  All in all, the literature reviewed in this section
-% does not inspire confidence in a DIE effect due
-% to a lack of applicability and lack of replication.
-%  Adding to our doubts are
-
-% Clutterbuck et al.~\cite{Clutterbuck09} studied a 5-month effort by a small-to-medium enterprise team developing a 71KLOEC web interface to a database application to implement 18 change requests-- see Figure~\ref{fig:clutterbuck} (note that these were for new and changed user requirements, not defects).
-Clutterbuck et al.
-found the cost of change to be relatively flat until the later phases, with much of the effort spent in analysis of the change requests~\cite{Clutterbuck09}.
-Note that in this study, the effort increased by only 60\% (see the start and end of the curve in Figure~\ref{fig:clutterbuck}).
-
-Some studies that report less-than large
- increase in the effort required to fix delayed issues. Royce~\cite{Royce98} studied  a million-line, safety-critical missile defense system
-(see  Figure~\ref{fig:royce}).
-Design changes (including architecture changes) required approximately twice the effort of implementation and test changes, and the cost-to-fix in implementation and test phases increased slowly.
-Boehm~\cite{Boehm10} attributes this success to the  development process, which focused on removing architecture risk early in the development lifecycle.
-Boehm~\cite{Boehm80} provides data suggesting that the cost-to-fix curve for small projects ,from two student projects of 2000 deliverable source instructions, is flatter than for large projects (the dashed line of Figure~\ref{fig:cost-to-fix}).
-Data from NASA's Johnson Space Flight Center, reported by Shull~\cite{Shull02}, found that the cost to fix certain non-critical classes of defects was fairly constant across lifecycle phases (1.2 hours on average early in the project, versus 1.5 hours late in the project).
-Again, the data points from these studies are not available, and thus the evidence both in favor of and contrary to the DIE is either quite old and/or is based on data that cannot be confirmed.
-
-\begin{figure}
-\begin{center}\includegraphics[width=4in]{img/Royce98.png}\end{center}
- \caption{Exception to the rule - The Royce study: cost-to-fix curve. From~\cite{Royce98}.}\label{fig:royce}
- \end{figure}
-
 \subsection{Early Onset of the DIE Effect}\label{sect:earlyonset}
 
-One feature of the results in the previous section is important to our subsequent discussion.  
-The DIE literature reports the onset of  DIE  {\em prior} to delivery.
+One feature of the the DIE literature is important to our subsequent discussion:   
+the onset of  DIE  {\em prior} to delivery.
 \bi
@@
 (for the larger projects) before the code is delivered 
-\item
-The projects surveyed in \fig{steck} report changes of a  
-\{37,40,40,50,51,70,75,250\}-fold increase in pre-deployment period when the software was being developed.
 \ei
@@
 \section{Delayed Issues are not  Harder  to Resolve}
-
+\label{sect:analysis}
 The above analysis motivates a more detailed look at the delayed issued effect.  
@@
 There were  some  medical device projects in  the US, France, Japan, and Germany as well  as a commercial computer-aided design systems, and embedded systems.
-A more thorough characterization of the projects providing data is provided in \S\ref{sec:data_character}.
+A more thorough characterization of the projects providing data is provided in \S\tion{data_character}.
 
@@
 \subsection{Data Collection and Definitions}
-\label{sec:data-collection}
+\label{sect:data-collection}
 
@@
 and 26,534 size logs. 
-In this paper, we exclude 41 of the 212 that had too few defects \todo{Define ``too few''}, leaving 171 projects included in the analysis.
+In this paper, we exclude 41 of the 212 that had too few defects (less than 30), leaving 171 projects included in the analysis.
 
@@
 
-\subsubsection{Definition: defects and time-to-fix}\label{tion:defin}
+\subsubsection{Definition: defects and time-to-fix}\label{sect:defin}
 In the TSP, a  \emph{defect} is any change to a product, after its construction, that is necessary to make the product correct.
@@
 \subsection{Data Integrity}
-\label{sec:data_integrity}
+\label{sect:data_integrity}
 
@@
 \subsection{Project Descriptive Characteristics}
-\label{sec:data_character}
+\label{sect:data_character}
 
@@
 The project contexts describe the conditions under which these measures were obtained, help determine relevance of the results, and may guide future data analysis with segmentation.
-
-While the processes and workflows among these projects can vary, the logical order is described section \ref{development_phase} is followed.
-Development tasks such as requirements development, design, or code, are typically followed by an appraisal phase such as personal review or inspection.
-Effort and effectiveness of these activities vary among projects and developers.
+Key attributes of the context include the business and application domains, product size, project duration,  work flows, team size, team management,  development and integration approaches, organization size, location or distribution, certifications, developer experience, programming languages and tools used.
 
@@
 Interest in the data sets by the community may encourage our sponsor to fund additional data mining.
+Nonetheless, much context is available from the project data  and we provide some additional context not included within the fact sheets.
 
@@
 The most common project application domains are banking, consumer applications, engineering design tools, and medical devices.   
+The data for programming languages is incomplete, with most projects using more than one language, but few reporting  programming language by component or size.
+The list of languages includes ABAP, ADA, Alpha, C, C++, C\#, ASP.net, Delphi, Gauss, Genexus, Hotware, HTML, Java, JavaScript, PHP, PLSQL, Ruby, SQL, and Visual Basic.
+
+ The specific process work flows and practices are developed by the development team personnel who have received specific training on defining work processes as part of their Personal Software Process training.
+The process data was collected by the team members to self-manage their personal and team work.
+Them members also exhibited self-management behavior by estimating planning and scheduling the work tasks.
+
+While the processes and work flows among these projects can vary, the logical order is described section \ref{development_phase} is followed.
+Development tasks such as requirements development, design, or code, are typically followed by an appraisal phase such as personal review or inspection.
+Effort and effectiveness of these activities vary among projects and developers.
 
@@
 
-The data for programming languages is incomplete, with most projects using more than one language, but few reporting  programming language by component or size.
-The list of languages includes ABAP, ADA, Alpha, C, C++, C\#, ASP.net, Delphi, Gauss, Genexus, Hotware, HTML, Java, JavaScript, PHP, PLSQL, Ruby, SQL, and Visual Basic.
-
  \begin{figure}[ht]
@@
 
- The teams were mostly of small size with a median size of 6 and a mean of 7.8. only 25\% of the teams were larger than 10 with a long tail on the distribution.
+How many of these teams could be classified as "agile" is not clear because actual practices in the agile world can vary.
+We did not ask teams to self-identify, however we offer the following observations regarding characteristics commonly associated with agile behavior.
+\begin{itemize}
+    \item all teams were self managed, defining work flows, practices, and schedules
+    \item teams meet at least weekly to evaluate progress and replan
+    \item most teams were small with a median size of 6 and a mean of 7.8. only 25\% of the teams were larger than 10 with a long tail on the distribution
+    \item the median project lasted only 60 days, suggesting limited scope for each integration
+\end{itemize}
 
@@
 \end{figure}
+
+\subsection{Statistical Analysis}\label{sect:stats}
+
+    In the following presentation of our results, three statistical methods were used to test for the delayed
+    issue effect: the Scott-Knott ranker;   bootstrap sampling (to test for statistical significantly
+    different results); and an effect size test (to reject any significant differences that are trivially small).
+
+    The Scott-Knott ranker 
+    recommended was   recommended  by  Mittas and Angelis in a
+    recent TSE'13 article~\cite{mittas13} and by Ghotra et al. in a recent
+    ICSE'15 article~\cite{ghotra2015icse}
+    Scott-Knott is a 
+    top-down clustering approach used to rank different treatments. If that 
+    clustering finds an ``interesting division'' of the data, then some 
+    statistical test is applied to the two divisions to check if they are 
+    statistically significant different. If so, Scott-Knott considers recurses into both 
+    halves.
+    Before Scott-Knot recurses,  however,  it apples some  statistical hypothesis test $H$ to check
+    if $m,n$ are significantly different.
+
+    To operationalize ``interesting'', Scott-Knott seeks the division of 
+    of $l$ treatments into subsets of size $m,n$ of sizes $ls,ms,ns$ and median values
+    $l.\mu, m.\mu, n.\mu$ (respectively) in order to maximize
+ $\frac{ms}{ls}abs(m.\mu - l.\mu)^2 + \frac{ns}{ls}abs(n.\mu - l.\mu)^2$.
+
+   To operationalize $H$, we use both  bootstrap sampling  and the A12 effect size test.
+   In other 
+    words, we divide the data if \textit{both} bootstrap sampling and effect 
+    size test agree that a division is statistically significant (with a 
+    confidence of 99\%) and not a small effect ($A12 \ge 0.6$).
+    For a justification of the use of non-parametric bootstrapping, see Efron 
+    \& Tibshirani~\cite[p220-223]{efron93}. For a justification of the use of 
+    effect size tests see Shepperd and MacDonell~\cite{shepperd12a}; 
+    Kampenes~\cite{kampenes07}; and Kocaguenli et 
+    al.~\cite{Kocaguneli2013:ep}. These researchers warn that even if a 
+    hypothesis test declares two populations to be ``significantly'' 
+    different, then that result is misleading if the ``effect size'' is very 
+    small. Hence, to assess the performance differences we first must rule out 
+    small effects using A12, a test   recently endorsed by Arcuri and 
+    Briand at ICSE'11~\cite{arcuri11}.
+
+    To  apply Scott-Knott, we divided data into the phases $P_0$ where issues are introduced.
+Next, for each division, we separated all the issues that were removed at different subsequent issues $P_r \in \{P_1,P_2,..\}$.
+    For each pair $P_0,P_r$, we build one treatment containing the  issue resolution times for     issues raises in $P_0$ and resolved in $P_r$. These treatments
+    were then ranked by Scott-Knott.
 
@@
 \end{center}
-\caption{Time to resolve issues seen in the SEI TSP data.
+\caption{Median times to resolve issues seen in the SEI TSP data.
 The main effect here is that there is no clear exponential growth in the time to resolve issues as issues are injected in some phase $P_0$ then resolved in some subsequent issue $P_r$.
-Shown where are the 50th and 90th percentiles of issue resolution times for issues injected in phase $P_O$ and resolved in phase $P_r$ (these values are calculated
-by sorting all resolution time, then reporting the values at 0.5 and 0.9).
+Shown where are the 50th percentiles of issue resolution times for issues injected in phase $P_O$ and resolved in phase $P_r$ (these values are calculated
+by sorting all resolution time, then reporting the middle values of that sort)
 The ``IQR'' column shows the ``inter-quartile range''; i.e. the range of values representing the 75th - 25th percentile range
-The left hand column shows the result of the Scott-Knott ranking procedure described in \tion{stats}. Note that
-most treatments achieved the same ranks i.e. they were found to be not statistically significantly different  (the one exception is the Planning:UnitTest results were UnitTests were ranked 2).
-The right hand side bars
-illustrate how these resolutions increases.
-The {\bf BLACK} and \textcolor{Gray}{{\bf GRAY}} bars show the increases for the  50th (median) and 90th percentile values, respectively.
+The left-hand-side ``rank'' column shows the result of the Scott-Knott ranking procedure described in \tion{stats}. Note that
+most treatments achieved the same ranks i.e. they were found to be insignificantly different  (the one exception is the Planning:UnitTest results 
+where UnitTests were ranked 2).
+The right-hand-side bars
+illustrate shows the releative sizes of the increases for the  50th (median)   percentile values.
 These increases are calculated with respect to the first value in each section ``Before, Planning, Reqts.
@@
 These right-hand-side bars are unitless since they are ratios. For example,
-on the last line, issues injected during coding and fixed in SysTest take 47 minutes (median) to resolve. This is 130\% more than the 35 minites (median) required
-to resolve coding issues during CodeReview.}
+on the last line, issues injected during coding and fixed in SysTest take 13 minutes (median) to resolve. This is 130\% more than the 10 minutes (median) required
+to resolve coding issues during CodeReview. The right-hand-side bar visually represents that 130\%.}
 \label{fig:raw}
@@
 
-\subsection{Statistical Analysis}\label{sect:stats}
-
-    In the following presentation of our results, three statistical methods were used to test for the delayed
-    issue effect: the Scott-Knott ranker;   bootstrap sampling (to test for statistical significantly
-    different results); and an effect size test (to reject any significant differences that are trivially small).
-
-    The Scott-Knott ranker 
-    recommended was   recommended  by  Mittas and Angelis in a
-    recent TSE'13 article~\cite{mittas13} and by Ghotra et al. in a recent
-    ICSE'15 article~\cite{ghotra2015icse}
-    Scott-Knott is a 
-    top-down clustering approach used to rank different treatments. If that 
-    clustering finds an ``interesting division'' of the data, then some 
-    statistical test is applied to the two divisions to check if they are 
-    statistically significant different. If so, Scott-Knott considers recurses into both 
-    halves.
-    Before Scott-Knot recurses,  however,  it apples some  statistical hypothesis test $H$ to check
-    if $m,n$ are significantly different.
-
-    To operationalize ``interesting'', Scott-Knott seeks the division of 
-    of $l$ treatments into subsets of size $m,n$ of sizes $ls,ms,ns$ and median values
-    $l.\mu, m.\mu, n.\mu$ (respectively) in order to maximize
- $\frac{ms}{ls}abs(m.\mu - l.\mu)^2 + \frac{ns}{ls}abs(n.\mu - l.\mu)^2$.
-
-   To operationalize $H$, we use both  bootstrap sampling  and the A12 effect size test.
-   In other 
-    words, we divide the data if \textit{both} bootstrap sampling and effect 
-    size test agree that a division is statistically significant (with a 
-    confidence of 99\%) and not a small effect ($A12 \ge 0.6$).
-    For a justification of the use of non-parametric bootstrapping, see Efron 
-    \& Tibshirani~\cite[p220-223]{efron93}. For a justification of the use of 
-    effect size tests see Shepperd and MacDonell~\cite{shepperd12a}; 
-    Kampenes~\cite{kampenes07}; and Kocaguenli et 
-    al.~\cite{Kocaguneli2013:ep}. These researchers warn that even if a 
-    hypothesis test declares two populations to be ``significantly'' 
-    different, then that result is misleading if the ``effect size'' is very 
-    small. Hence, to assess the performance differences we first must rule out 
-    small effects using A12, a test   recently endorsed by Arcuri and 
-    Briand at ICSE'11~\cite{arcuri11}.
-
-    To  apply Scott-Knott, we divided data into the phases $P_0$ where issues are introduced.
-Next, for each division, we separated all the issues that were removed at different subsequent issues $P_r \in \{P_1,P_2,..\}$.
-    For each pair $P_0,P_r$, we build one treatment containing the  issue resolution times for     issues raises in $P_0$ and resolved in $P_r$. These treatments
-    were then ranked by Scott-Knott.
-
-\subsection{Observations from 171 Projects}\label{tion:171projects}
-
-The distribution of defects found and fixed per phase in our data is shown in Figure~\ref{fig:fix-phase-dist}. 
-A high percentage of defects (44\%) were found and fixed in the early phases, i.e., requirements, high level design, and design reviews and inspections.
-\todo{Make sure that defect injection phases in the figure match those of \fig{waterfall}} This distribution is similar to that observed for other projects that emphasized investment in software engineering quality assurance practices.
+\subsection{Observations from 171 Projects}\label{sect:171projects}
+
+The distribution by phase in which defects were discovered (that is both found and fixed) in our data is shown in Figure~\ref{fig:fix-phase-dist}.
+Defects are counted only if they they escape the introduction phase unless a bad fix introduces a new defect.
+These secondary defects occur almost exclusively in test and very rarely in an inspection.
+A high percentage of defects (44\%) were found and fixed in the early phases, i.e., prior to coding.
+This distribution is similar to that observed for other projects that emphasized investment in software engineering quality assurance practices.
 For example, Jones and Bonsignour report 52\% of pretest defects removed before entering implementation, for large projects that focus on upfront defect removal techniques \cite{jones12}.
@@
 
-\fig{raw} shows the  50th and 90th percentile
+\fig{raw} shows the  50th  percentile
 of the time spent resolving issues
@@
 where there exists at least $N\ge 30$ examples\footnote{We selected 30
-for this threshold via the central limit theorem~\cite{maxwelldata}.} of issues {\em injected} in phase $X$ then
+for this threshold via the central limit theorem~\cite{maxwelldata}.} of issues {\em injected} in phase $P_0$ then
 removed in phase $P_r$.
-
-Nowhere in 
+The right hand side bars of that figure
+illustrate how these resolutions increase (expressed as ratios of resolution time in the phase where they were first injected).
+
+The twp key features of \fig{raw} are:
+\be
+\item Nowhere in
 these results do we see the kind of very large increases reported in the papers
-documenting DIE. Returning to
-the {\em before/integration test} result: 1.50 and 1.97 is far smaller
-than the scale ups seen in Figure~1.
-
-The right hand side bars of that figure
-illustrate how these resolutions increases (expressed as ratios of resolution time in the phase where they were first injected).
-The {\bf BLACK} and \textcolor{Gray}{{\bf Gray}} bars show the increases for the  50th (median) and 90th percentile values, respectively.
-Those bars offer little support for the 
-claim that early lifecycle errors are most expensive to fix.
-In all our results, the biggest scale ups are found with resolving
-issues injected in the {\em planning} phase. For example,
-resolving {\em planning} issues during unit testing was 3.00 and 1.51times (at the 50th and 90th percentile, respectively)
-slower than resolving issues earlier in the lifecycle.
-
-The left hand column of \fig{raw} shows the results of the Scott-Knott statistical tests. Note that nearly all
+documenting DIE. For example, consider the ratio of the issue resolution time
+between {\em Before/DesignInspect} and {\em Before/SysTest} result. That ratio is 1.11 which is far smaller
+than the scale ups seen in \fig{b81}.  
+\item
+Nearly all the supposed increases seen in \fig{raw} are insignificantly different
+to the other treatments.  The left hand column of \fig{raw} shows the results of the Scott-Knott statistical tests. Note that nearly all
 the treatments have the same rank (``1''); i.e. usually there is no statistically significant difference in the 
-time to resolve issues. The only exception here is Planning:UnitTest where is ranked ``2'' but ever here, the scale up is merely
+time to resolve issues. The only exception here is Planning:UnitTest which is ranked ``2'' but even here, the scale up is merely
 a factor of 3, and not the exponential increase promised by classic reports of the delayed issue effect.
-
-Before moving on, we commend on one quirk of \fig{raw} which is the counter-intuiative
-nature of some of the results. Consider, for example, the ``Reqts'' results were
+\ee
+One possible explanation for the lack of a DIE effect is that we are looking
+broadly at the entire data set but not at specific stratifications. To address that concern,
+we spent some time reproducing \fig{raw} for various subsets of our data.
+That proved to be an unfruitful-- no stratification was found that 
+contained an exponential expansion in the time to fix issues. The reason for this was  the small size
+of those stratifications  exacerbated the large IQR's seen in this data\footnote{Recall that in a sorted list of numbers,
+the inter-quartile range, or IQR, is difference between the 
+  75th to 25th percentile value.}.
+Our 171 projects   stratifies into subsets of varying sizes. Sorting those sizes, we get a list
+comprising of 17,12 then numerous much smaller stratifications.  Reasoning over such small samples
+is problematic in the general case and, in the case of our data, it is ever more problematic due to
+the large IQRs of the data
+(to see these large IQRs,  please consider the 50th and IQR columns of Figure~\ref{fig:raw} where  most of our IQRs were larger than tan the 50th percentile; i.e. software
+data exhibits large variances-- where are exacerbated by the smaller samples seen in the stratifications).
+Our conclusion from exploring the stratification's is, given the currently available data, we cannot check
+for a DIE effect in subsets of this data.
+
+Before moving on, we comment on some of the counter-intuitive
+results in \fig{raw}. Consider, for example, the ``Reqts'' results were
 the time required to fix issues actually tends to {\em decrease} the longer they are left in the system.
 In terms of explaining this result, the key thing is the left-hand-side statistically ranking: all these treatments were found to be statistically indistinguishable.
-In such a set of treatment, the observed
+In such a set of treatments, the observed
 difference may {\em not} be a causal effect; rather, it may just
@@
 \subsection{Conclusion Validity}
+\label{sect:conclusion_validity}
 Threats to conclusion validity are ``issues that affect the ability to draw the correct conclusion about relations between the treatment and the outcome''~\cite{wohlin2012}.
@@
 \textit{Low statistical power}: Our data set is comprised of 47,376 defect logs.
-Figure~\ref{tab:phaseSamples} shows the number of samples in each Injection-Removal phase pair from Figure~\ref{fig:fix-phase-dist}.
-\todo{more once Lucas' email is answered.}
-
-\textit{Reliability of measures}: The base measures in this study described in \S\ref{sec:data-collection} are \textit{defects} recorded in TSP defect logs and \textit{time} reported in time tracking logs.
+Our primary analysis in \fig{raw} is based on injection-removal phase pairs whose with sample size $>30$.
+The justification for the statistical techniques used in this paper are provided in \tion{stats}.
+
+\textit{Reliability of measures}: The base measures in this study described in \tion{data-collection} are \textit{defects} recorded in TSP defect logs and \textit{time} reported in time tracking logs.
 The primary threats to the reliability of these measures are: that the definition of a defect varies between projects and that time is not reported accurately or consistently.
-The reliability of the time reporting is discussed in \S\ref{sec:data_integrity}.
+The reliability of the time reporting is discussed in \tion{data_integrity}.
 Time is reported on a level of minutes.
@@
 A related issue is that we assume developers correctly report in which phases a defect was injected and corrected.
-One points of variation is the measurement framework that identifies process phases and joins the effort to a size measurement framework.
+One point of variation is the measurement framework that identifies process phases and joins the effort to a size measurement framework.
 Individual projects may choose to implement a different framework, for example adding phases for specific types of development (for example, adding static analysis or special testing or a non-standard size unit).
@@
 The guidance for logging time and defects is precisely defined.
-All tasks should be logged as the work is performed with a stopwatch.
+All tasks should be logged as the work is performed with a stopwatch like tool.
 All defects that escape a phase must be logged.
@@
 
-\begin{figure}[ht]
-%\renewcommand{\baselinestretch}{0.7}
-\scriptsize
-\centering
-\begin{tabular}{rrrrrr}
-  \hline
-                & BeforeDevelopment & Planning      & Reqts          & Design           & Code \\ 
-  \hline
-ReqtsReview     &   3               &  \textbf{80}  & \textbf{2221}  &   0              &   0 \\ 
-  ReqtsInspect  &   1               &   5           & \textbf{2439}  &   0              &   0 \\ 
-  DesignReview  &   6               &   7           &  \textbf{29}   & \textbf{3617}    &   0 \\ 
-  DesignInspect &  \textbf{24}      &  \textbf{12}  &  \textbf{58}   & \textbf{5243}    &   0 \\ 
-  CodeReview    &  \textbf{75}      &   2           &   8            & \textbf{250}     & \textbf{5916} \\ 
-  CodeInspect   & \textbf{161}      &  12           &  \textbf{11}   & \textbf{357}     & \textbf{6045} \\ 
-  UnitTest      & \textbf{240}      &   \textbf{9}  &  \textbf{27}   & \textbf{1789}    & \textbf{2082} \\ 
-  IntTest       &  \textbf{72}      &   0           &  \textbf{15}   &  \textbf{77}     & \textbf{841} \\ 
-  SysTest       &  \textbf{50}      &   3           &  \textbf{12}   & \textbf{116}     & \textbf{486} \\ 
-  AcceptTest    &   5               &   0           &  \textbf{12}   &  \textbf{13}     & \textbf{120} \\ 
-   \hline
-\end{tabular}
-\caption{phase delay samples. Boldface indicated that a cell is included in Fig. 8.
-\todo{XXX Lucas. Check figure reference}.} 
-\label{tab:phaseSamples}
-\end{figure}
+% \begin{figure}[ht]
+% %\renewcommand{\baselinestretch}{0.7}
+% \scriptsize
+% \centering
+% \begin{tabular}{rrrrrr}
+%   \hline
+%                 & BeforeDevelopment & Planning      & Reqts          & Design           & Code \\ 
+%   \hline
+% ReqtsReview     &   3               &  \textbf{80}  & \textbf{2221}  &   0              &   0 \\ 
+%   ReqtsInspect  &   1               &   5           & \textbf{2439}  &   0              &   0 \\ 
+%   DesignReview  &   6               &   7           &  \textbf{29}   & \textbf{3617}    &   0 \\ 
+%   DesignInspect &  \textbf{24}      &  \textbf{12}  &  \textbf{58}   & \textbf{5243}    &   0 \\ 
+%   CodeReview    &  \textbf{75}      &   2           &   8            & \textbf{250}     & \textbf{5916} \\ 
+%   CodeInspect   & \textbf{161}      &  12           &  \textbf{11}   & \textbf{357}     & \textbf{6045} \\ 
+%   UnitTest      & \textbf{240}      &   \textbf{9}  &  \textbf{27}   & \textbf{1789}    & \textbf{2082} \\ 
+%   IntTest       &  \textbf{72}      &   0           &  \textbf{15}   &  \textbf{77}     & \textbf{841} \\ 
+%   SysTest       &  \textbf{50}      &   3           &  \textbf{12}   & \textbf{116}     & \textbf{486} \\ 
+%   AcceptTest    &   5               &   0           &  \textbf{12}   &  \textbf{13}     & \textbf{120} \\ 
+%   \hline
+% \end{tabular}
+% \caption{phase delay samples. Boldface indicated that a cell is included in Fig. 8.
+% \label{tab:phaseSamples}
+% \end{figure}
 
@@
 \textit{Instrumentation}: The forms by which the TSP defect and time data are collected have been studied and matured over 20 years.
-Conversely, we do not find much documented evidence on how time and defects are reported for the original DIE papers (see the \S\ref{sect:doco}).
+Conversely, we do not find much documented evidence on how time and defects are reported for the original DIE papers (see \tion{belief}).
 Thus, we cannot be assured that reporting and data capture were not a significant influence on the delayed issue effect in the original papers.
 
-\textit{Interactions with selection}: As described in \S\ref{sec:data-collection}, all TSP teams are required to contribute time and defect data to the SEI, and thus there should be no selection bias in this sample compared to the overall population of TSP projects.
+\textit{Interactions with selection}: As described in \tion{data-collection}, all TSP teams are required to contribute time and defect data to the SEI, and thus there should be no selection bias in this sample compared to the overall population of TSP projects.
 However, there is likely selection bias in the teams that elect to use TSP compared to the entire population of software development teams.
@@
 
-\subsection{Construct validity} \iton{sect:construct}
+\subsection{Construct validity} \label{sect:construct}
 Construct validity concerns ``generalizing the result of the experiment to the concept or theory behind the experiment''~\cite{wohlin2012}.
@@
 
-\textit{Inadequate preoperational explication of constructs}: As described in \S\ref{sec:data-collection}, the measures of defect, time, and cost in the original DIE papers are not clearly defined.\footnote{In retrospect, empirical software engineering studies at that time were extremely rare, and guidance for reporting empirical case studies and experiments have improved substantially.
-One of the seminal books on quasi-experimention and reporting of validity concerns, Cook and Campbell~\cite{cook1979}, had not been published when most of the DIE papers were written.} Note that in Figure~\ref{fig:cost-to-fix}, the units of ``cost-to-fix'' are not expressed -- in the source references, cost appears as calendar time, effort, and price.
-In the TSP, a defect is defined as ``any change to a product, after its construction, that is necessaryto make the product correc'' and time to correct a defect includes ``the time to analyze, repair, and validate a defect fix''.
+\textit{Inadequate pre-operational explication of constructs}: As described in \tion{data-collection}, the measures of defect, time, and cost in the original DIE papers are not clearly defined.\footnote{In retrospect, empirical software engineering studies at that time were extremely rare, and guidance for reporting empirical case studies and experiments have improved substantially.
+One of the seminal books on quasi-experimentation and reporting of validity concerns, Cook and Campbell~\cite{cook1979}, had not been published when most of the DIE papers were written.} Note that in Figure~\ref{fig:cost-to-fix}, the units of ``cost-to-fix'' are not expressed -- in the source references, cost appears as calendar time, effort, and price.
+In the TSP, a defect is defined as ``any change to a product, after its construction, that is necessary to make the product correc'' and time to correct a defect includes ``the time to analyze, repair, and validate a defect fix''.
 Our analysis of DIE focuses on time as a measure of time-as-effort (persons * time).
@@
 scale-up seen in the {\em before} to {\em integration testing} 
-which is far smaller than the 37 to 250-fold increases reported in \fig{b81}, \fig{cost-to-fix},
-and \fig{steck}).
+which is far smaller than the 37 to 250-fold increases reported in \fig{b81} and \fig{cost-to-fix}).
 
@@
 High-severity defects are of particular concern in software projects, and even if the number of high-severity defects is low their cost to fix may be extremely large.
-If such outliers were common in our data, they would appear in the upper percentiles of results.
-As shown in the \textcolor{gray}{{\bf GRAY}} bars of \fig{raw}, no such effect appears in the 90th percentiles of \fig{raw}.
-That is, while such ``high risk'' issues may certainly exist, they were not a major effect in our data.
+Note that if such outliers were common in our data, they would appear in the upper percentiles of results.
+\fig{last} shows that no such effect occurs in the 90th percentile of our time to resolve
+issues.    That is, while such ``high risk'' issues may certainly exist, they were not a major effect in our data.
 
@@
 
+\begin{figure}[!t]
+% \renewcommand{\baselinestretch}{0.8}
+% \small
+\begin{center}
+\input{tex/figlast}
+\end{center}
+\caption{90th percentile tiles to resolve issues in the SEI TSP data. Same format as \fig{raw} (but here we look at 90th percentile
+outliers while   \fig{raw} explored the central tendencies of the data). The conclusion from this figure is the same
+as \fig{raw}: even when considering the 90th percentile outliers, there is no consistent patterns of 
+exponential growthe in the effort to fix issues delayed later in the lifecycle.}\label{fig:last}
+\end{figure}
+
 \subsection{External Validity}
+\label{sect:external_validity}
 External validity concerns the generalizability of findings~\cite{wohlin2012} beyond the context of the study.
@@
  \section{Discussion}
+ \label{sect:discussion}
 % Unexpected results such as this one require significant caution.
@@
 Edwards Deming~\cite{deming1986out}.
+Harter et al.
+find a statistical correlation between fewer high severity defects and rigourous process discipline in large or complex systems~\cite{harter2012does}.
 Technical advancements in processing power, storage, networking, and parallelism have combined with a deeper scientific understanding of software construction to enable a whole host of software assurance technologies, from early-phase requirements modeling to automated release testing.
@@
  \section{Conclusion}
+ \label{sect:conclusion}
 
@@
 Based on a  survey  of both researchers and practitioners, we  found that
-a specific form  of this effect (requirements errors are hardest to fix) is  commonly believed
-held in the community.
+a specific form  of this effect (requirements errors are hardest to fix) is  commonly believed in the community.
 
@@
 
-We do not claim that this effect {\em never} holds in software projects; just that it cannot be assumed to {\em always} hold. Our explanation of the observed lack-of-effect is five-fold:
+We do not claim that this effect {\em never} holds in software projects; just that it cannot be assumed to {\em always} hold.
+Our explanation of the observed lack-of-effect is five-fold.
+Each of the following explanations is essentially a hypothesis which should be tested against empirical data before we can effectively propose a new theory of the delayed issue effect.
 \be
-\item The effect might be an historical relic. Evidence:
+\item The effect might be an historical relic, which does not always hold on contemporary projects. Evidence:
 the effect was first described in the era of punch card computing and non-interactive environments.
-\item The effect might have been intermittent (rather than some fundamental law of software). Evidence: we can  found nearly
+\item The effect might be intermittent (rather than some fundamental law of software). Evidence: we can  found nearly
 as many papers reporting the effect~\cite{Boehm76,Boehm81,steck04,Fagan76,Stephenson76} as otherwise~\cite{Royce98,Boehm80,Shull02}.
@@
 \newpage
-\pagenumbering{arabic}
+\pagenumbering{roman}
  \setcounter{page}{1}

### 2016-06-17 tex/responses.tex
Updates from ShareLaTeX

@@
 those comments we have extensively revised and extended this draft.
-We trust you will find this draft gratefully improved}
+Specifically, the old version was 20 pages and this version is 27 pages (and the new
+material all comes from the clarification and correction points offered by the reviewers).
+We hope that  you will find this draft greatly improved.}
 \end{quote}
@@
 
-\begin{quote}{\em This new draft  now maps our work into the SE literature on theory building.}\end{quote}
+\begin{quote}{\em This new draft now maps our work into the SE literature on theory building. This new material appears in the introduction as well as being
+peppered throughout the document (in numerous places).}\end{quote}
 
@@
 \begin{quote}{\em
-This new draft  now maps not contains extensive notes on the origin and characteristics of our data.}\end{quote}
+This new draft contains much more detail on the origin and characteristics of our data (see \tion{data-collection} and \tion{data_character}).}\end{quote}
 
@@
 
-\begin{quote}{\em All our results are now augmented with statistical significance tests as well as effect size tests.
-Those tests greatly strength the overall message of this paper-- that there is no significant
+\begin{quote}{\em All our results are now augmented with statistical significance tests as well as effect size tests (see \tion{stats} and \tion{171projects}).
+Those tests strengthen the overall message of this paper-- that there is no significant
 phase delay effect.}\end{quote}
@@
 
-\begin{quote}{\em We have carefully read those comments  nd have made numerous changes
+\begin{quote}{\em We have carefully read those comments and have made numerous changes
 as directed by reviewer3.}\end{quote}
@@
 
-\begin{quote}{\em  We have added an example of the DIE in the Introduction, and more precisely defined how we measure this effect Section 6.\footnote{Lucas- where exactly?} }\end{quote}
+\begin{quote}{\em  The introduction now includes an example of DIE  (see Fig 1).
+Also there are now more precise  definitions on how we measure this effect in \tion{data-collection}}\end{quote}
 
  The authors do not relate their reassessment of an
-"old truism" to the growing set of papers on theory building
+``old truism'' to the growing set of papers on theory building
 in software engineering. With access to such large amounts
@@
 
-\begin{quote}{\em  fshull XXX }\end{quote}
+\begin{quote}{\em We now map our work to prior publications on theory-building in software engineering.
+That mapping appears in numerous comments through out the text (using the references offered by this reviewer) with the bulk of the comments being top of p2 and the conclusion.
+
+Notably, we now introduce the DIE as an example of such a theory (in the introduction), and explain that this is done on the basis of it providing a description of a Software development phenomenon accompanied by a rationale that explains why such behavior is expected.
+While we introduce in the conclusions several rationales that can explain why the effect called for by the theory was not observed in our dataset, we feel that future work is needed to collect data in support or contradiction of these explanations before a new theory can be formulated with confidence.
+}\end{quote}
 
@@
 
-Firstly, this draft now contains extensive notes on the different contexts of the
+First, this draft now contains extensive notes on the different contexts of the
 projects studied here. Those notes are offered in \tion{data_character} and \tion{171projects}.
 
-Secondly (and this was the ineffective step), we  searched the numerous   listed in \tion{data_character} and \tion{171projects}
-in order to check if any stratification contained the delayed issue effect. We failed to find a stratification
-containing an exponential expansion in the time to fix issues. The reason for this was due to the dataL those
-stratifications resulted in smaller sample sizes that exacerbated the large IQR's seen in this data
-(to see these large IQRs,  please consider the 50th and IQR columns of Figure~\ref{fig:fix}.IQR is the
-range 75th to 25th in the data. Note that most of our IQRs were larger than tan the 50th percentile; i.e. software
-data exhibits large variances-- where are exacerbated by the smaller samples seen in the stratifications
+Second (and this was the ineffective step), we  searched the categories   listed in \tion{data_character} and \tion{171projects}
+in order to check if any stratification contained the delayed issue effect. 
+That study   failed to find a stratification
+containing an exponential expansion in the time to fix issues. The reason for this was  the small size
+of those stratifications  exacerbated the large IQR's seen in this data\footnote{Recall that in a sorted list of numbers,
+the inter-quartile range, or IQR, is difference between the 
+  75th to 25th percentile value.}:
+\bi
+\item
+Our 171 projects   stratifies into subsets of varying sizes. Sorting those sizes, we get a list
+comprising of 17,12 then numerous much smaller stratifications.  Reasoning over such small samples
+is problematic in the general case and, in the case of our data, it is ever more problematic due to
+the large IQRS of the data:
+\item
+To see these large IQRs,  please consider the 50th and IQR columns of Figure~\ref{fig:raw} Note that most of our IQRs were larger than tan the 50th percentile; i.e. software
+data exhibits large variances-- where are exacerbated by the smaller samples seen in the stratifications.
+\ei
+Our conclusion from exploring the stratification's is, given the currently available data, we cannot check
+for a DIE effect in subsets of this data.
+
+Note that the above notes are now added to the paper at the end of Section 6.6.
  }\end{quote}
@@
 DETAILED REVIEW: The authors should expand on what they
-include in the "delayed issue effect" (DIE), in particular
+include in the ``delayed issue effect'' (DIE), in particular
 in relation to standard milestones in the life of an issue
@@
 \begin{quote} {\em
+We now provide a more explicit definition of what constitutes a defect (our study of issues is limited to defects) and the ``defect lifecycle'' phases we measure in \tion{data-collection}.
+These definitions greatly expand on the now-deleted list on 14:50 in teh previous version.
+In TSP, ``a defect is any change to a product, after its construction, that is necessary to make the product correct''.
+This includes both minor issues (typography) and severe issues (architecture changes, requirements errors).
+Time on a defect includes a) time to investigate/analyze a defect once discovered, b) time to craft and implement a fix, and c) time to verify and close the defect.
 
-We understand your comment here to be ``are we just reporting trivial errors?''
-As shown in \fig{dtypes}, it is true that around a quarter of the fixes were simple documentation changes.
-That said, 75\% of the changes are quote  elaborate;   e.g.
-fixes to function necessitates a careful reflection of the purpose of the code.
+Regarding the concern that we may only be reporting trivial defects, we have added \fig{dtypes}, which shows the distribution of defect types in our data.
+It is true that around a quarter of the fixes were documentation changes, but the other 75\% are function-altering behavior in the system.
 }\end{quote}
@@
 \begin{quote} {\em
-At your suggestion, we had added extensive notes on the nature of our data--
+At your suggestion, we have added extensive notes on the nature of our data--
 see \tion{data_character} and \tion{171projects}.
@@
 
-\begin{quote}{{\em 
+\begin{quote}{\em 
 Yes, there is undoubtedly some accuracy in the time-tracking.
-We provide some evidence on the TSP data set's integrity in Section 6.3, and a more extensive discussion on the issue of time-tracking and defect logging accuracy in Section 7.1 on Conclusion Validity.XXX lucas: please check the section numbers are currently
-correct.
+We provide some evidence on the rigor of the TSP data collection process in \tion{defin}, \tion{data_integrity}, and \tion{data_character}, and a more extensive discussion on the issue of time-tracking and defect logging accuracy in \tion{conclusion_validity} on Conclusion Validity.
+We note that the data being collected here was not collected
+for the purposes of this paper. Rather, it was being collected and used by the
+programmers to assist them in their day-to-day management of the project.
+Given this, we have confidence that the logging information is not widely inaccurate.
+
+Please also consider the sanity checks reported in a companion paper (referenced here \tion{data_integrity}). In summary, this data has been heavily vetted.
 }
@@
 series (GTSE). I suggest looking into the following
-references for a start: Sjøberg et al. (2008) "Building
-Theories in Software Engineering", Smolander and Päivärinta
-(2013) "Theorizing about software development practices",
+references for a start: Sj{\o}berg et al. (2008) ``Building
+Theories in Software Engineering'', Smolander and Paivarinta
+(2013) ``Theorizing about software development practices'',
 and Stol and Fitzgerald (2015) "Theory-oriented software
@@
  \begin{quote}{\em 
-XXX Fshull
+We thank the reviewer for this important observation and have connected our work to the literature on theory-building.
+However, our data supports a number of possible explanations for the observed intermittent appearance of the DIE.
+Since the current data is not suitable for choosing among the various hypothesized explanations we feel that formulating a new theory build from this data
+would be premature at this time. That said, we think the current paper would be useful
+as a data point which, when combined with other sources, could be useful for some new future theory of SE.
 }\end{quote}
@@
 \begin{quote}{\em 
- We have reviewed and cited the mentioned papers. We have incorporated many of the concerns raised in these papers into our expanded Section 7.4 on External Validity.
-XXX lucas: please check the section numbers are currently
-correct. XXX section 7.4 is a little... terse. can you add another 1/4 to 1/2 page of stuff? or is this sectionn just as large as it neeeds to be
+ We have reviewed and cited the mentioned papers. We have incorporated many of the concerns raised in these papers into the expanded \tion{external_validity} on External Validity.
+ There, we note that Sj{\o}berg et al.
+list three classes of threats to External Validity (the smallest number) and we address all three of them.
+In truth, the only way to address External Validity threats is to have a huge sample, broadly representative of your domain, over an extended period of time-- e,g.
+using the 171 projects from 2006 to 2014 studied in this paper.
+The issues external validity in observational studies raised in \cite{madigan2014} and \cite{carlson2009} are more applicable to studies that compare a treatment and a control, whereas we are observing an effect that leads us to question the theory that DIE is a truism.
 }\end{quote}
@@
 survey of software engineers, both practitioners and
-researchers. According to Fig. 2 (actually a table), the
+researchers. According to Fig. 2 (actually a table) [NOTE: Fig. 2 is now \fig{survey_results}], the
 number of respondents is 16 and 30 for practitioners and
@@
 strengthen the paper if the authors reported how they
-reviewed the literature. According to Page 19:34 only eight
+reviewed the literature. According to Page 19:34 [NOTE: the enumerated list in the Conclusion] only eight
 publications were identified.
 
-\begin{quote}{\em XXX
-Lucas: need to show that these were the only empirical
-results for DIE in tow
-}\end{quote}
+\begin{quote}{\em We agree with the reviewer - the phrasing of our claims and the supporting evidence were not congruent.
+Our intention is to argue from multiple angles that the evidence for DIE is sparse and, more importantly, that DIE persists as a popular belief.
+Our goal here is to motivate that we should re-investigate the DIE.
+We have toned down the language and combined our arguments on ``widely-believed'' and ``poorly-documented'' into a new \tion{belief} whose focus is to motivate our study.}\end{quote}
 
@@
 short discussion section could be extended). The phases of
-the linear development of the 80s, (such as in Fig. 1)
+the linear development of the 80s, (such as in Fig. 1) [NOTE: Fig. 1 remains \fig{b81}] 
 probably don't exist in many of the projects in the TSP
@@
 
-\begin{quote}{\em An astute observation. We provide more information on the types of projects included in our sample in Section 6.4. XXX Lucas. Check section number.
-XXX Lucas: can't see the "percent agile" in 6.4. Can you work this with Bill?
+\begin{quote}{\em An astute observation. We provide more information on the types of projects included in our sample in \tion{data_character}
 
@@
 The answer is - we don't know.
-A strong argument can be made that all iterative&incremental methodologies are meant to mitigate the DIE.
+A strong argument can be made that all iterative \& incremental methodologies were designed to mitigate the DIE.
 Further, huge advances in software engineering technology and process have occurred since DIE was first observed, many of which have precisely targeted the DIE (static analysis, test-first development, automated build).
@@
 We do not yet know the causes here, but certainly it appears the DIE should not be treated as a truism any more, but rather a variable that can be controlled.
-We discuss this in Section 8.
+We have added more discussion on this to \tion{discussion}.
 }\end{quote}
@@
 
-\begin{quote}{\em Those notes are offered in \tion{data_character} and \tion{171projects}. \end{quote}
+\begin{quote}{\em Much more context is now offered in \tion{data_character} and \tion{171projects}}. \end{quote}
 
-Fig 1: "Operation" dwarfs everything in Boehm's diagram.
+Fig 1 [NOTE: Still \fig{b81}]: "Operation" dwarfs everything in Boehm's diagram.
 Since you do not study anything post release in this paper,
@@
 you remove the rightmost bar and rescale accordingly, the
-plot better matches the findings you report in Fig. 10. You
+plot better matches the findings you report in Fig. 10 [NOTE: now \fig{raw}]. You
 still identify an interesting result though, but the
@@
 
-\begin{quote}{\em We recently published a text on classic SE papers. We know how to manage those issues.}\end{quote}
+\begin{quote}{\em We have removed all copyrighted images and replaced them with original drawings based on information from the cited source.}\end{quote}
 
@@
 
-\begin{quote}{\em  We agree: that reference has been deleted. }\end{quote}
+\begin{quote}{\em  We agree: that statement has been deleted. }\end{quote}
 
@@
 \begin{quote}{\em  Our policy is to get one publication out of paper
-before issueing the data in public.
+before issuing the data in public.
 If this paper is accepted, we will certainly add that link.
@@
 
-\begin{quote}{\em XXX Lucas. Not sure. but do you hanle all the following? }\end{quote}
-
 Page 18:27 - "We also survey the state of SW dev. practice /---/": given the size of the survey,
@@
 
+\begin{quote}{\em We have removed this statement.}\end{quote}
+
 TYPOS ETC.
+
 Sec 1, §1: [3] and [30] have been swapped?
+
 Page 14:46 - Spell out IV\&V the first (and only) time it appears.
+
 Page 3:24 - missing verb (is) Page 16:43 - "Did this study failed"
+
 Page 18:44 - Appache
+
 Page 19:19 - Al Ref 26 - oo --> OO
+
+\begin{quote}{\em  All of these have been corrected. }\end{quote}
 
@@
 \begin{quote}{\em 
-Those notes are offered in \tion{data_character} and \tion{171projects}.}\end{qupte}
+More description about the context and sample have been added in \tion{data_character} and \tion{171projects}.}\end{quote}
 
@@
  also be relevant. Accordingly, the formula for calculating
- variable "Total effort" in Fig. 7 could be explained, and
- Figure 9 made more readable (bigger size).
+ variable "Total effort" in Fig. 7 [NOTE: now \fig{dist}] could be explained, and
+ Figure 9 [NOTE: now \fig{fix-phase-dist}] made more readable (bigger size).
 
- \begin{quote}{\em We have added more description of the projects to Section 6.4.
-We have also added more precise definitions of our measures in Section 6.2.
-We have adjusted the size of figures throughout the paper to be more readable.XXX Lucas please check section references.
+ \begin{quote}{\em We have added more description of the projects in \tion{data_character}.
+We have also added more precise definitions of our measures in \tion{data-collection}.
+We have adjusted the size of figures throughout the paper to be more readable.
 We have added more text
@@
 \begin{quote}{\em 
-Those notes are offered in \tion{data_character} and \tion{171projects}.}\end{qupte}
+Those notes are offered in \tion{data_character} and \tion{171projects}.}\end{quote}
 
-Figure 10 is central to the paper's demonstration; its
+Figure 10 [NOTE: now \fig{raw}] is central to the paper's demonstration; its
 expressiveness could be enhanced: the formula for calculating the 50th
@@
 \begin{quote}{\em
-\fig{raw}'s caption now has extensive explanation text. e.g. units
+\fig{raw}'s caption (formerly Figure 10) now has extensive explanation text. e.g. units
 are added on line two. Also, an example of the calculations of the ratios is offered
-in the last line of the caption. }\end{quote}
+in the last line of the caption. Further,
+to simplify comprehension of this diagram, it has been split into two
+(and the more arcane discussion about the outliers, statistics gathered, and methods used is provided is also provided).}\end{quote}
 
@@
 \begin{quote}{\em
-Not explicitly, but \fig{dtypes} lists "interfaces" as a the fourth most
-common type of defect. Note that "interfaces" are changed by architectural
+Not explicitly, but \fig{dtypes} lists ``interfaces'' as a the fourth most
+common type of defect. Note that ``interfaces'' are changed by architectural
 issues. }\end{quote}
@@
 \begin{quote}{\em
-  Absolutely - there are potentially many explanations for why the DIE was NOT observed in our dataset. We discuss some more of the potential causes in Section 8.
-  XXX Lucas. Is the section reference right? Do we reference Harter (and if we do, mention that here).?
+  Absolutely - there are potentially many explanations for why the DIE was NOT observed in our dataset.
+We add more discussion on the potential causes in \tion{discussion}, supported by the reference to Harter et al.
  }\end{quote}
@@
 
+Minor remarks:
+
+- p.9, line 29 : " … reported by Shull [52], found that the cost to find certain non-critical classes of defects ..." => is it the "cost to find" or the "cost to FIX"?
+
+- p.10, line 21: duplicate word "there" -
+
+p.12, line19: missing word "One of THE guiding …"
+
+- p.16, line 33: "distinguish" instead of "distinguished"?
+
+- p.18, line 42: explain acronym MEAN
+
+- p.20, line 20: useless enumeration A1?
+
 \begin{quote}{\em
- XXX Lucas: can you retire the follwoing
+ All of these have been corrected or removed.
  }\end{quote}
 
-Minor remarks: - p.9, line 29 : " … reported by Shull [52], found that the cost to find certain non-critical classes of defects …" => is it the "cost to find" or the "cost to FIX"?  
-```
-
-- [X]  llayman It should be "cost to fix". We have corrected this.
-
-- p.10, line 21: duplicate word "there" -
-p.12, line19: missing word "One of THE guiding …" - p.16,
-line 33: "distinguish" instead of "distinguished"?  - p.18,
-line 42: explain acronym MEAN - p.20, line 20: useless
-enumeration A1?
-
-{\bf  Reviewer \#3's Comemnts}:
+{\bf  Reviewer \#3's Comments}:
 
@@
 
-The empirical analysis (Section 6) is a bit of a let down. I
+The empirical analysis (Section 6 [NOTE: now \tion{analysis}) is a bit of a let down. I
 would like to see some sort of more robust and rigorous
@@
 
-Section 2
+Section 2 [NOTE: still \tion{claims}]
 
@@
 general property of the measure) if $\sum_{1 <= i < i} m(i)
-< m(j)$. Then, it's up to you to make this a property of the
-measure, for example by using an existential "policy" ($m$
-has this property if there exists one phase $j$ where
-$\sum_{1 <= i < i} m(i) < m(j)$) or a universal one ($m$ has
-this property if for all phases $1<j$, $\sum_{1 <= i < i}
-m(i) < m(j)$).
+< m(j)$. 
 \begin{quote}{\em
-Thanks for that. We added the $m_j$ notation to Section 2, as you suggested.
+Thanks for that. We added the $m_j$ notation to \tion{claims}, as you suggested.
  }\end{quote}
@@
 of effort. The reasons why you introduce this definition
-become clearer only in Section 7.2, in the Threats to
+become clearer only in Section 7.2 [NOTE: now \tion{construct}], in the Threats to
 Validity, and that's too late. You should move some of the
@@
 \begin{quote}
-{\em Good idea! We did just that- see middle para in Section 2.}
+{\em Good idea! We did just that- see middle para in \tion{claims}.}
 \end{quote}
@@
 
-A very minor issue: why is Figure 2 a ... figure, instead of
+A very minor issue: why is Figure 2 [NOTE: now \fig{survey_results}] a ... figure, instead of
 a table? Same for all other tables ...
@@
 \begin{quote}
-{\em We now provide a more precise definition of defect in Section 6.2.
-XXXX= Lucas: check section number.
+{\em We now provide a more precise definition of defect in \tion{defin}.
 }
 \end{quote} 
-You do not explain that "QualTest" in Fig. 8 is.
+You do not explain that "QualTest" in Fig. 8 [NOTE: now \fig{waterfall}] is.
 \begin{quote}
-{\em Removed}
+{\em A mistake on our part that has been removed from the figure.}
 \end{quote}
-The definition of "time per defect - The total # of defects found in a plan item during a removal phase divided by the total time spent on that plan item in that phase." is not correct as it is, as this would basically be the number of defects per unit time, instead.
+The definition of "time per defect - The total \# of defects found in a plan item during a removal phase divided by the total time spent on that plan item in that phase." is not correct as it is, as this would basically be the number of defects per unit time, instead.
 The roles of time and
 defects should be reversed.
-\begin{quote}{\em We now provide a more precise definition of time measurement related to defects in Section 6.2.}\end{quote}
-
-Section 6.4
+\begin{quote}{\em We now provide a more precise definition of time measurement related to defects in \tion{data-collection}}\end{quote}
 
@@
 
-Figure 9. The caption says "Distribution of defects found
+Figure 9 [NOTE: now \fig{fix-phase-dist}]. The caption says "Distribution of defects found
 and fixed by phase." Is that the distribution of defects
@@
 our data is shown in Figure 9. A high percentage of defects
-(44\) were found and fixed in the early phases, i.e.,
+(44) were found and fixed in the early phases, i.e.,
 requirements, high level design, and design reviews and
@@
 \begin{quote}
-{\em XXXX Bill: one for you.}
+{\em 
+The TSP convention is to fix a defect in the phase in which it was identified, thus "find and fix".
+With some exceptions, defects are not counted as defects unless they escape that phase.
+The exception case is the secondary defect introduced during a fix that occurs with some frequency in test and rarely in inspection or review.
+We have clarified this distinction in the text.}
 \end{quote}
-Figure 10. The explanation of the data and histograms in the
+Figure 10 [NOTE: now \fig{raw}]. The explanation of the data and histograms in the
 figure is, at best, confusing. The caption says "50th and
@@
 \begin{quote}
-{\em \fig{raw} now has an extensive help text in the caption.
+{\em We agree-that figure was far too confusing. We have now split the 50th and 90th peercentile results
+into two charts, which simplifies both. Please see \fig{raw} and \fig{last}.
+Also, please note that that fit\fig{raw} now has an extensive help text in the caption.
 We used your comments there where we noted the ratios used in the
@@
 
-Section 7.1
-
 Maybe you could analyze the projects developed in a
@@
 life cycles (like Agile ones) help solve DIE. You do provide
-hints about this in Section 8 ("We also note that other
+hints about this in Section 8 [NOTE: now \tion{discussion}] ("We also note that other
 development practices have changed in ways that could
-mitigate the delayed issued effect") and Section 9, but
+mitigate the delayed issued effect") and Section 9 [NOTE: now \tion{conclusion}], but
 that's clearly not enough.
 \begin{quote}{\em
+
 If your comment is along the lines of ``look for stratifications to check if
-DIE exists in subsets'' then we offer the following failed study.
+DIE exists in subsets'' then we offer the following notes (which we added to the end of \tion{171projects}):
 
-We looked into those stratifications and have  failed to find a stratification
-containing an exponential expansion in the time to fix issues. The reason for this was due to the data in  those
-stratifications resulted in smaller sample sizes that exacerbated the large IQR's seen in this data
-(to see these large IQRs,  please consider the 50th and IQR columns of Figure~\ref{fig:fix}.IQR is the
-range 75th to 25th in the data. Note that most of our IQRs were larger than tan the 50th percentile; i.e. software
-data exhibits large variances-- where are exacerbated by the smaller samples seen in the stratifications.}
+We looked into   stratifications of the data and, so far,   failed to find a stratification
+containing an exponential expansion in the time to fix issues. The reason for this was  the small size
+of those stratifications  exacerbated the large IQR's seen in this data\footnote{Recall that in a sorted list of numbers,
+the inter-quartile range, or IQR, is difference between the 
+  75th to 25th percentile value.}:
+\bi
+\item
+Our 171 projects   stratifies into subsets of varying sizes. Sorting those sizes, we get a list
+comprising of 17,12 then numerous much smaller stratifications.  Reasoning over such small samples
+is problematic in the general case and, in the case of our data, it is ever more problematic due to
+the large IQRS of the data:
+\item
+To see these large IQRs,  please consider the 50th and IQR columns of Figure~\ref{fig:raw} Note that most of our IQRs were larger than tan the 50th percentile; i.e. software
+data exhibits large variances-- where are exacerbated by the smaller samples seen in the stratifications.
+\ei
+Our conclusion from exploring the stratifications is, given the currently available data, we cannot check
+for a DIE effect in subsets of this data.
+}
 \end{quote}
-Section 8
-
-- [ ] Verify that the Section #'s in our responses are accurate prior to submission.

### 2016-07-06 phaseDelay4Emse.tex
Updates from ShareLaTeX

@@
 
+\titlerunning{Are Delayed Issues Harder to Resolve? }
+
 \pagenumbering{arabic} 
@@
 \date{}
- \title{Are Delayed Issues Harder to Resolve?}
+ \title{Are Delayed Issues Harder to Resolve? Revisiting Cost-to-Fix of Defects throughout the Lifecycle}
 
@@
 Accordingly, this paper revisits
-the commonly held belief we label the {\em delayed issue effect} (hereafter, DIE):
+the commonly held theory we label the {\em delayed issue effect} (hereafter, DIE):
 % Later in this paper, we offer a precise definition for this effect.
@@
  the relative cost of fixing requirements issues at different phases of a project).
-
-This theory has been used as the basis for decision-making in software engineering. For example,  Basili and Boehm comment that, since the 1980s,  this effect
+As a falsifiable theory, the DIE can be compared to empirical data and, if inconsistencies are observed, refinements to the theory may be generated that better describe the phenomenon under observation ~\cite{Popper}.
+
+The DIE theory is worth examination since it has been used as the basis for decision-making in software engineering. For example,  Basili and Boehm comment that, since the 1980s,  this effect
  \begin{quote}
@@
 We examined the literature on the delayed issue effect through a combination of snowball sampling~\cite{wohlin2014guidelines} and database search.
-We began with \cite{boehm2012}, the most recent publication by Barry Boehm (at the time of this paper's first draft) that discusses the delayed issue effect.
-Boehm cites the seminal \textit{Software Engineering Economics}~\cite{Boehm81}.
-%, which is bases the original delayed issue effect on four primary sources in the late 70s from IBM~\cite{Fagan76}, TRW~\cite{Boehm76}, GTE~\cite{Daly77}, and Bell Labs~\cite{Stephenson76}. 
 We searched Google Scholar for terms such as ``cost to fix'' and ``defect cost'' and ``software quality cost''.
-The majority of the search results discuss quality measurements or quality improvement, but relatively few discuss cost-to-fix as a function of when the defect was injected or found.
-Our literature search is by no means exhaustive, but our search results yielded many popular sources such as~\cite{pressman2005software, boehm01,glass02,endres03}\footnote{These sources have a combined citation count of over 14,500 on Google Scholar}, all of which cite \textit{Software Engineering Economics}~\cite{Boehm81} or its derivative works.
+The majority of the search results discuss quality measurements, quality improvement, or the cost savings of phase-specific quality improvement efforts (e.g., heuristic test case selection vs.
+smoke testing).
+A systematic literature review of software quality cost research can be found in \cite{karg2011systematic}.
+Relatively few articles discuss cost-to-fix as a function of when the defect was injected or found.
+We also conducted a general Google search for the above terms.
+We found a number of website articles and blog postings on this topic, e.g., \cite{IfSQ2013,Soni2016,Parker2013,Gordon2016}.
+From these, we gathered additional citations for the delayed issue effect, the vast majority of which were secondary sources, e.g.,~\cite{Leffingwell96,mead2004software,mcconnell1996software,mcconnell01,Tassey2002,boehm2012}.
+Our literature search is not exhaustive, but our results yielded an obvious trend: nearly every citation to the delayed issue effect could be traced to the seminal \textit{Software Engineering Economics}~\cite{Boehm81} or its related works~\cite{boehm88,boehm01}.\footnote{For example, popular sources such as~\cite{pressman2005software, boehm01,glass02,endres03} have a combined citation count of over 14,500 on Google Scholar can all trace their evidence to \textit{Software Engineering Economics}~\cite{Boehm81}.}
+
+Ultimately, we identified nine sources of evidence for the delayed issue effect based on real project data: the original four \cite{Fagan76,Boehm76,Daly77,Stephenson76} reported in \textit{Software Engineering Economics}\cite{Boehm81}, a 1995 report by Baziuk~\cite{baziuk1995bnr} on repair costs at Nortel, a 1998 report by Willis et al.~\cite{willis1998hughes} on software projects at Hughes Aircraft, a 2002 experiment by Westland~\cite{westland2002cost} to fit regression lines to cost-to-fix of localization errors, a 2004 report by Stecklein et al.~\cite{steck04} on cost-to-fix in five NASA projects, and a 2007 survey by Reifer on CMMI Level 5 organization\cite{reifer2007profiles}.
 
@@
 The data points from these studies are not published for analysis.
-%These seminal studies are the basis for most DIE claims and that the cost to find and fix a defect monotonically increases with lifecycle phase.
-
-In the 40 years since these initial studies, few studies have explored the cost to resolve issues
-as a function of lifecycle phase. 
+Baziuk~\cite{baziuk1995bnr} reports an exponential increase in the cost to patch software in the field versus system test, and Stecklein et al.~\cite{steck04} produce a cost-to-fix curve (as price) that fits precisely with \fig{cost-to-fix}.
+Westland~\cite{westland2002cost} finds that the cost to fix engineering errors is exponentially related to the cost of the overall cost of a case study project.
+Reifer~\cite{reifer2007profiles} confirms the exponential increase in the DIE in 19 CMMI Level 5 organizations though this appears to be based on survey rather than empirical data.
+
+% \begin{figure}[!ht]
+% \centering
+% \begin{tabular}{lllll}
+%                 &   Inception   &   Elaboration &   Construction    &   Transition  \\\hline
+% Inception       &   25-100      &               &                   &               \\
+% Elaboration     &   100-500     &   50-250      &                   &               \\
+% Construction    &   500-1,000   &   250-1,500   &   75-500          &               \\
+% Transition      &   8,000-10,000  &  1,500-5,000   &   500-3,000        &   Not enough data\\
+% \end{tabular}
+% \caption{Range of cost to find and fix defects (\$US per defect) from~\cite{reifer2007profiles}.}
+% \label{fig:reifer}
+% \end{figure}
+
 Shull et al.~\cite{Shull02} conducted a literature survey and held a series of e-workshops with industry experts on fighting defects.
@@
 Some studies that report less-than large
- increase in the effort required to fix delayed issues.
+ increase in the effort required to fix delayed issues.  Boehm~\cite{Boehm80} provides data suggesting that the cost-to-fix curve for small projects
+ %, two student projects of 2000 deliverable source instructions, 
+ is flatter than for large projects (the dashed line of Figure~\ref{fig:cost-to-fix}).
+Data from NASA's Johnson Space Flight Center, reported by Shull~\cite{Shull02}, found that the cost to fix certain non-critical classes of defects was fairly constant across lifecycle phases (1.2 hours on average early in the project, versus 1.5 hours late in the project).
 Royce~\cite{Royce98} studied  a million-line, safety-critical missile defense system.
 Design changes (including architecture changes) required approximately twice the effort of implementation and test changes, and the cost-to-fix in implementation and test phases increased slowly.
-Boehm~\cite{Boehm10} attributes this success to the  development process, which focused on removing architecture risk early in the development lifecycle.
-Boehm~\cite{Boehm80} provides data suggesting that the cost-to-fix curve for small projects, two student projects of 2000 deliverable source instructions, is flatter than for large projects (the dashed line of Figure~\ref{fig:cost-to-fix}).
-Data from NASA's Johnson Space Flight Center, reported by Shull~\cite{Shull02}, found that the cost to fix certain non-critical classes of defects was fairly constant across lifecycle phases (1.2 hours on average early in the project, versus 1.5 hours late in the project).
+Boehm~\cite{Boehm10} attributes this success to a development process focused on removing architecture risk early in the lifecycle.
+Willis et al.
+(\cite{willis1998hughes}, page 54) provide tables summarizing the effort to fix over 66,000 defects as a function of lifecycle phase injected and removed from multiple projects.
+The tables are partly obscured, but seem to provide the first large scale evidence that a)~DIE need not be exponential and b)~DIE need not be monotonically increasing.
 Again, the data points from these studies are not available, and thus newer evidence both in favor of and contrary to the DIE cannot be evaluated.
@@
 Only one of these sources, \textit{Software Engineering Economics}~\cite{Boehm81}, is based on real project data.
-Despite the 40-year old evidence, the perception of the DIE persists today among both the software engineers sampled in our survey and in popular literature.
+Despite a lack of recent evidence, the perception of the DIE persists today among both the software engineers sampled in our survey and in popular literature.
 In the intervening years, many advances in software technology and processes have been made precisely to deal with risks such as the DIE.
@@
     The Scott-Knott ranker 
-    recommended was   recommended  by  Mittas and Angelis in a
+      was   recommended  by  Mittas and Angelis in a
     recent TSE'13 article~\cite{mittas13} and by Ghotra et al. in a recent
@@
 The left-hand-side ``rank'' column shows the result of the Scott-Knott ranking procedure described in \tion{stats}. Note that
-most treatments achieved the same ranks i.e. they were found to be insignificantly different  (the one exception is the Planning:UnitTest results 
+most treatments achieved the same ranks i.e. they were found to be insignificantly different  from each other (the one exception is the Planning:UnitTest results 
 where UnitTests were ranked 2).
@@
 
-The twp key features of \fig{raw} are:
+The two key features of \fig{raw} are:
 \be
@@
 data exhibits large variances-- where are exacerbated by the smaller samples seen in the stratifications).
-Our conclusion from exploring the stratification's is, given the currently available data, we cannot check
+Our conclusion from exploring the stratification's is that, given the currently available data, we cannot check
 for a DIE effect in subsets of this data.
@@
 The guidance for logging time and defects is precisely defined.
-All tasks should be logged as the work is performed with a stopwatch like tool.
+All tasks should be logged as the work is performed with a stopwatch tool.
 All defects that escape a phase must be logged.
@@
  of systems (ranging from e-commerce web portals to  banking systems) run in a variety of
- ways (agile or  waterfall or some combination of the two)These are realistic settings for contemporary software development teams, though perhaps not representative of all types of projects (see prior paragraph).
+ ways (agile or  waterfall or some combination of the two).
+These are realistic settings for contemporary software development teams, though perhaps not representative of all types of projects (see prior paragraph).
 
@@
 
-Earlier we noted that the delayed issue effect was first reported in 1976 in a era of punch card programming and non-interactive environments~\cite{Boehm76}.
+Earlier we noted that the delayed issue effect was first reported in 1976 in an era of punch card programming and non-interactive environments~\cite{Boehm76}.
 We also note that other development practices have changed in ways that could   mitigate the delayed issued effect.
@@
 % continually adjust  their modules in response to security issues~\cite{theisen2015}.
-Upgrades to the Microsft Windows operating system is moving from service patches and major releases to a stream of updates (so there will be no Windows 11- just a stream of continuous updates to what is currently called Windows 10)~\cite{bright15}.
+Upgrades to the Microsoft Windows operating system are moving from service patches and major releases to a stream of updates (so there will be no Windows 11- just a stream of continuous updates to what is currently called Windows 10)~\cite{bright15}.
 
@@
 
-Even organizations that build complex, high assurance systems are turning to agile development processes that purport to address the DIE. In 2010, the US Department of Defense mandated
-that all new software acquisitions must adopt agile methods~\cite{kim13}. 
-This change in DoD policy is enabled by a separation of baseline architecture
+Even organizations that build complex, high assurance systems are turning to agile development processes that purport to address the DIE.
+For example, agile methods have been advocated for software acquisitions within the US Department of Defense ~\cite{kim13}, and interest and adoption has been growing ~\cite{lapham11}.
+This change in DoD culture is enabled by a separation of baseline architecture
 (e.g., the design of an 
@@
 Many software engineering technologies have been created precisely to avoid the delayed issue effect by removing risk as early as possible.
-Boehm's spiral model~\cite{boehm1988spiral}, Humphrey's PSP~\cite{humphrey1995discipline} and TSP~\cite{tsp00}, the Unified Software Development Process~\cite{jacobson1999unified}, and agile methods~\cite{Beck2001a} all in part or in whole focus on removing risk early in the devleopment lifecycle.
+Boehm's spiral model~\cite{boehm1988spiral}, Humphrey's PSP~\cite{humphrey1995discipline} and TSP~\cite{tsp00}, the Unified Software Development Process~\cite{jacobson1999unified}, and agile methods~\cite{Beck2001a} all in part or in whole focus on removing risk early in the development lifecycle.
 Indeed, this idea is core to the whole history of iterative and incremental product development dating back to ``plan-do-study-act'' developed at Bell Labs in the 1930's~\cite{larman2003iterative} and popularized by W.
@@
 Our results are evidence that the software engineering community has been successful in meeting one of its over-arching goals.
-But our results raise an equally important point - should the DIE persist as a truism (see \S\ref{sect:belief}), or should is it a project outcome that can be controlled by software engineering process and technology?
+But our results raise an equally important point - should the DIE persist as a truism (see \S\ref{sect:belief}), or is it a project outcome that can be controlled by software engineering process and technology?
 
@@
 increases the difficulty of completing that  resolution).
-Several prominent SE researchers state this effect is a fundamental law of software engineering~\cite{boehm01,mcconnell01,boehm01,glass02}.
+Several prominent SE researchers state this effect is a fundamental law of software engineering~\cite{mcconnell01,boehm01,glass02}.
 Based on a  survey  of both researchers and practitioners, we  found that
@@
 
-We do not claim that this effect {\em never} holds in software projects; just that it cannot be assumed to {\em always} hold.
+We do not claim that this theory {\em never} holds in software projects; just that it cannot be assumed to {\em always} hold, as data have been found that falsify the general theory.
 Our explanation of the observed lack-of-effect is five-fold.

### 2016-07-06 phaseDelay4Emse.tex
typo

@@
 \begin{center}
-\includegraphics[,width=4in]{img/b81.png}
+\includegraphics[,width=4in]{b81.png}
  \end{center}
@@
 \begin{figure}[!t]
-% \begin{center}\includegraphics[width=4in]{img/boehm_cost-to-fix.png}\end{center}
-\begin{center}\includegraphics[width=4.5in]{img/boehm81.png}\end{center}
+% \begin{center}\includegraphics[width=4in]{boehm_cost-to-fix.png}\end{center}
+\begin{center}\includegraphics[width=4.5in]{boehm81.png}\end{center}
  \caption{Historical cost-to-fix curve. Adapted from~\cite{Boehm81}, p. 40.}\label{fig:cost-to-fix}
@@
 \begin{center}
-\includegraphics[width=2.5in]{img/defect_type_frequency.png}
+\includegraphics[width=2.5in]{defect_type_frequency.png}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=1.6in]{img/wfall.png}  
+\includegraphics[width=1.6in]{wfall.png}  
 \end{center}
@@
 \begin{center}
-\includegraphics[height=1.5in]{img/number_of_projects_per_org.png}
+\includegraphics[height=1.5in]{number_of_projects_per_org.png}
 \end{center} 
@@
                                                                    & N   & Min & Q1   & Median & Q3    & Max   & Mean   &  Distribution \\\hline
-Team size                                                          & 171 & 1   & 4    & 6      & 10    & 36    &  7.8    &  \includegraphics[width=15mm]{img/team_size.png} \\ 
-Duration {[}days{]}                                                & 171 & 7   & 33   & 61     & 118   & 1918  & 107    & \includegraphics[width=15mm]{img/duration.png} \\
-Added \& Modified LOC                                                 & 117 & 2   & 1125 & 4201   & 13092 & 88394 & 10259  & \includegraphics[width=15mm]{img/LOC.png} \\
-Components                                                         & 171 & 8   & 26   & 49     & 107   & 4170  & 116.2  & \includegraphics[width=15mm]{img/components.png} \\
-Year                                                               & 171 &     & 2011 & 2012   & 2013  & 2014  & 2011.9 & \includegraphics[width=15mm]{img/year.png}
+Team size                                                          & 171 & 1   & 4    & 6      & 10    & 36    &  7.8    &  \includegraphics[width=15mm]{team_size.png} \\ 
+Duration {[}days{]}                                                & 171 & 7   & 33   & 61     & 118   & 1918  & 107    & \includegraphics[width=15mm]{duration.png} \\
+Added \& Modified LOC                                                 & 117 & 2   & 1125 & 4201   & 13092 & 88394 & 10259  & \includegraphics[width=15mm]{LOC.png} \\
+Components                                                         & 171 & 8   & 26   & 49     & 107   & 4170  & 116.2  & \includegraphics[width=15mm]{components.png} \\
+Year                                                               & 171 &     & 2011 & 2012   & 2013  & 2014  & 2011.9 & \includegraphics[width=15mm]{year.png}
 \end{tabular}
@@
 \begin{center}
-\includegraphics[height=2.5in]{img/fix-phase-dist.pdf}
+\includegraphics[height=2.5in]{fix-phase-dist.pdf}
 \end{center} 
@@
 \begin{center}
-\input{tex/fig13}
+
+%\resizebox{\columnwidth}{!}{
+\scriptsize
+\begin{tabular}{c|lr|rr|rl}
+& \multicolumn{2}{c|}{ } & \multicolumn{2}{c|}{Percentiles} & \multicolumn{2}{c}{~ } \\ 
+  & \multicolumn{2}{c|}{ } & \multicolumn{2}{c|}{(units = } & \multicolumn{2}{c}{Growth with respect to earliest phase  } \\ 
+   & \multicolumn{2}{c|}{Phase} & \multicolumn{2}{c|}{  minutes)} & \multicolumn{2}{c}{(unitless  ratios of two time values) } \\\hline
+
+  rank & injected & removed & 50th &   IQR & \multicolumn{2}{c}{50th percentile growth}  \\ 
+\hline
+
+\multicolumn{7}{c}{~}  \\
+1 &  Before   
+      &   DesignInspect   & 10 &  14 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}}  \\
+1 &   &   CodeReview      & 8 &  14 & 0.80 & \textcolor{black}{\rule{8mm}{2mm}}  \\
+1 &   &   CodeInspect     & 10 &  16 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}}  \\
+1 &   &   UnitTest        & 12 &   21 & 1.20 & \textcolor{black}{\rule{12mm}{2mm}}  \\
+1 &   &   IntTest         & 15 &   31 & 1.50 & \textcolor{black}{\rule{15mm}{2mm}}  \\
+1 &   &   SysTest         & 11 &   22 & 1.10 & \textcolor{black}{\rule{12mm}{2mm}}  \\  
+\hline\multicolumn{7}{c}{~}  \\
+1 &  Planning     &   ReqtsReview     & 8 &  14 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}}  \\
+1 &               &   DesignInspect   & 11 &  13 & 1.38 & \textcolor{black}{\rule{13mm}{2mm}} \\
+2 &               &   UnitTest        & 24 &   25 & 3.00 & \textcolor{black}{\rule{30mm}{2mm}} \\
+\hline\multicolumn{7}{c}{~}  \\
+1 &  Reqts   &   ReqtsReview     & 13 &  20 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}} \\
+1 &          &   ReqtsInspect    & 12 &   18 & 0.92 & \textcolor{black}{\rule{9mm}{2mm}}  \\
+1 &          &   DesignReview    & 10 &   14 & 0.77 & \textcolor{black}{\rule{7mm}{2mm}} \\
+1 &          &   DesignInspect   & 9 &   15 & 0.69 & \textcolor{black}{\rule{6mm}{2mm}}   \\
+1 &          &   CodeInspect     & 13 &  24 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}} \\
+1 &          &   UnitTest        & 10 &  17 & 0.77 & \textcolor{black}{\rule{7mm}{2mm}}  \\
+1 &          &   IntTest         & 33 &   42 & 2.54 & \textcolor{black}{\rule{25mm}{2mm}}  \\
+1 &          &   SysTest         & 24 &  108 & 1.85 & \textcolor{black}{\rule{18mm}{2mm}}  \\
+\hline\multicolumn{7}{c}{~}  \\
+1 & Design   &   DesignReview    & 11 &   16 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}}  \\
+1 &   &   DesignInspect   & 8 &   12 & 0.73 & \textcolor{black}{\rule{7mm}{2mm}}  \\
+1 &   &   CodeReview      & 10 &   18 & 0.91 & \textcolor{black}{\rule{9mm}{2mm}}  \\
+1 &   &   CodeInspect     & 9 &  14 & 0.82 & \textcolor{black}{\rule{8mm}{2mm}}  \\
+1 &   &   UnitTest        & 11 &   18 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}}  \\
+1 &   &   IntTest         & 17 &   31 & 1.55 & \textcolor{black}{\rule{15mm}{2mm}}  \\
+1 &   &   AcceptTest      & 14 &   19 & 1.27 & \textcolor{black}{\rule{12mm}{2mm}}  \\
+1 &   &   SysTest         & 13 &   18 & 1.18 & \textcolor{black}{\rule{12mm}{2mm}}  \\
+\hline\multicolumn{7}{c}{~}  \\
+1 &  Code   &   CodeReview    & 10 &  16 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}}  \\
+1 &    &   CodeInspect   & 10 &   15 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}}  \\
+1 &   &   UnitTest      & 12 &   20 & 1.20 & \textcolor{black}{\rule{12mm}{2mm}}  \\
+1 &    &   IntTest       & 14 &   25 & 1.40 & \textcolor{black}{\rule{14mm}{2mm}}  \\
+1 &    &   AcceptTest    & 16 &   25 & 1.60 & \textcolor{black}{\rule{16mm}{2mm}}  \\
+1 &   &   SysTest       & 13 &   20 & 1.30 & \textcolor{black}{\rule{13mm}{2mm}}  \\
+
+\end{tabular}
+%}
 %
@@
 \begin{center}
-\input{tex/figlast}
+%\resizebox{\columnwidth}{!}{
+\scriptsize
+\begin{tabular}{c|lr|r|rl}
+  & \multicolumn{2}{c|}{ } &  Percentiles & \multicolumn{2}{c}{Growth with respect to earliest phase  } \\ 
+   & \multicolumn{2}{c|}{Phase} &  (units= miniutes) & \multicolumn{2}{c}{(unitless ratios of two time values) } \\\hline
+
+  rank & injected & removed &   90th   &  \multicolumn{2}{c}{90th percentile growth} \\ 
+\hline\multicolumn{6}{c}{~}  \\
+1 &  Before   &   DesignInspect     & 32  &  1.00 & \textcolor{black}{\rule{10mm}{2mm}}\\
+1 &   &   CodeReview      &   31   &    0.97 & \textcolor{black}{\rule{9mm}{2mm}}\\
+1 &   &   CodeInspect     &   32   &    1.00 & \textcolor{black}{\rule{10mm}{2mm}}\\
+1 &   &   UnitTest        &   45  &    1.41 & \textcolor{black}{\rule{14mm}{2mm}}\\
+1 &   &   IntTest         &   63  &   1.97 & \textcolor{black}{\rule{19mm}{2mm}}\\
+1 &   &   SysTest         &   46   &   1.44 & \textcolor{black}{\rule{14mm}{2mm}}\\  
+\hline\multicolumn{6}{c}{~}  \\
+1 &  Planning     &   ReqtsReview       & 35   &   1.00 & \textcolor{black}{\rule{10mm}{2mm}}\\
+1 &               &   DesignInspect     & 31  &   0.89 & \textcolor{black}{\rule{8mm}{2mm}}\\
+2 &               &   UnitTest          & 53 &   1.51 & \textcolor{black}{\rule{15mm}{2mm}}\\
+\hline\multicolumn{6}{c}{~}  \\
+1 &  Reqts   &   ReqtsReview      & 42 &    1.00 & \textcolor{black}{\rule{10mm}{2mm}}\\
+1 &          &   ReqtsInspect      & 40 &     0.95 & \textcolor{black}{\rule{9mm}{2mm}}\\
+1 &          &   DesignReview      & 34 &    0.81 & \textcolor{black}{\rule{8mm}{2mm}}\\
+1 &          &   DesignInspect     & 38 &     0.90 & \textcolor{black}{\rule{9mm}{2mm}}\\
+1 &          &   CodeInspect      & 45 &   1.07 & \textcolor{black}{\rule{10mm}{2mm}}\\
+1 &          &   UnitTest          & 40 &   0.95 & \textcolor{black}{\rule{9mm}{2mm}}\\
+1 &          &   IntTest          & 95 &      2.26 & \textcolor{black}{\rule{22mm}{2mm}}\\
+1 &          &   SysTest          & 126 &    3.00 & \textcolor{black}{\rule{30mm}{2mm}}\\
+\hline\multicolumn{6}{c}{~}  \\
+1 & Design   &   DesignReview     & 37 &   1.00 & \textcolor{black}{\rule{10mm}{2mm}}\\
+1 &          &   DesignInspect    & 28 &    0.76 & \textcolor{black}{\rule{7mm}{2mm}}\\
+1 &   &   CodeReview        & 40 &     1.08 & \textcolor{black}{\rule{10mm}{2mm}}\\
+1 &   &   CodeInspect      & 33 &     0.89 & \textcolor{black}{\rule{8mm}{2mm}}\\
+1 &   &   UnitTest          & 41 &     1.11 & \textcolor{black}{\rule{12mm}{2mm}}\\
+1 &   &   IntTest          & 75 &    2.03 & \textcolor{black}{\rule{20mm}{2mm}}\\
+1 &   &   AcceptTest       & 44 &     1.19 & \textcolor{black}{\rule{12mm}{2mm}}\\
+1 &   &   SysTest           & 40 &     1.08 & \textcolor{black}{\rule{10mm}{2mm}}\\
+\hline\multicolumn{6}{c}{~}  \\
+1 &  Code   &   CodeReview      & 35    & 1.00 & \textcolor{black}{\rule{10mm}{2mm}}\\
+1 &    &   CodeInspect    & 32   & 0.91 & \textcolor{black}{\rule{9mm}{2mm}}\\
+1 &   &   UnitTest      & 45  & 1.29 & \textcolor{black}{\rule{12mm}{2mm}}\\
+1 &    &   IntTest         & 58   & 1.66 & \textcolor{black}{\rule{16mm}{2mm}}\\
+1 &    &   AcceptTest      & 60 &   1.71 & \textcolor{black}{\rule{17mm}{2mm}}\\
+1 &   &   SysTest        & 47 &    1.34 & \textcolor{black}{\rule{13mm}{2mm}}\\
+
+\end{tabular}
+%}
 \end{center}
@@
 
-\input{tex/responses}
-
+ \section*{Response to Reviewers}
+
+{\bf Editor's comments}
+
+Dear authors,
+
+Three knowledgeable reviewers have scrutinized the manuscript
+and found it interesting and relevant,  although in need for
+certain improvements. Therefore, the authors are
+ invited to submit a revised version for a new review, where
+ the reviewers' comments
+are addressed.
+
+\begin{quote} {\em Thank you for the careful reviews. Using
+those comments we have extensively revised and extended this draft.
+Specifically, the old version was 20 pages and this version is 29 pages (and the new
+material all comes from the clarification and correction points offered by the reviewers).
+We hope that  you will find this draft greatly improved.}
+\end{quote}
+
+Particularly, this holds for:
+
+1) the connection to activities of theory building in ESE
+(reviewers 1).
+
+\begin{quote}{\em This new draft now maps our work into the SE literature on theory building. This new material appears in the introduction as well as throughout the document.}\end{quote}
+
+2) clarification about the data origin and
+characteristics (reviewer 1).
+
+\begin{quote}{\em
+This new draft contains much more detail on the origin and characteristics of our data (see \tion{data-collection} and \tion{data_character}).}\end{quote}
+
+3)  use of statistical analysis (reviewers 2 and 3)
+
+\begin{quote}{\em All our results are now augmented with statistical significance tests as well as effect size tests (see \tion{stats} and \tion{171projects}).
+Those tests strengthen the overall message of this paper-- that there is no significant
+phase delay effect.}\end{quote}
+
+- writing style, that would benefit form being more
+precise (reviewer 3)
+
+\begin{quote}{\em We have carefully read those comments and have made numerous changes
+as directed by reviewer3.}\end{quote}
+
+{\bf Reviewer \#1's Comments}
+
+The paper investigates whether a widely
+accepted "rule" in software engineering holds by looking at
+a large set of empirical data. The research topic is highly
+relevant and the results are interesting.   The discussion
+on reassessing old truism fits the scope of the journal very
+well.  The language in the paper is excellent, and the
+structure is (although non-standard) easy to follow.
+
+\begin{quote}{\em Thank you for those kind comments.}\end{quote}
+
+It is not easy for the reader to understand what parts
+of the "life of an issue" the authors claim are affected by
+DIE, and it is also not too easy to understand what data the
+authors have available.
+
+\begin{quote}{\em  The introduction now includes an example of DIE  (see Fig 1).
+Also there are now more precise  definitions on how we measure this effect in \tion{data-collection}}\end{quote}
+
+ The authors do not relate their reassessment of an
+``old truism'' to the growing set of papers on theory building
+in software engineering. With access to such large amounts
+of empirical evidence, it should be possible to take some
+steps toward improved DIE theories.
+
+\begin{quote}{\em We now map our work to prior publications on theory-building in software engineering, as well as theory falsification in general.
+That mapping appears in numerous comments through out the text (using the references offered by this reviewer) with the bulk of the comments being top of p2 and the conclusion.
+
+Notably, we now introduce the DIE as an example of such a theory (in the introduction), and explain that this is done on the basis of it providing a description of a Software development phenomenon accompanied by a rationale that explains why such behavior is expected.
+While we introduce in the conclusions several rationales that can explain why the effect called for by the theory was not observed in our dataset, we feel that future work is needed to collect data in support or contradiction of these explanations before a new theory can be formulated with confidence.
+}\end{quote}
+
+The different projects studied are all grouped together. I
+believe studying clusters of development contexts would be
+highly interesting, but unfortunately the current
+characterization of projects is inadequate.
+
+\begin{quote}{\em  We agree with this comment and so took two actions for this draft-- one that was effective and one that was ineffective.
+
+First, this draft now contains extensive notes on the different contexts of the
+projects studied here. Those notes are offered in \tion{data_character} and \tion{171projects}.
+
+Second (and this was the ineffective step), we  searched the categories   listed in \tion{data_character} and \tion{171projects}
+in order to check if any stratification contained the delayed issue effect. 
+That study   failed to find a stratification
+containing an exponential expansion in the time to fix issues. The reason for this was  the small size
+of those stratifications  exacerbated the large IQR's seen in this data\footnote{Recall that in a sorted list of numbers,
+the inter-quartile range, or IQR, is difference between the 
+  75th to 25th percentile value.}:
+\bi
+\item
+Our 171 projects   stratifies into subsets of varying sizes. Sorting those sizes, we get a list
+comprising of 17,12 then numerous much smaller stratifications.  Reasoning over such small samples
+is problematic in the general case and, in the case of our data, it is ever more problematic due to
+the large IQRS of the data:
+\item
+To see these large IQRs,  please consider the 50th and IQR columns of Figure~\ref{fig:raw} Note that most of our IQRs were larger than tan the 50th percentile; i.e. software
+data exhibits large variances-- where are exacerbated by the smaller samples seen in the stratifications.
+\ei
+Our conclusion from exploring the stratification's is, given the currently available data, we cannot check
+for a DIE effect in subsets of this data.
+
+Note that the above notes are now added to the paper at the end of Section 6.6.
+ }\end{quote}
+
+DETAILED REVIEW: The authors should expand on what they
+include in the ``delayed issue effect'' (DIE), in particular
+in relation to standard milestones in the life of an issue
+and the corresponding timespans.  I'm particularly
+interested in whether all "indirect" issue resolution steps
+are covered, e.g., change impact analysis, updating
+documentation, and regression testing. The effort involved
+in these steps clearly depends on the development context of
+the specific project. The short a-c) list on page 14:50
+suggest that the issues studied are restricted to minor
+fixes, i.e., no updates to architecture, changed hardware,
+recertification, updated user manuals etc.
+
+\begin{quote} {\em
+We now provide a more explicit definition of what constitutes a defect (our study of issues is limited to defects) and the ``defect lifecycle'' phases we measure in \tion{data-collection}.
+These definitions greatly expand on the now-deleted list on 14:50 in teh previous version.
+In TSP, ``a defect is any change to a product, after its construction, that is necessary to make the product correct''.
+This includes both minor issues (typography) and severe issues (architecture changes, requirements errors).
+Time on a defect includes a) time to investigate/analyze a defect once discovered, b) time to craft and implement a fix, and c) time to verify and close the defect.
+
+Regarding the concern that we may only be reporting trivial defects, we have added \fig{dtypes}, which shows the distribution of defect types in our data.
+It is true that around a quarter of the fixes were documentation changes, but the other 75\% are function-altering behavior in the system.
+}\end{quote}
+
+Section 6.3 describes the data, but some aspects
+should be further explained.
+
+\begin{quote} {\em
+At your suggestion, we have added extensive notes on the nature of our data--
+see \tion{data_character} and \tion{171projects}.
+}\end{quote}
+
+Logging interruption time
+sounds very useful, but I wonder how carefully the
+developers actually did this 
+ it must be really hard to keep up the discipline required.
+ If you are interrupted as a developer (e.g., phone calls,
+ urgent mail, someone asks a question) I don't think the
+ first thing you do is to stop a timer on your computer.
+ Moreover, developers often work on several defects in
+ parallel, and might interweave bug fixing with new
+ development. I don't think the "interruption time" captures
+ all such multi-tasking, and it should at least be properly
+ discussed in Section 7 "Threats to validity".
+
+\begin{quote}{\em 
+Yes, there is undoubtedly some accuracy in the time-tracking.
+We provide some evidence on the rigor of the TSP data collection process in \tion{defin}, \tion{data_integrity}, and \tion{data_character}, and a more extensive discussion on the issue of time-tracking and defect logging accuracy in \tion{conclusion_validity} on Conclusion Validity.
+We note that the data being collected here was not collected
+for the purposes of this paper. Rather, it was being collected and used by the
+programmers to assist them in their day-to-day management of the project.
+Given this, we have confidence that the logging information is not widely inaccurate.
+
+Please also consider the sanity checks reported in a companion paper (referenced here \tion{data_integrity}). In summary, this data has been heavily vetted.
+}
+\end{quote}
+
+The discussion on reassessing old truisms (Section 3) is
+interesting, but it should be complemented by the
+perspective of theory building in software engineering - An
+active research topic lately, with a dedicated workshop
+series (GTSE). I suggest looking into the following
+references for a start: Sj{\o}berg et al. (2008) ``Building
+Theories in Software Engineering'', Smolander and Paivarinta
+(2013) ``Theorizing about software development practices'',
+and Stol and Fitzgerald (2015) "Theory-oriented software
+engineering". Considering new empirical evidence is
+obviously critical to theory building, and discussing your
+new results in the light of an theory creation would be
+valuable. The authors mention that DIE appears to occur
+intermittently in certain kinds of projects - maybe the
+authors could elaborate on this idea and present an improved
+DIE theory based on what they now know? I believe the
+authors have the best available data to do so, and I would
+expect the paper to go beyond simply questioning the "old
+truth".
+
+ \begin{quote}{\em 
+We thank the reviewer for this important observation and have connected our work to the literature on theory-building.
+However, our data supports a number of possible explanations for the observed intermittent appearance of the DIE.
+Since the current data is not suitable for choosing among the various hypothesized explanations we feel that formulating a new theory build from this data
+would be premature at this time. That said, we think the current paper would be useful
+as a data point which, when combined with other sources, could be useful for some new future theory of SE.
+}\end{quote}
+
+The authors study three claims in this paper, and the
+third claim is the central one: "delayed issues are not
+harder to resolve". To study whether issues require longer
+resolution times in later phases, the authors analyze a
+large set of issue data from historical projects. Thus the
+manuscript reports from an observational study rather than
+an experiment with a controlled delivery of treatments.
+While I believe the authors' approach is practical, I would
+like to see a critical discussion on threats to validity of
+observational studies (e.g., Madigan et al., A Systematic
+Statistical Approach to Evaluating Evidence from
+Observational Studies, Annu. Rev. Stat. Appl. 2014. 1:11-39
+and Carlson and Morison, Study Design, Precision, and
+Validity in Observational Studies, J Palliat Med. 2009 Jan;
+12(1): 77-82). An alternative study design (although
+difficult to realize in a system of industrial size) would
+be to let different developers resolve the same issues for
+the same software system, i.e., one group resolves an issue
+during design, another during implementation, and a third
+during testing - some discussion along these lines would
+strengthen the validity section.
+
+\begin{quote}{\em 
+ We have reviewed and cited the mentioned papers. We have incorporated many of the concerns raised in these papers into the expanded \tion{external_validity} on External Validity.
+ There, we note that Sj{\o}berg et al.
+list three classes of threats to External Validity (the smallest number) and we address all three of them.
+In truth, the only way to address External Validity threats is to have a huge sample, broadly representative of your domain, over an extended period of time-- e,g.
+using the 171 projects from 2006 to 2014 studied in this paper.
+The issues external validity in observational studies raised in \cite{madigan2014} and \cite{carlson2009} are more applicable to studies that compare a treatment and a control, whereas we are observing an effect that leads us to question the theory that DIE is a truism.
+}\end{quote}
+
+The first and second claims are studied with much less
+rigor. "DIE is a commonly held belief" is studied using a
+survey of software engineers, both practitioners and
+researchers. According to Fig. 2 (actually a table) [NOTE: Fig. 2 is now \fig{survey_results}], the
+number of respondents is 16 and 30 for practitioners and
+researchers, respectively. Sixteen respondents from industry
+represent a tiny survey of a very general claim that any
+software engineer could respond to. Why were not more
+answers collected? The authors do not have much evidence to
+defend the first claim, and there is no discussion of the
+corresponding validity threats. The second claim, "DIE is
+poorly documented", is studied using a literature review.
+Unfortunately, the method underlying the literature review
+is not presented. Although it doesn't need to be an SLR of
+the most rigorous kind, the authors should report how the
+papers were identified. I suspect the terminology used to
+describe the DIE phenomenon is highly diverse, thus it would
+strengthen the paper if the authors reported how they
+reviewed the literature. According to Page 19:34 [NOTE: the enumerated list in the Conclusion] only eight
+publications were identified.
+
+\begin{quote}{\em We agree with the reviewer - the phrasing of our claims and the supporting evidence were not congruent.
+Our intention is to argue from multiple angles that the evidence for DIE is sparse and, more importantly, that DIE persists as a popular belief.
+Our goal here is to motivate that we should re-investigate the DIE.
+We have toned down the language and combined our arguments on ``widely-believed'' and ``poorly-documented'' into a new \tion{belief} whose focus is to motivate our study.
+We have added more details on our literature search and snowball sampling.
+We have also highlighted how blog articles often cite secondary sources, which in turn cite Software Engineering Economics, to illustrate how the DIE pehnemonon has perpetuated.}\end{quote}
+
+The iterative fashion of modern software development,
+with agile at the extreme end, is not fully discussed (the
+short discussion section could be extended). The phases of
+the linear development of the 80s, (such as in Fig. 1) [NOTE: Fig. 1 remains \fig{b81}] 
+probably don't exist in many of the projects in the TSP
+dataset, still the authors discuss the DIE-effect from the
+perspective of the 80s. Page 16:31 states "DIE has been used
+to justify many changes to early lifecycle software
+engineering" - does this mean the agile movement
+successfully mitigated DIE? This possibility is not fully
+considered in the paper.
+How many of the 171 projects were
+(more or less) agile? This appears to be an important
+characteristic of the included projects - very important to
+describe!
+
+\begin{quote}{\em We provide more information on the types of projects included in our sample in \tion{data_character}
+
+More to the point -- does agile help mitigate DIE?
+The answer is - we don't know.
+A strong argument can be made that all iterative \& incremental methodologies were designed to mitigate the DIE.
+Further, huge advances in software engineering technology and process have occurred since DIE was first observed, many of which have precisely targeted the DIE (static analysis, test-first development, automated build).
+Perhaps our analysis provides evidence that DIE can be defeated, at a large scale across multiple varied projects, using modern technologies.
+We do not yet know the causes here, but certainly it appears the DIE should not be treated as a truism any more, but rather a variable that can be controlled.
+We have added more discussion on this to \tion{discussion}.
+}\end{quote}
+
+Concerning characterization of the 171 projects, the
+paper needs to report much more detail. I would expect to
+see some descriptive statistics.   The paper does not report much
+characterization of the 171 projects. I strongly suggest the
+authors to dig deeper into the data, and analyzing for which
+types of projects the findings hold. What patterns are there
+to discover? I suggest Petersen and Wohlin, "Context in
+industrial software engineering research" (ESEM2009) for
+details on how to characterize development contexts.
+
+\begin{quote}{\em Much more context is now offered in \tion{data_character} and \tion{171projects}}. \end{quote}
+
+Fig 1 [NOTE: Still \fig{b81}]: "Operation" dwarfs everything in Boehm's diagram.
+Since you do not study anything post release in this paper,
+I think this figure skews the reader's mental picture. If
+you remove the rightmost bar and rescale accordingly, the
+plot better matches the findings you report in Fig. 10 [NOTE: now \fig{raw}]. You
+still identify an interesting result though, but the
+presentation turns fairer.
+
+\begin{quote}{\em We understand why you are making that suggestion.
+That figure is something of a classic. Therefore we opted not to chop the figure but hope the text is sufficiently clear}.
+\end{quote}
+
+MINOR COMMENTS: 
+Several figures are copied from previous work. Are all copyrights properly
+managed?
+
+\begin{quote}{\em We have removed all copyrighted images and replaced them with original drawings based on information from the cited source.}\end{quote}
+
+\fig{raw}: Black and red is not a good pair for
+grayscale printouts. Please use black and gray instead.
+\begin{quote}{\em  We agree: switched to black and gray.  }\end{quote}
+
+Some figures should be resized to better match the page width.
+
+\begin{quote}{\em  We agree: fixed. }\end{quote}
+
+Page 3:14 - "The above argument" Which argument? Could be precisely specified.
+
+\begin{quote}{\em  We agree: that statement has been deleted. }\end{quote}
+
+Page 3:22 - Please provide the full link to the dataset.
+
+\begin{quote}{\em  Our policy is to get one publication out of paper
+before issuing the data in public.
+If this paper is accepted, we will certainly add that link.
+Note that we have a long history of living up to that policy-- see http://openscience.us/repo.
+}\end{quote}
+
+Page 18:26 - First sentence: "Unexpected results such as this one". Ambiguous reference,
+please be specific.  
+\begin{quote}{\em  We agree. That back reference is removed. }\end{quote}
+
+Page 18:27 - "We also survey the state of SW dev. practice /---/": given the size of the survey,
+this statement feels a bit bold.
+
+\begin{quote}{\em We have removed this statement.}\end{quote}
+
+TYPOS ETC.
+
+Sec 1, §1: [3] and [30] have been swapped?
+
+Page 14:46 - Spell out IV\&V the first (and only) time it appears.
+
+Page 3:24 - missing verb (is) Page 16:43 - "Did this study failed"
+
+Page 18:44 - Appache
+
+Page 19:19 - Al Ref 26 - oo --> OO
+
+\begin{quote}{\em  All of these have been corrected. }\end{quote}
+
+{\bf Reviewer \#2's comments}
+
+The topic of this paper is the empirical
+evaluation of a largely admitted claim in software
+engineering, i.e. the exponential cost of correcting errors
+according to the phase in which the errors are discovered.
+The claim is first confirmed by surveying and interviewing
+practitioners and experts. Next, the authors use a large set
+of data from projects in SEI database. The results show
+that, strictly speaking, the claim does not hold (although
+some effects of delayed corrections can be noticed).
+
+Globally, the paper is well written and the authors develop
+a convincing demonstration. The topic of the paper is highly
+relevant for both practitioners and people from academia.
+Among many, I personally believed in this claim and had
+regularly taught it in my courses. Thus, reading this paper
+offers a refreshing perspective on our understandings of
+software engineering background and theoretical knowledge. I
+also like the suggested posture that insists on the
+necessary skepticism that we should have against such
+prevailing claims.
+
+Beyond this positive global impression, I have some concerns
+about the study reporting and analysis: -  It seems to me
+that more descriptive statistics about the sample would
+contribute to a better understanding of the scope of the
+study.
+
+\begin{quote}{\em 
+More description about the context and sample have been added in \tion{data_character} and \tion{171projects}.}\end{quote}
+
+I think in particular about the size (lines of codes and/or
+ number of software components); a duration histogram would
+ also be relevant. Accordingly, the formula for calculating
+ variable "Total effort" in Fig. 7 [NOTE: now \fig{dist}] could be explained, and
+ Figure 9 [NOTE: now \fig{fix-phase-dist}] made more readable (bigger size).
+
+ \begin{quote}{\em We have added more description of the projects in \tion{data_character}.
+We have also added more precise definitions of our measures in \tion{data-collection}.
+We have adjusted the size of figures throughout the paper to be more readable.
+We have added more text
+ to explain {\em effort} in \fig{dist}.}\end{quote}
+
+If certain attributes for the issues and errors are
+available in the data (e.g. severity, priority, etc.), they
+should also be brought to the readers knowledge.
+
+\begin{quote}{\em 
+Those notes are offered in \tion{data_character} and \tion{171projects}.}\end{quote}
+
+Figure 10 [NOTE: now \fig{raw}] is central to the paper's demonstration; its
+expressiveness could be enhanced: the formula for calculating the 50th
+and 95 ratio could be provided,  the unit of column
+"Percentile" could be mentioned.
+
+\begin{quote}{\em
+\fig{raw}'s caption (formerly Figure 10) now has extensive explanation text. e.g. units
+are added on line two. Also, an example of the calculations of the ratios is offered
+in the last line of the caption. Further,
+to simplify comprehension of this diagram, it has been split into two
+(and the more arcane discussion about the outliers, statistics gathered, and methods used is provided is also provided).}\end{quote}
+
+I like the discussion section, in particular, the idea
+of software architectures that could be a contributing
+factor for enhancing software evolution and reducing the
+cost of issues and errors fixing. Is there any available
+knowledge in the data set concerning any architectural
+design choices made in each project?
+
+\begin{quote}{\em
+Not explicitly, but \fig{dtypes} lists ``interfaces'' as a the fourth most
+common type of defect. Note that ``interfaces'' are changed by architectural
+issues. }\end{quote}
+
+In the same trend of ideas, I think what has
+fundamentally changed since early days of SE is the
+development of requirements engineering. In actual SE
+practices, problem and the solution spaces are
+systematically explored thanks to RE techniques; together
+with architecture, this could also explain why things have
+changed since the 70'. It could be, for example, that people
+tend to make less severe errors in early software project
+phases (thanks to RE techniques); a similar phenomenon was
+observed when SE practices become more mature (see Harter et
+al. 2012).
+
+References Harter D. E., Kemerer C. F., Slaughter S. (2012).
+Does Software Process Improvement Reduce the Severity of
+Defects? A Longitudinal Field Study. IEEE Transactions on
+Software Engineering, vol. 38, n° 4, p. 810-827.
+```
+
+\begin{quote}{\em
+  Absolutely - there are potentially many explanations for why the DIE was NOT observed in our dataset.
+We add more discussion on the potential causes in \tion{discussion}, supported by the reference to Harter et al.
+ }\end{quote}
+
+In the conclusion section, I feel uncomfortable with
+the assertion "That data held no trace of the delayed issued
+effect" (line 17). As mentioned elsewhere in the paper, the
+delayed issue is not absent; it is much less significant and
+systematic than what is usually claimed. Moreover, this
+reduction has been demonstrated for medium sized projects;
+we cannot generalize to larger projects.  
+\begin{quote}{\em
+  We agree -That line has been removed.
+ }\end{quote}
+
+Minor remarks:
+
+- p.9, line 29 : " … reported by Shull [52], found that the cost to find certain non-critical classes of defects ..." => is it the "cost to find" or the "cost to FIX"?
+
+- p.10, line 21: duplicate word "there" -
+
+p.12, line19: missing word "One of THE guiding …"
+
+- p.16, line 33: "distinguish" instead of "distinguished"?
+
+- p.18, line 42: explain acronym MEAN
+
+- p.20, line 20: useless enumeration A1?
+
+\begin{quote}{\em
+ All of these have been corrected or removed.
+ }\end{quote}
+
+{\bf  Reviewer \#3's Comments}:
+
+I like the idea of the paper and I certainly
+would like to see more empirical studies that periodically
+check if our beliefs about software engineering practices
+still hold (or even if they have ever held). The authors
+address one of the beliefs that are more entrenched in the
+population of researchers and practitioners, and a very
+important one too.
+
+\begin{quote}{\em
+Thanks for those kinds words.
+ }\end{quote}
+
+At the same time, I think that the paper needs to be
+improved before it can be published.
+
+The empirical analysis (Section 6 [NOTE: now \tion{analysis}) is a bit of a let down. I
+would like to see some sort of more robust and rigorous
+statistical analysis, based on statistical tests.
+
+\begin{quote}{\em
+We agree. \tion{stats} lists our stats methods and \fig{raw}
+shows the results (see the left-hand column: all those ``1''s mean
+that our stats could not find any difference between the fix times for
+delaying or not delaying.
+ }\end{quote}
+
+The paper seems to be written in a somewhat casual style,
+which makes the paper more pleasant to read, but less clear
+in some parts (see Detail Comments below).
+
+\begin{quote}{\em
+Thanks for those pointers. We have fixed the issue.
+ }\end{quote}
+
+DETAIL COMMENTS
+
+Section 2 [NOTE: still \tion{claims}]
+
+I'm not really sure what the authors mean by "We say that a
+measure collected in phase 1, ., i, .. j is very much more
+when that measure at phase j is larger than the sum of that
+measure in earlier phases 1 <= i < j." You are defining a
+property of a measure, but, the way this definition is
+written, it seems as if you are defining the property of a
+measure of "being much more," without any further
+qualifications.  I can guess that you mean that a measure
+$m$ (collected in phases, so you can denote by $m(i)$ the
+value of $m$ in phase $i$) has this property *in phase $j$*
+(so, it's a property m has in a specific phase and not a
+general property of the measure) if $\sum_{1 <= i < i} m(i)
+< m(j)$. 
+\begin{quote}{\em
+Thanks for that. We added the $m_j$ notation to \tion{claims}, as you suggested.
+ }\end{quote}
+
+Your definition of "difficult issue" is "Issues are more
+difficult when their resolution takes more time or costs
+more (e.g. needs expensive debugging tools or the skills of
+expensive developers)." That's not a very precise
+definition, though, because it has two different
+interpretations, one in terms of time and the other in terms
+of effort. The reasons why you introduce this definition
+become clearer only in Section 7.2 [NOTE: now \tion{construct}], in the Threats to
+Validity, and that's too late. You should move some of the
+discussion of Section 7.2 here.
+\begin{quote}
+{\em Good idea! We did just that- see middle para in \tion{claims}.}
+\end{quote}
+Why would the term "delayed issue effect" be a
+generalization of the rule "requirements errors are hardest
+to fix"? "Delayed" seems to be quite specific as it appears
+to refer to time, while "hardest to fix" may refer to other
+variables, like effort or cost.
+\begin{quote}
+{\em Not sure if this a reviewer requirement for something to be fixed
+prior to publication, or a clarification question. We used these
+terms since when we ran the survey, the audience clearly made
+that generalization. Does this reply adequately to your comment?}
+\end{quote}
+In Claim 3, "very much more harder" sounds like a
+little bit too much ... ("more harder"?)
+\begin{quote}
+{\em Our bad. We fixed that.}
+\end{quote}
+Section 4
+
+A very minor issue: why is Figure 2 [NOTE: now \fig{survey_results}] a ... figure, instead of
+a table? Same for all other tables ...
+\begin{quote}
+{\em Authors' preference. When a document has many tables and figures, navigating
+between them is hard. Calling them all "figures" makes navigation faster. }
+\end{quote}
+Section 5.1
+
+This is not a complete sentence "All the literature
+described above that reports the onset of DIE prior to delivery.''
+
+\begin{quote}
+{\em Fixed}
+\end{quote}
+
+Section 6.3
+
+You should rephrase your definition "a defect is any change," since a defect is what existed before the change was made and not the change itself.
+\begin{quote}
+{\em We now provide a more precise definition of defect in \tion{defin}.
+}
+\end{quote} 
+You do not explain that "QualTest" in Fig. 8 [NOTE: now \fig{waterfall}] is.
+\begin{quote}
+{\em A mistake on our part that has been removed from the figure.}
+\end{quote}
+The definition of "time per defect - The total \# of defects found in a plan item during a removal phase divided by the total time spent on that plan item in that phase." is not correct as it is, as this would basically be the number of defects per unit time, instead.
+The roles of time and
+defects should be reversed.
+\begin{quote}{\em We now provide a more precise definition of time measurement related to defects in \tion{data-collection}}\end{quote}
+
+Besides adding a more rigorous statistical analysis, you
+need improve your explanations.
+\begin{quote}{\em We agree. \tion{stats} lists our stats methods and \fig{raw}
+shows the results (see the left-hand column: all those ``1''s mean
+that our stats could not find any difference between the fix times for
+delaying or not delaying.}\end{quote}
+
+Figure 9 [NOTE: now \fig{fix-phase-dist}]. The caption says "Distribution of defects found
+and fixed by phase." Is that the distribution of defects
+found in some phase and fixed in that phase? The text says
+"The distribution of defects found and fixed per phase in
+our data is shown in Figure 9. A high percentage of defects
+(44) were found and fixed in the early phases, i.e.,
+requirements, high level design, and design reviews and
+inspections." which doesn't add much. If that's the
+distribution of defects introduced in a phase and fixed
+immediately in that phase, it is not clear why you introduce
+it, though.
+\begin{quote}
+{\em 
+The TSP convention is to fix a defect in the phase in which it was identified, thus "find and fix".
+With some exceptions, defects are not counted as defects unless they escape that phase.
+The exception case is the secondary defect introduced during a fix that occurs with some frequency in test and rarely in inspection or review.
+We have clarified this distinction in the text.}
+\end{quote}
+Figure 10 [NOTE: now \fig{raw}]. The explanation of the data and histograms in the
+figure is, at best, confusing. The caption says "50th and
+95th percentiles of issue resolution times" and the opening
+sentence of Section 6.4 says "Figure 10 shows the 50th and
+95th percentile of the time spent resolving issues ..."
+However, these are not "times," because you write "expressed
+as ratios of resolution time in the phase where they were
+first injected" a couple of paragraphs below. This seems to
+be the right interpretation, but right after that you write
+"The BLACK and RED bars show the increases for the 50th
+(median) and 95th percentile values, respectively," which
+would be a third interpretation of the data in Figure 10.
+\begin{quote}
+{\em We agree-that figure was far too confusing. We have now split the 50th and 90th peercentile results
+into two charts, which simplifies both. Please see \fig{raw} and \fig{last}.
+Also, please note that that fit\fig{raw} now has an extensive help text in the caption.
+We used your comments there where we noted the ratios used in the
+bard charts are unitless while the left-hand-side column numbers are
+times in minutes.
+}
+\end{quote}
+
+You need to at least mention (or, better, discuss) some of
+your results, because they appear to be "counterintuitive,"
+since they show that, for example, median resolution times
+can even decrease if issues are not fixed immediately, but
+in later phases. For example, looking at the Reqts section,
+most percentages are way below 1, which seems to indicate
+that some issues are actually much easier to fix at later
+stages (maybe because more or better information about the
+software system is available only later). That would be a
+very interesting result by itself, especially if you can
+provide some statistical support for it.
+\begin{quote}{\em
+The last paragraph of \tion{171projects} now comments
+on that effect. }
+\end{quote}
+
+Maybe you could analyze the projects developed in a
+"traditional" way and check if there is some sort of DIE
+there. For example, if you look at the projects developed
+with the waterfall life cycle, maybe you will find larger
+DIE than for the other life cycles, which might partially
+justify the claims of previous researchers about the
+existence of DIE. This would also show that newer kinds of
+life cycles (like Agile ones) help solve DIE. You do provide
+hints about this in Section 8 [NOTE: now \tion{discussion}] ("We also note that other
+development practices have changed in ways that could
+mitigate the delayed issued effect") and Section 9 [NOTE: now \tion{conclusion}], but
+that's clearly not enough.
+\begin{quote}{\em
+
+If your comment is along the lines of ``look for stratifications to check if
+DIE exists in subsets'' then we offer the following notes (which we added to the end of \tion{171projects}):
+
+We looked into   stratifications of the data and, so far,   failed to find a stratification
+containing an exponential expansion in the time to fix issues. The reason for this was  the small size
+of those stratifications  exacerbated the large IQR's seen in this data\footnote{Recall that in a sorted list of numbers,
+the inter-quartile range, or IQR, is difference between the 
+  75th to 25th percentile value.}:
+\bi
+\item
+Our 171 projects   stratifies into subsets of varying sizes. Sorting those sizes, we get a list
+comprising of 17,12 then numerous much smaller stratifications.  Reasoning over such small samples
+is problematic in the general case and, in the case of our data, it is ever more problematic due to
+the large IQRS of the data:
+\item
+To see these large IQRs,  please consider the 50th and IQR columns of Figure~\ref{fig:raw} Note that most of our IQRs were larger than tan the 50th percentile; i.e. software
+data exhibits large variances-- where are exacerbated by the smaller samples seen in the stratifications.
+\ei
+Our conclusion from exploring the stratifications is, given the currently available data, we cannot check
+for a DIE effect in subsets of this data.
+}
+\end{quote}
 \end{document}

### 2016-09-12 lucas.tex
Updates from ShareLaTeX

@@
 Our survey collected data on software engineers' views of commonly held software engineering ``laws''.
-One of the laws questioned is the a specific form of our delayed issue effect:  ``requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' (from Glass~\cite{glass02} p.71 who references Boehm \& Basili~\cite{boehm01}).
-We abbreviate this law as RqtsErr
-(and is
-illustrated in \fig{b81}).
+One of the laws questioned is a specific form of our delayed issue effect:  ``requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' (from Glass~\cite{glass02} p.71 who references Boehm \& Basili~\cite{boehm01}).
+We abbreviate this law (illustrated in \fig{b81}) as RqtsErr.
 
-(Technical aside: we    ask them  RqtsErr  since this issue typically needs no supportive explanatory
-text. On the other hand, if we had asked them about our more general term ``delayed issue
+(Technical aside: we use the RqtsErr formulation since this issue typically needs no supportive explanatory
+text. If we had asked respondents about our more general term ``delayed issue
 effect'', we would have had to burden our respondents with extra explanations).
 
-The survey was conducted in two phases using Amazon's Mechanical Turk. The first phase was conducted only with professional software engineers solicited through the Mechanical Turk,
-Professional software engineers were required to complete a pretest to verify their status as a professional or open source software developer and to confirm their knowledge of basic software engineering terminology and technology.
-The  second survey was conducted Program Committee members of the ESEC/FSE 2015 and ICSE 2014  conferences solicited via email.
+The survey was conducted in two phases using Amazon's Mechanical Turk. The first phase was conducted only with professional software engineers solicited through the Mechanical Turk;
+participants were required to complete a pretest to verify their status as a professional or open source software developer and to confirm their knowledge of basic software engineering terminology and technology.
+The  second survey was conducted with Program Committee members of the ESEC/FSE 2015 and ICSE 2014  conferences solicited via email.
 
@@
 only added them in as a way to calibrate responses to the RqtsErr question. All laws were drawn from \cite{glass02} and \cite{endres03}. 
-The PC member survey contained an additional question on ``In general, the longer errors are in the system (requirements errors, design errors, coding errors, etc.), the more expensive they are to fix'' (the Delayed Issue Effect, for short).
+The PC member survey contained an additional question on ``In general, the longer errors are in the system (requirements errors, design errors, coding errors, etc.), the more expensive they are to fix'' (the Delayed Issue Effect).
 Responses were recorded using the agreement question Likert scale.
@@
 
-Both  practitioners and researchers  strongly believed in RqtsErr (in both sets of responses, RqtsErr, received  scores higher than most
+Both  practitioners and researchers  strongly believed in RqtsErr: In both sets of responses, RqtsErr received  scores higher than most
 other laws. Further, in the case of practitioners, this law was rated
@@
 Further, any changes to the production %ystem have additional complexity (compared to changes while system in %evelopment) due to users, data existence, and architecture dependencies.
-Caveat: in the free reponse texts, we note that the researchers who disagreed with the law generally asserted that requirements change can be expensive, but that depends on the process used (e.g., agile vs.
+Caveat: in the free reponse texts, we note that the researchers who disagreed with the law generally asserted that requirements change can be expensive, but that the effect depends on the process used (e.g., agile vs.
 waterfall) and the adaptability of the system architecture.

### 2016-09-12 phaseDelay4Emse.tex
Updates from ShareLaTeX

@@
 Other researchers
-are just as adamant in asserting that that the delayed issue effect is a generally useful law of software engineering.
+are just as adamant in asserting that the delayed issue effect is a generally useful law of software engineering.
 For example, what we call the delayed issued effect was listed at \#1 by Boehm and Basili in their ``Top 10 list'' of ``objective and quantitative data, relationships,
@@
 delayed issue effect is both very sparse and very old.
+As  shown in this section,
+a goal of agile methods 
+is to reduce the difficulty associated with making changes later in the lifecycle~\cite{beck00}. Yet, as shown below,
+relatively little empirical data exists on this point.
 
@@
 Workshop participants from Toshiba and IBM reported cost-to-fix ratios between early lifecycle and post-delivery defects of 1:137 and 1:117 for large projects respectively~\cite{Shull02} -- but the raw data points were not provided and thus cannot be confirmed.
-A goal of agile methods
-is to reduce the difficulty associated with making changes later in the lifecycle~\cite{beck00}. Relatively little empirical data exists on this point.
 Elssamadisy and Schalliol~\cite{Elssamadisy02} offer an anecdotal report on the growing, high cost of rework in a 50 person, three-year, 500KLOC Extreme Programming project as the project grew in size and complexity-- but again we cannot access their
@@
 
-Some studies that report less-than large
+Some studies report less-than large
  increase in the effort required to fix delayed issues.  Boehm~\cite{Boehm80} provides data suggesting that the cost-to-fix curve for small projects
@@
 One of the authors (Nichols) has mentored software development teams and coaches around the world as they deploy TSP within their organizations since 2006.
-The  most recent completions were in 2014.
+The  most recent completions were in 2015.
 
@@
 In return the SEI agrees that  data must not be traceable to its source.
-The data is collected at major project events: launch, interim checkpoints, and at project completion.
+The data are collected at major project events: launch, interim checkpoints, and at project completion.
 The data from these TSP projects were collected and stored in the Software Engineering Measured Process Repository (SEMPR) at the SEI.
@@
 In our TSP data,  the relative frequencies of these defect types are shown in \fig{dtypes}. Around a quarter of the fixes were simple documentation changes.
-That said, 75\% of the changes are quote  elaborate;   e.g. fixes to function necessitates a careful reflection of the purpose of the code.
+That said, 75\% of the changes are quite elaborate;   e.g. fixes to function necessitates a careful reflection of the purpose of the code.
 
@@
 
-One or more defects are reported against a single plan item in the time tracking logs, e.g., a review session, an inspection meeting, a test execution.
-The time-tracking logs against the plan items for a defect includes the time required to (a)~collect data and realize there is an error,
+In the TSP, defect data includes the affected artifact, the estimated developer fix effort (find and fix), the lifecycle phases in which the defect was injected and removed, and the developer who implemented the fix.
+In the database, the task is associated with a plan item.
+Defects (one or more) are recorded in the defect log and associated with the plan item (task) in the time tracking logs.
+For example, a review session, an inspection meeting, or a test would be plan items associated with some product component.
+When defects are found and fixed, the time recorded in the time-tracking logs against the plan items includes the direct effort time (stop watch rather than wall clock time) required to (a)~collect data and realize there is an error,
 (b)~prepare a fix,  and (c)~apply some validation
-procedure to check the fix (e.g. discuss it with a colleague or execute some tests).
+procedure to check the fix (e.g.
+discuss it with a colleague or execute some tests).
+Although we have explicit estimates of "find and fix" effort for each defect, this fails to account for the full costs (e.g.
+meeting time or test execution).
+Because the vast majority of defects are removed in explicit removal phases, we chose to estimate defect cost using the entire time in removal phases divided by the number of defects.
+We recognize that this approach can exaggerate cost per defect for cases with few defects and large overhead effort, such large test suites or slow running tests that require continuous developer attention.
+Nonetheless, this approach provides a better comparison between early removals from inspections later removals from test.
+The result will be a time per defect that is greater than the directly measured "find and fix" time, but smaller than the wall clock or calendar time.
+
 Since multiple defects can be recorded against a plan item, the time-to-fix a defect is defined as:
@@
 %\todo{should we add a citation to Petersen and a few more words on context.}
-In this section we provide some descriptive statistics and discuss the of the projects from which this data was drawn and summarize some additional contextual information.
+In this section we provide some descriptive statistics, discuss the projects from which this data was drawn, summarize some additional contextual information.
 The project contexts describe the conditions under which these measures were obtained, help determine relevance of the results, and may guide future data analysis with segmentation.
@@
 While the development data was recorded in tools and submitted in a structured form, context data was collected in less structured project questionnaires, site questionnaires, team member surveys, launch presentations and reports, post mortem presentations and reports.
-This data has not yet been mined from the submissions 1) because of the cost and effort required, 2) we are obligated to avoid providing any data that can identify projects (that is, the data must remain anonymous), and 3) the unstructured data is not always  complete  submitted.
+This data has not yet been mined from the submissions 1) because of the cost and effort required, 2) we are obligated to avoid providing any data that can identify projects (that is, the data must remain anonymous), and 3) the unstructured data may not be complete when submitted.
 Gathering more projects will make it easier to anonymize the data and overcome missing data problems.
@@
 
-While the processes and work flows among these projects can vary, the logical order is described section \ref{development_phase} is followed.
+While the processes and work flows among these projects can vary, the logical order described in section \ref{development_phase} is followed.
 Development tasks such as requirements development, design, or code, are typically followed by an appraisal phase such as personal review or inspection.
@@
 \centering
+\renewcommand{\arraystretch}{1}%
 \begin{tabular}{lrrrrrrrl}
                                                                    & N   & Min & Q1   & Median & Q3    & Max   & Mean   &  Distribution \\\hline
-Team size                                                          & 171 & 1   & 4    & 6      & 10    & 36    &  7.8    &  \includegraphics[width=15mm]{team_size.png} \\ 
-Duration {[}days{]}                                                & 171 & 7   & 33   & 61     & 118   & 1918  & 107    & \includegraphics[width=15mm]{duration.png} \\
-Added \& Modified LOC                                                 & 117 & 2   & 1125 & 4201   & 13092 & 88394 & 10259  & \includegraphics[width=15mm]{LOC.png} \\
-Components                                                         & 171 & 8   & 26   & 49     & 107   & 4170  & 116.2  & \includegraphics[width=15mm]{components.png} \\
-Year                                                               & 171 &     & 2011 & 2012   & 2013  & 2014  & 2011.9 & \includegraphics[width=15mm]{year.png}
+Team size                                                          & 171 & 1   & 4    & 6      & 10    & 36    &  7.8    &  \includegraphics[width=25mm]{team_size.png} \\ 
+Duration {[}days{]}                                                & 171 & 7   & 33   & 61     & 118   & 1918  & 107    & \includegraphics[width=25mm]{duration.png} \\
+Added \& Modified LOC                                                 & 117 & 2   & 1125 & 4201   & 13092 & 88394 & 10259  & \includegraphics[width=25mm]{LOC.png} \\
+Components                                                         & 171 & 8   & 26   & 49     & 107   & 4170  & 116.2  & \includegraphics[width=25mm]{components.png} \\
+Project Initiation Year                                                       & 171 &    2006 & 2011 & 2012   & 2013  & 2014  & 2011.9 & \includegraphics[width=25mm]{year.png}
 \end{tabular}
@@
  Of the 171 projects in the sample, only 117 collected size data in lines of code, however all projects tracked effort and  the component counts with applied effort are provided.
-Other data is complete for all 171 projects.
+Other data are complete for all 171 projects.
 The projects were mostly of short duration and small to medium size.
-The median project began in 2012 lasted 61 days, produced 4200 Lines of Code, 49 components (modules or features).
-Duration ranged from 7 to 1918 days.
-Size ranged from minimal (this may represent a short maintenance project) to 88394.
+The median project began in 2012 lasted 61 days, produced 4,200 Lines of Code, 49 components (modules or features).
+Duration ranged from 7 to 1,918 days.
+Size ranged from minimal (this may represent a short maintenance project) to 88,394.
 The earliest project was in 2006 and the most recent in 2014.
@@
     \item all teams were self managed, defining work flows, practices, and schedules
-    \item teams meet at least weekly to evaluate progress and replan
-    \item most teams were small with a median size of 6 and a mean of 7.8. only 25\% of the teams were larger than 10 with a long tail on the distribution
+    \item teams met at least weekly to evaluate progress and re-plan
+    \item most teams were small with a median size of 6 and a mean of 7.8; only 25\% of the teams were larger than 10 with a long tail on the distribution
     \item the median project lasted only 60 days, suggesting limited scope for each integration
@@
 
-    The Scott-Knott ranker 
+    Scott-Knott allows for a direct answer to the following questions: given an issue
+    raised at phase $i$ and resolved at phase $\forall j,k \in \{i,i+1,i+2,...\}$, is it
+    true that that the time to resolve issues at $j$ is significantly different 
+    to the time to resolve issues at $k$? 
+    \be
+    \item Note that if $j,k$ times are significantly different, then we can compare
+    the median values to say (e.g.) resolution time at phase $k$ is 3 times slower than phase $j$.
+    \item Note also that if all times $j,k$ are {\em not} significantly different
+    then we say that the phases all {\em rank} the same (and we denote this by setting
+    all such {\em ranks} to 1).
+    \ee
+    In the following results, we nearly always encountered the second case;
+    i.e.   the times to resolve issues at different times were usually {\em not significantly}
+    different.
+
+    As to technical details of the Scott-Knott methods, this ranker 
       was   recommended  by  Mittas and Angelis in a
@@
     halves.
-    Before Scott-Knot recurses,  however,  it apples some  statistical hypothesis test $H$ to check
+    Before Scott-Knot recurses,  however,  it applies some  statistical hypothesis test $H$ to check
     if $m,n$ are significantly different.
@@
     To operationalize ``interesting'', Scott-Knott seeks the division of 
-    of $l$ treatments into subsets of size $m,n$ of sizes $ls,ms,ns$ and median values
+    $l$ treatments into subsets of size $m,n$ of sizes $ls,ms,ns$ and median values
     $l.\mu, m.\mu, n.\mu$ (respectively) in order to maximize
@@
 
-   To operationalize $H$, we use both  bootstrap sampling  and the A12 effect size test.
+   To operationalize $H$, we use both  bootstrap sampling  and Vargha and Delaney's A12 test A12 effect size test.
    In other 
@@
     small. Hence, to assess the performance differences we first must rule out 
-    small effects using A12, a test   recently endorsed by Arcuri and 
+    small effects using Vargha and Delaney's A12 test, a test   recently endorsed by Arcuri and 
     Briand at ICSE'11~\cite{arcuri11}.
@@
 Next, for each division, we separated all the issues that were removed at different subsequent issues $P_r \in \{P_1,P_2,..\}$.
-    For each pair $P_0,P_r$, we build one treatment containing the  issue resolution times for     issues raises in $P_0$ and resolved in $P_r$. These treatments
+    For each pair $P_0,P_r$, we build one treatment containing the  issue resolution times for     issues raised in $P_0$ and resolved in $P_r$. These treatments
     were then ranked by Scott-Knott.
@@
 1 &   &   IntTest         & 17 &   31 & 1.55 & \textcolor{black}{\rule{15mm}{2mm}}  \\
+1 &   &   SysTest         & 13 &   18 & 1.18 & \textcolor{black}{\rule{12mm}{2mm}}  \\
 1 &   &   AcceptTest      & 14 &   19 & 1.27 & \textcolor{black}{\rule{12mm}{2mm}}  \\
-1 &   &   SysTest         & 13 &   18 & 1.18 & \textcolor{black}{\rule{12mm}{2mm}}  \\
 \hline\multicolumn{7}{c}{~}  \\
@@
 1 &    &   IntTest       & 14 &   25 & 1.40 & \textcolor{black}{\rule{14mm}{2mm}}  \\
+1 &   &   SysTest       & 13 &   20 & 1.30 & \textcolor{black}{\rule{13mm}{2mm}}  \\
 1 &    &   AcceptTest    & 16 &   25 & 1.60 & \textcolor{black}{\rule{16mm}{2mm}}  \\
-1 &   &   SysTest       & 13 &   20 & 1.30 & \textcolor{black}{\rule{13mm}{2mm}}  \\
 
@@
 \end{center}
-\caption{Median times to resolve issues seen in the SEI TSP data.
-The main effect here is that there is no clear exponential growth in the time to resolve issues as issues are injected in some phase $P_0$ then resolved in some subsequent issue $P_r$.
-Shown where are the 50th percentiles of issue resolution times for issues injected in phase $P_O$ and resolved in phase $P_r$ (these values are calculated
-by sorting all resolution time, then reporting the middle values of that sort)
-The ``IQR'' column shows the ``inter-quartile range''; i.e. the range of values representing the 75th - 25th percentile range
-The left-hand-side ``rank'' column shows the result of the Scott-Knott ranking procedure described in \tion{stats}. Note that
-most treatments achieved the same ranks i.e. they were found to be insignificantly different  from each other (the one exception is the Planning:UnitTest results 
-where UnitTests were ranked 2).
-The right-hand-side bars
-illustrate shows the releative sizes of the increases for the  50th (median)   percentile values.
-These increases are calculated with respect to the first value in each section ``Before, Planning, Reqts.
-Design, Code''.
-These right-hand-side bars are unitless since they are ratios. For example,
-on the last line, issues injected during coding and fixed in SysTest take 13 minutes (median) to resolve. This is 130\% more than the 10 minutes (median) required
-to resolve coding issues during CodeReview. The right-hand-side bar visually represents that 130\%.}
+\caption{Times to resolve issues seen in the SEI TSP data. For an explanation of this figure,
+see \tion{171projects}.}
 \label{fig:raw}
@@
 
-In \fig{raw}, the data is split out according to issues that were fixed in phase $P_r$ after
-being introduced in phase $P_0$. The data is sub-divided into tables according to $P_0$;
+To understand that figure, we offer the following notes:
+\bi
+\item
+Shown here are the 50th percentiles of issue resolution times for issues injected in phase $P_O$ and resolved in phase $P_r$ (these values are calculated
+by sorting all resolution time, then reporting the middle values of that sort).
+\item
+The ``IQR'' column shows the ``inter-quartile range''; i.e. the range of values representing the 75th - 25th percentile range
+\item
+The results in that figure are  split out according to issues that were fixed in phase $P_r$ after
+being introduced in phase $P_0$. The data are sub-divided into tables according to $P_0$;
 i.e. according to {\em before, planning, requirements, design} or {\em  code}. 
-To ensure the representativeness of each line, we display examples
-where there exists at least $N\ge 30$ examples\footnote{We selected 30
+\item
+The left-hand-side ``rank'' column shows the result of the Scott-Knott ranking procedure described in \tion{stats}. These statistical results were applied separately
+to each group {\em Before, Planning, Reqts, Design, Code}.
+Recall from \S5.5 that if all the fix times within a group
+were sadistically insignificantly different, then they were all earn ``rank=1''.
+Note that
+most treatments achieved the same ranks i.e. they were found to be insignificantly different  from each other (the one exception is the Planning:UnitTest results 
+where UnitTests were ranked 2).
+\item
+The right-hand-side bars
+illustrate shows the releative sizes of the increases for the  50th (median)   percentile values.
+These increases are calculated with respect to the first value in each section ``Before, Planning, Reqts.
+Design, Code''.
+\item
+These right-hand-side bars are unitless since they are ratios. For example,
+on the last line, issues injected during coding and fixed in SysTest take 13 minutes (median) to resolve. This is 130\% more than the 10 minutes (median) required
+to resolve coding issues during CodeReview. The right-hand-side bar visually represents that 130\%.
+\ei
+Technical note: to ensure the representativeness of each line, we display examples
+where there exist at least $N\ge 30$ examples\footnote{We selected 30
 for this threshold via the central limit theorem~\cite{maxwelldata}.} of issues {\em injected} in phase $P_0$ then
 removed in phase $P_r$.
-The right hand side bars of that figure
-illustrate how these resolutions increase (expressed as ratios of resolution time in the phase where they were first injected).
 
@@
 data exhibits large variances-- where are exacerbated by the smaller samples seen in the stratifications).
-Our conclusion from exploring the stratification's is that, given the currently available data, we cannot check
+Our conclusion from exploring the stratifications is that, given the currently available data, we cannot check
 for a DIE effect in subsets of this data.
@@
 Time is reported on a level of minutes.
-We do not have an assessment of the error margin for time reporting.
+We do not have a precise assessment of the error margin for time reporting.
 Some developers are less precise with time or estimates.
-Nonetheless defect time-to-fix must sum to less than the total effort in the phase and time log time stamps must be consistent with the time stamps and phase in the defect log.
-Based on time stamp consistency and conformance with Benford distributions, our experience suggests greater than 90\% of the time log data is recorded in real time.
+Nonetheless, we have applied several tests to verify that the data is accurate.
+First we compare entries from the defect and time logs to verify that defect log times-to-fix sum to less than the total time log effort in the phase.
+Second, time log time stamps must be consistent with both the the time stamps and phase for defect in the defect log.
+Third, we applied a Benford test on the leading digits from the time log  and defect log times to estimate the number data entries that do not result from a natural process (that is, guessed or estimated rather than measured values) ~\cite{shirai14}.
+Based on these tests we believe that greater than 90\% of the time log data is recorded in real time.
 The fidelity and consistency of data will be subject of a future paper.
@@
 
-There are a number of reasons to believe that the data are are consistent between developers and between projects.
+There are a number of reasons to believe that the data are consistent between developers and between projects.
 First, developers receive PSP training, during which instructors focus on complete and accurate data gathering.
@@
 time and defect logs do not match, log entries have excessive rounding, or a developer is an outlier).
-Fourth, 15 of the projects received a TSP Certification in which process fidelity was evaluated independently by an observer and data analyst examining data internal consistency and consistency with  distributional properties known to consistent among all projects and team members.
-Fifth, all projects in this sample used the same data gathering tool.
+Fifth, 15 of the projects received a TSP Certification in which process fidelity was evaluated independently by an observer and data analyst examining data internal consistency and consistency with  distributional properties known to consistent among all projects and team members.
+Sixth, all projects in this sample used the same data gathering tool.
 Nonetheless, some variations exist.
@@
 Note that if such outliers were common in our data, they would appear in the upper percentiles of results.
-\fig{last} shows that no such effect occurs in the 90th percentile of our time to resolve
-issues.    That is, while such ``high risk'' issues may certainly exist, they were not a major effect in our data.
 
@@
 1 &   &   IntTest          & 75 &    2.03 & \textcolor{black}{\rule{20mm}{2mm}}\\
+1 &   &   SysTest           & 40 &     1.08 & \textcolor{black}{\rule{10mm}{2mm}}\\
 1 &   &   AcceptTest       & 44 &     1.19 & \textcolor{black}{\rule{12mm}{2mm}}\\
-1 &   &   SysTest           & 40 &     1.08 & \textcolor{black}{\rule{10mm}{2mm}}\\
 \hline\multicolumn{6}{c}{~}  \\
@@
 1 &    &   IntTest         & 58   & 1.66 & \textcolor{black}{\rule{16mm}{2mm}}\\
+1 &   &   SysTest        & 47 &    1.34 & \textcolor{black}{\rule{13mm}{2mm}}\\
 1 &    &   AcceptTest      & 60 &   1.71 & \textcolor{black}{\rule{17mm}{2mm}}\\
-1 &   &   SysTest        & 47 &    1.34 & \textcolor{black}{\rule{13mm}{2mm}}\\
 
@@
 as \fig{raw}: even when considering the 90th percentile outliers, there is no consistent patterns of 
-exponential growthe in the effort to fix issues delayed later in the lifecycle.}\label{fig:last}
+exponential growth in the effort to fix issues delayed later in the lifecycle.}\label{fig:last}
 \end{figure}
@@
 The purpose of this study is to draw attention to the notion that commonly-held belief of the delayed issue effect may not be a universal truth.
-This study adds to the evidence offered by the cased study in Royce~\cite{Royce98}.
+This study adds to the evidence offered by the case study in Royce~\cite{Royce98}.
 Our study invites a further explanation into the causal factors that mitigate DIE.
@@
 Foundation grants NSF-CISE 1302169 and CISE 1506586.
+
+This material is based upon work funded and supported by TSP Licensing under Contract No.
+FA8721-05-C-0003 with Carnegie Mellon University for the operation of the Software Engineering Institute, a federally funded research and development center sponsored by the United States Department of Defense.
+This material has been approved for public release and unlimited distribution.
+DM-0003956
+
 Personal Software Process$\textsuperscript{SM}$, Team Software Process$\textsuperscript{SM}$, and TSP$\textsuperscript{SM}$ are service marks of Carnegie Mellon University.
@@
 However, our data supports a number of possible explanations for the observed intermittent appearance of the DIE.
-Since the current data is not suitable for choosing among the various hypothesized explanations we feel that formulating a new theory build from this data
+Since the current data are not suitable for choosing among the various hypothesized explanations we feel that formulating a new theory build from this data
 would be premature at this time. That said, we think the current paper would be useful
@@
 found in some phase and fixed in that phase? The text says
-"The distribution of defects found and fixed per phase in
+"The distribution of defects  fixed per phase in
 our data is shown in Figure 9. A high percentage of defects
-(44) were found and fixed in the early phases, i.e.,
+(44) were "found and fixed" (that is, the necessary change was identified and implemented)   in the early phases, i.e.,
 requirements, high level design, and design reviews and
@@
 into two charts, which simplifies both. Please see \fig{raw} and \fig{last}.
-Also, please note that that fit\fig{raw} now has an extensive help text in the caption.
+Also, please note that fit\fig{raw} now has an extensive help text in the caption.
 We used your comments there where we noted the ratios used in the

### 2016-09-15 phaseDelay4Emse.tex
Updates from ShareLaTeX

--- 2016-07-06 phaseDelay4Emse.tex
@@
 only in  certain kinds of projects.  This is a significant result since it predicts that  new development
-processes that promise to faster retire more issues will have only a  limited return on investment, and that a long-held truth in software engineering can be defeated on a large scale. 
+processes that promise to faster retire more issues will not have a guaranteed return on investment (depending on the context where applied), and that a long-held truth in software engineering should not be considered a global truism.
 \end{abstract}
@@
 
-The dilemma of failing truths in the face of new evidence is not particular to software engineering. 
+The dilemma of updating truths in the face of new evidence is not particular to software engineering. 
 % To be fair, 
@@
 Devanbu et al.~\cite{prem16}  who examined responses from 564 Microsoft software developers from around
-the world, they found that  ``(a)~programmers do indeed have very
+the world; they found that  ``(a)~programmers do indeed have very
 strong beliefs on certain topics; (b)~their beliefs are primarily formed
@@
 research; (c)~beliefs can vary with each project, but do not necessarily
-correspond with actual evidence in that project''.
+correspond with actual evidence in that project.''
 Devanbu et al. further  comment that ``programmers give personal experience
-as the strongest influence in forming their opinions''. This is a troubling
+as the strongest influence in forming their opinions.'' This is a troubling
 result, especially given the above comments from Passos and  J{\o}rgensen et al.~\cite{passos11,jorgensen09} about how quickly practitioners form, freeze, and rarely revisit those opinions.
@@
 From these, we gathered additional citations for the delayed issue effect, the vast majority of which were secondary sources, e.g.,~\cite{Leffingwell96,mead2004software,mcconnell1996software,mcconnell01,Tassey2002,boehm2012}.
-Our literature search is not exhaustive, but our results yielded an obvious trend: nearly every citation to the delayed issue effect could be traced to the seminal \textit{Software Engineering Economics}~\cite{Boehm81} or its related works~\cite{boehm88,boehm01}.\footnote{For example, popular sources such as~\cite{pressman2005software, boehm01,glass02,endres03} have a combined citation count of over 14,500 on Google Scholar can all trace their evidence to \textit{Software Engineering Economics}~\cite{Boehm81}.}
+Our literature search is not exhaustive, but our results yielded an obvious trend: nearly every citation to the delayed issue effect could be traced to the seminal \textit{Software Engineering Economics}~\cite{Boehm81} or its related works~\cite{boehm88,boehm01}.\footnote{For example, popular sources such as~\cite{pressman2005software, boehm01,glass02,endres03}, with a combined citation count of over 14,500 on Google Scholar, can all trace their evidence to \textit{Software Engineering Economics}~\cite{Boehm81}.}
 
@@
 
-Some studies that report less-than large
- increase in the effort required to fix delayed issues.  Boehm~\cite{Boehm80} provides data suggesting that the cost-to-fix curve for small projects
+Some studies report smaller
+ increases in the effort required to fix delayed issues.  Boehm~\cite{Boehm80} provides data suggesting that the cost-to-fix curve for small projects
  %, two student projects of 2000 deliverable source instructions, 
@@
 The process data was collected by the team members to self-manage their personal and team work.
-Them members also exhibited self-management behavior by estimating planning and scheduling the work tasks.
-
-While the processes and work flows among these projects can vary, the logical order is described section \ref{development_phase} is followed.
+The members also exhibited self-management behavior by estimating planning and scheduling the work tasks.
+
+While the processes and work flows among these projects can vary, the logical order described in section \ref{development_phase} is followed.
 Development tasks such as requirements development, design, or code, are typically followed by an appraisal phase such as personal review or inspection.
@@
 Summary statistics and the year of project initiation are displayed in Figure~\ref{fig:Project-descriptive-stats}.
-From this table we can make some observations about the range of project characteristics .
+From this table we can make some observations about the range of project characteristics.
 
@@
 \centering
+\renewcommand{\arraystretch}{1}%
 \begin{tabular}{lrrrrrrrl}
                                         & N   & Min & Q1   & Median & Q3    & Max   & Mean   &  Distribution \\\hline
-Team size                                                          & 171 & 1   & 4    & 6      & 10    & 36    &  7.8    &  \includegraphics[width=15mm]{img/team_size.png} \\ 
-Duration {[}days{]}                                                & 171 & 7   & 33   & 61     & 118   & 1918  & 107    & \includegraphics[width=15mm]{img/duration.png} \\
-Added \& Modified LOC                                                 & 117 & 2   & 1125 & 4201   & 13092 & 88394 & 10259  & \includegraphics[width=15mm]{img/LOC.png} \\
-Components                                                         & 171 & 8   & 26   & 49     & 107   & 4170  & 116.2  & \includegraphics[width=15mm]{img/components.png} \\
-Year                                                               & 171 &     & 2011 & 2012   & 2013  & 2014  & 2011.9 & \includegraphics[width=15mm]{img/year.png}
+Team size                               & 171 & 1   & 4    & 6      & 10    & 36    &  7.8    &  \includegraphics[width=25mm]{team_size.png} \\ 
+Duration {[}days{]}                     & 171 & 7   & 33   & 61     & 118   & 1918  & 107    & \includegraphics[width=25mm]{duration.png} \\
+Added \& Modified LOC                   & 117 & 2   & 1125 & 4201   & 13092 & 88394 & 10259  & \includegraphics[width=25mm]{LOC.png} \\
+Defects Found \& Fixed                  & 171 & 1   & 28   & 95     & 278   & 4580  & 324.4  & \includegraphics[width=25mm]{defect_by_project.png} \\
+Components                              & 171 & 8   & 26   & 49     & 107   & 4170  & 116.2  & \includegraphics[width=25mm]{components.png} \\
+Project Initiation Year                 & 171 &    2006 & 2011 & 2012   & 2013  & 2014  & 2011.9 & \includegraphics[width=25mm]{year.png}
 \end{tabular}
@@
 
- Of the 171 projects in the sample, only 117 collected size data in lines of code, however all projects tracked effort and  the component counts with applied effort are provided.
-Other data is complete for all 171 projects.
+ Of the 171 projects in the sample, only 117 collected size data in lines of code.
+However all projects tracked effort and  the component counts with applied effort are provided.
+Other data are complete for all 171 projects.
 The projects were mostly of short duration and small to medium size.
-The median project began in 2012 lasted 61 days, produced 4200 Lines of Code, 49 components (modules or features).
-Duration ranged from 7 to 1918 days.
-Size ranged from minimal (this may represent a short maintenance project) to 88394.
+The median project began in 2012 lasted 61 days, produced 4,200 Lines of Code, 49 components (modules or features).
+Duration ranged from 7 to 1,918 days.
+Size ranged from minimal (this may represent a short maintenance project) to 88,394.
 The earliest project was in 2006 and the most recent in 2014.
@@
 \begin{center}
-\includegraphics[height=2.5in]{img/fix-phase-dist.pdf}
+\includegraphics[height=2.5in]{fix-phase-dist.pdf}
 \end{center} 
-\caption{Distribution of defects found and fixed by phase.}
+\caption{Distribution of defects by phase removed.}
 \label{fig:fix-phase-dist}
@@
     different results); and an effect size test (to reject any significant differences that are trivially small).
-
-    The Scott-Knott ranker 
+    Scott-Knott allows for a direct answer to the following questions:
+    \bi
+    \item
+    Given an issue
+    raised at phase $i$ and resolved at phase $\forall j,k \in \{i,i+1,i+2,...\}$,...
+    \item 
+    ... Is it
+    true that  the time to resolve issues in phase $j$ is significantly different 
+    to the time to resolve issues in phase $k$? 
+    \ei
+    Note that if $j,k$ times are significantly different, then we can compare
+    the median values to say (e.g.) resolution time at phase $k$ is 3 times slower than phase $j$.
+    Note also that if all times $j,k$ are {\em not} significantly different
+    then we say that the phases all {\em rank} the same (and we denote this by setting
+    all such {\em ranks} to 1).
+
+    In the following results, we nearly always encountered the second case;
+    i.e.   the times to resolve issues at different times were usually {\em not significantly}
+    different.
+
+    As to technical details of the Scott-Knott methods, this ranker 
       was   recommended  by  Mittas and Angelis in a
@@
     halves.
-    Before Scott-Knot recurses,  however,  it apples some  statistical hypothesis test $H$ to check
+    Before Scott-Knot recurses,  however,  it applies some  statistical hypothesis test $H$ to check
     if $m,n$ are significantly different. 
-
-    To operationalize ``interesting'', Scott-Knott seeks the division of 
-    of $l$ treatments into subsets of size $m,n$ of sizes $ls,ms,ns$ and median values
-    $l.\mu, m.\mu, n.\mu$ (respectively) in order to maximize
- $\frac{ms}{ls}abs(m.\mu - l.\mu)^2 + \frac{ns}{ls}abs(n.\mu - l.\mu)^2$.
-
-   To operationalize $H$, we use both  bootstrap sampling  and the A12 effect size test.
+    To operationalize ``interesting'', 
+    \bi
+    \item
+    Scott-Knott seeks the division of 
+    $l$ treatments into subsets of size $m,n$ (so if $n$ was 
+    appended to the  end of $m$ then that new list
+    would the same as $l$).
+    \item
+    We say that $l,m,n$ have
+    sizes $ls,ms,ns$ and median values
+    $l.\mu, m.\mu, n.\mu$ (respectively) 
+    \item
+    Scott-Knott tries all ways to split $l$ into $m,n$ and returns
+    the one that maximizes the differences in the mean values
+    before and after the splits; i.e. 
+ \[\frac{ms}{ls}abs(m.\mu - l.\mu)^2 + \frac{ns}{ls}abs(n.\mu - l.\mu)^2\]
+   \ei
+   To operationalize $H$, we use both  bootstrap sampling  and Vargha and Delaney's A12 effect size test.
    In other 
@@
     small. Hence, to assess the performance differences we first must rule out 
-    small effects using A12, a test   recently endorsed by Arcuri and 
+    small effects using Vargha and Delaney's A12 test, a test    endorsed by Arcuri and 
     Briand at ICSE'11~\cite{arcuri11}.
@@
 \begin{center}
-\input{tex/fig13}
+
+%\resizebox{\columnwidth}{!}{
+\scriptsize
+\begin{tabular}{c|lr|rr|rl}
+& \multicolumn{2}{c|}{ } & \multicolumn{2}{c|}{Percentiles} & \multicolumn{2}{c}{~ } \\ 
+  & \multicolumn{2}{c|}{ } & \multicolumn{2}{c|}{(units = } & \multicolumn{2}{c}{Growth with respect to earliest phase  } \\ 
+   & \multicolumn{2}{c|}{Phase} & \multicolumn{2}{c|}{  minutes)} & \multicolumn{2}{c}{(unitless  ratios of two time values) } \\\hline
+
+  rank & injected & removed & 50th &   IQR & \multicolumn{2}{c}{50th percentile growth}  \\ 
+\hline
+
+\multicolumn{7}{c}{~}  \\
+1 &  Before   
+      &   DesignInspect   & 10 &  14 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}}  \\
+1 &   &   CodeReview      & 8 &  14 & 0.80 & \textcolor{black}{\rule{8mm}{2mm}}  \\
+1 &   &   CodeInspect     & 10 &  16 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}}  \\
+1 &   &   UnitTest        & 12 &   21 & 1.20 & \textcolor{black}{\rule{12mm}{2mm}}  \\
+1 &   &   IntTest         & 15 &   31 & 1.50 & \textcolor{black}{\rule{15mm}{2mm}}  \\
+1 &   &   SysTest         & 11 &   22 & 1.10 & \textcolor{black}{\rule{12mm}{2mm}}  \\  
+\hline\multicolumn{7}{c}{~}  \\
+1 &  Planning     &   ReqtsReview     & 8 &  14 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}}  \\
+1 &               &   DesignInspect   & 11 &  13 & 1.38 & \textcolor{black}{\rule{13mm}{2mm}} \\
+2 &               &   UnitTest        & 24 &   25 & 3.00 & \textcolor{black}{\rule{30mm}{2mm}} \\
+\hline\multicolumn{7}{c}{~}  \\
+1 &  Reqts   &   ReqtsReview     & 13 &  20 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}} \\
+1 &          &   ReqtsInspect    & 12 &   18 & 0.92 & \textcolor{black}{\rule{9mm}{2mm}}  \\
+1 &          &   DesignReview    & 10 &   14 & 0.77 & \textcolor{black}{\rule{7mm}{2mm}} \\
+1 &          &   DesignInspect   & 9 &   15 & 0.69 & \textcolor{black}{\rule{6mm}{2mm}}   \\
+1 &          &   CodeInspect     & 13 &  24 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}} \\
+1 &          &   UnitTest        & 10 &  17 & 0.77 & \textcolor{black}{\rule{7mm}{2mm}}  \\
+1 &          &   IntTest         & 33 &   42 & 2.54 & \textcolor{black}{\rule{25mm}{2mm}}  \\
+1 &          &   SysTest         & 24 &  108 & 1.85 & \textcolor{black}{\rule{18mm}{2mm}}  \\
+\hline\multicolumn{7}{c}{~}  \\
+1 & Design   &   DesignReview    & 11 &   16 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}}  \\
+1 &   &   DesignInspect   & 8 &   12 & 0.73 & \textcolor{black}{\rule{7mm}{2mm}}  \\
+1 &   &   CodeReview      & 10 &   18 & 0.91 & \textcolor{black}{\rule{9mm}{2mm}}  \\
+1 &   &   CodeInspect     & 9 &  14 & 0.82 & \textcolor{black}{\rule{8mm}{2mm}}  \\
+1 &   &   UnitTest        & 11 &   18 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}}  \\
+1 &   &   IntTest         & 17 &   31 & 1.55 & \textcolor{black}{\rule{15mm}{2mm}}  \\
+1 &   &   SysTest         & 13 &   18 & 1.18 & \textcolor{black}{\rule{12mm}{2mm}}  \\
+1 &   &   AcceptTest      & 14 &   19 & 1.27 & \textcolor{black}{\rule{12mm}{2mm}}  \\
+\hline\multicolumn{7}{c}{~}  \\
+1 &  Code   &   CodeReview    & 10 &  16 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}}  \\
+1 &    &   CodeInspect   & 10 &   15 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}}  \\
+1 &   &   UnitTest      & 12 &   20 & 1.20 & \textcolor{black}{\rule{12mm}{2mm}}  \\
+1 &    &   IntTest       & 14 &   25 & 1.40 & \textcolor{black}{\rule{14mm}{2mm}}  \\
+1 &   &   SysTest       & 13 &   20 & 1.30 & \textcolor{black}{\rule{13mm}{2mm}}  \\
+1 &    &   AcceptTest    & 16 &   25 & 1.60 & \textcolor{black}{\rule{16mm}{2mm}}  \\
+
+\end{tabular}
+%}
 %
@@
 \end{center}
-\caption{Median times to resolve issues seen in the SEI TSP data.
-The main effect here is that there is no clear exponential growth in the time to resolve issues as issues are injected in some phase $P_0$ then resolved in some subsequent issue $P_r$.
-Shown where are the 50th percentiles of issue resolution times for issues injected in phase $P_O$ and resolved in phase $P_r$ (these values are calculated
-by sorting all resolution time, then reporting the middle values of that sort)
-The ``IQR'' column shows the ``inter-quartile range''; i.e. the range of values representing the 75th - 25th percentile range
-The left-hand-side ``rank'' column shows the result of the Scott-Knott ranking procedure described in \tion{stats}. Note that
-most treatments achieved the same ranks i.e. they were found to be insignificantly different  from each other (the one exception is the Planning:UnitTest results 
-where UnitTests were ranked 2).
-The right-hand-side bars
-illustrate shows the releative sizes of the increases for the  50th (median)   percentile values.
-These increases are calculated with respect to the first value in each section ``Before, Planning, Reqts.
-Design, Code''.
-These right-hand-side bars are unitless since they are ratios. For example,
-on the last line, issues injected during coding and fixed in SysTest take 13 minutes (median) to resolve. This is 130\% more than the 10 minutes (median) required
-to resolve coding issues during CodeReview. The right-hand-side bar visually represents that 130\%.}
+\caption{Median times to resolve issues seen in the SEI TSP data. For an explanation of this figure,
+see \tion{171projects}.}
 \label{fig:raw}
 \end{figure}
-
+ \begin{figure}[!t]
+% \renewcommand{\baselinestretch}{0.8}
+% \small
+\begin{center}
+%\resizebox{\columnwidth}{!}{
+\scriptsize
+\begin{tabular}{c|lr|r|rl}
+  & \multicolumn{2}{c|}{ } &  Percentiles & \multicolumn{2}{c}{Growth with respect to earliest phase  } \\ 
+   & \multicolumn{2}{c|}{Phase} &  (units= miniutes) & \multicolumn{2}{c}{(unitless ratios of two time values) } \\\hline
+
+  rank & injected & removed &   90th   &  \multicolumn{2}{c}{90th percentile growth} \\ 
+\hline\multicolumn{6}{c}{~}  \\
+1 &  Before   &   DesignInspect     & 32  &  1.00 & \textcolor{black}{\rule{10mm}{2mm}}\\
+1 &   &   CodeReview      &   31   &    0.97 & \textcolor{black}{\rule{9mm}{2mm}}\\
+1 &   &   CodeInspect     &   32   &    1.00 & \textcolor{black}{\rule{10mm}{2mm}}\\
+1 &   &   UnitTest        &   45  &    1.41 & \textcolor{black}{\rule{14mm}{2mm}}\\
+1 &   &   IntTest         &   63  &   1.97 & \textcolor{black}{\rule{19mm}{2mm}}\\
+1 &   &   SysTest         &   46   &   1.44 & \textcolor{black}{\rule{14mm}{2mm}}\\  
+\hline\multicolumn{6}{c}{~}  \\
+1 &  Planning     &   ReqtsReview       & 35   &   1.00 & \textcolor{black}{\rule{10mm}{2mm}}\\
+1 &               &   DesignInspect     & 31  &   0.89 & \textcolor{black}{\rule{8mm}{2mm}}\\
+2 &               &   UnitTest          & 53 &   1.51 & \textcolor{black}{\rule{15mm}{2mm}}\\
+\hline\multicolumn{6}{c}{~}  \\
+1 &  Reqts   &   ReqtsReview      & 42 &    1.00 & \textcolor{black}{\rule{10mm}{2mm}}\\
+1 &          &   ReqtsInspect      & 40 &     0.95 & \textcolor{black}{\rule{9mm}{2mm}}\\
+1 &          &   DesignReview      & 34 &    0.81 & \textcolor{black}{\rule{8mm}{2mm}}\\
+1 &          &   DesignInspect     & 38 &     0.90 & \textcolor{black}{\rule{9mm}{2mm}}\\
+1 &          &   CodeInspect      & 45 &   1.07 & \textcolor{black}{\rule{10mm}{2mm}}\\
+1 &          &   UnitTest          & 40 &   0.95 & \textcolor{black}{\rule{9mm}{2mm}}\\
+1 &          &   IntTest          & 95 &      2.26 & \textcolor{black}{\rule{22mm}{2mm}}\\
+1 &          &   SysTest          & 126 &    3.00 & \textcolor{black}{\rule{30mm}{2mm}}\\
+\hline\multicolumn{6}{c}{~}  \\
+1 & Design   &   DesignReview     & 37 &   1.00 & \textcolor{black}{\rule{10mm}{2mm}}\\
+1 &          &   DesignInspect    & 28 &    0.76 & \textcolor{black}{\rule{7mm}{2mm}}\\
+1 &   &   CodeReview        & 40 &     1.08 & \textcolor{black}{\rule{10mm}{2mm}}\\
+1 &   &   CodeInspect      & 33 &     0.89 & \textcolor{black}{\rule{8mm}{2mm}}\\
+1 &   &   UnitTest          & 41 &     1.11 & \textcolor{black}{\rule{12mm}{2mm}}\\
+1 &   &   IntTest          & 75 &    2.03 & \textcolor{black}{\rule{20mm}{2mm}}\\
+1 &   &   SysTest           & 40 &     1.08 & \textcolor{black}{\rule{10mm}{2mm}}\\
+1 &   &   AcceptTest       & 44 &     1.19 & \textcolor{black}{\rule{12mm}{2mm}}\\
+\hline\multicolumn{6}{c}{~}  \\
+1 &  Code   &   CodeReview      & 35    & 1.00 & \textcolor{black}{\rule{10mm}{2mm}}\\
+1 &    &   CodeInspect    & 32   & 0.91 & \textcolor{black}{\rule{9mm}{2mm}}\\
+1 &   &   UnitTest      & 45  & 1.29 & \textcolor{black}{\rule{12mm}{2mm}}\\
+1 &    &   IntTest         & 58   & 1.66 & \textcolor{black}{\rule{16mm}{2mm}}\\
+1 &   &   SysTest        & 47 &    1.34 & \textcolor{black}{\rule{13mm}{2mm}}\\
+1 &    &   AcceptTest      & 60 &   1.71 & \textcolor{black}{\rule{17mm}{2mm}}\\
+
+\end{tabular}
+%}
+\end{center}
+\caption{90th percentile times to resolve issues seen in the SEI TSP data.   Same format as \fig{raw} (but here we look at 90th percentile
+outliers while   \fig{raw} explored the central tendencies of the data).}\label{fig:last}
+\end{figure}
 \subsection{Observations from 171 Projects}\label{sect:171projects}
 
-The distribution by phase in which defects were discovered (that is both found and fixed) in our data is shown in Figure~\ref{fig:fix-phase-dist}.
+The count by phase in which defects were removed is shown in Figure~\ref{fig:fix-phase-dist}.
 Defects are counted only if they they escape the introduction phase unless a bad fix introduces a new defect.
@@
 
-\fig{raw} shows the  50th  percentile
+\fig{raw} and \fig{last} show the  50th and 90th  percentile (respectively)
 of the time spent resolving issues
@@
 
-In \fig{raw}, the data is split out according to issues that were fixed in phase $P_r$ after
-being introduced in phase $P_0$. The data is sub-divided into tables according to $P_0$;
+To understand that figure, we offer the following notes:
+\bi
+\item
+Shown here are the 50th/90th percentiles of issue resolution times for issues injected in phase $P_O$ and resolved in phase $P_r$ (these values are calculated
+by sorting all resolution time, then reporting the middle values of that sort).
+\item
+The ``IQR'' column shows the ``inter-quartile range''; i.e. the range of values representing the 75th - 25th percentile range
+\item
+The results in that figure are  split out according to issues that were fixed in phase $P_r$ after
+being introduced in phase $P_0$. The data are sub-divided into tables according to $P_0$;
 i.e. according to {\em before, planning, requirements, design} or {\em  code}. 
-To ensure the representativeness of each line, we display examples
-where there exists at least $N\ge 30$ examples\footnote{We selected 30
+\item
+The left-hand-side ``rank'' column shows the result of the Scott-Knott ranking procedure described in \tion{stats}. These statistical results were applied separately
+to each group {\em Before, Planning, Reqts, Design, Code}.
+Recall from \S5.5 that if all the fix times within a group
+were statistically insignificantly different, then they all earn ``rank=1''.
+Note that
+most treatments achieved the same ranks i.e. they were found to be insignificantly different  from each other (the one exception is  within the  Planning:UnitTest results 
+where UnitTests were ranked 2).
+\item
+The right-hand-side bars
+show the relative sizes of the increases for the  50th (median)   percentile values.
+These increases are calculated with respect to the first value in each section ``Before, Planning, Reqts.
+Design, Code''.
+\item
+These right-hand-side bars are unitless since they are ratios. For example,
+on the last line of \fig{raw}, issues injected during coding and fixed in SysTest take 13 minutes (median) to resolve. This is 130\% more than the 10 minutes (median) required
+to resolve coding issues during CodeReview. The right-hand-side bar visually represents that 130\%.
+\ei
+Technical note: to ensure   representativeness, we display examples
+where there exist at least $N\ge 30$ examples\footnote{We selected 30
 for this threshold via the central limit theorem~\cite{maxwelldata}.} of issues {\em injected} in phase $P_0$ then
 removed in phase $P_r$.
-The right hand side bars of that figure
-illustrate how these resolutions increase (expressed as ratios of resolution time in the phase where they were first injected).
-
-The two key features of \fig{raw} are:
+
+The two key features of \fig{raw} and  \fig{last} are:
 \be
@@
 these results do we see the kind of very large increases reported in the papers
-documenting DIE. For example, consider the ratio of the issue resolution time
-between {\em Before/DesignInspect} and {\em Before/SysTest} result. That ratio is 1.11 which is far smaller
+documenting DIE; neither in the median fix times of \fig{raw}
+or at the 90th percentile level of \fig{last}. For example, consider the ratio of the issue resolution time
+between {\em Before/DesignInspect} and {\em Before/SysTest} result of \fig{raw}. That ratio is 1.11 which is far smaller
 than the scale ups seen in \fig{b81}.  
 \item
-Nearly all the supposed increases seen in \fig{raw} are insignificantly different
+Nearly all the supposed increases seen in \fig{raw} and \fig{last} are insignificantly different
 to the other treatments.  The left hand column of \fig{raw} shows the results of the Scott-Knott statistical tests. Note that nearly all
@@
 broadly at the entire data set but not at specific stratifications. To address that concern,
-we spent some time reproducing \fig{raw} for various subsets of our data.
+we spent some time reproducing these figures for various subsets of our data.
 That proved to be an unfruitful-- no stratification was found that 
@@
 of those stratifications  exacerbated the large IQR's seen in this data\footnote{Recall that in a sorted list of numbers,
-the inter-quartile range, or IQR, is difference between the 
-  75th to 25th percentile value.}.
-Our 171 projects   stratifies into subsets of varying sizes. Sorting those sizes, we get a list
-comprising of 17,12 then numerous much smaller stratifications.  Reasoning over such small samples
-is problematic in the general case and, in the case of our data, it is ever more problematic due to
-the large IQRs of the data
-(to see these large IQRs,  please consider the 50th and IQR columns of Figure~\ref{fig:raw} where  most of our IQRs were larger than tan the 50th percentile; i.e. software
-data exhibits large variances-- where are exacerbated by the smaller samples seen in the stratifications).
-Our conclusion from exploring the stratification's is that, given the currently available data, we cannot check
+the inter-quartile range, or IQR, is the difference between the 
+  75th and 25th percentile value.}.
+Our 171 projects   stratify into subsets of varying sizes. 
+The two largest subsets contained only 17 and 12 projects, with numerous much smaller stratifications.
+Reasoning over such small samples
+is problematic in the general case and, in the case of our data, it is even more problematic due to
+the large IQRs of the data.
+(To see these large IQRs,  please compare the 50th percentile and IQR columns of Figure~\ref{fig:raw}, where  most of the IQRs are larger than the 50th percentile; i.e. software
+data exhibits large variances, which in this case are exacerbated by the smaller samples seen in the stratifications).
+Our conclusion from exploring the stratifications is that, given the currently available data, we cannot check
 for a DIE effect in subsets of this data.
@@
 Before moving on, we comment on some of the counter-intuitive
-results in \fig{raw}. Consider, for example, the ``Reqts'' results were
+results in these figures. Consider, for example, the ``Reqts'' results of  \fig{raw} where
 the time required to fix issues actually tends to {\em decrease} the longer they are left in the system.
-In terms of explaining this result, the key thing is the left-hand-side statistically ranking: all these treatments were found to be statistically indistinguishable.
+In terms of explaining this result, the key thing is the left-hand-side statistical ranking: all these treatments were found to be statistically indistinguishable.
 In such a set of treatments, the observed
@@
 Our primary analysis in \fig{raw} is based on injection-removal phase pairs whose with sample size $>30$.
-The justification for the statistical techniques used in this paper are provided in \tion{stats}.
+The justification for the statistical techniques used in this paper is provided in \tion{stats}.
 
@@
 also be doubted.
-In TSP, developers are trained and supplied templates for defect and time tracking, all data entry is double-checked by the team TSP coach, and TSP demands that developers analyze their data to make process improvements.
+In TSP, developers are trained and supplied with templates for defect and time tracking, all data entry is double-checked by the team TSP coach, and developers are required to analyze their data to make process improvements.
 That is, TSP developers are always testing
 if their project insights are accurate. In such an environment,
-it is more likely that they will better identify the injection phase.
+it is more likely that they will accurately identify the injection phase.
 
@@
 One of the seminal books on quasi-experimentation and reporting of validity concerns, Cook and Campbell~\cite{cook1979}, had not been published when most of the DIE papers were written.} Note that in Figure~\ref{fig:cost-to-fix}, the units of ``cost-to-fix'' are not expressed -- in the source references, cost appears as calendar time, effort, and price.
-In the TSP, a defect is defined as ``any change to a product, after its construction, that is necessary to make the product correc'' and time to correct a defect includes ``the time to analyze, repair, and validate a defect fix''.
-Our analysis of DIE focuses on time as a measure of time-as-effort (persons * time).
+In the TSP, a defect is defined as ``any change to a product, after its construction, that is necessary to make the product correc'' and time to correct a defect includes ``the time to analyze, repair, and validate a defect fix.'' Our analysis of DIE focuses on time as a measure of time-as-effort (persons * time).
 
@@
 For example, a low time-to-fix may come at the expense of a more robust solution, i.e., a quick and dirty fix instead of an elegant repair.
-
-\begin{figure}[!t]
-% \renewcommand{\baselinestretch}{0.8}
-% \small
-\begin{center}
-\input{tex/figlast}
-\end{center}
-\caption{90th percentile tiles to resolve issues in the SEI TSP data. Same format as \fig{raw} (but here we look at 90th percentile
-outliers while   \fig{raw} explored the central tendencies of the data). The conclusion from this figure is the same
-as \fig{raw}: even when considering the 90th percentile outliers, there is no consistent patterns of 
-exponential growthe in the effort to fix issues delayed later in the lifecycle.}\label{fig:last}
-\end{figure}
 
@@
 most software systems were large, monolithic, and ''write once and maintain
-forever''.
-Today, even large software systems are trending toward DevOps and cloud-based deployment.
+forever.'' Today, even large software systems are trending toward DevOps and cloud-based deployment.
 Advances in network communications, CPU processing power, memory storage, virtualization, and cloud architectures have enabled faster changes to software, even for large systems.
@@
 Harter et al.
-find a statistical correlation between fewer high severity defects and rigourous process discipline in large or complex systems~\cite{harter2012does}.
+find a statistical correlation between fewer high severity defects and rigorous process discipline in large or complex systems~\cite{harter2012does}.
 Technical advancements in processing power, storage, networking, and parallelism have combined with a deeper scientific understanding of software construction to enable a whole host of software assurance technologies, from early-phase requirements modeling to automated release testing.
@@
 
-\input{tex/responses}
+ \section*{Response to Reviewers}
+
+{\bf Editor's comments}
+
+The reviewers have carefully scrutinized the revised version, and all of the found the new version being globally improved with respect to the major concerns, and thus found the manuscript being worth publishing.
+At the same time, changes have introduced some inconsistencies and minor issues (such as a figure caption being too extensive) that the authors should fix them before submitting the final version.
+
+\begin{quote}{\em  That you for your comments.
+We have removed that excessively long figure text and fixed the typos and other small inconsistencies mentioned by the reviewers.
+We hope this revision meets with your approval.
+}\end{quote}
+
+{\bf Reviewer1}
+
+Thank you for carefully revising the manuscript.
+All my comments have been addressed - either elaborated in the paper or explained in the rejoinder.
+The manuscript is mature and my remaining comments are all minor.
+Unfortunately, it is not easy to refer to specific comments in the rejoinder, but I do my best below.
+
+I think the manuscript has improved considerably, especially in regard to three key aspects.
+First, the data and the overall development context has greatly improved.
+This will help future readers understand the results and allow further interpretation.
+Second, the paper now connects to previous research on theory building in software engineering - already in the introduction section.
+While no new theories are put forward in this paper, I think the discussion is important, and the pointers to future work could inspire more work.
+Third, I appreciate the authors' efforts to stratify the data.
+Although there are no new findings, I believe the discussion belongs in the paper - I missed it in the previous version.
+
+R1.1: Note that there is no Section 6.6 in the manuscript though, make sure you didn't forget anything.
+
+\begin{quote}{\em Reply1.1: Our bad. Wrong section reference. One more Latex compile and that bad ref went away.
+}\end{quote}
+
+Minor comments:
+
+R1.2Section 4: I think the paragraphing could be improved - the highly interesting point about agile is hidden.
+"A goal of agile methods is to reduce /---/ little empirical data exist" appears in the middle of a paragraph headed by "Shull et al.
+conducted a literature survey and held a series of e-workshops".
+Consider restructuring this section to bring forward the agile perspective.
+
+\begin{quote}{\em 
+Reply1.2: We agree that this section "buries the lead". We have moved those critical sentences to the front of the section.
+}\end{quote}
+
+R1.3: I don't fully understand Section 5.2.2 paragraph 4, and the relation between plan items and the time tracking logs.
+"One or more defects are reported against a single plan item, e.g., a review session, an inspection meeting, a test execution".
+Multiple defects can be mapped to a plan item?
+But a plan item can also be "resolving a defect" (page 11:46).
+Also, the time tracking logs include time to collect data, prepare a fix, and its validation (page 12:50)… But what if a plan item includes both running a test suite and resolving the identified defect?
+The equation presented on page 13 suggests that the "time-to-fix a defect" could be inflated by e.g.
+running slow test cases.
+I'm sure the authors have all this covered, but the text confuses me.
+Could it possibly be revised?
+Maybe another figure could clarify the situation?
+
+\begin{quote}{\em 
+Reply1.3: We apologize for the confusion.
+We have rewritten this paragraph to be more understandable.
+The key point is that we measure defects both directly and by summing effort in defect removal phases.
+The total removal phases produce a greater cost estimate by including overhead.
+}\end{quote}
+
+R1.4: Section 5.2.3: There is a minor discrepancy between Figure 7 and the text.
+In the running text you list system tests before acceptance tests, but the opposite order is depicted in the figure.
+Make sure the order is correct, also in Figure 13.
+
+\begin{quote}{\em 
+Reply1.4: Our bad. System comes before acceptance. Changed throughout. Good catch!
+}\end{quote}
+
+R1.5: Figure 10 - I like this figure, especially the final distribution column!
+But please clarify the year… I'm sure many projects are not completed within a calendar year.
+Does the year show when the project started or finished (released)?
+
+\begin{quote}{\em 
+Reply 1.5: Thank you for catching this ambiguity and for the positive comment on the distributions. We reported the start year and have modified the figure to clarify.
+}\end{quote}
+
+R1.6: Figure 12 - Found and fixed? Does this mean each defect appears twice in the figure? Could you separate them? The current figure is hard to penetrate.
+
+\begin{quote}{\em 
+Reply 1.6: Each defect is counted only once.
+The term "found and fixed" has a specific meaning, but this meaning has not come through clearly to more than one reviewer.
+To TSP practitioners, "found and fixed" means that the required change was "identified and implemented".
+We modified the text to clarify this that this operational term is a single activity.
+}\end{quote}
+
+R1.7 Figure 13 - The caption is far too long. Could parts of it be moved to a paragraph in the running text?
+
+\begin{quote}{\em 
+Reply1.7: Moved into text
+}\end{quote}
+
+R1.8: Section 6.1 - It is not obvious to me how to interpret "conformance with Benford distribution". Could you please elaborate somewhat?
+
+\begin{quote}{\em 
+Reply 1.8: A Benford test is used in forensic accounting to detect human manipulation of data entries.
+The distribution is not quite, but vaguely log-normal and occurs naturally in frequency distributions, such as log tables and data entries.
+Human alterations (guesses) cause deviations from the expected distribution.
+We have applied this test to log entries to estimate those recorded in real time vs those estimated or guessed.
+We have re-written the paragraph and included a citation to clarify the meaning.
+}\end{quote}
+
+R1.9: Section 6.1 paragraph 7 - "Fourth" appears twice.
+
+\begin{quote}{\em 
+Reply 1.9: Fixed
+}\end{quote}
+
+R1.10: Section 6.3 paragraph 5 - "Figure 14 shows that no such effect occurs…" How? Severity is not presented in the figure. Moreover, did your stratification cover high-severity defects?
+
+\begin{quote}{\em 
+Reply 1.10: You are correct. That is an incorrect inference. We have deleted that sentence.
+}\end{quote}
+
+R1.11:
+
+Details:
+\item Page 24:12 - "the cased study"
+\item Page 7:24 - "Some studies that report less-than" Remove extra "that".
+\item Page 12:41 - "quote elaborate" quite?
+\item Page 14:30 - "the of the"
+\item Page 14:45 - "complete submitted"
+\item Page 15-16 - No thousands separators.
+\item Page 16:48 - "only 25\% of the teams…" Was this supposed to be a separate item in the bullet list?
+\item Page 18:51 - "raises" -\> "raised"
+
+\begin{quote}{\em 
+Reply 1.11: Fixed. Thanks!
+}\end{quote}
+
+{\bf Reviewer2}
+
+Reviewer #2: This revised version (R1) of the paper contains many corrections, changes, and enhancements.
+The authors have considered all the reviewer's comments and have provided precise responses.
+My own observations (i.e.
+Reviewer #2) have been globally satisfied, and thanks to comments from reviewers #1 and #3, the paper has been significantly enhanced.
+
+\begin{quote}{\em 
+Thank you for that comment.
+}\end{quote}
+
+Nonetheless, there are still some issues that need some clarifications.
+
+R2.1 My main concerns are related to data and statistics sections:
+
+As requested, the authors have added descriptive statistics about the data.
+However, some of these are not very relevant here, i.e.
+number of projects per organization (Fig.
+9), year distribution (Fig.11 last row).
+Moreover, Fig.
+11 could be complemented with the number of defects distribution.
+
+\begin{quote}{\em 
+Reply 2.1: We have added a defect count distribution as suggested.
+}\end{quote}
+
+R2.2- The statistical analysis section (§5.5) can be enhanced by better explicating the calculations.
+The Scott-Knott algorithm is a particular clustering algorithm, and the reader needs some clarification how it contributes to the demonstration.
+
+\begin{quote}{\em 
+Reply 2.2: You are quite correct- our description of Scott-Knott was opaque. We have added more text at the start of 5.5.
+}\end{quote}
+
+2.3 Moreover, Fig.13 is still insufficiently clear.
+The authors added a lengthy legend; I tend to think this would better be placed in the text.
+The column on the left hand (i.e.
+"rank") is still unclear to me.
+
+\begin{quote}{\em 
+Reply 2.3: . You are correct: that explanation text in the figure caption was confusing. By moving it to to section 5.6, it can be expanded into a dot list (easier to read) and extra details
+can be added (for example, as you suggest, we can discuss more the meaning of the "rank" column).
+}\end{quote}
+
+2.4 - I am troubled by Fig.
+12, how many defects have in total been fixed in the same phase in which they have been found?
+What is the proportion from the initial 47,376 defect logs that are concerned in the statistics in Fig.
+13?
+
+\begin{quote}{\em 
+Reply 2.4: "Find and fix" is a single sub-task, so all defects are "found and fixed" in the same phase. This is distinct from injected and found.
+}\end{quote}
+
+- Finally, looking at Fig.6 and defect type distribution, wouldn't be relevant to explore any possible effect of defect type on the results in Fig.
+13?
+As defect severity is absent, may be defect type has some influence.
+It could be that certain categories of defect can be corrected any time during the project, it could even be that these defects, in particular, are delayed for later (because they are easy to repair)/
+
+\begin{quote}{\em 
+Reply 2.5: Indeed, some activities are more likely to find certain types of defects. This was explored in a pair of TSP Symposium papers using PSP data:
+
+D. Vallespir and W. Nichols, “An Analysis of Code Defect Injection and Removal in PSP,” in Proceedings of the TSP Symposium 2012, 2012. , and
+
+D. Vallespir, “Analysis of Design Defect Injection and Removal in PSP,” in TSP Symposium 2011, 2011, pp. 1–28.
+
+However, exploring this with data from so many distinct projects presents challenges and analysis beyond the scope of this paper.
+This is an interesting question that should be pursued in future work.
+Although some types of defect are clearly more challenging than the others (syntax is often simple), the discovery phase and activity is a much larger effect.
+Examining this relationship is a very good idea, but we judged that to be a significant future effort to elaborate on the more basic finding.
+}\end{quote}
+
+2.6 Minor remarks:
+
+\item p3, line 1: "that" repeated
+
+\item Section 5.2: the numbers about the TSP database can be regrouped in a table
+\item Fig.11: the presentation of the table could be enhanced by reducing the size of columns N, Min, etc. and enlarging the column Distribution to make the graphics visible.
+\item p. 17: the full name should be mention for the "A12 effect size test" (i.e. Vargha and Delaney's A12 test, cf. [1])
+\item p.17, line 36: "it applies"
+\item p.17, line 38: "of" repeated
+\item Section 5.6: the last paragraphs about the possible explanations could be regrouped in a separate subsection
+\item p.21, line 20: "are" repeated
+
+\begin{quote}{\em 
+Reply 2.6: Fixed
+}\end{quote}
+
+{\bf Reviewer3}
+
+Reviewer #3: The authors have adequately addressed my observations. Below is a list minor comments (with one exception, all of them are typos).
+
+R3.1: Section 5.2.2
+"quote elaborate" should be "quite elaborate"
+
+Section 5.4
+"discuss the of the projects" should probably be "discuss those of the projects"
+
+"the logical order is described section 5.2.3 is followed" should be "the logical order described in Section 5.2.3 is followed"
+
+In Fig. 11, the "min" Year is not reported (even though the text says it's 2006).
+
+"teams meet at least weekly" should be "teams met at least weekly" (everything else is in the past tense)
+
+Section 5.5
+
+"it apples some statistical hypothesis test" should be "it applies some statistical hypothesis test"
+
+"the division of of l treatments" should be "the division of l treatments"
+
+\begin{quote}{\em 
+Reply 3.1: These items have been fixed. Thanks!
+}\end{quote}
+
+3.2 The description of the Scott-Knott ranker is not terribly clear.
+You write "Scott-Knott seeks the division of of l treatments into subsets of size m, n of sizes ls,ms, ns and median values l.μ,m.μ, n.μ (respectively) in order to maximize \[ms/ls* abs(m.μ − l.μ)^2 + ns/ls* abs(n.μ − l.μ)^2\] but it's not clear to me what "subsets of size m, n of sizes ls,ms, ns and median values l.μ,m.μ, n.μ (respectively)" means.
+Are these 5 subsets?
+How are they related?
+
+\begin{quote}{\em 
+Reply 3.2: That text was unclear and we have fixed it. See the second dot list in Section 5.5.
+}\end{quote}
+
+3.3 In the caption of Figure 13, you write "(these values are calculated by sorting all resolution time, then reporting the middle values of that sort) The", which should be "(these values are calculated by sorting all resolution times, then reporting the middle values of that sort).
+The"
+
+Section 5.6
+"examples where there exists at least N >= 30 examples" should be "examples where there exist at least N >= 30 examples"
+
+Should "stratification's" be "stratifications"?
+
+Section 6.1
+"the data are are consistent" should be "the data is consistent" (you've always used "data" as a singular noun everywhere else in the paper)
+
+Section 6.4
+"cased study" should be "case study"
+
+"the data are" should be "the data is"
+
+\begin{quote}{\em 
+Reply 3.3: fixed
+}\end{quote}
 

### 2016-09-15 phaseDelay4Emse.tex
typo

@@
 Details:
+\bi
 \item Page 24:12 - "the cased study"
@@
 \item Page 16:48 - "only 25\% of the teams…" Was this supposed to be a separate item in the bullet list?
-\item Page 18:51 - "raises" -\> "raised"
-
+\item Page 18:51 - "raises" to "raised"
+\ei
 \begin{quote}{\em 
@@
 
-Reviewer #2: This revised version (R1) of the paper contains many corrections, changes, and enhancements.
+Reviewer \#2: This revised version (R1) of the paper contains many corrections, changes, and enhancements.
 The authors have considered all the reviewer's comments and have provided precise responses.
 My own observations (i.e.
-Reviewer #2) have been globally satisfied, and thanks to comments from reviewers #1 and #3, the paper has been significantly enhanced.
+Reviewer \#2) have been globally satisfied, and thanks to comments from reviewers \#1 and \#3, the paper has been significantly enhanced.
 
@@
 
+\bi
 \item p3, line 1: "that" repeated
@@
 \item p.21, line 20: "are" repeated
-
+\ei
 \begin{quote}{\em 
@@
 
-Reviewer #3: The authors have adequately addressed my observations. Below is a list minor comments (with one exception, all of them are typos).
+Reviewer \#3: The authors have adequately addressed my observations. Below is a list minor comments (with one exception, all of them are typos).
 

### 2016-09-16 old_reviewer_response.tex
Updates from ShareLaTeX

 \section*{Response to Reviewers}

{\bf Editor's comments}

Dear authors,

Three knowledgeable reviewers have scrutinized the manuscript
and found it interesting and relevant,  although in need for
certain improvements. Therefore, the authors are
 invited to submit a revised version for a new review, where
 the reviewers' comments
are addressed.

\begin{quote} {\em Thank you for the careful reviews. Using
those comments we have extensively revised and extended this draft.
Specifically, the old version was 20 pages and this version is 29 pages (and the new
material all comes from the clarification and correction points offered by the reviewers).
We hope that  you will find this draft greatly improved.}
\end{quote}

Particularly, this holds for:

1) the connection to activities of theory building in ESE
(reviewers 1).

\begin{quote}{\em This new draft now maps our work into the SE literature on theory building. This new material appears in the introduction as well as throughout the document.}\end{quote}

2) clarification about the data origin and
characteristics (reviewer 1).

\begin{quote}{\em
This new draft contains much more detail on the origin and characteristics of our data (see \tion{data-collection} and \tion{data_character}).}\end{quote}

3)  use of statistical analysis (reviewers 2 and 3)

\begin{quote}{\em All our results are now augmented with statistical significance tests as well as effect size tests (see \tion{stats} and \tion{171projects}). Those tests strengthen the overall message of this paper-- that there is no significant
phase delay effect.}\end{quote}

- writing style, that would benefit form being more
precise (reviewer 3)

\begin{quote}{\em We have carefully read those comments and have made numerous changes
as directed by reviewer3.}\end{quote}

{\bf Reviewer \#1's Comments}

The paper investigates whether a widely
accepted "rule" in software engineering holds by looking at
a large set of empirical data. The research topic is highly
relevant and the results are interesting.   The discussion
on reassessing old truism fits the scope of the journal very
well.  The language in the paper is excellent, and the
structure is (although non-standard) easy to follow.

\begin{quote}{\em Thank you for those kind comments.}\end{quote}

It is not easy for the reader to understand what parts
of the "life of an issue" the authors claim are affected by
DIE, and it is also not too easy to understand what data the
authors have available.

\begin{quote}{\em  The introduction now includes an example of DIE  (see Fig 1).
Also there are now more precise  definitions on how we measure this effect in \tion{data-collection}}\end{quote}

 The authors do not relate their reassessment of an
``old truism'' to the growing set of papers on theory building
in software engineering. With access to such large amounts
of empirical evidence, it should be possible to take some
steps toward improved DIE theories.

\begin{quote}{\em We now map our work to prior publications on theory-building in software engineering, as well as theory falsification in general. That mapping appears in numerous comments through out the text (using the references offered by this reviewer) with the bulk of the comments being top of p2 and the conclusion.

Notably, we now introduce the DIE as an example of such a theory (in the introduction), and explain that this is done on the basis of it providing a description of a Software development phenomenon accompanied by a rationale that explains why such behavior is expected. While we introduce in the conclusions several rationales that can explain why the effect called for by the theory was not observed in our dataset, we feel that future work is needed to collect data in support or contradiction of these explanations before a new theory can be formulated with confidence. }\end{quote}

The different projects studied are all grouped together. I
believe studying clusters of development contexts would be
highly interesting, but unfortunately the current
characterization of projects is inadequate.

\begin{quote}{\em  We agree with this comment and so took two actions for this draft-- one that was effective and one that was ineffective.

First, this draft now contains extensive notes on the different contexts of the
projects studied here. Those notes are offered in \tion{data_character} and \tion{171projects}.

Second (and this was the ineffective step), we  searched the categories   listed in \tion{data_character} and \tion{171projects}
in order to check if any stratification contained the delayed issue effect. 
That study   failed to find a stratification
containing an exponential expansion in the time to fix issues. The reason for this was  the small size
of those stratifications  exacerbated the large IQRs seen in this data\footnote{Recall that in a sorted list of numbers,
the inter-quartile range, or IQR, is difference between the 
  75th to 25th percentile value.}:
\bi
\item
Our 171 projects   stratify into subsets of varying sizes. 
The two largest subsets contained only 17 and 12 projects, with numerous much smaller stratifications.
Reasoning over such small samples
is problematic in the general case and, in the case of our data, it is ever more problematic due to
the large IQRs of the data.
\item
To see these large IQRs,  please consider the 50th and IQR columns of Figure~\ref{fig:raw} Note that most of our IQRs were larger than the 50th percentile; i.e. software
data exhibits large variances, which are only exacerbated by the smaller samples seen in the stratifications.
\ei
Our conclusion from exploring the stratification's is, given the currently available data, we cannot check
for a DIE effect in subsets of this data.

Note that the above notes are now added to the paper at the end of Section 6.6.
 }\end{quote}

DETAILED REVIEW: The authors should expand on what they
include in the ``delayed issue effect'' (DIE), in particular
in relation to standard milestones in the life of an issue
and the corresponding timespans.  I'm particularly
interested in whether all "indirect" issue resolution steps
are covered, e.g., change impact analysis, updating
documentation, and regression testing. The effort involved
in these steps clearly depends on the development context of
the specific project. The short a-c) list on page 14:50
suggest that the issues studied are restricted to minor
fixes, i.e., no updates to architecture, changed hardware,
recertification, updated user manuals etc.

\begin{quote} {\em
We now provide a more explicit definition of what constitutes a defect (our study of issues is limited to defects) and the ``defect lifecycle'' phases we measure in \tion{data-collection}. These definitions greatly expand on the now-deleted list on 14:50 in teh previous version. In TSP, ``a defect is any change to a product, after its construction, that is necessary to make the product correct''. This includes both minor issues (typography) and severe issues (architecture changes, requirements errors). Time on a defect includes a) time to investigate/analyze a defect once discovered, b) time to craft and implement a fix, and c) time to verify and close the defect.

Regarding the concern that we may only be reporting trivial defects, we have added \fig{dtypes}, which shows the distribution of defect types in our data. It is true that around a quarter of the fixes were documentation changes, but the other 75\% are function-altering behavior in the system. 
}\end{quote}

Section 6.3 describes the data, but some aspects
should be further explained.

\begin{quote} {\em
At your suggestion, we have added extensive notes on the nature of our data--
see \tion{data_character} and \tion{171projects}.
}\end{quote}

Logging interruption time
sounds very useful, but I wonder how carefully the
developers actually did this 
 it must be really hard to keep up the discipline required.
 If you are interrupted as a developer (e.g., phone calls,
 urgent mail, someone asks a question) I don't think the
 first thing you do is to stop a timer on your computer.
 Moreover, developers often work on several defects in
 parallel, and might interweave bug fixing with new
 development. I don't think the "interruption time" captures
 all such multi-tasking, and it should at least be properly
 discussed in Section 7 "Threats to validity".

\begin{quote}{\em 
Yes, there is undoubtedly some accuracy in the time-tracking. We provide some evidence on the rigor of the TSP data collection process in \tion{defin}, \tion{data_integrity}, and \tion{data_character}, and a more extensive discussion on the issue of time-tracking and defect logging accuracy in \tion{conclusion_validity} on Conclusion Validity. We note that the data being collected here was not collected
for the purposes of this paper. Rather, it was being collected and used by the
programmers to assist them in their day-to-day management of the project.
Given this, we have confidence that the logging information is not widely inaccurate.

Please also consider the sanity checks reported in a companion paper (referenced here \tion{data_integrity}). In summary, this data has been heavily vetted.
}
\end{quote}

The discussion on reassessing old truisms (Section 3) is
interesting, but it should be complemented by the
perspective of theory building in software engineering - An
active research topic lately, with a dedicated workshop
series (GTSE). I suggest looking into the following
references for a start: Sj{\o}berg et al. (2008) ``Building
Theories in Software Engineering'', Smolander and Paivarinta
(2013) ``Theorizing about software development practices'',
and Stol and Fitzgerald (2015) "Theory-oriented software
engineering". Considering new empirical evidence is
obviously critical to theory building, and discussing your
new results in the light of an theory creation would be
valuable. The authors mention that DIE appears to occur
intermittently in certain kinds of projects - maybe the
authors could elaborate on this idea and present an improved
DIE theory based on what they now know? I believe the
authors have the best available data to do so, and I would
expect the paper to go beyond simply questioning the "old
truth".

 \begin{quote}{\em 
We thank the reviewer for this important observation and have connected our work to the literature on theory-building. However, our data supports a number of possible explanations for the observed intermittent appearance of the DIE. Since the current data are not suitable for choosing among the various hypothesized explanations we feel that formulating a new theory build from this data
would be premature at this time. That said, we think the current paper would be useful
as a data point which, when combined with other sources, could be useful for some new future theory of SE.
}\end{quote}

The authors study three claims in this paper, and the
third claim is the central one: "delayed issues are not
harder to resolve". To study whether issues require longer
resolution times in later phases, the authors analyze a
large set of issue data from historical projects. Thus the
manuscript reports from an observational study rather than
an experiment with a controlled delivery of treatments.
While I believe the authors' approach is practical, I would
like to see a critical discussion on threats to validity of
observational studies (e.g., Madigan et al., A Systematic
Statistical Approach to Evaluating Evidence from
Observational Studies, Annu. Rev. Stat. Appl. 2014. 1:11-39
and Carlson and Morison, Study Design, Precision, and
Validity in Observational Studies, J Palliat Med. 2009 Jan;
12(1): 77-82). An alternative study design (although
difficult to realize in a system of industrial size) would
be to let different developers resolve the same issues for
the same software system, i.e., one group resolves an issue
during design, another during implementation, and a third
during testing - some discussion along these lines would
strengthen the validity section.

\begin{quote}{\em 
 We have reviewed and cited the mentioned papers. We have incorporated many of the concerns raised in these papers into the expanded \tion{external_validity} on External Validity.
 There, we note that Sj{\o}berg et al. list three classes of threats to External Validity (the smallest number) and we address all three of them. In truth, the only way to address External Validity threats is to have a huge sample, broadly representative of your domain, over an extended period of time-- e,g. using the 171 projects from 2006 to 2014 studied in this paper. The issues external validity in observational studies raised in \cite{madigan2014} and \cite{carlson2009} are more applicable to studies that compare a treatment and a control, whereas we are observing an effect that leads us to question the theory that DIE is a truism.
}\end{quote}

The first and second claims are studied with much less
rigor. "DIE is a commonly held belief" is studied using a
survey of software engineers, both practitioners and
researchers. According to Fig. 2 (actually a table) [NOTE: Fig. 2 is now \fig{survey_results}], the
number of respondents is 16 and 30 for practitioners and
researchers, respectively. Sixteen respondents from industry
represent a tiny survey of a very general claim that any
software engineer could respond to. Why were not more
answers collected? The authors do not have much evidence to
defend the first claim, and there is no discussion of the
corresponding validity threats. The second claim, "DIE is
poorly documented", is studied using a literature review.
Unfortunately, the method underlying the literature review
is not presented. Although it doesn't need to be an SLR of
the most rigorous kind, the authors should report how the
papers were identified. I suspect the terminology used to
describe the DIE phenomenon is highly diverse, thus it would
strengthen the paper if the authors reported how they
reviewed the literature. According to Page 19:34 [NOTE: the enumerated list in the Conclusion] only eight
publications were identified.

\begin{quote}{\em We agree with the reviewer - the phrasing of our claims and the supporting evidence were not congruent. Our intention is to argue from multiple angles that the evidence for DIE is sparse and, more importantly, that DIE persists as a popular belief. Our goal here is to motivate that we should re-investigate the DIE. We have toned down the language and combined our arguments on ``widely-believed'' and ``poorly-documented'' into a new \tion{belief} whose focus is to motivate our study. We have added more details on our literature search and snowball sampling. We have also highlighted how blog articles often cite secondary sources, which in turn cite Software Engineering Economics, to illustrate how the DIE pehnemonon has perpetuated.}\end{quote}

The iterative fashion of modern software development,
with agile at the extreme end, is not fully discussed (the
short discussion section could be extended). The phases of
the linear development of the 80s, (such as in Fig. 1) [NOTE: Fig. 1 remains \fig{b81}] 
probably don't exist in many of the projects in the TSP
dataset, still the authors discuss the DIE-effect from the
perspective of the 80s. Page 16:31 states "DIE has been used
to justify many changes to early lifecycle software
engineering" - does this mean the agile movement
successfully mitigated DIE? This possibility is not fully
considered in the paper.
How many of the 171 projects were
(more or less) agile? This appears to be an important
characteristic of the included projects - very important to
describe!

\begin{quote}{\em We provide more information on the types of projects included in our sample in \tion{data_character}

More to the point -- does agile help mitigate DIE? The answer is - we don't know. A strong argument can be made that all iterative \& incremental methodologies were designed to mitigate the DIE. Further, huge advances in software engineering technology and process have occurred since DIE was first observed, many of which have precisely targeted the DIE (static analysis, test-first development, automated build). Perhaps our analysis provides evidence that DIE can be defeated, at a large scale across multiple varied projects, using modern technologies. We do not yet know the causes here, but certainly it appears the DIE should not be treated as a truism any more, but rather a variable that can be controlled. We have added more discussion on this to \tion{discussion}.
}\end{quote}

Concerning characterization of the 171 projects, the
paper needs to report much more detail. I would expect to
see some descriptive statistics.   The paper does not report much
characterization of the 171 projects. I strongly suggest the
authors to dig deeper into the data, and analyzing for which
types of projects the findings hold. What patterns are there
to discover? I suggest Petersen and Wohlin, "Context in
industrial software engineering research" (ESEM2009) for
details on how to characterize development contexts.

\begin{quote}{\em Much more context is now offered in \tion{data_character} and \tion{171projects}}. \end{quote}

Fig 1 [NOTE: Still \fig{b81}]: "Operation" dwarfs everything in Boehm's diagram.
Since you do not study anything post release in this paper,
I think this figure skews the reader's mental picture. If
you remove the rightmost bar and rescale accordingly, the
plot better matches the findings you report in Fig. 10 [NOTE: now \fig{raw}]. You
still identify an interesting result though, but the
presentation turns fairer.

\begin{quote}{\em We understand why you are making that suggestion.
That figure is something of a classic. Therefore we opted not to chop the figure but hope the text is sufficiently clear}.
\end{quote}

MINOR COMMENTS: 
Several figures are copied from previous work. Are all copyrights properly
managed?

\begin{quote}{\em We have removed all copyrighted images and replaced them with original drawings based on information from the cited source.}\end{quote}

\fig{raw}: Black and red is not a good pair for
grayscale printouts. Please use black and gray instead.
\begin{quote}{\em  We agree: switched to black and gray.  }\end{quote}

Some figures should be resized to better match the page width.

\begin{quote}{\em  We agree: fixed. }\end{quote}

Page 3:14 - "The above argument" Which argument? Could be precisely specified.

\begin{quote}{\em  We agree: that statement has been deleted. }\end{quote}

Page 3:22 - Please provide the full link to the dataset.

\begin{quote}{\em  Our policy is to get one publication out of paper
before issuing the data in public. If this paper is accepted, we will certainly add that link. Note that we have a long history of living up to that policy-- see http://openscience.us/repo. }\end{quote}

Page 18:26 - First sentence: "Unexpected results such as this one". Ambiguous reference,
please be specific.  
\begin{quote}{\em  We agree. That back reference is removed. }\end{quote}

Page 18:27 - "We also survey the state of SW dev. practice /---/": given the size of the survey,
this statement feels a bit bold.

\begin{quote}{\em We have removed this statement.}\end{quote}

TYPOS ETC.

Sec 1, §1: [3] and [30] have been swapped?

Page 14:46 - Spell out IV\&V the first (and only) time it appears.

Page 3:24 - missing verb (is) Page 16:43 - "Did this study failed"

Page 18:44 - Appache

Page 19:19 - Al Ref 26 - oo --> OO

\begin{quote}{\em  All of these have been corrected. }\end{quote}

{\bf Reviewer \#2's comments}

The topic of this paper is the empirical
evaluation of a largely admitted claim in software
engineering, i.e. the exponential cost of correcting errors
according to the phase in which the errors are discovered.
The claim is first confirmed by surveying and interviewing
practitioners and experts. Next, the authors use a large set
of data from projects in SEI database. The results show
that, strictly speaking, the claim does not hold (although
some effects of delayed corrections can be noticed).

Globally, the paper is well written and the authors develop
a convincing demonstration. The topic of the paper is highly
relevant for both practitioners and people from academia.
Among many, I personally believed in this claim and had
regularly taught it in my courses. Thus, reading this paper
offers a refreshing perspective on our understandings of
software engineering background and theoretical knowledge. I
also like the suggested posture that insists on the
necessary skepticism that we should have against such
prevailing claims.

Beyond this positive global impression, I have some concerns
about the study reporting and analysis: -  It seems to me
that more descriptive statistics about the sample would
contribute to a better understanding of the scope of the
study.

\begin{quote}{\em 
More description about the context and sample have been added in \tion{data_character} and \tion{171projects}.}\end{quote}

I think in particular about the size (lines of codes and/or
 number of software components); a duration histogram would
 also be relevant. Accordingly, the formula for calculating
 variable "Total effort" in Fig. 7 [NOTE: now \fig{dist}] could be explained, and
 Figure 9 [NOTE: now \fig{fix-phase-dist}] made more readable (bigger size).

 \begin{quote}{\em We have added more description of the projects in \tion{data_character}. We have also added more precise definitions of our measures in \tion{data-collection}. We have adjusted the size of figures throughout the paper to be more readable. We have added more text 
 to explain {\em effort} in \fig{dist}.}\end{quote}

If certain attributes for the issues and errors are
available in the data (e.g. severity, priority, etc.), they
should also be brought to the readers knowledge.

\begin{quote}{\em 
Those notes are offered in \tion{data_character} and \tion{171projects}.}\end{quote}

Figure 10 [NOTE: now \fig{raw}] is central to the paper's demonstration; its
expressiveness could be enhanced: the formula for calculating the 50th
and 95 ratio could be provided,  the unit of column
"Percentile" could be mentioned.

\begin{quote}{\em
\fig{raw}'s caption (formerly Figure 10) now has extensive explanation text. e.g. units
are added on line two. Also, an example of the calculations of the ratios is offered
in the last line of the caption. Further,
to simplify comprehension of this diagram, it has been split into two
(and the more arcane discussion about the outliers, statistics gathered, and methods used is provided is also provided).}\end{quote}

I like the discussion section, in particular, the idea
of software architectures that could be a contributing
factor for enhancing software evolution and reducing the
cost of issues and errors fixing. Is there any available
knowledge in the data set concerning any architectural
design choices made in each project?

\begin{quote}{\em
Not explicitly, but \fig{dtypes} lists ``interfaces'' as a the fourth most
common type of defect. Note that ``interfaces'' are changed by architectural
issues. }\end{quote}

In the same trend of ideas, I think what has
fundamentally changed since early days of SE is the
development of requirements engineering. In actual SE
practices, problem and the solution spaces are
systematically explored thanks to RE techniques; together
with architecture, this could also explain why things have
changed since the 70'. It could be, for example, that people
tend to make less severe errors in early software project
phases (thanks to RE techniques); a similar phenomenon was
observed when SE practices become more mature (see Harter et
al. 2012).

References Harter D. E., Kemerer C. F., Slaughter S. (2012).
Does Software Process Improvement Reduce the Severity of
Defects? A Longitudinal Field Study. IEEE Transactions on
Software Engineering, vol. 38, n° 4, p. 810-827.
```

\begin{quote}{\em
  Absolutely - there are potentially many explanations for why the DIE was NOT observed in our dataset. We add more discussion on the potential causes in \tion{discussion}, supported by the reference to Harter et al.
 }\end{quote}

In the conclusion section, I feel uncomfortable with
the assertion "That data held no trace of the delayed issued
effect" (line 17). As mentioned elsewhere in the paper, the
delayed issue is not absent; it is much less significant and
systematic than what is usually claimed. Moreover, this
reduction has been demonstrated for medium sized projects;
we cannot generalize to larger projects.  
\begin{quote}{\em
  We agree -That line has been removed.
 }\end{quote}

Minor remarks:

- p.9, line 29 : " … reported by Shull [52], found that the cost to find certain non-critical classes of defects ..." => is it the "cost to find" or the "cost to FIX"?

- p.10, line 21: duplicate word "there" -

p.12, line19: missing word "One of THE guiding …"

- p.16, line 33: "distinguish" instead of "distinguished"?

- p.18, line 42: explain acronym MEAN

- p.20, line 20: useless enumeration A1?

\begin{quote}{\em
 All of these have been corrected or removed.
 }\end{quote}

{\bf  Reviewer \#3's Comments}:

I like the idea of the paper and I certainly
would like to see more empirical studies that periodically
check if our beliefs about software engineering practices
still hold (or even if they have ever held). The authors
address one of the beliefs that are more entrenched in the
population of researchers and practitioners, and a very
important one too.

\begin{quote}{\em
Thanks for those kinds words.
 }\end{quote}

At the same time, I think that the paper needs to be
improved before it can be published.

The empirical analysis (Section 6 [NOTE: now \tion{analysis}) is a bit of a let down. I
would like to see some sort of more robust and rigorous
statistical analysis, based on statistical tests.

\begin{quote}{\em
We agree. \tion{stats} lists our stats methods and \fig{raw}
shows the results (see the left-hand column: all those ``1''s mean
that our stats could not find any difference between the fix times for
delaying or not delaying.
 }\end{quote}

The paper seems to be written in a somewhat casual style,
which makes the paper more pleasant to read, but less clear
in some parts (see Detail Comments below).

\begin{quote}{\em
Thanks for those pointers. We have fixed the issue.
 }\end{quote}

DETAIL COMMENTS

Section 2 [NOTE: still \tion{claims}]

I'm not really sure what the authors mean by "We say that a
measure collected in phase 1, ., i, .. j is very much more
when that measure at phase j is larger than the sum of that
measure in earlier phases 1 <= i < j." You are defining a
property of a measure, but, the way this definition is
written, it seems as if you are defining the property of a
measure of "being much more," without any further
qualifications.  I can guess that you mean that a measure
$m$ (collected in phases, so you can denote by $m(i)$ the
value of $m$ in phase $i$) has this property *in phase $j$*
(so, it's a property m has in a specific phase and not a
general property of the measure) if $\sum_{1 <= i < i} m(i)
< m(j)$. 
\begin{quote}{\em
Thanks for that. We added the $m_j$ notation to \tion{claims}, as you suggested.
 }\end{quote}

Your definition of "difficult issue" is "Issues are more
difficult when their resolution takes more time or costs
more (e.g. needs expensive debugging tools or the skills of
expensive developers)." That's not a very precise
definition, though, because it has two different
interpretations, one in terms of time and the other in terms
of effort. The reasons why you introduce this definition
become clearer only in Section 7.2 [NOTE: now \tion{construct}], in the Threats to
Validity, and that's too late. You should move some of the
discussion of Section 7.2 here.
\begin{quote}
{\em Good idea! We did just that- see middle para in \tion{claims}.}
\end{quote}
Why would the term "delayed issue effect" be a
generalization of the rule "requirements errors are hardest
to fix"? "Delayed" seems to be quite specific as it appears
to refer to time, while "hardest to fix" may refer to other
variables, like effort or cost.
\begin{quote}
{\em Not sure if this a reviewer requirement for something to be fixed
prior to publication, or a clarification question. We used these
terms since when we ran the survey, the audience clearly made
that generalization. Does this reply adequately to your comment?}
\end{quote}
In Claim 3, "very much more harder" sounds like a
little bit too much ... ("more harder"?)
\begin{quote}
{\em Our bad. We fixed that.}
\end{quote}
Section 4

A very minor issue: why is Figure 2 [NOTE: now \fig{survey_results}] a ... figure, instead of
a table? Same for all other tables ...
\begin{quote}
{\em Authors' preference. When a document has many tables and figures, navigating
between them is hard. Calling them all "figures" makes navigation faster. }
\end{quote}
Section 5.1

This is not a complete sentence "All the literature
described above that reports the onset of DIE prior to delivery.''

\begin{quote}
{\em Fixed}
\end{quote}

Section 6.3

You should rephrase your definition "a defect is any change," since a defect is what existed before the change was made and not the change itself.
\begin{quote}
{\em We now provide a more precise definition of defect in \tion{defin}.
}
\end{quote} 
You do not explain that "QualTest" in Fig. 8 [NOTE: now \fig{waterfall}] is.
\begin{quote}
{\em A mistake on our part that has been removed from the figure.}
\end{quote}
The definition of "time per defect - The total \# of defects found in a plan item during a removal phase divided by the total time spent on that plan item in that phase." is not correct as it is, as this would basically be the number of defects per unit time, instead. The roles of time and
defects should be reversed.
\begin{quote}{\em We now provide a more precise definition of time measurement related to defects in \tion{data-collection}}\end{quote}

Besides adding a more rigorous statistical analysis, you
need improve your explanations.
\begin{quote}{\em We agree. \tion{stats} lists our stats methods and \fig{raw}
shows the results (see the left-hand column: all those ``1''s mean
that our stats could not find any difference between the fix times for
delaying or not delaying.}\end{quote}

Figure 9 [NOTE: now \fig{fix-phase-dist}]. The caption says "Distribution of defects found
and fixed by phase." Is that the distribution of defects
found in some phase and fixed in that phase? The text says
"The distribution of defects  fixed per phase in
our data is shown in Figure 9. A high percentage of defects
(44) were "found and fixed" (that is, the necessary change was identified and implemented)   in the early phases, i.e.,
requirements, high level design, and design reviews and
inspections." which doesn't add much. If that's the
distribution of defects introduced in a phase and fixed
immediately in that phase, it is not clear why you introduce
it, though.
\begin{quote}
{\em 
The TSP convention is to fix (that is remove) a defect in the phase in which it was identified.  The term "find and fix" refers to the effort to isolate the necessary change and implement that change. With some exceptions, defects are not counted as defects unless they escape that phase. The exception case is the secondary defect introduced during a fix that occurs with some frequency in test and rarely in inspection or review. We have clarified this distinction in the text.}
\end{quote}
Figure 10 [NOTE: now \fig{raw}]. The explanation of the data and histograms in the
figure is, at best, confusing. The caption says "50th and
95th percentiles of issue resolution times" and the opening
sentence of Section 6.4 says "Figure 10 shows the 50th and
95th percentile of the time spent resolving issues ..."
However, these are not "times," because you write "expressed
as ratios of resolution time in the phase where they were
first injected" a couple of paragraphs below. This seems to
be the right interpretation, but right after that you write
"The BLACK and RED bars show the increases for the 50th
(median) and 95th percentile values, respectively," which
would be a third interpretation of the data in Figure 10.
\begin{quote}
{\em We agree-that figure was far too confusing. We have now split the 50th and 90th peercentile results
into two charts, which simplifies both. Please see \fig{raw} and \fig{last}.
Also, please note that fit\fig{raw} now has an extensive help text in the caption.
We used your comments there where we noted the ratios used in the
bard charts are unitless while the left-hand-side column numbers are
times in minutes.
}
\end{quote}

You need to at least mention (or, better, discuss) some of
your results, because they appear to be "counterintuitive,"
since they show that, for example, median resolution times
can even decrease if issues are not fixed immediately, but
in later phases. For example, looking at the Reqts section,
most percentages are way below 1, which seems to indicate
that some issues are actually much easier to fix at later
stages (maybe because more or better information about the
software system is available only later). That would be a
very interesting result by itself, especially if you can
provide some statistical support for it.
\begin{quote}{\em
The last paragraph of \tion{171projects} now comments
on that effect. }
\end{quote}

Maybe you could analyze the projects developed in a
"traditional" way and check if there is some sort of DIE
there. For example, if you look at the projects developed
with the waterfall life cycle, maybe you will find larger
DIE than for the other life cycles, which might partially
justify the claims of previous researchers about the
existence of DIE. This would also show that newer kinds of
life cycles (like Agile ones) help solve DIE. You do provide
hints about this in Section 8 [NOTE: now \tion{discussion}] ("We also note that other
development practices have changed in ways that could
mitigate the delayed issued effect") and Section 9 [NOTE: now \tion{conclusion}], but
that's clearly not enough.
\begin{quote}{\em

If your comment is along the lines of ``look for stratifications to check if
DIE exists in subsets'' then we offer the following notes (which we added to the end of \tion{171projects}):

We looked into   stratifications of the data and, so far,   failed to find a stratification
containing an exponential expansion in the time to fix issues. The reason for this was  the small size
of those stratifications  exacerbated the large IQR's seen in this data\footnote{Recall that in a sorted list of numbers,
the inter-quartile range, or IQR, is difference between the 
  75th to 25th percentile value.}:
\bi
\item
Our 171 projects   stratifies into subsets of varying sizes. Sorting those sizes, we get a list
comprising of 17,12 then numerous much smaller stratifications.  Reasoning over such small samples
is problematic in the general case and, in the case of our data, it is ever more problematic due to
the large IQRS of the data:
\item
To see these large IQRs,  please consider the 50th and IQR columns of Figure~\ref{fig:raw} Note that most of our IQRs were larger than tan the 50th percentile; i.e. software
data exhibits large variances-- where are exacerbated by the smaller samples seen in the stratifications.
\ei
Our conclusion from exploring the stratifications is, given the currently available data, we cannot check
for a DIE effect in subsets of this data.
}
\end{quote}
\end{document}

### 2016-09-16 lucasRelatedWork.tex
Updates from ShareLaTeX

@@
 The first data on the difficulties of resolving delayed issues as a function of lifecycle phase date back to large systems in the late 70s from IBM~\cite{Fagan76}, TRW~\cite{Boehm76}, GTE~\cite{Daly77}, and Bell Labs~\cite{Stephenson76} (Figure~\ref{fig:cost-to-fix}).
+These studies are most often cited by secondary sources regarding the delayed issue effect.
+We note that it is unclear from the text in ~\cite{Daly77} and \cite{Boehm76} if cost is defined in terms of effort, or in actual cost (i.e., labor, materiel, travel, etc).
+
 These studies suggest that the difficulty (in terms of effort) to find and fix an error monotonically increases with lifecycle phase.
@@
 {\em any} of those results since the links in the references of
-that  survey are all broken (``page not foind'' errors).
+that  survey are all broken (``page not found'' errors).
 
@@
  issues. 
-Clutterbuck et al.~\cite{Clutterbuck09} studied 5-month effort by a small-to-medium enterprise team developing a 71KLOEC web interface to a database application to implement 18 change requests-- see Figure~\ref{fig:clutterbuck} (note that these were for new and changed user requirements, not defects).
+Clutterbuck et al.~\cite{Clutterbuck09} studied a 5-month effort by a small-to-medium enterprise team developing a 71KLOEC web interface to a database application to implement 18 change requests-- see Figure~\ref{fig:clutterbuck} (note that these were for new and changed user requirements, not defects).
 Clutterbuck et al.

### 2016-09-16 tex/fig13.tex
Updates from ShareLaTeX

@@
-\resizebox{\columnwidth}{!}{\begin{tabular}{c@{~}|l@{~}r@{~}|rrr|r@{~}l|r@{~}l}
-  & \multicolumn{2}{c|}{ } & \multicolumn{3}{c|}{Percentiles} & \multicolumn{4}{c}{Growth with respect to earliest phase  } \\ 
-   & \multicolumn{2}{c|}{Phase} & \multicolumn{3}{c|}{(units= time in miniutes)} & \multicolumn{4}{c}{(unitless, since these are ratios of two time values) } \\\hline
+%\resizebox{\columnwidth}{!}{
+\scriptsize
+\begin{tabular}{c|lr|rr|rl}
+& \multicolumn{2}{c|}{ } & \multicolumn{2}{c|}{Percentiles} & \multicolumn{2}{c}{~ } \\ 
+  & \multicolumn{2}{c|}{ } & \multicolumn{2}{c|}{(units = } & \multicolumn{2}{c}{Growth with respect to earliest phase  } \\ 
+   & \multicolumn{2}{c|}{Phase} & \multicolumn{2}{c|}{  minutes)} & \multicolumn{2}{c}{(unitless  ratios of two time values) } \\\hline
 
-  rank & injected & removed & 50th & 90th &  IQR & \multicolumn{2}{c|}{50th percentile growth} & \multicolumn{2}{c}{90th percentile growth} \\ 
-\hline\multicolumn{10}{c}{~}  \\
-1 &  Before   &   DesignInspect   & 10 & 32 & 14 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}} & 1.00 & \textcolor{Gray}{\rule{10mm}{2mm}}\\
-1 &   &   CodeReview      & 8 & 31 & 14 & 0.80 & \textcolor{black}{\rule{8mm}{2mm}} & 0.97 & \textcolor{Gray}{\rule{9mm}{2mm}}\\
-1 &   &   CodeInspect     & 10 & 32 & 16 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}} & 1.00 & \textcolor{Gray}{\rule{10mm}{2mm}}\\
-1 &   &   UnitTest        & 12 & 45 & 21 & 1.20 & \textcolor{black}{\rule{12mm}{2mm}} & 1.41 & \textcolor{Gray}{\rule{14mm}{2mm}}\\
-1 &   &   IntTest         & 15 & 63 & 31 & 1.50 & \textcolor{black}{\rule{15mm}{2mm}} & 1.97 & \textcolor{Gray}{\rule{19mm}{2mm}}\\
-1 &   &   SysTest         & 11 & 46 & 22 & 1.10 & \textcolor{black}{\rule{12mm}{2mm}} & 1.44 & \textcolor{Gray}{\rule{14mm}{2mm}}\\  
-\hline\multicolumn{10}{c}{~}  \\
-1 &  Planning     &   ReqtsReview     & 8 & 35 & 14 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}} & 1.00 & \textcolor{Gray}{\rule{10mm}{2mm}}\\
-1 &               &   DesignInspect   & 11 & 31 & 13 & 1.38 & \textcolor{black}{\rule{13mm}{2mm}} & 0.89 & \textcolor{Gray}{\rule{8mm}{2mm}}\\
-2 &               &   UnitTest        & 24 & 53 & 25 & 3.00 & \textcolor{black}{\rule{30mm}{2mm}} & 1.51 & \textcolor{Gray}{\rule{15mm}{2mm}}\\
-\hline\multicolumn{10}{c}{~}  \\
-1 &  Reqts   &   ReqtsReview     & 13 & 42 & 20 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}} & 1.00 & \textcolor{Gray}{\rule{10mm}{2mm}}\\
-1 &          &   ReqtsInspect    & 12 & 40 & 18 & 0.92 & \textcolor{black}{\rule{9mm}{2mm}} & 0.95 & \textcolor{Gray}{\rule{9mm}{2mm}}\\
-1 &          &   DesignReview    & 10 & 34 & 14 & 0.77 & \textcolor{black}{\rule{7mm}{2mm}} & 0.81 & \textcolor{Gray}{\rule{8mm}{2mm}}\\
-1 &          &   DesignInspect   & 9 & 38 & 15 & 0.69 & \textcolor{black}{\rule{6mm}{2mm}} & 0.90 & \textcolor{Gray}{\rule{9mm}{2mm}}\\
-1 &          &   CodeInspect     & 13 & 45 & 24 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}} & 1.07 & \textcolor{Gray}{\rule{10mm}{2mm}}\\
-1 &          &   UnitTest        & 10 & 40 & 17 & 0.77 & \textcolor{black}{\rule{7mm}{2mm}} & 0.95 & \textcolor{Gray}{\rule{9mm}{2mm}}\\
-1 &          &   IntTest         & 33 & 95 & 42 & 2.54 & \textcolor{black}{\rule{25mm}{2mm}} & 2.26 & \textcolor{Gray}{\rule{22mm}{2mm}}\\
-1 &          &   SysTest         & 24 & 126 & 108 & 1.85 & \textcolor{black}{\rule{18mm}{2mm}} & 3.00 & \textcolor{Gray}{\rule{30mm}{2mm}}\\
-\hline\multicolumn{10}{c}{~}  \\
-1 & Design   &   DesignReview    & 11 & 37 & 16 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}} & 1.00 & \textcolor{Gray}{\rule{10mm}{2mm}}\\
-1 &   &   DesignInspect   & 8 & 28 & 12 & 0.73 & \textcolor{black}{\rule{7mm}{2mm}} & 0.76 & \textcolor{Gray}{\rule{7mm}{2mm}}\\
-1 &   &   CodeReview      & 10 & 40 & 18 & 0.91 & \textcolor{black}{\rule{9mm}{2mm}} & 1.08 & \textcolor{Gray}{\rule{10mm}{2mm}}\\
-1 &   &   CodeInspect     & 9 & 33 & 14 & 0.82 & \textcolor{black}{\rule{8mm}{2mm}} & 0.89 & \textcolor{Gray}{\rule{8mm}{2mm}}\\
-1 &   &   UnitTest        & 11 & 41 & 18 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}} & 1.11 & \textcolor{Gray}{\rule{12mm}{2mm}}\\
-1 &   &   IntTest         & 17 & 75 & 31 & 1.55 & \textcolor{black}{\rule{15mm}{2mm}} & 2.03 & \textcolor{Gray}{\rule{20mm}{2mm}}\\
-1 &   &   AcceptTest      & 14 & 44 & 19 & 1.27 & \textcolor{black}{\rule{12mm}{2mm}} & 1.19 & \textcolor{Gray}{\rule{12mm}{2mm}}\\
-1 &   &   SysTest         & 13 & 40 & 18 & 1.18 & \textcolor{black}{\rule{12mm}{2mm}} & 1.08 & \textcolor{Gray}{\rule{10mm}{2mm}}\\
-\hline\multicolumn{10}{c}{~}  \\
-1 &  Code   &   CodeReview    & 10 & 35 & 16 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}} & 1.00 & \textcolor{Gray}{\rule{10mm}{2mm}}\\
-1 &    &   CodeInspect   & 10 & 32 & 15 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}} & 0.91 & \textcolor{Gray}{\rule{9mm}{2mm}}\\
-1 &   &   UnitTest      & 12 & 45 & 20 & 1.20 & \textcolor{black}{\rule{12mm}{2mm}} & 1.29 & \textcolor{Gray}{\rule{12mm}{2mm}}\\
-1 &    &   IntTest       & 14 & 58 & 25 & 1.40 & \textcolor{black}{\rule{14mm}{2mm}} & 1.66 & \textcolor{Gray}{\rule{16mm}{2mm}}\\
-1 &    &   AcceptTest    & 16 & 60 & 25 & 1.60 & \textcolor{black}{\rule{16mm}{2mm}} & 1.71 & \textcolor{Gray}{\rule{17mm}{2mm}}\\
-1 &   &   SysTest       & 13 & 47 & 20 & 1.30 & \textcolor{black}{\rule{13mm}{2mm}} & 1.34 & \textcolor{Gray}{\rule{13mm}{2mm}}\\
+  rank & injected & removed & 50th &   IQR & \multicolumn{2}{c}{50th percentile growth}  \\ 
+\hline
 
-\end{tabular}}
+\multicolumn{7}{c}{~}  \\
+1 &  Before   
+      &   DesignInspect   & 10 &  14 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}}  \\
+1 &   &   CodeReview      & 8 &  14 & 0.80 & \textcolor{black}{\rule{8mm}{2mm}}  \\
+1 &   &   CodeInspect     & 10 &  16 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}}  \\
+1 &   &   UnitTest        & 12 &   21 & 1.20 & \textcolor{black}{\rule{12mm}{2mm}}  \\
+1 &   &   IntTest         & 15 &   31 & 1.50 & \textcolor{black}{\rule{15mm}{2mm}}  \\
+1 &   &   SysTest         & 11 &   22 & 1.10 & \textcolor{black}{\rule{12mm}{2mm}}  \\  
+\hline\multicolumn{7}{c}{~}  \\
+1 &  Planning     &   ReqtsReview     & 8 &  14 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}}  \\
+1 &               &   DesignInspect   & 11 &  13 & 1.38 & \textcolor{black}{\rule{13mm}{2mm}} \\
+2 &               &   UnitTest        & 24 &   25 & 3.00 & \textcolor{black}{\rule{30mm}{2mm}} \\
+\hline\multicolumn{7}{c}{~}  \\
+1 &  Reqts   &   ReqtsReview     & 13 &  20 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}} \\
+1 &          &   ReqtsInspect    & 12 &   18 & 0.92 & \textcolor{black}{\rule{9mm}{2mm}}  \\
+1 &          &   DesignReview    & 10 &   14 & 0.77 & \textcolor{black}{\rule{7mm}{2mm}} \\
+1 &          &   DesignInspect   & 9 &   15 & 0.69 & \textcolor{black}{\rule{6mm}{2mm}}   \\
+1 &          &   CodeInspect     & 13 &  24 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}} \\
+1 &          &   UnitTest        & 10 &  17 & 0.77 & \textcolor{black}{\rule{7mm}{2mm}}  \\
+1 &          &   IntTest         & 33 &   42 & 2.54 & \textcolor{black}{\rule{25mm}{2mm}}  \\
+1 &          &   SysTest         & 24 &  108 & 1.85 & \textcolor{black}{\rule{18mm}{2mm}}  \\
+\hline\multicolumn{7}{c}{~}  \\
+1 & Design   &   DesignReview    & 11 &   16 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}}  \\
+1 &   &   DesignInspect   & 8 &   12 & 0.73 & \textcolor{black}{\rule{7mm}{2mm}}  \\
+1 &   &   CodeReview      & 10 &   18 & 0.91 & \textcolor{black}{\rule{9mm}{2mm}}  \\
+1 &   &   CodeInspect     & 9 &  14 & 0.82 & \textcolor{black}{\rule{8mm}{2mm}}  \\
+1 &   &   UnitTest        & 11 &   18 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}}  \\
+1 &   &   IntTest         & 17 &   31 & 1.55 & \textcolor{black}{\rule{15mm}{2mm}}  \\
+1 &   &   AcceptTest      & 14 &   19 & 1.27 & \textcolor{black}{\rule{12mm}{2mm}}  \\
+1 &   &   SysTest         & 13 &   18 & 1.18 & \textcolor{black}{\rule{12mm}{2mm}}  \\
+\hline\multicolumn{7}{c}{~}  \\
+1 &  Code   &   CodeReview    & 10 &  16 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}}  \\
+1 &    &   CodeInspect   & 10 &   15 & 1.00 & \textcolor{black}{\rule{10mm}{2mm}}  \\
+1 &   &   UnitTest      & 12 &   20 & 1.20 & \textcolor{black}{\rule{12mm}{2mm}}  \\
+1 &    &   IntTest       & 14 &   25 & 1.40 & \textcolor{black}{\rule{14mm}{2mm}}  \\
+1 &    &   AcceptTest    & 16 &   25 & 1.60 & \textcolor{black}{\rule{16mm}{2mm}}  \\
+1 &   &   SysTest       & 13 &   20 & 1.30 & \textcolor{black}{\rule{13mm}{2mm}}  \\
+
+\end{tabular}
+%}

### 2016-09-16 data.tex

@@
 One the guiding principles of TSP was that  people performing the work are  responsible for planning and tracking the work. That is,  all the data collected here was entered
-by local developers. This data was then check by local {\em coaches} before being sent to the SEI
+by local developers. This data was then checked by local coaches before being sent to the SEI
 databases. Coaches are certified by demonstrating competent use of the TSP process with the artifacts and data.
@@
 One special feature of  \fig{waterfall} is
-the {\em before} where the   TSP Launch Preparation scripts and checklists (see TSP: Leading a development Team~\cite{Humphrey:2005}) require the TSP team to assure that management and business have clearly identified cost, schedule, and scope goals appropriate to the upcoming development activities, often including a conceptual model.
+the {\em before} phase, in which the TSP team assures that management has clearly identified cost, schedule, and scope goals appropriate to the upcoming development activities, often including a conceptual model ~\cite{Humphrey:2005}.
 For example an architecture team must have sufficient requirements to reason about, prototype, and specify an architecture~\cite{Bachmann13} while a coding only team within a larger project would have more precisely defined requirements and high level design.
@@
 In TSP  reviews,  individuals perform personal reviews of their work products prior to the peer review
-(which TSP calls the inspection(.
+(which TSP calls the inspection).
 Also, Figure~\ref{fig:waterfall} divides  testing     as follows.
 Developers perform unit test prior to code complete.
-After code complete a standard phase is the integration, which combines program units into workable system ready for system test.
+After code complete a standard phase is integration, which combines program units into a workable system ready for system test.
 Integration,  system test, and acceptance test are often performed by another group.

### 2016-09-16 phaseDelay.tex

@@
 Such a large and growing effort should be managed and optimized via  well-researched conclusions.  
-
-It is standard practice
+To assist in achieving this, there has been a growing recognition within the software engineering research community 
+of the importance of theory building [Sjoberg08, Paivarinta15, Stol15]. A good theory allows empirical research to go beyond simply
+reporting observations and instead provides explanations for why results are observed [Stol15]. Theory-building needs to be an iterative process,
+in which results from practice are used to refine theories and theories are used to inform future observation and data collection [Paivarinta15, Stol15]. 
+It is no coincidence that it  
+is standard practice
 in other fields, such as medicine,
-to continually revisit old conclusions~\cite{prasad13}.
+to continually revisit old conclusions in the light of new theories and new understanding~\cite{prasad13}.
+
 Accordingly, this paper revisits
-the widely-held belief of a {\em delayed issue effect} (hereafter, DIE).
+a fundamental software engineering theory, the {\em delayed issue effect} (hereafter, DIE).
 Later in this paper, we offer a precise definition for this effect.
@@
 
- Basili \& Boehm comment that since the 1980s,  this effect
+This theory has been used as the basis for decision-making in software engineering. For example, Basili \& Boehm comment that since the 1980s,  this effect
  \begin{quote}
@@
  %One hypothesized explanation that can be ruled out is that the lack of a phase delay effect is due to using the TSP method. 
-Our results beg the question: why did an idea, with so little support, become so widespread in the software engineering literature?
+Our results raise the question: why did an idea, with so little support, become so widespread in the software engineering literature?
 No doubt the original evidence was compelling at the time, but much has changed in the realm of software development in the subsequent 40 years.

### 2016-09-16 tex/responses.tex

@@
 those comments we have extensively revised and extended this draft.
-Specifically, the old version was 20 pages and this version is 27 pages (and the new
+Specifically, the old version was 20 pages and this version is 29 pages (and the new
 material all comes from the clarification and correction points offered by the reviewers).
@@
 
-\begin{quote}{\em This new draft now maps our work into the SE literature on theory building. This new material appears in the introduction as well as being
-peppered throughout the document (in numerous places).}\end{quote}
+\begin{quote}{\em This new draft now maps our work into the SE literature on theory building. This new material appears in the introduction as well as throughout the document.}\end{quote}
 
@@
 
-\begin{quote}{\em We now map our work to prior publications on theory-building in software engineering.
+\begin{quote}{\em We now map our work to prior publications on theory-building in software engineering, as well as theory falsification in general.
 That mapping appears in numerous comments through out the text (using the references offered by this reviewer) with the bulk of the comments being top of p2 and the conclusion.
@@
 Our goal here is to motivate that we should re-investigate the DIE.
-We have toned down the language and combined our arguments on ``widely-believed'' and ``poorly-documented'' into a new \tion{belief} whose focus is to motivate our study.}\end{quote}
+We have toned down the language and combined our arguments on ``widely-believed'' and ``poorly-documented'' into a new \tion{belief} whose focus is to motivate our study.
+We have added more details on our literature search and snowball sampling.
+We have also highlighted how blog articles often cite secondary sources, which in turn cite Software Engineering Economics, to illustrate how the DIE pehnemonon has perpetuated.}\end{quote}
 
@@
 
-\begin{quote}{\em An astute observation. We provide more information on the types of projects included in our sample in \tion{data_character}
+\begin{quote}{\em We provide more information on the types of projects included in our sample in \tion{data_character}
 
@@
 \begin{quote}{\em We understand why you are making that suggestion.
-That figure is something of a classic. It is appropriate we chop it like that?}.
+That figure is something of a classic. Therefore we opted not to chop the figure but hope the text is sufficiently clear}.
 \end{quote}
@@
 \begin{quote}{\em
-Thanks for those pointers. We will fix.
+Thanks for those pointers. We have fixed the issue.
  }\end{quote}
@@
 {\em Authors' preference. When a document has many tables and figures, navigating
-between them is hard. Calling them all "figures" makes navigation faster. 
-But do you want us to change that?}
+between them is hard. Calling them all "figures" makes navigation faster. }
 \end{quote}
